### 2017-05-22 PISA Writeup.Rmd
All files in the repo now

---
title: "PISA Writeup"
author: "Jorge Cimentada"
date: "3/14/2017"
output: html_document
---

```{r}
knitr::opts_chunk$set(
  warning = T,
  message = T
)
```

```{r, cache = T}
library(haven)
library(PISA2000lite)
library(PISA2003lite)
library(PISA2006lite)
library(PISA2009lite)
library(PISA2012lite)
library(intsvy)
library(cimentadaj) # For pisa_countrynames
library(countrycode) # For region variable
library(tidyverse)
library(car)
library(readr)
library(SAScii)
library(ggthemes)
library(lme4)

# For PISA_2015
pisa2015_conf <- list(variables = list(pvlabelpref = "PV",
                                       pvlabelsuff = "READ",
                                       weightFinal = "W_FSTUWT",
                                       weightBRR = "W_FSTURWT"),
                      parameters = list(cutoffs = c(357.77, 420.07, 482.38, 544.68, 606.99, 669.30),
                                        percentiles = c(5, 10, 25, 75, 90, 95),
                                        PVreps = 10,
                                        BRRreps = 80,
                                        weights = "BRR",
                                        replication_scheme = 'pisa')
)

quantiles <- c(10, 50, 90)
pisa_merged <- read_rds("./data/pisa_list2.Rdata")
```

The work of Sean Reardon has caused a great deal of interest in me. It talks about the evolution of cognitive inequalities in the United States. Interestingly, he explores patterns of the cognitive gap between rich and poor children and finds that the gap has been increasing since we have data on it. Although it could be argued that this pattern is an artifact of aggregating statistics, or that the pattern is the product of other things, such as income inequality or the racial gap, he shows convincing evidence in favour that dismisses both explanations.

In an elegant way, Reardon shows how the racial gap has been diminishing since the 1940's (although not at a great pace; other research shows that this gap has decreased little to nothing) while the income cognitive gap has been increasing. Without going into technicalities, I was fascinated with his work. I thought that more research was needed on this topic, specially at a greater scale and not U.S centric.

I had the idea of looking for a similar pattern in the PISA datasets. I know beforehand that this is not at all very similar to Reardon's work for two reasons. First, Reardon has data starting from 1943 to about 2010. The PISA data has only 5 time points, starting in the year 2000 and ending in 2015, every 3 years. Second, Reardon's aggregate pattern includes children from all ages, while the PISA only concentrates on 15 year olds. Despite this, I though it might be an interesting idea to explore this.

The achievement gap can be measured in very different ways. First, it could be measured simply as the difference in test scores between children coming from parents with ISCED 5-6 (college or above) and children coming from parents with ISCED 1-2 (less than high school). Van der Werfhorst and colleagues
<!-- Cite Van der Werfhorst correctly -->
also discuss other ways, such as the gap between the 90th and 10th percentile in the distribution of cognitive skills. In order to stay close to the design of Reardon, I calculate the difference between the 90th and 10th percentile, as well as the 80th and 20th percentile. I also compare these results to the gap of children from high educated parents and children from low educated parents.

***

After discussion of 90/10/80/10/high/low educated gaps.

I then proceeded to disagreggate the pattern even more. Because the no pattern was evidence from above, I updated my beliefs and though that if any pattern was to be evidence, it should in percentiles within class groups. So, for example, I'd like to compare the evolution of the gap of the 90th percentile from the higher educated against the 90th percentile of the lower educated. Furthermore, I'd like to compare the same strategy for:

Name    High educated - Low educated
            90              90
            90              10
            10              90
            50              50
            90              50
                            50/10
<!-- Create table with all computed variables -->

These groups compose a pattern of extremes, specially the type of groups that are interesting for policy making, because they are those which need the biggest help.

```{r}
unnest_rename <- function(df, col_unnest, var_rename) {
  unnested <-
    df %>%
    unnest_(col_unnest)

  if(sum(names(unnested) %in% var_rename) > 1) {
    warning(paste(as.name(var_rename), "not a unique name in unnested df. Renaming the first instance"))
  }

  unnested %>%
    rename_(se_mean = var_rename)
}

pisa_ci <- function(df, col_unnest, var_rename, mean_var) {
  dots <- setNames(list(
    lazyeval::interp(~ x - y, x = as.name(mean_var), y = quote(se_twice)),
    lazyeval::interp(~ x + y, x = as.name(mean_var), y = quote(se_twice))
  ), c("low_ci", "upper_ci"))

  unnest_rename(pisa_merged, col_unnest, var_rename) %>%
    mutate(
      se_twice = 2 * se_mean,
      region = countrycode(country, "country.name", "region"),
      continent = countrycode(country, "country.name", "continent")
    ) %>%
    mutate_(.dots = dots) %>%
    filter(!is.na(continent))
}

pisa_quant_ci <-
  pisa_ci(pisa_merged, "quantile_mean", "`Std. err.`", "Score")

pisa_gaps <-
  pisa_quant_ci %>%
  unite(edu_perc, high_edu, Percentiles) %>%
  select(year:Score) %>%
  spread(edu_perc, Score) %>%
  mutate(`srich/spoor` = `3_90` - `1_90`,
         `srich/dpoor` = `3_90` - `1_10`,
         `drich/spoor` = `3_10` - `1_90`,
         `srich/drich` = `3_90` - `3_10`,
         `avgrich/avgpoor` = `3_50` - `1_50`,
         `srich/avgpoor` = `3_90` - `1_50`,
         `avgpoor/dpoor` = `1_50`- `1_10`) %>%
  select(year, country, contains("rich"), contains("poor")) %>%
  gather(diff, value, -year, -country)

gap_df <-
  pisa_gaps %>%
  mutate(region = countrycode(country, "country.name", "region"),
         continent = countrycode(country, "country.name", "continent"))

gap_groups <-
  c("50_poor/10_poor",
    "90_rich/50_poor",
    "10_rich/90_poor",
    "90_rich/50_poor",
    "90_rich/10_poor",
    "90_rich/10_rich",
    "90_rich/90_poor")

gap_df %>%
  group_by(year, region, diff) %>%
  summarize(avg = mean(value, na.rm = T)) %>%
  ggplot(aes(x = year, avg, group = diff, colour = diff)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ region) +
  labs(x = "Year", y = "Average difference") +
  scale_colour_discrete(name = "Gap groups", labels = gap_groups) +
  theme_minimal()
```
One pattern is clearly evident: with the exception of Western Asia, all regions have a plane line. For all of these categories not a lot of changes have ocurred in the last 15 years. There are other interesting patterns such as the gap between the the 50th percentile of the lower educated and the 10th percentile of the lower educated is very similar to the gap between the 90th percentile of the higher educated and the 50th percentile of the lower educated. In a similar vein, the gap between the 90th percentile of both high and low educated are very similar as well. We shouldn't allow this to cause a big surprise because the absolution position of each group in the 'ranking' of scores is probably very different, despite the gaps being very similar.

Even if we take this to the continent level, we find that not that many patterns emerge. One slight exception is the Americas (mostly driven by South America, if we look at the previous plot). Taking a closer look it seems that most gaps have a negative slope, specially the 90th percentile from high educated and the 10th percentile of the low educated. However, because we're not including any confidence intervals we're not sure on how uncertain these estimates are.
```{r}
gap_df %>%
  group_by(year, continent, diff) %>%
  summarize(avg = mean(value, na.rm = T)) %>%
  ggplot(aes(x = year, avg, group = diff, colour = diff)) +
  geom_line() +
  geom_point() +
  facet_wrap(~ continent) +
  labs(x = "Year", y = "Average difference") +
  scale_colour_discrete(name = "Gap groups", labels = gap_groups) +
  theme_minimal()
```

```{r}

gap_df %>%
  filter(diff == "drich/spoor") %>%
  split(.$year, drop = T) %>%
  map(~ tidy(lmer(value ~ 1 + (1 | region), data = .x))) %>%
  map(~ .x[grep("sd", .x$term), ]) %>%
  enframe %>%
  unnest(value) %>%
  ggplot(aes(name, estimate, fill = term)) + geom_col(position = "dodge")

```

<!-- Do so multilevel exploration -->
<!-- Adjust for family activities such as reading, etc -->
<!-- Adjust for gender -->
<!-- Calculate the gap as percentages. So the gap of this is 50% the gap of the other -->
<!-- Add confidence intervals -->
<!-- Take reardon's strategy for prediction in his appendix -->

### 2017-05-22 Datasets_PISA.R

@@ -1,0 +1,73 @@
+if (!require(devtools)) {
+    install.packages("devtools")
+    require(devtools)
+}
+# install_github("pbiecek/PISA2012lite")
+# load it
+
+library("PISA2012lite")
+library(Hmisc)
+cntrylist <- list(student=student2012, parent=parent2012, school=school2012, cognitive=scoredItem2012)
+
+means <- by(student2012[, c("PV1MATH", "W_FSTUWT")], student2012$CNT,
+            function(x) wtd.quantile(x[,1], x[,2], probs = c(0.25,0.75)))
+
+meansx <- by(student2012[, c("PV1MATH", "W_FSTUWT")], student2012$CNT,
+            function(x) wtd.mean(x[,1], x[,2]))
+
+library(car)
+library(arm)
+library(purrr)
+library(dplyr)
+library(tidyr)
+
+student2012$ST17Q01 <- as.numeric(student2012$ST17Q01)
+student2012$ST17Q01 <- recode(student2012$ST17Q01,"7:8 = NA; 1:2 = 1")
+student2012$ST17Q01 <- as.factor(student2012$ST17Q01)
+
+
+r.sq2 <- by(student2012[, c("PV1MATH", "ST17Q01")], student2012$CNT,
+   function(x) summary(lm(PV1MATH ~ ST17Q01, x))$adj.r.squared)
+
+
+means2 <- as.data.frame(do.call(rbind, means))
+means2$cntr <- row.names(means2); means2 <- means2[,c("cntr","25%","75%")]
+row.names(means2) <- 1:nrow(means2)
+
+means2 <- cbind(means2, mean=as.numeric(meansx), r.sq=as.numeric(r.sq2))
+means2$Diff <- means2[,"75%"] - means2[,"25%"]
+
+means2 <- means2[order(means2$Diff), ]
+
+
+means3 <- means2$Diff; names(means3) <- row.names(means2)
+dotchart(means3, pch = 19)
+
+means4 <- means2$r.sq; names(means4) <- means2$cntr;
+dotchart(sort(means4), pch=19)
+
+
+
+ggplot(means2, aes(mean, Diff, color=r.sq)) + geom_smooth(colour="blue",method="lm", se=F) +
+    geom_point(size=4, alpha=0.5) +
+    geom_smooth(colour="red",se=F) +
+    scale_color_continuous(low="red",high = "green") +
+    annotate("text", x=375, y=175, label=paste("R of",
[…]

### 2017-05-22 inequality_explr.R

@@ -1,0 +1,426 @@
+library(saves)
+library(haven)
+library(PISA2000lite)
+library(PISA2003lite)
+library(PISA2006lite)
+library(PISA2009lite)
+library(PISA2012lite)
+library(intsvy)
+library(cimentadaj) # For pisa_countrynames
+library(countrycode) # For region variable
+library(tidyverse)
+library(car)
+library(readr)
+library(SAScii)
+
+# Download PISA 2015 data
+temp_dir <- tempdir()
+url <- "http://vs-web-fs-1.oecd.org/pisa/PUF_SPSS_COMBINED_CMB_STU_QQQ.zip"
+dir_pisa <- file.path(temp_dir, "pisa.zip")
+download.file(url, destfile = dir_pisa)
+
+unzip(dir_pisa, exdir = temp_dir)
+
+pisa_2015 <- read_spss(file.path(temp_dir, "CY6_MS_CMB_STU_QQQ.sav"))
+
+# Download PISA 2012 files
+url <- "http://www.oecd.org/pisa/pisaproducts/INT_STU12_DEC03.zip"
+pisa2012_syntax <- "http://www.oecd.org/pisa/pisaproducts/PISA2012_SAS_student.sas"
+
+# Download the SAS syntax to determine column width
+download.file(pisa2012_syntax, file.path(temp_dir, "PISA2012_SAS_student.sas"))
+dic_student <- parse.SAScii(sas_ri = file.path(temp_dir, 'PISA2012_SAS_student.sas'))
+
+# Download PISA 2012 actual data
+dir_pisa <- file.path(temp_dir, "pisa2012.zip")
+download.file(url, destfile = dir_pisa)
+unzip(dir_pisa, exdir = temp_dir)
+
+# Read the data using the fixed width from the SAS syntax
+student2012 <- read_fwf(file = file.path(temp_dir, 'INT_STU12_DEC03.txt'),
+                        col_positions = fwf_widths(dic_student$width), progress = T)
+colnames(student2012) <- dic_student$varname
+
+# Remove temporary files
+file.remove(file.path(temp_dir, list.files(temp_dir)))
+
+student2015 <- pisa_2015 # toy dataset
+
+# # If PISA2012 starts working, then use this code:
+# # Vector with database names except student2015
+# databases <- c("math2000", paste0("student", seq(2003, 2012, 3)))
+#
+# # Create a list with all data sets
+# pisa_list <- map(c(databases, "student2015"), get)
+#
[…]

### 2017-05-22 new_inequality.R

@@ -1,0 +1,574 @@
+library(arm)
+library(saves)
+library(haven)
+library(PISA2000lite)
+library(PISA2003lite)
+library(PISA2006lite)
+library(PISA2009lite)
+library(PISA2012lite)
+library(intsvy)
+library(cimentadaj) # For pisa_countrynames
+library(countrycode) # For region variable
+library(tidyverse)
+library(car)
+library(readr)
+library(SAScii)
+library(inequalityintsvy)
+library(lme4)
+library(modelr)
+
+# For PISA_2015
+pisa2015_conf <- list(variables = list(pvlabelpref = "PV",
+                                       pvlabelsuff = "READ",
+                                       weightFinal = "W_FSTUWT",
+                                       weightBRR = "W_FSTURWT"),
+                      parameters = list(cutoffs = c(357.77, 420.07, 482.38, 544.68, 606.99, 669.30),
+                                        percentiles = c(5, 10, 25, 75, 90, 95),
+                                        PVreps = 10,
+                                        BRRreps = 80,
+                                        weights = "BRR",
+                                        replication_scheme = 'pisa')
+)
+
+pisa_all <- read_rds("./data/pisa_listcol.Rdata")
+pisa_all2 <- pisa_all
+
+db <- paste0("pisa", seq(2000, 2015, 3))
+pisa_all2$value <- map2(pisa_all2$value, db, ~ { .x$wave <- .y; .x})
+pisa_all2$value[[1]]$CNT <- pisa_all2$value[[1]]$COUNTRY
+
+pisa_all2$value <- map(pisa_all2$value, ~ {
+  .x$father_edu <- car::recode(as.numeric(.x$FISCED), "8:9 = NA")
+  .x$mother_edu <- car::recode(as.numeric(.x$MISCED), "8:9 = NA")
+  .x$high_edu_broad <- pmax(.x$father_edu, .x$mother_edu)
+  .x$country <- pisa_countrynames[.x$CNT]
+
+  if (any(unique(.x$wave) %in% c("pisa2012", "pisa2015"))) {
+    .x$father_edu <- .x$father_edu + 1
+    .x$mother_edu <- .x$mother_edu + 1
+    .x$high_edu_broad <- .x$high_edu_broad + 1
+  }
+  .x
+})
+
[…]

### 2017-05-22 new_inequality_two.R

@@ -1,0 +1,540 @@
+  library(arm)
+  library(saves)
+  library(haven)
+  library(PISA2000lite)
+  library(PISA2003lite)
+  library(PISA2006lite)
+  library(PISA2009lite)
+  library(PISA2012lite)
+  library(intsvy)
+  library(cimentadaj) # For pisa_countrynames
+  library(countrycode) # For region variable
+  library(tidyverse)
+  library(car)
+  library(readr)
+  library(SAScii)
+  library(inequalityintsvy)
+  library(lme4)
+  library(modelr)
+  
+  # Conf for PISA_2015
+  pisa2015_conf <- list(variables = list(pvlabelpref = "PV",
+                                         pvlabelsuff = "READ",
+                                         weightFinal = "W_FSTUWT",
+                                         weightBRR = "W_FSTURWT"),
+                        parameters = list(cutoffs = c(357.77, 420.07, 482.38, 544.68, 606.99, 669.30),
+                                          percentiles = c(5, 10, 25, 75, 90, 95),
+                                          PVreps = 10,
+                                          BRRreps = 80,
+                                          weights = "BRR",
+                                          replication_scheme = 'pisa')
+  )
+  
+  pisa_all <- read_rds("./data/pisa_listcol.Rdata")
+  pisa_all2 <- pisa_all
+  
+  years <- seq(2000, 2015, 3)
+  countries <- c("Germany", "United States", "Denmark", "Sweden", "United Kingdom",
+                 "Spain", "Italy", "Canada", "Australia")
+  
+  db <- paste0("pisa", years)
+  pisa_all2$value <- map2(pisa_all2$value, db, ~ { .x$wave <- .y; .x})
+  pisa_all2$value[[1]]$CNT <- pisa_all2$value[[1]]$COUNTRY
+  
+  pisa_all2$value <- map(pisa_all2$value, ~ {
+    .x$father_edu <- car::recode(as.numeric(.x$FISCED), "8:9 = NA")
+    .x$mother_edu <- car::recode(as.numeric(.x$MISCED), "8:9 = NA")
+    .x$high_edu_broad <- pmax(.x$father_edu, .x$mother_edu)
+    .x$country <- pisa_countrynames[as.character(.x$CNT)]
+  
+    if (any(unique(.x$wave) %in% c("pisa2012", "pisa2015"))) {
+      .x$father_edu <- .x$father_edu + 1
+      .x$mother_edu <- .x$mother_edu + 1
[…]

### 2017-05-22 inequality_explr.R
Recoded 9997 to NA in all waves when calc dv

@@ -14,101 +14,101 @@ library(readr)
 library(SAScii)
 
 # Download PISA 2015 data
-temp_dir <- tempdir()
-url <- "http://vs-web-fs-1.oecd.org/pisa/PUF_SPSS_COMBINED_CMB_STU_QQQ.zip"
-dir_pisa <- file.path(temp_dir, "pisa.zip")
-download.file(url, destfile = dir_pisa)
-
-unzip(dir_pisa, exdir = temp_dir)
-
-pisa_2015 <- read_spss(file.path(temp_dir, "CY6_MS_CMB_STU_QQQ.sav"))
-
-# Download PISA 2012 files
-url <- "http://www.oecd.org/pisa/pisaproducts/INT_STU12_DEC03.zip"
-pisa2012_syntax <- "http://www.oecd.org/pisa/pisaproducts/PISA2012_SAS_student.sas"
-
-# Download the SAS syntax to determine column width
-download.file(pisa2012_syntax, file.path(temp_dir, "PISA2012_SAS_student.sas"))
-dic_student <- parse.SAScii(sas_ri = file.path(temp_dir, 'PISA2012_SAS_student.sas'))
-
-# Download PISA 2012 actual data
-dir_pisa <- file.path(temp_dir, "pisa2012.zip")
-download.file(url, destfile = dir_pisa)
-unzip(dir_pisa, exdir = temp_dir)
-
-# Read the data using the fixed width from the SAS syntax
-student2012 <- read_fwf(file = file.path(temp_dir, 'INT_STU12_DEC03.txt'),
-                        col_positions = fwf_widths(dic_student$width), progress = T)
-colnames(student2012) <- dic_student$varname
-
-# Remove temporary files
-file.remove(file.path(temp_dir, list.files(temp_dir)))
-
-student2015 <- pisa_2015 # toy dataset
-
-# # If PISA2012 starts working, then use this code:
+# temp_dir <- tempdir()
+# url <- "http://vs-web-fs-1.oecd.org/pisa/PUF_SPSS_COMBINED_CMB_STU_QQQ.zip"
+# dir_pisa <- file.path(temp_dir, "pisa.zip")
+# download.file(url, destfile = dir_pisa)
+# 
+# unzip(dir_pisa, exdir = temp_dir)
+# 
+# pisa_2015 <- read_spss(file.path(temp_dir, "CY6_MS_CMB_STU_QQQ.sav"))
+# 
+# # Download PISA 2012 files
[…]

### 2017-05-22 new_inequality_two.R

@@ -364,9 +364,13 @@
                AGE)
       
       message(paste(unique(.x$wave), "data ready"))
+
+
+      test_vars <- paste0("PV", seq_len(conf$parameters$PVreps), test)
+      .x[test_vars] <- map(.x[test_vars], ~ ifelse(.x == 9997, NA, .x))
       
       # Calculate median math score of all PV's
-      .x$dv <- apply(.x[paste0("PV", seq_len(conf$parameters$PVreps), test)], 1, median, na.rm = T)
+      .x$dv <- apply(.x[test_vars], 1, median, na.rm = T)
       
       # Should I estimate the model separately by country?
       mod1 <- lm(dv ~ AGE,

### 2017-05-22 new_inequality_two.R
didn't change anything - everything tested to work

@@ -432,7 +432,6 @@
   
   # Graph the gap by using both the top and bottom instead of the absolute diff
   reduced_data %>%
-    # filter(country %in% countries) %>%
     ggplot(aes(as.character(wave), Mean, group = escs_dummy, colour = escs_dummy)) +
     geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
     geom_line() +
@@ -465,7 +464,7 @@
            lower = Mean - 1.96 * se_diff,
            upper = Mean + 1.96 * se_diff)
   
-  walk(countries, ~ {
+  map(countries, ~ {
     reduced %>%
     filter(!is.na(continent), !is.na(region), country == .x) %>%
     ggplot(aes(as.factor(wave), Mean, group = country, colour = country)) +

### 2017-05-23 new_inequality_two.R
Comments on country trends available

@@ -417,11 +417,11 @@
 }
 
   results_math <- test_diff(pisa_all2$value, reliability_pisa, "MATH")
-  results_read <- test_diff(pisa_all2$value, reliability_pisa, "READ")
+  # results_read <- test_diff(pisa_all2$value, reliability_pisa, "READ")
   # US is missing for reading
   
   reduced_data <-
-    map2(results_read, years, function(.x, .y) {
+    map2(results_math, years, function(.x, .y) {
       .x %>%
         mutate(wave = .y) %>%
         filter(!is.na(escs_dummy))
@@ -464,7 +464,7 @@
            lower = Mean - 1.96 * se_diff,
            upper = Mean + 1.96 * se_diff)
   
-  map(countries, ~ {
+  walk(countries, ~ {
     reduced %>%
     filter(!is.na(continent), !is.na(region), country == .x) %>%
     ggplot(aes(as.factor(wave), Mean, group = country, colour = country)) +
@@ -480,30 +480,30 @@
     ggsave(paste0(.x, "_read.png"), path = "./plots/")
   })
   
-# Increase:  
-# Sweden - Increase in both tests
-# Austria - Increase for math, not sure for reading (2006 is also an outlier, just as in the U.S)
-# Canada - Slight increase in math and static in reading
-# Finland - Sharp increase in both tests but year 2006 needs to be checked out
-# France - Sharpe increase in both, although 2006 problem.
-# Japan - Increase in both tests, although no CI. Whats up?
-# Netherlands - Increase in both tests (2006 problem)
-# Spain - Slight increase in both
-  
-  
-# Decrease:
-# US - slight decrease in math and more pronounced decrease in reading (checkout year 2006, drastic outlier)
-# Chile - slight decrease in math also unstable for reading (2006 as well)
 
+# Increase:  
+# Sweden - steady increase in both tests
+# Austria - increase in math - slight increase in read
+# Finland - very sharp increase in both
+# France - very sharp increase in both
+# Netherlands - sharp increase in both
+
+# Decrease:
+# US - decrease in both tests
+# Chile - decrease in both tests
+  
 # No change:
-# UK - no change in both tests (slight decrease in reading)
-# Belgium - no change but checkout year 2006 for reading (Also belgium and US)
-# Czcheck Republic - unstable
-# Germany - No change although 2006 problem
-# Italy - no change
-# Denmark - slight increase in math but very unstable due to 2006
-# Norway - No change
-# Poland - very unstable
[…]

### 2017-05-23 new_inequality_two.R
Introd error calculating quantiles - fixed

@@ -299,11 +299,11 @@
     
     quan_lower <- try(Hmisc::wtd.quantile(df_lower$escs_trend,
                                           weights = df_lower[[weights]],
-                                          probs = c(0.90)))
+                                          probs = c(0.10)))
     
     quan_upper <- try(Hmisc::wtd.quantile(df_upper$escs_trend,
                                     weights = df_upper[[weights]],
-                                    probs = c(0.10)))
+                                    probs = c(0.90)))
     
     if (any("try-error" %in% c(class(quan_lower), class(quan_upper)))) {
       return(c(NA, NA))
@@ -475,9 +475,9 @@
     # geom_smooth(aes(group = continent), colour = "blue", method = "lm") +
     coord_cartesian(ylim = c(0, 2.5)) +
     facet_wrap(~ country) +
-    ggtitle(paste0(.x, "_read"))
+    ggtitle(paste0(.x, "_math"))
     
-    ggsave(paste0(.x, "_read.png"), path = "./plots/")
+    ggsave(paste0(.x, "_math.png"), path = "./plots/")
   })
   
 

### 2017-05-23 new_inequality_two.R
Revert "Introd error calculating quantiles - fixed"
This reverts commit db5a8c80fa8e4f7ecead1057bd1a517f9e37f7d6.

@@ -299,11 +299,11 @@
     
     quan_lower <- try(Hmisc::wtd.quantile(df_lower$escs_trend,
                                           weights = df_lower[[weights]],
-                                          probs = c(0.10)))
+                                          probs = c(0.90)))
     
     quan_upper <- try(Hmisc::wtd.quantile(df_upper$escs_trend,
                                     weights = df_upper[[weights]],
-                                    probs = c(0.90)))
+                                    probs = c(0.10)))
     
     if (any("try-error" %in% c(class(quan_lower), class(quan_upper)))) {
       return(c(NA, NA))
@@ -475,9 +475,9 @@
     # geom_smooth(aes(group = continent), colour = "blue", method = "lm") +
     coord_cartesian(ylim = c(0, 2.5)) +
     facet_wrap(~ country) +
-    ggtitle(paste0(.x, "_math"))
+    ggtitle(paste0(.x, "_read"))
     
-    ggsave(paste0(.x, "_math.png"), path = "./plots/")
+    ggsave(paste0(.x, "_read.png"), path = "./plots/")
   })
   
 

### 2017-05-23 new_inequality_two.R
Changed quantiles to be top/bottom 30%

@@ -299,11 +299,11 @@
     
     quan_lower <- try(Hmisc::wtd.quantile(df_lower$escs_trend,
                                           weights = df_lower[[weights]],
-                                          probs = c(0.90)))
+                                          probs = c(0.30)))
     
     quan_upper <- try(Hmisc::wtd.quantile(df_upper$escs_trend,
                                     weights = df_upper[[weights]],
-                                    probs = c(0.10)))
+                                    probs = c(0.70)))
     
     if (any("try-error" %in% c(class(quan_lower), class(quan_upper)))) {
       return(c(NA, NA))
@@ -312,7 +312,7 @@
     }
   }
   
-  pisa_all2$value <-
+  adapted_year_data <-
     map(pisa_all2$value, ~ {
       if (unique(.x$wave) == "pisa2000") {
         # pisa2000 has a different coding so here I recode 6 to 7 so that in all waves the top edu
@@ -321,14 +321,13 @@
           mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `6` = 7)))
       } else {
         .x <-
-          mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `2` = 1)))
+          mutate(.x, new_hisced = as.character(high_edu_broad))
       }
       .x
   })
   
   # Producing the plot to get the difference between the top 5% of the high educated
   # vs the bottom 5% of the low educated.
-  
   test_diff <- function(df, reliability, test) {
   
     map2(df, reliability, function(.x, .y) {
@@ -416,8 +415,8 @@
     })
 }
 
-  results_math <- test_diff(pisa_all2$value, reliability_pisa, "MATH")
-  # results_read <- test_diff(pisa_all2$value, reliability_pisa, "READ")
+  results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH")
+  # results_read <- test_diff(adapted_year_data, reliability_pisa, "READ")
   # US is missing for reading
   
   reduced_data <-
@@ -432,6 +431,7 @@
   
   # Graph the gap by using both the top and bottom instead of the absolute diff
   reduced_data %>%
+    filter(country %in% c("United States")) %>%
     ggplot(aes(as.character(wave), Mean, group = escs_dummy, colour = escs_dummy)) +
[…]

### 2017-05-23 new_inequality_two.R
Included to-do list comments

@@ -451,9 +451,7 @@
   countries <- c("Chile", "Austria", "Belgium", "Canada", "Czech Republic", "Denmark",
                  "Finland", "France", "Germany", "Italy", "Japan", "Netherlands", "Norway",
                  "Poland", "Spain", "Sweden", "United Kingdom", "United States")
-  
-  countries <- c("United States", "United Kingdom")
-  
+
   reduced_data %>%
     select(country, escs_dummy, Mean, wave) %>%
     spread(key = escs_dummy, value = Mean) %>%
@@ -474,6 +472,23 @@
     coord_cartesian(ylim = c(0, 4)) +
     facet_wrap(~ country) +
     ggtitle("math")
+  
+# Next steps:
+# Making country groups of the graphs showing that some countries are increasing, others are steady
+# While others are decreasing. Do this with a few countries, but end by showing all countries.
+
+# Continue by studying what's happening in those countries where it's increasing/decreasing
+# In the gap graph
+  
+# Continue by doing the multilevel models to see what explains what. Include
+# all indicators from the reardon/russian girl paper.
+  
+# Graph the increase in each country vs the increase/decrease of the economic inequality indicators
+# Specially the 90/10
+
+# Calculate how big is the gap between reading and math
+  
+# Continue with the PIRLS to see if there are specific patterns in 4th and 8th graders gap.
   
   walk(countries, ~ {
     reduced %>%

### 2017-06-01 new_inequality_two.R
Added to-do comments

@@ -490,6 +490,10 @@
   
 # Continue with the PIRLS to see if there are specific patterns in 4th and 8th graders gap.
   
+# Get each country trendline adjusted for the inequality indicators and place in the same country graph.
+
+# Should I add the parent's education in the lm model to see how trends change adjusted for that?
+  
   walk(countries, ~ {
     reduced %>%
     filter(!is.na(continent), !is.na(region), country == .x) %>%

### 2017-06-01 paper/paper.Rnw
Created paper folder with beginning of paper

\documentclass[11pt, a4paper]{article}
\bibliographystyle{apalike}
\pagestyle{headings}

\usepackage{pdflscape}
\usepackage{subcaption}
\usepackage{graphics}
\usepackage{graphicx}
\usepackage[round, colon]{natbib}
\usepackage[colorlinks]{hyperref}
\AtBeginDocument{%
  \hypersetup{
    citecolor=blue,
    linkcolor=blue,   
    urlcolor=blue}}

\title{First paper - Draft}
\author{Jorge Cimentada}

\begin{document}
\setlength{\parindent}{2em}
\setlength{\parskip}{1em}
\showboxdepth=5
\showboxbreadth=5

\maketitle

\tableofcontents

\section{Analysis}

<<packages_conf, echo = F>>=
  library(arm)
  library(saves)
  library(haven)
  library(PISA2000lite)
  library(PISA2003lite)
  library(PISA2006lite)
  library(PISA2009lite)
  library(PISA2012lite)
  library(intsvy)
  library(cimentadaj) # For pisa_countrynames
  library(countrycode) # For region variable
  library(tidyverse)
  library(car)
  library(readr)
  library(SAScii)
  library(inequalityintsvy)
  library(lme4)
  library(modelr)

  # Conf for PISA_2015
  pisa2015_conf <- list(variables = list(pvlabelpref = "PV",
                                         pvlabelsuff = "READ",
                                         weightFinal = "W_FSTUWT",
                                         weightBRR = "W_FSTURWT"),
                        parameters = list(cutoffs = c(357.77, 420.07, 482.38, 544.68, 606.99, 669.30),
                                          percentiles = c(5, 10, 25, 75, 90, 95),
                                          PVreps = 10,
                                          BRRreps = 80,
                                          weights = "BRR",
                                          replication_scheme = 'pisa')
  )

@

<<>>=
pisa_all <- read_rds("./data/pisa_listcol.Rdata")
pisa_all2 <- pisa_all

years <- seq(2000, 2015, 3)
countries <- c("Germany", "United States", "Denmark", "Sweden", "United Kingdom",
                 "Spain", "Italy", "Canada", "Australia")

db <- paste0("pisa", years)
pisa_all2$value <- map2(pisa_all2$value, db, ~ { .x$wave <- .y; .x})
pisa_all2$value[[1]]$CNT <- pisa_all2$value[[1]]$COUNTRY

pisa_all2$value <- map(pisa_all2$value, ~ {
  .x$father_edu <- car::recode(as.numeric(.x$FISCED), "8:9 = NA")
  .x$mother_edu <- car::recode(as.numeric(.x$MISCED), "8:9 = NA")
  .x$high_edu_broad <- pmax(.x$father_edu, .x$mother_edu)
  .x$country <- pisa_countrynames[as.character(.x$CNT)]

  if (any(unique(.x$wave) %in% c("pisa2012", "pisa2015"))) {
    .x$father_edu <- .x$father_edu + 1
    .x$mother_edu <- .x$mother_edu + 1
    .x$high_edu_broad <- .x$high_edu_broad + 1
  }
  .x
})

reliability_pisa <-
  c("2000" = 0.81,
    "2003" = 0.85,
    "2006" = 0.78,
    "2009" = 0.74,
    "2012" = 0.82,
    "2015" = 0.74) # 2015 imputed

@

\end{document}

### 2017-06-01 transform_data.R
Moved data creation to separate file

@@ -1,0 +1,103 @@
+library(saves)
+library(haven)
+library(PISA2000lite)
+library(PISA2003lite)
+library(PISA2006lite)
+library(PISA2009lite)
+library(PISA2012lite)
+library(intsvy)
+library(cimentadaj) # For pisa_countrynames
+library(countrycode) # For region variable
+library(tidyverse)
+library(car)
+library(readr)
+library(SAScii)
+
+# Download PISA 2015 data
+
+temp_dir <- tempdir()
+url <- "http://vs-web-fs-1.oecd.org/pisa/PUF_SPSS_COMBINED_CMB_STU_QQQ.zip"
+dir_pisa <- file.path(temp_dir, "pisa.zip")
+download.file(url, destfile = dir_pisa)
+
+unzip(dir_pisa, exdir = temp_dir)
+
+pisa_2015 <- read_spss(file.path(temp_dir, "CY6_MS_CMB_STU_QQQ.sav"))
+
+# Download PISA 2012 files
+url <- "http://www.oecd.org/pisa/pisaproducts/INT_STU12_DEC03.zip"
+pisa2012_syntax <- "http://www.oecd.org/pisa/pisaproducts/PISA2012_SAS_student.sas"
+
+# Download the SAS syntax to determine column width
+download.file(pisa2012_syntax, file.path(temp_dir, "PISA2012_SAS_student.sas"))
+dic_student <- parse.SAScii(sas_ri = file.path(temp_dir, 'PISA2012_SAS_student.sas'))
+
+# Download PISA 2012 actual data
+dir_pisa <- file.path(temp_dir, "pisa2012.zip")
+download.file(url, destfile = dir_pisa)
+unzip(dir_pisa, exdir = temp_dir)
+
+# Read the data using the fixed width from the SAS syntax
+student2012 <- read_fwf(file = file.path(temp_dir, 'INT_STU12_DEC03.txt'),
+                        col_positions = fwf_widths(dic_student$width), progress = T)
+colnames(student2012) <- dic_student$varname
+
+# Remove temporary files
+file.remove(file.path(temp_dir, list.files(temp_dir)))
+
+student2015 <- pisa_2015 # toy dataset
+
+# Provisional code for separate PISA2012
+# Vector with database names except student2015
+databases <- c("math2000", paste0("student", seq(2003, 2009, 3)))
+
+# Create a list with all data sets
+pisa_list <- map(c(databases, paste0("student", c(2012, 2015))), get)
+
+# Give each dataset their own name
[…]

### 2017-06-01 inequality_explr.R
Adapted both scrip to contain the data transformation script at first

@@ -1,114 +1,13 @@
 library(saves)
 library(haven)
-library(PISA2000lite)
-library(PISA2003lite)
-library(PISA2006lite)
-library(PISA2009lite)
-library(PISA2012lite)
 library(intsvy)
 library(cimentadaj) # For pisa_countrynames
 library(countrycode) # For region variable
 library(tidyverse)
 library(car)
 library(readr)
-library(SAScii)
 
-# Download PISA 2015 data
-# temp_dir <- tempdir()
-# url <- "http://vs-web-fs-1.oecd.org/pisa/PUF_SPSS_COMBINED_CMB_STU_QQQ.zip"
-# dir_pisa <- file.path(temp_dir, "pisa.zip")
-# download.file(url, destfile = dir_pisa)
-# 
-# unzip(dir_pisa, exdir = temp_dir)
-# 
-# pisa_2015 <- read_spss(file.path(temp_dir, "CY6_MS_CMB_STU_QQQ.sav"))
-# 
-# # Download PISA 2012 files
-# url <- "http://www.oecd.org/pisa/pisaproducts/INT_STU12_DEC03.zip"
-# pisa2012_syntax <- "http://www.oecd.org/pisa/pisaproducts/PISA2012_SAS_student.sas"
-# 
-# # Download the SAS syntax to determine column width
-# download.file(pisa2012_syntax, file.path(temp_dir, "PISA2012_SAS_student.sas"))
-# dic_student <- parse.SAScii(sas_ri = file.path(temp_dir, 'PISA2012_SAS_student.sas'))
-# 
-# # Download PISA 2012 actual data
-# dir_pisa <- file.path(temp_dir, "pisa2012.zip")
-# download.file(url, destfile = dir_pisa)
-# unzip(dir_pisa, exdir = temp_dir)
-# 
-# # Read the data using the fixed width from the SAS syntax
-# student2012 <- read_fwf(file = file.path(temp_dir, 'INT_STU12_DEC03.txt'),
-#                         col_positions = fwf_widths(dic_student$width), progress = T)
-# colnames(student2012) <- dic_student$varname
-# 
-# # Remove temporary files
-# file.remove(file.path(temp_dir, list.files(temp_dir)))
-# 
-# student2015 <- pisa_2015 # toy dataset
-# 
-# # # If PISA2012 starts working, then use this code:
-# # # Vector with database names except student2015
-# # databases <- c("math2000", paste0("student", seq(2003, 2012, 3)))
[…]

### 2017-06-01 new_inequality_two.R

@@ -17,6 +17,8 @@
   library(lme4)
   library(modelr)
   
+  source("./transform_data.R")
+  
   # Conf for PISA_2015
   pisa2015_conf <- list(variables = list(pvlabelpref = "PV",
                                          pvlabelsuff = "READ",
@@ -237,57 +239,57 @@
   #     results
   #   })
   
-  df_lower <-
-    pisa_all2$value[[1]] %>%
-    mutate(new_hisced = as.character(dplyr::recode(as.numeric(HISCED),
-                                                   `0` = 1))) %>%
-    filter(country == "United States", new_hisced == 1)
-  
-  df_upper <- 
-    pisa_all2$value[[6]] %>%
-    mutate(new_hisced = as.character(dplyr::recode(as.numeric(HISCED),
-                                                   `0` = 1))) %>%
-    filter(country == "United States", new_hisced == 6)
-  
-  qp_lower <- Hmisc::wtd.quantile(df_lower$hisei,
-                      weights = df_lower[["spftw0"]],
-                      probs = c(0.10))
-  
-  qp_upper <- Hmisc::wtd.quantile(df_upper$hisei,
-                                  weights = df_upper[["spftw0"]],
-                                  probs = c(0.95))
-  
-  df %>%
-    ggplot(aes(hisei, colour = as.character(new_hisced))) +
-    geom_freqpoly() +
-    geom_vline(xintercept = c(qp_lower, qp_upper)) +
-    scale_x_continuous(breaks = seq(min(df$hisei, na.rm = T),
-                                    max(df$hisei, na.rm = T), 2))
-  
-  df %>%
-    mutate(really_high_really_low = case_when(.$hisei >= qp_upper[1] ~ "1",
-                                              .$hisei <= qp_lower[1] ~ "0")) %>%
-    count(country, really_high_really_low) %>%
-    filter(!is.na(really_high_really_low)) %>%
-    arrange(n)
-  
-  qp <- Hmisc::wtd.quantile(df$escs_trend,
-                            weights = df[["spftw0"]],
-                            probs = c(0.05, 0.95))
-  
-  df %>%
-    ggplot(aes(escs_trend, colour = new_hisced)) +
-    geom_freqpoly() +
-    geom_vline(xintercept = qp) +
-    scale_x_continuous(breaks = seq(min(pisa_all2$value[[6]]$escs_trend, na.rm = T),
[…]

### 2017-06-01 paper/paper.Rnw
Moving script to paper - modelling part done

@@
 
+  # source("./transform_data.R")
+
   # Conf for PISA_2015
@@
 
-<<>>=
+<<loading_data/recoding, echo = F, cache = TRUE>>=
 pisa_all <- read_rds("./data/pisa_listcol.Rdata")
@@
 
+<<escs_trend, echo = F, cache = TRUE>>=
+  # Rescaled trend ESCS data to merge.
+  # This only has data for seq(2000, 2012, 3) because
+  # PISA 2015 has the ESCS trend variable.
+  dir <- tempdir()
+  file_name <- "escs_trend.zip"
+  download.file("http://vs-web-fs-1.oecd.org/pisa/trend_escs_SPSS.zip",
+                destfile = file.path(dir, file_name))
+  unzip(file.path(dir, file_name), exdir = dir)
+  escs_trend <- map(file.path(dir, list.files(dir, pattern = ".sav")), haven::read_spss)
+  file.remove(file.path(dir, list.files(dir)))
+
+  escs_trend <-
+    map(escs_trend, ~ {
+    mutate(.x, cnt = pisa_countrynames[cnt]) %>%
+    rename(country = cnt)
+  })
+
+@
+
+<<merge_escs_pisa, echo = F, cache = TRUE>>=
+   # Next we'll merge the ESCS data with the PISA data. As explained above, the 6th data (PISA
+  # 2015) doesn't need to be merged so I exclude it with this vector
+  exclude <- -6
+
+  # Loop in parallel to the PISA data, the ESCS data and the year vector (which is seq(2012, 2015, 3))
+  pisa_all2$value[exclude] <-
+    pmap(list(pisa_all2$value[exclude], escs_trend, years[exclude]), function(.x, .y, .z) {
+
+    # The escs data needs to have the key variables the same class as the
+    # same data.
+    escs <-
+      .y %>% mutate(schoolid = as.numeric(schoolid),
+                    stidstd = as.numeric(stidstd))
+
+    # .z is the corresponding year that will be created as a column
+    # And perform the same transformation of the key variables as in the ESCS data
+    data_trend <-
+      .x %>%
+        mutate(
+          year = .z,
+          schoolid = as.numeric(as.character(SCHOOLID)),
+          stidstd = as.numeric(as.character(STIDSTD))
+          ) %>%
+   left_join(escs,
+              by = c("country", "schoolid", "stidstd"))
+
+    message(paste(unique(.x$wave), "done"))
+
+    data_trend
+  })
+
+  pisa_all2$value[[6]] <-
+    pisa_all2$value[[6]] %>%
+    rename(escs_trend = ESCS)
+@
+
+<<functions_for_modelling, echo = F, cache = TRUE>>=
+
+# Function calculates the bottom 30th quantile for the bottom educated and the 70th quantile
+# for the top educated. If the quantiles can't be estimated, it returns two NA's instead
+quantile_missing <- function(df, weights) {
+
+    df_lower <- filter(df, new_hisced == 1)
+    df_upper <- filter(df, new_hisced == 7)
+
+    quan_lower <- try(Hmisc::wtd.quantile(df_lower$escs_trend,
+                                          weights = df_lower[[weights]],
+                                          probs = c(0.30)))
+
+    quan_upper <- try(Hmisc::wtd.quantile(df_upper$escs_trend,
+                                    weights = df_upper[[weights]],
+                                    probs = c(0.70)))
+
+    if (any("try-error" %in% c(class(quan_lower), class(quan_upper)))) {
+      return(c(NA, NA))
+      } else {
+     return(c(quan_lower[1], quan_upper[1]))
+    }
+}
+
+# Producing the plot to get the difference between the top 30% of the high educated
+# vs the bottom 30% of the low educated. This function loops through each dataset/country
+# and survey reliability and estimates the difference while also calculating the
+# joint standard error of the difference.
+
+# It returns a dataframe for each survey with all countries and respective coefficients and
+# standard errors.
+test_diff <- function(df, reliability, test) {
+
+    map2(df, reliability, function(.x, .y) {
+
+      conf <- if (unique(.x$wave) == "pisa2015") pisa2015_conf else pisa_conf
+      weights_var <- conf$variables$weightFinal
+
+      country_split <- split(.x, .x$country)
+
+      country_list <- map(country_split, function(country) {
+        print(unique(country$country))
+
+        # In some countries the quan can't be estimated because of very few obs.
+        # The function doesn't stop but returns two NA's.
+        quan <- quantile_missing(country, weights_var)
+
+        # It's very important to create a variable that returns the number of observations of this dummy
+        # For each country. Possibly to weight by the number of observations.
+        country$escs_dummy <-
+          with(country, case_when(escs_trend >= quan[2] ~ 1,
+                                  escs_trend <= quan[1] ~ 0))
+        country
+      })
+
+      .x <-
+        enframe(country_list) %>%
+        unnest(value)
+
+      .x <-
+        .x %>%
+        select(wave, matches(paste0("^PV.*", test, "$")), escs_dummy,
+               country, one_of(weights_var),
+               AGE)
+
+      message(paste(unique(.x$wave), "data ready"))
+
+      test_vars <- paste0("PV", seq_len(conf$parameters$PVreps), test)
+      .x[test_vars] <- map(.x[test_vars], ~ ifelse(.x == 9997, NA, .x))
+
+      # Calculate median math score of all PV's
+      .x$dv <- apply(.x[test_vars], 1, median, na.rm = T)
+
+      # Should I estimate the model separately by country?
+      mod1 <- lm(dv ~ AGE,
+                 weights = .x[[weights_var]],
+                 data = .x,
+                 na.action = "na.exclude")
+
+      # Take residuals of model and divide by rmse. Multiply that by
+      # 1 / sqrt(reliability of each survey), which is .y in the loop.
+      .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)
+
+      mod2 <-
+        lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
+             data = .x,
+             weights = .x[[weights_var]])
+
+      # Take the country coefficients (absolute coefficients)
+      country_coef <-
+        coef(mod2)$country %>%
+        rownames_to_column() %>%
+        gather(escs_dummy, Mean, -rowname) %>%
+        mutate(escs_dummy = dplyr::recode(escs_dummy,
+                                          `(Intercept)` = "0",
+                                          `escs_dummy` = "1"))
+
+      # Take the absolute country standard errors
+      se <-
+        se.coef(mod2)$country %>%
+        as.data.frame() %>%
+        rownames_to_column() %>%
+        gather(escs_dummy, s.e., -rowname) %>%
+        mutate(escs_dummy = dplyr::recode(escs_dummy,
+                                          `(Intercept)` = "0",
+                                          `escs_dummy` = "1"))
+
+      results <-
+        inner_join(country_coef, se, by = c("rowname", "escs_dummy")) %>%
+        rename(country = rowname) %>%
+        arrange(country, escs_dummy)
+
+      message(paste0(unique(.x$wave), " modeling done"))
+      results
+    })
+}
+@
+
+<<modelling, echo = F, cache = T>>=
+adapted_year_data <-
+    map(pisa_all2$value, ~ {
+      if (unique(.x$wave) == "pisa2000") {
+        # pisa2000 has a different coding so here I recode 6 to 7 so that in all waves the top edu
+        # is 7 and the bottom is 1
+        .x <- 
+          mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `6` = 7)))
+      } else {
+        .x <-
+          mutate(.x, new_hisced = as.character(high_edu_broad))
+      }
+      .x
+})
+
+results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH")
+results_read <- test_diff(adapted_year_data, reliability_pisa, "READ")
+# US is missing for reading
+
+@
 \end{document}

### 2017-06-01 paper/paper.Rnw
Finished moving script

@@
 
-<<functions_for_modelling, echo = F, cache = TRUE>>=
+<<functions_for_modelling, echo = F>>=
 
@@
 # vs the bottom 30% of the low educated. This function loops through each dataset/country
-# and survey reliability and estimates the difference while also calculating the
-# joint standard error of the difference.
+# and survey reliability and estimates the difference while also extracting the s.e. of each
+# difference.
 
@@
 # US is missing for reading
+@
+
+<<ci_for_difference, echo = F>>=
+reduced_data <-
+    map2(results_math, years, function(.x, .y) {
+      .x %>%
+        mutate(wave = .y) %>%
+        filter(!is.na(escs_dummy))
+    }) %>%
+    bind_rows() %>%
+    mutate(lower = Mean - 1.96 * s.e.,
+           upper = Mean + 1.96 * s.e.)
+@
+
+<<graphin differences>>=
+  reduced_data %>%
+    filter(country %in% c("United States")) %>%
+    ggplot(aes(as.character(wave), Mean, group = escs_dummy, colour = escs_dummy)) +
+    geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
+    geom_line() +
+    facet_wrap(~ country)
+@
+
+<<>>=
+# Calculate the joint standard error of the different
+  se_data <-
+    reduced_data %>%
+    select(country, escs_dummy, wave, s.e.) %>%
+    spread(escs_dummy, s.e.) %>%
+    transmute(country, wave,
+              se_diff = sqrt(abs(`1`^2 - `0`^2)))
 
 @
+
+<<>>=
+
+# Calculate the different between the gap and together with it's joint s.e graph
+# the absolut difference.
+
+countries <- c("Chile", "Austria", "Belgium", "Canada", "Czech Republic", "Denmark",
+                 "Finland", "France", "Germany", "Italy", "Japan", "Netherlands", "Norway",
+                 "Poland", "Spain", "Sweden", "United Kingdom", "United States")
+
+reduced_data %>%
+    select(country, escs_dummy, Mean, wave) %>%
+    spread(key = escs_dummy, value = Mean) %>%
+    mutate(diff = `1` - `0`) %>%
+    gather(escs_dummy, Mean, -country, -wave, -`1`, -`0`) %>%
+    left_join(se_data) %>%
+    mutate(continent = countrycode(country, "country.name", "continent"),
+           region = countrycode(country, "country.name", "region"),
+           lower = Mean - 1.96 * se_diff,
+           upper = Mean + 1.96 * se_diff) %>%
+    filter(!is.na(continent), !is.na(region), country %in% countries) %>%
+    ggplot(aes(as.factor(wave), Mean, group = country, colour = country)) +
+    geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
+    geom_line() +
+    geom_point(size = 0.5) +
+    scale_colour_discrete(guide = F) +
+    # geom_smooth(aes(group = region), colour = "blue", method = "lm") +
+    coord_cartesian(ylim = c(0, 4)) +
+    facet_wrap(~ country) +
+    ggtitle("math")
+
+# Increase:  
+# Sweden - steady increase in both tests
+# Austria - increase in math - slight increase in read
+# Finland - very sharp increase in both
+# France - very sharp increase in both
+# Netherlands - sharp increase in both
+
+# Decrease:
+# US - decrease in both tests
+# Chile - decrease in both tests
+
+# No change:
+# Canada - stable red - increase math
+# UK - slight decrease red - stable math
+# Belgium - no change
+# Czech republic - no change
+# Denmark no change
+# Germany - no change
+# ITaly - no change
+# Japan -  no change
+# Norway - no change
+# Poland - no change
+# Spain - no change
+
+@
+
+<<>>=
+# Next steps:
+
+# Show country patterns with reading and math together into one graph
+
+# Making country groups of the graphs showing that some countries are increasing, others are steady
+# While others are decreasing. Do this with a few countries, but end by showing all countries.
+
+# Continue by studying what's happening in those countries where it's increasing/decreasing
+# In the gap graph
+
+# Continue by doing the multilevel models to see what explains what. Include
+# all indicators from the reardon/russian girl paper.
+
+# Graph the increase in each country vs the increase/decrease of the economic inequality indicators
+# Specially the 90/10
+
+# Calculate how big is the gap between reading and math
+
+# Continue with the PIRLS to see if there are specific patterns in 4th and 8th graders gap.
+
+# Get each country trendline adjusted for the inequality indicators and place in the same country graph.
+
+# Should I add the parent's education in the lm model to see how trends change adjusted for that?
+
+@
+
+Try one!
 \end{document}

### 2017-06-01 new_inequality_two.R

@@ -17,7 +17,7 @@
   library(lme4)
   library(modelr)
   
-  source("./transform_data.R")
+# source("./transform_data.R")
   
   # Conf for PISA_2015
   pisa2015_conf <- list(variables = list(pvlabelpref = "PV",

### 2017-06-01 inequality_explr.R
Forgot to include data_transform in the preamble of script

@@ -7,7 +7,7 @@ library(tidyverse)
 library(car)
 library(readr)
 
-source("./transform_data.R")
+# source("./transform_data.R")
 
 # For PISA_2015
 pisa2015_conf <- list(variables = list(pvlabelpref = "PV",

### 2017-06-02 paper/paper.Rnw
Ran paper with code - all working

@@
 
-\tableofcontents
+<<working directory, echo = F>>=
+opts_knit$set(root.dir = '..')
 
-\section{Analysis}
+opts_chunk$set(echo = F,
+               message = F,
+               warning = F,
+               include = F,
+               cache.lazy = F,
+               results = 'asis')
+@
 
 <<packages_conf, echo = F>>=
+  library(knitr)
   library(arm)
@@
   library(countrycode) # For region variable
-  library(tidyverse)
   library(car)
@@
   library(modelr)
+  library(tidyverse)
 
@@
 
-<<loading_data/recoding, echo = F, cache = TRUE>>=
+\tableofcontents
+
+\section{Analysis}
+
+<<loading_data-recoding, cache = TRUE>>=
 pisa_all <- read_rds("./data/pisa_listcol.Rdata")
@@
 
-<<escs_trend, echo = F, cache = TRUE>>=
+<<escs_trend, cache = TRUE>>=
   # Rescaled trend ESCS data to merge.
@@
 
-<<merge_escs_pisa, echo = F, cache = TRUE>>=
+<<merge_escs_pisa, cache = TRUE>>=
    # Next we'll merge the ESCS data with the PISA data. As explained above, the 6th data (PISA
@@
 
-<<functions_for_modelling, echo = F>>=
+<<functions_for_modelling>>=
 
@@
 
-<<modelling, echo = F, cache = T>>=
+<<modeling, cache = T>>=
 adapted_year_data <-
@@
 
-<<ci_for_difference, echo = F>>=
+<<ci_for_difference>>=
 reduced_data <-
@@
 
-<<graphin differences>>=
+<<graphin differences, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
   reduced_data %>%
@@
 
-<<>>=
+<<include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
 
@@
 
+# The gap between the top 90th and the 50th, did it grow?
+# The gap between the top 50th and the 10th, did it grow?
+
+# A weak welfare system, together with income inequality, what's their pattern?
+# What if we put the school differentiation/tracking aspect in? Are there country groups based on these
+# patterns.
+
+# Show that within top and bottom educated categories, there is reasonable variation of SES index (to justify that your approach of suing the top and bottom percentiles within each category is the right approach)
+
+# In countries where there is high differentiation/tracking, is there a jump in the evolution of the gap between PIRLS/TIMSS and PISA?
+
 @
 
-Try one!
+Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years.
+The vast literature on educational inequality has mainly concentrated on the United States (reardon here).
+First, because it is the only country where cognitive testing is very widespread across surveys.
+And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day.
+Using this information, Reardon (2011) is the first to investigate the evolution of the cognitive gap and the results are very surprising.
+Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap).
+
+The widening of the achievement gap has been happening in parallel to the growth of income inequality.
+Although very suggestive, it is hard to link both things causally.
+The downside of this research study is that it is mainly US-based, and not a lot of evidence is available in an international context.
+To the best of my knowledge, the only attempt to take this to an international context is the work by Chmielewski and Reardon (2016).
+They find that there is considerable variation in the achievement gap between top and bottom earning families.
+However, their design is quite different from the one from Reardon (2011).
+Reardon (2011) takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds.
+Although the author adjusts for age, the trends can't be generalized to age-specific, but rather in overall terms.} and makes them comparable across time.
+That is, he builds a trend of achievement gap.
+Chmielewski and Reardon (2016) use the international surveys PISA and PIRLS to compare the achievement gap across countries.
+Unfortunately, very few countries can be compared across time in PISA.
+
+This paper introduces one novelty in the literature which is to evaluate the evolution of the High/Low SES achievement gap in the past 15 years for all PISA participant countries.
+This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for each country.
+
+Do we need to distinguish between age-group achievement gaps?
+
 \end{document}

### 2017-06-02 paper/paper.tex
Ran paper with code - all working

--- 2017-06-01 paper/paper.Rnw
@@
-\documentclass[11pt, a4paper]{article}
+\documentclass[11pt, a4paper]{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+\let\hlipl\hlkwb
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \bibliographystyle{apalike}
@@
 \author{Jorge Cimentada}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 
-<<packages_conf, echo = F>>=
-  library(arm)
-  library(saves)
-  library(haven)
-  library(PISA2000lite)
-  library(PISA2003lite)
-  library(PISA2006lite)
-  library(PISA2009lite)
-  library(PISA2012lite)
-  library(intsvy)
-  library(cimentadaj) # For pisa_countrynames
-  library(countrycode) # For region variable
-  library(tidyverse)
-  library(car)
-  library(readr)
-  library(SAScii)
-  library(inequalityintsvy)
-  library(lme4)
-  library(modelr)
+{\centering \includegraphics[width=5.5in,height=5in]{figure/graphin_differences-1}
 
-  # source("./transform_data.R")
-
-  # Conf for PISA_2015
-  pisa2015_conf <- list(variables = list(pvlabelpref = "PV",
-                                         pvlabelsuff = "READ",
-                                         weightFinal = "W_FSTUWT",
-                                         weightBRR = "W_FSTURWT"),
-                        parameters = list(cutoffs = c(357.77, 420.07, 482.38, 544.68, 606.99, 669.30),
-                                          percentiles = c(5, 10, 25, 75, 90, 95),
-                                          PVreps = 10,
-                                          BRRreps = 80,
-                                          weights = "BRR",
-                                          replication_scheme = 'pisa')
-  )
-
-@
-
-<<loading_data/recoding, echo = F, cache = TRUE>>=
-pisa_all <- read_rds("./data/pisa_listcol.Rdata")
-pisa_all2 <- pisa_all
-
-years <- seq(2000, 2015, 3)
-countries <- c("Germany", "United States", "Denmark", "Sweden", "United Kingdom",
-                 "Spain", "Italy", "Canada", "Australia")
-
-db <- paste0("pisa", years)
-pisa_all2$value <- map2(pisa_all2$value, db, ~ { .x$wave <- .y; .x})
-pisa_all2$value[[1]]$CNT <- pisa_all2$value[[1]]$COUNTRY
-
-pisa_all2$value <- map(pisa_all2$value, ~ {
-  .x$father_edu <- car::recode(as.numeric(.x$FISCED), "8:9 = NA")
-  .x$mother_edu <- car::recode(as.numeric(.x$MISCED), "8:9 = NA")
-  .x$high_edu_broad <- pmax(.x$father_edu, .x$mother_edu)
-  .x$country <- pisa_countrynames[as.character(.x$CNT)]
-
-  if (any(unique(.x$wave) %in% c("pisa2012", "pisa2015"))) {
-    .x$father_edu <- .x$father_edu + 1
-    .x$mother_edu <- .x$mother_edu + 1
-    .x$high_edu_broad <- .x$high_edu_broad + 1
-  }
-  .x
-})
-
-reliability_pisa <-
-  c("2000" = 0.81,
-    "2003" = 0.85,
-    "2006" = 0.78,
-    "2009" = 0.74,
-    "2012" = 0.82,
-    "2015" = 0.74) # 2015 imputed
-
-@
-
-<<escs_trend, echo = F, cache = TRUE>>=
-  # Rescaled trend ESCS data to merge.
-  # This only has data for seq(2000, 2012, 3) because
-  # PISA 2015 has the ESCS trend variable.
-  dir <- tempdir()
-  file_name <- "escs_trend.zip"
-  download.file("http://vs-web-fs-1.oecd.org/pisa/trend_escs_SPSS.zip",
-                destfile = file.path(dir, file_name))
-  unzip(file.path(dir, file_name), exdir = dir)
-  escs_trend <- map(file.path(dir, list.files(dir, pattern = ".sav")), haven::read_spss)
-  file.remove(file.path(dir, list.files(dir)))
-
-  escs_trend <-
-    map(escs_trend, ~ {
-    mutate(.x, cnt = pisa_countrynames[cnt]) %>%
-    rename(country = cnt)
-  })
-
-@
-
-<<merge_escs_pisa, echo = F, cache = TRUE>>=
-   # Next we'll merge the ESCS data with the PISA data. As explained above, the 6th data (PISA
-  # 2015) doesn't need to be merged so I exclude it with this vector
-  exclude <- -6
-
-  # Loop in parallel to the PISA data, the ESCS data and the year vector (which is seq(2012, 2015, 3))
-  pisa_all2$value[exclude] <-
-    pmap(list(pisa_all2$value[exclude], escs_trend, years[exclude]), function(.x, .y, .z) {
-
-    # The escs data needs to have the key variables the same class as the
-    # same data.
-    escs <-
-      .y %>% mutate(schoolid = as.numeric(schoolid),
-                    stidstd = as.numeric(stidstd))
-
-    # .z is the corresponding year that will be created as a column
-    # And perform the same transformation of the key variables as in the ESCS data
-    data_trend <-
-      .x %>%
-        mutate(
-          year = .z,
-          schoolid = as.numeric(as.character(SCHOOLID)),
-          stidstd = as.numeric(as.character(STIDSTD))
-          ) %>%
-   left_join(escs,
-              by = c("country", "schoolid", "stidstd"))
-
-    message(paste(unique(.x$wave), "done"))
-
-    data_trend
-  })
-
-  pisa_all2$value[[6]] <-
-    pisa_all2$value[[6]] %>%
-    rename(escs_trend = ESCS)
-@
-
-<<functions_for_modelling, echo = F, cache = TRUE>>=
-
-# Function calculates the bottom 30th quantile for the bottom educated and the 70th quantile
-# for the top educated. If the quantiles can't be estimated, it returns two NA's instead
-quantile_missing <- function(df, weights) {
-
-    df_lower <- filter(df, new_hisced == 1)
-    df_upper <- filter(df, new_hisced == 7)
-
-    quan_lower <- try(Hmisc::wtd.quantile(df_lower$escs_trend,
-                                          weights = df_lower[[weights]],
-                                          probs = c(0.30)))
-
-    quan_upper <- try(Hmisc::wtd.quantile(df_upper$escs_trend,
-                                    weights = df_upper[[weights]],
-                                    probs = c(0.70)))
-
-    if (any("try-error" %in% c(class(quan_lower), class(quan_upper)))) {
-      return(c(NA, NA))
-      } else {
-     return(c(quan_lower[1], quan_upper[1]))
-    }
 }
 
-# Producing the plot to get the difference between the top 30% of the high educated
-# vs the bottom 30% of the low educated. This function loops through each dataset/country
-# and survey reliability and estimates the difference while also calculating the
-# joint standard error of the difference.
+{\centering \includegraphics[width=5.5in,height=5in]{figure/unnamed-chunk-2-1}
 
-# It returns a dataframe for each survey with all countries and respective coefficients and
-# standard errors.
-test_diff <- function(df, reliability, test) {
+}
 
-    map2(df, reliability, function(.x, .y) {
+Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years.
+The vast literature on educational inequality has mainly concentrated on the United States (reardon here).
+First, because it is the only country where cognitive testing is very widespread across surveys.
+And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day.
+Using this information, Reardon (2011) is the first to investigate the evolution of the cognitive gap and the results are very surprising.
+Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap).
 
-      conf <- if (unique(.x$wave) == "pisa2015") pisa2015_conf else pisa_conf
-      weights_var <- conf$variables$weightFinal
+The widening of the achievement gap has been happening in parallel to the growth of income inequality.
+Although very suggestive, it is hard to link both things causally.
+The downside of this research study is that it is mainly US-based, and not a lot of evidence is available in an international context.
+To the best of my knowledge, the only attempt to take this to an international context is the work by Chmielewski and Reardon (2016).
+They find that there is considerable variation in the achievement gap between top and bottom earning families.
+However, their design is quite different from the one from Reardon (2011).
+Reardon (2011) takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds.
+Although the author adjusts for age, the trends can't be generalized to age-specific, but rather in overall terms.} and makes them comparable across time.
+That is, he builds a trend of achievement gap.
+Chmielewski and Reardon (2016) use the international surveys PISA and PIRLS to compare the achievement gap across countries.
+Unfortunately, very few countries can be compared across time in PISA.
 
-      country_split <- split(.x, .x$country)
+This paper introduces one novelty in the literature which is to evaluate the evolution of the High/Low SES achievement gap in the past 15 years for all PISA participant countries.
+This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for each country.
 
-      country_list <- map(country_split, function(country) {
-        print(unique(country$country))
+Do we need to distinguish between age-group achievement gaps?
 
-        # In some countries the quan can't be estimated because of very few obs.
-        # The function doesn't stop but returns two NA's.
-        quan <- quantile_missing(country, weights_var)
-
-        # It's very important to create a variable that returns the number of observations of this dummy
-        # For each country. Possibly to weight by the number of observations.
-        country$escs_dummy <-
-          with(country, case_when(escs_trend >= quan[2] ~ 1,
-                                  escs_trend <= quan[1] ~ 0))
-        country
-      })
-
-      .x <-
-        enframe(country_list) %>%
-        unnest(value)
-
-      .x <-
-        .x %>%
-        select(wave, matches(paste0("^PV.*", test, "$")), escs_dummy,
-               country, one_of(weights_var),
-               AGE)
-
-      message(paste(unique(.x$wave), "data ready"))
-
-      test_vars <- paste0("PV", seq_len(conf$parameters$PVreps), test)
-      .x[test_vars] <- map(.x[test_vars], ~ ifelse(.x == 9997, NA, .x))
-
-      # Calculate median math score of all PV's
-      .x$dv <- apply(.x[test_vars], 1, median, na.rm = T)
-
-      # Should I estimate the model separately by country?
-      mod1 <- lm(dv ~ AGE,
-                 weights = .x[[weights_var]],
-                 data = .x,
-                 na.action = "na.exclude")
-
-      # Take residuals of model and divide by rmse. Multiply that by
-      # 1 / sqrt(reliability of each survey), which is .y in the loop.
-      .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)
-
-      mod2 <-
-        lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
-             data = .x,
-             weights = .x[[weights_var]])
-
-      # Take the country coefficients (absolute coefficients)
-      country_coef <-
-        coef(mod2)$country %>%
-        rownames_to_column() %>%
-        gather(escs_dummy, Mean, -rowname) %>%
-        mutate(escs_dummy = dplyr::recode(escs_dummy,
-                                          `(Intercept)` = "0",
-                                          `escs_dummy` = "1"))
-
-      # Take the absolute country standard errors
-      se <-
-        se.coef(mod2)$country %>%
-        as.data.frame() %>%
-        rownames_to_column() %>%
-        gather(escs_dummy, s.e., -rowname) %>%
-        mutate(escs_dummy = dplyr::recode(escs_dummy,
-                                          `(Intercept)` = "0",
-                                          `escs_dummy` = "1"))
-
-      results <-
-        inner_join(country_coef, se, by = c("rowname", "escs_dummy")) %>%
-        rename(country = rowname) %>%
-        arrange(country, escs_dummy)
-
-      message(paste0(unique(.x$wave), " modeling done"))
-      results
-    })
-}
-@
-
-<<modelling, echo = F, cache = T>>=
-adapted_year_data <-
-    map(pisa_all2$value, ~ {
-      if (unique(.x$wave) == "pisa2000") {
-        # pisa2000 has a different coding so here I recode 6 to 7 so that in all waves the top edu
-        # is 7 and the bottom is 1
-        .x <- 
-          mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `6` = 7)))
-      } else {
-        .x <-
-          mutate(.x, new_hisced = as.character(high_edu_broad))
-      }
-      .x
-})
-
-results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH")
-results_read <- test_diff(adapted_year_data, reliability_pisa, "READ")
-# US is missing for reading
-
-@
 \end{document}

### 2017-06-02 paper/paper.Rnw
Started building descriptive tables of sample size

@@
 
-<<loading_data-recoding, cache = TRUE>>=
+<<loading_data-recoding>>=
 pisa_all <- read_rds("./data/pisa_listcol.Rdata")
@@
 
-<<modeling, cache = T>>=
-adapted_year_data <-
-    map(pisa_all2$value, ~ {
-      if (unique(.x$wave) == "pisa2000") {
-        # pisa2000 has a different coding so here I recode 6 to 7 so that in all waves the top edu
-        # is 7 and the bottom is 1
-        .x <- 
-          mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `6` = 7)))
-      } else {
-        .x <-
-          mutate(.x, new_hisced = as.character(high_edu_broad))
+<<modeling>>=
+# adapted_year_data <-
+#     map(pisa_all2$value, ~ {
+#       if (unique(.x$wave) == "pisa2000") {
+#         # pisa2000 has a different coding so here I recode 6 to 7 so that in all waves the top edu
+#         # is 7 and the bottom is 1
+#         .x <- 
+#           mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `6` = 7)))
+#       } else {
+#         .x <-
+#           mutate(.x, new_hisced = as.character(high_edu_broad))
+#       }
+#       .x
+# })
+# 
+# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH")
+# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ")
+# US is missing for reading
+
+# Cache is not working properly for the code above, so I just load the saved cached file
+load("./paper/cache/modeling_9a0b38d1d53fa243b0242580f0672fa5.RData")
+
+@
+
+<<>>=
+i <- 3
+vals <- intersect(unique(adapted_year_data[[1]]$country), unique(adapted_year_data[[2]]$country))
+
+while (i < 7) {
+  vals <- intersect(vals, unique(adapted_year_data[[i]]$country))
+  i <- i + 1
 }
+
+subset_cnts <- map(adapted_year_data, ~ filter(.x, country %in% vals))
+
+cnt_to_bind <-
+  map(subset_cnts, function(df) {
+
+      print(unique(df$wave))
+      conf <- if (unique(df$wave) == "pisa2015") pisa2015_conf else pisa_conf
+      weights_var <- conf$variables$weightFinal
+
+      split_df <- split(df, df$country)
+
+      split_df_two <-
+        map(split_df, ~ {
+      # In some countries the quan can't be estimated because of very few obs.
+      # The function doesn't stop but returns two NA's.
+      quan <- quantile_missing(.x, weights_var)
+
+      # It's very important to create a variable that returns the number of observations of this dummy
+      # For each country. Possibly to weight by the number of observations.
+      .x$escs_dummy <-
+        with(.x, case_when(escs_trend >= quan[2] ~ 1,
+                                  escs_trend <= quan[1] ~ 0))
       .x
     })
+  unsplit_df <- split_df_two %>% enframe() %>% unnest(value)
+  unsplit_df
+})
 
-results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH")
-results_read <- test_diff(adapted_year_data, reliability_pisa, "READ")
-# US is missing for reading
+selected_cnt <- map(cnt_to_bind, ~ select(.x, country, high_edu_broad, new_hisced, escs_dummy))
+
+full_cnt <- 
+  selected_cnt %>%
+  enframe() %>%
+  unnest(value) %>%
+  mutate(year = as.character(dplyr::recode(.$name,
+                `1` = 2000,
+                `2` = 2003,
+                `3` = 2006,
+                `4` = 2009,
+                `5` = 2012,
+                `6` = 2015)),
+         escs_dummy = as.character(escs_dummy))
+
+sample_size <-
+  full_cnt %>%
+  count(year, country)
+
+high_low_sample <-
+  full_cnt %>%
+  count(year, country, new_hisced) %>%
+  filter(new_hisced %in% c(1, 7)) %>%
+  spread(new_hisced, n) %>%
+  rename(low_isced = `1`, high_isced = `7`)
+
+ses_sample <-
+  full_cnt %>%
+  count(year, country, new_hisced, escs_dummy) %>%
+  filter(new_hisced == 7 & escs_dummy != 0 | new_hisced == 1 & escs_dummy != 1) %>%
+  ungroup() %>%
+  select(-new_hisced) %>%
+  spread(escs_dummy, n) %>%
+  rename(low_low_isced = `0`, high__high_isced = `1`)
+
+left_join(sample_size, high_low_sample) %>%
+  left_join(ses_sample)
+
+@
+
+<<eval = F>>=
+# In this chunk you can join reading and math datasets
+descrip_math <- map(results_math, ~ rename(.x, mean_math = Mean, se_math = s.e.))
+descrip_read <- map(results_read, ~ rename(.x, mean_read = Mean, se_read = s.e.))
+
+map2(results_math, results_read, inner_join, by = "country")
 @

### 2017-06-02 paper/paper.Rnw
Sample size was too low with current specs - changed to have bigger sample

@@
 
-    df_lower <- filter(df, new_hisced == 1)
+    df_lower <- filter(df, new_hisced %in% c(1, 2))
     df_upper <- filter(df, new_hisced == 7)
@@
                                           weights = df_lower[[weights]],
-                                          probs = c(0.30)))
+                                          probs = c(0.50)))
 
@@
 
-left_join(sample_size, high_low_sample) %>%
-  left_join(ses_sample)
+m <- left_join(sample_size, high_low_sample) %>%
+  left_join(ses_sample) %>%
+  arrange(country, year)
 

### 2017-06-04 paper.R
Code from the Rmarkdown file with some changes

@@ -1,0 +1,648 @@
+## ----working directory, echo = F-----------------------------------------
+opts_knit$set(root.dir = '..')
+
+opts_chunk$set(echo = F,
+               message = F,
+               warning = F,
+               include = F,
+               cache.lazy = F,
+               results = 'asis')
+
+## ----packages_conf, echo = F---------------------------------------------
+  library(knitr)
+  library(arm)
+  library(saves)
+  library(haven)
+  library(PISA2000lite)
+  library(PISA2003lite)
+  library(PISA2006lite)
+  library(PISA2009lite)
+  library(PISA2012lite)
+  library(intsvy)
+  library(cimentadaj) # For pisa_countrynames
+  library(countrycode) # For region variable
+  library(car)
+  library(readr)
+  library(SAScii)
+  library(inequalityintsvy)
+  library(lme4)
+  library(modelr)
+  library(tidyverse)
+  
+  # source("./transform_data.R")
+
+  # Conf for PISA_2015
+  pisa2015_conf <- list(variables = list(pvlabelpref = "PV",
+                                         pvlabelsuff = "READ",
+                                         weightFinal = "W_FSTUWT",
+                                         weightBRR = "W_FSTURWT"),
+          parameters = list(cutoffs = c(357.77, 420.07, 482.38, 544.68, 606.99, 669.30),
+                                          percentiles = c(5, 10, 25, 75, 90, 95),
+                                          PVreps = 10,
+                                          BRRreps = 80,
+                                          weights = "BRR",
+                                          replication_scheme = 'pisa')
+  )
+
+
+## ----loading_data-recoding-----------------------------------------------
+pisa_all <- read_rds("./data/pisa_listcol.Rdata")
+pisa_all2 <- pisa_all
+
+years <- seq(2000, 2015, 3)
+countries <- c("Germany", "United States", "Denmark", "Sweden", "United Kingdom",
+                 "Spain", "Italy", "Canada", "Australia")
+  
+db <- paste0("pisa", years)
+pisa_all2$value <- map2(pisa_all2$value, db, ~ { .x$wave <- .y; .x})
+pisa_all2$value[[1]]$CNT <- pisa_all2$value[[1]]$COUNTRY
+  
+pisa_all2$value <- map(pisa_all2$value, ~ {
+  
+# 2000 to 2015
+# The coding is from 0 to 6, where 0 is no schooling and 6 is
+# BA or above.
+
+# When turning 0:6 to numeric, it becomes 1:7 that's why
[…]

### 2017-06-04 paper/paper.Rnw
Played around with the Rnw file but wasn't getting it -- moved to only code file first

@@
 
-    df_lower <- filter(df, new_hisced %in% c(1, 2))
+    df_lower <- filter(df, new_hisced %in% 1)
     df_upper <- filter(df, new_hisced == 7)
@@
 <<modeling>>=
-# adapted_year_data <-
-#     map(pisa_all2$value, ~ {
-#       if (unique(.x$wave) == "pisa2000") {
-#         # pisa2000 has a different coding so here I recode 6 to 7 so that in all waves the top edu
-#         # is 7 and the bottom is 1
-#         .x <- 
-#           mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `6` = 7)))
-#       } else {
-#         .x <-
-#           mutate(.x, new_hisced = as.character(high_edu_broad))
-#       }
-#       .x
-# })
-# 
-# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH")
+adapted_year_data <-
+    map(pisa_all2$value, ~ {
+      if (unique(.x$wave) == "pisa2000") {
+        # pisa2000 has a different coding so here I recode 6 to 7 so that in all waves the top edu
+        # is 7 and the bottom is 1
+        .x <-
+          mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `6` = 7)))
+      } else {
+        .x <-
+          mutate(.x, new_hisced = as.character(high_edu_broad))
+      }
+      .x
+})
+
+results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH")
 # results_read <- test_diff(adapted_year_data, reliability_pisa, "READ")
@@
 # Cache is not working properly for the code above, so I just load the saved cached file
-load("./paper/cache/modeling_9a0b38d1d53fa243b0242580f0672fa5.RData")
+# load("./paper/cache/modeling_9a0b38d1d53fa243b0242580f0672fa5.RData")
 
@@
 <<>>=
-i <- 3
-vals <- intersect(unique(adapted_year_data[[1]]$country), unique(adapted_year_data[[2]]$country))
+vals <- c("Australia",
+          "Germany",
+          "Denmark",
+          "Spain",
+          "France",
+          "Italy",
+          "Netherlands",
+          "Sweden",
+          "United States")
+# i <- 3
+# vals <- intersect(unique(adapted_year_data[[1]]$country), unique(adapted_year_data[[2]]$country))
 
-while (i < 7) {
-  vals <- intersect(vals, unique(adapted_year_data[[i]]$country))
-  i <- i + 1
-}
+# while (i < 7) {
+#   vals <- intersect(vals, unique(adapted_year_data[[i]]$country))
+#   i <- i + 1
+# }
 
-subset_cnts <- map(adapted_year_data, ~ filter(.x, country %in% vals))
+sample_size_calc <- function(df, selected = F, cnts) {
+
+if (selected) df <- map(df, ~ filter(.x, country %in% cnts))
 
 cnt_to_bind <-
-  map(subset_cnts, function(df) {
+  map(df, function(df) {
 
@@
 
-m <- left_join(sample_size, high_low_sample) %>%
+selected_cnts <-
+  left_join(sample_size, high_low_sample) %>%
   left_join(ses_sample) %>%
-  arrange(country, year)
+  arrange(year, country)
+
+selected_cnts
+}
+
+selected_cnts <- sample_size_calc(adapted_year_data, vals)  
+xtable::xtable(selected_cnts)
+@
+
+<<>>=
+all_cnts <- sample_size_calc(adapted_year_data)
+@
+
+<<>>=
+selected_cnt <- map(adapted_year_data, ~ select(.x, country, high_edu_broad, new_hisced, escs_dummy))
+
+full_cnt <- 
+  selected_cnt %>%
+  enframe() %>%
+  unnest(value) %>%
+  mutate(year = as.character(dplyr::recode(.$name,
+                `1` = 2000,
+                `2` = 2003,
+                `3` = 2006,
+                `4` = 2009,
+                `5` = 2012,
+                `6` = 2015)),
+         escs_dummy = as.character(escs_dummy))
+
+sample_size <-
+  full_cnt %>%
+  count(year, country)
+
+high_low_sample <-
+  full_cnt %>%
+  count(year, country, new_hisced) %>%
+  filter(new_hisced %in% c(1, 7)) %>%
+  spread(new_hisced, n) %>%
+  rename(low_isced = `1`, high_isced = `7`)
+
+ses_sample <-
+  full_cnt %>%
+  count(year, country, new_hisced, escs_dummy) %>%
+  filter(new_hisced == 7 & escs_dummy != 0 | new_hisced == 1 & escs_dummy != 1) %>%
+  ungroup() %>%
+  select(-new_hisced) %>%
+  spread(escs_dummy, n) %>%
+  rename(low_low_isced = `0`, high__high_isced = `1`)
+
+selected_cnts <-
+  left_join(sample_size, high_low_sample) %>%
+  left_join(ses_sample) %>%
+  arrange(year, country)
 
@@
 reduced_data %>%
-    filter(country %in% c("United States")) %>%
+    filter(country %in% c("United Kingdom")) %>%
     ggplot(aes(as.character(wave), Mean, group = escs_dummy, colour = escs_dummy)) +
@@
            upper = Mean + 1.96 * se_diff) %>%
-    filter(!is.na(continent), !is.na(region), country %in% countries) %>%
+    filter(!is.na(continent), !is.na(region), country %in% vals) %>%
     ggplot(aes(as.factor(wave), Mean, group = country, colour = country)) +

### 2017-06-04 inequality_explr.R
Ignored cache/figure folders of paper + moved transforming data to other script

@@ -7,7 +7,7 @@ library(tidyverse)
 library(car)
 library(readr)
 
-# source("./transform_data.R")
+source("./transform_data.R")
 
 # For PISA_2015
 pisa2015_conf <- list(variables = list(pvlabelpref = "PV",

### 2017-06-04 paper.R
Adjusting read/math scores to be together - adj both gaps + standard errors

@@ -379,7 +379,7 @@ results_read <- test_diff(adapted_year_data, reliability_pisa, "READ")
 
 
 ## ------------------------------------------------------------------------
-vals <- c("Australia",
+countries_subset <- c("Australia",
           "Germany",
           "Denmark",
           "Spain",
@@ -470,8 +470,8 @@ selected_cnts <-
 selected_cnts
 }
 
-selected_cnts <- sample_size_calc(adapted_year_data, vals)  
-xtable::xtable(selected_cnts)
+selected_cnts <- sample_size_calc(adapted_year_data, selected = TRUE, countries_subset)  
+# xtable::xtable(selected_cnts)
 
 ## ------------------------------------------------------------------------
 all_cnts <- sample_size_calc(adapted_year_data)
@@ -519,49 +519,90 @@ selected_cnts <-
 
 
 ## ----eval = F------------------------------------------------------------
-## # In this chunk you can join reading and math datasets
-## descrip_math <- map(results_math, ~ rename(.x, mean_math = Mean, se_math = s.e.))
-## descrip_read <- map(results_read, ~ rename(.x, mean_read = Mean, se_read = s.e.))
-## 
-## map2(results_math, results_read, inner_join, by = "country")
+# In this chunk you can join reading and math datasets
+descrip_math <- map(results_math, ~ rename(.x, mean_math = Mean, se_math = s.e.))
+descrip_read <- map(results_read, ~ rename(.x, mean_read = Mean, se_read = s.e.))
+
+both_tests <- map2(descrip_math, descrip_read, inner_join, by = c("country", "escs_dummy"))
 
 ## ----ci_for_difference---------------------------------------------------
 reduced_data <-
-    map2(results_math, years, function(.x, .y) {
+    map2(both_tests, years, function(.x, .y) {
       .x %>%
         mutate(wave = .y) %>%
         filter(!is.na(escs_dummy))
     }) %>%
     bind_rows() %>%
-    mutate(lower = Mean - 1.96 * s.e.,
-           upper = Mean + 1.96 * s.e.)
+    as_tibble() %>%
+    mutate(lower_math = mean_math - 1.96 * se_math,
+           upper_math = mean_math + 1.96 * se_math,
+           lower_read = mean_read - 1.96 * se_read,
+           upper_read = mean_read + 1.96 * se_read)
+
+test_data <-
+  reduced_data %>%
[…]

### 2017-06-04 paper.R
Finished merging read/math into one single line plot for all countries

@@ -387,7 +387,8 @@ countries_subset <- c("Australia",
           "Italy",
           "Netherlands",
           "Sweden",
-          "United States")
+          "United States",
+          "United Kingdom")
 # i <- 3
 # vals <- intersect(unique(adapted_year_data[[1]]$country), unique(adapted_year_data[[2]]$country))
 
@@ -604,26 +605,62 @@ se_data <- left_join(math_se_data, read_se_data)
 # Calculate the different between the gap and together with it's joint s.e graph
 # the absolut difference.
 
-reduced_data %>%
-    select(country, escs_dummy, Mean, wave) %>%
-    spread(key = escs_dummy, value = Mean) %>%
-    mutate(diff = `1` - `0`) %>%
-    gather(escs_dummy, Mean, -country, -wave, -`1`, -`0`) %>%
-    left_join(se_data) %>%
-    mutate(continent = countrycode(country, "country.name", "continent"),
-           region = countrycode(country, "country.name", "region"),
-           lower = Mean - 1.96 * se_diff,
-           upper = Mean + 1.96 * se_diff) %>%
-    filter(!is.na(continent), !is.na(region), country %in% vals) %>%
-    ggplot(aes(as.factor(wave), Mean, group = country, colour = country)) +
+math_diff <-
+  reduced_data %>%
+  select(wave, country, escs_dummy, mean_math) %>%
+  spread(escs_dummy, mean_math) %>%
+  transmute(wave, country, diff_math = `1` - `0`)
+
+read_diff <-
+  reduced_data %>%
+  select(wave, country, escs_dummy, mean_read) %>%
+  spread(escs_dummy, mean_read) %>%
+  transmute(wave, country, diff_read = `1` - `0`)
+
+data_summaries <-
+  math_diff %>%
+  left_join(read_diff) %>%
+  left_join(se_data) %>%
+  transmute(wave, country, diff_math, diff_read,
+           lower_math = diff_math - 1.96 * se_diff_math,
+           lower_read = diff_read - 1.96 * se_diff_read,
+           upper_math = diff_math + 1.96 * se_diff_math,
+           upper_read = diff_read + 1.96 * se_diff_read)
+
+differences <-
+  data_summaries %>%
+  select(wave, country, diff_math, diff_read) %>%
+  gather(test, difference, starts_with("diff")) %>%
+  mutate(type_test = ifelse(.$test == "diff_math", "math", "read"))
+
+
+bounds_lower <-
[…]

### 2017-06-05 paper.R
Added rate of change graph + rate of closing gap

@@ -28,6 +28,7 @@ opts_chunk$set(echo = F,
   library(lme4)
   library(modelr)
   library(tidyverse)
+  library(ggrepel)
   
   # source("./transform_data.R")
 
@@ -387,6 +388,7 @@ countries_subset <- c("Australia",
           "Italy",
           "Netherlands",
           "Sweden",
+          "Finland",
           "United States",
           "United Kingdom")
 # i <- 3
@@ -526,19 +528,24 @@ descrip_read <- map(results_read, ~ rename(.x, mean_read = Mean, se_read = s.e.)
 
 both_tests <- map2(descrip_math, descrip_read, inner_join, by = c("country", "escs_dummy"))
 
-## ----ci_for_difference---------------------------------------------------
+
+## -- correlation with indicators --------
 reduced_data <-
-    map2(both_tests, years, function(.x, .y) {
-      .x %>%
-        mutate(wave = .y) %>%
-        filter(!is.na(escs_dummy))
-    }) %>%
-    bind_rows() %>%
-    as_tibble() %>%
-    mutate(lower_math = mean_math - 1.96 * se_math,
-           upper_math = mean_math + 1.96 * se_math,
-           lower_read = mean_read - 1.96 * se_read,
-           upper_read = mean_read + 1.96 * se_read)
+  map2(both_tests, years, function(.x, .y) {
+    .x %>%
+      mutate(wave = .y) %>%
+      filter(!is.na(escs_dummy))
+  }) %>%
+  bind_rows() %>%
+  as_tibble() %>%
+  mutate(lower_math = mean_math - 1.96 * se_math,
+         upper_math = mean_math + 1.96 * se_math,
+         lower_read = mean_read - 1.96 * se_read,
+         upper_read = mean_read + 1.96 * se_read)
+
+
+## ----ci_for_difference---------------------------------------------------
+
 
 test_data <-
   reduced_data %>%
@@ -560,26 +567,6 @@ read_data <-
   right_join(filter(test_data, test == "mean_read"))
 
 all_data <- bind_rows(math_data, read_data)
-
-## ----graphin differences, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'----
-
-# to visualize the gap use the specific read_*, math_* datasets.
-
-math_data %>%
-  spread(test_bound, bound) %>%
-  filter(country %in% c("United Kingdom")) %>%
-  ggplot(aes(as.character(wave), score, group = escs_dummy, colour = escs_dummy)) +
[…]

### 2017-06-05 paper.R
Added gap-gini correlations - non linearity spotted

@@ -543,9 +543,38 @@ reduced_data <-
          lower_read = mean_read - 1.96 * se_read,
          upper_read = mean_read + 1.96 * se_read)
 
+# When leaving the GINI and the avg difference as is, no correlation
+reduced_data %>%
+  select(country, wave, escs_dummy, mean_math) %>%
+  spread(escs_dummy, mean_math) %>%
+  transmute(country, wave = as.character(wave), avg_diff = `1` - `0`) %>%
+  left_join(inequality_data, by = c("wave" = "year", "country")) %>%
+  filter(indicators == "GINI") %>%
+  group_by(country) %>%
+  summarize(avg_diff = mean(avg_diff, na.rm = T),
+            avg_value = mean(value, na.rm = T)) %>%
+  ggplot(aes(avg_value, avg_diff)) +
+  geom_point() +
+  geom_smooth(method = "lm")
+
+
+# But if I eclude some outliers, the relationship is non linear
+reduced_data %>%
+  select(country, wave, escs_dummy, mean_math) %>%
+  spread(escs_dummy, mean_math) %>%
+  transmute(country, wave = as.character(wave), avg_diff = `1` - `0`) %>%
+  left_join(inequality_data, by = c("wave" = "year", "country")) %>%
+  filter(indicators == "GINI") %>%
+  group_by(country) %>%
+  summarize(avg_diff = mean(avg_diff, na.rm = T),
+            avg_value = mean(value, na.rm = T),
+            high_value = avg_value > 0.3) %>%
+  filter(avg_diff <= 2.7) %>%
+  ggplot(aes(avg_value, avg_diff, colour = high_value)) +
+  geom_point() +
+  geom_smooth(method = "lm")
 
 ## ----ci_for_difference---------------------------------------------------
-
 
 test_data <-
   reduced_data %>%

### 2017-06-05 paper.R
Making script only to try different gap options

@@ -152,17 +152,17 @@ reliability_pisa <-
 # for the top educated. If the quantiles can't be estimated, it returns two NA's instead
 quantile_missing <- function(df, weights) {
     
-    df_lower <- filter(df, new_hisced == 1)
+    df_lower <- filter(df, new_hisced %in% c(1, 2))
     df_upper <- filter(df, new_hisced == 7)
     
     
     quan_lower <- try(Hmisc::wtd.quantile(df_lower$escs_trend,
                                           weights = df_lower[[weights]],
-                                          probs = c(0.50)))
+                                          probs = c(0.10)))
     
     quan_upper <- try(Hmisc::wtd.quantile(df_upper$escs_trend,
                                     weights = df_upper[[weights]],
-                                    probs = c(0.70)))
+                                    probs = c(0.90)))
     
     if (any("try-error" %in% c(class(quan_lower), class(quan_upper)))) {
       return(c(NA, NA))
@@ -280,99 +280,8 @@ adapted_year_data <-
       .x
 })
 
-## ----distributions---------
-theme_pub <- function (base_size = 12, base_family = "") {
-  
-  theme_grey(base_size = base_size, 
-             base_family = base_family) %+replace% 
-    
-    theme(# Set text size
-      plot.title = element_text(size = 18),
-      axis.title.x = element_text(size = 16),
-      axis.title.y = element_text(size = 16, 
-                                  angle = 90),
-      
-      axis.text.x = element_text(size = 14),
-      axis.text.y = element_text(size = 14),
-      
-      strip.text.x = element_text(size = 15),
-      strip.text.y = element_text(size = 15,
-                                  angle = -90),
-      
-      # Legend text
-      legend.title = element_text(size = 15),
-      legend.text = element_text(size = 15),
-      
-      # Configure lines and axes
-      axis.ticks.x = element_line(colour = "black"), 
-      axis.ticks.y = element_line(colour = "black"), 
-      
-      # Plot background
-      panel.background = element_rect(fill = "white"),
-      panel.grid.major = element_line(colour = "grey83",
-                                      size = 0.2),
-      # panel.grid.minor = element_line(colour = "grey88", 
[…]

### 2017-06-05 paper.R
Cleaned unnecesary options

@@ -1,13 +1,3 @@
-## ----working directory, echo = F-----------------------------------------
-opts_knit$set(root.dir = '..')
-
-opts_chunk$set(echo = F,
-               message = F,
-               warning = F,
-               include = F,
-               cache.lazy = F,
-               results = 'asis')
-
 ## ----packages_conf, echo = F---------------------------------------------
   library(knitr)
   library(arm)

### 2017-06-14 paper.R
Estimated SES gaps with escs for all gaps (top/mid/bottom) and everything working

@@ -140,24 +140,18 @@ reliability_pisa <-
 
 # Function calculates the bottom 30th quantile for the bottom educated and the 70th quantile
 # for the top educated. If the quantiles can't be estimated, it returns two NA's instead
-quantile_missing <- function(df, weights) {
+quantile_missing <- function(df, weights, probs) {
     
-    df_lower <- filter(df, new_hisced %in% c(1, 2))
-    df_upper <- filter(df, new_hisced == 7)
-    
-    
-    quan_lower <- try(Hmisc::wtd.quantile(df_lower$escs_trend,
-                                          weights = df_lower[[weights]],
-                                          probs = c(0.10)))
-    
-    quan_upper <- try(Hmisc::wtd.quantile(df_upper$escs_trend,
-                                    weights = df_upper[[weights]],
-                                    probs = c(0.90)))
-    
-    if (any("try-error" %in% c(class(quan_lower), class(quan_upper)))) {
+    quan <- try(Hmisc::wtd.quantile(
+      df$escs_trend,
+      weights = df[[weights]],
+      probs = probs
+      ))
+
+    if (any("try-error" %in% class(quan))) {
       return(c(NA, NA))
       } else {
-     return(c(quan_lower[1], quan_upper[1]))
+     return(c(quan[1], quan[2]))
     }
 }
   
@@ -168,21 +162,19 @@ quantile_missing <- function(df, weights) {
   
 # It returns a dataframe for each survey with all countries and respective coefficients and
 # standard errors.
-test_diff <- function(df, reliability, test) {
+test_diff <- function(df, reliability, test, probs) {
   
     map2(df, reliability, function(.x, .y) {
       
       conf <- if (unique(.x$wave) == "pisa2015") pisa2015_conf else pisa_conf
       weights_var <- conf$variables$weightFinal
-      
+
       country_split <- split(.x, .x$country)
       
       country_list <- map(country_split, function(country) {
         print(unique(country$country))
         
-        # In some countries the quan can't be estimated because of very few obs.
-        # The function doesn't stop but returns two NA's.
-        quan <- quantile_missing(country, weights_var)
+        quan <- quantile_missing(country, weights_var, probs)
[…]

### 2017-06-15 paper.R
Estimated SES gaps and made graphs for 90/50/10 gaps - everything working

@@ -301,8 +301,8 @@ countries_subset <- c("Australia",
 
 pisa_preparer <- function(df_math, df_read) {
 
-descrip_math <- map(results_math, ~ rename(.x, mean_math = Mean, se_math = s.e.))
-descrip_read <- map(results_read, ~ rename(.x, mean_read = Mean, se_read = s.e.))
+descrip_math <- map(df_math, ~ rename(.x, mean_math = Mean, se_math = s.e.))
+descrip_read <- map(df_read, ~ rename(.x, mean_read = Mean, se_read = s.e.))
 
 ## -- correlation with indicators --------
 reduced_data_math <-
@@ -486,14 +486,25 @@ complete_data_midbottom <- mutate(complete_data_midbottom, type = "50th/10th SES
 
 complete_data_topbottom %>%
   bind_rows(complete_data_topmid) %>%
-  bind_rows(midbottom) %>%
-  filter(country %in% c("United States", "Denmark")) %>%
+  bind_rows(complete_data_midbottom) %>%
+  filter(country %in% c("United States", "Denmark", "France")) %>%
+  mutate(type = factor(type,
+                       levels = c("90th/10th SES gap", "90th/50th SES gap", "50th/10th SES gap"),
+                       ordered = TRUE)) %>%
   ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
   geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
+  geom_hline(yintercept = 0, linetype = "longdash") +
   geom_line() +
   geom_point(size = 0.5) +
-  coord_cartesian(ylim = c(0, 4)) +
+  coord_cartesian(ylim = c(-0.5, 3)) +
   facet_grid(country ~ type)
+
+# Show plot for selected countries
+# Show plot of gaps evolving for even more selected countries
+# Show table with % increase decrease over time
+# Show plot comparing coefficients of which countries is the top decreasing/increasing faster
+# Get sample size for these calculations
+# Save top/mid datasets for not having to load them again
 
 # Increase:  
 # Sweden - steady increase in both tests

### 2017-06-15 paper.R
Calculated overall trend for each country and graph in scatterplot

@@ -298,6 +298,8 @@ countries_subset <- c("Australia",
 
 ## ----eval = F------------------------------------------------------------
 # In this chunk you can join reading and math datasets
+df_math <- results_math
+df_read <- results_read
 
 pisa_preparer <- function(df_math, df_read) {
 
@@ -470,9 +472,34 @@ bounds_upper <-
 #   select(wave, country, contains("upper")) %>%
 #   gather(upper_bound, upper, upper_math)
 
+# Getting the original data in
+df %>% 
+  gather(variable, value, -(month:student)) %>%
+  unite(temp, student, variable) %>%
+  spread(temp, value)
+
+original_math <-
+  reduced_data_math %>%
+  select(wave, everything(), -se_math) %>%
+  gather(metric, value, -(wave:escs_dummy)) %>%
+  unite(combination, escs_dummy, metric, sep = "_") %>%
+  spread(combination, value) %>%
+  mutate(type_test = "math")
+
+original_read <-
+  reduced_data_read %>%
+  select(wave, everything(), -se_read) %>%
+  gather(metric, value, -(wave:escs_dummy)) %>%
+  unite(combination, escs_dummy, metric, sep = "_") %>%
+  spread(combination, value) %>%
+  mutate(type_test = "read")
+
+# final data
 complete_data <-
   left_join(differences, bounds_lower) %>%
-  left_join(bounds_upper)
+  left_join(bounds_upper) %>%
+  left_join(original_math) %>%
+  left_join(original_read)
 }
 
 complete_data_topbottom <- pisa_preparer(results_math, results_read)
@@ -484,6 +511,19 @@ complete_data_topmid <- mutate(complete_data_topmid, type = "90th/50th SES gap")
 complete_data_midbottom <- mutate(complete_data_midbottom, type = "50th/10th SES gap")
 
 
+# 90/10 gaps acros countries
+complete_data_topbottom %>%
+  filter(country %in% c("United States", "Netherlands", "France",
+                        "Germany", "Poland", "Finland")) %>%
+  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
+  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
+  geom_hline(yintercept = 0, linetype = "longdash") +
+  geom_line() +
+  geom_point(size = 0.5) +
+  coord_cartesian(ylim = c(-0.5, 3)) +
[…]

### 2017-06-15 paper.R
Small changes so that I can run script from start - all wrking now

@@ -267,14 +267,17 @@ adapted_year_data <-
 
 # results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH")
 # results_read <- test_diff(adapted_year_data, reliability_pisa, "READ")
+# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
+# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
+# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
+# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
+
 results_math <- read_rds("./data/delete.Rdata")
 results_read <- read_rds("./data/delete_read.Rdata")
-
-results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
-results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
-
-results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
-results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
+results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
+results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
+results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
+results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
 
 
 # US is missing for reading
@@ -298,8 +301,6 @@ countries_subset <- c("Australia",
 
 ## ----eval = F------------------------------------------------------------
 # In this chunk you can join reading and math datasets
-df_math <- results_math
-df_read <- results_read
 
 pisa_preparer <- function(df_math, df_read) {
 
@@ -473,11 +474,6 @@ bounds_upper <-
 #   gather(upper_bound, upper, upper_math)
 
 # Getting the original data in
-df %>% 
-  gather(variable, value, -(month:student)) %>%
-  unite(temp, student, variable) %>%
-  spread(temp, value)
-
 original_math <-
   reduced_data_math %>%
   select(wave, everything(), -se_math) %>%
[…]

### 2017-06-15 paper.R
Got sample size for escs_dummy for each country

@@ -279,7 +279,38 @@ results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
 results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
 results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
 
+sample_size_calc <- function(df, probs, selected = F, cnts) {
+  
+  if (selected) df <- map(df, ~ filter(.x, country %in% cnts))
+  
+  cnt_to_bind <-
+    map(df, function(df) {
+      
+      print(unique(df$wave))
+      conf <- if (unique(df$wave) == "pisa2015") pisa2015_conf else pisa_conf
+      weights_var <- conf$variables$weightFinal
+      
+      split_df <- split(df, df$country)
+      
+      split_df_two <-
+        map(split_df, ~ {
+          # In some countries the quan can't be estimated because of very few obs.
+          # The function doesn't stop but returns two NA's.
+          quan <- quantile_missing(.x, weights_var, probs)
+          
+          # It's very important to create a variable that returns the number of observations of this dummy
+          # For each country. Possibly to weight by the number of observations.
+          .x$escs_dummy <-
+            with(.x, case_when(escs_trend >= quan[2] ~ 1,
+                               escs_trend <= quan[1] ~ 0))
+          .x
+        })
+      unsplit_df <- split_df_two %>% enframe() %>% unnest(value)
+      unsplit_df
+    })
+}
 
+sample_tables <- sample_size_calc(adapted_year_data, c(.1, .9))
 # US is missing for reading
 
 # Cache is not working properly for the code above, so I just load the saved cached file
@@ -636,7 +667,7 @@ full_data %>%
             fill = rect_data$colour,
             alpha = 0.2,
             inherit.aes = FALSE) +
-  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.2, colour = "blue", size = 1) +
+  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1) +
   geom_point(alpha = 0.2) +
   geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
   geom_text_repel(data = filter(full_data, continent == "my_cnt"), aes(label = country), box.padding = unit(2.7, "lines")) +

### 2017-06-15 paper.R
Computed table only for selected countries + added percentage column next to n()

@@ -279,8 +279,9 @@ results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
 results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
 results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
 
-sample_size_calc <- function(df, probs, selected = F, cnts) {
+sample_size_calc <- function(df, probs, selected = F, cnts = NULL) {
   
+  stopifnot(selected & !is.null(NULL))
   if (selected) df <- map(df, ~ filter(.x, country %in% cnts))
   
   cnt_to_bind <-
@@ -306,11 +307,20 @@ sample_size_calc <- function(df, probs, selected = F, cnts) {
           .x
         })
       unsplit_df <- split_df_two %>% enframe() %>% unnest(value)
-      unsplit_df
+      
+      unsplit_df %>%
+        count(country, escs_dummy) %>%
+        filter(!is.na(escs_dummy)) %>%
+        left_join(summarize(group_by(unsplit_df, country), total_n = n()), by = "country") %>%
+        mutate(perc = paste0(round(n / total_n * 100, 0), "%")) %>%
+        select(-total_n)
     })
+  setNames(cnt_to_bind, seq(2000, 2015, 3)) %>%
+    enframe() %>%
+    unnest()
 }
 
-sample_tables <- sample_size_calc(adapted_year_data, c(.1, .9))
+sample_tables <- sample_size_calc(adapted_year_data, c(.1, .9), selected = TRUE, countries)
 # US is missing for reading
 
 # Cache is not working properly for the code above, so I just load the saved cached file

### 2017-06-15 paper.R
Calculated last/first year percentage change for each country

@@ -278,7 +278,9 @@ results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
 results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
 results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
 results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
+# US is missing for reading
 
+# Get sample counts for each dummy
 sample_size_calc <- function(df, probs, selected = F, cnts = NULL) {
   
   stopifnot(selected & !is.null(NULL))
@@ -319,9 +321,7 @@ sample_size_calc <- function(df, probs, selected = F, cnts = NULL) {
     enframe() %>%
     unnest()
 }
-
 sample_tables <- sample_size_calc(adapted_year_data, c(.1, .9), selected = TRUE, countries)
-# US is missing for reading
 
 # Cache is not working properly for the code above, so I just load the saved cached file
 # load("./paper/cache/modeling_9a0b38d1d53fa243b0242580f0672fa5.RData")

### 2017-06-15 paper.R
Calculated last/first year percentage change for each country - ignore previous

@@ -597,7 +597,6 @@ complete_data_topbottom %>%
   facet_grid(country ~ type)
 
 # Show table with % increase decrease over time
-# Get sample size for these calculations
 
 # Increase:  
 # Sweden - steady increase in both tests
@@ -694,7 +693,46 @@ full_data %>%
 ## ------------------------------------------------------------------------
 
 # Show the rates at which is increasing/decreasing
-mutate(full_data, diff = high_increase - low_increase)
+perc_increase_fun <- function(df) {
+  
+  # Average standard deviation increase
+  data_ready <-
+    df %>%
+    select(wave, country, type_test, difference) %>%
+    group_by(type_test) %>%
+    split(.$country) %>%
+    map(~ {
+      .x <-
+        spread(.x, wave, difference) %>%
+        ungroup()
+  
+      year_vars <- sum(map_dbl(.x, is.numeric)) - 1
+      years_subtract <- names(.x)[c(ncol(.x) - year_vars, ncol(.x))]
+      years_subtract <- lapply(years_subtract, as.name)
+
+      last_year <- rlang::new_quosure(years_subtract[[2]], env = .GlobalEnv)
+      first_year <- rlang::new_quosure(years_subtract[[1]], env = .GlobalEnv)
+      
+      years_available <-
+        .x %>%
+        gather(year, val, -(country:type_test)) %>%
+        group_by(type_test) %>%
+        summarise(yr_avaible = sum(!is.na(val))) %>%
+        pull(yr_avaible)
+      
+      .x %>%
+        ungroup() %>%
+        transmute(type_test,
+                  country,
+                  diff = round(((!!last_year) - (!!first_year)) / (!!first_year) * 100, 1),
+                  years_available = years_available)
+    })
+  data_ready
+}
+
+top_bottom_perc <- perc_increase_fun(complete_data_topbottom)
+top_mid_perc <- perc_increase_fun(complete_data_topmid)
+
 
 # Gap is closing at an average of the variable diff per year.
 

### 2017-06-15 paper.R
Added uncertainty and uncertainty intervals for % increase tables

@@ -720,11 +720,21 @@ perc_increase_fun <- function(df) {
         summarise(yr_avaible = sum(!is.na(val))) %>%
         pull(yr_avaible)
       
+      year_sd <-
+        .x %>%
+        gather(year, val, -(country:type_test)) %>%
+        group_by(type_test) %>%
+        summarise(sd_year = mad(val, na.rm = T)) %>%
+        pull(sd_year)
+      
       .x %>%
         ungroup() %>%
         transmute(type_test,
                   country,
                   diff = round(((!!last_year) - (!!first_year)) / (!!first_year) * 100, 1),
+                  sd_year = year_sd,
+                  diff_lower = diff - 1 * year_sd * 100,
+                  diff_upper = diff + 1 * year_sd * 100,
                   years_available = years_available)
     })
   data_ready

### 2017-06-16 paper.R
Added bootstrap standard deviation for percentage change

@@ -35,6 +35,21 @@
                                           replication_scheme = 'pisa')
   )
 
+# Adapted from: https://github.com/jtleek/slipper/blob/master/R/slipper.R
+bootstrapper <- function(df, expr, B = 100, n = nrow(df), replacement = TRUE) {
+  bootstrapper_(df, lazyeval::lazy(expr), B, n, replacement)
+}
+bootstrapper_ <- function(df, expr, B = 500, n = nrow(df), replacement = TRUE) {
+  obs_val = lazyeval::lazy_eval(expr, data = df)
+  boot_val = replicate(B, {
+    newdata = sample_n(df, n, replace = replacement)
+    lazyeval::lazy_eval(expr, data = newdata)
+  })
+  out = tibble(type = c("observed", "bootstrap"), 
+                   value = c(obs_val, mean(boot_val, na.rm = T)))
+  return(out)
+}
+
 
 ## ----loading_data-recoding-----------------------------------------------
 pisa_all <- read_rds("./data/pisa_listcol.Rdata")
@@ -723,18 +738,18 @@ perc_increase_fun <- function(df) {
       year_sd <-
         .x %>%
         gather(year, val, -(country:type_test)) %>%
-        group_by(type_test) %>%
-        summarise(sd_year = mad(val, na.rm = T)) %>%
-        pull(sd_year)
+        split(.$type_test) %>%
+        map_dbl(~ bootstrapper(.x, mad(val, na.rm = T), B = 100) %>% .[[2, 2]]) %>%
+        round(2) * 100
+
       
       .x %>%
-        ungroup() %>%
         transmute(type_test,
                   country,
-                  diff = round(((!!last_year) - (!!first_year)) / (!!first_year) * 100, 1),
+                  perc_diff = round(((!!last_year) - (!!first_year)) / (!!first_year) * 100, 1),
                   sd_year = year_sd,
-                  diff_lower = diff - 1 * year_sd * 100,
-                  diff_upper = diff + 1 * year_sd * 100,
+                  diff_lower = perc_diff - 1 * year_sd,
+                  diff_upper = perc_diff + 1 * year_sd,
                   years_available = years_available)
     })
   data_ready
@@ -742,7 +757,7 @@ perc_increase_fun <- function(df) {
 
 top_bottom_perc <- perc_increase_fun(complete_data_topbottom)
 top_mid_perc <- perc_increase_fun(complete_data_topmid)
-
[…]

### 2017-06-16 paper.R
Fixed NULL argument from sample function

@@ -35,22 +35,6 @@
                                           replication_scheme = 'pisa')
   )
 
-# Adapted from: https://github.com/jtleek/slipper/blob/master/R/slipper.R
-bootstrapper <- function(df, expr, B = 100, n = nrow(df), replacement = TRUE) {
-  bootstrapper_(df, lazyeval::lazy(expr), B, n, replacement)
-}
-bootstrapper_ <- function(df, expr, B = 500, n = nrow(df), replacement = TRUE) {
-  obs_val = lazyeval::lazy_eval(expr, data = df)
-  boot_val = replicate(B, {
-    newdata = sample_n(df, n, replace = replacement)
-    lazyeval::lazy_eval(expr, data = newdata)
-  })
-  out = tibble(type = c("observed", "bootstrap"), 
-                   value = c(obs_val, mean(boot_val, na.rm = T)))
-  return(out)
-}
-
-
 ## ----loading_data-recoding-----------------------------------------------
 pisa_all <- read_rds("./data/pisa_listcol.Rdata")
 pisa_all2 <- pisa_all
@@ -777,9 +761,6 @@ mid_bottom_perc <- perc_increase_fun(complete_data_midbottom)
 # Get each country trendline adjusted for the inequality indicators and place in the same country graph.
 
 # Should I add the parent's education in the lm model to see how trends change adjusted for that?
- 
-# The gap between the top 90th and the 50th, did it grow?
-# The gap between the top 50th and the 10th, did it grow?
 
 # A weak welfare system, together with income inequality, what's their pattern?
 # What if we put the school differentiation/tracking aspect in? Are there country groups based on these

### 2017-06-16 paper.R
Finalized all code chunks - ready for moving to paper

@@ -1,4 +1,4 @@
-## ----packages_conf, echo = F---------------------------------------------
+## ----default_conf, echo = F---------------------------------------------
   library(knitr)
   library(arm)
   library(saves)
@@ -34,14 +34,25 @@
                                           weights = "BRR",
                                           replication_scheme = 'pisa')
   )
+  
+  countries <- c("Australia",
+                 "Germany",
+                 "Denmark",
+                 "Spain",
+                 "France",
+                 "Italy",
+                 "Netherlands",
+                 "Sweden",
+                 "Finland",
+                 "United States",
+                 "United Kingdom")
+
 
 ## ----loading_data-recoding-----------------------------------------------
 pisa_all <- read_rds("./data/pisa_listcol.Rdata")
 pisa_all2 <- pisa_all
 
 years <- seq(2000, 2015, 3)
-countries <- c("Germany", "United States", "Denmark", "Sweden", "United Kingdom",
-                 "Spain", "Italy", "Canada", "Australia")
   
 db <- paste0("pisa", years)
 pisa_all2$value <- map2(pisa_all2$value, db, ~ { .x$wave <- .y; .x})
@@ -80,6 +91,7 @@ reliability_pisa <-
     "2015" = 0.74) # 2015 imputed
 
 
+
 ## ----escs_trend, cache = TRUE--------------------------------------------
   # Rescaled trend ESCS data to merge.
   # This only has data for seq(2000, 2012, 3) because
@@ -97,6 +109,7 @@ reliability_pisa <-
     mutate(.x, cnt = pisa_countrynames[cnt]) %>%
     rename(country = cnt)
   })
+
 
 
 ## ----merge_escs_pisa, cache = TRUE---------------------------------------
@@ -134,6 +147,7 @@ reliability_pisa <-
   pisa_all2$value[[6]] <-
     pisa_all2$value[[6]] %>%
     rename(escs_trend = ESCS)
+
 
 ## ----functions_for_modelling---------------------------------------------
 
@@ -249,7 +263,28 @@ test_diff <- function(df, reliability, test, probs) {
     })
 }
 
-## ----modeling------------------------------------------------------------
+
+# Adapted from: https://github.com/jtleek/slipper/blob/master/R/slipper.R
+# Returns a tibble with the actual expr + the bootstrapped expr.
[…]

### 2017-06-16 paper.R
Added a space between code chunks

@@ -48,6 +48,7 @@
                  "United Kingdom")
 
 
+
 ## ----loading_data-recoding-----------------------------------------------
 pisa_all <- read_rds("./data/pisa_listcol.Rdata")
 pisa_all2 <- pisa_all
@@ -92,6 +93,7 @@ reliability_pisa <-
 
 
 
+
 ## ----escs_trend, cache = TRUE--------------------------------------------
   # Rescaled trend ESCS data to merge.
   # This only has data for seq(2000, 2012, 3) because
@@ -109,6 +111,7 @@ reliability_pisa <-
     mutate(.x, cnt = pisa_countrynames[cnt]) %>%
     rename(country = cnt)
   })
+
 
 
 
@@ -147,6 +150,7 @@ reliability_pisa <-
   pisa_all2$value[[6]] <-
     pisa_all2$value[[6]] %>%
     rename(escs_trend = ESCS)
+
 
 
 ## ----functions_for_modelling---------------------------------------------
@@ -284,6 +288,7 @@ bootstrapper_ <- function(df, expr, B = 500, n = nrow(df), replacement = TRUE) {
 # bootstrapper(mtcars, mean(mpg), B = 200)
 
 
+
 ## ----modeling, echo = F------------------------------------------------------------
 adapted_year_data <-
     map(pisa_all2$value, ~ {
@@ -316,6 +321,7 @@ results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
 
 # Cache is not working properly for the code above, so I just load the saved cached file
 # load("./paper/cache/modeling_9a0b38d1d53fa243b0242580f0672fa5.RData")
+
 
 
 ## ----sample_size, echo = F------------------------------------------------------------
@@ -365,6 +371,7 @@ sample_size_calc <- function(df, probs, selected = F, cnts = NULL) {
 sample_tables_topbottom <- sample_size_calc(adapted_year_data, c(.1, .9), selected = TRUE, countries)
 sample_tables_topmid <- sample_size_calc(adapted_year_data, c(.5, .9), selected = TRUE, countries)
 sample_tables_midbottom <- sample_size_calc(adapted_year_data, c(.1, .5), selected = TRUE, countries)
+
 
 
 ## ----merge_math_read, echo = F------------------------------------------------------------
@@ -523,6 +530,7 @@ complete_data_topmid <- mutate(complete_data_topmid, type = "90th/50th SES gap")
[…]

### 2017-06-16 paper/paper.Rnw
Moving code into paper - 30% done

@@
 
-<<packages_conf, echo = F>>=
+<<default_conf>>=
   library(knitr)
@@
   library(tidyverse)
+  library(ggrepel)
 
@@
 
+  countries <- c("Australia",
+                 "Germany",
+                 "Denmark",
+                 "Spain",
+                 "France",
+                 "Italy",
+                 "Netherlands",
+                 "Sweden",
+                 "Finland",
+                 "United States",
+                 "United Kingdom")
 @
@@
 
-<<loading_data-recoding>>=
+<<loading_data-recoding, cache = TRUE>>=
 pisa_all <- read_rds("./data/pisa_listcol.Rdata")
@@
 years <- seq(2000, 2015, 3)
-countries <- c("Germany", "United States", "Denmark", "Sweden", "United Kingdom",
-                 "Spain", "Italy", "Canada", "Australia")
 
@@
 pisa_all2$value <- map(pisa_all2$value, ~ {
+
+# 2000 to 2015
+# The coding is from 0 to 6, where 0 is no schooling and 6 is
+# BA or above.
+
+# When turning 0:6 to numeric, it becomes 1:7 that's why
+# I recode 8:9 to NA. This, however, didn't work for last two surveys
+
   .x$father_edu <- car::recode(as.numeric(.x$FISCED), "8:9 = NA")
@@
   if (any(unique(.x$wave) %in% c("pisa2012", "pisa2015"))) {
+    # These two surveys were from 0:6 so I had to add + 1
+    # so that it equals 1:7 as all other surveys.
     .x$father_edu <- .x$father_edu + 1
@@
     "2015" = 0.74) # 2015 imputed
-
 @
@@
   })
-
 @
@@
 # for the top educated. If the quantiles can't be estimated, it returns two NA's instead
-quantile_missing <- function(df, weights) {
+quantile_missing <- function(df, weights, probs) {
 
-    df_lower <- filter(df, new_hisced %in% 1)
-    df_upper <- filter(df, new_hisced == 7)
+    quan <- try(Hmisc::wtd.quantile(
+      df$escs_trend,
+      weights = df[[weights]],
+      probs = probs
+      ))
 
-    quan_lower <- try(Hmisc::wtd.quantile(df_lower$escs_trend,
-                                          weights = df_lower[[weights]],
-                                          probs = c(0.50)))
-
-    quan_upper <- try(Hmisc::wtd.quantile(df_upper$escs_trend,
-                                    weights = df_upper[[weights]],
-                                    probs = c(0.70)))
-
-    if (any("try-error" %in% c(class(quan_lower), class(quan_upper)))) {
+    if (any("try-error" %in% class(quan))) {
       return(c(NA, NA))
       } else {
-     return(c(quan_lower[1], quan_upper[1]))
+     return(c(quan[1], quan[2]))
     }
@@
 # standard errors.
-test_diff <- function(df, reliability, test) {
+test_diff <- function(df, reliability, test, probs) {
 
@@
 
-        # In some countries the quan can't be estimated because of very few obs.
-        # The function doesn't stop but returns two NA's.
-        quan <- quantile_missing(country, weights_var)
+        quan <- quantile_missing(country, weights_var, probs)
 
@@
         .x %>%
-        select(wave, matches(paste0("^PV.*", test, "$")), escs_dummy,
-               country, one_of(weights_var),
+        dplyr::select(wave,
+                      matches(paste0("^PV.*", test, "$")),
+                      escs_dummy,
+                      country,
+                      one_of(weights_var),
                       AGE)
@@
 }
+
+# Adapted from: https://github.com/jtleek/slipper/blob/master/R/slipper.R
+# Returns a tibble with the actual expr + the bootstrapped expr.
+bootstrapper <- function(df, expr, B = 100, n = nrow(df), replacement = TRUE) {
+  bootstrapper_(df, lazyeval::lazy(expr), B, n, replacement)
+}
+
+bootstrapper_ <- function(df, expr, B = 500, n = nrow(df), replacement = TRUE) {
+  obs_val = lazyeval::lazy_eval(expr, data = df)
+  boot_val = replicate(B, {
+    newdata = sample_n(df, n, replace = replacement)
+    lazyeval::lazy_eval(expr, data = newdata)
+  })
+  out = tibble(type = c("observed", "bootstrap"), 
+               value = c(obs_val, mean(boot_val, na.rm = T)))
+  return(out)
+}
+
+# For example
+# bootstrapper(mtcars, mean(mpg), B = 200)
 @
+
+% This chunk needs to have cache = TRUE once you want to run it for like 2 hours until 
+% the results_* models have been created once and the cache can save them
 
@@
 
-results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH")
+# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH")
 # results_read <- test_diff(adapted_year_data, reliability_pisa, "READ")
+# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
+# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
+# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
+# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
+
+results_math <- read_rds("./data/delete.Rdata")
+results_read <- read_rds("./data/delete_read.Rdata")
+results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
+results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
+results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
+results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
 # US is missing for reading
@@
 # load("./paper/cache/modeling_9a0b38d1d53fa243b0242580f0672fa5.RData")
-
 @
 
-<<>>=
-vals <- c("Australia",
-          "Germany",
-          "Denmark",
-          "Spain",
-          "France",
-          "Italy",
-          "Netherlands",
-          "Sweden",
-          "United States")
-# i <- 3
-# vals <- intersect(unique(adapted_year_data[[1]]$country), unique(adapted_year_data[[2]]$country))
+% Figure out how to ouput data frames as latex tables.
+<<sample_size>>=
 
-# while (i < 7) {
-#   vals <- intersect(vals, unique(adapted_year_data[[i]]$country))
-#   i <- i + 1
-# }
+# Get sample counts for each dummy
+sample_size_calc <- function(df, probs, selected = F, cnts = NULL) {
 
-sample_size_calc <- function(df, selected = F, cnts) {
+  stopifnot(selected & !is.null(cnts))
 
@@
           # The function doesn't stop but returns two NA's.
-      quan <- quantile_missing(.x, weights_var)
+          quan <- quantile_missing(.x, weights_var, probs)
 
@@
       unsplit_df <- split_df_two %>% enframe() %>% unnest(value)
-  unsplit_df
+
+      unsplit_df %>%
+        count(country, escs_dummy) %>%
+        filter(!is.na(escs_dummy)) %>%
+        left_join(summarize(group_by(unsplit_df, country), total_n = n()), by = "country") %>%
+        mutate(perc = paste0(round(n / total_n * 100, 0), "%")) %>%
+        select(-total_n)
     })
-
-selected_cnt <- map(cnt_to_bind, ~ select(.x, country, high_edu_broad, new_hisced, escs_dummy))
-
-full_cnt <- 
-  selected_cnt %>%
+  setNames(cnt_to_bind, seq(2000, 2015, 3)) %>%
     enframe() %>%
-  unnest(value) %>%
-  mutate(year = as.character(dplyr::recode(.$name,
-                `1` = 2000,
-                `2` = 2003,
-                `3` = 2006,
-                `4` = 2009,
-                `5` = 2012,
-                `6` = 2015)),
-         escs_dummy = as.character(escs_dummy))
-
-sample_size <-
-  full_cnt %>%
-  count(year, country)
-
-high_low_sample <-
-  full_cnt %>%
-  count(year, country, new_hisced) %>%
-  filter(new_hisced %in% c(1, 7)) %>%
-  spread(new_hisced, n) %>%
-  rename(low_isced = `1`, high_isced = `7`)
-
-ses_sample <-
-  full_cnt %>%
-  count(year, country, new_hisced, escs_dummy) %>%
-  filter(new_hisced == 7 & escs_dummy != 0 | new_hisced == 1 & escs_dummy != 1) %>%
-  ungroup() %>%
-  select(-new_hisced) %>%
-  spread(escs_dummy, n) %>%
-  rename(low_low_isced = `0`, high__high_isced = `1`)
-
-selected_cnts <-
-  left_join(sample_size, high_low_sample) %>%
-  left_join(ses_sample) %>%
-  arrange(year, country)
-
-selected_cnts
+    unnest()
 }
 
-selected_cnts <- sample_size_calc(adapted_year_data, vals)  
-xtable::xtable(selected_cnts)
-@
-
-<<>>=
-all_cnts <- sample_size_calc(adapted_year_data)
+sample_tables_topbottom <- sample_size_calc(adapted_year_data, c(.1, .9), selected = TRUE, countries)
+sample_tables_topmid <- sample_size_calc(adapted_year_data, c(.5, .9), selected = TRUE, countries)
+sample_tables_midbottom <- sample_size_calc(adapted_year_data, c(.1, .5), selected = TRUE, countries)
 @

### 2017-06-16 paper/paper.Rnw
Moved code into paper - everything good

@@
   library(intsvy)
-  library(cimentadaj) # For pisa_countrynames
+  library(cimentadaj) # # devtools::install_github("cimentadaj/cimentadaj")
   library(countrycode) # For region variable
@@
   library(SAScii)
-  library(inequalityintsvy)
+  library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
   library(lme4)
@@
 
-<<>>=
-selected_cnt <- map(adapted_year_data, ~ select(.x, country, high_edu_broad, new_hisced, escs_dummy))
+<<merge_math_read>>=
 
-full_cnt <- 
-  selected_cnt %>%
-  enframe() %>%
-  unnest(value) %>%
-  mutate(year = as.character(dplyr::recode(.$name,
-                `1` = 2000,
-                `2` = 2003,
-                `3` = 2006,
-                `4` = 2009,
-                `5` = 2012,
-                `6` = 2015)),
-         escs_dummy = as.character(escs_dummy))
+# Function does a lot of things, but in short:
 
-sample_size <-
-  full_cnt %>%
-  count(year, country)
+# Calculate the difference between the gap and together with it's joint s.e
+# Also uncertainty intervals and returns a tibble with the difference between
+# SES gaps with the adjusted SE difference + uncertainty intervals + the original
+# data (the absolute numbers before the differences)
 
-high_low_sample <-
-  full_cnt %>%
-  count(year, country, new_hisced) %>%
-  filter(new_hisced %in% c(1, 7)) %>%
-  spread(new_hisced, n) %>%
-  rename(low_isced = `1`, high_isced = `7`)
+pisa_preparer <- function(df_math, df_read) {
 
-ses_sample <-
-  full_cnt %>%
-  count(year, country, new_hisced, escs_dummy) %>%
-  filter(new_hisced == 7 & escs_dummy != 0 | new_hisced == 1 & escs_dummy != 1) %>%
-  ungroup() %>%
-  select(-new_hisced) %>%
-  spread(escs_dummy, n) %>%
-  rename(low_low_isced = `0`, high__high_isced = `1`)
+descrip_math <- map(df_math, ~ rename(.x, mean_math = Mean, se_math = s.e.))
+descrip_read <- map(df_read, ~ rename(.x, mean_read = Mean, se_read = s.e.))
 
-selected_cnts <-
-  left_join(sample_size, high_low_sample) %>%
-  left_join(ses_sample) %>%
-  arrange(year, country)
-
-@
-
-<<eval = F>>=
-# In this chunk you can join reading and math datasets
-descrip_math <- map(results_math, ~ rename(.x, mean_math = Mean, se_math = s.e.))
-descrip_read <- map(results_read, ~ rename(.x, mean_read = Mean, se_read = s.e.))
-
-map2(results_math, results_read, inner_join, by = "country")
-@
-
-<<ci_for_difference>>=
-reduced_data <-
-    map2(results_math, years, function(.x, .y) {
+reduced_data_math <-
+  map2(descrip_math, years, function(.x, .y) {
     .x %>%
@@
   bind_rows() %>%
-    mutate(lower = Mean - 1.96 * s.e.,
-           upper = Mean + 1.96 * s.e.)
-@
+  as_tibble() %>%
+  mutate(lower_math = mean_math - 1.96 * se_math,
+         upper_math = mean_math + 1.96 * se_math)
 
-<<graphin differences, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+reduced_data_read <-
+  map2(descrip_read, years, function(.x, .y) {
+    .x %>%
+      mutate(wave = .y) %>%
+      filter(!is.na(escs_dummy))
+  }) %>%
+  bind_rows() %>%
+  as_tibble() %>%
+  mutate(lower_read = mean_read - 1.96 * se_read,
+         upper_read = mean_read + 1.96 * se_read)
+
+reduced_data <- left_join(reduced_data_math,
+                          reduced_data_read, by = c("country", "escs_dummy", "wave"))
+
+# Merging math and reading data
+test_data <-
   reduced_data %>%
-    filter(country %in% c("United Kingdom")) %>%
-    ggplot(aes(as.character(wave), Mean, group = escs_dummy, colour = escs_dummy)) +
-    geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
-    geom_line() +
-    facet_wrap(~ country)
-@
+  select(country, wave, escs_dummy, contains("mean")) %>%
+  gather(test, score, contains("mean"))
 
-<<>>=
-# Calculate the joint standard error of the different
-  se_data <-
+math_data <-
   reduced_data %>%
-    select(country, escs_dummy, wave, s.e.) %>%
-    spread(escs_dummy, s.e.) %>%
+  select(country, wave, escs_dummy, contains("math")) %>%
+  gather(test_bound, bound, contains("lower"), contains("upper")) %>%
+  select(-contains("math")) %>%
+  right_join(filter(test_data, test == "mean_math"))
+
+read_data <-
+  reduced_data %>%
+  select(country, wave, escs_dummy, contains("read")) %>%
+  gather(test_bound, bound, contains("lower"), contains("upper")) %>%
+  select(-contains("read")) %>%
+  right_join(filter(test_data, test == "mean_read"))
+
+all_data <- bind_rows(math_data, read_data)
+
+# Calculate the joint standard error of the difference
+math_se_data <-
+  reduced_data %>%
+  select(country, escs_dummy, wave, se_math) %>%
+  spread(escs_dummy, se_math) %>%
     transmute(country, wave,
-              se_diff = sqrt(abs(`1`^2 - `0`^2)))
+              se_diff_math = sqrt(abs(`1`^2 - `0`^2)))
 
-@
+read_se_data <-
+  reduced_data %>%
+  select(country, escs_dummy, wave, se_read) %>%
+  spread(escs_dummy, se_read) %>%
+  transmute(country, wave,
+            se_diff_read = sqrt(abs(`1`^2 - `0`^2)))
 
-<<include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+se_data <- left_join(math_se_data, read_se_data)
 
@@
 
-countries <- c("Chile", "Austria", "Belgium", "Canada", "Czech Republic", "Denmark",
-                 "Finland", "France", "Germany", "Italy", "Japan", "Netherlands", "Norway",
-                 "Poland", "Spain", "Sweden", "United Kingdom", "United States")
+math_diff <-
+  reduced_data %>%
+  select(wave, country, escs_dummy, mean_math) %>%
+  spread(escs_dummy, mean_math) %>%
+  transmute(wave, country, diff_math = `1` - `0`)
 
+read_diff <-
   reduced_data %>%
-    select(country, escs_dummy, Mean, wave) %>%
-    spread(key = escs_dummy, value = Mean) %>%
-    mutate(diff = `1` - `0`) %>%
-    gather(escs_dummy, Mean, -country, -wave, -`1`, -`0`) %>%
+  select(wave, country, escs_dummy, mean_read) %>%
+  spread(escs_dummy, mean_read) %>%
+  transmute(wave, country, diff_read = `1` - `0`)
+
+data_summaries <-
+  math_diff %>%
+  left_join(read_diff) %>%
   left_join(se_data) %>%
-    mutate(continent = countrycode(country, "country.name", "continent"),
-           region = countrycode(country, "country.name", "region"),
-           lower = Mean - 1.96 * se_diff,
-           upper = Mean + 1.96 * se_diff) %>%
-    filter(!is.na(continent), !is.na(region), country %in% vals) %>%
-    ggplot(aes(as.factor(wave), Mean, group = country, colour = country)) +
+  transmute(wave, country, diff_math, diff_read,
+           lower_math = diff_math - 1.96 * se_diff_math,
+           lower_read = diff_read - 1.96 * se_diff_read,
+           upper_math = diff_math + 1.96 * se_diff_math,
+           upper_read = diff_read + 1.96 * se_diff_read)
+
+differences <-
+  data_summaries %>%
+  select(wave, country, diff_math, diff_read) %>%
+  gather(test, difference, starts_with("diff")) %>%
+  mutate(type_test = ifelse(.$test == "diff_math", "math", "read"))
+
+bounds_lower <-
+  data_summaries %>%
+  select(wave, country, contains("lower")) %>%
+  gather(lower_bound, lower, lower_math, lower_read) %>%
+  mutate(type_test = ifelse(grepl("math", .$lower_bound), "math", "read"))
+
+bounds_upper <-
+  data_summaries %>%
+  select(wave, country, contains("upper")) %>%
+  gather(upper_bound, upper, upper_math, upper_read) %>%
+  mutate(type_test = ifelse(grepl("math", .$upper_bound), "math", "read"))
+
+# Getting the original data in
+original_math <-
+  reduced_data_math %>%
+  select(wave, everything(), -se_math) %>%
+  gather(metric, value, -(wave:escs_dummy)) %>%
+  unite(combination, escs_dummy, metric, sep = "_") %>%
+  spread(combination, value) %>%
+  mutate(type_test = "math")
+
+original_read <-
+  reduced_data_read %>%
+  select(wave, everything(), -se_read) %>%
+  gather(metric, value, -(wave:escs_dummy)) %>%
+  unite(combination, escs_dummy, metric, sep = "_") %>%
+  spread(combination, value) %>%
+  mutate(type_test = "read")
+
+# final data
+complete_data <-
+  left_join(differences, bounds_lower) %>%
+  left_join(bounds_upper) %>%
+  left_join(original_math) %>%
+  left_join(original_read)
+}
+
+complete_data_topbottom <- pisa_preparer(results_math, results_read)
+complete_data_topmid <- pisa_preparer(results_math_topmid, results_read_topmid)
+complete_data_midbottom <- pisa_preparer(results_math_midbottom, results_read_midbottom)
+
+complete_data_topbottom <- mutate(complete_data_topbottom, type = "90th/10th SES gap")
+complete_data_topmid <- mutate(complete_data_topmid, type = "90th/50th SES gap")
+complete_data_midbottom <- mutate(complete_data_midbottom, type = "50th/10th SES gap")
+@
+
+<<correlation_incomeineq, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+complete_data_topbottom %>%
+  mutate(wave = as.character(wave)) %>%
+  left_join(inequalityintsvy::economic_inequality, by = c("wave" = "year", "country")) %>%
+  filter(indicators == "GINI") %>%
+  group_by(country) %>%
+  summarize(avg_diff = mean(difference, na.rm = T),
+            avg_value = mean(value, na.rm = T)) %>%
+  filter(avg_value < 8) %>%
+  ggplot(aes(avg_value, avg_diff)) +
+  geom_point() +
+  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), linetype = "longdash", se = F)
+@
+
+<<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+
+# 90/10 gaps acros countries
+complete_data_topbottom %>%
+  filter(country %in% c("United States", "Netherlands", "France",
+                        "Germany", "Poland", "Finland")) %>%
+  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
   geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
+  geom_hline(yintercept = 0, linetype = "longdash") +
   geom_line() +
   geom_point(size = 0.5) +
-    scale_colour_discrete(guide = F) +
-    # geom_smooth(aes(group = region), colour = "blue", method = "lm") +
-    coord_cartesian(ylim = c(0, 4)) +
-    facet_wrap(~ country) +
-    ggtitle("math")
+  coord_cartesian(ylim = c(-0.5, 3)) +
+  facet_wrap(~ country)
+@
+
+<<graphing_allgaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+
+# Comparing all gaps across countries
+complete_data_topbottom %>%
+  bind_rows(complete_data_topmid) %>%
+  bind_rows(complete_data_midbottom) %>%
+  filter(country %in% c("United States", "Denmark", "France")) %>%
+  mutate(type = factor(type,
+                       levels = c("90th/10th SES gap", "90th/50th SES gap", "50th/10th SES gap"),
+                       ordered = TRUE)) %>%
+  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
+  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
+  geom_hline(yintercept = 0, linetype = "longdash") +
+  geom_line() +
+  geom_point(size = 0.5) +
+  coord_cartesian(ylim = c(-0.5, 3)) +
+  facet_grid(country ~ type)
+@
+
+<<graphing_ses_growth, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+# Graphing how the top/bottom are evolving over time instead of absolute difference
+
+complete_data_topbottom %>%
+  bind_rows(complete_data_topmid) %>%
+  bind_rows(complete_data_midbottom) %>%
+  filter(country %in% c("Germany", "Denmark", "France"), type_test == "math") %>%
+  select(wave, country, type_test, type, contains("math")) %>%
+  mutate(type = factor(type,
+                       levels = c("90th/10th SES gap", "90th/50th SES gap", "50th/10th SES gap"),
+                       ordered = TRUE)) %>%
+  gather(score, value, -(wave:type)) %>%
+  separate(score, c("ses", "score"), sep = 2) %>%
+  spread(score, value) %>%
+  ggplot(aes(as.factor(wave), mean_math, group = ses, colour = ses)) +
+  geom_errorbar(aes(ymin = lower_math, ymax = upper_math), width = 0.1) +
+  geom_hline(yintercept = 0, linetype = "longdash") +
+  geom_line() +
+  geom_point(size = 0.5) +
+  coord_cartesian(ylim = c(-0.5, 3)) +
+  facet_grid(country ~ type)
 
@@
 # Spain - no change
-
 @
 
-<<>>=
+% How to output data frames are latex tables?
+<<rate_change>>=
+
+avg_increase_fun <- function(df, class) {
+
+# Average standard deviation increase
+  data_ready <-
+    df %>%
+    select(wave, country, type_test, contains("mean_math")) %>%
+    gather(metric, value, -(wave:type_test)) %>%
+    separate(metric, c("ses", "test"), sep = 2) %>%
+    spread(test, value) %>%
+    mutate(ses = gsub("_", "", ses)) %>%
+    filter(type_test == "math", ses == class) %>%
+    split(.$country) %>%
+    map(~ mutate(.,
+                 diff = c(diff(mean_math, lag = 1), NA),
+                 perc = round(diff / mean_math, 2) * 100,
+                 perc_pos = mean(perc > 0, na.rm = T))) %>%
+    enframe() %>%
+    unnest(value) %>%
+    split(.$country)
+
+    map2(data_ready, names(data_ready), ~ {
+
+      print(.y)
+
+      mean_df <-
+        bootstrapper(.x, mean(diff, na.rm = T), B = 500) %>%
+        filter(type == "bootstrap") %>%
+        rename(mean = value)
+
+      sd_df <-
+        bootstrapper(.x, sd(diff, na.rm = T), B = 500) %>%
+        filter(type == "bootstrap") %>%
+        rename(sd = value)
+
+      suppressMessages(
+        left_join(mean_df, sd_df) %>%
+        mutate(lower_bound = mean - 1 * sd,
+               upper_bound = mean + 1 * sd)
+      )
+    }) %>%
+    enframe() %>%
+    unnest(value)
+}
+
+avg_sd_increase_high <- avg_increase_fun(complete_data_topbottom, 1)
+avg_sd_increase_low <- avg_increase_fun(complete_data_topbottom, 0)
+@
+
+<<rate_change_graph, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+
+full_data <-
+  left_join(avg_sd_increase_high, avg_sd_increase_low, by = "country") %>%
+  mutate(continent = ifelse(country %in% countries, "my_cnt", "other_cnt"))
+
+colnames(full_data) <- c("country", "high_increase", "low_increase", "continent")
+
+lims <- list(xlim = c(-0.15, 0.25), ylim = c(-0.25, 0.25))
+
+rect_data <- tibble(xst = c(lims$xlim[1], 0),
+                    xen = c(0.0, lims$xlim[2]),
+                    yst = c(0.0, lims$ylim[1]),
+                    yen = c(lims$ylim[2], 0),
+                    colour = c("red", "green"))
+
+full_data %>%
+  ggplot(aes(low_increase, high_increase), alpha = 0.2) +
+  geom_rect(data = rect_data, aes(xmin = xst,
+                                  xmax = xen,
+                                  ymin = yst,
+                                  ymax = yen),
+            fill = rect_data$colour,
+            alpha = 0.2,
+            inherit.aes = FALSE) +
+  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1) +
+  geom_point(alpha = 0.2) +
+  geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
+  geom_text_repel(data = filter(full_data, continent == "my_cnt"),
+                  aes(label = country), box.padding = unit(2.7, "lines")) +
+  geom_vline(xintercept = 0, alpha = 0.5) +
+  geom_hline(yintercept = 0, alpha = 0.5) +
+  xlim(lims$xlim) +
+  ylim(lims$ylim) +
+  coord_cartesian(expand = FALSE) +
+  annotate(geom = "text", x = 0.15, y = -0.2,
+           label = "Low SES are catching up \n faster than High SES",
+           fontface = 2, size = 3) +
+  annotate(geom = "text", x = -0.05, y = 0.20,
+           label = "High SES are increasing  \n faster than Low SES",
+           fontface = 2, size = 3) +
+  labs(x = "Average increase of low SES in SD", y = "Average increase of high SES in SD") +
+  theme_minimal()
+@
+
+% How to output data frames are latex tables?
+<<perc_increase_tables>>=
+
+# Show the rates at which is increasing/decreasing
+perc_increase_fun <- function(df) {
+
+  # Average standard deviation increase
+  data_ready <-
+    df %>%
+    select(wave, country, type_test, difference) %>%
+    group_by(type_test) %>%
+    split(.$country) %>%
+    map(~ {
+      .x <-
+        spread(.x, wave, difference) %>%
+        ungroup()
+
+      year_vars <- sum(map_dbl(.x, is.numeric)) - 1
+      years_subtract <- names(.x)[c(ncol(.x) - year_vars, ncol(.x))]
+      years_subtract <- lapply(years_subtract, as.name)
+
+      last_year <- rlang::new_quosure(years_subtract[[2]], env = .GlobalEnv)
+      first_year <- rlang::new_quosure(years_subtract[[1]], env = .GlobalEnv)
+
+      years_available <-
+        .x %>%
+        gather(year, val, -(country:type_test)) %>%
+        group_by(type_test) %>%
+        summarise(yr_avaible = sum(!is.na(val))) %>%
+        pull(yr_avaible)
+
+      year_sd <-
+        .x %>%
+        gather(year, val, -(country:type_test)) %>%
+        split(.$type_test) %>%
+        map_dbl(~ bootstrapper(.x, mad(val, na.rm = T), B = 100) %>% .[[2, 2]]) %>%
+        round(2) * 100
+
+      .x %>%
+        transmute(type_test,
+                  country,
+                  perc_diff = round(((!!last_year) - (!!first_year)) / (!!first_year) * 100, 1),
+                  sd_year = year_sd,
+                  diff_lower = perc_diff - 1 * year_sd,
+                  diff_upper = perc_diff + 1 * year_sd,
+                  years_available = years_available)
+    })
+  data_ready
+}
+
+top_bottom_perc <- perc_increase_fun(complete_data_topbottom)
+top_mid_perc <- perc_increase_fun(complete_data_topmid)
+mid_bottom_perc <- perc_increase_fun(complete_data_midbottom)
+
+# Gap is closing at an average of the variable diff per year.
+@
+
+<<next_steps>>=
 # Next steps:
-
-# Show country patterns with reading and math together into one graph
-
-# Making country groups of the graphs showing that some countries are increasing, others are steady
-# While others are decreasing. Do this with a few countries, but end by showing all countries.
-
-# Continue by studying what's happening in those countries where it's increasing/decreasing
-# In the gap graph
 
@@
 
-# The gap between the top 90th and the 50th, did it grow?
-# The gap between the top 50th and the 10th, did it grow?
-
 # A weak welfare system, together with income inequality, what's their pattern?
@@
 
-# Show that within top and bottom educated categories, there is reasonable variation of SES index (to justify that your approach of suing the top and bottom percentiles within each category is the right approach)
-
 # In countries where there is high differentiation/tracking, is there a jump in the evolution of the gap between PIRLS/TIMSS and PISA?
-
 @

### 2017-06-16 paper.R

@@ -49,6 +49,7 @@
 
 
 
+  
 ## ----loading_data-recoding-----------------------------------------------
 pisa_all <- read_rds("./data/pisa_listcol.Rdata")
 pisa_all2 <- pisa_all
@@ -90,6 +91,7 @@ reliability_pisa <-
     "2009" = 0.74,
     "2012" = 0.82,
     "2015" = 0.74) # 2015 imputed
+
 
 
 
@@ -376,7 +378,7 @@ sample_tables_midbottom <- sample_size_calc(adapted_year_data, c(.1, .5), select
 
 ## ----merge_math_read, echo = F------------------------------------------------------------
 
-# Function does a lot of things, but in short,
+# Function does a lot of things, but in short:
 
 # Calculate the difference between the gap and together with it's joint s.e
 # Also uncertainty intervals and returns a tibble with the difference between

### 2017-06-16 paper.R
Deleted separate code file - all code is in the paper

@@ -1,819 +1,0 @@
-## ----default_conf, echo = F---------------------------------------------
-  library(knitr)
-  library(arm)
-  library(saves)
-  library(haven)
-  library(PISA2000lite)
-  library(PISA2003lite)
-  library(PISA2006lite)
-  library(PISA2009lite)
-  library(PISA2012lite)
-  library(intsvy)
-  library(cimentadaj) # For pisa_countrynames
-  library(countrycode) # For region variable
-  library(car)
-  library(readr)
-  library(SAScii)
-  library(inequalityintsvy)
-  library(lme4)
-  library(modelr)
-  library(tidyverse)
-  library(ggrepel)
-  
-  # source("./transform_data.R")
-
-  # Conf for PISA_2015
-  pisa2015_conf <- list(variables = list(pvlabelpref = "PV",
-                                         pvlabelsuff = "READ",
-                                         weightFinal = "W_FSTUWT",
-                                         weightBRR = "W_FSTURWT"),
-          parameters = list(cutoffs = c(357.77, 420.07, 482.38, 544.68, 606.99, 669.30),
-                                          percentiles = c(5, 10, 25, 75, 90, 95),
-                                          PVreps = 10,
-                                          BRRreps = 80,
-                                          weights = "BRR",
-                                          replication_scheme = 'pisa')
-  )
-  
-  countries <- c("Australia",
-                 "Germany",
-                 "Denmark",
-                 "Spain",
-                 "France",
-                 "Italy",
-                 "Netherlands",
-                 "Sweden",
-                 "Finland",
-                 "United States",
-                 "United Kingdom")
-
-
-
-  
-## ----loading_data-recoding-----------------------------------------------
-pisa_all <- read_rds("./data/pisa_listcol.Rdata")
-pisa_all2 <- pisa_all
-
-years <- seq(2000, 2015, 3)
-  
-db <- paste0("pisa", years)
-pisa_all2$value <- map2(pisa_all2$value, db, ~ { .x$wave <- .y; .x})
-pisa_all2$value[[1]]$CNT <- pisa_all2$value[[1]]$COUNTRY
-  
-pisa_all2$value <- map(pisa_all2$value, ~ {
-  
-# 2000 to 2015
-# The coding is from 0 to 6, where 0 is no schooling and 6 is
-# BA or above.
-
[…]

### 2017-06-17 paper/paper.Rnw
Updated paper code - ran paper and everything worked

@@
 <<working directory, echo = F>>=
-opts_knit$set(root.dir = '..')
+# opts_knit$set(root.dir = '..')
 
@@
 
-<<modeling>>=
+<<modeling, cache = TRUE>>=
 adapted_year_data <-
@@
 
-# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH")
-# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ")
-# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
-# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
-# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
-# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
+results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
+results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
+results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
+results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
+results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
+results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
 
-results_math <- read_rds("./data/delete.Rdata")
-results_read <- read_rds("./data/delete_read.Rdata")
-results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
-results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
-results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
-results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
+# results_math <- read_rds("./data/delete.Rdata")
+# results_read <- read_rds("./data/delete_read.Rdata")
+# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
+# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
+# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
+# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
 # US is missing for reading
@@
 % Figure out how to ouput data frames as latex tables.
-<<sample_size>>=
+<<sample_size, cache = TRUE>>=
 
@@
 
-<<merge_math_read>>=
+<<merge_math_read, cache = TRUE>>=
 
@@
 % How to output data frames are latex tables?
-<<rate_change>>=
+<<rate_change, cache = TRUE>>=
 
@@
 full_data <-
-  left_join(avg_sd_increase_high, avg_sd_increase_low, by = "country") %>%
-  mutate(continent = ifelse(country %in% countries, "my_cnt", "other_cnt"))
+  left_join(select(avg_sd_increase_high, name, mean),
+            select(avg_sd_increase_low, name, mean), by = "name") %>%
+  mutate(continent = ifelse(name %in% countries, "my_cnt", "other_cnt"))
 
@@
 % How to output data frames are latex tables?
-<<perc_increase_tables>>=
+<<perc_increase_tables, cache = TRUE>>=
 

### 2017-06-17 paper/paper.tex
Updated paper code - ran paper and everything worked

@@
 
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphin_differences-1}
+% This chunk needs to have cache = TRUE once you want to run it for like 2 hours until 
+% the results_* models have been created once and the cache can save them
+
+% Figure out how to ouput data frames as latex tables.
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/correlation_incomeineq-1}
 
@@
 
-{\centering \includegraphics[width=5.5in,height=5in]{figure/unnamed-chunk-2-1}
+{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_9010gaps-1}
 
 }
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_allgaps-1}
+
+}
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_ses_growth-1}
+
+}
+
+% How to output data frames are latex tables?
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/rate_change_graph-1}
+
+}
+
+% How to output data frames are latex tables?
 

### 2017-06-19 paper/paper.Rnw
Wrote a bit of literature review + removed changes that belonged to timeseries branch

@@
 
+\section(Literature Review)
+
 Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years.
-The vast literature on educational inequality has mainly concentrated on the United States (reardon here).
-First, because it is the only country where cognitive testing is very widespread across surveys.
+The literature on educational inequality has mainly concentrated on the United States (reardon here) but other international evidence is emerging with other similar findings for other countries.
+Despite this, the foremost and most important study on the subject is Reardon (2011) with American data.
+This is the cases, firstly, because it is the only country where cognitive testing is very widespread across surveys.
 And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day.
@@
 Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap).
-
 The widening of the achievement gap has been happening in parallel to the growth of income inequality.
 Although very suggestive, it is hard to link both things causally.
-The downside of this research study is that it is mainly US-based, and not a lot of evidence is available in an international context.
-To the best of my knowledge, the only attempt to take this to an international context is the work by Chmielewski and Reardon (2016).
-They find that there is considerable variation in the achievement gap between top and bottom earning families.
-However, their design is quite different from the one from Reardon (2011).
+
+Many authors have taken this analysis to an international context in order to discover between-country trends.
+The comprehensive work of Corak and Waldfogel () shows studies, Australia, United Kingdom, United States and Canada.
+Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is the only study that uses panel data to study achievement gaps, let alone to do this between country}.
+The core finding behind the book is that the American achievement gap is much wider than in any other country.
+More specifically, the American achievement gap has been seen to be around ~ 1.25/1.50 standard deviations and doesn't seem to change much across the lifetime.
+That is, they find that the achievement gap is very stable all the way from early education to secondary school (although their studies cover up until age 14).
+Despite their rigurosity, the four surveys have significant differences and cannot be readily compared.
+This makes their results more suggestive rather than definitive.
+Which is why authors such as Chmielewski and Reardon (2016) and Chmielewski (2017) have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries.
+Chmielewski and Reardon (2016), using the Programme for International Student Assessment (PISA), the Trends in International Mathematics and Science Study (TIMSS) and PIRLS (the Progress in International Reading Literacy Study), the assess whether there are differences in the achievement gap between countries.
+Although there is potential to study countries across years using these studies, given their researchn design it's impossible and only 3 countries have enough data points to be studied longitudinally.
+None the less, they find that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
+They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, etc ...
+The correlations are indeed very suggestive but clearly we must be extremely cautios in drawing causal changes from these two variables.
+
+Their design is quite different from the work of Reardon (2011).
 Reardon (2011) takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds.
@@
 Chmielewski and Reardon (2016) use the international surveys PISA and PIRLS to compare the achievement gap across countries.
-Unfortunately, very few countries can be compared across time in PISA.
+
+Chmielewski (2017), building on the work of Chmielewski and Reardon (2016) and Reardon and Portilla (2015) pooled together all the previous data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries.
+With over 50 years of data, and over 100 countries, Chmielewski (2017) finds that there seems to be a general world increase in the achievement gap.
+However, once she disentangles the relationship by country, she finds a reasonable amount of heterogeneity, with some countries seeing the achievement gap closing, others no change at all, while others record a steady increase.
+One clear limitation of their study (as well as Reardon (2011)) is that the adjust for the age of each child in all studies.
+Although for their purposes is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
+However, as we've noted before, Corak and Waldfogel (2015) find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as Reardon (2011) did.
+
+The evolution of High/Low SES gaps for preschool children might be much less marked than the same gap for high school children.
+The explanation, although very debated, has been gaining much support in recent years.
+In countries with high levels of curricular differentiation, the transition from early schooling into the tracking system has been found to increase inequality of learning (Wossman and Hanushek diff-in-diff).
+Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs (Van der Werfhorst and Mijs 2011).
+Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as elimination tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
 
@@
 
+\section(Research design)
+
 \end{document}

### 2017-06-19 paper/paper.Rnw
Added next-to write

@@
 
-\section(Literature Review)
+\section{Literature Review}
 
@@
 The correlations are indeed very suggestive but clearly we must be extremely cautios in drawing causal changes from these two variables.
-
-Their design is quite different from the work of Reardon (2011).
+But it is important to stress that their design is quite different from the work of Reardon (2011).
 Reardon (2011) takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds.
 Although the author adjusts for age, the trends can't be generalized to age-specific, but rather in overall terms.} and makes them comparable across time.
-That is, he builds a trend of achievement gap.
-Chmielewski and Reardon (2016) use the international surveys PISA and PIRLS to compare the achievement gap across countries.
+This gives the author a very long time series to build a reliable achievement gap (over 40 years).
+In their study, Chmielewski and Reardon (2016) change the aim of the study to model between-country differences from a cross-sectional perspective.
 
-Chmielewski (2017), building on the work of Chmielewski and Reardon (2016) and Reardon and Portilla (2015) pooled together all the previous data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries.
+Chmielewski (2017), building on the work of Chmielewski and Reardon (2016) and Reardon and Portilla (2015) pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries.
 With over 50 years of data, and over 100 countries, Chmielewski (2017) finds that there seems to be a general world increase in the achievement gap.
@@
 
-This paper introduces one novelty in the literature which is to evaluate the evolution of the High/Low SES achievement gap in the past 15 years for all PISA participant countries.
+With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the High/Low SES achievement gap in the past 15 years for all PISA participant countries.
 This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for each country.
+The advantages of this study are numerous.
+First, we concentrate on the evolution of the gap for only 15 year olds.
+This is different from all the work of Chmielewski and Reardon, in which they pool all ages to get average changes, to study the properties of specific age-groups.
+As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
+Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.
 
-Do we need to distinguish between age-group achievement gaps?
+Talk about reardon and portilla and how the achievement gap has been closing in the last 15 years.
+Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
+Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
+Talk more about how countries with high social mobility has been linked to smallest achievement gaps
+Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.
 
-\section(Research design)
+\section{Research design}
 

### 2017-06-19 paper/paper.tex
Added next-to write

@@
 Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap).
-
 The widening of the achievement gap has been happening in parallel to the growth of income inequality.
 Although very suggestive, it is hard to link both things causally.
-The downside of this research study is that it is mainly US-based, and not a lot of evidence is available in an international context.
-To the best of my knowledge, the only attempt to take this to an international context is the work by Chmielewski and Reardon (2016).
-They find that there is considerable variation in the achievement gap between top and bottom earning families.
-However, their design is quite different from the one from Reardon (2011).
+
+Many authors have taken this analysis to an international context in order to discover between-country trends.
+The comprehensive work of Corak and Waldfogel () shows studies, Australia, United Kingdom, United States and Canada.
+Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is the only study that uses panel data to study achievement gaps, let alone to do this between country}.
+The core finding behind the book is that the American achievement gap is much wider than in any other country.
+More specifically, the American achievement gap has been seen to be around ~ 1.25/1.50 standard deviations and doesn't seem to change much across the lifetime.
+That is, they find that the achievement gap is very stable all the way from early education to secondary school (although their studies cover up until age 14).
+Despite their rigurosity, the four surveys have significant differences and cannot be readily compared.
+This makes their results more suggestive rather than definitive.
+Which is why authors such as Chmielewski and Reardon (2016) and Chmielewski (2017) have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries.
+Chmielewski and Reardon (2016), using the Programme for International Student Assessment (PISA), the Trends in International Mathematics and Science Study (TIMSS) and PIRLS (the Progress in International Reading Literacy Study), the assess whether there are differences in the achievement gap between countries.
+Although there is potential to study countries across years using these studies, given their researchn design it's impossible and only 3 countries have enough data points to be studied longitudinally.
+None the less, they find that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
+They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, etc ...
+The correlations are indeed very suggestive but clearly we must be extremely cautios in drawing causal changes from these two variables.
+But it is important to stress that their design is quite different from the work of Reardon (2011).
 Reardon (2011) takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds.
 Although the author adjusts for age, the trends can't be generalized to age-specific, but rather in overall terms.} and makes them comparable across time.
-That is, he builds a trend of achievement gap.
-Chmielewski and Reardon (2016) use the international surveys PISA and PIRLS to compare the achievement gap across countries.
-Unfortunately, very few countries can be compared across time in PISA.
+This gives the author a very long time series to build a reliable achievement gap (over 40 years).
+In their study, Chmielewski and Reardon (2016) change the aim of the study to model between-country differences from a cross-sectional perspective.
 
-This paper introduces one novelty in the literature which is to evaluate the evolution of the High/Low SES achievement gap in the past 15 years for all PISA participant countries.
+Chmielewski (2017), building on the work of Chmielewski and Reardon (2016) and Reardon and Portilla (2015) pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries.
+With over 50 years of data, and over 100 countries, Chmielewski (2017) finds that there seems to be a general world increase in the achievement gap.
+However, once she disentangles the relationship by country, she finds a reasonable amount of heterogeneity, with some countries seeing the achievement gap closing, others no change at all, while others record a steady increase.
+One clear limitation of their study (as well as Reardon (2011)) is that the adjust for the age of each child in all studies.
+Although for their purposes is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
+However, as we've noted before, Corak and Waldfogel (2015) find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as Reardon (2011) did.
+
+The evolution of High/Low SES gaps for preschool children might be much less marked than the same gap for high school children.
+The explanation, although very debated, has been gaining much support in recent years.
+In countries with high levels of curricular differentiation, the transition from early schooling into the tracking system has been found to increase inequality of learning (Wossman and Hanushek diff-in-diff).
+Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs (Van der Werfhorst and Mijs 2011).
+Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as elimination tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
+
+With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the High/Low SES achievement gap in the past 15 years for all PISA participant countries.
 This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for each country.
+The advantages of this study are numerous.
+First, we concentrate on the evolution of the gap for only 15 year olds.
+This is different from all the work of Chmielewski and Reardon, in which they pool all ages to get average changes, to study the properties of specific age-groups.
+As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
+Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.
 
-Do we need to distinguish between age-group achievement gaps?
+Talk about reardon and portilla and how the achievement gap has been closing in the last 15 years.
+Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
+Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
+Talk more about how countries with high social mobility has been linked to smallest achievement gaps
+
+\section(Research design)
 

### 2017-06-19 paper/paper.Rnw
Implemented graphs for % increase/decrease + sample size tables

@@
 \tableofcontents
-
-\section{Analysis}
 
@@
 
+\section{Descriptives}
+
 % Figure out how to ouput data frames as latex tables.
@@
 
+\section{Analysis}
+
 <<correlation_incomeineq, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
@@
 <<perc_increase_tables, cache = TRUE>>=
-
+df <- split(
+  group_by(complete_data_topmid, type_test) %>% select(wave, country, type_test, difference), complete_data_topmid$country)
+.x <- df$`United States`
 # Show the rates at which is increasing/decreasing
@@
       .x %>%
+        map_if(is_double, round, 3) %>%
+        as_tibble() %>%
         transmute(type_test,
                   country,
-                  perc_diff = round(((!!last_year) - (!!first_year)) / (!!first_year) * 100, 1),
+                  perc_diff = round(((!!last_year) - (!!first_year)) * 100, 1),
                   sd_year = year_sd,
@@
 # Gap is closing at an average of the variable diff per year.
+@
+
+<<graphing_perc_increase, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+perc_graph <- function(df) {
+  df %>%
+  enframe(name = "country") %>%
+  unnest(value) %>%
+  filter(country %in% countries) %>%
+  select(-country1, -years_available) %>%
+  mutate(diff_95_lower = perc_diff - 2*sd_year,
+         diff_95_upper = perc_diff + 2*sd_year) %>%
+  setNames(c("Country", "Type of test", "Average % difference", "Average SD",
+             "Lower 50% bound", "Upper 50% bound", "Lower 95% bound", "Upper 95% bound")) %>%
+  filter(`Type of test` == "math") %>%
+  arrange(`Average % difference`) %>%
+  mutate(Country = ordered(forcats::as_factor(Country)),
+         Country_num = as.numeric(Country)) %>%
+  ggplot(aes(Country, `Average % difference`)) +
+  geom_point() +
+  geom_ribbon(aes(x = Country_num, ymin = `Lower 95% bound`, ymax = `Upper 95% bound`), alpha = 0.5) +
+  geom_ribbon(aes(x = Country_num, ymin = `Lower 50% bound`, ymax = `Upper 50% bound`), alpha = 0.5) +
+  geom_hline(yintercept = 0, linetype = "longdash") +
+  scale_y_continuous(breaks = seq(-160, 160, 40)) +
+  coord_cartesian(ylim = c(-160, 160)) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+}
+
+m1 <- perc_graph(top_bottom_perc)
+m2 <- perc_graph(top_mid_perc)
+m3 <- perc_graph(mid_bottom_perc)
+
+gridExtra::grid.arrange(m1, m2, m3, nrow = 1)
+@
+
+<<table_sample>>=
+sample_tables_topbottom %>%
+  filter(name %in% c(2000, 2015)) %>%
+  transmute(name, country, escs_dummy, N = paste0(n, " (", perc, ")")) %>%
+  spread(escs_dummy, N) %>%
+  rename(Year = name,
+         Country = country,
+         `10th SES` = `0`,
+         `90th SES` = `1`) %>%
+  select(Country, Year, everything()) %>%
+  arrange(Country, Year) %>%
+  xtable::xtable()
 @

### 2017-06-20 paper/paper.Rnw
Continue writing - literature review

@@
 Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years.
-The literature on educational inequality has mainly concentrated on the United States (reardon here) but other international evidence is emerging with other similar findings for other countries.
-Despite this, the foremost and most important study on the subject is Reardon (2011) with American data.
-This is the cases, firstly, because it is the only country where cognitive testing is very widespread across surveys.
+The literature on educational inequality has mainly concentrated on the United States (reardon here) but other international evidence is emerging with other similar findings for other count.
+The United States is usually the case study of interest, firstly, because it is the only country where cognitive testing is very widespread across surveys.
 And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day.
@@
 
-Many authors have taken this analysis to an international context in order to discover between-country trends.
-The comprehensive work of Corak and Waldfogel () shows studies, Australia, United Kingdom, United States and Canada.
-Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is the only study that uses panel data to study achievement gaps, let alone to do this between country}.
-The core finding behind the book is that the American achievement gap is much wider than in any other country.
-More specifically, the American achievement gap has been seen to be around ~ 1.25/1.50 standard deviations and doesn't seem to change much across the lifetime.
-That is, they find that the achievement gap is very stable all the way from early education to secondary school (although their studies cover up until age 14).
-Despite their rigurosity, the four surveys have significant differences and cannot be readily compared.
-This makes their results more suggestive rather than definitive.
-Which is why authors such as Chmielewski and Reardon (2016) and Chmielewski (2017) have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries.
-Chmielewski and Reardon (2016), using the Programme for International Student Assessment (PISA), the Trends in International Mathematics and Science Study (TIMSS) and PIRLS (the Progress in International Reading Literacy Study), the assess whether there are differences in the achievement gap between countries.
-Although there is potential to study countries across years using these studies, given their researchn design it's impossible and only 3 countries have enough data points to be studied longitudinally.
-None the less, they find that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
+Reardon (2011) finds that the increase has occured predominantely from the 1970's until the 2000's.
+In fact, hard numbers suggest that the gap increased by about ~ 40-50\%.
+The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%.
+Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footenote{Even after weighting appropriately.}.
+Finally, Reardon (2011) concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.
+Corak and Waldfogel (2015) don't find results which are very far away from this.
+They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25.
+Duncan and Magnuson find similarly to the previous studies and confirms a gap of ~ 1.25/1.50 standard deviations.
+
+One important drawback of these studies is that they don't present the uncertainty of this estimate.
+Not necessarily to gauge their statistical significance, but to simply asses how much we can trust these numbers.
+
+With very recent data, Reardon and Portilla (2015) have unexpectedly uncovered a new finding: the reversal of the trend.
+From 1998 until 2010, it seems like the income achievement gap \footnote{The income achievement gap is the difference between the 90th and 10th income percentile in the test of interest} decreased at a rate of 0.01 and 0.008 standard deviations per year.
+In contrast to Reardon (2011), we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something not dramatically different from the previous estimates.
+The American situation is a very distinctive one relative to Europe.
+The reasons why we're seeing a reversal in the trend could be numerous and should be studied very close to context.
+For example, the American educational system lacks any formal curricular differentiation \emph{à la} European style.
+In contrast, Reardon and Portilla (2015) suggest that the reversal is likely due to the high increase of preschool enrollment.
+They suggest that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
+
+Some authors have taken this analysis to an international context in order to discover between-country trends.
+The comprehensive work of Corak and Waldfogel () performs a comparative analysis of Australia, United Kingdom, United States and Canada.
+Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is the only study that uses panel data to study achievement gaps, let alone to do this between countries}.
+The core finding behind the book is that the American achievement gap is much wider than in any other comparison countries.
+As mentioned earlier, they find that once the achievement gap is present in early school entry, it doesn't seem to narrow very much over the life course.
+They find that the achievement gap is very stable all the way from early education to secondary school (although their studies cover up until age 14).
+
+Say more above Corak and Waldfogel above (book).
+
+Despite their high level of rigurosity, their analysis is based on four surveys that have significant differences and cannot be easily compared.
+Their findings are very reliable but should be taken as suggestive.
+For this reason we should also pay particular attention to studies such as Chmielewski and Reardon (2016) and Chmielewski (2017) which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries.
+These studies tackle a completely different question from the above, but they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
+
+Chmielewski and Reardon (2016), using the Programme for International Student Assessment (PISA), the Trends in International Mathematics and Science Study (TIMSS) and PIRLS (the Progress in International Reading Literacy Study), the assess whether there are patterns of cross-national variation in the achievement gap.
+In other words, is the size of achievement gap different for many countries?
+Although there is potential to study countries over time using these surveys, they choose not to.
+They do so because their question of interest (income categories) was only asked in three waves; in the end the only had three countries repeated in all three waves.
+
+Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
+In comparison to the literature on achievement gaps, they find that the U.S has a gap a little over 1 standard deviation in the year 2005 (for 15 year olds) and Germany has a decreasing gap from around ~ 1.25 to ~ 1 standard deviation.
 They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, etc ...
@@
 
+Make sure PISA variation is constant over time
 \end{document}

### 2017-06-20 paper/paper.Rnw
Changed small thing in paper +  added all other paper docs

@@
 <<perc_increase_tables, cache = TRUE>>=
-df <- split(
-  group_by(complete_data_topmid, type_test) %>% select(wave, country, type_test, difference), complete_data_topmid$country)
-.x <- df$`United States`
 # Show the rates at which is increasing/decreasing
@@
 
+Large inequalities in parent's education leads to large inequalities in children's education (Breen & Jonsson 2007)
+
+The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th. (Micklewright & Schnepf, 2007)
+
+Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (Mickleweight & Schnepf did it).
+They also find that France is a lower dispersion country (but in my results this is changing a lot).
+They also find that Germany and US are very dispersed countries (quite the opposite)
+
+Mickleweight & Schnepf find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
+Mickleweight & Schnpef find
+that average achievement is negatively related to dispersion.
+
+Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
+
 Talk about reardon and portilla and how the achievement gap has been closing in the last 15 years.

### 2017-06-20 paper/paper.tex
Changed small thing in paper +  added all other paper docs

@@
 
-\section{Analysis}
-
 % This chunk needs to have cache = TRUE once you want to run it for like 2 hours until 
@@
 
+\section{Descriptives}
+
 % Figure out how to ouput data frames as latex tables.
+
+\section{Analysis}
 
@@
 
-\section(Literature Review)
+{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_perc_increase-1}
+
+}
+
+\section{Literature Review}
 
 Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years.
-The literature on educational inequality has mainly concentrated on the United States (reardon here) but other international evidence is emerging with other similar findings for other countries.
-Despite this, the foremost and most important study on the subject is Reardon (2011) with American data.
-This is the cases, firstly, because it is the only country where cognitive testing is very widespread across surveys.
+The literature on educational inequality has mainly concentrated on the United States (reardon here) but other international evidence is emerging with other similar findings for other count.
+The United States is usually the case study of interest, firstly, because it is the only country where cognitive testing is very widespread across surveys.
 And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day.
@@
 Talk more about how countries with high social mobility has been linked to smallest achievement gaps
+Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.
 
-\section(Research design)
+\section{Research design}
 
+Make sure PISA variation is constant over time
 \end{document}

### 2017-06-20 paper/paper.Rnw
First try - no changes
Plotted cor between dispersion and achievement - very high cor

@@
 # Spain - no change
+@
+
+<<>>=
+
+country_scores <-
+  map(adapted_year_data, ~ {
+  if (unique(.x$wave) == "pisa2015") {
+    intsvy::pisa2015.mean.pv("MATH", by = "country", data = .x)
+  } else {
+    intsvy::pisa.mean.pv("MATH", by = "country", data = .x)
+  }
+}) %>% setNames(seq(2000, 2015), 3) %>%
+
+country_scores <-
+  enframe(country_scores, name = "wave") %>%
+  unnest() %>%
+  mutate(wave = as.double(wave))
+
+data_to_plot <-
+  left_join(complete_data_topbottom, country_scores, by = c("wave", "country")) %>%
+  select(wave, country, type_test, difference, Mean) %>%
+  filter(type_test == "math")
+
+data_to_plot %>%
+  ggplot(aes(scale(Mean), difference)) +
+  geom_point() +
+  geom_smooth()
+
+data_to_plot %>%
+  filter(country %in% countries) %>%
+  ggplot(aes(as.character(wave), difference, group = country)) +
+  geom_point(colour = "blue") +
+  geom_line(colour = "blue") +
+  geom_point(aes(y = scale(Mean)), colour = "red") +
+  geom_line(aes(y = scale(Mean)), colour = "red") +
+  facet_wrap(~ country)
+
+data_to_plot %>%
+  semi_join(data_to_plot %>% group_by(country) %>% count() %>% filter(n > 3), by = "country") %>%
+  group_by(country) %>%
+  summarize(cor = cor(scale(Mean), difference, use = 'complete.obs')) %>%
+  ggplot(aes(reorder(country, cor), cor, fill = cor > 0)) + geom_col() + coord_flip()
+
 @
@@
 One important drawback of these studies is that they don't present the uncertainty of this estimate.
-Not necessarily to gauge their statistical significance, but to simply asses how much we can trust these numbers.
+Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
+Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education (Breen & Jonsson 2007).
+Virtually in all countries, be it developed or developing, there is inequality of opportunity.
+But there is considerable variation in the magnitud of this effect.
+For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries (Gosta and his phd student, Breen and other people, Shavit first study).
 
@@
 
-Say more above Corak and Waldfogel above (book).
-
-Despite their high level of rigurosity, their analysis is based on four surveys that have significant differences and cannot be easily compared.
-Their findings are very reliable but should be taken as suggestive.
-For this reason we should also pay particular attention to studies such as Chmielewski and Reardon (2016) and Chmielewski (2017) which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries.
+One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula.
+These four countries have no major jump in school selection, quite the opposite to the European average countries.
+A thorough review by Van der Werfhorst and Mijs (2010) sheds some light on the subject.
+First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps.
+Hanushek and Woessman (tracking study), use the TIMSS and PISA surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
+Moreover, Mickleweight & Schnepf (2007) using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
+In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
+
+Despite their high level of rigurosity, their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
+Their findings are very reliable but should be taken as suggestive at best.
+For this reason we should also pay particular attention to studies such as Chmielewski and Reardon (2016) and Chmielewski (2017) which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries using similar studies.
 These studies tackle a completely different question from the above, but they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.

### 2017-06-20 paper/paper.Rnw
Added PISA variance across years graph

@@
 
-\section{Descriptives}
+\section{Literature Review}
+
+Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years.
+The literature on educational inequality has mainly concentrated on the United States (reardon here) but other international evidence is emerging with other similar findings for other count.
+The United States is usually the case study of interest, firstly, because it is the only country where cognitive testing is very widespread across surveys.
+And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day.
+Using this information, Reardon (2011) is the first to investigate the evolution of the cognitive gap and the results are very surprising.
+Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap).
+The widening of the achievement gap has been happening in parallel to the growth of income inequality.
+Although very suggestive, it is hard to link both things causally.
+
+Reardon (2011) finds that the increase has occured predominantely from the 1970's until the 2000's.
+In fact, hard numbers suggest that the gap increased by about ~ 40-50\%.
+The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%.
+Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footenote{Even after weighting appropriately.}.
+Finally, Reardon (2011) concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.
+Corak and Waldfogel (2015) don't find results which are very far away from this.
+They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25.
+Duncan and Magnuson find similarly to the previous studies and confirms a gap of ~ 1.25/1.50 standard deviations.
+
+One important drawback of these studies is that they don't present the uncertainty of this estimate.
+Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
+Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education (Breen & Jonsson 2007).
+Virtually in all countries, be it developed or developing, there is inequality of opportunity.
+But there is considerable variation in the magnitud of this effect.
+For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries (Gosta and his phd student, Breen and other people, Shavit first study).
+
+With very recent data, Reardon and Portilla (2015) have unexpectedly uncovered a new finding: the reversal of the trend.
+From 1998 until 2010, it seems like the income achievement gap \footnote{The income achievement gap is the difference between the 90th and 10th income percentile in the test of interest} decreased at a rate of 0.01 and 0.008 standard deviations per year.
+In contrast to Reardon (2011), we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something not dramatically different from the previous estimates.
+The American situation is a very distinctive one relative to Europe.
+The reasons why we're seeing a reversal in the trend could be numerous and should be studied very close to context.
+For example, the American educational system lacks any formal curricular differentiation \emph{à la} European style.
+In contrast, Reardon and Portilla (2015) suggest that the reversal is likely due to the high increase of preschool enrollment.
+They suggest that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
+
+Some authors have taken this analysis to an international context in order to discover between-country trends.
+The comprehensive work of Corak and Waldfogel () performs a comparative analysis of Australia, United Kingdom, United States and Canada.
+Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is the only study that uses panel data to study achievement gaps, let alone to do this between countries}.
+The core finding behind the book is that the American achievement gap is much wider than in any other comparison countries.
+As mentioned earlier, they find that once the achievement gap is present in early school entry, it doesn't seem to narrow very much over the life course.
+They find that the achievement gap is very stable all the way from early education to secondary school (although their studies cover up until age 14).
+
+One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula.
+These four countries have no major jump in school selection, quite the opposite to the European average countries.
+A thorough review by Van der Werfhorst and Mijs (2010) sheds some light on the subject.
+First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps.
+Hanushek and Woessman (tracking study), use the TIMSS and PISA surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
+Moreover, Mickleweight & Schnepf (2007) using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
+In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
+
+Despite their high level of rigurosity, their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
+Their findings are very reliable but should be taken as suggestive at best.
+For this reason we should also pay particular attention to studies such as Chmielewski and Reardon (2016) and Chmielewski (2017) which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries using similar studies.
+These studies tackle a completely different question from the above, but they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
+
+Chmielewski and Reardon (2016), using the Programme for International Student Assessment (PISA), the Trends in International Mathematics and Science Study (TIMSS) and PIRLS (the Progress in International Reading Literacy Study), the assess whether there are patterns of cross-national variation in the achievement gap.
+In other words, is the size of achievement gap different for many countries?
+Although there is potential to study countries over time using these surveys, they choose not to.
+They do so because their question of interest (income categories) was only asked in three waves; in the end the only had three countries repeated in all three waves.
+
+Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
+In comparison to the literature on achievement gaps, they find that the U.S has a gap a little over 1 standard deviation in the year 2005 (for 15 year olds) and Germany has a decreasing gap from around ~ 1.25 to ~ 1 standard deviation.
+They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, etc ...
+The correlations are indeed very suggestive but clearly we must be extremely cautios in drawing causal changes from these two variables.
+But it is important to stress that their design is quite different from the work of Reardon (2011).
+Reardon (2011) takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds.
+Although the author adjusts for age, the trends can't be generalized to age-specific, but rather in overall terms.} and makes them comparable across time.
+This gives the author a very long time series to build a reliable achievement gap (over 40 years).
+In their study, Chmielewski and Reardon (2016) change the aim of the study to model between-country differences from a cross-sectional perspective.
+
+Chmielewski (2017), building on the work of Chmielewski and Reardon (2016) and Reardon and Portilla (2015) pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries.
+With over 50 years of data, and over 100 countries, Chmielewski (2017) finds that there seems to be a general world increase in the achievement gap.
+However, once she disentangles the relationship by country, she finds a reasonable amount of heterogeneity, with some countries seeing the achievement gap closing, others no change at all, while others record a steady increase.
+One clear limitation of their study (as well as Reardon (2011)) is that the adjust for the age of each child in all studies.
+Although for their purposes is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
+However, as we've noted before, Corak and Waldfogel (2015) find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as Reardon (2011) did.
+
+The evolution of High/Low SES gaps for preschool children might be much less marked than the same gap for high school children.
+The explanation, although very debated, has been gaining much support in recent years.
+In countries with high levels of curricular differentiation, the transition from early schooling into the tracking system has been found to increase inequality of learning (Wossman and Hanushek diff-in-diff).
+Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs (Van der Werfhorst and Mijs 2011).
+Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as elimination tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
+
+With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the High/Low SES achievement gap in the past 15 years for all PISA participant countries.
+This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for each country.
+The advantages of this study are numerous.
+First, we concentrate on the evolution of the gap for only 15 year olds.
+This is different from all the work of Chmielewski and Reardon, in which they pool all ages to get average changes, to study the properties of specific age-groups.
+As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
+Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.
+
+The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th. (Micklewright & Schnepf, 2007)
+
+Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (Mickleweight & Schnepf did it).
+They also find that France is a lower dispersion country (but in my results this is changing a lot).
+They also find that Germany and US are very dispersed countries (quite the opposite)
+
+Mickleweight & Schnepf find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
+Mickleweight & Schnpef find
+that average achievement is negatively related to dispersion.
+
+Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
+
+Talk about reardon and portilla and how the achievement gap has been closing in the last 15 years.
+Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
+Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
+Talk more about how countries with high social mobility has been linked to smallest achievement gaps
+Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.
+
+\section{Methods}
+
+\subsection{Data}
+
+\subsection{Data analysis}
+
+\section{Analysis}
+
+\subsection{Descriptives}
 
@@
 
-\section{Analysis}
+<<variance_pisa>>=
+
+var_years <-
+  map(adapted_year_data, function(.x) {
+  .x %>%
+  transmute(avg_cogn = matrixStats::rowMedians(as.matrix(select(.x, matches("^PV*.MATH$"))))) %>%
+  summarize(var = sd(avg_cogn, na.rm = T),
+            var_of_var = ) %>%
+  pull(var)
+})
+
+enframer <- function(df) {
+  df %>%
+    enframe() %>%
+    unnest()
+}
+
+var_years %>%
+  setNames(seq(2000, 2015, 3)) %>%
+  enframer() %>%
+  ggplot(aes(name, value)) + geom_col()
+@
+\subsection{Results}
 
@@
 
-\section{Literature Review}
-
-Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years.
-The literature on educational inequality has mainly concentrated on the United States (reardon here) but other international evidence is emerging with other similar findings for other count.
-The United States is usually the case study of interest, firstly, because it is the only country where cognitive testing is very widespread across surveys.
-And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day.
-Using this information, Reardon (2011) is the first to investigate the evolution of the cognitive gap and the results are very surprising.
-Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap).
-The widening of the achievement gap has been happening in parallel to the growth of income inequality.
-Although very suggestive, it is hard to link both things causally.
-
-Reardon (2011) finds that the increase has occured predominantely from the 1970's until the 2000's.
-In fact, hard numbers suggest that the gap increased by about ~ 40-50\%.
-The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%.
-Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footenote{Even after weighting appropriately.}.
-Finally, Reardon (2011) concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.
-Corak and Waldfogel (2015) don't find results which are very far away from this.
-They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25.
-Duncan and Magnuson find similarly to the previous studies and confirms a gap of ~ 1.25/1.50 standard deviations.
-
-One important drawback of these studies is that they don't present the uncertainty of this estimate.
-Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
-Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education (Breen & Jonsson 2007).
-Virtually in all countries, be it developed or developing, there is inequality of opportunity.
-But there is considerable variation in the magnitud of this effect.
-For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries (Gosta and his phd student, Breen and other people, Shavit first study).
-
-With very recent data, Reardon and Portilla (2015) have unexpectedly uncovered a new finding: the reversal of the trend.
-From 1998 until 2010, it seems like the income achievement gap \footnote{The income achievement gap is the difference between the 90th and 10th income percentile in the test of interest} decreased at a rate of 0.01 and 0.008 standard deviations per year.
-In contrast to Reardon (2011), we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something not dramatically different from the previous estimates.
-The American situation is a very distinctive one relative to Europe.
-The reasons why we're seeing a reversal in the trend could be numerous and should be studied very close to context.
-For example, the American educational system lacks any formal curricular differentiation \emph{à la} European style.
-In contrast, Reardon and Portilla (2015) suggest that the reversal is likely due to the high increase of preschool enrollment.
-They suggest that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
-
-Some authors have taken this analysis to an international context in order to discover between-country trends.
-The comprehensive work of Corak and Waldfogel () performs a comparative analysis of Australia, United Kingdom, United States and Canada.
-Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is the only study that uses panel data to study achievement gaps, let alone to do this between countries}.
-The core finding behind the book is that the American achievement gap is much wider than in any other comparison countries.
-As mentioned earlier, they find that once the achievement gap is present in early school entry, it doesn't seem to narrow very much over the life course.
-They find that the achievement gap is very stable all the way from early education to secondary school (although their studies cover up until age 14).
-
-One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula.
-These four countries have no major jump in school selection, quite the opposite to the European average countries.
-A thorough review by Van der Werfhorst and Mijs (2010) sheds some light on the subject.
-First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps.
-Hanushek and Woessman (tracking study), use the TIMSS and PISA surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
-Moreover, Mickleweight & Schnepf (2007) using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
-In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
-
-Despite their high level of rigurosity, their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
-Their findings are very reliable but should be taken as suggestive at best.
-For this reason we should also pay particular attention to studies such as Chmielewski and Reardon (2016) and Chmielewski (2017) which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries using similar studies.
-These studies tackle a completely different question from the above, but they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
-
-Chmielewski and Reardon (2016), using the Programme for International Student Assessment (PISA), the Trends in International Mathematics and Science Study (TIMSS) and PIRLS (the Progress in International Reading Literacy Study), the assess whether there are patterns of cross-national variation in the achievement gap.
-In other words, is the size of achievement gap different for many countries?
-Although there is potential to study countries over time using these surveys, they choose not to.
-They do so because their question of interest (income categories) was only asked in three waves; in the end the only had three countries repeated in all three waves.
-
-Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
-In comparison to the literature on achievement gaps, they find that the U.S has a gap a little over 1 standard deviation in the year 2005 (for 15 year olds) and Germany has a decreasing gap from around ~ 1.25 to ~ 1 standard deviation.
-They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, etc ...
-The correlations are indeed very suggestive but clearly we must be extremely cautios in drawing causal changes from these two variables.
-But it is important to stress that their design is quite different from the work of Reardon (2011).
-Reardon (2011) takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds.
-Although the author adjusts for age, the trends can't be generalized to age-specific, but rather in overall terms.} and makes them comparable across time.
-This gives the author a very long time series to build a reliable achievement gap (over 40 years).
-In their study, Chmielewski and Reardon (2016) change the aim of the study to model between-country differences from a cross-sectional perspective.
-
-Chmielewski (2017), building on the work of Chmielewski and Reardon (2016) and Reardon and Portilla (2015) pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries.
-With over 50 years of data, and over 100 countries, Chmielewski (2017) finds that there seems to be a general world increase in the achievement gap.
-However, once she disentangles the relationship by country, she finds a reasonable amount of heterogeneity, with some countries seeing the achievement gap closing, others no change at all, while others record a steady increase.
-One clear limitation of their study (as well as Reardon (2011)) is that the adjust for the age of each child in all studies.
-Although for their purposes is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
-However, as we've noted before, Corak and Waldfogel (2015) find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as Reardon (2011) did.
-
-The evolution of High/Low SES gaps for preschool children might be much less marked than the same gap for high school children.
-The explanation, although very debated, has been gaining much support in recent years.
-In countries with high levels of curricular differentiation, the transition from early schooling into the tracking system has been found to increase inequality of learning (Wossman and Hanushek diff-in-diff).
-Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs (Van der Werfhorst and Mijs 2011).
-Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as elimination tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
-
-With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the High/Low SES achievement gap in the past 15 years for all PISA participant countries.
-This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for each country.
-The advantages of this study are numerous.
-First, we concentrate on the evolution of the gap for only 15 year olds.
-This is different from all the work of Chmielewski and Reardon, in which they pool all ages to get average changes, to study the properties of specific age-groups.
-As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
-Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.
-
-Large inequalities in parent's education leads to large inequalities in children's education (Breen & Jonsson 2007)
-
-The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th. (Micklewright & Schnepf, 2007)
-
-Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (Mickleweight & Schnepf did it).
-They also find that France is a lower dispersion country (but in my results this is changing a lot).
-They also find that Germany and US are very dispersed countries (quite the opposite)
-
-Mickleweight & Schnepf find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
-Mickleweight & Schnpef find
-that average achievement is negatively related to dispersion.
-
-Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
-
-Talk about reardon and portilla and how the achievement gap has been closing in the last 15 years.
-Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
-Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
-Talk more about how countries with high social mobility has been linked to smallest achievement gaps
-Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.
-
-\section{Research design}
-
-Make sure PISA variation is constant over time
 \end{document}

### 2017-06-20 paper/paper.Rnw
Continued writing the data description

@@
 
+\usepackage{geometry}
+ \geometry{
+ a4paper,
+ total={170mm,257mm},
+ left=20mm,
+ top=20mm,
+ }
 \usepackage{pdflscape}
@@
 The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%.
-Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footenote{Even after weighting appropriately.}.
+Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footnote{Even after weighting appropriately.}.
 Finally, Reardon (2011) concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.
@@
 Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
-Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education (Breen & Jonsson 2007).
+Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education (Breen \& Jonsson 2007).
 Virtually in all countries, be it developed or developing, there is inequality of opportunity.
@@
 Hanushek and Woessman (tracking study), use the TIMSS and PISA surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
-Moreover, Mickleweight & Schnepf (2007) using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
+Moreover, Mickleweight \& Schnepf (2007) using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
 In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
@@
 
-The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th. (Micklewright & Schnepf, 2007)
-
-Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (Mickleweight & Schnepf did it).
+The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th. (Micklewright \& Schnepf, 2007)
+
+Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (Mickleweight \& Schnepf did it).
 They also find that France is a lower dispersion country (but in my results this is changing a lot).
@@
 
-Mickleweight & Schnepf find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
-Mickleweight & Schnpef find
+Mickleweight \& Schnepf find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
+Mickleweight \& Schnpef find
 that average achievement is negatively related to dispersion.
@@
 
-Talk about reardon and portilla and how the achievement gap has been closing in the last 15 years.
 Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
@@
 
+\section{Research questions}
+
 \section{Methods}
@@
 
+<<country_sample_numeric_vec>>=
+country_rows <- map_dbl(adapted_year_data, nrow)
+@
+
+In order the study the above mentioned questions I will be using the Programme for International Student Assessment (PISA).
+PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students.
+Currrently, PISA has six waves starting in the year 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 countries, both developed and developing ones.
+
+PISA collects data through a two-stage stratified sampling design.
+With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests.
+The sample size for PISA 2000 is \Sexpr{country_rows[1]}, for PISA 2003, \Sexpr{country_rows[2]}, for PISA 2006, \Sexpr{country_rows[3]}, for PISA 2009, \Sexpr{country_rows[4]}, for PISA 2012, \Sexpr{country_rows[5]}, and for PISA 2015 \Sexpr{country_rows[6]}.
+Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance.
+With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough time to visualize changes in the structure of an educational system.
+
+In order to measure family background PISA includes several variables.
+Firstly, the ask student's their parent's educational level.
+Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education (Reardon 2011).
+This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding.
+Moreover, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time.
+For example, until PISA 2009, the ISCED 1997 classification was currently used, but the next wave already used the new ISCED 2011 classification.
+Both these classification schemes have equivalent look-up tables, yet this requires a detailed inspection of the codings.
+
+Another social background variable PISA collects is the International Socio-Economic Index of Occupational Status (ISEI).
+This variable captures the economic status of the family, the closest it can without asking for income information.
+It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature(Anna study about world-wide inequality).
+PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.
+
+But the most importat variable for our study is the is a composite SES index created by the PISA team.
+The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI); the highest level of education of the student’s parents, converted into years of schooling; the PISA index of family wealth; the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (Education at a Glance, OECD, Paris, 2002, Glossary).
+This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable measures across waves.
+In fact, we use the newly-released ESCS index (% http://www.oecd-ilibrary.org/education/pisa-2015-results-volume-i/pisa-2015-technical-background_9789264266490-13-en
+) which rescaled all ESCS indexes to make them suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.}.
+
+Talk about plausible values, what they mean and how they're measured.
+
 \subsection{Data analysis}
+
+The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries.
+To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps(Fryer and Levitt 2004).
+The actual calculation of the achievement gap varies substantially and different strategies have been implemented.
+For example, Mickleweight \& Funsche calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution.
+Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy.
+Guan Kung Saw (2015), for instance, used parental education as a proxy of SES, whereas Soo-yong Byun and Kyung-keun Kim (2010) use a similar SES index as ours, but created by them.
+
+Reardon (2011) and subsequently (all reardon papers), used a different method developed by Reardon (2011) which we partially adopt adopt in this paper.
+SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.
+
+We start by standardizing achievement for each country, year and subject.
+Because PISA does not provide a single achievement indicator, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.
+We then fit a linear model for each wave where we regress the single achievement score on the age variable.
+We take the adjusted test scores and divide them by the root mean square error and divide it by the square root of the reliability of the survey for that year.
+
+Due to the standardization of achievement described above, gaps will also be attenuated in years where achievement is less reliably measured.
+Therefore, gaps are adjusted according to each country’s test reliability for each study, as published in the corresponding technical reports, as well as according to the estimated reliability of each SES report.
+For studies where both students and parents reported the same SES variable, reliability can be calculated from the microdata.
+These reliabilities are then applied to all other years.
+
+% To remember the models and formula for the above paragraph.
+      % mod1 <- lm(dv ~ AGE,
+      %            weights = .x[[weights_var]],
+      %            data = .x,
+      %            na.action = "na.exclude")
+      % 
+      % # Take residuals of model and divide by rmse. Multiply that by
+      % # 1 / sqrt(reliability of each survey), which is .y in the loop.
+      % .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)
+      % 
+      % mod2 <-
+      %   lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
+      %        data = .x,
+      %        weights = .x[[weights_var]])
 
@@
 
-<<variance_pisa>>=
+<<variance_pisa, cache = TRUE>>=
 
@@
 
-<<>>=
+<<graphing_achievement_disparity, cache = T>>=
 
@@
   }
-}) %>% setNames(seq(2000, 2015), 3) %>%
+}) %>% setNames(seq(2000, 2015, 3))
 

### 2017-06-20 paper/paper.tex
Continued writing the data description

@@
 
-\section{Descriptives}
-
-% Figure out how to ouput data frames as latex tables.
-
-\section{Analysis}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/correlation_incomeineq-1}
-
-}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_9010gaps-1}
-
-}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_allgaps-1}
-
-}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_ses_growth-1}
-
-}
-
-% How to output data frames are latex tables?
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/rate_change_graph-1}
-
-}
-
-% How to output data frames are latex tables?
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_perc_increase-1}
-
-}
-
 \section{Literature Review}
@@
 One important drawback of these studies is that they don't present the uncertainty of this estimate.
-Not necessarily to gauge their statistical significance, but to simply asses how much we can trust these numbers.
+Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
+Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education (Breen \& Jonsson 2007).
+Virtually in all countries, be it developed or developing, there is inequality of opportunity.
+But there is considerable variation in the magnitud of this effect.
+For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries (Gosta and his phd student, Breen and other people, Shavit first study).
 
@@
 
-Say more above Corak and Waldfogel above (book).
+One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula.
+These four countries have no major jump in school selection, quite the opposite to the European average countries.
+A thorough review by Van der Werfhorst and Mijs (2010) sheds some light on the subject.
+First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps.
+Hanushek and Woessman (tracking study), use the TIMSS and PISA surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
+Moreover, Mickleweight \& Schnepf (2007) using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
+In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
 
-Despite their high level of rigurosity, their analysis is based on four surveys that have significant differences and cannot be easily compared.
-Their findings are very reliable but should be taken as suggestive.
-For this reason we should also pay particular attention to studies such as Chmielewski and Reardon (2016) and Chmielewski (2017) which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries.
+Despite their high level of rigurosity, their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
+Their findings are very reliable but should be taken as suggestive at best.
+For this reason we should also pay particular attention to studies such as Chmielewski and Reardon (2016) and Chmielewski (2017) which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries using similar studies.
 These studies tackle a completely different question from the above, but they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
@@
 
-Talk about reardon and portilla and how the achievement gap has been closing in the last 15 years.
+The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th. (Micklewright \& Schnepf, 2007)
+
+Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (Mickleweight \& Schnepf did it).
+They also find that France is a lower dispersion country (but in my results this is changing a lot).
+They also find that Germany and US are very dispersed countries (quite the opposite)
+
+Mickleweight \& Schnepf find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
+Mickleweight \& Schnpef find
+that average achievement is negatively related to dispersion.
+
+Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
+
 Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
@@
 
-\section{Research design}
+\section{Research questions}
 
-Make sure PISA variation is constant over time
+\section{Methods}
+
+\subsection{Data}
+
+In order the study the above mentioned questions I will be using the Programme for International Student Assessment (PISA).
+PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students.
+Currrently, PISA has six waves starting in the year 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 countries, both developed and developing ones.
+
+PISA collects data through a two-stage stratified sampling design.
+With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests.
+The sample size for PISA 2000 is \ensuremath{1.27388\times 10^{5}}, for PISA 2003, \ensuremath{2.76165\times 10^{5}}, for PISA 2006, \ensuremath{3.9875\times 10^{5}}, for PISA 2009, \ensuremath{5.15958\times 10^{5}}, for PISA 2012, \ensuremath{4.8081\times 10^{5}}, and for PISA 2015 \ensuremath{5.19334\times 10^{5}}.
+Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance.
+With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough time to visualize changes in the structure of an educational system.
+
+In order to measure family background PISA includes several variables.
+Firstly, the ask student's their parent's educational level.
+Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education (Reardon 2011).
+This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding.
+Moreover, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time.
+For example, until PISA 2009, the ISCED 1997 classification was currently used, but the next wave already used the new ISCED 2011 classification.
+Both these classification schemes have equivalent look-up tables, yet this requires a detailed inspection of the codings.
+
+Another social background variable PISA collects is the International Socio-Economic Index of Occupational Status (ISEI).
+This variable captures the economic status of the family, the closest it can without asking for income information.
+It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature(Anna study about world-wide inequality).
+PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.
+
+But the most importat variable for our study is the is a composite SES index created by the PISA team.
+The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI); the highest level of education of the student’s parents, converted into years of schooling; the PISA index of family wealth; the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (Education at a Glance, OECD, Paris, 2002, Glossary).
+This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable measures across waves.
+In fact, we use the newly-released ESCS index (% http://www.oecd-ilibrary.org/education/pisa-2015-results-volume-i/pisa-2015-technical-background_9789264266490-13-en
+) which rescaled all ESCS indexes to make them suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.}.
+
+Talk about plausible values, what they mean and how they're measured.
+
+\subsection{Data analysis}
+
+The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries.
+To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps.
+The actual calculation of the achievement gap varies substantially and different strategies have been implemented.
+For example, Mickleweight & Funsche calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution.
+Although in principle you should be able to capture some type of SES effect, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education.
+\section{Analysis}
+
+\subsection{Descriptives}
+
+% Figure out how to ouput data frames as latex tables.
+
+\subsection{Results}
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/correlation_incomeineq-1}
+
+}
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_9010gaps-1}
+
+}
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_allgaps-1}
+
+}
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_ses_growth-1}
+
+}
+
+% How to output data frames are latex tables?
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/rate_change_graph-1}
+
+}
+
+% How to output data frames are latex tables?
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_perc_increase-1}
+
+}
+
 \end{document}

### 2017-06-20 paper/paper.Rnw
Continued writing data section + started adding bibliography

@@
     "2015" = 0.74) # 2015 imputed
+
 @
@@
 Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years.
-The literature on educational inequality has mainly concentrated on the United States (reardon here) but other international evidence is emerging with other similar findings for other count.
+The literature on educational inequality has mainly concentrated on the United States \citep{reardon2011} but other international evidence is emerging with other similar findings for other count.
 The United States is usually the case study of interest, firstly, because it is the only country where cognitive testing is very widespread across surveys.
 And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day.
-Using this information, Reardon (2011) is the first to investigate the evolution of the cognitive gap and the results are very surprising.
-Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap).
+Using this information, \citep{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
+Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap -waldfogel study).
 The widening of the achievement gap has been happening in parallel to the growth of income inequality.
@@
 
-Reardon (2011) finds that the increase has occured predominantely from the 1970's until the 2000's.
+\citet{reardon2011} finds that the increase has occured predominantely from the 1970's until the 2000's.
 In fact, hard numbers suggest that the gap increased by about ~ 40-50\%.
@@
 Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footnote{Even after weighting appropriately.}.
-Finally, Reardon (2011) concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.
-Corak and Waldfogel (2015) don't find results which are very far away from this.
+Finally, \citet{readon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.
+\citet{bradbury2015} don't find results which are very far away from this.
 They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25.
-Duncan and Magnuson find similarly to the previous studies and confirms a gap of ~ 1.25/1.50 standard deviations.
+\citet{duncan2011} find similarly to the previous studies and confirm a gap of ~ 1.25/1.50 standard deviations.
 
@@
 Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
-Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education (Breen \& Jonsson 2007).
+Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education \citep{breen2007}.
 Virtually in all countries, be it developed or developing, there is inequality of opportunity.
 But there is considerable variation in the magnitud of this effect.
-For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries (Gosta and his phd student, Breen and other people, Shavit first study).
-
-With very recent data, Reardon and Portilla (2015) have unexpectedly uncovered a new finding: the reversal of the trend.
+For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries \citep{esping2012, breen2007, shavit1993}.
+
+With very recent data, \citet{reardon_portilla} have unexpectedly uncovered a new finding: the reversal of the trend.
 From 1998 until 2010, it seems like the income achievement gap \footnote{The income achievement gap is the difference between the 90th and 10th income percentile in the test of interest} decreased at a rate of 0.01 and 0.008 standard deviations per year.
-In contrast to Reardon (2011), we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something not dramatically different from the previous estimates.
+In contrast to \citep{reardon2011}, we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something not dramatically different from the previous estimates.
 The American situation is a very distinctive one relative to Europe.
@@
 For example, the American educational system lacks any formal curricular differentiation \emph{à la} European style.
-In contrast, Reardon and Portilla (2015) suggest that the reversal is likely due to the high increase of preschool enrollment.
+In contrast, \citet{reardon2011} suggest that the reversal is likely due to the high increase of preschool enrollment.
 They suggest that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
@@
 Some authors have taken this analysis to an international context in order to discover between-country trends.
-The comprehensive work of Corak and Waldfogel () performs a comparative analysis of Australia, United Kingdom, United States and Canada.
+The comprehensive work of \citet{bradbury2015} performs a comparative analysis of Australia, United Kingdom, United States and Canada.
 Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is the only study that uses panel data to study achievement gaps, let alone to do this between countries}.
@@
 
+% You left documenting the bibliography here
 One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula.
 These four countries have no major jump in school selection, quite the opposite to the European average countries.
-A thorough review by Van der Werfhorst and Mijs (2010) sheds some light on the subject.
+A thorough review by \citet{werfhorst_mijs} sheds some light on the subject.
 First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps.
@@
 Talk more about how countries with high social mobility has been linked to smallest achievement gaps
+
 Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.
@@
 Because PISA does not provide a single achievement indicator, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.
-We then fit a linear model for each wave where we regress the single achievement score on the age variable.
-We take the adjusted test scores and divide them by the root mean square error and divide it by the square root of the reliability of the survey for that year.
-
-Due to the standardization of achievement described above, gaps will also be attenuated in years where achievement is less reliably measured.
-Therefore, gaps are adjusted according to each country’s test reliability for each study, as published in the corresponding technical reports, as well as according to the estimated reliability of each SES report.
-For studies where both students and parents reported the same SES variable, reliability can be calculated from the microdata.
-These reliabilities are then applied to all other years.
-
-% To remember the models and formula for the above paragraph.
-      % mod1 <- lm(dv ~ AGE,
-      %            weights = .x[[weights_var]],
-      %            data = .x,
-      %            na.action = "na.exclude")
-      % 
-      % # Take residuals of model and divide by rmse. Multiply that by
-      % # 1 / sqrt(reliability of each survey), which is .y in the loop.
-      % .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)
-      % 
-      % mod2 <-
-      %   lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
-      %        data = .x,
-      %        weights = .x[[weights_var]])
+We then fit a linear model for each wave where we regress the single achievement score on the age variable weighted by the appropriate student sample weights.
+We take the residuals and divide them by the root mean square error of the model and divide that by one over the square root of the reliability of the survey for that year-subject (implement this all in formula).
+\footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair (Anna analysis of wordwide gap).
+Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own at the moment of writing this paper.
+For this reason we implement the analysis following the original work of Reardon (2011)}
+
+This implementation has several drawbacks (talk about the assumptions that reardong makes).
+
+We estimate the SES gap in a very simple fashion.
+For each country-wave, we estimate the weighted 90th and 10th percentile scores for the SES composite index, and then we generate a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.
+Finally, for each wave, we fit a mixed model with the adjusted test score as the dependent variable and the SES index as the only covariate.
+We allow both the intercept and the dummy variable to vary by country, and weight the model by the appropriate student weights.
+
+We take the random effects of each country together with its standard error as the achievement gap.
+We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size.
+Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible.
+
+% Show the properties of the ESCS dummy variable
+% % in high educated
+% % of home possessions
+% % of ISEI
 
@@
 
+\bibliography{mybibliography.bib}
+
 \end{document}

### 2017-06-21 paper/paper.Rnw
Wrote research design section + plausible values

@@
 
+This paper is interested in two questions.
+As explained in the literature review, we already know that U.S achievement gaps have been growing over time, and some international evidence suggests that it might be increasing in other countries as well.
+All of these studies pool all students together from different ages and estimate global trends.
+
+We take a different approach and we want to study age-specific trends in achievement gaps.
+We're interested in studying a) the size of the achievement in a comparative perspective and its relationship to overall achievement levels, b) which countries are experiencing changes in the achievement gap, c) the rate at which each country-gap is widening/narrowing and d) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
+
+We develop each question separately for more detail.
+
+\begin{description}
+  \item[We want to test the notion that better performing countries have lower leves of achievement gap.
+\citep{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
+This pattern is not so obvious.
+For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score.
+But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with widening achievement gap.]
+  \item[The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
+After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
+This lends credibility that our estimations will show variation over time.
+As usual, their estimations are only for children going deom 4th to 8th grade.
+We concentrate on age-specific groups, in this case, 15 year olds.]
+  \item[We want to compare the percentage change at which the gap widened/narrowed from the first to the last year.
+This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
+Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders.}]
+  \item[The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
+The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind.
+We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.
+This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving.
+If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.]
+\end{description}
+
 \section{Methods}
@@
 
-But the most importat variable for our study is the is a composite SES index created by the PISA team.
+But one of the most relevant variables for our study is the is a composite SES index created by the PISA team.
 The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI); the highest level of education of the student’s parents, converted into years of schooling; the PISA index of family wealth; the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (Education at a Glance, OECD, Paris, 2002, Glossary).
@@
 
-Talk about plausible values, what they mean and how they're measured.
+Aside from SES, the other most relevant variables are test scores.
+PISA provides plausible values instead of one single indicator of test score performance.
+As explained in the PISA manual (PISA 2012, 2015), these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured.
+Suppose we have µi, the average student test score in mathematics for student i.
+Instead of estimating µi alone, plausible values estimate a distribution of possible µ’s, for student i, together with the likelihood of each µi based on the respondents answers on the test.
+This is defined as the posterior distributions of µ’s for student i.
+The reason why we use this procedure is because estimating a single estimate µi is plagued with measurement error, among other types of bias (see Wu, 2005).
+The plausible values in the PIAAC data are effectively five (ten for PISA 2015) random draws from this distribution.
+These variable is are continuous, ranging from 0 to 500, with a mean of 250.
 
@@
 SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.
+
+% Transform all of this text into equations
 
@@
 
-This implementation has several drawbacks (talk about the assumptions that reardong makes).
-
-We estimate the SES gap in a very simple fashion.
+Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves.
+However, if the variance of academic achievement changes over time, then standardizing the overall score at each wave of testing confounds changes in the gap with changes in the variance of test scores.
+That is, by standardizing we're forcing the variance of test scores to be zero across all waves.
+But if the true variance of academic achievement grows over time, then the estimated trend in the achievement gaps will be underestimated, and vice versa.
+
+% Show graph of variance over time
+% I'm showing standard deviations, is it better to show variance?
+<<variance_pisa, cache = TRUE>>=
+
+var_years <-
+  map(adapted_year_data, function(.x) {
+  .x %>%
+  transmute(avg_cogn = matrixStats::rowMedians(as.matrix(select(.x, matches("^PV*.MATH$"))))) %>%
+  summarize(var = sd(avg_cogn, na.rm = T),
+            var_of_var = ) %>%
+  pull(var)
+})
+
+enframer <- function(df) {
+  df %>%
+    enframe() %>%
+    unnest()
+}
+
+var_years %>%
+  setNames(seq(2000, 2015, 3)) %>%
+  enframer() %>%
+  ggplot(aes(name, value)) + geom_col()
+@
+
+Inspection of the data suggests that it's something we shouldn't be deeply concerned with.
+The standard deviation of each wave seems to be following a very similar pattern with an exception of the year 2000.
+
+Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap.
+This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citet{reardon2011}.
+The PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but still we have to adjust for this imprecision.
+
+In order to correct gap estimates for measurement error, I weight the student test scores by the P where P is the reliability of the test for each wave.
+Each PISA study provided a global reliability indicator which we use accordingly to each wave.
+This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.
+
+After constructing the adjusted test score measurement, we estimate the SES gap in a very simple fashion.
 For each country-wave, we estimate the weighted 90th and 10th percentile scores for the SES composite index, and then we generate a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.
@@
 
-<<variance_pisa, cache = TRUE>>=
-
-var_years <-
-  map(adapted_year_data, function(.x) {
-  .x %>%
-  transmute(avg_cogn = matrixStats::rowMedians(as.matrix(select(.x, matches("^PV*.MATH$"))))) %>%
-  summarize(var = sd(avg_cogn, na.rm = T),
-            var_of_var = ) %>%
-  pull(var)
-})
-
-enframer <- function(df) {
-  df %>%
-    enframe() %>%
-    unnest()
-}
-
-var_years %>%
-  setNames(seq(2000, 2015, 3)) %>%
-  enframer() %>%
-  ggplot(aes(name, value)) + geom_col()
-@
 \subsection{Results}

### 2017-06-21 paper/paper.Rnw
Included about 80% of equations - haven't tested yet

@@
 Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footnote{Even after weighting appropriately.}.
-Finally, \citet{readon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.
+Finally, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.
 \citet{bradbury2015} don't find results which are very far away from this.
@@
 The correlations are indeed very suggestive but clearly we must be extremely cautios in drawing causal changes from these two variables.
-But it is important to stress that their design is quite different from the work of Reardon (2011).
-Reardon (2011) takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds.
+But it is important to stress that their design is quite different from the work of \citet{reardon2011}.
+\citet{reardon2011} takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds.
 Although the author adjusts for age, the trends can't be generalized to age-specific, but rather in overall terms.} and makes them comparable across time.
@@
 However, once she disentangles the relationship by country, she finds a reasonable amount of heterogeneity, with some countries seeing the achievement gap closing, others no change at all, while others record a steady increase.
-One clear limitation of their study (as well as Reardon (2011)) is that the adjust for the age of each child in all studies.
+One clear limitation of their study (as well as \citet{reardon2011}) is that the adjust for the age of each child in all studies.
 Although for their purposes is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
-However, as we've noted before, Corak and Waldfogel (2015) find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as Reardon (2011) did.
+However, as we've noted before, Corak and Waldfogel (2015) find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as \citet{reardon2011} did.
 
@@
 
-\begin{description}
-  \item[We want to test the notion that better performing countries have lower leves of achievement gap.
+a) We want to test the notion that better performing countries have lower leves of achievement gap.
 \citep{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
@@
 For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score.
-But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with widening achievement gap.]
-  \item[The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
+But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with widening achievement gap.
+
+b) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
 After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
 This lends credibility that our estimations will show variation over time.
-As usual, their estimations are only for children going deom 4th to 8th grade.
-We concentrate on age-specific groups, in this case, 15 year olds.]
-  \item[We want to compare the percentage change at which the gap widened/narrowed from the first to the last year.
+As usual, their estimations are only for children going from 4th to 8th grade.
+We concentrate on age-specific groups, in this case, 15 year olds.
+
+c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year.
 This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
-Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders.}]
-  \item[The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
+Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders.}
+
+d) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
 The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind.
@@
 This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving.
-If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.]
-\end{description}
+If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.
 
@@
 Firstly, the ask student's their parent's educational level.
-Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education (Reardon 2011).
+Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}.
 This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding.
@@
 
-Reardon (2011) and subsequently (all reardon papers), used a different method developed by Reardon (2011) which we partially adopt adopt in this paper.
+\citet{reardon2011} and subsequently (all reardon papers), used a different method developed by \citet{reardon2011} which we partially adopt adopt in this paper.
 SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.
-
-% Transform all of this text into equations
 
@@
 Because PISA does not provide a single achievement indicator, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.
-We then fit a linear model for each wave where we regress the single achievement score on the age variable weighted by the appropriate student sample weights.
-We take the residuals and divide them by the root mean square error of the model and divide that by one over the square root of the reliability of the survey for that year-subject (implement this all in formula).
-\footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair (Anna analysis of wordwide gap).
-Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own at the moment of writing this paper.
-For this reason we implement the analysis following the original work of Reardon (2011)}
+
+We then fit a linear model
+
+\begin{equation}
+Y_i = \alpha + \beta_1 * AGE_i + \epsilon_i, \epsilon_i ~ N(0, \sigma^2)
+\end{equation}
+
+for each wave where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (because all children should be 15 year olds) weighted by the student sample weights.
+
+We then adjust \begin{math} \hat{Y_i} \end{math} by
+
+% I think the denominator could be wrong! Because I include y_hat and that's what the equation is estimating
+\begin{equation}
+\hat{Y_i} = \frac{\hat{\epsilon_i}}{\sqrt{1/n \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
+\end{equation}
+
+% # Take residuals of model and divide by rmse. Multiply that by
+% # 1 / sqrt(reliability of each survey), which is .y in the loop.
+% .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)
+
+Where \begin{math}\epsilon_i\end{math} is the residual for student \begin{math}i\end{math} and the denominator is the root mean square error of the model.
 
@@
 Inspection of the data suggests that it's something we shouldn't be deeply concerned with.
-The standard deviation of each wave seems to be following a very similar pattern with an exception of the year 2000.
+The standard deviation of each wave seems to be following a very similar pattern with not so drastic exception of the year 2000.
 
@@
 This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citet{reardon2011}.
-The PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but still we have to adjust for this imprecision.
-
-In order to correct gap estimates for measurement error, I weight the student test scores by the P where P is the reliability of the test for each wave.
-Each PISA study provided a global reliability indicator which we use accordingly to each wave.
+The PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but we still have to adjust for this imprecision.
+
+In order to correct gap estimates for measurement error, I adjust student test scores by
+
+\begin{equation}
+\hat{Y_i} = \hat{Y_i} * \frac{1}{\sqrt{r}}
+\end{equation}
+
+where \begin{math}r\end{math} is the reliability score of the wave.
+\footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair (Anna analysis of wordwide gap).
+Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own at the moment of writing this paper.
+For this reason we implement the analysis following the original work of \citet{reardon2011}}.
+Each PISA survey provides a reliability indicator which we use accordingly.
 This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.
 
-After constructing the adjusted test score measurement, we estimate the SES gap in a very simple fashion.
-For each country-wave, we estimate the weighted 90th and 10th percentile scores for the SES composite index, and then we generate a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.
-Finally, for each wave, we fit a mixed model with the adjusted test score as the dependent variable and the SES index as the only covariate.
-We allow both the intercept and the dummy variable to vary by country, and weight the model by the appropriate student weights.
+After constructing the adjusted test score measurement, we estimate the SES dummy by calculating the weighted 90th and 10th percentile scores for the SES composite index, and then generating a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.
+
+We then fit
+
+\begin{equation}
+\hat{Y_i} = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i, \rlap{\quad \text{for} \(i\) \text{= 1, 2, ...,} \(n\) \text{within country} \(j\)}
+\end{equation}
+
+% mod2 <-
+%   lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
+%        data = .x,
+%        weights = .x[[weights_var]])
+
+where \(SES_i\) is whether the student is in the top/bottom SES dummy and we allow both \(\alpha_i\) and \(\beta_i\) to vary by country \(j\).
+We implement this separately for each wave and weighting by the student sample weights.
+
+Finally,
 

### 2017-06-21 paper/paper.tex
Included about 80% of equations - haven't tested yet

@@
 Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footnote{Even after weighting appropriately.}.
-Finally, Reardon (2011) concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.
-Corak and Waldfogel (2015) don't find results which are very far away from this.
+Finally, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.
+\citet{bradbury2015} don't find results which are very far away from this.
 They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25.
-Duncan and Magnuson find similarly to the previous studies and confirms a gap of ~ 1.25/1.50 standard deviations.
+\citet{duncan2011} find similarly to the previous studies and confirm a gap of ~ 1.25/1.50 standard deviations.
 
@@
 Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
-Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education (Breen \& Jonsson 2007).
+Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education \citep{breen2007}.
 Virtually in all countries, be it developed or developing, there is inequality of opportunity.
 But there is considerable variation in the magnitud of this effect.
-For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries (Gosta and his phd student, Breen and other people, Shavit first study).
+For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries \citep{esping2012, breen2007, shavit1993}.
 
-With very recent data, Reardon and Portilla (2015) have unexpectedly uncovered a new finding: the reversal of the trend.
+With very recent data, \citet{reardon_portilla} have unexpectedly uncovered a new finding: the reversal of the trend.
 From 1998 until 2010, it seems like the income achievement gap \footnote{The income achievement gap is the difference between the 90th and 10th income percentile in the test of interest} decreased at a rate of 0.01 and 0.008 standard deviations per year.
-In contrast to Reardon (2011), we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something not dramatically different from the previous estimates.
+In contrast to \citep{reardon2011}, we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something not dramatically different from the previous estimates.
 The American situation is a very distinctive one relative to Europe.
@@
 \section{Research questions}
+
+This paper is interested in two questions.
+As explained in the literature review, we already know that U.S achievement gaps have been growing over time, and some international evidence suggests that it might be increasing in other countries as well.
+All of these studies pool all students together from different ages and estimate global trends.
+
+We take a different approach and we want to study age-specific trends in achievement gaps.
+We're interested in studying a) the size of the achievement in a comparative perspective and its relationship to overall achievement levels, b) which countries are experiencing changes in the achievement gap, c) the rate at which each country-gap is widening/narrowing and d) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
+
+We develop each question separately for more detail.
+
+a) We want to test the notion that better performing countries have lower leves of achievement gap.
+\citep{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
+This pattern is not so obvious.
+For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score.
+But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with widening achievement gap.
+
+b) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
+After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
+This lends credibility that our estimations will show variation over time.
+As usual, their estimations are only for children going from 4th to 8th grade.
+We concentrate on age-specific groups, in this case, 15 year olds.
+
+c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year.
+This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
+Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders.}
+
+d) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
+The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind.
+We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.
+This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving.
+If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.
 
@@
 The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries.
-To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps.
+To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps(Fryer and Levitt 2004).
 The actual calculation of the achievement gap varies substantially and different strategies have been implemented.
-For example, Mickleweight & Funsche calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution.
-Although in principle you should be able to capture some type of SES effect, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education.
+For example, Mickleweight \& Funsche calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution.
+Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy.
+Guan Kung Saw (2015), for instance, used parental education as a proxy of SES, whereas Soo-yong Byun and Kyung-keun Kim (2010) use a similar SES index as ours, but created by them.
+
+\citet{reardon2011} and subsequently (all reardon papers), used a different method developed by \citet{reardon2011} which we partially adopt adopt in this paper.
+SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.
+
+% Transform all of this text into equations
+
+\[ x^n + y^n = z^n \]
+
+\begin{math}
+E=mc^2
+\end{math}
+
+We start by standardizing achievement for each country, year and subject.
+Because PISA does not provide a single achievement indicator, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.
+We then fit a linear model for each wave where we regress the single achievement score on the age variable weighted by the appropriate student sample weights.
+We take the residuals and divide them by the root mean square error of the model and divide that by one over the square root of the reliability of the survey for that year-subject (implement this all in formula).
+\footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair (Anna analysis of wordwide gap).
+Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own at the moment of writing this paper.
+For this reason we implement the analysis following the original work of \citet{reardon2011}}
+
+Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves.
+However, if the variance of academic achievement changes over time, then standardizing the overall score at each wave of testing confounds changes in the gap with changes in the variance of test scores.
+That is, by standardizing we're forcing the variance of test scores to be zero across all waves.
+But if the true variance of academic achievement grows over time, then the estimated trend in the achievement gaps will be underestimated, and vice versa.
+
+% Show graph of variance over time
+% I'm showing standard deviations, is it better to show variance?
+
+Inspection of the data suggests that it's something we shouldn't be deeply concerned with.
+The standard deviation of each wave seems to be following a very similar pattern with an exception of the year 2000.
+
+Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap.
+This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citet{reardon2011}.
+The PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but still we have to adjust for this imprecision.
+
+In order to correct gap estimates for measurement error, I weight the student test scores by the P where P is the reliability of the test for each wave.
+Each PISA study provided a global reliability indicator which we use accordingly to each wave.
+This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.
+
+After constructing the adjusted test score measurement, we estimate the SES gap in a very simple fashion.
+For each country-wave, we estimate the weighted 90th and 10th percentile scores for the SES composite index, and then we generate a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.
+Finally, for each wave, we fit a mixed model with the adjusted test score as the dependent variable and the SES index as the only covariate.
+We allow both the intercept and the dummy variable to vary by country, and weight the model by the appropriate student weights.
+
+We take the random effects of each country together with its standard error as the achievement gap.
+We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size.
+Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible.
+
+% Show the properties of the ESCS dummy variable
+% % in high educated
+% % of home possessions
+% % of ISEI
+
 \section{Analysis}

### 2017-06-21 paper/paper.Rnw
Updated + tested equations - everything working

@@
 \usepackage{pdflscape}
+\usepackage{amsmath}
 \usepackage{subcaption}
@@
 
-for each wave where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (because all children should be 15 year olds) weighted by the student sample weights.
+for each wave,  where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (because all children should be 15 year olds) weighted by the student sample weights.
 
@@
 \begin{equation}
-\hat{Y_i} = \frac{\hat{\epsilon_i}}{\sqrt{1/n \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
+\hat{Y_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
 \end{equation}
@@
 Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap.
-This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citet{reardon2011}.
-The PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but we still have to adjust for this imprecision.
+This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citep{reardon2011}.
+PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but we still have to adjust for this imprecision.
 
@@
 \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair (Anna analysis of wordwide gap).
-Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own at the moment of writing this paper.
+Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own.
+At the moment of writing this paper, they were unavailable.
 For this reason we implement the analysis following the original work of \citet{reardon2011}}.
@@
 \begin{equation}
-\hat{Y_i} = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i, \rlap{\quad \text{for} \(i\) \text{= 1, 2, ...,} \(n\) \text{within country} \(j\)}
+\hat{Y_i} = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
 \end{equation}
@@
 where \(SES_i\) is whether the student is in the top/bottom SES dummy and we allow both \(\alpha_i\) and \(\beta_i\) to vary by country \(j\).
-We implement this separately for each wave and weighting by the student sample weights.
-
-Finally,
-
-We take the random effects of each country together with its standard error as the achievement gap.
+We implement this separately for each wave and weight by the student sample weights.
+
+Finally, we calculate the SES achievement gap for each country by extracting the fitted \(\alpha\) and \(\beta\) for each country \(j\) and calculating the difference of the predicted score of High and Low SES.
+
+\begin{equation}
+\begin{split}
+\quad \text{High SES}_j = \alpha_j + \beta_j \\
+\quad \text{Low SES}_j = \alpha_j \\
+\quad \text{SES gap}_j = \text{High SES}_j - \text{Low SES}_j
+\end{split}
+\end{equation}
+
+We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends.
 We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size.

### 2017-06-21 paper/paper.tex
Updated + tested equations - everything working

@@
 
-% Transform all of this text into equations
-
-\[ x^n + y^n = z^n \]
-
-\begin{math}
-E=mc^2
-\end{math}
-
 We start by standardizing achievement for each country, year and subject.
 Because PISA does not provide a single achievement indicator, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.
-We then fit a linear model for each wave where we regress the single achievement score on the age variable weighted by the appropriate student sample weights.
-We take the residuals and divide them by the root mean square error of the model and divide that by one over the square root of the reliability of the survey for that year-subject (implement this all in formula).
-\footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair (Anna analysis of wordwide gap).
-Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own at the moment of writing this paper.
-For this reason we implement the analysis following the original work of \citet{reardon2011}}
+
+We then fit a linear model
+
+\begin{equation}
+Y_i = \alpha + \beta_1 * AGE_i + \epsilon_i, \epsilon_i ~ N(0, \sigma^2)
+\end{equation}
+
+for each wave,  where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (because all children should be 15 year olds) weighted by the student sample weights.
+
+We then adjust \begin{math} \hat{Y_i} \end{math} by
+
+% I think the denominator could be wrong! Because I include y_hat and that's what the equation is estimating
+\begin{equation}
+\hat{Y_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
+\end{equation}
+
+% # Take residuals of model and divide by rmse. Multiply that by
+% # 1 / sqrt(reliability of each survey), which is .y in the loop.
+% .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)
+
+Where \begin{math}\epsilon_i\end{math} is the residual for student \begin{math}i\end{math} and the denominator is the root mean square error of the model.
 
@@
 Inspection of the data suggests that it's something we shouldn't be deeply concerned with.
-The standard deviation of each wave seems to be following a very similar pattern with an exception of the year 2000.
+The standard deviation of each wave seems to be following a very similar pattern with not so drastic exception of the year 2000.
 
 Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap.
-This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citet{reardon2011}.
-The PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but still we have to adjust for this imprecision.
+This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citep{reardon2011}.
+PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but we still have to adjust for this imprecision.
 
-In order to correct gap estimates for measurement error, I weight the student test scores by the P where P is the reliability of the test for each wave.
-Each PISA study provided a global reliability indicator which we use accordingly to each wave.
+In order to correct gap estimates for measurement error, I adjust student test scores by
+
+\begin{equation}
+\hat{Y_i} = \hat{Y_i} * \frac{1}{\sqrt{r}}
+\end{equation}
+
+where \begin{math}r\end{math} is the reliability score of the wave.
+\footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair (Anna analysis of wordwide gap).
+Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own.
+At the moment of writing this paper, they were unavailable.
+For this reason we implement the analysis following the original work of \citet{reardon2011}}.
+Each PISA survey provides a reliability indicator which we use accordingly.
 This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.
 
-After constructing the adjusted test score measurement, we estimate the SES gap in a very simple fashion.
-For each country-wave, we estimate the weighted 90th and 10th percentile scores for the SES composite index, and then we generate a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.
-Finally, for each wave, we fit a mixed model with the adjusted test score as the dependent variable and the SES index as the only covariate.
-We allow both the intercept and the dummy variable to vary by country, and weight the model by the appropriate student weights.
+After constructing the adjusted test score measurement, we estimate the SES dummy by calculating the weighted 90th and 10th percentile scores for the SES composite index, and then generating a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.
 
-We take the random effects of each country together with its standard error as the achievement gap.
+We then fit
+
+\begin{equation}
+\hat{Y_i} = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
+\end{equation}
+
+% mod2 <-
+%   lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
+%        data = .x,
+%        weights = .x[[weights_var]])
+
+where \(SES_i\) is whether the student is in the top/bottom SES dummy and we allow both \(\alpha_i\) and \(\beta_i\) to vary by country \(j\).
+We implement this separately for each wave and weight by the student sample weights.
+
+Finally, we calculate the SES achievement gap for each country by extracting the fitted \(\alpha\) and \(\beta\) for each country \(j\) and calculating the difference of the predicted score of High and Low SES.
+
+\begin{equation}
+\begin{split}
+\quad \text{High SES}_j = \alpha_j + \beta_j \\
+\quad \text{Low SES}_j = \alpha_j \\
+\quad \text{SES gap}_j = \text{High SES}_j - \text{Low SES}_j
+\end{split}
+\end{equation}
+
+We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends.
 We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size.

### 2017-06-21 paper/paper.Rnw
About 80% of citations

@@
 Hanushek and Woessman (tracking study), use the TIMSS and PISA surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
-Moreover, Mickleweight \& Schnepf (2007) using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
+Moreover, \citet{micklewright} using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
 In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
@@
 Their findings are very reliable but should be taken as suggestive at best.
-For this reason we should also pay particular attention to studies such as Chmielewski and Reardon (2016) and Chmielewski (2017) which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries using similar studies.
+For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citep{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries using similar studies.
 These studies tackle a completely different question from the above, but they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
 
-Chmielewski and Reardon (2016), using the Programme for International Student Assessment (PISA), the Trends in International Mathematics and Science Study (TIMSS) and PIRLS (the Progress in International Reading Literacy Study), the assess whether there are patterns of cross-national variation in the achievement gap.
+\citet{anna2016}, using the Programme for International Student Assessment (PISA), the Trends in International Mathematics and Science Study (TIMSS) and PIRLS (the Progress in International Reading Literacy Study), the assess whether there are patterns of cross-national variation in the achievement gap.
 In other words, is the size of achievement gap different for many countries?
@@
 This gives the author a very long time series to build a reliable achievement gap (over 40 years).
-In their study, Chmielewski and Reardon (2016) change the aim of the study to model between-country differences from a cross-sectional perspective.
-
-Chmielewski (2017), building on the work of Chmielewski and Reardon (2016) and Reardon and Portilla (2015) pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries.
-With over 50 years of data, and over 100 countries, Chmielewski (2017) finds that there seems to be a general world increase in the achievement gap.
+In their study, \citet{anna2016} change the aim of the study to model between-country differences from a cross-sectional perspective.
+
+\citet{anna2016_global}, building on the work of \citet{anna2016} and \citet{reardon_portilla} pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries.
+With over 50 years of data, and over 100 countries, she finds that there seems to be a general world increase in the achievement gap.
 However, once she disentangles the relationship by country, she finds a reasonable amount of heterogeneity, with some countries seeing the achievement gap closing, others no change at all, while others record a steady increase.
@@
 Although for their purposes is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
-However, as we've noted before, Corak and Waldfogel (2015) find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as \citet{reardon2011} did.
+However, as we've noted before, \citet{bradbury2015} find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as \citet{reardon2011} did.
 
@@
 The explanation, although very debated, has been gaining much support in recent years.
-In countries with high levels of curricular differentiation, the transition from early schooling into the tracking system has been found to increase inequality of learning (Wossman and Hanushek diff-in-diff).
-Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs (Van der Werfhorst and Mijs 2011).
+In countries with high levels of curricular differentiation, the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}.
+Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}.
 Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as elimination tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
@@
 First, we concentrate on the evolution of the gap for only 15 year olds.
-This is different from all the work of Chmielewski and Reardon, in which they pool all ages to get average changes, to study the properties of specific age-groups.
+This is different from all the work of Anna Chmielewski and Sean Reardon, in which they pool all ages to get average changes, to study the properties of specific age-groups.
 As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
@@
 
-The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th. (Micklewright \& Schnepf, 2007)
-
-Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (Mickleweight \& Schnepf did it).
+The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th \citep{micklewright}.
+
+Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
 They also find that France is a lower dispersion country (but in my results this is changing a lot).
@@
 
-Mickleweight \& Schnepf find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
-Mickleweight \& Schnpef find
+\citet{micklewright} find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
+\citet{micklewright} find
 that average achievement is negatively related to dispersion.
@@
 This variable captures the economic status of the family, the closest it can without asking for income information.
-It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature(Anna study about world-wide inequality).
+It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature \citep{anna2016_global.
 PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.
@@
 But one of the most relevant variables for our study is the is a composite SES index created by the PISA team.
-The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI); the highest level of education of the student’s parents, converted into years of schooling; the PISA index of family wealth; the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (Education at a Glance, OECD, Paris, 2002, Glossary).
+The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI); the highest level of education of the student’s parents, converted into years of schooling; the PISA index of family wealth; the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home \citep{oecd_glance_2002}.
 This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable measures across waves.
-In fact, we use the newly-released ESCS index (% http://www.oecd-ilibrary.org/education/pisa-2015-results-volume-i/pisa-2015-technical-background_9789264266490-13-en
-) which rescaled all ESCS indexes to make them suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.}.
+In fact, we use the newly-released ESCS index \citep{pisa_2015_results} which rescaled all ESCS indexes to make them suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.}.
 
@@
 PISA provides plausible values instead of one single indicator of test score performance.
-As explained in the PISA manual (PISA 2012, 2015), these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured.
+As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured.
 Suppose we have µi, the average student test score in mathematics for student i.
@@
 
-Where \begin{math}\epsilon_i\end{math} is the residual for student \begin{math}i\end{math} and the denominator is the root mean square error of the model.
+Where \begin{math}\hat{\epsilon_i\end}{math} is the residual for student \begin{math}i\end{math} and the denominator is the root mean square error of the model.
 
@@
 \begin{equation}
-\hat{Y_i} = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
+Y_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
 \end{equation}

### 2017-06-21 paper/paper.Rnw
Continue writing literature review + formatted sample sizes to have commas

@@
 <<working directory, echo = F>>=
-# opts_knit$set(root.dir = '..')
-
+options(scipen = 213141)
 opts_chunk$set(echo = F,
@@
 
+\section{Introduction}
+
 \section{Literature Review}
-
-Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years.
-The literature on educational inequality has mainly concentrated on the United States \citep{reardon2011} but other international evidence is emerging with other similar findings for other count.
-The United States is usually the case study of interest, firstly, because it is the only country where cognitive testing is very widespread across surveys.
-And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day.
-Using this information, \citep{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
-Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap -waldfogel study).
-The widening of the achievement gap has been happening in parallel to the growth of income inequality.
-Although very suggestive, it is hard to link both things causally.
-
-\citet{reardon2011} finds that the increase has occured predominantely from the 1970's until the 2000's.
-In fact, hard numbers suggest that the gap increased by about ~ 40-50\%.
-The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%.
-Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footnote{Even after weighting appropriately.}.
-Finally, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.
-\citet{bradbury2015} don't find results which are very far away from this.
-They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25.
-\citet{duncan2011} find similarly to the previous studies and confirm a gap of ~ 1.25/1.50 standard deviations.
-
-One important drawback of these studies is that they don't present the uncertainty of this estimate.
-Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
-Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education \citep{breen2007}.
+% You have a bunch of citations to take from here
+Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years.
+Are we destined to belong to the same social status as our parents?
+Is meritocracy over, if it ever existed?
+These questions have motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century.
+When James Colem(1966) released his famous Coleman report it raised the topic to starlight by suggesting that family SES and student performance are tightly linked.
+Fast-forward today, we've understood the relationship much better, although far from being clear.
+The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school (cite heckman).
+Moreover, these inequalities can be narrowed significantly if the children coming from unfavourable conditions are exposed to high quality learning environments.
+One particular finding coming from (cite heckman skill begets skill) is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\).
+Although this sounds logical, it is often not acknowledged properly.
+
+The implications of this cummulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\).
+That is why the whole debate about narrowing inequality between polarized families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns.
+Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status, is present in virtually all empirical studies of social mobility and inequalities (cite breen and goldthorpe, cite waldfogel what children need, cite bradbury 2015, cite chetty for recent evidence).
+What we don't know is how and why this relationship is formed from such an early age.
+Moreover, many have suggested different explanations and the topic has been very contested (duncan and brooks-gun 1997).
+
+Social mobility has deepend our knowledge on the odds and relationship between intergenerational transmission of SES.
+The field of economics of education has also helped to understand that education is by far the investment that yields the best outcomes not only for the child and his/her family, but to the economy as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth (cite hanushk and woesmann study).
+We've become so intertwined with the subject that we've virtually spent most of our time studying the mechanisms through which this inequality comes about.
+Naturally, we want to do that to be able to reverse it and help disadvantaged children reach their fullest potential.
+Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities.
+In addition, since most research has concentrated on the American case, we haven't been able to learn which countries have the biggest or smallests gaps, and whether they've been narrowing or closing over time.
+
 Virtually in all countries, be it developed or developing, there is inequality of opportunity.
@@
 
-With very recent data, \citet{reardon_portilla} have unexpectedly uncovered a new finding: the reversal of the trend.
-From 1998 until 2010, it seems like the income achievement gap \footnote{The income achievement gap is the difference between the 90th and 10th income percentile in the test of interest} decreased at a rate of 0.01 and 0.008 standard deviations per year.
-In contrast to \citep{reardon2011}, we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something not dramatically different from the previous estimates.
+Taking the Danish example, we learned a great deal about how to improve social mobility by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility for less advantaged children (Bjorklund and Jantti, 2009; Jaeger and Holm, 2007).
+But the important finding came when we discovered the main reason behind their social elevator: its' educational system.
+For example, research by Esping and Waldfogel and Gosta's phd student on Denmark, show that the Danish early education system is particularly efficient at giving free high quality education, regardless of socio-economic status.
+Moreover, it tracks students into different curricula very late compared to other Europen countries Europe (age 16).
+The importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
+
+Recent research on educational inequality has found that achievement gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years.
+Literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape.
+The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940.
+This trend has allowed researchers to study achievement gaps in a lengthy period of time.
+In fact, using this information, \citep{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
+Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap -waldfogel study).
+Surprisingly, the gaps have reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap.
+The widening of the achievement gap has been happening in parallel to the growth of income inequality, although very suggestive, it is hard to link both things causally.
+However, a recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979.
+If income inequality would be a strong a direct function of the achievement gap, we should expect for the white-black achievement gap to be widening rather than narrowing, as it is now (Magnuson and Waldfogel 2008) \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
+
+\citet{reardon2011} finds that the increase has occured predominantely from the 1970's until the 2000's.
+In fact, the hard numbers suggest that the gap widened by ~ 40-50\%.
+The author also estimates the rate of change using data from the 1940's and finds an even higher increase of ~ 75\%.
+Given that the studies before 1970 are less reliablle in terms of comparability and sampling design, the author computes all results for both before/after 1970.
+To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2010.
+Using longitudinal data, \citet{bradbury2015} find similar results.
+Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25.
+Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of ~ 1.25/1.50 standard deviations.
+\footnote{One important drawback of these studies is that they don't present the uncertainty of this estimates.
+Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
+It could be that the gap is at ~ 1.25 standard deviations, varying up to 1.75 and down to 0.80.
+Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.}
+
+In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend.
+This follow-up study concentrated solely on the Kindergarten gap in which they took a snapshot of the achievement gap in Kindergarten in 1998, 2006 and 2010 and found that the 90th/10th income gap in readiness closed modestly.
+Furthermore, using data from fall and spring in the same kindergarden year, it seems like the gap in narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010.
+In contrast to \citep{reardon2011}, we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates.
+Their results not only hold for the income achievement gap, but they also recorded a particular decline in the white-hispanic gap.
+
 The American situation is a very distinctive one relative to Europe.
-The reasons why we're seeing a reversal in the trend could be numerous and should be studied very close to context.
+The reasons the authors find a reversal in the trend could be numerous and should be studied very close.
 For example, the American educational system lacks any formal curricular differentiation \emph{à la} European style.
-In contrast, \citet{reardon2011} suggest that the reversal is likely due to the high increase of preschool enrollment.
-They suggest that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
-
-Some authors have taken this analysis to an international context in order to discover between-country trends.
-The comprehensive work of \citet{bradbury2015} performs a comparative analysis of Australia, United Kingdom, United States and Canada.
-Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is the only study that uses panel data to study achievement gaps, let alone to do this between countries}.
-The core finding behind the book is that the American achievement gap is much wider than in any other comparison countries.
-As mentioned earlier, they find that once the achievement gap is present in early school entry, it doesn't seem to narrow very much over the life course.
-They find that the achievement gap is very stable all the way from early education to secondary school (although their studies cover up until age 14).
-
-% You left documenting the bibliography here
+It is true nonetheless that \citet{reardon2011} did not cover from the period 2000 onwards, which is specifically the trend they studied.
+In contrast, \citet{reardon2011} suggests that the reversal is likely due to the high increase of preschool enrollment.
+They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
+This explanation still needs further confirmation.
+
+Motivated by these recent results, some authors have taken this analysis to an international context in order to discover between-country trends.
+The unique work of \citet{bradbury2015} employs a unique comparative analysis of Australia, United Kingdom, United States and Canada.
+Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is not only the first study that uses panel data to study achievement gaps, but to also do it between countries}.
+The core finding behind the book is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada.
+They find that once the achievement gap is present in early school entry, it doesn't seem to narrow or neither widen very much over the life course.
+In fact, they calculate that the schooling environment once they enter school can only explain about ~ 30\%-40\% of the high school SES gap.
+This suggests that once the achievement gap widens before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous RCT's show that high quality schooling can indeed help ease the gap, in some instances even close it.}.
+Once exception is the United Kingdom, which they find closed the gap in early primary years.
+This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health.
+
 One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula.
-These four countries have no major jump in school selection, quite the opposite to the European average countries.
+These four countries have no major jump in school selection, quite the opposite to the average European country.
 A thorough review by \citet{werfhorst_mijs} sheds some light on the subject.
 First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps.
-Hanushek and Woessman (tracking study), use the TIMSS and PISA surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
-Moreover, \citet{micklewright} using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
+\citet{hanushek_woesmann_tracking}, use the Progress in International Reading Literacy Study (PIRLS), the Trends in International Mathematics and Science Study (TIMSS) and the Programme for International Student Assessment (PISA) surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
+The results suggest that early tracking increases educational inequality.
+While less clear, there is also a tendency for early tracking to reduce mean performance.
+Moreover, \citet{micklewright} using similar data but a different empirical strategy, finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
 In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
 
-Despite their high level of rigurosity, their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
+Another limitation of their study is that their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
 Their findings are very reliable but should be taken as suggestive at best.
-For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citep{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries using similar studies.
-These studies tackle a completely different question from the above, but they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
-
-\citet{anna2016}, using the Programme for International Student Assessment (PISA), the Trends in International Mathematics and Science Study (TIMSS) and PIRLS (the Progress in International Reading Literacy Study), the assess whether there are patterns of cross-national variation in the achievement gap.
-In other words, is the size of achievement gap different for many countries?
-Although there is potential to study countries over time using these surveys, they choose not to.
-They do so because their question of interest (income categories) was only asked in three waves; in the end the only had three countries repeated in all three waves.
-
+For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citep{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries using much more comparable studies.
+These studies tackle a completely different question from the above, namely to use study cross-sectional differences between many countries, instead of over-time analysis of student gaps.
+However, they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
+
+\citet{anna2016}, again using PIRLS, TIMSS and PISA, assess whether there are patterns of cross-national variation in the achievement gap.
+In other words, does the achievement gap differ between countries?
 Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
-In comparison to the literature on achievement gaps, they find that the U.S has a gap a little over 1 standard deviation in the year 2005 (for 15 year olds) and Germany has a decreasing gap from around ~ 1.25 to ~ 1 standard deviation.
-They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, etc ...
-The correlations are indeed very suggestive but clearly we must be extremely cautios in drawing causal changes from these two variables.
-But it is important to stress that their design is quite different from the work of \citet{reardon2011}.
-\citet{reardon2011} takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds.
-Although the author adjusts for age, the trends can't be generalized to age-specific, but rather in overall terms.} and makes them comparable across time.
-This gives the author a very long time series to build a reliable achievement gap (over 40 years).
-In their study, \citet{anna2016} change the aim of the study to model between-country differences from a cross-sectional perspective.
+In comparison to the literature on achievement gaps, they find that the U.S has a gap of ~ 1.20 SD in 2001 which increase to around ~ 1.30 in the year 2006 while Germany has a decreasing gap from around ~ 1.25 to ~ 1 standard deviation in the same year-span.
+However, these numbers vary a lot and carry a great deal of overlapping uncertainty.
+
+They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, among other things.
+The correlations are very suggestive  as explanatory mechanisms but they are very cautios in drawing causality, specially because they are all reasonably correlated.
+However, one interesting question is whether these country gaps have evolved over time.
+With their data, they only have 3 countries which are present in all waves but they also have very few waves because their question of interest (income categories) was only asked in three waves.
+That is why their study is more about between country gaps rather than the magnitude and evolution of the gaps.
 
 \citet{anna2016_global}, building on the work of \citet{anna2016} and \citet{reardon_portilla} pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries.
-With over 50 years of data, and over 100 countries, she finds that there seems to be a general world increase in the achievement gap.
-However, once she disentangles the relationship by country, she finds a reasonable amount of heterogeneity, with some countries seeing the achievement gap closing, others no change at all, while others record a steady increase.
+With over 50 years of data, and over 100 countries, she finds that there seems to be a general pattern of increasing achievement gap.
+However, once she disentangles the relationship by country, she finds a sizable amount of heterogeneity, with some countries seeing the achievement gap narrowing, others no change at all, while others record a steady increase.
+This is revealing because it doesn't really pay off to look at general average once each country has their own distinctive gap and evolution.
+Another important finding is that the achievement gap is clearly no universal and should be studied in context.
+
 One clear limitation of their study (as well as \citet{reardon2011}) is that the adjust for the age of each child in all studies.
-Although for their purposes is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
+Although for their modeling purposes this is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
 However, as we've noted before, \citet{bradbury2015} find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as \citet{reardon2011} did.
-
-The evolution of High/Low SES gaps for preschool children might be much less marked than the same gap for high school children.
+We clearly see in \citet{reardon_portilla} that paying attention to age-specific gaps varies reasonably even in short periods of time.
+
+In fact, the evolution of high/low SES gaps for preschool children might be much less marked than the same gap for high school children.
 The explanation, although very debated, has been gaining much support in recent years.
@@
 
-With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the High/Low SES achievement gap in the past 15 years for all PISA participant countries.
-This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for each country.
+With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries.
+This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for each developed countries.
 The advantages of this study are numerous.
-First, we concentrate on the evolution of the gap for only 15 year olds.
-This is different from all the work of Anna Chmielewski and Sean Reardon, in which they pool all ages to get average changes, to study the properties of specific age-groups.
+First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the Kindergarten gap in \citet{reardon_portilla}.
 As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
@@
 
-The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th \citep{micklewright}.
-
-Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
-They also find that France is a lower dispersion country (but in my results this is changing a lot).
-They also find that Germany and US are very dispersed countries (quite the opposite)
-
-\citet{micklewright} find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
-\citet{micklewright} find
-that average achievement is negatively related to dispersion.
-
+Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
 Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
-
-Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
 Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
 Talk more about how countries with high social mobility has been linked to smallest achievement gaps
-
 Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.
@@
 <<country_sample_numeric_vec>>=
-country_rows <- map_dbl(adapted_year_data, nrow)
+country_rows <- 
+  map_dbl(adapted_year_data, nrow) %>%
+  format(big.mark = ",")
 @
@@
 This variable captures the economic status of the family, the closest it can without asking for income information.
-It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature \citep{anna2016_global.
+It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature \citep{anna2016_global}.
 PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.
@@
 As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured.
-Suppose we have µi, the average student test score in mathematics for student i.
-Instead of estimating µi alone, plausible values estimate a distribution of possible µ’s, for student i, together with the likelihood of each µi based on the respondents answers on the test.
-This is defined as the posterior distributions of µ’s for student i.
-The reason why we use this procedure is because estimating a single estimate µi is plagued with measurement error, among other types of bias (see Wu, 2005).
+Suppose we have \(\mu_i\), the average student test score in mathematics for student \(i\).
+Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\), for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test.
+This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\).
+The reason why we use this procedure is because estimating a single estimate \(\mu_i\) is plagued with measurement error, among other types of bias \citep[see][]{wu2005}.
 The plausible values in the PIAAC data are effectively five (ten for PISA 2015) random draws from this distribution.
@@
 The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries.
-To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps(Fryer and Levitt 2004).
+To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004}.
 The actual calculation of the achievement gap varies substantially and different strategies have been implemented.
-For example, Mickleweight \& Funsche calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution.
+For example, \citet{micklewright} calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution.
 Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy.
-Guan Kung Saw (2015), for instance, used parental education as a proxy of SES, whereas Soo-yong Byun and Kyung-keun Kim (2010) use a similar SES index as ours, but created by them.
-
-\citet{reardon2011} and subsequently (all reardon papers), used a different method developed by \citet{reardon2011} which we partially adopt adopt in this paper.
+\citet{saw2016}, for instance, used parental education as a proxy of SES, whereas \citet{byun2010} use a similar SES index as ours, but created by them.
+
+\citet{reardon_portilla, anna2016, anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt adopt in this paper.
 SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.
+The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th \citep{micklewright} while the literature has been following the standard set by \citet{reardon2011} to use 90th/10th percentiles.
 
@@
 
-Where \begin{math}\hat{\epsilon_i\end}{math} is the residual for student \begin{math}i\end{math} and the denominator is the root mean square error of the model.
+Where \begin{math}\hat{\epsilon_i}\end{math} is the residual for student \begin{math}i\end{math} and the denominator is the root mean square error of the model.
 
@@
 \subsection{Results}
+
+Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
+They also find that France is a lower dispersion country (but in my results this is changing a lot).
+They also find that Germany and US are very dispersed countries (quite the opposite)
+
+\citet{micklewright} find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
+\citet{micklewright} find
+that average achievement is negatively related to dispersion.
+
+It seems that social origin effects have actually increased in France  (Ballarino and Bernardi, 2016)
 

### 2017-06-21 paper/paper.tex
Continue writing literature review + formatted sample sizes to have commas

@@
 
+\section{Introduction}
+
 \section{Literature Review}
+% You have a bunch of citations to take from here
+Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years.
+Are we destined to belong to the same social status as our parents?
+Is meritocracy over, if it ever existed?
+These questions have motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century.
+When James Colem(1966) released his famous Coleman report it raised the topic to starlight by suggesting that family SES and student performance are tightly linked.
+Fast-forward today, we've understood the relationship much better, although far from being clear.
+The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school (cite heckman).
+Moreover, these inequalities can be narrowed significantly if the children coming from unfavourable conditions are exposed to high quality learning environments.
+One particular finding coming from (cite heckman skill begets skill) is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\).
+Although this sounds logical, it is often not acknowledged properly.
 
-Recent research on educational inequality has found that differences in test peformance between High-SES and Low-SES kids has been growing quickly over the years.
-The literature on educational inequality has mainly concentrated on the United States \citep{reardon2011} but other international evidence is emerging with other similar findings for other count.
-The United States is usually the case study of interest, firstly, because it is the only country where cognitive testing is very widespread across surveys.
-And secondly, this trend has allowed to have testing records as early as 1940 (check reardon) until present day.
-Using this information, \citep{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
-Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap -waldfogel study).
-The widening of the achievement gap has been happening in parallel to the growth of income inequality.
-Although very suggestive, it is hard to link both things causally.
+The implications of this cummulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\).
+That is why the whole debate about narrowing inequality between polarized families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns.
+Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status, is present in virtually all empirical studies of social mobility and inequalities (cite breen and goldthorpe, cite waldfogel what children need, cite bradbury 2015, cite chetty for recent evidence).
+What we don't know is how and why this relationship is formed from such an early age.
+Moreover, many have suggested different explanations and the topic has been very contested (duncan and brooks-gun 1997).
 
-\citet{reardon2011} finds that the increase has occured predominantely from the 1970's until the 2000's.
-In fact, hard numbers suggest that the gap increased by about ~ 40-50\%.
-The author also estimates the rate of change using data from the 1940's and finds an even higher increase of about ~ 75\%.
-Since the studies before the 1970's lack a lot of rigor in comparability and sampling, the author computes everything for the sample before/after 1970's separately \footnote{Even after weighting appropriately.}.
-Finally, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations.
-\citet{bradbury2015} don't find results which are very far away from this.
-They find that for 14 year olds in the United States, the gap is above 1 SD but lower than 1.25.
-\citet{duncan2011} find similarly to the previous studies and confirm a gap of ~ 1.25/1.50 standard deviations.
+Social mobility has deepend our knowledge on the odds and relationship between intergenerational transmission of SES.
+The field of economics of education has also helped to understand that education is by far the investment that yields the best outcomes not only for the child and his/her family, but to the economy as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth (cite hanushk and woesmann study).
+We've become so intertwined with the subject that we've virtually spent most of our time studying the mechanisms through which this inequality comes about.
+Naturally, we want to do that to be able to reverse it and help disadvantaged children reach their fullest potential.
+Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities.
+In addition, since most research has concentrated on the American case, we haven't been able to learn which countries have the biggest or smallests gaps, and whether they've been narrowing or closing over time.
 
-One important drawback of these studies is that they don't present the uncertainty of this estimate.
-Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
-Despite the recent surge in understanding achievement gaps as a function of social background, sociologists have already amassed a great deal of findings concerning inter-generational inequalities, one notable example being that it's very clear that large inequalities in parent's education tend to reproduce large inequalities in children's education \citep{breen2007}.
 Virtually in all countries, be it developed or developing, there is inequality of opportunity.
@@
 
-With very recent data, \citet{reardon_portilla} have unexpectedly uncovered a new finding: the reversal of the trend.
-From 1998 until 2010, it seems like the income achievement gap \footnote{The income achievement gap is the difference between the 90th and 10th income percentile in the test of interest} decreased at a rate of 0.01 and 0.008 standard deviations per year.
-In contrast to \citep{reardon2011}, we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something not dramatically different from the previous estimates.
+Taking the Danish example, we learned a great deal about how to improve social mobility by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility for less advantaged children (Bjorklund and Jantti, 2009; Jaeger and Holm, 2007).
+But the important finding came when we discovered the main reason behind their social elevator: its' educational system.
+For example, research by Esping and Waldfogel and Gosta's phd student on Denmark, show that the Danish early education system is particularly efficient at giving free high quality education, regardless of socio-economic status.
+Moreover, it tracks students into different curricula very late compared to other Europen countries Europe (age 16).
+The importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
+
+Recent research on educational inequality has found that achievement gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years.
+Literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape.
+The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940.
+This trend has allowed researchers to study achievement gaps in a lengthy period of time.
+In fact, using this information, \citep{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
+Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap -waldfogel study).
+Surprisingly, the gaps have reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap.
+The widening of the achievement gap has been happening in parallel to the growth of income inequality, although very suggestive, it is hard to link both things causally.
+However, a recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979.
+If income inequality would be a strong a direct function of the achievement gap, we should expect for the white-black achievement gap to be widening rather than narrowing, as it is now (Magnuson and Waldfogel 2008) \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
+
+\citet{reardon2011} finds that the increase has occured predominantely from the 1970's until the 2000's.
+In fact, the hard numbers suggest that the gap widened by ~ 40-50\%.
+The author also estimates the rate of change using data from the 1940's and finds an even higher increase of ~ 75\%.
+Given that the studies before 1970 are less reliablle in terms of comparability and sampling design, the author computes all results for both before/after 1970.
+To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2010.
+Using longitudinal data, \citet{bradbury2015} find similar results.
+Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25.
+Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of ~ 1.25/1.50 standard deviations.
+\footnote{One important drawback of these studies is that they don't present the uncertainty of this estimates.
+Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
+It could be that the gap is at ~ 1.25 standard deviations, varying up to 1.75 and down to 0.80.
+Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.}
+
+In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend.
+This follow-up study concentrated solely on the Kindergarten gap in which they took a snapshot of the achievement gap in Kindergarten in 1998, 2006 and 2010 and found that the 90th/10th income gap in readiness closed modestly.
+Furthermore, using data from fall and spring in the same kindergarden year, it seems like the gap in narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010.
+In contrast to \citep{reardon2011}, we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates.
+Their results not only hold for the income achievement gap, but they also recorded a particular decline in the white-hispanic gap.
+
 The American situation is a very distinctive one relative to Europe.
-The reasons why we're seeing a reversal in the trend could be numerous and should be studied very close to context.
+The reasons the authors find a reversal in the trend could be numerous and should be studied very close.
 For example, the American educational system lacks any formal curricular differentiation \emph{à la} European style.
-In contrast, \citet{reardon2011} suggest that the reversal is likely due to the high increase of preschool enrollment.
-They suggest that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
+It is true nonetheless that \citet{reardon2011} did not cover from the period 2000 onwards, which is specifically the trend they studied.
+In contrast, \citet{reardon2011} suggests that the reversal is likely due to the high increase of preschool enrollment.
+They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
+This explanation still needs further confirmation.
 
-Some authors have taken this analysis to an international context in order to discover between-country trends.
-The comprehensive work of \citet{bradbury2015} performs a comparative analysis of Australia, United Kingdom, United States and Canada.
-Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is the only study that uses panel data to study achievement gaps, let alone to do this between countries}.
-The core finding behind the book is that the American achievement gap is much wider than in any other comparison countries.
-As mentioned earlier, they find that once the achievement gap is present in early school entry, it doesn't seem to narrow very much over the life course.
-They find that the achievement gap is very stable all the way from early education to secondary school (although their studies cover up until age 14).
+Motivated by these recent results, some authors have taken this analysis to an international context in order to discover between-country trends.
+The unique work of \citet{bradbury2015} employs a unique comparative analysis of Australia, United Kingdom, United States and Canada.
+Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is not only the first study that uses panel data to study achievement gaps, but to also do it between countries}.
+The core finding behind the book is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada.
+They find that once the achievement gap is present in early school entry, it doesn't seem to narrow or neither widen very much over the life course.
+In fact, they calculate that the schooling environment once they enter school can only explain about ~ 30\%-40\% of the high school SES gap.
+This suggests that once the achievement gap widens before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous RCT's show that high quality schooling can indeed help ease the gap, in some instances even close it.}.
+Once exception is the United Kingdom, which they find closed the gap in early primary years.
+This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health.
 
-% You left documenting the bibliography here
 One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula.
-These four countries have no major jump in school selection, quite the opposite to the European average countries.
+These four countries have no major jump in school selection, quite the opposite to the average European country.
 A thorough review by \citet{werfhorst_mijs} sheds some light on the subject.
 First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps.
-Hanushek and Woessman (tracking study), use the TIMSS and PISA surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
-Moreover, Mickleweight \& Schnepf (2007) using similar data but different design finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
+\citet{hanushek_woesmann_tracking}, use the Progress in International Reading Literacy Study (PIRLS), the Trends in International Mathematics and Science Study (TIMSS) and the Programme for International Student Assessment (PISA) surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
+The results suggest that early tracking increases educational inequality.
+While less clear, there is also a tendency for early tracking to reduce mean performance.
+Moreover, \citet{micklewright} using similar data but a different empirical strategy, finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
 In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
 
-Despite their high level of rigurosity, their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
+Another limitation of their study is that their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
 Their findings are very reliable but should be taken as suggestive at best.
-For this reason we should also pay particular attention to studies such as Chmielewski and Reardon (2016) and Chmielewski (2017) which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries using similar studies.
-These studies tackle a completely different question from the above, but they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
+For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citep{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries using much more comparable studies.
+These studies tackle a completely different question from the above, namely to use study cross-sectional differences between many countries, instead of over-time analysis of student gaps.
+However, they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
 
-Chmielewski and Reardon (2016), using the Programme for International Student Assessment (PISA), the Trends in International Mathematics and Science Study (TIMSS) and PIRLS (the Progress in International Reading Literacy Study), the assess whether there are patterns of cross-national variation in the achievement gap.
-In other words, is the size of achievement gap different for many countries?
-Although there is potential to study countries over time using these surveys, they choose not to.
-They do so because their question of interest (income categories) was only asked in three waves; in the end the only had three countries repeated in all three waves.
+\citet{anna2016}, again using PIRLS, TIMSS and PISA, assess whether there are patterns of cross-national variation in the achievement gap.
+In other words, does the achievement gap differ between countries?
+Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
+In comparison to the literature on achievement gaps, they find that the U.S has a gap of ~ 1.20 SD in 2001 which increase to around ~ 1.30 in the year 2006 while Germany has a decreasing gap from around ~ 1.25 to ~ 1 standard deviation in the same year-span.
+However, these numbers vary a lot and carry a great deal of overlapping uncertainty.
 
-Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
-In comparison to the literature on achievement gaps, they find that the U.S has a gap a little over 1 standard deviation in the year 2005 (for 15 year olds) and Germany has a decreasing gap from around ~ 1.25 to ~ 1 standard deviation.
-They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, etc ...
-The correlations are indeed very suggestive but clearly we must be extremely cautios in drawing causal changes from these two variables.
-But it is important to stress that their design is quite different from the work of \citet{reardon2011}.
-\citet{reardon2011} takes studies in the U.S starting from the 1940's until 2015 \footnote{Each study is independent of each other meaning that the it might've been given to 6 year olds as well as to 12 year olds.
-Although the author adjusts for age, the trends can't be generalized to age-specific, but rather in overall terms.} and makes them comparable across time.
-This gives the author a very long time series to build a reliable achievement gap (over 40 years).
-In their study, Chmielewski and Reardon (2016) change the aim of the study to model between-country differences from a cross-sectional perspective.
+They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, among other things.
+The correlations are very suggestive  as explanatory mechanisms but they are very cautios in drawing causality, specially because they are all reasonably correlated.
+However, one interesting question is whether these country gaps have evolved over time.
+With their data, they only have 3 countries which are present in all waves but they also have very few waves because their question of interest (income categories) was only asked in three waves.
+That is why their study is more about between country gaps rather than the magnitude and evolution of the gaps.
 
-Chmielewski (2017), building on the work of Chmielewski and Reardon (2016) and Reardon and Portilla (2015) pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries.
-With over 50 years of data, and over 100 countries, Chmielewski (2017) finds that there seems to be a general world increase in the achievement gap.
-However, once she disentangles the relationship by country, she finds a reasonable amount of heterogeneity, with some countries seeing the achievement gap closing, others no change at all, while others record a steady increase.
+\citet{anna2016_global}, building on the work of \citet{anna2016} and \citet{reardon_portilla} pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries.
+With over 50 years of data, and over 100 countries, she finds that there seems to be a general pattern of increasing achievement gap.
+However, once she disentangles the relationship by country, she finds a sizable amount of heterogeneity, with some countries seeing the achievement gap narrowing, others no change at all, while others record a steady increase.
+This is revealing because it doesn't really pay off to look at general average once each country has their own distinctive gap and evolution.
+Another important finding is that the achievement gap is clearly no universal and should be studied in context.
+
 One clear limitation of their study (as well as \citet{reardon2011}) is that the adjust for the age of each child in all studies.
-Although for their purposes is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
-However, as we've noted before, Corak and Waldfogel (2015) find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as \citet{reardon2011} did.
+Although for their modeling purposes this is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
+However, as we've noted before, \citet{bradbury2015} find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as \citet{reardon2011} did.
+We clearly see in \citet{reardon_portilla} that paying attention to age-specific gaps varies reasonably even in short periods of time.
 
-The evolution of High/Low SES gaps for preschool children might be much less marked than the same gap for high school children.
+In fact, the evolution of high/low SES gaps for preschool children might be much less marked than the same gap for high school children.
 The explanation, although very debated, has been gaining much support in recent years.
-In countries with high levels of curricular differentiation, the transition from early schooling into the tracking system has been found to increase inequality of learning (Wossman and Hanushek diff-in-diff).
-Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs (Van der Werfhorst and Mijs 2011).
+In countries with high levels of curricular differentiation, the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}.
+Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}.
 Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as elimination tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
 
-With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the High/Low SES achievement gap in the past 15 years for all PISA participant countries.
-This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for each country.
+With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries.
+This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for each developed countries.
 The advantages of this study are numerous.
-First, we concentrate on the evolution of the gap for only 15 year olds.
-This is different from all the work of Chmielewski and Reardon, in which they pool all ages to get average changes, to study the properties of specific age-groups.
+First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the Kindergarten gap in \citet{reardon_portilla}.
 As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
@@
 
-The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th. (Micklewright \& Schnepf, 2007)
-
-Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (Mickleweight \& Schnepf did it).
-They also find that France is a lower dispersion country (but in my results this is changing a lot).
-They also find that Germany and US are very dispersed countries (quite the opposite)
-
-Mickleweight \& Schnepf find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
-Mickleweight \& Schnpef find
-that average achievement is negatively related to dispersion.
-
+Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
 Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
-
-Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
 Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
 Talk more about how countries with high social mobility has been linked to smallest achievement gaps
-
 Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.
@@
 With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests.
-The sample size for PISA 2000 is \ensuremath{1.27388\times 10^{5}}, for PISA 2003, \ensuremath{2.76165\times 10^{5}}, for PISA 2006, \ensuremath{3.9875\times 10^{5}}, for PISA 2009, \ensuremath{5.15958\times 10^{5}}, for PISA 2012, \ensuremath{4.8081\times 10^{5}}, and for PISA 2015 \ensuremath{5.19334\times 10^{5}}.
+The sample size for PISA 2000 is 127,388, for PISA 2003, 276,165, for PISA 2006, 398,750, for PISA 2009, 515,958, for PISA 2012, 480,810, and for PISA 2015 519,334.
 Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance.
@@
 This variable captures the economic status of the family, the closest it can without asking for income information.
-It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature(Anna study about world-wide inequality).
+It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature \citep{anna2016_global}.
 PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.
@@
 PISA provides plausible values instead of one single indicator of test score performance.
-As explained in the PISA manual (PISA 2012, 2015), these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured.
-Suppose we have µi, the average student test score in mathematics for student i.
-Instead of estimating µi alone, plausible values estimate a distribution of possible µ’s, for student i, together with the likelihood of each µi based on the respondents answers on the test.
-This is defined as the posterior distributions of µ’s for student i.
-The reason why we use this procedure is because estimating a single estimate µi is plagued with measurement error, among other types of bias (see Wu, 2005).
+As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured.
+Suppose we have \(\mu_i\), the average student test score in mathematics for student \(i\).
+Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\), for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test.
+This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\).
+The reason why we use this procedure is because estimating a single estimate \(\mu_i\) is plagued with measurement error, among other types of bias \citep[see][]{wu2005}.
 The plausible values in the PIAAC data are effectively five (ten for PISA 2015) random draws from this distribution.
@@
 The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries.
-To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps(Fryer and Levitt 2004).
+To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004}.
 The actual calculation of the achievement gap varies substantially and different strategies have been implemented.
-For example, Mickleweight \& Funsche calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution.
+For example, \citet{micklewright} calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution.
 Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy.
-Guan Kung Saw (2015), for instance, used parental education as a proxy of SES, whereas Soo-yong Byun and Kyung-keun Kim (2010) use a similar SES index as ours, but created by them.
+\citet{saw2016}, for instance, used parental education as a proxy of SES, whereas \citet{byun2010} use a similar SES index as ours, but created by them.
 
-\citet{reardon2011} and subsequently (all reardon papers), used a different method developed by \citet{reardon2011} which we partially adopt adopt in this paper.
+\citet{reardon_portilla, anna2016, anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt adopt in this paper.
 SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.
+The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th \citep{micklewright} while the literature has been following the standard set by \citet{reardon2011} to use 90th/10th percentiles.
 
@@
 
-Where \begin{math}\epsilon_i\end{math} is the residual for student \begin{math}i\end{math} and the denominator is the root mean square error of the model.
+Where \begin{math}\hat{\epsilon_i}\end{math} is the residual for student \begin{math}i\end{math} and the denominator is the root mean square error of the model.
 
@@
 
+Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
+They also find that France is a lower dispersion country (but in my results this is changing a lot).
+They also find that Germany and US are very dispersed countries (quite the opposite)
+
+\citet{micklewright} find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
+\citet{micklewright} find
+that average achievement is negatively related to dispersion.
+
+It seems that social origin effects have actually increased in France  (Ballarino and Bernardi, 2016)
+
 {\centering \includegraphics[width=5.5in,height=5in]{figure/correlation_incomeineq-1}

### 2017-06-22 paper/paper.Rnw
Finished reviewing research and data sections

@@
 
-With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries.
-This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for each developed countries.
-The advantages of this study are numerous.
-First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the Kindergarten gap in \citet{reardon_portilla}.
-As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
-Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.
-
 Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
@@
 
-This paper is interested in two questions.
-As explained in the literature review, we already know that U.S achievement gaps have been growing over time, and some international evidence suggests that it might be increasing in other countries as well.
-All of these studies pool all students together from different ages and estimate global trends.
-
-We take a different approach and we want to study age-specific trends in achievement gaps.
-We're interested in studying a) the size of the achievement in a comparative perspective and its relationship to overall achievement levels, b) which countries are experiencing changes in the achievement gap, c) the rate at which each country-gap is widening/narrowing and d) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
+With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries.
+This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for many countries.
+The advantages of this study are twofold.
+
+First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the Kindergarten gap in \citet{reardon_portilla}.
+As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
+Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.
+
+Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States.
+This will help asses the mangitude of the inequality as well as the rate of growth/decline.
+Thirdly, this paper not only documents the size and changes of the SES gap, but studies with particular attention the source of the changes.
+That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.
+
+More formally, We're interested in studying
+a) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
+b) Which countries are experiencing changes in the achievement gap;
+c) the rate at which each country-gap is widening/narrowing and;
+d) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
 
@@
 
-a) We want to test the notion that better performing countries have lower leves of achievement gap.
+a) We want to test the notion that better performing countries have a smaller degree of inequality than other countries.
 \citep{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
-This pattern is not so obvious.
+This pattern is not so obvious, however.
 For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score.
-But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with widening achievement gap.
+But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.
 
@@
 After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
-This lends credibility that our estimations will show variation over time.
-As usual, their estimations are only for children going from 4th to 8th grade.
-We concentrate on age-specific groups, in this case, 15 year olds.
-
-c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year.
+This shows that it is important to study the changes in the achievement gap over time.
+This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.
+
+c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year available.
 This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
@@
 
-In order the study the above mentioned questions I will be using the Programme for International Student Assessment (PISA).
+To investigate the above mentioned questions I will be using the Programme for International Student Assessment (PISA).
 PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students.
-Currrently, PISA has six waves starting in the year 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 countries, both developed and developing ones.
+Currrently, PISA has six waves starting which started in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 countries, both developed and developing ones.
 
@@
 With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests.
-The sample size for PISA 2000 is \Sexpr{country_rows[1]}, for PISA 2003, \Sexpr{country_rows[2]}, for PISA 2006, \Sexpr{country_rows[3]}, for PISA 2009, \Sexpr{country_rows[4]}, for PISA 2012, \Sexpr{country_rows[5]}, and for PISA 2015 \Sexpr{country_rows[6]}.
+The sample size for PISA 2000 is \Sexpr{country_rows[1]}, \Sexpr{country_rows[2]} for PISA 2003, \Sexpr{country_rows[3]} for PISA 2006, \Sexpr{country_rows[4]} for PISA 2009, \Sexpr{country_rows[5]} for PISA 2012, and \Sexpr{country_rows[6]} for PISA 2015.
 Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance.
-With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough time to visualize changes in the structure of an educational system.
-
-In order to measure family background PISA includes several variables.
-Firstly, the ask student's their parent's educational level.
-Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}.
+With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system.
+None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.
+
+To identify a family's socio-economic status PISA collects several variables that measure different dimensions.
+Classically, they ask student's their parent's educational level.
+Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}(cite more here).
 This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding.
-Moreover, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time.
-For example, until PISA 2009, the ISCED 1997 classification was currently used, but the next wave already used the new ISCED 2011 classification.
-Both these classification schemes have equivalent look-up tables, yet this requires a detailed inspection of the codings.
+In spite of this, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time.
+For example, until PISA 2009, the preferred framework was ISCED 1997, whereas the next wave switched to the newly developed ISCED 2011 classification.
+Both these classification schemes have equivalent look-up tables, but this requires a detailed inspection of the codings.
+Aside from all these limitations, this is very reliable and important variable that measures an important factor of socio-economic status.
 
@@
 This variable captures the economic status of the family, the closest it can without asking for income information.
+This index variable was developed by Ganzeboom (cite) and it attempts to measure occupational status using a continuous measure.
+This indicator has been found to ourperform classical occupational classifications such as the Erikson-Goldthorpe-Portocarero schema (cite).
 It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature \citep{anna2016_global}.
@@
 
-But one of the most relevant variables for our study is the is a composite SES index created by the PISA team.
-The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI); the highest level of education of the student’s parents, converted into years of schooling; the PISA index of family wealth; the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home \citep{oecd_glance_2002}.
-This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable measures across waves.
-In fact, we use the newly-released ESCS index \citep{pisa_2015_results} which rescaled all ESCS indexes to make them suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.}.
-
-Aside from SES, the other most relevant variables are test scores.
-PISA provides plausible values instead of one single indicator of test score performance.
-As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured.
-Suppose we have \(\mu_i\), the average student test score in mathematics for student \(i\).
-Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\), for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test.
+Yet one of the most relevant variables for our study is a composite SES index created by the PISA team.
+The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measured the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}.
+This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable metrics across waves.
+To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.
+
+Aside from SES, the other most relevant variables are test scores for Mathematics, Literacy and Science.
+PISA does not provide a single test result for each respondent.
+Instead, it provides a \emph{series} of 'plausible values' that the child could actually score.
+As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured (the true distribution of the possible scores a student can achieve).
+Suppose we have \(\mu_i\), the average student test score in Mathematics for student \(i\).
+Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\) for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test.
 This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\).
 The reason why we use this procedure is because estimating a single estimate \(\mu_i\) is plagued with measurement error, among other types of bias \citep[see][]{wu2005}.
-The plausible values in the PIAAC data are effectively five (ten for PISA 2015) random draws from this distribution.
-These variable is are continuous, ranging from 0 to 500, with a mean of 250.
+The number of plausible values for PISA waves are usually five (although ten for PISA 2015) random draws from this distribution.
+In practice each student has 5 scores for each test, that resembles this distribution.
+They are continuous, ranging from 0 to 500, with a mean of 250.
 
@@
 The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries.
-To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004}.
+To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004, hanushek_woesmann_tracking, saw2016, bradbury2015, byun2010}.
 The actual calculation of the achievement gap varies substantially and different strategies have been implemented.
-For example, \citet{micklewright} calculate the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution.
+For example, \citet{micklewright} calculates the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution.
 Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy.
@@
 
-\citet{reardon_portilla, anna2016, anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt adopt in this paper.
+\citet{reardon_portilla, anna2016, anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt in this paper.
 SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.
@@
 
-We start by standardizing achievement for each country, year and subject.
-Because PISA does not provide a single achievement indicator, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.
-
-We then fit a linear model
+For each country in each study, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps.
+
+The original strategy of \citet{reardon2011} is as follows: first, achievement is standardized (see below for a mathematical explanation of the standardization).
+We then use it to calculate the mean achievement (and standard error) for each category of the SES variable of interest (parent's education, income categories, etc..).
+"Category means are plotted at their percentile ranks and cubic models are fit through the points using weighted least squares.
+Finally, achievement at each country’s 90th and 10th SES percentiles is interpolated from the model" \citep{anna2016_global}.
+The result is a SES gap from an ordinal variable of interest.
+
+As mentioned before, PISA does not provide a single achievement indicator.
+Instead, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.
+
+To standardize the test score we fit a linear model
 
@@
 
+This new standardized variable has a mean of zero.
 Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves.
-However, if the variance of academic achievement changes over time, then standardizing the overall score at each wave of testing confounds changes in the gap with changes in the variance of test scores.
-That is, by standardizing we're forcing the variance of test scores to be zero across all waves.
-But if the true variance of academic achievement grows over time, then the estimated trend in the achievement gaps will be underestimated, and vice versa.
+However, if the variance of academic achievement changes over time, then standardizing the overall score at each country-wave pair actually makes the variable more biased.
+That is, by standardizing we're forcing the standard deviation of test scores to be zero across all waves.
+But if the true deviations of the median academic achievement changes over time, then the estimated trend in the SES gaps will be underestimated, or vice versa.
 
@@
 % I'm showing standard deviations, is it better to show variance?
-<<variance_pisa, cache = TRUE>>=
+<<variance_pisa, cache = TRUE, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
 
@@
 Inspection of the data suggests that it's something we shouldn't be deeply concerned with.
-The standard deviation of each wave seems to be following a very similar pattern with not so drastic exception of the year 2000.
+The standard deviation of each wave seems to be following a very similar pattern with a not so drastic exception of the year 2000.
 
@@
 where \begin{math}r\end{math} is the reliability score of the wave.
-\footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair (Anna analysis of wordwide gap).
+\footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}.
 Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own.
@@
 
-After constructing the adjusted test score measurement, we estimate the SES dummy by calculating the weighted 90th and 10th percentile scores for the SES composite index, and then generating a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.
+After constructing the adjusted test score measurement, we estimate the SES dummy by calculating the weighted 90th and 10th quantiles for the SES composite index, and then generating a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.
 
@@
 where \(SES_i\) is whether the student is in the top/bottom SES dummy and we allow both \(\alpha_i\) and \(\beta_i\) to vary by country \(j\).
-We implement this separately for each wave and weight by the student sample weights.
+We implement this separately for each wave and weight by the wave-specific student sample weights.
 
@@
 
+\section{Analysis}
+
+\subsection{Descriptives}
+
 % Show the properties of the ESCS dummy variable
@@
 % % of ISEI
-
-\section{Analysis}
-
-\subsection{Descriptives}
 

### 2017-06-22 paper/paper.Rnw
Updated countries for graphs + updated graph - first results writeup

@@
 
-  countries <- c("Australia",
+  countries <- c("Finland",
+                 "France",
+                 "New Zealand",
+                 "Austria",
+                 "Australia",
+                 "Sweden",
+                 "Czech Republic",
+                 "Canada",
+                 "Hungary",
+                 "Iceland", "Netherlands",
+                 "Spain",
+                 "Belgium",
+                 "Italy", "Norway",
+                 "United Kingdom",
+                 "Greece",
+                 "Denmark",
+                 "Israel",
+                 "Poland",
+                 "United States",
                  "Germany",
-                 "Denmark",
-                 "Spain",
-                 "France",
-                 "Italy",
-                 "Netherlands",
-                 "Sweden",
-                 "Finland",
-                 "United States",
-                 "United Kingdom")
+                 "Turkey",
+                 "Russia")
 @
@@
 
-results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
-results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
-results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
-results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
-results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
-results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
-
-# results_math <- read_rds("./data/delete.Rdata")
-# results_read <- read_rds("./data/delete_read.Rdata")
-# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
-# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
-# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
-# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
+# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
+# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
+# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
+# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
+# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
+# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
+
+results_math <- read_rds("./data/delete.Rdata")
+results_read <- read_rds("./data/delete_read.Rdata")
+results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
+results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
+results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
+results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
 # US is missing for reading
@@
 
-enframer <- function(df) {
+enframer <- function(df, col_name = "name") {
   df %>%
-    enframe() %>%
+    enframe(name = col_name) %>%
     unnest()
@@
   setNames(seq(2000, 2015, 3)) %>%
-  enframer() %>%
+  enframer(name) %>%
   ggplot(aes(name, value)) + geom_col()
@@
 Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible.
+For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.
 
@@
 
+<<escsdummy_charachteristics, cache = TRUE, eval = F>>=
+.x <- adapted_year_data[[1]]
+country <- filter(.x, country == "United States")
+probs <- c(0.1, 0.9)
+
+escs_dummy_creator <- function(df, probs) {
+
+    map(df, function(.x) {
+
+      conf <- if (unique(.x$wave) == "pisa2015") pisa2015_conf else pisa_conf
+      weights_var <- conf$variables$weightFinal
+
+      country_split <- split(.x, .x$country)
+
+      country_list <- map(country_split, function(country) {
+        print(unique(country$country))
+
+        quan <- quantile_missing(country, weights_var, probs)
+
+        # It's very important to create a variable that returns the number of observations of this dummy
+        # For each country. Possibly to weight by the number of observations.
+        country$escs_dummy <-
+          with(country, case_when(escs_trend >= quan[2] ~ 1,
+                                  escs_trend <= quan[1] ~ 0))
+        country
+      })
+
+      .x <-
+        enframe(country_list) %>%
+        unnest(value)
+
+      message(paste(unique(.x$wave), "data ready"))
+
+      .x
+    })
+}
+
+escs_data <- escs_dummy_creator(adapted_year_data, c(0.1, 0.9))
+
+# Make recode the education variable into three categories
+map_lgl(escs_data, ~ "HISEI" %in% names(.x))
+
+escs_data[[1]] %>%
+  select(country, escs_dummy, high_edu_broad, ISEI) %>%
+  filter(country %in% countries) %>%
+  count(country, escs_dummy, high_edu_broad) %>%
+  group_by(country, escs_dummy) %>%
+  mutate(total_n = sum(n),
+         perc = n / total_n * 100) %>%
+  select(country, escs_dummy, high_edu_broad, perc) %>%
+  map_if(is_double, round, 2) %>%
+  as_tibble() %>%
+  filter(!is.na(escs_dummy)) %>%
+  spread(high_edu_broad, perc)
+
+escs_data[[1]] %>%
+  select(country, escs_dummy, high_edu_broad, ISEI) %>%
+  filter(country %in% countries) %>%
+  count(country, escs_dummy, isei_group = cut_number(ISEI, 4)) %>%
+  group_by(country, isei_group) %>%
+  mutate(total_n = sum(n),
+         perc = n / total_n * 100) %>%
+  select(country, escs_dummy, isei_group, perc) %>%
+  map_if(is_double, round, 2) %>%
+  as_tibble() %>%
+  filter(!is.na(escs_dummy)) %>%
+  spread(isei_group, perc) %>%
+  left_join(m, by = c("country", "escs_dummy"))
+
+@
+
 % Figure out how to ouput data frames as latex tables.
 <<sample_size, cache = TRUE>>=
-
 # Get sample counts for each dummy
@@
 @
-
-\subsection{Results}
-
-Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
-They also find that France is a lower dispersion country (but in my results this is changing a lot).
-They also find that Germany and US are very dispersed countries (quite the opposite)
-
-\citet{micklewright} find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
-\citet{micklewright} find
-that average achievement is negatively related to dispersion.
-
-It seems that social origin effects have actually increased in France  (Ballarino and Bernardi, 2016)
 
@@
 
+\subsection{Results}
+
+Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
+
 <<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
-
 # 90/10 gaps acros countries
+diff_increase_fun <- function(df) {
+
+  # Average standard deviation increase
+  data_ready <-
+    df %>%
+    select(wave, country, type_test, difference) %>%
+    group_by(type_test) %>%
+    split(.$country) %>%
+    map(~ {
+      .x <-
+        spread(.x, wave, difference) %>%
+        ungroup()
+
+      year_vars <- sum(map_dbl(.x, is.numeric)) - 1
+      years_subtract <- names(.x)[c(ncol(.x) - year_vars, ncol(.x))]
+      years_subtract <- lapply(years_subtract, as.name)
+
+      last_year <- rlang::new_quosure(years_subtract[[2]], env = .GlobalEnv)
+      first_year <- rlang::new_quosure(years_subtract[[1]], env = .GlobalEnv)
+
+      years_available <-
+        .x %>%
+        gather(year, val, -(country:type_test)) %>%
+        group_by(type_test) %>%
+        summarise(yr_avaible = sum(!is.na(val))) %>%
+        pull(yr_avaible)
+
+      year_sd <-
+        .x %>%
+        gather(year, val, -(country:type_test)) %>%
+        split(.$type_test) %>%
+        map_dbl(~ bootstrapper(.x, mad(val, na.rm = T), B = 100) %>% .[[2, 2]]) %>%
+        round(2) * 100
+
+      .x %>%
+        map_if(is_double, round, 3) %>%
+        as_tibble() %>%
+        transmute(type_test,
+                  country,
+                  diff = round(((!!last_year) - (!!first_year)), 1),
+                  sd_year = year_sd,
+                  diff_lower = diff - 1 * year_sd,
+                  diff_upper = diff + 1 * year_sd,
+                  years_available = years_available)
+    })
+  data_ready
+}
+
+diff_data <-
+  diff_increase_fun(complete_data_topbottom) %>%
+  enframer("country") %>%
+  filter(country %in% countries) %>%
+  select(country, type_test, diff) %>%
+  split(.$type_test) %>%
+  map(~ .x %>% select(-type_test) %>% deframe())
+
+lm_data <- function(df) {
+  lm(log(difference) ~ wave, data = df) %>%
+    broom::tidy() %>%
+    mutate(estimate = exp(estimate))
+}
+
+ordered_cnt <-
   complete_data_topbottom %>%
-  filter(country %in% c("United States", "Netherlands", "France",
-                        "Germany", "Poland", "Finland")) %>%
+  filter(type_test == "math") %>%
+  select(wave, country, difference) %>%
+  filter(country %in% countries) %>%
+  split(.$country) %>%
+  map(lm_data) %>%
+  enframer("country") %>%
+  filter(term == "wave") %>%
+  arrange(-estimate) %>%
+  pull(country)
+
+complete_data_topbottom %>%
+  filter(country %in% countries) %>%
+  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
   ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
-  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
+  geom_line(stat = "smooth", method = "lm", aes(group = 1),
+            formula = y ~ splines::ns(x, 3), linetype = "longdash",
+            colour = "black") +
+  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
   geom_hline(yintercept = 0, linetype = "longdash") +
-  geom_line() +
-  geom_point(size = 0.5) +
-  coord_cartesian(ylim = c(-0.5, 3)) +
+  geom_line(alpha = 0.4) +
+  geom_point(size = 0.5, alpha = 0.4) +
+  coord_cartesian(ylim = c(0, 3)) +
   facet_wrap(~ country)
 @
+
+We start by look at the achievement for some countries.
+We plot both mathematics and reading gaps for each country and also a cubic trend spline for both test scores pooled.
+Some countries have increased their achievement very strongly.
+For example, France, Austria and surprisingly Sweden.
+In mathematics, France increase the gap by roughly \Sexpr{diff_data$math[["France"]]}.
+This increase happens similarly for literacy with an increase of \Sexpr{diff_data$read[["France"]]}.
+For such a short period of time, the magnitud of the increase is reasonably big.
+
+Given that no one has estimated the evolution of the gap we can't cross-check how other empirical estimations put France.
+However, we can make some educated inferences.
+For example, \citet{micklewright} also finds that France is a low dispersion country, specifically in the year 2000.
+They also find that Germany and US are very dispersed countries (quite the opposite).
+
+\citet{micklewright} find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
+\citet{micklewright} find
+that average achievement is negatively related to dispersion.
+
+It seems that social origin effects have actually increased in France  (Ballarino and Bernardi, 2016)
 

### 2017-06-22 paper/paper.tex
Updated countries for graphs + updated graph - first results writeup

@@
 
-This paper is interested in two questions.
-As explained in the literature review, we already know that U.S achievement gaps have been growing over time, and some international evidence suggests that it might be increasing in other countries as well.
-All of these studies pool all students together from different ages and estimate global trends.
+With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries.
+This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for many countries.
+The advantages of this study are twofold.
 
-We take a different approach and we want to study age-specific trends in achievement gaps.
-We're interested in studying a) the size of the achievement in a comparative perspective and its relationship to overall achievement levels, b) which countries are experiencing changes in the achievement gap, c) the rate at which each country-gap is widening/narrowing and d) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
+First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the Kindergarten gap in \citet{reardon_portilla}.
+As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
+Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.
+
+Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States.
+This will help asses the mangitude of the inequality as well as the rate of growth/decline.
+Thirdly, this paper not only documents the size and changes of the SES gap, but studies with particular attention the source of the changes.
+That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.
+
+More formally, We're interested in studying
+a) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
+b) Which countries are experiencing changes in the achievement gap;
+c) the rate at which each country-gap is widening/narrowing and;
+d) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
 
@@
 After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
-This lends credibility that our estimations will show variation over time.
-As usual, their estimations are only for children going from 4th to 8th grade.
-We concentrate on age-specific groups, in this case, 15 year olds.
+This shows that it is important to study the changes in the achievement gap over time.
+This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.
 
-c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year.
+c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year available.
 This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
@@
 With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests.
-The sample size for PISA 2000 is 127,388, for PISA 2003, 276,165, for PISA 2006, 398,750, for PISA 2009, 515,958, for PISA 2012, 480,810, and for PISA 2015 519,334.
+The sample size for PISA 2000 is 127,388, 276,165 for PISA 2003, 398,750 for PISA 2006, 515,958 for PISA 2009, 480,810 for PISA 2012, and 519,334 for PISA 2015.
 Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance.
-With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough time to visualize changes in the structure of an educational system.
+With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system.
+None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.
 
-In order to measure family background PISA includes several variables.
-Firstly, the ask student's their parent's educational level.
-Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}.
+To identify a family's socio-economic status PISA collects several variables that measure different dimensions.
+Classically, they ask student's their parent's educational level.
+Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}(cite more here).
 This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding.
-Moreover, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time.
-For example, until PISA 2009, the ISCED 1997 classification was currently used, but the next wave already used the new ISCED 2011 classification.
-Both these classification schemes have equivalent look-up tables, yet this requires a detailed inspection of the codings.
+In spite of this, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time.
+For example, until PISA 2009, the preferred framework was ISCED 1997, whereas the next wave switched to the newly developed ISCED 2011 classification.
+Both these classification schemes have equivalent look-up tables, but this requires a detailed inspection of the codings.
+Aside from all these limitations, this is very reliable and important variable that measures an important factor of socio-economic status.
 
@@
 
-But one of the most relevant variables for our study is the is a composite SES index created by the PISA team.
-The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI); the highest level of education of the student’s parents, converted into years of schooling; the PISA index of family wealth; the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home \citep{oecd_glance_2002}.
-This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable measures across waves.
-In fact, we use the newly-released ESCS index \citep{pisa_2015_results} which rescaled all ESCS indexes to make them suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.}.
+Yet one of the most relevant variables for our study is a composite SES index created by the PISA team.
+The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measured the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}.
+This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable metrics across waves.
+To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.
 
-Aside from SES, the other most relevant variables are test scores.
-PISA provides plausible values instead of one single indicator of test score performance.
-As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured.
-Suppose we have \(\mu_i\), the average student test score in mathematics for student \(i\).
-Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\), for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test.
+Aside from SES, the other most relevant variables are test scores for Mathematics, Literacy and Science.
+PISA does not provide a single test result for each respondent.
+Instead, it provides a \emph{series} of 'plausible values' that the child could actually score.
+As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured (the true distribution of the possible scores a student can achieve).
+Suppose we have \(\mu_i\), the average student test score in Mathematics for student \(i\).
+Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\) for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test.
 This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\).
 The reason why we use this procedure is because estimating a single estimate \(\mu_i\) is plagued with measurement error, among other types of bias \citep[see][]{wu2005}.
-The plausible values in the PIAAC data are effectively five (ten for PISA 2015) random draws from this distribution.
-These variable is are continuous, ranging from 0 to 500, with a mean of 250.
+The number of plausible values for PISA waves are usually five (although ten for PISA 2015) random draws from this distribution.
+In practice each student has 5 scores for each test, that resembles this distribution.
+They are continuous, ranging from 0 to 500, with a mean of 250.
 
@@
 
-We start by standardizing achievement for each country, year and subject.
-Because PISA does not provide a single achievement indicator, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.
+For each country in each study, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps.
 
-We then fit a linear model
+The original strategy of \citet{reardon2011} is as follows: first, achievement is standardized (see below for a mathematical explanation of the standardization).
+We then use it to calculate the mean achievement (and standard error) for each category of the SES variable of interest (parent's education, income categories, etc..).
+"Category means are plotted at their percentile ranks and cubic models are fit through the points using weighted least squares.
+Finally, achievement at each country’s 90th and 10th SES percentiles is interpolated from the model" \citep{anna2016_global}.
+The result is a SES gap from an ordinal variable of interest.
+
+As mentioned before, PISA does not provide a single achievement indicator.
+Instead, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.
+
+To standardize the test score we fit a linear model
 
@@
 
+{\centering \includegraphics[width=5.5in,height=5in]{figure/variance_pisa-1}
+
+}
+
 Inspection of the data suggests that it's something we shouldn't be deeply concerned with.
-The standard deviation of each wave seems to be following a very similar pattern with not so drastic exception of the year 2000.
+The standard deviation of each wave seems to be following a very similar pattern with a not so drastic exception of the year 2000.
 
@@
 
-\section{Analysis}
+% Figure out how to ouput data frames as latex tables.
 
-\subsection{Descriptives}
+{\centering \includegraphics[width=5.5in,height=5in]{figure/correlation_incomeineq-1}
 
-% Figure out how to ouput data frames as latex tables.
+}
 
@@
 It seems that social origin effects have actually increased in France  (Ballarino and Bernardi, 2016)
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/correlation_incomeineq-1}
-
-}
 

### 2017-06-26 paper/paper.Rnw
Made relative achievement gap - error bars missing

@@
 In mathematics, France increase the gap by roughly \Sexpr{diff_data$math[["France"]]}.
-This increase happens similarly for literacy with an increase of \Sexpr{diff_data$read[["France"]]}.
+This pattern happens similarly for literacy with an increase of \Sexpr{diff_data$read[["France"]]}.
 For such a short period of time, the magnitud of the increase is reasonably big.
 
-Given that no one has estimated the evolution of the gap we can't cross-check how other empirical estimations put France.
+Given that no one has estimated the evolution of the gap we can't cross-check how other empirical estimations put France at.
 However, we can make some educated inferences.
 For example, \citet{micklewright} also finds that France is a low dispersion country, specifically in the year 2000.
-They also find that Germany and US are very dispersed countries (quite the opposite).
-
-\citet{micklewright} find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
-\citet{micklewright} find
-that average achievement is negatively related to dispersion.
-
-It seems that social origin effects have actually increased in France  (Ballarino and Bernardi, 2016)
+
+However, we don't know what happens over time.
+But Ballarino and Bernardi (2016) did study the relationship of social origin effects and found that they increased very much in France in the last decades.
+
+Other countries have reasonable increases such as Austria and Finland.
+Aside from these countries, some other countries have no change at all.
+Specially, Canada, Netherlands and Spain.
+Canada excels here not only because the gap has been stable over time, but because it has the smallest gap of all countries presented here.
+It's nearly 0.5 standard deviations in 2000 and it increased only by \Sexpr{diff_data$math[["Canada"]]} in 2015.
+We can't really say Italy has had a significant since it has a sort of wiggly pattern that looks like it cancels itself out.
+
+Furthermore, we do have some countries which show a decrease in the SES achievement gap.
+United Kingdom, Israel, Poland, United States and Germany show some steep downward slopes.
+Most notably are the U.S and German cases.
+These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations.
+However, both countries reduced the gaps by \Sexpr{diff_data$math[["United States"]]} and \Sexpr{diff_data$math[["Germany"]]} respectively.
+Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}.
+
+Analyzing this graph we might naively conclude that these trends are not very steep and they should not be very important to consider.
+However, remember that the Y axis is measured in standard deviations.
+Small changes are actually very huge in practical terms.
+Take the case of Sweden.
+The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation.
+The actual difference is of about \Sexpr{diff_data$math[["Sweden"]]}.
+Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.
+
+Something reassuring is that mathematics and reading follow basically the same trend across all countries.
+This means that the result is not an artifact of chance.
+However, it's important to disentangle where is this gap originating from.
+Is this because the top are improving why the bottom decreases?
+Or is it that the bottom is catching up?
+Up next we plot the same graph only
+
+<<>>=
+# You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
+complete_data_topbottom %>%
+  filter(country %in% countries) %>%
+  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
+  select(wave, country, matches("*._math$")) %>%
+  gather(ses, gap_size, matches("^\\d_mean_math$")) %>%
+  separate(ses, c("ses", "delete"), sep = 2) %>%
+  mutate(ses = gsub("_", "", ses)) %>%
+  filter(complete.cases(.)) %>%
+  # gather(lower_bound, lower, matches("\\d_lower")) %>%
+  # gather(upper_bound, upper, matches("\\d_upper"))
+  ggplot(aes(as.factor(wave), gap_size, group = ses, colour = ses)) +
+  geom_point(size = 0.5, alpha = 0.4) +
+  geom_line(alpha = 0.4) +
+  geom_line(stat = "smooth", method = "lm", aes(group = 1),
+            formula = y ~ splines::ns(x, 3), linetype = "longdash",
+            colour = "black") +
+  # geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
+  # geom_hline(yintercept = 0, linetype = "longdash") +
+  # coord_cartesian(ylim = c(0, 3)) +
+  facet_wrap(~ country)
+@
+
+<<>>=
+
+lower <- seq(13, 1, -1)
+upper <- seq(100, 220, 10)
+
+df <- tibble(country = "fake_country",
+             years = rep(seq(2000, 2025, 2), 2),
+             ses = rep(c("lower", "upper"), each = 13),
+             value = c(lower, upper))
+
+# You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
+df %>%
+  ggplot(aes(as.factor(years), value, group = ses, colour = ses)) +
+  geom_point(size = 0.5, alpha = 0.4) +
+  geom_line(alpha = 0.4) +
+  geom_line(stat = "smooth", method = "lm", aes(group = 1),
+            formula = y ~ splines::ns(x, 3), linetype = "longdash",
+            colour = "black") +
+  # geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
+  # geom_hline(yintercept = 0, linetype = "longdash") +
+  # coord_cartesian(ylim = c(0, 3)) +
+  facet_wrap(~ country)
+@
 

### 2017-06-28 paper/paper.Rnw
Comented 90th/50th/10th gaps for selected countries

@@
   library(intsvy)
-  library(cimentadaj) # # devtools::install_github("cimentadaj/cimentadaj")
+  library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
   library(countrycode) # For region variable
   library(car)
-  library(readr)
   library(SAScii)
@@
 Or is it that the bottom is catching up?
-Up next we plot the same graph only
+Up next we plot the same graph but show the divergent patterns between high/low SES origins.
 
@@
   geom_line(stat = "smooth", method = "lm", aes(group = 1),
-            formula = y ~ splines::ns(x, 3), linetype = "longdash",
+            formula = y ~ splines::ns(x, 1), linetype = "longdash",
             colour = "black") +
@@
 
+The top line belongs to the high SES group whereas the bottom line belongs to the low SES group.
+The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other.
+We find plenty of variation in these patterns.
+For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up.
+The U.K seems to be following the same path as the United States as well.
+On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster.
+The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different.
+We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.
+
+Finally, we pay attention to the flat summary lines.
+These lines should be interpreted with caution because it doesn't mean that the gap is not increasing.
+Denmark, Belgium, Netherlands, Norway and Spain show a flat line because the gap is growing very little, if at all.
+On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate.
+there's a group of distinctive groups where both low/high SES rates are very great resulting in the summary line being flat.
+
+Ths source of this gap can be deconstructed into two broader gaps.
+We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group.
+Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap.
+For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th?
+Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier.
+Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous plots \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.
+
+% This plot might not be necessary any more
 <<>>=
@@
   bind_rows(complete_data_midbottom) %>%
-  filter(country %in% c("United States", "Denmark", "France")) %>%
+  filter(country %in% c("United States", "Denmark", "France", "Germany")) %>%
   mutate(type = factor(type,
@@
 
+We start off with Denmark.
+Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger for both subjects.
+Furthermore, we find, just as in the 90th/10th SES gap, that there isn't a significant decrease, with the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation.
+France shows a similar pattern as Denmark in that the 50th/10th SES gap is the main contributer and that the three gaps have the similar trend.
+
+Germany and U.S, on the other hand, show a very different landscape.
+For example, Germany, shows that the 90th/10th gap is primarily driven by the 50th/10th SES gap and that the slopes of the 90th/50th gap is much weaker than the other two gaps.
+This means that Top/Middle SES groups have shrunk much less than the other groups.
+The big catching up is actually coming from the Middle/Bottom SES groups.
+However, we don't know if this is because the bottom is catching up or that the middle is coming down.
+However, based on the gap plots from above, we can infer that it is because the bottom is catching up rapidly because against the 90th SES group, it is increasing steeply every year.
+The U.S shows a different pattern from all other countries.
+Both the 90th/50th/10th gaps seem to contribute rather similarly to the overall 90th/10th gap (although 50th/10th is marginally higher) but more interestingly, the 90th/50th gap has stagnated in the last 15 years.
+The top/middle groups have a relatively big gap and it hasn't changed at all.
+On the other hand, the steep decrease we saw for the U.S is coming exclusively from the 50th/10th gap shrinking.
+The reason, nonetheless, might be different from the German case, where it's probable that the bottom is catching up.
+Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group.
+
 <<graphing_ses_growth, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
@@
   bind_rows(complete_data_midbottom) %>%
-  filter(country %in% c("Germany", "Denmark", "France"), type_test == "math") %>%
+  filter(country %in% c("United States", "Denmark", "France", "Germany"), type_test == "math") %>%
   select(wave, country, type_test, type, contains("math")) %>%

### 2017-06-28 paper/paper.Rnw
Improved gap graph

@@
 
+<<>>=
+
+ggplot(complete_data_topbottom, aes(as.character(wave), country, fill = difference)) +
+  geom_tile() +
+  viridis::scale_fill_viridis()
+
+complete_data_topbottom %>%
+  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
+  filter(type_test == "math", !is.na(country)) %>%
+  ggplot(aes(as.character(wave), difference)) +
+  geom_point() +
+  geom_linerange(aes(ymin = 0, ymax = difference))+
+  geom_line(stat = "smooth", method = "lm", aes(group = 1),
+            formula = y ~ splines::ns(x, 2), size = 0.7,
+            colour = "red") +
+  facet_wrap(~ country, ncol = 12) +
+  # viridis::scale_fill_viridis(option = "B") +
+  scale_y_continuous(expand = c(0, 0), lim = c(0, 3)) +
+  ggthemes::theme_few() +
+  theme(panel.spacing = unit(1, "lines"),
+      axis.text.x = element_text(angle = 90, hjust = 1))
+
+@
 The top line belongs to the high SES group whereas the bottom line belongs to the low SES group.

### 2017-07-01 paper/paper.Rnw
Commented all graphs - First review of the paper will now begin

--- 2017-06-28 paper/paper.Rnw
@@
                  "France",
-                 "New Zealand",
+                 # "New Zealand",
                  "Austria",
@@
                  "Sweden",
-                 "Czech Republic",
+                 # "Czech Republic",
                  "Canada",
                  "Hungary",
-                 "Iceland", "Netherlands",
+                 # "Iceland",
+                 "Netherlands",
                  "Spain",
-                 "Belgium",
-                 "Italy", "Norway",
+                 # "Belgium",
+                 "Italy",
+                 # "Norway",
                  "United Kingdom",
-                 "Greece",
+                 # "Greece",
                  "Denmark",
-                 "Israel",
+                 # "Israel",
                  "Poland",
                  "United States",
-                 "Germany",
-                 "Turkey",
-                 "Russia")
+                 "Germany"
+                 # "Turkey",
+                 # "Russia"
+                 )
 @
@@
 
-# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
-# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
-# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
-# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
-# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
-# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
-
-results_math <- read_rds("./data/delete.Rdata")
-results_read <- read_rds("./data/delete_read.Rdata")
-results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
-results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
-results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
-results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
+results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
+results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
+results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
+results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
+results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
+results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
+
+# results_math <- read_rds("./data/delete.Rdata")
+# results_read <- read_rds("./data/delete_read.Rdata")
+# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
+# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
+# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
+# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
 # US is missing for reading
@@
   setNames(seq(2000, 2015, 3)) %>%
-  enframer(name) %>%
+  enframer("name") %>%
   ggplot(aes(name, value)) + geom_col()
@@
 
+<<>>=
+
+complete_data_topbottom %>%
+  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
+  filter(type_test == "math", !is.na(country)) %>%
+  ggplot(aes(as.character(wave), difference)) +
+  geom_point() +
+  geom_linerange(aes(ymin = 0, ymax = difference))+
+  geom_line(stat = "smooth", method = "lm", aes(group = 1),
+            formula = y ~ splines::ns(x, 2), size = 0.7,
+            colour = "red") +
+  facet_wrap(~ country, ncol = 12) +
+  viridis::scale_fill_viridis(option = "B") +
+  scale_y_continuous(expand = c(0, 0), lim = c(0, 3)) +
+  ggthemes::theme_few() +
+  theme(panel.spacing = unit(1, "lines"),
+      axis.text.x = element_text(angle = 90, hjust = 1))
+
+@
 The top line belongs to the high SES group whereas the bottom line belongs to the low SES group.
@@
 % This plot might not be necessary any more
-<<>>=
-
-lower <- seq(13, 1, -1)
-upper <- seq(100, 220, 10)
-
-df <- tibble(country = "fake_country",
-             years = rep(seq(2000, 2025, 2), 2),
-             ses = rep(c("lower", "upper"), each = 13),
-             value = c(lower, upper))
-
-# You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
-df %>%
-  ggplot(aes(as.factor(years), value, group = ses, colour = ses)) +
-  geom_point(size = 0.5, alpha = 0.4) +
-  geom_line(alpha = 0.4) +
-  geom_line(stat = "smooth", method = "lm", aes(group = 1),
-            formula = y ~ splines::ns(x, 3), linetype = "longdash",
-            colour = "black") +
-  # geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
-  # geom_hline(yintercept = 0, linetype = "longdash") +
-  # coord_cartesian(ylim = c(0, 3)) +
-  facet_wrap(~ country)
-@
+% <<>>=
+% 
+% lower <- seq(13, 1, -1)
+% upper <- seq(100, 220, 10)
+% 
+% df <- tibble(country = "fake_country",
+%              years = rep(seq(2000, 2025, 2), 2),
+%              ses = rep(c("lower", "upper"), each = 13),
+%              value = c(lower, upper))
+% 
+% # You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
+% df %>%
+%   ggplot(aes(as.factor(years), value, group = ses, colour = ses)) +
+%   geom_point(size = 0.5, alpha = 0.4) +
+%   geom_line(alpha = 0.4) +
+%   geom_line(stat = "smooth", method = "lm", aes(group = 1),
+%             formula = y ~ splines::ns(x, 3), linetype = "longdash",
+%             colour = "black") +
+%   # geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
+%   # geom_hline(yintercept = 0, linetype = "longdash") +
+%   # coord_cartesian(ylim = c(0, 3)) +
+%   facet_wrap(~ country)
+% @
 
@@
 We start off with Denmark.
-Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger for both subjects.
+Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger for both tests.
 Furthermore, we find, just as in the 90th/10th SES gap, that there isn't a significant decrease, with the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation.
@@
 Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group.
+
+But, similar as before, we don't know which group is causing the decrease/increase of the gap in the 90th/50th and 50th/10th gap. For that we plot the evolution of these gaps below.
 
@@
   geom_errorbar(aes(ymin = lower_math, ymax = upper_math), width = 0.1) +
-  geom_hline(yintercept = 0, linetype = "longdash") +
   geom_line() +
@@
 
+With Denmark there's no average change across all years.
+However, for France, we see that the gap is growing much faster for the 50th/10th SES gap.
+For Germany we see that the gap is closing in all three groups although much more strongly for the 50th/10th SES gap.
+It looks like the 50th/10th gap is mainly driving the changes in the gaps for most of the countries that actually see a change over time.
+For the U.S, there's very little change in the 90th/50th SES gap, with most of the change coming from the 50th/10th gap.
+
+Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's an efficiency/equality tradeoff and if it stands over time.
+One particular topic that has been researched in the past is whether improving the overall cognitive development of a country compromises the distribution of the countries overall performance.
+For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance.
+This, however, can come at a price.
+If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance.
+
 <<graphing_achievement_disparity, cache = T>>=
@@
   geom_smooth()
-
+@
+
+For all country-years pooled we see a very strong correlation but contrary to what we would expect.
+We find that countries with high performance have also a very low level of inequality.
+In fact, even if we separate it by year.
+
+<<>>=
+data_to_plot %>%
+  ggplot(aes(scale(Mean), difference)) +
+  geom_point() +
+  geom_smooth(method = "lm") +
+  facet_wrap(~ wave)
+@
+
+<<>>=
 data_to_plot %>%
@@
   facet_wrap(~ country)
-
+@
+
+In fact, we see a strong correspondence.
+In countries like Germany and France we see quite a functional form between both inequality and performance.
+This pattern is also seen in in the UK, Austria, Australia and New Zealand.
+Poland and Italy are quite interesting cases because it shows how inequality has been decreasing that much whereas average performance has been improving notably.
+On another hand, only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality.
+
+<<>>=
 data_to_plot %>%
@@
   ggplot(aes(reorder(country, cor), cor, fill = cor > 0)) + geom_col() + coord_flip()
-
 @
@@
             inherit.aes = FALSE) +
-  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1) +
+  geom_line(stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "grey", size = 1,
+            linetype = "longdash") +
   geom_point(alpha = 0.2) +
   geom_point(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.7) +
+  geom_line(data = filter(full_data, continent == "my_cnt"), colour = "red", alpha = 0.4,
+            stat="smooth", method = "lm", se = FALSE, alpha = 0.5, colour = "red", size = 1) +
   geom_text_repel(data = filter(full_data, continent == "my_cnt"),
@@
   annotate(geom = "text", x = 0.15, y = -0.2,
-           label = "Low SES are catching up \n faster than High SES",
+           label = "Low SES is catching up \n faster than High SES",
            fontface = 2, size = 3) +
   annotate(geom = "text", x = -0.05, y = 0.20,
-           label = "High SES are increasing  \n faster than Low SES",
+           label = "High SES is increasing  \n faster than Low SES",
            fontface = 2, size = 3) +
@@
 @
+
+In a more convenient way, we can see that there is a lot of variance in the rate at which top/bottom groups are changing.
+We see that there are more countries where the inequality is increasing (red panel) but there's also a fair share where Low SES is catching up faster, such as United States and Germany.
+We see a negative relationship where the faster the low SES catch up the slower the High SES decrease.
+Quite an interesting pattern because we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel).
+On the other hand, we do see more ocuntries where the both groups are increasing (top right panel).
+We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.
 
@@
 <<graphing_perc_increase, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
-perc_graph <- function(df) {
+perc_graph <- function(df, title) {
   df %>%
@@
   coord_cartesian(ylim = c(-160, 160)) +
+  ggtitle(title) +
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
@@
 
-m1 <- perc_graph(top_bottom_perc)
-m2 <- perc_graph(top_mid_perc)
-m3 <- perc_graph(mid_bottom_perc)
+m1 <- perc_graph(top_bottom_perc, "90/10")
+m2 <- perc_graph(top_mid_perc, "90/50")
+m3 <- perc_graph(mid_bottom_perc, "50/10")
 

### 2017-07-01 paper/paper.tex
Commented all graphs - First review of the paper will now begin

@@
 Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
-They also find that France is a lower dispersion country (but in my results this is changing a lot).
-They also find that Germany and US are very dispersed countries (quite the opposite)
-
-\citet{micklewright} find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
-\citet{micklewright} find
-that average achievement is negatively related to dispersion.
-
-It seems that social origin effects have actually increased in France  (Ballarino and Bernardi, 2016)
 
@@
 
+We start by look at the achievement for some countries.
+We plot both mathematics and reading gaps for each country and also a cubic trend spline for both test scores pooled.
+Some countries have increased their achievement very strongly.
+For example, France, Austria and surprisingly Sweden.
+In mathematics, France increase the gap by roughly 0.9.
+This pattern happens similarly for literacy with an increase of 0.6.
+For such a short period of time, the magnitud of the increase is reasonably big.
+
+Given that no one has estimated the evolution of the gap we can't cross-check how other empirical estimations put France at.
+However, we can make some educated inferences.
+For example, \citet{micklewright} also finds that France is a low dispersion country, specifically in the year 2000.
+
+However, we don't know what happens over time.
+But Ballarino and Bernardi (2016) did study the relationship of social origin effects and found that they increased very much in France in the last decades.
+
+Other countries have reasonable increases such as Austria and Finland.
+Aside from these countries, some other countries have no change at all.
+Specially, Canada, Netherlands and Spain.
+Canada excels here not only because the gap has been stable over time, but because it has the smallest gap of all countries presented here.
+It's nearly 0.5 standard deviations in 2000 and it increased only by 0.2 in 2015.
+We can't really say Italy has had a significant since it has a sort of wiggly pattern that looks like it cancels itself out.
+
+Furthermore, we do have some countries which show a decrease in the SES achievement gap.
+United Kingdom, Israel, Poland, United States and Germany show some steep downward slopes.
+Most notably are the U.S and German cases.
+These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations.
+However, both countries reduced the gaps by -0.6 and -0.9 respectively.
+Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}.
+
+Analyzing this graph we might naively conclude that these trends are not very steep and they should not be very important to consider.
+However, remember that the Y axis is measured in standard deviations.
+Small changes are actually very huge in practical terms.
+Take the case of Sweden.
+The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation.
+The actual difference is of about 0.6.
+Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.
+
+Something reassuring is that mathematics and reading follow basically the same trend across all countries.
+This means that the result is not an artifact of chance.
+However, it's important to disentangle where is this gap originating from.
+Is this because the top are improving why the bottom decreases?
+Or is it that the bottom is catching up?
+Up next we plot the same graph but show the divergent patterns between high/low SES origins.
+
+The top line belongs to the high SES group whereas the bottom line belongs to the low SES group.
+The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other.
+We find plenty of variation in these patterns.
+For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up.
+The U.K seems to be following the same path as the United States as well.
+On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster.
+The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different.
+We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.
+
+Finally, we pay attention to the flat summary lines.
+These lines should be interpreted with caution because it doesn't mean that the gap is not increasing.
+Denmark, Belgium, Netherlands, Norway and Spain show a flat line because the gap is growing very little, if at all.
+On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate.
+there's a group of distinctive groups where both low/high SES rates are very great resulting in the summary line being flat.
+
+Ths source of this gap can be deconstructed into two broader gaps.
+We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group.
+Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap.
+For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th?
+Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier.
+Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous plots \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.
+
+% This plot might not be necessary any more
+% <<>>=
+% 
+% lower <- seq(13, 1, -1)
+% upper <- seq(100, 220, 10)
+% 
+% df <- tibble(country = "fake_country",
+%              years = rep(seq(2000, 2025, 2), 2),
+%              ses = rep(c("lower", "upper"), each = 13),
+%              value = c(lower, upper))
+% 
+% # You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
+% df %>%
+%   ggplot(aes(as.factor(years), value, group = ses, colour = ses)) +
+%   geom_point(size = 0.5, alpha = 0.4) +
+%   geom_line(alpha = 0.4) +
+%   geom_line(stat = "smooth", method = "lm", aes(group = 1),
+%             formula = y ~ splines::ns(x, 3), linetype = "longdash",
+%             colour = "black") +
+%   # geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
+%   # geom_hline(yintercept = 0, linetype = "longdash") +
+%   # coord_cartesian(ylim = c(0, 3)) +
+%   facet_wrap(~ country)
+% @
+
 {\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_allgaps-1}
@@
 
+We start off with Denmark.
+Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger for both tests.
+Furthermore, we find, just as in the 90th/10th SES gap, that there isn't a significant decrease, with the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation.
+France shows a similar pattern as Denmark in that the 50th/10th SES gap is the main contributer and that the three gaps have the similar trend.
+
+Germany and U.S, on the other hand, show a very different landscape.
+For example, Germany, shows that the 90th/10th gap is primarily driven by the 50th/10th SES gap and that the slopes of the 90th/50th gap is much weaker than the other two gaps.
+This means that Top/Middle SES groups have shrunk much less than the other groups.
+The big catching up is actually coming from the Middle/Bottom SES groups.
+However, we don't know if this is because the bottom is catching up or that the middle is coming down.
+However, based on the gap plots from above, we can infer that it is because the bottom is catching up rapidly because against the 90th SES group, it is increasing steeply every year.
+The U.S shows a different pattern from all other countries.
+Both the 90th/50th/10th gaps seem to contribute rather similarly to the overall 90th/10th gap (although 50th/10th is marginally higher) but more interestingly, the 90th/50th gap has stagnated in the last 15 years.
+The top/middle groups have a relatively big gap and it hasn't changed at all.
+On the other hand, the steep decrease we saw for the U.S is coming exclusively from the 50th/10th gap shrinking.
+The reason, nonetheless, might be different from the German case, where it's probable that the bottom is catching up.
+Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group.
+
+But, similar as before, we don't know which group is causing the decrease/increase of the gap in the 90th/50th and 50th/10th gap. For that we plot the evolution of these gaps below.
+
 {\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_ses_growth-1}
@@
 }
+
+With Denmark there's no average change across all years.
+However, for France, we see that the gap is growing much faster for the 50th/10th SES gap.
+For Germany we see that the gap is closing in all three groups although much more strongly for the 50th/10th SES gap.
+It looks like the 50th/10th gap is mainly driving the changes in the gaps for most of the countries that actually see a change over time.
+For the U.S, there's very little change in the 90th/50th SES gap, with most of the change coming from the 50th/10th gap.
+
+Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's an efficiency/equality tradeoff and if it stands over time.
+One particular topic that has been researched in the past is whether improving the overall cognitive development of a country compromises the distribution of the countries overall performance.
+For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance.
+This, however, can come at a price.
+If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance.
+
+For all country-years pooled we see a very strong correlation but contrary to what we would expect.
+We find that countries with high performance have also a very low level of inequality.
+In fact, even if we separate it by year.
+
+In fact, we see a strong correspondence.
+In countries like Germany and France we see quite a functional form between both inequality and performance.
+This pattern is also seen in in the UK, Austria, Australia and New Zealand.
+Poland and Italy are quite interesting cases because it shows how inequality has been decreasing that much whereas average performance has been improving notably.
+On another hand, only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality.
 
@@
 }
+
+In a more convenient way, we can see that there is a lot of variance in the rate at which top/bottom groups are changing.
+We see that there are more countries where the inequality is increasing (red panel) but there's also a fair share where Low SES is catching up faster, such as United States and Germany.
+We see a negative relationship where the faster the low SES catch up the slower the High SES decrease.
+Quite an interesting pattern because we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel).
+On the other hand, we do see more ocuntries where the both groups are increasing (top right panel).
+We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.
 

### 2017-07-01 paper/paper.Rnw
Rewrote everything except analysis + spell check

@@
 Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years.
-Are we destined to belong to the same social status as our parents?
-Is meritocracy over, if it ever existed?
-These questions have motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century.
-When James Colem(1966) released his famous Coleman report it raised the topic to starlight by suggesting that family SES and student performance are tightly linked.
-Fast-forward today, we've understood the relationship much better, although far from being clear.
-The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school (cite heckman).
-Moreover, these inequalities can be narrowed significantly if the children coming from unfavourable conditions are exposed to high quality learning environments.
-One particular finding coming from (cite heckman skill begets skill) is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\).
-Although this sounds logical, it is often not acknowledged properly.
-
-The implications of this cummulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\).
-That is why the whole debate about narrowing inequality between polarized families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns.
-Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status, is present in virtually all empirical studies of social mobility and inequalities (cite breen and goldthorpe, cite waldfogel what children need, cite bradbury 2015, cite chetty for recent evidence).
+The idea of meritocracy and intergenerational transfers have motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century.
+When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family SES and student performance are tightly linked.
+The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements.
+For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.
+But fast-forward until today, we've understood the relationship much better.
+
+The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
+This means that inequality starts strong before the schooling of the children being and schools can do so much to close it.
+Moreover, these inequalities can be narrowed significantly if the child coming from unfavourable conditions is exposed to high quality learning environments.
+But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\).
+Although this sounds logical, it's implications are not assesed properly.
+
+The implications of this cumulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\).
+That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns.
+Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.
 What we don't know is how and why this relationship is formed from such an early age.
-Moreover, many have suggested different explanations and the topic has been very contested (duncan and brooks-gun 1997).
-
-Social mobility has deepend our knowledge on the odds and relationship between intergenerational transmission of SES.
-The field of economics of education has also helped to understand that education is by far the investment that yields the best outcomes not only for the child and his/her family, but to the economy as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth (cite hanushk and woesmann study).
+Moreover, many have suggested different explanations and the topic has been very contested \citep{brooks1997}.
+
+It is incomplete to tusy educational inequalities without considering the previlance of intergenerational transmission of human and social capital.
+Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity.
+The field of economics of education has also helped to understand that education is by far the investment that yields the best returns.
+This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
 We've become so intertwined with the subject that we've virtually spent most of our time studying the mechanisms through which this inequality comes about.
-Naturally, we want to do that to be able to reverse it and help disadvantaged children reach their fullest potential.
+Naturally, we want to do that in order to reverse it and help every child reach its fullest potential.
 Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities.
-In addition, since most research has concentrated on the American case, we haven't been able to learn which countries have the biggest or smallests gaps, and whether they've been narrowing or closing over time.
+We want to know if the gap is big or small relative to other countries.
+Furthermore we want to know if this gap has reponded to our policy efforts to reduce it.
 
@@
 
-Taking the Danish example, we learned a great deal about how to improve social mobility by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility for less advantaged children (Bjorklund and Jantti, 2009; Jaeger and Holm, 2007).
-But the important finding came when we discovered the main reason behind their social elevator: its' educational system.
-For example, research by Esping and Waldfogel and Gosta's phd student on Denmark, show that the Danish early education system is particularly efficient at giving free high quality education, regardless of socio-economic status.
+For Denmark, for example, we learned a great deal about how to improve social mobility by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility for less advantaged children \citep{bjorklund2009, jaeger2007}.
+But the important finding came when we discovered the main reason behind their social escalator: its' educational system.
+For example, research by \citet{esping2012} and \citet{bauchmuller2015}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential, specially disadvantaged children.
 Moreover, it tracks students into different curricula very late compared to other Europen countries Europe (age 16).
-The importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
-
-Recent research on educational inequality has found that achievement gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years.
-Literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape.
+These two traits make the Danish system uniquely effective.
+The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable.
+In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
+
+The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years.
+The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape.
 The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940.
 This trend has allowed researchers to study achievement gaps in a lengthy period of time.
-In fact, using this information, \citep{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
-Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap -waldfogel study).
-Surprisingly, the gaps have reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap.
-The widening of the achievement gap has been happening in parallel to the growth of income inequality, although very suggestive, it is hard to link both things causally.
-However, a recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979.
-If income inequality would be a strong a direct function of the achievement gap, we should expect for the white-black achievement gap to be widening rather than narrowing, as it is now (Magnuson and Waldfogel 2008) \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
-
-\citet{reardon2011} finds that the increase has occured predominantely from the 1970's until the 2000's.
-In fact, the hard numbers suggest that the gap widened by ~ 40-50\%.
-The author also estimates the rate of change using data from the 1940's and finds an even higher increase of ~ 75\%.
+Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
+Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citet{magnuson2008}.
+These gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
+
+\citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's.
+In fact, the hard numbers suggest that the gap widened by \~40-50\%.
+The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of \~75\%.
 Given that the studies before 1970 are less reliablle in terms of comparability and sampling design, the author computes all results for both before/after 1970.
@@
 Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25.
-Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of ~ 1.25/1.50 standard deviations.
-\footnote{One important drawback of these studies is that they don't present the uncertainty of this estimates.
+Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of \~1.25-1.50 standard deviations.
+One important drawback of these studies is that they don't present the uncertainty of this estimates.
 Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
-It could be that the gap is at ~ 1.25 standard deviations, varying up to 1.75 and down to 0.80.
-Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.}
+It could be that the gap is at \~1.25 standard deviations, varying up to 1.75 and down to 0.80.
+Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.
+
+Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation.
+However likely, this is a highly contested topic.
+A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality.
+However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011, magnuson2008}
+\footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
 
 In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend.
-This follow-up study concentrated solely on the Kindergarten gap in which they took a snapshot of the achievement gap in Kindergarten in 1998, 2006 and 2010 and found that the 90th/10th income gap in readiness closed modestly.
-Furthermore, using data from fall and spring in the same kindergarden year, it seems like the gap in narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010.
-In contrast to \citep{reardon2011}, we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates.
-Their results not only hold for the income achievement gap, but they also recorded a particular decline in the white-hispanic gap.
-
-The American situation is a very distinctive one relative to Europe.
+This follow-up study concentrated solely on the U.S Kindergarten gap in which they took a snapshot of the achievement gap in Kindergarten in 1998, 2006 and 2010.
+They found that the 90th/10th income gap in readiness closed modestly.
+Furthermore, using data from fall and spring in the same kindergarden year, they calculated that the gap narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010.
+They also calculated the same changes for a number of personality traits such as self-control and externalizing behaviour and found similar results.
+In contrast to \citet{reardon2011}, he finds that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates.
+Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap).
+
+The American social landscape is very distinctive relative to Europe.
 The reasons the authors find a reversal in the trend could be numerous and should be studied very close.
-For example, the American educational system lacks any formal curricular differentiation \emph{à la} European style.
-It is true nonetheless that \citet{reardon2011} did not cover from the period 2000 onwards, which is specifically the trend they studied.
-In contrast, \citet{reardon2011} suggests that the reversal is likely due to the high increase of preschool enrollment.
+For example, the American educational system lacks any formal curricular differentiation \emph{\à la} European style.
+It is true nonetheless that \citet{reardon2011} did not cover from the period 2000 onwards, which is specifically the trend they studied so we have no other comparison.
+\citet{reardonportilla} suggests that the reversal is likely due to the high increase of preschool enrollment.
 They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
@@
 
-Motivated by these recent results, some authors have taken this analysis to an international context in order to discover between-country trends.
-The unique work of \citet{bradbury2015} employs a unique comparative analysis of Australia, United Kingdom, United States and Canada.
+Motivated by these recent results, other authors have taken this analysis to an international context in order to discover between-country trends.
+The work of \citet{bradbury2015} employs a unique comparative analysis of the achievement gap between Australia, United Kingdom, United States and Canada.
 Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is not only the first study that uses panel data to study achievement gaps, but to also do it between countries}.
-The core finding behind the book is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada.
+The core finding behind their study is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada.
 They find that once the achievement gap is present in early school entry, it doesn't seem to narrow or neither widen very much over the life course.
-In fact, they calculate that the schooling environment once they enter school can only explain about ~ 30\%-40\% of the high school SES gap.
-This suggests that once the achievement gap widens before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous RCT's show that high quality schooling can indeed help ease the gap, in some instances even close it.}.
-Once exception is the United Kingdom, which they find closed the gap in early primary years.
-This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health.
+In fact, they calculate that the schooling environment once they enter school can only explain about \~30-40\% of the high school SES gap.
+This suggests that once the achievement gap widens before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous RCT's show that high quality schooling can indeed help ease the gap, in some instances even close it \citep{campbell2002}.}.
+One exception is the United Kingdom, which they found to help close the gap in early primary years.
+This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health and welfare support.
 
@@
 Another limitation of their study is that their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
-Their findings are very reliable but should be taken as suggestive at best.
-For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citep{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries using much more comparable studies.
-These studies tackle a completely different question from the above, namely to use study cross-sectional differences between many countries, instead of over-time analysis of student gaps.
-However, they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
+They do an amazing work at makgin the 4 surveys comparable and their findings are very reliable but should be taken as suggestive at best.
+For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citet{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in the gap using much more comparable surveys.
+However it should be noted that these studies tackle a completely different question from the above, namely to study cross-sectional differences between many countries, instead of over-time analysis of student gaps.
+Nonetheless, they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
 
@@
 Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
-In comparison to the literature on achievement gaps, they find that the U.S has a gap of ~ 1.20 SD in 2001 which increase to around ~ 1.30 in the year 2006 while Germany has a decreasing gap from around ~ 1.25 to ~ 1 standard deviation in the same year-span.
+In comparison to the literature on achievement gaps, they find that the U.S has a gap of \~ 1.20 SD in 2001 which increase to around ~ 1.30 in the year 2006 while Germany has a decreasing gap from around \~ 1.25 to \~ 1 standard deviation in the same year-span.
 However, these numbers vary a lot and carry a great deal of overlapping uncertainty.
 
-They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, among other things.
-The correlations are very suggestive  as explanatory mechanisms but they are very cautios in drawing causality, specially because they are all reasonably correlated.
-However, one interesting question is whether these country gaps have evolved over time.
-With their data, they only have 3 countries which are present in all waves but they also have very few waves because their question of interest (income categories) was only asked in three waves.
+They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation and central exams, among other things.
+The correlations are very suggestive  as explanatory mechanisms but they are very cautios in drawing causality, specially because they are all reasonably correlated among each other.
+
+One interesting question is still missing is whether these country gaps have evolved over time.
+With their data, they only have 3 countries which are present in all waves but they also have very few waves because their question of interest (income categories) was only asked in three time points.
 That is why their study is more about between country gaps rather than the magnitude and evolution of the gaps.
@@
 Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}.
-Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as elimination tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
-
-Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
-Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
-Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
-Talk more about how countries with high social mobility has been linked to smallest achievement gaps
-Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.
+Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as eliminating tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
+
+% Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
+% Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
+% Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
+% Talk more about how countries with high social mobility has been linked to smallest achievement gaps
+% Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.
 
@@
 
-With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries.
-This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for many countries.
+This study seeks to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries.
+This is different from previous work because it concentrates solely on 15 year old children, something not done for the achievement gap, and it attempts to capture the evolution of the achievement gap for countries with different educational systems.
+Furthermore, we study the relationship between the gap and the average performance of each country in order to detech if there's a tradeoff between efficiency and equality.
 The advantages of this study are twofold.
@@
 This will help asses the mangitude of the inequality as well as the rate of growth/decline.
-Thirdly, this paper not only documents the size and changes of the SES gap, but studies with particular attention the source of the changes.
+This study not only documents the size and changes of the SES gap, but pays particular attention to the source of the changes.
 That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.
@@
 a) We want to test the notion that better performing countries have a smaller degree of inequality than other countries.
-\citep{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
+\citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
 This pattern is not so obvious, however.
@@
 This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
-Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders.}
+Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}
 
@@
 PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students.
-Currrently, PISA has six waves starting which started in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 countries, both developed and developing ones.
+Currrently, PISA has six waves starting in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 developed/developing countries.
 
@@
 Classically, they ask student's their parent's educational level.
-Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}(cite more here).
-This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding.
-In spite of this, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time.
+Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}.
+This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have small differences.
+In spite of this, another limitation is the fact that parent's education is measured using the ISCED classification, something that has changed over time.
 For example, until PISA 2009, the preferred framework was ISCED 1997, whereas the next wave switched to the newly developed ISCED 2011 classification.
 Both these classification schemes have equivalent look-up tables, but this requires a detailed inspection of the codings.
-Aside from all these limitations, this is very reliable and important variable that measures an important factor of socio-economic status.
-
-Another social background variable PISA collects is the International Socio-Economic Index of Occupational Status (ISEI).
-This variable captures the economic status of the family, the closest it can without asking for income information.
-This index variable was developed by Ganzeboom (cite) and it attempts to measure occupational status using a continuous measure.
-This indicator has been found to ourperform classical occupational classifications such as the Erikson-Goldthorpe-Portocarero schema (cite).
-It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature \citep{anna2016_global}.
+
+Another social background variable is the International Socio-Economic Index of Occupational Status (ISEI).
+This variable attempts to capture the social status of the family, without asking for income information.
+This index variable was developed by \citet{ganzeboom1996, ganzeboom2010} and it attempts to measure occupational status using a continuous measure.
+This indicator has been found to ourperform classical occupational classifications such as the Erikson-Goldthorpe-Portocarero classification \citet{erikson1979}.
+It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}.
 PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.
@@
 The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measured the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}.
-This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable metrics across waves.
+This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming all mentioned variables into comparable metrics across waves.
 To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.
 
-Aside from SES, the other most relevant variables are test scores for Mathematics, Literacy and Science.
+Aside from SES, the other most relevant variables are test scores for Mathematics and Literacy.
 PISA does not provide a single test result for each respondent.
@@
 The number of plausible values for PISA waves are usually five (although ten for PISA 2015) random draws from this distribution.
-In practice each student has 5 scores for each test, that resembles this distribution.
-They are continuous, ranging from 0 to 500, with a mean of 250.
+In practice each student has 5 scores for each test, that resemble their distribution.
+Those values are continuous, ranging from 0 to 500, with a mean of 250.
 
@@
 
-\citet{reardon_portilla, anna2016, anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt in this paper.
+\citet{reardon_portilla}, \citet{anna2016} and \citet{anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt in this paper.
 SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.
-The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th \citep{micklewright} while the literature has been following the standard set by \citet{reardon2011} to use 90th/10th percentiles.
-
-For each country in each study, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps.
-
+The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, as others have used, for example, the 5th, 50th and 95th \citep{micklewright}.
+We use the conventional cutoffs as the literature has been following the standard set by \citet{reardon2011} to use 90th/10th percentiles.
+
+For each country in each wave, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps.
 The original strategy of \citet{reardon2011} is as follows: first, achievement is standardized (see below for a mathematical explanation of the standardization).
@@
 \begin{equation}
-Y_i = \alpha + \beta_1 * AGE_i + \epsilon_i, \epsilon_i ~ N(0, \sigma^2)
+Y_i = \alpha + \beta_1 * AGE_i + \epsilon_i, \quad \epsilon_i ~ N(0, \sigma^2)
 \end{equation}
@@
 This new standardized variable has a mean of zero.
-Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves.
-However, if the variance of academic achievement changes over time, then standardizing the overall score at each country-wave pair actually makes the variable more biased.
+Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves because the test scores have now the same metric caross time.
+However, if the variance of academic achievement changes over time, then standardizing the overall score at each country-wave pair actually makes the transformation biased.
 That is, by standardizing we're forcing the standard deviation of test scores to be zero across all waves.
@@
 
-Inspection of the data suggests that it's something we shouldn't be deeply concerned with.
+A plotting ot the standard deviation of the data suggests that it's something we shouldn't be deeply concerned with.
 The standard deviation of each wave seems to be following a very similar pattern with a not so drastic exception of the year 2000.
@@
 
-In order to correct gap estimates for measurement error, I adjust student test scores by
+In order to correct gap estimates for measurement error, I adjust student test scores(\begin{math} \hat{Y_i} \end{math}) by
 
@@
 
-where \begin{math}r\end{math} is the reliability score of the wave.
-\footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}.
+where \begin{math}r\end{math} is the reliability score of the wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}.
 Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own.

### 2017-07-03 paper/paper.Rnw
Playing around with latex tables

@@
 options(scipen = 213141)
-opts_chunk$set(echo = F,
-               message = F,
-               warning = F,
-               include = F,
-               cache.lazy = F,
+
+opts_chunk$set(echo = FALSE,
+               message = FALSE,
+               cache = TRUE,
+               warning = FALSE,
+               include = FALSE,
+               cache.lazy = FALSE,
                results = 'asis')
@@
   library(ggrepel)
+  library(xtable)
 
@@
 pisa_all <- read_rds("./data/pisa_listcol.Rdata")
+
 pisa_all2 <- pisa_all
@@
 
-<<escs_trend, cache = TRUE>>=
+<<escs_trend>>=
   # Rescaled trend ESCS data to merge.
@@
 
-<<merge_escs_pisa, cache = TRUE>>=
+<<merge_escs_pisa>>=
    # Next we'll merge the ESCS data with the PISA data. As explained above, the 6th data (PISA
@@
 
-<<modeling, cache = TRUE>>=
+<<modeling>>=
 adapted_year_data <-
@@
 But the important finding came when we discovered the main reason behind their social escalator: its' educational system.
-For example, research by \citet{esping2012} and \citet{bauchmuller2015}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential, specially disadvantaged children.
+For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential, specially disadvantaged children.
 Moreover, it tracks students into different curricula very late compared to other Europen countries Europe (age 16).
@@
 \citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's.
-In fact, the hard numbers suggest that the gap widened by \~40-50\%.
-The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of \~75\%.
+In fact, the hard numbers suggest that the gap widened by \~  40-50\%.
+The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of \~  75\%.
 Given that the studies before 1970 are less reliablle in terms of comparability and sampling design, the author computes all results for both before/after 1970.
@@
 Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25.
-Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of \~1.25-1.50 standard deviations.
+Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of \~ 1.25-1.50 standard deviations.
 One important drawback of these studies is that they don't present the uncertainty of this estimates.
 Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
-It could be that the gap is at \~1.25 standard deviations, varying up to 1.75 and down to 0.80.
+It could be that the gap is at \~ 1.25 standard deviations, varying up to 1.75 and down to 0.80.
 Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.
@@
 The reasons the authors find a reversal in the trend could be numerous and should be studied very close.
-For example, the American educational system lacks any formal curricular differentiation \emph{\à la} European style.
+For example, the American educational system lacks any formal curricular differentiation \emph{\'a la} European style.
 It is true nonetheless that \citet{reardon2011} did not cover from the period 2000 onwards, which is specifically the trend they studied so we have no other comparison.
-\citet{reardonportilla} suggests that the reversal is likely due to the high increase of preschool enrollment.
+\citet{reardon_portilla} suggests that the reversal is likely due to the high increase of preschool enrollment.
 They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
@@
 They find that once the achievement gap is present in early school entry, it doesn't seem to narrow or neither widen very much over the life course.
-In fact, they calculate that the schooling environment once they enter school can only explain about \~30-40\% of the high school SES gap.
+In fact, they calculate that the schooling environment once they enter school can only explain about \~ 30-40\% of the high school SES gap.
 This suggests that once the achievement gap widens before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous RCT's show that high quality schooling can indeed help ease the gap, in some instances even close it \citep{campbell2002}.}.
@@
 More formally, We're interested in studying
-a) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
-b) Which countries are experiencing changes in the achievement gap;
-c) the rate at which each country-gap is widening/narrowing and;
-d) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
+a) Which countries are experiencing changes in the achievement gap;
+b) the rate at which each country-gap is widening/narrowing and;
+c) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
+d) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
 
@@
 
-a) We want to test the notion that better performing countries have a smaller degree of inequality than other countries.
+a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
+After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
+This shows that it is important to study the changes in the achievement gap over time.
+This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.
+
+b) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year available.
+This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
+Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}
+
+c) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
+The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind.
+We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.
+This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving.
+If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.
+
+d) We want to test the notion that better performing countries have a smaller degree of inequality than other countries.
 \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
@@
 But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.
-
-b) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
-After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
-This shows that it is important to study the changes in the achievement gap over time.
-This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.
-
-c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year available.
-This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
-Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}
-
-d) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
-The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind.
-We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.
-This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving.
-If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.
 
@@
 This new standardized variable has a mean of zero.
-Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves because the test scores have now the same metric caross time.
+Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves because the test scores have now the same metric across time.
 However, if the variance of academic achievement changes over time, then standardizing the overall score at each country-wave pair actually makes the transformation biased.
@@
 % I'm showing standard deviations, is it better to show variance?
-<<variance_pisa, cache = TRUE, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+
+\begin{figure}
+\begin{center}
+<<variance_pisa, include = T, out.height = '3in', out.width = '3.5in', fig.align = 'center'>>=
 
@@
 @
+\caption{Standard deviation of test scores across all waves.}
+\end{center}
+\end{figure}
 
@@
 This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citep{reardon2011}.
-PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but we still have to adjust for this imprecision.
+PISA has tried to make sure the tests are comparable across waves \citep{pisa2012_technical}, but we still have to adjust for this imprecision.
 
@@
 
-Finally, we calculate the SES achievement gap for each country by extracting the fitted \(\alpha\) and \(\beta\) for each country \(j\) and calculating the difference of the predicted score of High and Low SES.
+Finally, we calculate the SES achievement gap for each country by extracting the fitted \(\alpha\) and \(\beta\) for each country \(j\) and calculating the difference of the predicted score of high and low SES.
 
@@
 
-<<escsdummy_charachteristics, cache = TRUE, eval = F>>=
-.x <- adapted_year_data[[1]]
-country <- filter(.x, country == "United States")
-probs <- c(0.1, 0.9)
-
+<<escsdummy_charachteristics>>=
 escs_dummy_creator <- function(df, probs) {
@@
 
-escs_data[[1]] %>%
-  select(country, escs_dummy, high_edu_broad, ISEI) %>%
-  filter(country %in% countries) %>%
+escs_data[[1]]$high_edu_broad <- dplyr::recode(escs_data[[1]]$high_edu_broad, `6` = 7)
+
+edu_calc <- function(df) {
+  df %>%
+  select(country, escs_dummy, high_edu_broad) %>%
   count(country, escs_dummy, high_edu_broad) %>%
@@
          perc = n / total_n * 100) %>%
+  filter(country %in% countries, high_edu_broad %in% c(1, 7)) %>%
   select(country, escs_dummy, high_edu_broad, perc) %>%
@@
   spread(high_edu_broad, perc) 
-
-escs_data[[1]] %>%
-  select(country, escs_dummy, high_edu_broad, ISEI) %>%
-  filter(country %in% countries) %>%
-  count(country, escs_dummy, isei_group = cut_number(ISEI, 4)) %>%
-  group_by(country, isei_group) %>%
-  mutate(total_n = sum(n),
-         perc = n / total_n * 100) %>%
-  select(country, escs_dummy, isei_group, perc) %>%
-  map_if(is_double, round, 2) %>%
-  as_tibble() %>%
-  filter(!is.na(escs_dummy)) %>%
-  spread(isei_group, perc) %>%
-  left_join(m, by = c("country", "escs_dummy"))
-
+}
+
+summary_data <-
+  map(escs_data, edu_calc) %>%
+  setNames(years) %>%
+  enframer()
 @
@@
 % Figure out how to ouput data frames as latex tables.
-<<sample_size, cache = TRUE>>=
+<<sample_size, results = 'tex'>>=
 # Get sample counts for each dummy
@@
 sample_tables_midbottom <- sample_size_calc(adapted_year_data, c(.1, .5), selected = TRUE, countries)
-@
-
-<<merge_math_read, cache = TRUE>>=
+
+summary_table <-
+  left_join(summary_data, sample_tables_topbottom) %>%
+  filter(name %in% c("2000", "2015")) %>%
+  arrange(country, name)
+
+table_coming <-
+  summary_table %>%
+  select(-perc) %>%
+  gather(cats, vals, -(name:escs_dummy)) %>%
+  unite(all_vals, escs_dummy, cats, sep = "_") %>%
+  spread(all_vals, vals) %>%
+  map_at(c(3, 4, 6, 7), ~ {
+    .x[is.na(.x)] <- 0
+    paste0(round(.x, 0), "%")
+  }) %>%
+  map_if(is_double, as.character) %>%
+  data.frame() %>%
+  xtable::xtable()
+
+addtorow <- list()
+addtorow$pos <- list(0)
+addtorow$command <- paste("\\hline \\
+                          & & & Low SES & & & High SES \\\\
+                          \\cmidrule(l){3-5}
+                          \\cmidrule(l){6-8}",
+                          paste0(
+                            c("Year",
+                              "Countries",
+                              "\\% Low Edu",
+                              "\\% High Edu",
+                              "Sample size",
+                              "\\% Low Edu",
+                              "\\% High Edu",
+                              "Sample size"),
+                            collapse = " & "),
+                          "\\\\ \\hline")
+
+print(table_coming,
+      add.to.row = addtorow,
+      include.colnames = FALSE,
+      include.rownames = FALSE,
+      hline.after = NULL,
+      caption.placement = "top",
+      caption = "SES sample size and ISCED composition")
+@
+
+<<merge_math_read>>=
 
@@
 % How to output data frames are latex tables?
-<<rate_change, cache = TRUE>>=
+<<rate_change>>=
 
@@
   annotate(geom = "text", x = -0.05, y = 0.20,
-           label = "High SES is increasing  \n faster than Low SES",
+           label = "High SES is increasing  \n faster than Low SES is catching up",
            fontface = 2, size = 3) +
@@
 In a more convenient way, we can see that there is a lot of variance in the rate at which top/bottom groups are changing.
-We see that there are more countries where the inequality is increasing (red panel) but there's also a fair share where Low SES is catching up faster, such as United States and Germany.
+We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as United States and Germany.
 We see a negative relationship where the faster the low SES catch up the slower the High SES decrease.
@@
 % How to output data frames are latex tables?
-<<perc_increase_tables, cache = TRUE>>=
+<<perc_increase_tables>>=
 # Show the rates at which is increasing/decreasing

### 2017-07-03 paper/paper.tex
Playing around with latex tables

@@
 Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years.
-Are we destined to belong to the same social status as our parents?
-Is meritocracy over, if it ever existed?
-These questions have motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century.
-When James Colem(1966) released his famous Coleman report it raised the topic to starlight by suggesting that family SES and student performance are tightly linked.
-Fast-forward today, we've understood the relationship much better, although far from being clear.
-The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school (cite heckman).
-Moreover, these inequalities can be narrowed significantly if the children coming from unfavourable conditions are exposed to high quality learning environments.
-One particular finding coming from (cite heckman skill begets skill) is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\).
-Although this sounds logical, it is often not acknowledged properly.
+The idea of meritocracy and intergenerational transfers have motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century.
+When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family SES and student performance are tightly linked.
+The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements.
+For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.
+But fast-forward until today, we've understood the relationship much better.
 
-The implications of this cummulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\).
-That is why the whole debate about narrowing inequality between polarized families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns.
-Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status, is present in virtually all empirical studies of social mobility and inequalities (cite breen and goldthorpe, cite waldfogel what children need, cite bradbury 2015, cite chetty for recent evidence).
+The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
+This means that inequality starts strong before the schooling of the children being and schools can do so much to close it.
+Moreover, these inequalities can be narrowed significantly if the child coming from unfavourable conditions is exposed to high quality learning environments.
+But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\).
+Although this sounds logical, it's implications are not assesed properly.
+
+The implications of this cumulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\).
+That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns.
+Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.
 What we don't know is how and why this relationship is formed from such an early age.
-Moreover, many have suggested different explanations and the topic has been very contested (duncan and brooks-gun 1997).
+Moreover, many have suggested different explanations and the topic has been very contested \citep{brooks1997}.
 
-Social mobility has deepend our knowledge on the odds and relationship between intergenerational transmission of SES.
-The field of economics of education has also helped to understand that education is by far the investment that yields the best outcomes not only for the child and his/her family, but to the economy as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth (cite hanushk and woesmann study).
+It is incomplete to tusy educational inequalities without considering the previlance of intergenerational transmission of human and social capital.
+Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity.
+The field of economics of education has also helped to understand that education is by far the investment that yields the best returns.
+This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
 We've become so intertwined with the subject that we've virtually spent most of our time studying the mechanisms through which this inequality comes about.
-Naturally, we want to do that to be able to reverse it and help disadvantaged children reach their fullest potential.
+Naturally, we want to do that in order to reverse it and help every child reach its fullest potential.
 Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities.
-In addition, since most research has concentrated on the American case, we haven't been able to learn which countries have the biggest or smallests gaps, and whether they've been narrowing or closing over time.
+We want to know if the gap is big or small relative to other countries.
+Furthermore we want to know if this gap has reponded to our policy efforts to reduce it.
 
@@
 
-Taking the Danish example, we learned a great deal about how to improve social mobility by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility for less advantaged children (Bjorklund and Jantti, 2009; Jaeger and Holm, 2007).
-But the important finding came when we discovered the main reason behind their social elevator: its' educational system.
-For example, research by Esping and Waldfogel and Gosta's phd student on Denmark, show that the Danish early education system is particularly efficient at giving free high quality education, regardless of socio-economic status.
+For Denmark, for example, we learned a great deal about how to improve social mobility by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility for less advantaged children \citep{bjorklund2009, jaeger2007}.
+But the important finding came when we discovered the main reason behind their social escalator: its' educational system.
+For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential, specially disadvantaged children.
 Moreover, it tracks students into different curricula very late compared to other Europen countries Europe (age 16).
-The importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
+These two traits make the Danish system uniquely effective.
+The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable.
+In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
 
-Recent research on educational inequality has found that achievement gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years.
-Literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape.
+The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years.
+The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape.
 The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940.
 This trend has allowed researchers to study achievement gaps in a lengthy period of time.
-In fact, using this information, \citep{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
-Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap -waldfogel study).
-Surprisingly, the gaps have reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap.
-The widening of the achievement gap has been happening in parallel to the growth of income inequality, although very suggestive, it is hard to link both things causally.
-However, a recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979.
-If income inequality would be a strong a direct function of the achievement gap, we should expect for the white-black achievement gap to be widening rather than narrowing, as it is now (Magnuson and Waldfogel 2008) \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
+Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
+Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citet{magnuson2008}.
+These gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
 
-\citet{reardon2011} finds that the increase has occured predominantely from the 1970's until the 2000's.
-In fact, the hard numbers suggest that the gap widened by ~ 40-50\%.
-The author also estimates the rate of change using data from the 1940's and finds an even higher increase of ~ 75\%.
+\citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's.
+In fact, the hard numbers suggest that the gap widened by \~  40-50\%.
+The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of \~  75\%.
 Given that the studies before 1970 are less reliablle in terms of comparability and sampling design, the author computes all results for both before/after 1970.
@@
 Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25.
-Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of ~ 1.25/1.50 standard deviations.
-\footnote{One important drawback of these studies is that they don't present the uncertainty of this estimates.
+Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of \~ 1.25-1.50 standard deviations.
+One important drawback of these studies is that they don't present the uncertainty of this estimates.
 Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
-It could be that the gap is at ~ 1.25 standard deviations, varying up to 1.75 and down to 0.80.
-Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.}
+It could be that the gap is at \~ 1.25 standard deviations, varying up to 1.75 and down to 0.80.
+Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.
+
+Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation.
+However likely, this is a highly contested topic.
+A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality.
+However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011, magnuson2008}
+\footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
 
 In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend.
-This follow-up study concentrated solely on the Kindergarten gap in which they took a snapshot of the achievement gap in Kindergarten in 1998, 2006 and 2010 and found that the 90th/10th income gap in readiness closed modestly.
-Furthermore, using data from fall and spring in the same kindergarden year, it seems like the gap in narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010.
-In contrast to \citep{reardon2011}, we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates.
-Their results not only hold for the income achievement gap, but they also recorded a particular decline in the white-hispanic gap.
+This follow-up study concentrated solely on the U.S Kindergarten gap in which they took a snapshot of the achievement gap in Kindergarten in 1998, 2006 and 2010.
+They found that the 90th/10th income gap in readiness closed modestly.
+Furthermore, using data from fall and spring in the same kindergarden year, they calculated that the gap narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010.
+They also calculated the same changes for a number of personality traits such as self-control and externalizing behaviour and found similar results.
+In contrast to \citet{reardon2011}, he finds that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates.
+Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap).
 
-The American situation is a very distinctive one relative to Europe.
+The American social landscape is very distinctive relative to Europe.
 The reasons the authors find a reversal in the trend could be numerous and should be studied very close.
-For example, the American educational system lacks any formal curricular differentiation \emph{à la} European style.
-It is true nonetheless that \citet{reardon2011} did not cover from the period 2000 onwards, which is specifically the trend they studied.
-In contrast, \citet{reardon2011} suggests that the reversal is likely due to the high increase of preschool enrollment.
+For example, the American educational system lacks any formal curricular differentiation \emph{\'a la} European style.
+It is true nonetheless that \citet{reardon2011} did not cover from the period 2000 onwards, which is specifically the trend they studied so we have no other comparison.
+\citet{reardon_portilla} suggests that the reversal is likely due to the high increase of preschool enrollment.
 They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
@@
 
-Motivated by these recent results, some authors have taken this analysis to an international context in order to discover between-country trends.
-The unique work of \citet{bradbury2015} employs a unique comparative analysis of Australia, United Kingdom, United States and Canada.
+Motivated by these recent results, other authors have taken this analysis to an international context in order to discover between-country trends.
+The work of \citet{bradbury2015} employs a unique comparative analysis of the achievement gap between Australia, United Kingdom, United States and Canada.
 Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is not only the first study that uses panel data to study achievement gaps, but to also do it between countries}.
-The core finding behind the book is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada.
+The core finding behind their study is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada.
 They find that once the achievement gap is present in early school entry, it doesn't seem to narrow or neither widen very much over the life course.
-In fact, they calculate that the schooling environment once they enter school can only explain about ~ 30\%-40\% of the high school SES gap.
-This suggests that once the achievement gap widens before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous RCT's show that high quality schooling can indeed help ease the gap, in some instances even close it.}.
-Once exception is the United Kingdom, which they find closed the gap in early primary years.
-This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health.
+In fact, they calculate that the schooling environment once they enter school can only explain about \~ 30-40\% of the high school SES gap.
+This suggests that once the achievement gap widens before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous RCT's show that high quality schooling can indeed help ease the gap, in some instances even close it \citep{campbell2002}.}.
+One exception is the United Kingdom, which they found to help close the gap in early primary years.
+This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health and welfare support.
 
@@
 Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
-In comparison to the literature on achievement gaps, they find that the U.S has a gap of ~ 1.20 SD in 2001 which increase to around ~ 1.30 in the year 2006 while Germany has a decreasing gap from around ~ 1.25 to ~ 1 standard deviation in the same year-span.
+In comparison to the literature on achievement gaps, they find that the U.S has a gap of \~  1.20 SD in 2001 which increase to around ~ 1.30 in the year 2006 while Germany has a decreasing gap from around \~  1.25 to \~  1 standard deviation in the same year-span.
 However, these numbers vary a lot and carry a great deal of overlapping uncertainty.
 
-They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, among other things.
-The correlations are very suggestive  as explanatory mechanisms but they are very cautios in drawing causality, specially because they are all reasonably correlated.
-However, one interesting question is whether these country gaps have evolved over time.
-With their data, they only have 3 countries which are present in all waves but they also have very few waves because their question of interest (income categories) was only asked in three waves.
+They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation and central exams, among other things.
+The correlations are very suggestive  as explanatory mechanisms but they are very cautios in drawing causality, specially because they are all reasonably correlated among each other.
+
+One interesting question is still missing is whether these country gaps have evolved over time.
+With their data, they only have 3 countries which are present in all waves but they also have very few waves because their question of interest (income categories) was only asked in three time points.
 That is why their study is more about between country gaps rather than the magnitude and evolution of the gaps.
@@
 Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}.
-Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as elimination tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
+Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as eliminating tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
 
-Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
-Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
-Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
-Talk more about how countries with high social mobility has been linked to smallest achievement gaps
-Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.
+% Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
+% Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
+% Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
+% Talk more about how countries with high social mobility has been linked to smallest achievement gaps
+% Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.
 
@@
 
-a) We want to test the notion that better performing countries have a smaller degree of inequality than other countries.
-\citep{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
-This pattern is not so obvious, however.
-For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score.
-But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.
-
-b) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
+a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
 After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
@@
 
-c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year available.
+b) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year available.
 This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
-Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders.}
+Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}
 
-d) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
+c) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
 The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind.
@@
 If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.
+
+d) We want to test the notion that better performing countries have a smaller degree of inequality than other countries.
+\citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
+This pattern is not so obvious, however.
+For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score.
+But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.
 
@@
 Classically, they ask student's their parent's educational level.
-Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}(cite more here).
-This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding.
-In spite of this, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time.
+Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}.
+This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have small differences.
+In spite of this, another limitation is the fact that parent's education is measured using the ISCED classification, something that has changed over time.
 For example, until PISA 2009, the preferred framework was ISCED 1997, whereas the next wave switched to the newly developed ISCED 2011 classification.
 Both these classification schemes have equivalent look-up tables, but this requires a detailed inspection of the codings.
-Aside from all these limitations, this is very reliable and important variable that measures an important factor of socio-economic status.
 
-Another social background variable PISA collects is the International Socio-Economic Index of Occupational Status (ISEI).
-This variable captures the economic status of the family, the closest it can without asking for income information.
-This index variable was developed by Ganzeboom (cite) and it attempts to measure occupational status using a continuous measure.
-This indicator has been found to ourperform classical occupational classifications such as the Erikson-Goldthorpe-Portocarero schema (cite).
-It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature \citep{anna2016_global}.
+Another social background variable is the International Socio-Economic Index of Occupational Status (ISEI).
+This variable attempts to capture the social status of the family, without asking for income information.
+This index variable was developed by \citet{ganzeboom1996, ganzeboom2010} and it attempts to measure occupational status using a continuous measure.
+This indicator has been found to ourperform classical occupational classifications such as the Erikson-Goldthorpe-Portocarero classification \citet{erikson1979}.
+It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}.
 PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.
@@
 
-\citet{reardon_portilla, anna2016, anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt in this paper.
+\citet{reardon_portilla}, \citet{anna2016} and \citet{anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt in this paper.
 SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.
-The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th \citep{micklewright} while the literature has been following the standard set by \citet{reardon2011} to use 90th/10th percentiles.
+The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, as others have used, for example, the 5th, 50th and 95th \citep{micklewright}.
+We use the conventional cutoffs as the literature has been following the standard set by \citet{reardon2011} to use 90th/10th percentiles.
 
-For each country in each study, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps.
-
+For each country in each wave, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps.
 The original strategy of \citet{reardon2011} is as follows: first, achievement is standardized (see below for a mathematical explanation of the standardization).
@@
 This new standardized variable has a mean of zero.
-Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves.
-However, if the variance of academic achievement changes over time, then standardizing the overall score at each country-wave pair actually makes the variable more biased.
+Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves because the test scores have now the same metric across time.
+However, if the variance of academic achievement changes over time, then standardizing the overall score at each country-wave pair actually makes the transformation biased.
 That is, by standardizing we're forcing the standard deviation of test scores to be zero across all waves.
@@
 
-{\centering \includegraphics[width=5.5in,height=5in]{figure/variance_pisa-1}
+\begin{figure}
+\begin{center}
+
+{\centering \includegraphics[width=3.5in,height=3in]{figure/variance_pisa-1}
 
@@
 
-Inspection of the data suggests that it's something we shouldn't be deeply concerned with.
+\caption{Standard deviation of test scores across all waves.}
+\end{center}
+\end{figure}
+
+A plotting ot the standard deviation of the data suggests that it's something we shouldn't be deeply concerned with.
 The standard deviation of each wave seems to be following a very similar pattern with a not so drastic exception of the year 2000.
@@
 This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citep{reardon2011}.
-PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but we still have to adjust for this imprecision.
+PISA has tried to make sure the tests are comparable across waves \citep{pisa2012_technical}, but we still have to adjust for this imprecision.
 
-In order to correct gap estimates for measurement error, I adjust student test scores by
+In order to correct gap estimates for measurement error, I adjust student test scores(\begin{math} \hat{Y_i} \end{math}) by
 

### 2017-07-04 paper/paper.Rnw
Replaced evolution of achievement graph

@@
 % Figure out how to ouput data frames as latex tables.
-<<sample_size, results = 'tex'>>=
+<<sample_size, include = TRUE, out.width = '4in'>>=
 # Get sample counts for each dummy
@@
 print(table_coming,
+      type = "latex",
       add.to.row = addtorow,
@@
 complete_data_topbottom %>%
-  filter(country %in% countries) %>%
   mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
-  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
+  filter(type_test == "math", !is.na(country)) %>%
+  ggplot(aes(as.character(wave), difference)) +
+  geom_point() +
+  geom_linerange(aes(ymin = 0, ymax = difference))+
   geom_line(stat = "smooth", method = "lm", aes(group = 1),
-            formula = y ~ splines::ns(x, 3), linetype = "longdash",
-            colour = "black") +
-  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
-  geom_hline(yintercept = 0, linetype = "longdash") +
-  geom_line(alpha = 0.4) +
-  geom_point(size = 0.5, alpha = 0.4) +
-  coord_cartesian(ylim = c(0, 3)) +
-  facet_wrap(~ country)
+            formula = y ~ splines::ns(x, 2), size = 0.7,
+            colour = "red") +
+  facet_wrap(~ country, ncol = 5) +
+  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
+  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
+  coord_fixed() +
+  ggthemes::theme_few() +
+  theme(panel.spacing = unit(1, "lines"),
+        panel.grid.major.y = element_line(colour = "grey")) +
+  ggtitle("Evolution of the 90/10 achievement gap")
+
 @
@@
 
-<<>>=
-
-complete_data_topbottom %>%
-  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
-  filter(type_test == "math", !is.na(country)) %>%
-  ggplot(aes(as.character(wave), difference)) +
-  geom_point() +
-  geom_linerange(aes(ymin = 0, ymax = difference))+
-  geom_line(stat = "smooth", method = "lm", aes(group = 1),
-            formula = y ~ splines::ns(x, 2), size = 0.7,
-            colour = "red") +
-  facet_wrap(~ country, ncol = 12) +
-  viridis::scale_fill_viridis(option = "B") +
-  scale_y_continuous(expand = c(0, 0), lim = c(0, 3)) +
-  ggthemes::theme_few() +
-  theme(panel.spacing = unit(1, "lines"),
-      axis.text.x = element_text(angle = 90, hjust = 1))
-
-@
 The top line belongs to the high SES group whereas the bottom line belongs to the low SES group.

### 2017-07-04 paper/paper.tex
Replaced evolution of achievement graph

--- 2017-06-22 paper/paper.tex
@@
-\documentclass[11pt, a4paper]{article}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-\let\hlipl\hlkwb
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
+\documentclass[11pt, a4paper]{article}
 \bibliographystyle{apalike}
@@
 \author{Jorge Cimentada}
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+
 \begin{document}
@@
 % the results_* models have been created once and the cache can save them
-
-\section{Introduction}
-
-\section{Literature Review}
-% You have a bunch of citations to take from here
-Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years.
-Are we destined to belong to the same social status as our parents?
-Is meritocracy over, if it ever existed?
-These questions have motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century.
-When James Colem(1966) released his famous Coleman report it raised the topic to starlight by suggesting that family SES and student performance are tightly linked.
-Fast-forward today, we've understood the relationship much better, although far from being clear.
-The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school (cite heckman).
-Moreover, these inequalities can be narrowed significantly if the children coming from unfavourable conditions are exposed to high quality learning environments.
-One particular finding coming from (cite heckman skill begets skill) is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\).
-Although this sounds logical, it is often not acknowledged properly.
-
-The implications of this cummulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\).
-That is why the whole debate about narrowing inequality between polarized families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns.
-Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status, is present in virtually all empirical studies of social mobility and inequalities (cite breen and goldthorpe, cite waldfogel what children need, cite bradbury 2015, cite chetty for recent evidence).
-What we don't know is how and why this relationship is formed from such an early age.
-Moreover, many have suggested different explanations and the topic has been very contested (duncan and brooks-gun 1997).
-
-Social mobility has deepend our knowledge on the odds and relationship between intergenerational transmission of SES.
-The field of economics of education has also helped to understand that education is by far the investment that yields the best outcomes not only for the child and his/her family, but to the economy as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth (cite hanushk and woesmann study).
-We've become so intertwined with the subject that we've virtually spent most of our time studying the mechanisms through which this inequality comes about.
-Naturally, we want to do that to be able to reverse it and help disadvantaged children reach their fullest potential.
-Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities.
-In addition, since most research has concentrated on the American case, we haven't been able to learn which countries have the biggest or smallests gaps, and whether they've been narrowing or closing over time.
-
-Virtually in all countries, be it developed or developing, there is inequality of opportunity.
-But there is considerable variation in the magnitud of this effect.
-For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries \citep{esping2012, breen2007, shavit1993}.
-
-Taking the Danish example, we learned a great deal about how to improve social mobility by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility for less advantaged children (Bjorklund and Jantti, 2009; Jaeger and Holm, 2007).
-But the important finding came when we discovered the main reason behind their social elevator: its' educational system.
-For example, research by Esping and Waldfogel and Gosta's phd student on Denmark, show that the Danish early education system is particularly efficient at giving free high quality education, regardless of socio-economic status.
-Moreover, it tracks students into different curricula very late compared to other Europen countries Europe (age 16).
-The importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
-
-Recent research on educational inequality has found that achievement gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years.
-Literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape.
-The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940.
-This trend has allowed researchers to study achievement gaps in a lengthy period of time.
-In fact, using this information, \citep{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
-Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap (cite study of white-black gap -waldfogel study).
-Surprisingly, the gaps have reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap.
-The widening of the achievement gap has been happening in parallel to the growth of income inequality, although very suggestive, it is hard to link both things causally.
-However, a recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979.
-If income inequality would be a strong a direct function of the achievement gap, we should expect for the white-black achievement gap to be widening rather than narrowing, as it is now (Magnuson and Waldfogel 2008) \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
-
-\citet{reardon2011} finds that the increase has occured predominantely from the 1970's until the 2000's.
-In fact, the hard numbers suggest that the gap widened by ~ 40-50\%.
-The author also estimates the rate of change using data from the 1940's and finds an even higher increase of ~ 75\%.
-Given that the studies before 1970 are less reliablle in terms of comparability and sampling design, the author computes all results for both before/after 1970.
-To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2010.
-Using longitudinal data, \citet{bradbury2015} find similar results.
-Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25.
-Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of ~ 1.25/1.50 standard deviations.
-\footnote{One important drawback of these studies is that they don't present the uncertainty of this estimates.
-Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
-It could be that the gap is at ~ 1.25 standard deviations, varying up to 1.75 and down to 0.80.
-Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.}
-
-In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend.
-This follow-up study concentrated solely on the Kindergarten gap in which they took a snapshot of the achievement gap in Kindergarten in 1998, 2006 and 2010 and found that the 90th/10th income gap in readiness closed modestly.
-Furthermore, using data from fall and spring in the same kindergarden year, it seems like the gap in narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010.
-In contrast to \citep{reardon2011}, we find that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates.
-Their results not only hold for the income achievement gap, but they also recorded a particular decline in the white-hispanic gap.
-
-The American situation is a very distinctive one relative to Europe.
-The reasons the authors find a reversal in the trend could be numerous and should be studied very close.
-For example, the American educational system lacks any formal curricular differentiation \emph{à la} European style.
-It is true nonetheless that \citet{reardon2011} did not cover from the period 2000 onwards, which is specifically the trend they studied.
-In contrast, \citet{reardon2011} suggests that the reversal is likely due to the high increase of preschool enrollment.
-They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
-This explanation still needs further confirmation.
-
-Motivated by these recent results, some authors have taken this analysis to an international context in order to discover between-country trends.
-The unique work of \citet{bradbury2015} employs a unique comparative analysis of Australia, United Kingdom, United States and Canada.
-Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is not only the first study that uses panel data to study achievement gaps, but to also do it between countries}.
-The core finding behind the book is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada.
-They find that once the achievement gap is present in early school entry, it doesn't seem to narrow or neither widen very much over the life course.
-In fact, they calculate that the schooling environment once they enter school can only explain about ~ 30\%-40\% of the high school SES gap.
-This suggests that once the achievement gap widens before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous RCT's show that high quality schooling can indeed help ease the gap, in some instances even close it.}.
-Once exception is the United Kingdom, which they find closed the gap in early primary years.
-This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health.
-
-One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula.
-These four countries have no major jump in school selection, quite the opposite to the average European country.
-A thorough review by \citet{werfhorst_mijs} sheds some light on the subject.
-First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps.
-\citet{hanushek_woesmann_tracking}, use the Progress in International Reading Literacy Study (PIRLS), the Trends in International Mathematics and Science Study (TIMSS) and the Programme for International Student Assessment (PISA) surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
-The results suggest that early tracking increases educational inequality.
-While less clear, there is also a tendency for early tracking to reduce mean performance.
-Moreover, \citet{micklewright} using similar data but a different empirical strategy, finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
-In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
-
-Another limitation of their study is that their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
-Their findings are very reliable but should be taken as suggestive at best.
-For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citep{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in educational inequality in many countries using much more comparable studies.
-These studies tackle a completely different question from the above, namely to use study cross-sectional differences between many countries, instead of over-time analysis of student gaps.
-However, they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
-
-\citet{anna2016}, again using PIRLS, TIMSS and PISA, assess whether there are patterns of cross-national variation in the achievement gap.
-In other words, does the achievement gap differ between countries?
-Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
-In comparison to the literature on achievement gaps, they find that the U.S has a gap of ~ 1.20 SD in 2001 which increase to around ~ 1.30 in the year 2006 while Germany has a decreasing gap from around ~ 1.25 to ~ 1 standard deviation in the same year-span.
-However, these numbers vary a lot and carry a great deal of overlapping uncertainty.
-
-They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation, central exams, among other things.
-The correlations are very suggestive  as explanatory mechanisms but they are very cautios in drawing causality, specially because they are all reasonably correlated.
-However, one interesting question is whether these country gaps have evolved over time.
-With their data, they only have 3 countries which are present in all waves but they also have very few waves because their question of interest (income categories) was only asked in three waves.
-That is why their study is more about between country gaps rather than the magnitude and evolution of the gaps.
-
-\citet{anna2016_global}, building on the work of \citet{anna2016} and \citet{reardon_portilla} pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries.
-With over 50 years of data, and over 100 countries, she finds that there seems to be a general pattern of increasing achievement gap.
-However, once she disentangles the relationship by country, she finds a sizable amount of heterogeneity, with some countries seeing the achievement gap narrowing, others no change at all, while others record a steady increase.
-This is revealing because it doesn't really pay off to look at general average once each country has their own distinctive gap and evolution.
-Another important finding is that the achievement gap is clearly no universal and should be studied in context.
-
-One clear limitation of their study (as well as \citet{reardon2011}) is that the adjust for the age of each child in all studies.
-Although for their modeling purposes this is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
-However, as we've noted before, \citet{bradbury2015} find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as \citet{reardon2011} did.
-We clearly see in \citet{reardon_portilla} that paying attention to age-specific gaps varies reasonably even in short periods of time.
-
-In fact, the evolution of high/low SES gaps for preschool children might be much less marked than the same gap for high school children.
-The explanation, although very debated, has been gaining much support in recent years.
-In countries with high levels of curricular differentiation, the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}.
-Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}.
-Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as elimination tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
-
-Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
-Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
-Note how many of these studies haven't really concentrated on who is getting better or worse: top or bottom?
-Talk more about how countries with high social mobility has been linked to smallest achievement gaps
-Talk about Durpiez and Dumay and how there's not relationship between inequality income - inequality achievement in contrast to reardon and anna who find some relationship.
-
-\section{Research questions}
-
-With this being said, this paper introduces one novelty in the literature which is to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries.
-This is different from previous work because it concentrates solely on 15 year old children, and it attempts to capture the evolution of the achievement gap for many countries.
-The advantages of this study are twofold.
-
-First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the Kindergarten gap in \citet{reardon_portilla}.
-As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
-Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.
-
-Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States.
-This will help asses the mangitude of the inequality as well as the rate of growth/decline.
-Thirdly, this paper not only documents the size and changes of the SES gap, but studies with particular attention the source of the changes.
-That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.
-
-More formally, We're interested in studying
-a) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
-b) Which countries are experiencing changes in the achievement gap;
-c) the rate at which each country-gap is widening/narrowing and;
-d) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
-
-We develop each question separately for more detail.
-
-a) We want to test the notion that better performing countries have a smaller degree of inequality than other countries.
-\citep{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
-This pattern is not so obvious, however.
-For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score.
-But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.
-
-b) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
-After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
-This shows that it is important to study the changes in the achievement gap over time.
-This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.
-
-c) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year available.
-This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
-Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders.}
-
-d) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
-The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind.
-We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.
-This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving.
-If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.
-
-\section{Methods}
-
-\subsection{Data}
-
-To investigate the above mentioned questions I will be using the Programme for International Student Assessment (PISA).
-PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students.
-Currrently, PISA has six waves starting which started in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 countries, both developed and developing ones.
-
-PISA collects data through a two-stage stratified sampling design.
-With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests.
-The sample size for PISA 2000 is 127,388, 276,165 for PISA 2003, 398,750 for PISA 2006, 515,958 for PISA 2009, 480,810 for PISA 2012, and 519,334 for PISA 2015.
-Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance.
-With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system.
-None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.
-
-To identify a family's socio-economic status PISA collects several variables that measure different dimensions.
-Classically, they ask student's their parent's educational level.
-Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}(cite more here).
-This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have a different coding.
-In spite of this, another limitation is the fact that parent's education is measured with the ISCED classification, something that has changed over time.
-For example, until PISA 2009, the preferred framework was ISCED 1997, whereas the next wave switched to the newly developed ISCED 2011 classification.
-Both these classification schemes have equivalent look-up tables, but this requires a detailed inspection of the codings.
-Aside from all these limitations, this is very reliable and important variable that measures an important factor of socio-economic status.
-
-Another social background variable PISA collects is the International Socio-Economic Index of Occupational Status (ISEI).
-This variable captures the economic status of the family, the closest it can without asking for income information.
-This index variable was developed by Ganzeboom (cite) and it attempts to measure occupational status using a continuous measure.
-This indicator has been found to ourperform classical occupational classifications such as the Erikson-Goldthorpe-Portocarero schema (cite).
-It has been scaled for comparability between waves and some authors have used it for inequality studies finding expected results to be consistent with social origin literature \citep{anna2016_global}.
-PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.
-
-Yet one of the most relevant variables for our study is a composite SES index created by the PISA team.
-The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measured the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}.
-This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable metrics across waves.
-To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.
-
-Aside from SES, the other most relevant variables are test scores for Mathematics, Literacy and Science.
-PISA does not provide a single test result for each respondent.
-Instead, it provides a \emph{series} of 'plausible values' that the child could actually score.
-As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured (the true distribution of the possible scores a student can achieve).
-Suppose we have \(\mu_i\), the average student test score in Mathematics for student \(i\).
-Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\) for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test.
-This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\).
-The reason why we use this procedure is because estimating a single estimate \(\mu_i\) is plagued with measurement error, among other types of bias \citep[see][]{wu2005}.
-The number of plausible values for PISA waves are usually five (although ten for PISA 2015) random draws from this distribution.
-In practice each student has 5 scores for each test, that resembles this distribution.
-They are continuous, ranging from 0 to 500, with a mean of 250.
-
-\subsection{Data analysis}
-
-The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries.
-To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004, hanushek_woesmann_tracking, saw2016, bradbury2015, byun2010}.
-The actual calculation of the achievement gap varies substantially and different strategies have been implemented.
-For example, \citet{micklewright} calculates the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution.
-Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy.
-\citet{saw2016}, for instance, used parental education as a proxy of SES, whereas \citet{byun2010} use a similar SES index as ours, but created by them.
-
-\citet{reardon_portilla, anna2016, anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt in this paper.
-SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.
-The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, others have used, for example, the 5th, 50th and 95th \citep{micklewright} while the literature has been following the standard set by \citet{reardon2011} to use 90th/10th percentiles.
-
-For each country in each study, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps.
-
-The original strategy of \citet{reardon2011} is as follows: first, achievement is standardized (see below for a mathematical explanation of the standardization).
-We then use it to calculate the mean achievement (and standard error) for each category of the SES variable of interest (parent's education, income categories, etc..).
-"Category means are plotted at their percentile ranks and cubic models are fit through the points using weighted least squares.
-Finally, achievement at each country’s 90th and 10th SES percentiles is interpolated from the model" \citep{anna2016_global}.
-The result is a SES gap from an ordinal variable of interest.
-
-As mentioned before, PISA does not provide a single achievement indicator.
-Instead, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.
-
-To standardize the test score we fit a linear model
-
-\begin{equation}
-Y_i = \alpha + \beta_1 * AGE_i + \epsilon_i, \epsilon_i ~ N(0, \sigma^2)
-\end{equation}
-
-for each wave,  where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (because all children should be 15 year olds) weighted by the student sample weights.
-
-We then adjust \begin{math} \hat{Y_i} \end{math} by
-
-% I think the denominator could be wrong! Because I include y_hat and that's what the equation is estimating
-\begin{equation}
-\hat{Y_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
-\end{equation}
-
-% # Take residuals of model and divide by rmse. Multiply that by
-% # 1 / sqrt(reliability of each survey), which is .y in the loop.
-% .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)
-
-Where \begin{math}\hat{\epsilon_i}\end{math} is the residual for student \begin{math}i\end{math} and the denominator is the root mean square error of the model.
-
-This new standardized variable has a mean of zero.
-Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves.
-However, if the variance of academic achievement changes over time, then standardizing the overall score at each country-wave pair actually makes the variable more biased.
-That is, by standardizing we're forcing the standard deviation of test scores to be zero across all waves.
-But if the true deviations of the median academic achievement changes over time, then the estimated trend in the SES gaps will be underestimated, or vice versa.
-
-% Show graph of variance over time
-% I'm showing standard deviations, is it better to show variance?
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/variance_pisa-1}
-
-}
-
-Inspection of the data suggests that it's something we shouldn't be deeply concerned with.
-The standard deviation of each wave seems to be following a very similar pattern with a not so drastic exception of the year 2000.
-
-Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap.
-This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citep{reardon2011}.
-PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but we still have to adjust for this imprecision.
-
-In order to correct gap estimates for measurement error, I adjust student test scores by
-
-\begin{equation}
-\hat{Y_i} = \hat{Y_i} * \frac{1}{\sqrt{r}}
-\end{equation}
-
-where \begin{math}r\end{math} is the reliability score of the wave.
-\footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}.
-Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own.
-At the moment of writing this paper, they were unavailable.
-For this reason we implement the analysis following the original work of \citet{reardon2011}}.
-Each PISA survey provides a reliability indicator which we use accordingly.
-This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.
-
-After constructing the adjusted test score measurement, we estimate the SES dummy by calculating the weighted 90th and 10th quantiles for the SES composite index, and then generating a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.
-
-We then fit
-
-\begin{equation}
-Y_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
-\end{equation}
-
-% mod2 <-
-%   lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
-%        data = .x,
-%        weights = .x[[weights_var]])
-
-where \(SES_i\) is whether the student is in the top/bottom SES dummy and we allow both \(\alpha_i\) and \(\beta_i\) to vary by country \(j\).
-We implement this separately for each wave and weight by the wave-specific student sample weights.
-
-Finally, we calculate the SES achievement gap for each country by extracting the fitted \(\alpha\) and \(\beta\) for each country \(j\) and calculating the difference of the predicted score of High and Low SES.
-
-\begin{equation}
-\begin{split}
-\quad \text{High SES}_j = \alpha_j + \beta_j \\
-\quad \text{Low SES}_j = \alpha_j \\
-\quad \text{SES gap}_j = \text{High SES}_j - \text{Low SES}_j
-\end{split}
-\end{equation}
-
-We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends.
-We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size.
-Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible.
-
-\section{Analysis}
-
-\subsection{Descriptives}
-
-% Show the properties of the ESCS dummy variable
-% % in high educated
-% % of home possessions
-% % of ISEI
-
-% Figure out how to ouput data frames as latex tables.
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/correlation_incomeineq-1}
-
-}
-
-\subsection{Results}
-
-Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
-They also find that France is a lower dispersion country (but in my results this is changing a lot).
-They also find that Germany and US are very dispersed countries (quite the opposite)
-
-\citet{micklewright} find that France has a very small gap in 2000, something we find very clearly in our data, and the U.S has a very big gap in 2000, something we also find in our data.
-\citet{micklewright} find
-that average achievement is negatively related to dispersion.
-
-It seems that social origin effects have actually increased in France  (Ballarino and Bernardi, 2016)
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_9010gaps-1}
-
-}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_allgaps-1}
-
-}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_ses_growth-1}
-
-}
-
-% How to output data frames are latex tables?
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/rate_change_graph-1}
-
-}
-
-% How to output data frames are latex tables?
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_perc_increase-1}
-
-}
-
-\bibliography{mybibliography.bib}
-
-\end{document}

### 2017-07-05 paper/paper.Rnw
All graphs and tables visible in tex. Rearranging and recommenting

--- 2017-07-03 paper/paper.Rnw
@@
 \usepackage[colorlinks]{hyperref}
+\usepackage{booktabs}
 \AtBeginDocument{%
@@
   library(xtable)
+  library(ggthemes)
 
@@
 % Figure out how to ouput data frames as latex tables.
-<<sample_size, results = 'tex'>>=
+<<sample_size>>=
 # Get sample counts for each dummy
@@
 sample_tables_midbottom <- sample_size_calc(adapted_year_data, c(.1, .5), selected = TRUE, countries)
-
+@
+
+<<table_escs, include = TRUE>>=
 summary_table <-
@@
 print(table_coming,
+      scalebox = '0.75',
+      floating = FALSE,
       add.to.row = addtorow,
@@
       include.rownames = FALSE,
-      hline.after = NULL,
+      hline.after = nrow(table_coming),
       caption.placement = "top",
@@
 
-<<correlation_incomeineq, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+<<correlation_incomeineq, include = F, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
 complete_data_topbottom %>%
@@
 
+\begin{figure}
+\begin{center}
 <<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
@@
 complete_data_topbottom %>%
-  filter(country %in% countries) %>%
   mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
-  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
+  filter(type_test == "read", !is.na(country)) %>%
+  ggplot(aes(as.character(wave), difference)) +
+  geom_point() +
+  geom_linerange(aes(ymin = 0, ymax = difference))+
   geom_line(stat = "smooth", method = "lm", aes(group = 1),
-            formula = y ~ splines::ns(x, 3), linetype = "longdash",
-            colour = "black") +
-  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
-  geom_hline(yintercept = 0, linetype = "longdash") +
-  geom_line(alpha = 0.4) +
-  geom_point(size = 0.5, alpha = 0.4) +
-  coord_cartesian(ylim = c(0, 3)) +
-  facet_wrap(~ country)
-@
+            formula = y ~ splines::ns(x, 2), size = 0.7,
+            colour = "red") +
+  facet_wrap(~ country, ncol = 5) +
+  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
+  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
+  ggthemes::theme_few() +
+  theme(panel.spacing = unit(1, "lines"),
+        panel.grid.major.y = element_line(colour = "grey")) +
+  ggtitle("Evolution of the 90/10 achievement gap")
+@
+\end{center}
+\end{figure}
 
 We start by look at the achievement for some countries.
-We plot both mathematics and reading gaps for each country and also a cubic trend spline for both test scores pooled.
+We plot only mathematics for each country and also a quadratic trend spline for both test scores pooled.
 Some countries have increased their achievement very strongly.
@@
 
-However, we don't know what happens over time.
-But Ballarino and Bernardi (2016) did study the relationship of social origin effects and found that they increased very much in France in the last decades.
+However, we don't know what happens over time. But \citet{bernardi2016} did study the relationship of social origin effects and found that they increased very much in France in the last decades.
 
@@
 Take the case of Sweden.
-The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation.
+The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015.
 The actual difference is of about \Sexpr{diff_data$math[["Sweden"]]}.
@@
 
-Something reassuring is that mathematics and reading follow basically the same trend across all countries.
+The plot below takes a more direct approach and looks at the percentage change from the first and last timepoint available.
+Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.
+
+% How to output data frames are latex tables?
+<<perc_increase_tables>>=
+# Show the rates at which is increasing/decreasing
+perc_increase_fun <- function(df) {
+
+  # Average standard deviation increase
+  data_ready <-
+    df %>%
+    select(wave, country, type_test, difference) %>%
+    group_by(type_test) %>%
+    split(.$country) %>%
+    map(~ {
+      .x <-
+        spread(.x, wave, difference) %>%
+        ungroup()
+
+      year_vars <- sum(map_dbl(.x, is.numeric)) - 1
+      years_subtract <- names(.x)[c(ncol(.x) - year_vars, ncol(.x))]
+      years_subtract <- lapply(years_subtract, as.name)
+
+      last_year <- rlang::new_quosure(years_subtract[[2]], env = .GlobalEnv)
+      first_year <- rlang::new_quosure(years_subtract[[1]], env = .GlobalEnv)
+
+      years_available <-
+        .x %>%
+        gather(year, val, -(country:type_test)) %>%
+        group_by(type_test) %>%
+        summarise(yr_avaible = sum(!is.na(val))) %>%
+        pull(yr_avaible)
+
+      year_sd <-
+        .x %>%
+        gather(year, val, -(country:type_test)) %>%
+        split(.$type_test) %>%
+        map_dbl(~ bootstrapper(.x, mad(val, na.rm = T), B = 100) %>% .[[2, 2]]) %>%
+        round(2) * 100
+
+      .x %>%
+        map_if(is_double, round, 3) %>%
+        as_tibble() %>%
+        transmute(type_test,
+                  country,
+                  perc_diff = round(((!!last_year) - (!!first_year)) * 100, 1),
+                  sd_year = year_sd,
+                  diff_lower = perc_diff - 1 * year_sd,
+                  diff_upper = perc_diff + 1 * year_sd,
+                  years_available = years_available)
+    })
+  data_ready
+}
+
+top_bottom_perc <- perc_increase_fun(complete_data_topbottom)
+top_mid_perc <- perc_increase_fun(complete_data_topmid)
+mid_bottom_perc <- perc_increase_fun(complete_data_midbottom)
+
+# Gap is closing at an average of the variable diff per year.
+@
+
+\begin{figure}
+\begin{center}
+<<graphing_perc_increase, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+perc_graph <- function(df, test, title, subtitle = NULL) {
+  df %>%
+  enframe(name = "country") %>%
+  unnest(value) %>%
+  filter(country %in% countries) %>%
+  select(-country1, -years_available) %>%
+  mutate(diff_95_lower = perc_diff - 2*sd_year,
+         diff_95_upper = perc_diff + 2*sd_year) %>%
+  setNames(c("Country", "Type of test", "Average % difference", "Average SD",
+             "Lower 50% bound", "Upper 50% bound", "Lower 95% bound", "Upper 95% bound")) %>%
+  filter(`Type of test` == test) %>%
+  arrange(`Average % difference`) %>%
+  mutate(Country = ordered(forcats::as_factor(Country)),
+         Country_num = as.numeric(Country)) %>%
+  ggplot(aes(Country, `Average % difference`, fill = `Average % difference` > 0)) +
+  geom_hline(yintercept = 0, linetype = "longdash") +
+  geom_point() +
+  geom_ribbon(aes(x = Country_num, ymin = `Lower 95% bound`, ymax = `Upper 95% bound`), alpha = 0.3) +
+  geom_ribbon(aes(x = Country_num, ymin = `Lower 50% bound`, ymax = `Upper 50% bound`), alpha = 0.2) +
+  geom_linerange(aes(ymin = 0, ymax = `Average % difference`,
+                     colour = `Average % difference` > 0), alpha = 0.4) +
+  scale_y_continuous(name = "Avg % increase/decrease",breaks = seq(-160, 160, 40)) +
+  scale_fill_discrete(guide = FALSE) +
+  scale_colour_discrete(guide = FALSE) +
+  coord_cartesian(ylim = c(-160, 160)) +
+  ggtitle(title, subtitle) +
+  ggthemes::theme_few() +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+}
+
+perc_graph(top_bottom_perc, "math", "90/10 achievement gap", "Percentage change from 2000 to 2015")
+# m2 <- perc_graph(top_mid_perc, "90/50")
+# m3 <- perc_graph(mid_bottom_perc, "50/10")
+
+# gridExtra::grid.arrange(m1, m2, m3, nrow = 1)
+@
+\end{center}
+\end{figure}
+
+Generally speaking, we see that most countries increased their achievement gap.
+France had an average increase of about 80\% since 2000, whereas Germany had a similar figure but decreasing.
+Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap.
+In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%.
+
+<<evolution_reading, eval = F>>=
+# See below that reading graph is upon request. This is the one.
+complete_data_topbottom %>%
+  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
+  filter(type_test == "read", !is.na(country)) %>%
+  ggplot(aes(as.character(wave), difference)) +
+  geom_point() +
+  geom_linerange(aes(ymin = 0, ymax = difference))+
+  geom_line(stat = "smooth", method = "lm", aes(group = 1),
+            formula = y ~ splines::ns(x, 2), size = 0.7,
+            colour = "red") +
+  facet_wrap(~ country, ncol = 5) +
+  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
+  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
+  coord_fixed() +
+  ggthemes::theme_few() +
+  theme(panel.spacing = unit(1, "lines"),
+        panel.grid.major.y = element_line(colour = "grey")) +
+  ggtitle("Evolution of the 90/10 achievement gap")
+
+perc_graph(top_bottom_perc, "read", "90/10 achievement gap", "Percentage change from 2000 to 2015")
+
+@
+
+Something reassuring is that mathematics and reading follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase.
+The previos two graphs for reading are available upon request}.
 This means that the result is not an artifact of chance.
@@
 
-<<>>=
+\begin{figure}
+\begin{center}
+<<rate_increase/decrease, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
 # You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
@@
             colour = "black") +
+  scale_y_continuous(name = "Standardized test scores (mean 0)") +
+  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
+  scale_colour_discrete(name = NULL, labels = c("Low SES", "High SES")) +
+  ggtitle("Evolution of the achievement gap by top/bottom groups") +
   # geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
@@
   # coord_cartesian(ylim = c(0, 3)) +
-  facet_wrap(~ country)
-@
-
-<<>>=
-
-complete_data_topbottom %>%
-  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
-  filter(type_test == "math", !is.na(country)) %>%
-  ggplot(aes(as.character(wave), difference)) +
-  geom_point() +
-  geom_linerange(aes(ymin = 0, ymax = difference))+
-  geom_line(stat = "smooth", method = "lm", aes(group = 1),
-            formula = y ~ splines::ns(x, 2), size = 0.7,
-            colour = "red") +
-  facet_wrap(~ country, ncol = 12) +
-  viridis::scale_fill_viridis(option = "B") +
-  scale_y_continuous(expand = c(0, 0), lim = c(0, 3)) +
-  ggthemes::theme_few() +
+  facet_wrap(~ country, ncol = 5) +
+  theme_few() +
   theme(panel.spacing = unit(1, "lines"),
-      axis.text.x = element_text(angle = 90, hjust = 1))
-
-@
+        legend.position = "bottom")
+@
+\end{center}
+\end{figure}
+
 The top line belongs to the high SES group whereas the bottom line belongs to the low SES group.
@@
 For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th?
-Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier.
-Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous plots \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.
+Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier?
+Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous graphs \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.
 
@@
 
+\begin{figure}
+\begin{center}
 <<graphing_allgaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
@@
   geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
-  geom_hline(yintercept = 0, linetype = "longdash") +
-  geom_line() +
-  geom_point(size = 0.5) +
-  coord_cartesian(ylim = c(-0.5, 3)) +
-  facet_grid(country ~ type)
-@
+  geom_line(alpha = 0.4) +
+  geom_point(size = 0.5, alpha = 0.4) +
+  coord_cartesian(ylim = c(0, 3)) +
+  scale_y_continuous(name = "Standardized test scores (mean 0)") +
+  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
+  scale_colour_discrete(name = NULL, labels = c("Low SES", "High SES")) +
+  ggtitle("Comparison of the achievement gap for three socio-economic groups") +
+  facet_grid(country ~ type) +
+  theme_few() +
+  theme(panel.spacing = unit(1, "lines"),
+        legend.position = "bottom")
+@
+\end{center}
+\end{figure}
 
@@
 
-But, similar as before, we don't know which group is causing the decrease/increase of the gap in the 90th/50th and 50th/10th gap. For that we plot the evolution of these gaps below.
-
-<<graphing_ses_growth, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
-# Graphing how the top/bottom are evolving over time instead of absolute difference
-
-complete_data_topbottom %>%
-  bind_rows(complete_data_topmid) %>%
-  bind_rows(complete_data_midbottom) %>%
-  filter(country %in% c("United States", "Denmark", "France", "Germany"), type_test == "math") %>%
-  select(wave, country, type_test, type, contains("math")) %>%
-  mutate(type = factor(type,
-                       levels = c("90th/10th SES gap", "90th/50th SES gap", "50th/10th SES gap"),
-                       ordered = TRUE)) %>%
-  gather(score, value, -(wave:type)) %>%
-  separate(score, c("ses", "score"), sep = 2) %>%
-  spread(score, value) %>%
-  ggplot(aes(as.factor(wave), mean_math, group = ses, colour = ses)) +
-  geom_errorbar(aes(ymin = lower_math, ymax = upper_math), width = 0.1) +
-  geom_line() +
-  geom_point(size = 0.5) +
-  coord_cartesian(ylim = c(-0.5, 3)) +
-  facet_grid(country ~ type)
-
-# Increase:  
-# Sweden - steady increase in both tests
-# Austria - increase in math - slight increase in read
-# Finland - very sharp increase in both
-# France - very sharp increase in both
-# Netherlands - sharp increase in both
-
-# Decrease:
-# US - decrease in both tests
-# Chile - decrease in both tests
-
-# No change:
-# Canada - stable red - increase math
-# UK - slight decrease red - stable math
-# Belgium - no change
-# Czech republic - no change
-# Denmark no change
-# Germany - no change
-# ITaly - no change
-# Japan -  no change
-# Norway - no change
-# Poland - no change
-# Spain - no change
-@
-
-With Denmark there's no average change across all years.
-However, for France, we see that the gap is growing much faster for the 50th/10th SES gap.
-For Germany we see that the gap is closing in all three groups although much more strongly for the 50th/10th SES gap.
-It looks like the 50th/10th gap is mainly driving the changes in the gaps for most of the countries that actually see a change over time.
-For the U.S, there's very little change in the 90th/50th SES gap, with most of the change coming from the 50th/10th gap.
-
-Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's an efficiency/equality tradeoff and if it stands over time.
-One particular topic that has been researched in the past is whether improving the overall cognitive development of a country compromises the distribution of the countries overall performance.
-For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance.
-This, however, can come at a price.
-If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance.
-
-<<graphing_achievement_disparity, cache = T>>=
-
-country_scores <-
-  map(adapted_year_data, ~ {
-  if (unique(.x$wave) == "pisa2015") {
-    intsvy::pisa2015.mean.pv("MATH", by = "country", data = .x)
-  } else {
-    intsvy::pisa.mean.pv("MATH", by = "country", data = .x)
-  }
-}) %>% setNames(seq(2000, 2015, 3))
-
-country_scores <-
-  enframe(country_scores, name = "wave") %>%
-  unnest() %>%
-  mutate(wave = as.double(wave))
-
-data_to_plot <-
-  left_join(complete_data_topbottom, country_scores, by = c("wave", "country")) %>%
-  select(wave, country, type_test, difference, Mean) %>%
-  filter(type_test == "math")
-
-data_to_plot %>%
-  ggplot(aes(scale(Mean), difference)) +
-  geom_point() +
-  geom_smooth()
-@
-
-For all country-years pooled we see a very strong correlation but contrary to what we would expect.
-We find that countries with high performance have also a very low level of inequality.
-In fact, even if we separate it by year.
-
-<<>>=
-data_to_plot %>%
-  ggplot(aes(scale(Mean), difference)) +
-  geom_point() +
-  geom_smooth(method = "lm") +
-  facet_wrap(~ wave)
-@
-
-<<>>=
-data_to_plot %>%
-  filter(country %in% countries) %>%
-  ggplot(aes(as.character(wave), difference, group = country)) +
-  geom_point(colour = "blue") +
-  geom_line(colour = "blue") +
-  geom_point(aes(y = scale(Mean)), colour = "red") +
-  geom_line(aes(y = scale(Mean)), colour = "red") +
-  facet_wrap(~ country)
-@
-
-In fact, we see a strong correspondence.
-In countries like Germany and France we see quite a functional form between both inequality and performance.
-This pattern is also seen in in the UK, Austria, Australia and New Zealand.
-Poland and Italy are quite interesting cases because it shows how inequality has been decreasing that much whereas average performance has been improving notably.
-On another hand, only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality.
-
-<<>>=
-data_to_plot %>%
-  semi_join(data_to_plot %>% group_by(country) %>% count() %>% filter(n > 3), by = "country") %>%
-  group_by(country) %>%
-  summarize(cor = cor(scale(Mean), difference, use = 'complete.obs')) %>%
-  ggplot(aes(reorder(country, cor), cor, fill = cor > 0)) + geom_col() + coord_flip()
-@
+In a more convenient way, we can inspect whether there is a lot of variance in the rate at which top/bottom groups are changing.
 
@@
 
+\begin{figure}
+\begin{center}
 <<rate_change_graph, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
@@
 @
-
-In a more convenient way, we can see that there is a lot of variance in the rate at which top/bottom groups are changing.
+\end{center}
+\end{figure}
+
 We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as United States and Germany.
@@
 
-% How to output data frames are latex tables?
-<<perc_increase_tables>>=
-# Show the rates at which is increasing/decreasing
-perc_increase_fun <- function(df) {
-
-  # Average standard deviation increase
-  data_ready <-
-    df %>%
-    select(wave, country, type_test, difference) %>%
-    group_by(type_test) %>%
-    split(.$country) %>%
-    map(~ {
-      .x <-
-        spread(.x, wave, difference) %>%
-        ungroup()
-
-      year_vars <- sum(map_dbl(.x, is.numeric)) - 1
-      years_subtract <- names(.x)[c(ncol(.x) - year_vars, ncol(.x))]
-      years_subtract <- lapply(years_subtract, as.name)
-
-      last_year <- rlang::new_quosure(years_subtract[[2]], env = .GlobalEnv)
-      first_year <- rlang::new_quosure(years_subtract[[1]], env = .GlobalEnv)
-
-      years_available <-
-        .x %>%
-        gather(year, val, -(country:type_test)) %>%
-        group_by(type_test) %>%
-        summarise(yr_avaible = sum(!is.na(val))) %>%
-        pull(yr_avaible)
-
-      year_sd <-
-        .x %>%
-        gather(year, val, -(country:type_test)) %>%
-        split(.$type_test) %>%
-        map_dbl(~ bootstrapper(.x, mad(val, na.rm = T), B = 100) %>% .[[2, 2]]) %>%
-        round(2) * 100
-
-      .x %>%
-        map_if(is_double, round, 3) %>%
-        as_tibble() %>%
-        transmute(type_test,
-                  country,
-                  perc_diff = round(((!!last_year) - (!!first_year)) * 100, 1),
-                  sd_year = year_sd,
-                  diff_lower = perc_diff - 1 * year_sd,
-                  diff_upper = perc_diff + 1 * year_sd,
-                  years_available = years_available)
-    })
-  data_ready
+But, similar as before, we don't know which group is causing the decrease/increase of the gap in the 90th/50th and 50th/10th gap. For that we plot the evolution of these gaps below.
+
+\begin{figure}
+\begin{center}
+<<graphing_ses_growth, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+# Graphing how the top/bottom are evolving over time instead of absolute difference
+
+complete_data_topbottom %>%
+  bind_rows(complete_data_topmid) %>%
+  bind_rows(complete_data_midbottom) %>%
+  filter(country %in% c("United States", "Denmark", "France", "Germany"), type_test == "math") %>%
+  select(wave, country, type_test, type, contains("math")) %>%
+  mutate(type = factor(type,
+                       levels = c("90th/10th SES gap", "90th/50th SES gap", "50th/10th SES gap"),
+                       ordered = TRUE)) %>%
+  gather(score, value, -(wave:type)) %>%
+  separate(score, c("ses", "score"), sep = 2) %>%
+  spread(score, value) %>%
+  ggplot(aes(as.factor(wave), mean_math, group = ses, colour = ses)) +
+  geom_errorbar(aes(ymin = lower_math, ymax = upper_math), width = 0.1) +
+  geom_line(alpha = 0.4) +
+  geom_point(size = 0.5, alpha = 0.4) +
+  coord_cartesian(ylim = c(-1, 2)) +
+  scale_y_continuous(name = "Standardized test scores (mean 0)") +
+  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
+  scale_colour_discrete(name = NULL, labels = c("Low SES", "High SES")) +
+  ggtitle("Evolution of the top/bottom groups by different gaps") +
+  facet_grid(country ~ type) +
+  theme_few() +
+  theme(panel.spacing = unit(1, "lines"),
+        legend.position = "bottom")
+@
+\end{center}
+\end{figure}
+
+With Denmark there's no average change across all years.
+However, for France, we see that the gap is growing much faster for the 50th/10th SES gap.
+For Germany we see that the gap is closing in all three groups although much more strongly for the 50th/10th SES gap.
+It looks like the 50th/10th gap is mainly driving the changes in the gaps for most of the countries that actually see a change over time.
+For the U.S, there's very little change in the 90th/50th SES gap, with most of the change coming from the 50th/10th gap.
+
+Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's an efficiency/equality tradeoff and if it stands over time.
+One particular topic that has been researched in the past is whether improving the overall cognitive development of a country compromises the distribution of the countries overall performance.
+For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance.
+This, however, can come at a price.
+If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance.
+
+<<graphing_achievement_disparity, cache = T>>=
+
+country_scores <-
+  map(adapted_year_data, ~ {
+  if (unique(.x$wave) == "pisa2015") {
+    intsvy::pisa2015.mean.pv("MATH", by = "country", data = .x)
+  } else {
+    intsvy::pisa.mean.pv("MATH", by = "country", data = .x)
   }
-
-top_bottom_perc <- perc_increase_fun(complete_data_topbottom)
-top_mid_perc <- perc_increase_fun(complete_data_topmid)
-mid_bottom_perc <- perc_increase_fun(complete_data_midbottom)
-
-# Gap is closing at an average of the variable diff per year.
-@
-
-<<graphing_perc_increase, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
-perc_graph <- function(df, title) {
-  df %>%
-  enframe(name = "country") %>%
-  unnest(value) %>%
+}) %>% setNames(seq(2000, 2015, 3))
+
+country_scores <-
+  enframe(country_scores, name = "wave") %>%
+  unnest() %>%
+  mutate(wave = as.double(wave))
+
+data_to_plot <-
+  left_join(complete_data_topbottom, country_scores, by = c("wave", "country")) %>%
+  select(wave, country, type_test, difference, Mean) %>%
+  filter(type_test == "math")
+
+data_to_plot %>%
+  ggplot(aes(scale(Mean), difference)) +
+  geom_point() +
+  geom_smooth()
+@
+
+<<correlation_performance_inequality>>=
+correlation <-
+  data_to_plot %>% group_by(wave) %>%
+  summarize(cor = cor(difference, Mean, use = 'complete.obs')) %>%
+  pull(cor)
+@
+
+\begin{figure}
+\begin{center}
+<<avgperformance_inequality, include = T, out.height = '5in', out.width = '8.5in', fig.align = 'center'>>=
+data_to_plot %>%
+  ggplot(aes(scale(Mean), difference)) +
+  geom_smooth(method = "lm") +
+  geom_point() +
+  scale_y_continuous(name = "90/10 achievement gap") +
+  scale_x_continuous(name = "Average performance standardized") +
+  ggtitle("Relationship between average performance and achievement gap") +
+  facet_wrap(~ wave, nrow = 2) +
+  theme_few()
+@
+\end{center}
+\end{figure}
+
+For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}.
+The correlations are between \Sexpr{max(correlation)} and \Sexpr{min(correlation)}.
+This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}.
+However, what they didn't really go in depth was whether this corelation is strongest for some countries over time.
+With the data available we can look at the evolution separately for many countries.
+
+\begin{figure}
+\begin{center}
+<<country_achievement_and_inequality, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+data_to_plot %>%
+  mutate(Mean = scale(Mean),
+         country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
+  gather(type, value, difference, Mean) %>%
   filter(country %in% countries) %>%
-  select(-country1, -years_available) %>%
-  mutate(diff_95_lower = perc_diff - 2*sd_year,
-         diff_95_upper = perc_diff + 2*sd_year) %>%
-  setNames(c("Country", "Type of test", "Average % difference", "Average SD",
-             "Lower 50% bound", "Upper 50% bound", "Lower 95% bound", "Upper 95% bound")) %>%
-  filter(`Type of test` == "math") %>%
-  arrange(`Average % difference`) %>%
-  mutate(Country = ordered(forcats::as_factor(Country)),
-         Country_num = as.numeric(Country)) %>%
-  ggplot(aes(Country, `Average % difference`)) +
-  geom_point() +
-  geom_ribbon(aes(x = Country_num, ymin = `Lower 95% bound`, ymax = `Upper 95% bound`), alpha = 0.5) +
-  geom_ribbon(aes(x = Country_num, ymin = `Lower 50% bound`, ymax = `Upper 50% bound`), alpha = 0.5) +
-  geom_hline(yintercept = 0, linetype = "longdash") +
-  scale_y_continuous(breaks = seq(-160, 160, 40)) +
-  coord_cartesian(ylim = c(-160, 160)) +
-  ggtitle(title) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-}
-
-m1 <- perc_graph(top_bottom_perc, "90/10")
-m2 <- perc_graph(top_mid_perc, "90/50")
-m3 <- perc_graph(mid_bottom_perc, "50/10")
-
-gridExtra::grid.arrange(m1, m2, m3, nrow = 1)
-@
+  ggplot(aes(as.character(wave), value, group = type, colour = type)) +
+  geom_point(alpha = 0.4) +
+  geom_line(alpha = 0.4) +
+  scale_y_continuous(name = "Standardized test scores (mean 0)") +
+  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
+  scale_colour_discrete(name = NULL, labels = c("Achievement gap", "Avg performance")) +
+  ggtitle("Evolution of the top/bottom groups by different gaps") +
+  facet_wrap(~ country, ncol = 5) +
+  theme_few() +
+  theme(panel.spacing = unit(1, "lines"),
+        legend.position = "bottom")
+@
+\end{center}
+\end{figure}
+
+In fact, we see a strong correspondence.
+In countries like France we see quite a functional form between both inequality and performance.
+As average performance plummets, the achievement gap is increasing.
+This pattern is also seen in in the Austria, Australia and Germany.
+The other obvious pattern is the opposite trend.
+We see countries like Poland, Finland, Canada and much less pronounced Italy that this pattern is present.
+The surprising case here is the United States.
+We see that as the average performance has been decreasing, the average achievement gap has been decreasing as well.
+In fact, among the few countries we should find something that resembles this pattern should be the U.S because it doesn't have a strict institutionalized tracking system.
+However, the explanation behind this pattern is very speculative.
+Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap.
+Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality.
 

### 2017-07-05 paper/paper.tex
All graphs and tables visible in tex. Rearranging and recommenting

--- 2017-07-03 paper/paper.tex
@@
 
-{\centering \includegraphics[width=5.5in,height=5in]{figure/correlation_incomeineq-1}
-
+% latex table generated in R 3.4.0 by xtable 1.8-2 package
+% Wed Jul  5 09:41:15 2017
+\scalebox{0.75}{
+\begin{tabular}{llllllll}
+   \hline \
+                          & & & Low SES & & & High SES \\
+                          \cmidrule(l){3-5}
+                          \cmidrule(l){6-8} Year & Countries & \% Low Edu & \% High Edu & Sample size & \% Low Edu & \% High Edu & Sample size \\ \hline2000 & Australia & 5\% & 1\% & 294 & 0\% & 97\% & 270 \\
+  2000 & Austria & 4\% & 0\% & 294 & 0\% & 99\% & 235 \\ 
+  2000 & Canada & 3\% & 1\% & 1970 & 0\% & 99\% & 1357 \\ 
+  2000 & Denmark & 4\% & 0\% & 230 & 0\% & 97\% & 228 \\ 
+  2000 & Finland & 2\% & 0\% & 261 & 0\% & 97\% & 267 \\ 
+  2000 & France & 14\% & 0\% & 253 & 0\% & 97\% & 258 \\ 
+  2000 & Germany & 2\% & 0\% & 250 & 0\% & 93\% & 303 \\ 
+  2000 & Hungary & 1\% & 0\% & 291 & 0\% & 99\% & 254 \\ 
+  2000 & Italy & 3\% & 0\% & 278 & 0\% & 99\% & 278 \\ 
+  2000 & Netherlands & 9\% & 0\% & 129 & 0\% & 99\% & 158 \\ 
+  2000 & Poland & 0\% & 0\% &  & 0\% & 92\% & 166 \\ 
+  2000 & Spain & 15\% & 0\% & 318 & 0\% & 98\% & 340 \\ 
+  2000 & Sweden & 4\% & 1\% & 241 & 0\% & 98\% & 246 \\ 
+  2000 & United Kingdom & 2\% & 0\% & 531 & 0\% & 95\% & 415 \\ 
+  2000 & United States & 5\% & 0\% & 222 & 0\% & 97\% & 164 \\ 
+  2015 & Australia & 2\% & 2\% & 1694 & 0\% & 94\% & 1235 \\ 
+  2015 & Austria & 4\% & 0\% & 676 & 0\% & 96\% & 705 \\ 
+  2015 & Canada & 1\% & 3\% & 2215 & 0\% & 97\% & 1863 \\ 
+  2015 & Denmark & 4\% & 0\% & 1013 & 0\% & 99\% & 595 \\ 
+  2015 & Finland & 2\% & 1\% & 575 & 0\% & 99\% & 584 \\ 
+  2015 & France & 3\% & 0\% & 570 & 0\% & 94\% & 615 \\ 
+  2015 & Germany & 5\% & 0\% & 545 & 0\% & 96\% & 582 \\ 
+  2015 & Hungary & 1\% & 0\% & 466 & 0\% & 99\% & 589 \\ 
+  2015 & Italy & 2\% & 0\% & 947 & 0\% & 96\% & 1043 \\ 
+  2015 & Netherlands & 7\% & 0\% & 521 & 0\% & 83\% & 525 \\ 
+  2015 & Poland & 0\% & 0\% &  & 0\% & 97\% & 448 \\ 
+  2015 & Spain & 10\% & 0\% & 608 & 0\% & 99\% & 703 \\ 
+  2015 & Sweden & 7\% & 2\% & 527 & 0\% & 97\% & 542 \\ 
+  2015 & United Kingdom & 2\% & 1\% & 1387 & 0\% & 93\% & 1388 \\ 
+  2015 & United States & 10\% & 0\% & 585 & 0\% & 97\% & 544 \\ 
+   \hline
+\end{tabular}
 }
@@
 
+\begin{figure}
+\begin{center}
+
 {\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_9010gaps-1}
@@
 
+\end{center}
+\end{figure}
+
 We start by look at the achievement for some countries.
-We plot both mathematics and reading gaps for each country and also a cubic trend spline for both test scores pooled.
+We plot only mathematics for each country and also a quadratic trend spline for both test scores pooled.
 Some countries have increased their achievement very strongly.
@@
 Take the case of Sweden.
-The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation.
+The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015.
 The actual difference is of about 0.6.
@@
 
-Something reassuring is that mathematics and reading follow basically the same trend across all countries.
+The plot below takes a more direct approach and looks at the percentage change from the first and last timepoint available.
+Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.
+
+% How to output data frames are latex tables?
+
+\begin{figure}
+\begin{center}
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_perc_increase-1}
+
+}
+
+\end{center}
+\end{figure}
+
+Generally speaking, we see that most countries increased their achievement gap.
+France had an average increase of about 80\% since 2000, whereas Germany had a similar figure but decreasing.
+Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap.
+In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%.
+
+Something reassuring is that mathematics and reading follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase.
+The previos two graphs for reading are available upon request}.
 This means that the result is not an artifact of chance.
@@
 Up next we plot the same graph but show the divergent patterns between high/low SES origins.
+
+\begin{figure}
+\begin{center}
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/rate_increase/decrease-1}
+
+}
+
+\end{center}
+\end{figure}
 
@@
 
+\begin{figure}
+\begin{center}
+
 {\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_allgaps-1}
@@
 }
+
+\end{center}
+\end{figure}
 
@@
 
+In a more convenient way, we can inspect whether there is a lot of variance in the rate at which top/bottom groups are changing.
+
+% How to output data frames are latex tables?
+
+\begin{figure}
+\begin{center}
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/rate_change_graph-1}
+
+}
+
+\end{center}
+\end{figure}
+
+We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as United States and Germany.
+We see a negative relationship where the faster the low SES catch up the slower the High SES decrease.
+Quite an interesting pattern because we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel).
+On the other hand, we do see more ocuntries where the both groups are increasing (top right panel).
+We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.
+
 But, similar as before, we don't know which group is causing the decrease/increase of the gap in the 90th/50th and 50th/10th gap. For that we plot the evolution of these gaps below.
+
+\begin{figure}
+\begin{center}
 
@@
 
-For all country-years pooled we see a very strong correlation but contrary to what we would expect.
-We find that countries with high performance have also a very low level of inequality.
-In fact, even if we separate it by year.
+\begin{figure}
+\begin{center}
 
-In fact, we see a strong correspondence.
-In countries like Germany and France we see quite a functional form between both inequality and performance.
-This pattern is also seen in in the UK, Austria, Australia and New Zealand.
-Poland and Italy are quite interesting cases because it shows how inequality has been decreasing that much whereas average performance has been improving notably.
-On another hand, only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality.
-
-% How to output data frames are latex tables?
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/rate_change_graph-1}
+{\centering \includegraphics[width=8.5in,height=5in]{figure/avgperformance_inequality-1}
 
@@
 
-In a more convenient way, we can see that there is a lot of variance in the rate at which top/bottom groups are changing.
-We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as United States and Germany.
-We see a negative relationship where the faster the low SES catch up the slower the High SES decrease.
-Quite an interesting pattern because we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel).
-On the other hand, we do see more ocuntries where the both groups are increasing (top right panel).
-We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.
+\end{center}
+\end{figure}
 
-% How to output data frames are latex tables?
+For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}.
+The correlations are between -0.7343267 and -0.8781625.
+This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}.
+However, what they didn't really go in depth was whether this corelation is strongest for some countries over time.
+With the data available we can look at the evolution separately for many countries.
 
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_perc_increase-1}
+\begin{figure}
+\begin{center}
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/country_achievement_and_inequality-1}
 
 }
+
+\end{center}
+\end{figure}
+
+In fact, we see a strong correspondence.
+In countries like France we see quite a functional form between both inequality and performance.
+As average performance plummets, the achievement gap is increasing.
+This pattern is also seen in in the Austria, Australia and Germany.
+The other obvious pattern is the opposite trend.
+We see countries like Poland, Finland, Canada and much less pronounced Italy that this pattern is present.
+The surprising case here is the United States.
+We see that as the average performance has been decreasing, the average achievement gap has been decreasing as well.
+In fact, among the few countries we should find something that resembles this pattern should be the U.S because it doesn't have a strict institutionalized tracking system.
+However, the explanation behind this pattern is very speculative.
+Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap.
+Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality.
 

### 2017-07-05 paper/paper.Rnw
Revised text and about 70% of document

@@
 Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities.
-We want to know if the gap is big or small relative to other countries.
-Furthermore we want to know if this gap has reponded to our policy efforts to reduce it.
+We still don't know, in comparative terms, which countries have big or small cognitive gaps resulting from SES origins.
+And even more importantly, we haven't assesed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.
 
@@
 
-For Denmark, for example, we learned a great deal about how to improve social mobility by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility for less advantaged children \citep{bjorklund2009, jaeger2007}.
-But the important finding came when we discovered the main reason behind their social escalator: its' educational system.
-For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential, specially disadvantaged children.
-Moreover, it tracks students into different curricula very late compared to other Europen countries Europe (age 16).
+From Denmark, for example, the literature has learned a great deal about how to improve social mobility.
+But that was possible by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility specially for less advantaged children \citep{bjorklund2009, jaeger2007}.
+The important finding came when we discovered the main reason behind their social escalator: it's educational system.
+For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential.
+It does so very strongly for children which otherwise would be doomed to have a bleak future.
+Moreover, it tracks students into different curricula very late compared to other Europen countries (age 16), something which has been linked to less educational inequality.
 These two traits make the Danish system uniquely effective.
@@
 The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940.
-This trend has allowed researchers to study achievement gaps in a lengthy period of time.
+This time-trend has allowed researchers to study achievement gaps in a lengthy period of time.
 Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
-Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citet{magnuson2008}.
-These gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
+Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}.
+The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
 
 \citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's.
-In fact, the hard numbers suggest that the gap widened by \~  40-50\%.
-The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of \~  75\%.
-Given that the studies before 1970 are less reliablle in terms of comparability and sampling design, the author computes all results for both before/after 1970.
-To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2010.
+In fact, the hard numbers suggest that the gap widened by 40-50\%.
+The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of   75\%.
+Given that the studies before 1970 are less reliable in terms of comparability and sampling design, the author computes all results for both before/after 1970.
+To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2001.
 Using longitudinal data, \citet{bradbury2015} find similar results.
 Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25.
-Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of \~ 1.25-1.50 standard deviations.
+Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of  1.25-1.50 standard deviations.
 One important drawback of these studies is that they don't present the uncertainty of this estimates.
 Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
-It could be that the gap is at \~ 1.25 standard deviations, varying up to 1.75 and down to 0.80.
+It could be that the gap is at  1.25 standard deviations, varying up to 1.75 and down to 0.80.
 Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.
@@
 A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality.
-However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011, magnuson2008}
+However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008}.
+
 \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
@@
 In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend.
-This follow-up study concentrated solely on the U.S Kindergarten gap in which they took a snapshot of the achievement gap in Kindergarten in 1998, 2006 and 2010.
+This follow-up study concentrated solely on the kindergarten children in which they took a snapshot of the achievement gap in for 1998, 2006 and 2010.
 They found that the 90th/10th income gap in readiness closed modestly.
@@
 Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap).
-
-The American social landscape is very distinctive relative to Europe.
-The reasons the authors find a reversal in the trend could be numerous and should be studied very close.
-For example, the American educational system lacks any formal curricular differentiation \emph{\'a la} European style.
-It is true nonetheless that \citet{reardon2011} did not cover from the period 2000 onwards, which is specifically the trend they studied so we have no other comparison.
-\citet{reardon_portilla} suggests that the reversal is likely due to the high increase of preschool enrollment.
+It should be noted that perhaps the reversal of trend in \citet{reardon2011} would be obvious if data were available for years after 2000,  which is specifically the time that \citet{reardon_portilla} find the reversal.
+
+The reasons the authors find a reversal in the trend could be numerous and should be studied very closely.
+The authors incorporate a number of country-level indicators to try and explaind this change and suggest that the reversal is likely due to the high increase of preschool enrollment.
 They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
-This explanation still needs further confirmation.
+Their conclusions are very speculative and have no empirical support, which is why this is still an open question.
 
@@
 The core finding behind their study is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada.
-They find that once the achievement gap is present in early school entry, it doesn't seem to narrow or neither widen very much over the life course.
-In fact, they calculate that the schooling environment once they enter school can only explain about \~ 30-40\% of the high school SES gap.
-This suggests that once the achievement gap widens before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous RCT's show that high quality schooling can indeed help ease the gap, in some instances even close it \citep{campbell2002}.}.
-One exception is the United Kingdom, which they found to help close the gap in early primary years.
-This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health and welfare support.
+They find that once the achievement gap is present in early school entry, it doesn't seem to narrow or widen very much over the life course.
+In fact, they calculate that the quality of early education can only explain about  30-40\% of the high school SES gap.
+This suggests that once the achievement gap is present before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous RCT's show that high quality schooling can indeed help ease the gap, in some instances even close it \citep{campbell2002}.}.
+One exception is the United Kingdom, which they found to be a country that helps close the gap in early primary years.
+This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health and income support.
 
@@
 While less clear, there is also a tendency for early tracking to reduce mean performance.
-Moreover, \citet{micklewright} using similar data but a different empirical strategy, finds that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
+\citet{micklewright} using PISA but a different empirical strategy, find that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
 In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
@@
 Another limitation of their study is that their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
-They do an amazing work at makgin the 4 surveys comparable and their findings are very reliable but should be taken as suggestive at best.
+They do an amazing work at making the four surveys comparable and their findings are indeed very reliable.
+But we should be careful at intepreting these findings causally and should be taken as suggestive at best.
 For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citet{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in the gap using much more comparable surveys.
@@
 Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
-In comparison to the literature on achievement gaps, they find that the U.S has a gap of \~  1.20 SD in 2001 which increase to around ~ 1.30 in the year 2006 while Germany has a decreasing gap from around \~  1.25 to \~  1 standard deviation in the same year-span.
+In comparison to the literature on achievement gaps, they find that the U.S has a gap of   1.20 SD in 2001 which increase to around ~ 1.30 in the year 2006 while Germany has a decreasing gap from around   1.25 to   1 standard deviation in the same year-span.
 However, these numbers vary a lot and carry a great deal of overlapping uncertainty.
@@
 They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation and central exams, among other things.
-The correlations are very suggestive  as explanatory mechanisms but they are very cautios in drawing causality, specially because they are all reasonably correlated among each other.
-
-One interesting question is still missing is whether these country gaps have evolved over time.
+The correlations are very suggestive  as explanatory mechanisms but they are very cautious in drawing causality, specially because they are all reasonably correlated among each other.
+
+One interesting question that is still missing from the literature is whether these country gaps have evolved over time.
 With their data, they only have 3 countries which are present in all waves but they also have very few waves because their question of interest (income categories) was only asked in three time points.
@@
 However, once she disentangles the relationship by country, she finds a sizable amount of heterogeneity, with some countries seeing the achievement gap narrowing, others no change at all, while others record a steady increase.
-This is revealing because it doesn't really pay off to look at general average once each country has their own distinctive gap and evolution.
+This is revealing because it doesn't really pay off to look at a general average once each country has their own distinctive gap and evolution.
 Another important finding is that the achievement gap is clearly no universal and should be studied in context.
 
-One clear limitation of their study (as well as \citet{reardon2011}) is that the adjust for the age of each child in all studies.
-Although for their modeling purposes this is the right thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
-However, as we've noted before, \citet{bradbury2015} find that the achievement gap is very stable across the life time}, they are masking age-specific achievement gaps by controlling for age, such as \citet{reardon2011} did.
-We clearly see in \citet{reardon_portilla} that paying attention to age-specific gaps varies reasonably even in short periods of time.
+One clear limitation of their study (as well as \citet{reardon2011}) is that they adjust for the age of each child in all studies.
+Although for their modeling purposes this is the correct thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
+However, as we've noted before, \citet{bradbury2015} find that the achievement gap is very stable across the life time}, these modeling strategies are masking age-specific achievement gaps by controlling for age.
+We clearly see in \citet{reardon_portilla} that there are age-specific gaps, and they do change at a fast pace in very little time.
 
@@
 The explanation, although very debated, has been gaining much support in recent years.
-In countries with high levels of curricular differentiation, the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}.
+If we remember correctly, countries with high levels of curricular differentiation, the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}.
 Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}.
@@
 This is different from previous work because it concentrates solely on 15 year old children, something not done for the achievement gap, and it attempts to capture the evolution of the achievement gap for countries with different educational systems.
-Furthermore, we study the relationship between the gap and the average performance of each country in order to detech if there's a tradeoff between efficiency and equality.
+In addition, extend the previous findings by studying the relationship between the gap and the average performance of each country in order to detect if there's a tradeoff between efficiency and equality over time separately for each country.
 The advantages of this study are twofold.
 
-First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the Kindergarten gap in \citet{reardon_portilla}.
+First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the kindergarten gap in \citet{reardon_portilla}.
 As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
@@
 This index variable was developed by \citet{ganzeboom1996, ganzeboom2010} and it attempts to measure occupational status using a continuous measure.
-This indicator has been found to ourperform classical occupational classifications such as the Erikson-Goldthorpe-Portocarero classification \citet{erikson1979}.
+This indicator is a very reliable alternative to the  classical Erikson-Goldthorpe-Portocarero classification \citet{erikson1979}.
 It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}.
@@
 As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured (the true distribution of the possible scores a student can achieve).
-Suppose we have \(\mu_i\), the average student test score in Mathematics for student \(i\).
+
+A more intuitivity is explanation is this: suppose we have \(\mu_i\), the average student test score in Mathematics for student \(i\).
 Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\) for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test.
@@
 The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, as others have used, for example, the 5th, 50th and 95th \citep{micklewright}.
-We use the conventional cutoffs as the literature has been following the standard set by \citet{reardon2011} to use 90th/10th percentiles.
+We use the conventional 90/10 cutoffs in the literature following the standard set by \citet{reardon2011}.
 
@@
   enframer("name") %>%
-  ggplot(aes(name, value)) + geom_col()
+  ggplot(aes(name, value)) +
+  geom_col() +
+  scale_y_continuous(name = "Standard deviation of raw test scores") +
+  scale_x_discrete(name = NULL) +
+  ggtitle("Standard deviation of test scores over time") +
+  theme_few()
 @
@@
 
-\section{Analysis}
+\section{Analysis and results}
 
@@
 
-\subsection{Results}
-
-Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
+\subsection{Evolution of the achievement gap}
+% Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
 
@@
 @
+\caption{Evolution of the achievement gap for selected countries}
 \end{center}
@@
 
-We start by look at the achievement for some countries.
-We plot only mathematics for each country and also a quadratic trend spline for both test scores pooled.
+We start by look at the achievement for some countries in figure 1.
+We plot only mathematics for each country and also a quadratic trend spline for \emph{both mathematics and literacy} pooled.
 Some countries have increased their achievement very strongly.
-For example, France, Austria and surprisingly Sweden.
-In mathematics, France increase the gap by roughly \Sexpr{diff_data$math[["France"]]}.
-This pattern happens similarly for literacy with an increase of \Sexpr{diff_data$read[["France"]]}.
-For such a short period of time, the magnitud of the increase is reasonably big.
+For example, France, Austria and surprisingly Sweden have very steep slopes.
+France increasde the gap by roughly \Sexpr{diff_data$math[["France"]]} standard deviations, Austria by \Sexpr{diff_data$math[["Austria"]]} and Sweden by \Sexpr{diff_data$math[["Sweden"]]}.
+This pattern happens similarly for literacy.
+For example, France has an increase of \Sexpr{diff_data$read[["France"]]}.
+For such a short period of time, the magnitud of these increases are reasonably big.
 
 Given that no one has estimated the evolution of the gap we can't cross-check how other empirical estimations put France at.
-However, we can make some educated inferences.
-For example, \citet{micklewright} also finds that France is a low dispersion country, specifically in the year 2000.
-
-However, we don't know what happens over time. But \citet{bernardi2016} did study the relationship of social origin effects and found that they increased very much in France in the last decades.
-
-Other countries have reasonable increases such as Austria and Finland.
-Aside from these countries, some other countries have no change at all.
-Specially, Canada, Netherlands and Spain.
+However, we can take \citet{micklewright} as the closest reference which also finds that France was a low dispersion country in 2000.
+However, there's no evidence on what happened over time.
+Luckily, the work of \citet{bernardi2016} did study the relationship of social origin effects (broadly speaking, not in terms of achievement gaps) in France and found that they increased in the last decades.
+
+Other countries have reasonable increases such as Finland and Hungary, with increases of nearly \Sexpr{diff_data$math[["Finland"]]} and \Sexpr{diff_data$math[["Hungary"]]}, respectively.
+Aside from these countries, there are other cases that have no change at all, specifically, Canada, Netherlands and Spain.
 Canada excels here not only because the gap has been stable over time, but because it has the smallest gap of all countries presented here.
@@
 
-Furthermore, we do have some countries which show a decrease in the SES achievement gap.
-United Kingdom, Israel, Poland, United States and Germany show some steep downward slopes.
-Most notably are the U.S and German cases.
+On the other hand, we do have some countries which show a decrease in the SES achievement gap.
+United Kingdom decreased by about \Sexpr{diff_data$math[["United Kingdom"]]}, Poland by \Sexpr{diff_data$math[["Poland"]]}, and Denmark by \Sexpr{diff_data$math[["Denmark"]]}.
+However, the most notable cases are the United States and Germany.
 These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations.
-However, both countries reduced the gaps by \Sexpr{diff_data$math[["United States"]]} and \Sexpr{diff_data$math[["Germany"]]} respectively.
+But in the 15-year time trend both countries reduced the gaps by \Sexpr{diff_data$math[["United States"]]} and \Sexpr{diff_data$math[["Germany"]]} respectively.
 Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}.
+This corroborates the findings of \citet{reardon_portilla}, which found a decreasing gap for kindergarteners.
+This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy.
 
@@
 
-The plot below takes a more direct approach and looks at the percentage change from the first and last timepoint available.
+<<us_sd_start>>=
+us_math_2000 <-
+  filter(complete_data_topbottom, wave == 2000, country == "United States", type_test == "math") %>%
+  .[["difference"]]
+
+us_math_2000_lower <-
+  filter(complete_data_topbottom, wave == 2000, country == "United States", type_test == "math") %>%
+  .[["lower"]]
+
+us_math_2000_upper <-
+  filter(complete_data_topbottom, wave == 2000, country == "United States", type_test == "math") %>%
+  .[["upper"]]
+@
+We find that the initial gap for the U.S in 2000 is \Sexpr{us_math_2000} varying between \Sexpr{us_math_2000_lower} and \Sexpr{us_math_2000_upper}.
+This gap is much higher than found in previous studies, but again, we don't know much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely.
+All country-level studies pool many studies together and control for age, removing the age-specific gap effect.
+
+Figure 2 below takes a more direct approach and looks at the percentage change from the first and last timepoint available.
 Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.
@@
 @
+\caption{Percentage change in the achievement gap from 2000 to 2015}
 \end{center}
@@
 
-Generally speaking, we see that most countries increased their achievement gap.
-France had an average increase of about 80\% since 2000, whereas Germany had a similar figure but decreasing.
+Generally speaking, we see that most countries increased their achievement gap over time.
+France had an average increase of about 80\% since 2000 varying down to 40\%, whereas Germany had a similar figure but decreasing.
 Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap.
 In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%.
+The benefit of presenting these estimates this way is that we can actually asses the uncertainty of each calculation, and we do find that some of these estimates have wide variability.
+Despite this, most countries show a clear sign of either decreasing or increasing.
 
@@
 
-Something reassuring is that mathematics and reading follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase.
+Something reassuring is that mathematics and reading (not presented) follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase.
 The previos two graphs for reading are available upon request}.
@@
 Up next we plot the same graph but show the divergent patterns between high/low SES origins.
+
+\subsection{Source of the achievement gap}
 
@@
   # gather(upper_bound, upper, matches("\\d_upper"))
-  ggplot(aes(as.factor(wave), gap_size, group = ses, colour = ses)) +
-  geom_point(size = 0.5, alpha = 0.4) +
+  ggplot(aes(as.factor(wave), gap_size, group = ses, colour = ses, shape = ses)) +
+  geom_point(size = 2, alpha = 0.4) +
   geom_line(alpha = 0.4) +
@@
   scale_colour_discrete(name = NULL, labels = c("Low SES", "High SES")) +
+  scale_shape_discrete(name = NULL, labels = c("Low SES", "High SES")) +
   ggtitle("Evolution of the achievement gap by top/bottom groups") +
@@
 @
+\caption{Evolution of the achievement gap for separate SES groups}
 \end{center}
@@
 
-The top line belongs to the high SES group whereas the bottom line belongs to the low SES group.
+Figure 3 shows the evolution for both Low and High SES groups separetly.
 The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other.
@@
 On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate.
-there's a group of distinctive groups where both low/high SES rates are very great resulting in the summary line being flat.
-
-Ths source of this gap can be deconstructed into two broader gaps.
+These results highlight the importance of not only summarizing average achievement gaps.
+The source of these gaps vary greatly between countries and we see how each of these patterns contibutes to the overall inequality of a country.
+
+Ths source of this gap can be deconstructed further into two broader gaps.
 We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group.
-Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap.
 For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th?
 Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier?
+Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap.
 Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous graphs \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.
@@
   bind_rows(complete_data_midbottom) %>%
-  filter(country %in% c("United States", "Denmark", "France", "Germany")) %>%
+  filter(country %in% c("United States", "Denmark", "France", "Germany"),
+         type_test == "math") %>%
   mutate(type = factor(type,
@@
                        ordered = TRUE)) %>%
-  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test)) +
+  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test, shape = type_test)) +
   geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
@@
   facet_grid(country ~ type) +
-  theme_few() +
+  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
+  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
+  ggthemes::theme_few() +
   theme(panel.spacing = unit(1, "lines"),
-        legend.position = "bottom")
+        panel.grid.major.y = element_line(colour = "grey"))
 @
@@
 \begin{center}
-<<avgperformance_inequality, include = T, out.height = '5in', out.width = '8.5in', fig.align = 'center'>>=
+<<avgperformance_inequality, include = T, out.height = '5in', out.width = '7in', fig.align = 'center'>>=
 data_to_plot %>%

### 2017-07-05 paper/paper.Rnw
Fixed all graphs but table has wrong edu levels. Next gonne deleted and only add avg performance

@@
 The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
-This means that inequality starts strong before the schooling of the children being and schools can do so much to close it.
-Moreover, these inequalities can be narrowed significantly if the child coming from unfavourable conditions is exposed to high quality learning environments.
+This means that inequality starts strong before the schooling of a child begins and schools can do so much to close it.
+These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions.
 But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\).
-Although this sounds logical, it's implications are not assesed properly.
+Although this sounds logical, it's importance is not understood entirely.
 
@@
 
-It is incomplete to tusy educational inequalities without considering the previlance of intergenerational transmission of human and social capital.
+It is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital.
 Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity.
@@
 This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
-We've become so intertwined with the subject that we've virtually spent most of our time studying the mechanisms through which this inequality comes about.
+Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about.
 Naturally, we want to do that in order to reverse it and help every child reach its fullest potential.
@@
 
-A plotting ot the standard deviation of the data suggests that it's something we shouldn't be deeply concerned with.
+We plot the standard deviation of the mathematics test score for all waves in figure 1.
+The plot suggests that it's something we shouldn't be deeply concerned with.
 The standard deviation of each wave seems to be following a very similar pattern with a not so drastic exception of the year 2000.
@@
 \subsection{Descriptives}
-
-% Show the properties of the ESCS dummy variable
-% % in high educated
-% % of home possessions
-% % of ISEI
 
@@
 
-% Figure out how to ouput data frames as latex tables.
 <<sample_size>>=
@@
 
-We start by look at the achievement for some countries in figure 1.
+We start by look at the achievement for some countries in figure 2.
 We plot only mathematics for each country and also a quadratic trend spline for \emph{both mathematics and literacy} pooled.
@@
 
-Figure 2 below takes a more direct approach and looks at the percentage change from the first and last timepoint available.
+Figure 3 below takes a more direct approach and looks at the percentage change from the first and last timepoint available.
 Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.
@@
 
-Figure 3 shows the evolution for both Low and High SES groups separetly.
+Figure 4 shows the evolution for both Low and High SES groups separetly.
 The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other.
@@
 Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier?
-Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap.
+Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap in figure 5.
 Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous graphs \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.
-
-% This plot might not be necessary any more
-% <<>>=
-% 
-% lower <- seq(13, 1, -1)
-% upper <- seq(100, 220, 10)
-% 
-% df <- tibble(country = "fake_country",
-%              years = rep(seq(2000, 2025, 2), 2),
-%              ses = rep(c("lower", "upper"), each = 13),
-%              value = c(lower, upper))
-% 
-% # You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
-% df %>%
-%   ggplot(aes(as.factor(years), value, group = ses, colour = ses)) +
-%   geom_point(size = 0.5, alpha = 0.4) +
-%   geom_line(alpha = 0.4) +
-%   geom_line(stat = "smooth", method = "lm", aes(group = 1),
-%             formula = y ~ splines::ns(x, 3), linetype = "longdash",
-%             colour = "black") +
-%   # geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
-%   # geom_hline(yintercept = 0, linetype = "longdash") +
-%   # coord_cartesian(ylim = c(0, 3)) +
-%   facet_wrap(~ country)
-% @
 
@@
                        ordered = TRUE)) %>%
-  ggplot(aes(as.factor(wave), difference, group = type_test, colour = type_test, shape = type_test)) +
+  ggplot(aes(as.factor(wave), difference, group = type_test)) +
   geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
-  geom_line(alpha = 0.4) +
+  geom_linerange(aes(ymin = 0, ymax = difference)) +
+  geom_line(stat = "smooth", method = "lm", aes(group = 1),
+            formula = y ~ splines::ns(x, 2), size = 0.7,
+            colour = "red") +
   geom_point(size = 0.5, alpha = 0.4) +
+  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
+  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
   coord_cartesian(ylim = c(0, 3)) +
-  scale_y_continuous(name = "Standardized test scores (mean 0)") +
-  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
-  scale_colour_discrete(name = NULL, labels = c("Low SES", "High SES")) +
   ggtitle("Comparison of the achievement gap for three socio-economic groups") +
   facet_grid(country ~ type) +
-  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
-  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
   ggthemes::theme_few() +
-  theme(panel.spacing = unit(1, "lines"),
-        panel.grid.major.y = element_line(colour = "grey"))
-@
+  theme(panel.spacing = unit(1, "lines"))
+@
+\caption{Evolution of 90/50/10 SES gaps for selected countries}
 \end{center}
@@
 We start off with Denmark.
-Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger for both tests.
-Furthermore, we find, just as in the 90th/10th SES gap, that there isn't a significant decrease, with the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation.
+Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger in mathematics.
+Furthermore, we find, just as in the 90th/10th SES gap, that there isn't a significant decrease \footnote{With the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation.}.
 France shows a similar pattern as Denmark in that the 50th/10th SES gap is the main contributer and that the three gaps have the similar trend.
@@
 For example, Germany, shows that the 90th/10th gap is primarily driven by the 50th/10th SES gap and that the slopes of the 90th/50th gap is much weaker than the other two gaps.
-This means that Top/Middle SES groups have shrunk much less than the other groups.
-The big catching up is actually coming from the Middle/Bottom SES groups.
+This means that top/middle SES groups have shrunk much less than the other groups.
+The big catching up is actually coming from the middle/bottom SES groups.
 However, we don't know if this is because the bottom is catching up or that the middle is coming down.
-However, based on the gap plots from above, we can infer that it is because the bottom is catching up rapidly because against the 90th SES group, it is increasing steeply every year.
+Based on the gap plots from above, we can infer that it is because the bottom is catching up rapidly because against the 90th SES group, it is increasing steeply every year.
+
 The U.S shows a different pattern from all other countries.
@@
 Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group.
-
-In a more convenient way, we can inspect whether there is a lot of variance in the rate at which top/bottom groups are changing.
+The big takeaway from the U.S is that the decrease of inequality is coming from the shrinking of the 50/10 gap.
+
+In a more convenient way, we can inspect whether there is a lot of variance in the rate at which top/bottom groups are changing in figure 6.
 
@@
 @
+\caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
 \end{center}
@@
 
-We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as United States and Germany.
+We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany.
 We see a negative relationship where the faster the low SES catch up the slower the High SES decrease.
-Quite an interesting pattern because we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel).
+This is very interesting because it suggests that there's a trend for countries to have the bottom SES groups are climbing upwards, a trait of meritocracy.
+Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel).
 On the other hand, we do see more ocuntries where the both groups are increasing (top right panel).
@@
 
-But, similar as before, we don't know which group is causing the decrease/increase of the gap in the 90th/50th and 50th/10th gap. For that we plot the evolution of these gaps below.
+But, similar as before, we don't know which group is causing the decrease/increase of the gap in the 90th/50th and 50th/10th gap. For that we plot the evolution of these gaps in figure 7.
 
@@
 If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance.
+We plot this relationship in figure 8.
 
@@
 \begin{center}
-<<avgperformance_inequality, include = T, out.height = '5in', out.width = '7in', fig.align = 'center'>>=
+<<avgperformance_inequality, include = T, out.height = '5in', out.width = '6.5in', fig.align = 'center'>>=
 data_to_plot %>%
@@
 @
+\caption{Relationship between 90/10 achievement gap and average performance in mathematics}
 \end{center}
@@
 With the data available we can look at the evolution separately for many countries.
+We plot it up next in figure 9.
 
@@
   filter(country %in% countries) %>%
-  ggplot(aes(as.character(wave), value, group = type, colour = type)) +
+  ggplot(aes(as.character(wave), value, group = type, colour = type, shape = type)) +
   geom_point(alpha = 0.4) +
@@
   scale_colour_discrete(name = NULL, labels = c("Achievement gap", "Avg performance")) +
+  scale_shape_discrete(name = NULL, labels = c("Achievement gap", "Avg performance")) +
   ggtitle("Evolution of the top/bottom groups by different gaps") +
@@
 @
+\caption{Time trends of achievement gap and average achiement between 2000 and 2015}
 \end{center}
@@
 
-In fact, we see a strong correspondence.
+Looking at figure 9, we see a strong correspondence between the two variables.
 In countries like France we see quite a functional form between both inequality and performance.
 As average performance plummets, the achievement gap is increasing.
-This pattern is also seen in in the Austria, Australia and Germany.
+This pattern is also seen in Austria, Australia and Germany.
 The other obvious pattern is the opposite trend.
-We see countries like Poland, Finland, Canada and much less pronounced Italy that this pattern is present.
+We see countries like Poland, Finland, Canada and less pronounced, Italy, where this pattern is evident.
 The surprising case here is the United States.
@@
 However, the explanation behind this pattern is very speculative.
+
 Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap.
-Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality.
+Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentially, are countries that have the lower leven of inequalities among the ones presented here.
 

### 2017-07-05 paper/paper.Rnw
Fixed table and graphs. Everything ready.

@@
 
-\title{First paper - Draft}
+\title{The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015}
 \author{Jorge Cimentada}
@@
 
+% <<table_escs, include = TRUE>>=
+% summary_table <-
+%   left_join(summary_data, sample_tables_topbottom) %>%
+%   filter(name %in% c("2000", "2015")) %>%
+%   arrange(country, name)
+% 
+% table_coming <-
+%   summary_table %>%
+%   select(-perc) %>%
+%   gather(cats, vals, -(name:escs_dummy)) %>%
+%   unite(all_vals, escs_dummy, cats, sep = "_") %>%
+%   spread(all_vals, vals) %>%
+%   map_at(c(3, 4, 6, 7), ~ {
+%     .x[is.na(.x)] <- 0
+%     paste0(round(.x, 0), "%")
+%   }) %>%
+%   map_if(is_double, as.character) %>%
+%   data.frame() %>%
+%   xtable::xtable()
+% 
+% addtorow <- list()
+% addtorow$pos <- list(0)
+% addtorow$command <- paste("\\hline \\
+%                           & & & Low SES & & & High SES \\\\
+%                           \\cmidrule(l){3-5}
+%                           \\cmidrule(l){6-8}",
+%                           paste0(
+%                             c("Year",
+%                               "Countries",
+%                               "\\% Low Edu",
+%                               "\\% High Edu",
+%                               "Sample size",
+%                               "\\% Low Edu",
+%                               "\\% High Edu",
+%                               "Sample size"),
+%                             collapse = " & "),
+%                           "\\\\ \\hline")
+% 
+% print(table_coming,
+%       scalebox = '0.75',
+%       floating = FALSE,
+%       add.to.row = addtorow,
+%       include.colnames = FALSE,
+%       include.rownames = FALSE,
+%       hline.after = nrow(table_coming),
+%       caption.placement = "top",
+%       caption = "SES sample size and ISCED composition")
+% @
+
 <<table_escs, include = TRUE>>=
+
+avg_performance <-
+  list(results_math[[1]], results_math[[6]]) %>%
+  set_names(c(2000, 2015)) %>%
+  enframe() %>%
+  unnest()
+
 summary_table <-
-  left_join(summary_data, sample_tables_topbottom) %>%
+  sample_tables_topbottom %>%
+  mutate(escs_dummy = as.character(escs_dummy)) %>%
   filter(name %in% c("2000", "2015")) %>%
+  left_join(avg_performance) %>%
   arrange(country, name)
@@
   spread(all_vals, vals) %>%
-  map_at(c(3, 4, 6, 7), ~ {
-    .x[is.na(.x)] <- 0
-    paste0(round(.x, 0), "%")
-  }) %>%
-  map_if(is_double, as.character) %>%
-  data.frame() %>%
+  select(name, country, `0_n`, `0_Mean`, `0_s.e.`, `1_n`, `1_Mean`, `1_s.e.`) %>%
+  mutate(`0_n` = as.character(`0_n`),
+         `1_n` = as.character(`1_n`)) %>%
   xtable::xtable()
@@
                               "Countries",
-                              "\\% Low Edu",
-                              "\\% High Edu",
-                              "Sample size",
-                              "\\% Low Edu",
-                              "\\% High Edu",
-                              "Sample size"),
+                              "N",
+                              "Avg score",
+                              "S.E",
+                              "N",
+                              "Avg score",
+                              "S.E"),
                             collapse = " & "),
@@
 @
+
+The first table shows a description of the sample size and mean score of both top and bottom SES groups for only the first and last time point.
+One main concern from the planned analysis is that getting the top 90th percentile and bottom 10th percentile would result in a small sample size.
+The table suggests that we have a reasonable number of respondents to actually estimate gaps accurately.
+Moreover, we can see that in all instances the bottom SES group has a lower score than all top SES groups.
+
+We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting their very equal countries.
+However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average, which is 0.
+We also see some countries with major changes from year 2000 to 2015, with, for example, Australia decreasing the low SES gap from 0.22 standard deviations to -0.15, well below the country average.
+In the next section we take a look at this in a more detailed fashion, considering all years in between.
 
@@
 @
-We find that the initial gap for the U.S in 2000 is \Sexpr{us_math_2000} varying between \Sexpr{us_math_2000_lower} and \Sexpr{us_math_2000_upper}.
+We find that the initial gap for the U.S in 2000 is \Sexpr{round(us_math_2000, 2)} varying between \Sexpr{round(us_math_2000_lower, 2)} and \Sexpr{round(us_math_2000_upper, 2)}.
 This gap is much higher than found in previous studies, but again, we don't know much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely.
@@
            fontface = 2, size = 3) +
-  annotate(geom = "text", x = -0.05, y = 0.20,
+  annotate(geom = "text", x = -0.075, y = 0.20,
            label = "High SES is increasing  \n faster than Low SES is catching up",
@@
 
+\subsection{Efficiency-equality tradeoff}
 <<graphing_achievement_disparity, cache = T>>=
@@
 For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}.
-The correlations are between \Sexpr{max(correlation)} and \Sexpr{min(correlation)}.
+The correlations are between \Sexpr{round(max(correlation), 2)} and \Sexpr{round(min(correlation), 2)}.
 This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}.

### 2017-07-05 paper/paper.tex
Fixed table and graphs. Everything ready.

@@
 This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
-We've become so intertwined with the subject that we've virtually spent most of our time studying the mechanisms through which this inequality comes about.
+Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about.
 Naturally, we want to do that in order to reverse it and help every child reach its fullest potential.
 Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities.
-We want to know if the gap is big or small relative to other countries.
-Furthermore we want to know if this gap has reponded to our policy efforts to reduce it.
+We still don't know, in comparative terms, which countries have big or small cognitive gaps resulting from SES origins.
+And even more importantly, we haven't assesed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.
 
@@
 Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap).
+It should be noted that perhaps the reversal of trend in \citet{reardon2011} would be obvious if data were available for years after 2000,  which is specifically the time that \citet{reardon_portilla} find the reversal.
 
-The American social landscape is very distinctive relative to Europe.
-The reasons the authors find a reversal in the trend could be numerous and should be studied very close.
-For example, the American educational system lacks any formal curricular differentiation \emph{\'a la} European style.
-It is true nonetheless that \citet{reardon2011} did not cover from the period 2000 onwards, which is specifically the trend they studied so we have no other comparison.
-\citet{reardon_portilla} suggests that the reversal is likely due to the high increase of preschool enrollment.
+The reasons the authors find a reversal in the trend could be numerous and should be studied very closely.
+The authors incorporate a number of country-level indicators to try and explaind this change and suggest that the reversal is likely due to the high increase of preschool enrollment.
 They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
-This explanation still needs further confirmation.
+Their conclusions are very speculative and have no empirical support, which is why this is still an open question.
 
@@
 They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation and central exams, among other things.
-The correlations are very suggestive  as explanatory mechanisms but they are very cautios in drawing causality, specially because they are all reasonably correlated among each other.
+The correlations are very suggestive  as explanatory mechanisms but they are very cautious in drawing causality, specially because they are all reasonably correlated among each other.
 
-One interesting question is still missing is whether these country gaps have evolved over time.
+One interesting question that is still missing from the literature is whether these country gaps have evolved over time.
 With their data, they only have 3 countries which are present in all waves but they also have very few waves because their question of interest (income categories) was only asked in three time points.
@@
 
-% Show the properties of the ESCS dummy variable
-% % in high educated
-% % of home possessions
-% % of ISEI
-
-% Figure out how to ouput data frames as latex tables.
+% <<table_escs, include = TRUE>>=
+% summary_table <-
+%   left_join(summary_data, sample_tables_topbottom) %>%
+%   filter(name %in% c("2000", "2015")) %>%
+%   arrange(country, name)
+% 
+% table_coming <-
+%   summary_table %>%
+%   select(-perc) %>%
+%   gather(cats, vals, -(name:escs_dummy)) %>%
+%   unite(all_vals, escs_dummy, cats, sep = "_") %>%
+%   spread(all_vals, vals) %>%
+%   map_at(c(3, 4, 6, 7), ~ {
+%     .x[is.na(.x)] <- 0
+%     paste0(round(.x, 0), "%")
+%   }) %>%
+%   map_if(is_double, as.character) %>%
+%   data.frame() %>%
+%   xtable::xtable()
+% 
+% addtorow <- list()
+% addtorow$pos <- list(0)
+% addtorow$command <- paste("\\hline \\
+%                           & & & Low SES & & & High SES \\\\
+%                           \\cmidrule(l){3-5}
+%                           \\cmidrule(l){6-8}",
+%                           paste0(
+%                             c("Year",
+%                               "Countries",
+%                               "\\% Low Edu",
+%                               "\\% High Edu",
+%                               "Sample size",
+%                               "\\% Low Edu",
+%                               "\\% High Edu",
+%                               "Sample size"),
+%                             collapse = " & "),
+%                           "\\\\ \\hline")
+% 
+% print(table_coming,
+%       scalebox = '0.75',
+%       floating = FALSE,
+%       add.to.row = addtorow,
+%       include.colnames = FALSE,
+%       include.rownames = FALSE,
+%       hline.after = nrow(table_coming),
+%       caption.placement = "top",
+%       caption = "SES sample size and ISCED composition")
+% @
 
 % latex table generated in R 3.4.0 by xtable 1.8-2 package
-% Wed Jul  5 09:41:15 2017
+% Wed Jul  5 17:51:02 2017
 \scalebox{0.75}{
-\begin{tabular}{llllllll}
+\begin{tabular}{lllrrlrr}
    \hline \
@@
                           \cmidrule(l){3-5}
-                          \cmidrule(l){6-8} Year & Countries & \% Low Edu & \% High Edu & Sample size & \% Low Edu & \% High Edu & Sample size \\ \hline2000 & Australia & 5\% & 1\% & 294 & 0\% & 97\% & 270 \\
-  2000 & Austria & 4\% & 0\% & 294 & 0\% & 99\% & 235 \\ 
-  2000 & Canada & 3\% & 1\% & 1970 & 0\% & 99\% & 1357 \\ 
-  2000 & Denmark & 4\% & 0\% & 230 & 0\% & 97\% & 228 \\ 
-  2000 & Finland & 2\% & 0\% & 261 & 0\% & 97\% & 267 \\ 
-  2000 & France & 14\% & 0\% & 253 & 0\% & 97\% & 258 \\ 
-  2000 & Germany & 2\% & 0\% & 250 & 0\% & 93\% & 303 \\ 
-  2000 & Hungary & 1\% & 0\% & 291 & 0\% & 99\% & 254 \\ 
-  2000 & Italy & 3\% & 0\% & 278 & 0\% & 99\% & 278 \\ 
-  2000 & Netherlands & 9\% & 0\% & 129 & 0\% & 99\% & 158 \\ 
-  2000 & Poland & 0\% & 0\% &  & 0\% & 92\% & 166 \\ 
-  2000 & Spain & 15\% & 0\% & 318 & 0\% & 98\% & 340 \\ 
-  2000 & Sweden & 4\% & 1\% & 241 & 0\% & 98\% & 246 \\ 
-  2000 & United Kingdom & 2\% & 0\% & 531 & 0\% & 95\% & 415 \\ 
-  2000 & United States & 5\% & 0\% & 222 & 0\% & 97\% & 164 \\ 
-  2015 & Australia & 2\% & 2\% & 1694 & 0\% & 94\% & 1235 \\ 
-  2015 & Austria & 4\% & 0\% & 676 & 0\% & 96\% & 705 \\ 
-  2015 & Canada & 1\% & 3\% & 2215 & 0\% & 97\% & 1863 \\ 
-  2015 & Denmark & 4\% & 0\% & 1013 & 0\% & 99\% & 595 \\ 
-  2015 & Finland & 2\% & 1\% & 575 & 0\% & 99\% & 584 \\ 
-  2015 & France & 3\% & 0\% & 570 & 0\% & 94\% & 615 \\ 
-  2015 & Germany & 5\% & 0\% & 545 & 0\% & 96\% & 582 \\ 
-  2015 & Hungary & 1\% & 0\% & 466 & 0\% & 99\% & 589 \\ 
-  2015 & Italy & 2\% & 0\% & 947 & 0\% & 96\% & 1043 \\ 
-  2015 & Netherlands & 7\% & 0\% & 521 & 0\% & 83\% & 525 \\ 
-  2015 & Poland & 0\% & 0\% &  & 0\% & 97\% & 448 \\ 
-  2015 & Spain & 10\% & 0\% & 608 & 0\% & 99\% & 703 \\ 
-  2015 & Sweden & 7\% & 2\% & 527 & 0\% & 97\% & 542 \\ 
-  2015 & United Kingdom & 2\% & 1\% & 1387 & 0\% & 93\% & 1388 \\ 
-  2015 & United States & 10\% & 0\% & 585 & 0\% & 97\% & 544 \\ 
+                          \cmidrule(l){6-8} Year & Countries & N & Avg score & S.E & N & Avg score & S.E \\ \hline2000 & Australia & 294 & 0.22 & 0.07 & 270 & 1.22 & 0.09 \\ 
+  2000 & Austria & 294 & -0.04 & 0.11 & 235 & 1.15 & 0.15 \\ 
+  2000 & Canada & 1970 & 0.30 & 0.05 & 1357 & 0.90 & 0.07 \\ 
+  2000 & Denmark & 230 & -0.01 & 0.13 & 228 & 1.17 & 0.17 \\ 
+  2000 & Finland & 261 & 0.42 & 0.12 & 267 & 0.89 & 0.15 \\ 
+  2000 & France & 253 & 0.04 & 0.04 & 258 & 1.21 & 0.05 \\ 
+  2000 & Germany & 250 & -0.64 & 0.04 & 303 & 1.81 & 0.05 \\ 
+  2000 & Hungary & 291 & -0.46 & 0.09 & 254 & 1.66 & 0.12 \\ 
+  2000 & Italy & 278 & -0.31 & 0.05 & 278 & 0.81 & 0.06 \\ 
+  2000 & Netherlands & 129 & 0.46 & 0.08 & 158 & 1.25 & 0.11 \\ 
+  2000 & Poland & 209 & -0.31 & 0.04 & 166 & 1.25 & 0.06 \\ 
+  2000 & Spain & 318 & -0.27 & 0.05 & 340 & 1.09 & 0.07 \\ 
+  2000 & Sweden & 241 & 0.12 & 0.10 & 246 & 1.06 & 0.13 \\ 
+  2000 & United Kingdom & 531 & 0.16 & 0.04 & 415 & 1.33 & 0.06 \\ 
+  2000 & United States & 222 & -0.43 & 0.02 & 164 & 1.74 & 0.03 \\ 
+  2015 & Australia & 1694 & -0.15 & 0.04 & 1235 & 1.28 & 0.05 \\ 
+  2015 & Austria & 676 & -0.30 & 0.07 & 705 & 1.53 & 0.10 \\ 
+  2015 & Canada & 2215 & 0.20 & 0.03 & 1863 & 0.99 & 0.05 \\ 
+  2015 & Denmark & 1013 & 0.09 & 0.08 & 595 & 1.08 & 0.10 \\ 
+  2015 & Finland & 575 & 0.07 & 0.08 & 584 & 1.16 & 0.11 \\ 
+  2015 & France & 570 & -0.37 & 0.02 & 615 & 1.66 & 0.03 \\ 
+  2015 & Germany & 545 & -0.09 & 0.02 & 582 & 1.45 & 0.03 \\ 
+  2015 & Hungary & 466 & -0.72 & 0.07 & 589 & 1.79 & 0.09 \\ 
+  2015 & Italy & 947 & -0.20 & 0.03 & 1043 & 1.23 & 0.04 \\ 
+  2015 & Netherlands & 521 & 0.13 & 0.04 & 525 & 1.24 & 0.06 \\ 
+  2015 & Poland & 446 & 0.10 & 0.03 & 448 & 1.24 & 0.05 \\ 
+  2015 & Spain & 608 & -0.25 & 0.03 & 703 & 1.24 & 0.04 \\ 
+  2015 & Sweden & 527 & -0.20 & 0.06 & 542 & 1.32 & 0.09 \\ 
+  2015 & United Kingdom & 1387 & -0.03 & 0.03 & 1388 & 1.15 & 0.04 \\ 
+  2015 & United States & 585 & -0.39 & 0.01 & 544 & 1.22 & 0.01 \\ 
    \hline
@@
 
-\subsection{Results}
+The first table shows a description of the sample size and mean score of both top and bottom SES groups for only the first and last time point.
+One main concern from the planned analysis is that getting the top 90th percentile and bottom 10th percentile would result in a small sample size.
+The table suggests that we have a reasonable number of respondents to actually estimate gaps accurately.
+Moreover, we can see that in all instances the bottom SES group has a lower score than all top SES groups.
 
-Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
+We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting their very equal countries.
+However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average, which is 0.
+We also see some countries with major changes from year 2000 to 2015, with, for example, Australia decreasing the low SES gap from 0.22 standard deviations to -0.15, well below the country average.
+In the next section we take a look at this in a more detailed fashion, considering all years in between.
+
+\subsection{Evolution of the achievement gap}
+% Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
 
@@
 
+\caption{Evolution of the achievement gap for selected countries}
 \end{center}
@@
 
-We start by look at the achievement for some countries.
-We plot only mathematics for each country and also a quadratic trend spline for both test scores pooled.
+We start by look at the achievement for some countries in figure 2.
+We plot only mathematics for each country and also a quadratic trend spline for \emph{both mathematics and literacy} pooled.
 Some countries have increased their achievement very strongly.
-For example, France, Austria and surprisingly Sweden.
-In mathematics, France increase the gap by roughly 0.9.
-This pattern happens similarly for literacy with an increase of 0.6.
-For such a short period of time, the magnitud of the increase is reasonably big.
+For example, France, Austria and surprisingly Sweden have very steep slopes.
+France increasde the gap by roughly 0.9 standard deviations, Austria by 0.6 and Sweden by 0.6.
+This pattern happens similarly for literacy.
+For example, France has an increase of 0.6.
+For such a short period of time, the magnitud of these increases are reasonably big.
 
 Given that no one has estimated the evolution of the gap we can't cross-check how other empirical estimations put France at.
-However, we can make some educated inferences.
-For example, \citet{micklewright} also finds that France is a low dispersion country, specifically in the year 2000.
+However, we can take \citet{micklewright} as the closest reference which also finds that France was a low dispersion country in 2000.
+However, there's no evidence on what happened over time.
+Luckily, the work of \citet{bernardi2016} did study the relationship of social origin effects (broadly speaking, not in terms of achievement gaps) in France and found that they increased in the last decades.
 
-However, we don't know what happens over time. But \citet{bernardi2016} did study the relationship of social origin effects and found that they increased very much in France in the last decades.
-
-Other countries have reasonable increases such as Austria and Finland.
-Aside from these countries, some other countries have no change at all.
-Specially, Canada, Netherlands and Spain.
+Other countries have reasonable increases such as Finland and Hungary, with increases of nearly 0.6 and 0.4, respectively.
+Aside from these countries, there are other cases that have no change at all, specifically, Canada, Netherlands and Spain.
 Canada excels here not only because the gap has been stable over time, but because it has the smallest gap of all countries presented here.
@@
 
-Furthermore, we do have some countries which show a decrease in the SES achievement gap.
-United Kingdom, Israel, Poland, United States and Germany show some steep downward slopes.
-Most notably are the U.S and German cases.
+On the other hand, we do have some countries which show a decrease in the SES achievement gap.
+United Kingdom decreased by about 0, Poland by -0.4, and Denmark by -0.2.
+However, the most notable cases are the United States and Germany.
 These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations.
-However, both countries reduced the gaps by -0.6 and -0.9 respectively.
+But in the 15-year time trend both countries reduced the gaps by -0.6 and -0.9 respectively.
 Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}.
+This corroborates the findings of \citet{reardon_portilla}, which found a decreasing gap for kindergarteners.
+This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy.
 
@@
 
-The plot below takes a more direct approach and looks at the percentage change from the first and last timepoint available.
+We find that the initial gap for the U.S in 2000 is 2.17 varying between 2.13 and 2.21.
+This gap is much higher than found in previous studies, but again, we don't know much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely.
+All country-level studies pool many studies together and control for age, removing the age-specific gap effect.
+
+Figure 3 below takes a more direct approach and looks at the percentage change from the first and last timepoint available.
 Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.
@@
 
+\caption{Percentage change in the achievement gap from 2000 to 2015}
 \end{center}
@@
 
-Generally speaking, we see that most countries increased their achievement gap.
-France had an average increase of about 80\% since 2000, whereas Germany had a similar figure but decreasing.
+Generally speaking, we see that most countries increased their achievement gap over time.
+France had an average increase of about 80\% since 2000 varying down to 40\%, whereas Germany had a similar figure but decreasing.
 Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap.
 In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%.
+The benefit of presenting these estimates this way is that we can actually asses the uncertainty of each calculation, and we do find that some of these estimates have wide variability.
+Despite this, most countries show a clear sign of either decreasing or increasing.
 
-Something reassuring is that mathematics and reading follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase.
+Something reassuring is that mathematics and reading (not presented) follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase.
 The previos two graphs for reading are available upon request}.
@@
 
+\caption{Evolution of the achievement gap for separate SES groups}
 \end{center}
@@
 
-The top line belongs to the high SES group whereas the bottom line belongs to the low SES group.
+Figure 4 shows the evolution for both Low and High SES groups separetly.
 The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other.
@@
 On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate.
-there's a group of distinctive groups where both low/high SES rates are very great resulting in the summary line being flat.
+These results highlight the importance of not only summarizing average achievement gaps.
+The source of these gaps vary greatly between countries and we see how each of these patterns contibutes to the overall inequality of a country.
 
-Ths source of this gap can be deconstructed into two broader gaps.
+Ths source of this gap can be deconstructed further into two broader gaps.
 We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group.
-Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap.
 For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th?
 Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier?
+Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap in figure 5.
 Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous graphs \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.
-
-% This plot might not be necessary any more
-% <<>>=
-% 
-% lower <- seq(13, 1, -1)
-% upper <- seq(100, 220, 10)
-% 
-% df <- tibble(country = "fake_country",
-%              years = rep(seq(2000, 2025, 2), 2),
-%              ses = rep(c("lower", "upper"), each = 13),
-%              value = c(lower, upper))
-% 
-% # You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
-% df %>%
-%   ggplot(aes(as.factor(years), value, group = ses, colour = ses)) +
-%   geom_point(size = 0.5, alpha = 0.4) +
-%   geom_line(alpha = 0.4) +
-%   geom_line(stat = "smooth", method = "lm", aes(group = 1),
-%             formula = y ~ splines::ns(x, 3), linetype = "longdash",
-%             colour = "black") +
-%   # geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1, alpha = 0.4) +
-%   # geom_hline(yintercept = 0, linetype = "longdash") +
-%   # coord_cartesian(ylim = c(0, 3)) +
-%   facet_wrap(~ country)
-% @
 
@@
 
+\caption{Evolution of 90/50/10 SES gaps for selected countries}
 \end{center}
@@
 Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group.
+The big takeaway from the U.S is that the decrease of inequality is coming from the shrinking of the 50/10 gap.
 
-In a more convenient way, we can inspect whether there is a lot of variance in the rate at which top/bottom groups are changing.
+In a more convenient way, we can inspect whether there is a lot of variance in the rate at which top/bottom groups are changing in figure 6.
 
@@
 
+\caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
 \end{center}
@@
 If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance.
+We plot this relationship in figure 8.
+
+\subsection{Efficiency-equality tradeoff}
 
@@
 
-{\centering \includegraphics[width=8.5in,height=5in]{figure/avgperformance_inequality-1}
+{\centering \includegraphics[width=6.5in,height=5in]{figure/avgperformance_inequality-1}
 
@@
 
+\caption{Relationship between 90/10 achievement gap and average performance in mathematics}
 \end{center}
@@
 For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}.
-The correlations are between -0.7343267 and -0.8781625.
+The correlations are between -0.73 and -0.88.
 This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}.
@@
 
+\caption{Time trends of achievement gap and average achiement between 2000 and 2015}
 \end{center}

### 2017-07-05 paper/paper.Rnw
Fixed shapes of two graphs

@@
   ggtitle("Standard deviation of test scores over time") +
-  theme_few()
+  theme_few() +
+  theme(text = element_text(size = 12))
 @
@@
   spread(score, value) %>%
-  ggplot(aes(as.factor(wave), mean_math, group = ses, colour = ses)) +
-  geom_errorbar(aes(ymin = lower_math, ymax = upper_math), width = 0.1) +
+  ggplot(aes(as.factor(wave), mean_math, group = ses, colour = ses, shape = ses)) +
   geom_line(alpha = 0.4) +
-  geom_point(size = 0.5, alpha = 0.4) +
+  geom_point(size = 2, alpha = 0.4) +
   coord_cartesian(ylim = c(-1, 2)) +
@@
   scale_colour_discrete(name = NULL, labels = c("Low SES", "High SES")) +
+  scale_shape_discrete(name = NULL, labels = c("Low SES", "High SES")) +
   ggtitle("Evolution of the top/bottom groups by different gaps") +
@@
 @
+\caption{Evolution of the SES gaps for top/bottom groups}
 \end{center}
@@
 \begin{center}
-<<avgperformance_inequality, include = T, out.height = '5in', out.width = '6.5in', fig.align = 'center'>>=
+<<avgperformance_inequality, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
 data_to_plot %>%
@@
   ggplot(aes(as.character(wave), value, group = type, colour = type, shape = type)) +
-  geom_point(alpha = 0.4) +
+  geom_point(size = 2, alpha = 0.4) +
   geom_line(alpha = 0.4) +
-  scale_y_continuous(name = "Standardized test scores (mean 0)") +
+  scale_y_continuous(name = "Test scores in SD") +
   scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +

### 2017-07-05 paper/paper.tex
Fixed shapes of two graphs

@@
 
+\caption{Evolution of the SES gaps for top/bottom groups}
 \end{center}
@@
 
-{\centering \includegraphics[width=6.5in,height=5in]{figure/avgperformance_inequality-1}
+{\centering \includegraphics[width=5.5in,height=5in]{figure/avgperformance_inequality-1}
 

### 2017-07-05 paper/paper.Rnw
Updated README with step-by-step of repo + deleted all useless scripts from before

@@
+% Things to remember:
+% The working directory is not where the data is.
+% Change working directory to "/Users/cimentadaj/Downloads/gitrepo/Achievement_Gap_PISA/paper"
+% for everything below to work.
+
+% There's a code chunk named modeling which lasts about 3-4 hours to run. I explain how to get
+% around that in a comment above the code chunk. For down there.
+
 \documentclass[11pt, a4paper]{article}
@@
 
+% pisa_listcol.Rdata is a dataset produced by the source() expression from the above
+% code chunk
 <<loading_data-recoding>>=
@@
 
-% This chunk needs to have cache = TRUE once you want to run it for like 2 hours until 
-% the results_* models have been created once and the cache can save them
-
+% This chunk takes about 3-4 hours to run. Once I had them ready I saved them so you can read them
+% as rds. If you run the 'compile pdf' this chunk is already cached (DONT CHANGE ANYTHING FROM
+% THIS CHUNK, CAUSE OTHERWISE IT WILL RERUN FOR 3-4 HOURS). If you want to run the code without
+% compiling sweave, comment the test_diff expressions and read as rds.
+
+% Remember to uncomment when compiling sweave again, so the chunk remains intact.
 <<modeling>>=
@@
 
-% Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
+% TODO: Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
 % Is efficiency related to opportunity here? Is efficiency what explains the inequality of opportunity?
@@
 But if the true deviations of the median academic achievement changes over time, then the estimated trend in the SES gaps will be underestimated, or vice versa.
-
-% Show graph of variance over time
-% I'm showing standard deviations, is it better to show variance?
 
@@
   theme_few() +
-  theme(text = element_text(size = 12))
+  theme(text = element_text(size = 14))
 @
@@
 @
+
+% Previous table that had % of low and high education for low and high SES dummies
 
@@
 
+% New table only has sample size, average score and s.e.
 <<table_escs, include = TRUE>>=
@@
 \subsection{Evolution of the achievement gap}
-% Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
+% TODO: Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
 
@@
 
+% Code chunks for the same plots as above for reading. Available upon request.
 <<evolution_reading, eval = F>>=
@@
 
-% How to output data frames are latex tables?
 <<rate_change>>=
@@
 
-<<table_sample>>=
-sample_tables_topbottom %>%
-  filter(name %in% c(2000, 2015)) %>%
-  transmute(name, country, escs_dummy, N = paste0(n, " (", perc, ")")) %>%
-  spread(escs_dummy, N) %>%
-  rename(Year = name,
-         Country = country,
-         `10th SES` = `0`,
-         `90th SES` = `1`) %>%
-  select(Country, Year, everything()) %>%
-  arrange(Country, Year) %>%
-  xtable::xtable()
-@
-
 <<next_steps>>=

### 2017-07-05 Datasets_PISA.R

@@ -1,73 +1,0 @@
-if (!require(devtools)) {
-    install.packages("devtools")
-    require(devtools)
-}
-# install_github("pbiecek/PISA2012lite")
-# load it
-
-library("PISA2012lite")
-library(Hmisc)
-cntrylist <- list(student=student2012, parent=parent2012, school=school2012, cognitive=scoredItem2012)
-
-means <- by(student2012[, c("PV1MATH", "W_FSTUWT")], student2012$CNT,
-            function(x) wtd.quantile(x[,1], x[,2], probs = c(0.25,0.75)))
-
-meansx <- by(student2012[, c("PV1MATH", "W_FSTUWT")], student2012$CNT,
-            function(x) wtd.mean(x[,1], x[,2]))
-
-library(car)
-library(arm)
-library(purrr)
-library(dplyr)
-library(tidyr)
-
-student2012$ST17Q01 <- as.numeric(student2012$ST17Q01)
-student2012$ST17Q01 <- recode(student2012$ST17Q01,"7:8 = NA; 1:2 = 1")
-student2012$ST17Q01 <- as.factor(student2012$ST17Q01)
-
-
-r.sq2 <- by(student2012[, c("PV1MATH", "ST17Q01")], student2012$CNT,
-   function(x) summary(lm(PV1MATH ~ ST17Q01, x))$adj.r.squared)
-
-
-means2 <- as.data.frame(do.call(rbind, means))
-means2$cntr <- row.names(means2); means2 <- means2[,c("cntr","25%","75%")]
-row.names(means2) <- 1:nrow(means2)
-
-means2 <- cbind(means2, mean=as.numeric(meansx), r.sq=as.numeric(r.sq2))
-means2$Diff <- means2[,"75%"] - means2[,"25%"]
-
-means2 <- means2[order(means2$Diff), ]
-
-
-means3 <- means2$Diff; names(means3) <- row.names(means2)
-dotchart(means3, pch = 19)
-
-means4 <- means2$r.sq; names(means4) <- means2$cntr;
-dotchart(sort(means4), pch=19)
-
-
-
-ggplot(means2, aes(mean, Diff, color=r.sq)) + geom_smooth(colour="blue",method="lm", se=F) +
-    geom_point(size=4, alpha=0.5) +
-    geom_smooth(colour="red",se=F) +
-    scale_color_continuous(low="red",high = "green") +
-    annotate("text", x=375, y=175, label=paste("R of",
[…]

### 2017-07-05 inequality_explr.R

@@ -1,325 +1,0 @@
-library(saves)
-library(haven)
-library(intsvy)
-library(cimentadaj) # For pisa_countrynames
-library(countrycode) # For region variable
-library(tidyverse)
-library(car)
-library(readr)
-
-source("./transform_data.R")
-
-# For PISA_2015
-pisa2015_conf <- list(variables = list(pvlabelpref = "PV",
-                                       pvlabelsuff = "READ",
-                                       weightFinal = "W_FSTUWT",
-                                       weightBRR = "W_FSTURWT"),
-                      parameters = list(cutoffs = c(357.77, 420.07, 482.38, 544.68, 606.99, 669.30),
-                                        percentiles = c(5, 10, 25, 75, 90, 95),
-                                        PVreps = 10,
-                                        BRRreps = 80,
-                                        weights = "BRR",
-                                        replication_scheme = 'pisa')
-)
-
-pisa_all <- read_rds("./data/pisa_listcol.Rdata")
-
-pisa_all2 <- pisa_all
-quantiles <- c(10, 50, 90)
-
-# Function computes arbitrary descriptives with fun for a specific class(topic).
-# Using the specified configuration for pisa survey.
-
-pisa_fun_switch <- function(fun, topic, groups, data, quant, config = pisa_conf) {
-  if (fun == "quantile" && missing(quant)) stop("Quantiles not supplied")
-  
-  topic_mean <- paste0("PV", seq_len(config$parameters$PVreps), topic) # Create variable name for mean
-  
-  # Depending on the type of function, switch to the appropriate function
-  switch(fun,
-  "mean" = intsvy.mean.pv(pvnames = topic_mean, by = groups, data = data, config = config),
-  "quantile" = intsvy.per.pv(pvlabel = topic, by = groups, per = quant, data = data, config = config)
-  )
-}
-
-# Function loops over df, applies the fun for the chosen topic and the groups specified.
-# quant is an optional argument for specifiying quantiles.
-
-pisa_yearly_mean <- function(fun, topic, groups, df, quant) {
-  stopifnot(class(df[[2]]) == "list")
-  
[…]

### 2017-07-05 new_inequality.R

@@ -1,574 +1,0 @@
-library(arm)
-library(saves)
-library(haven)
-library(PISA2000lite)
-library(PISA2003lite)
-library(PISA2006lite)
-library(PISA2009lite)
-library(PISA2012lite)
-library(intsvy)
-library(cimentadaj) # For pisa_countrynames
-library(countrycode) # For region variable
-library(tidyverse)
-library(car)
-library(readr)
-library(SAScii)
-library(inequalityintsvy)
-library(lme4)
-library(modelr)
-
-# For PISA_2015
-pisa2015_conf <- list(variables = list(pvlabelpref = "PV",
-                                       pvlabelsuff = "READ",
-                                       weightFinal = "W_FSTUWT",
-                                       weightBRR = "W_FSTURWT"),
-                      parameters = list(cutoffs = c(357.77, 420.07, 482.38, 544.68, 606.99, 669.30),
-                                        percentiles = c(5, 10, 25, 75, 90, 95),
-                                        PVreps = 10,
-                                        BRRreps = 80,
-                                        weights = "BRR",
-                                        replication_scheme = 'pisa')
-)
-
-pisa_all <- read_rds("./data/pisa_listcol.Rdata")
-pisa_all2 <- pisa_all
-
-db <- paste0("pisa", seq(2000, 2015, 3))
-pisa_all2$value <- map2(pisa_all2$value, db, ~ { .x$wave <- .y; .x})
-pisa_all2$value[[1]]$CNT <- pisa_all2$value[[1]]$COUNTRY
-
-pisa_all2$value <- map(pisa_all2$value, ~ {
-  .x$father_edu <- car::recode(as.numeric(.x$FISCED), "8:9 = NA")
-  .x$mother_edu <- car::recode(as.numeric(.x$MISCED), "8:9 = NA")
-  .x$high_edu_broad <- pmax(.x$father_edu, .x$mother_edu)
-  .x$country <- pisa_countrynames[.x$CNT]
-
-  if (any(unique(.x$wave) %in% c("pisa2012", "pisa2015"))) {
-    .x$father_edu <- .x$father_edu + 1
-    .x$mother_edu <- .x$mother_edu + 1
-    .x$high_edu_broad <- .x$high_edu_broad + 1
-  }
-  .x
-})
-
[…]

### 2017-07-05 new_inequality_two.R

@@ -1,576 +1,0 @@
-  library(arm)
-  library(saves)
-  library(haven)
-  library(PISA2000lite)
-  library(PISA2003lite)
-  library(PISA2006lite)
-  library(PISA2009lite)
-  library(PISA2012lite)
-  library(intsvy)
-  library(cimentadaj) # For pisa_countrynames
-  library(countrycode) # For region variable
-  library(tidyverse)
-  library(car)
-  library(readr)
-  library(SAScii)
-  library(inequalityintsvy)
-  library(lme4)
-  library(modelr)
-  
-# source("./transform_data.R")
-  
-  # Conf for PISA_2015
-  pisa2015_conf <- list(variables = list(pvlabelpref = "PV",
-                                         pvlabelsuff = "READ",
-                                         weightFinal = "W_FSTUWT",
-                                         weightBRR = "W_FSTURWT"),
-                        parameters = list(cutoffs = c(357.77, 420.07, 482.38, 544.68, 606.99, 669.30),
-                                          percentiles = c(5, 10, 25, 75, 90, 95),
-                                          PVreps = 10,
-                                          BRRreps = 80,
-                                          weights = "BRR",
-                                          replication_scheme = 'pisa')
-  )
-  
-  pisa_all <- read_rds("./data/pisa_listcol.Rdata")
-  pisa_all2 <- pisa_all
-  
-  years <- seq(2000, 2015, 3)
-  countries <- c("Germany", "United States", "Denmark", "Sweden", "United Kingdom",
-                 "Spain", "Italy", "Canada", "Australia")
-  
-  db <- paste0("pisa", years)
-  pisa_all2$value <- map2(pisa_all2$value, db, ~ { .x$wave <- .y; .x})
-  pisa_all2$value[[1]]$CNT <- pisa_all2$value[[1]]$COUNTRY
-  
-  pisa_all2$value <- map(pisa_all2$value, ~ {
-    .x$father_edu <- car::recode(as.numeric(.x$FISCED), "8:9 = NA")
-    .x$mother_edu <- car::recode(as.numeric(.x$MISCED), "8:9 = NA")
-    .x$high_edu_broad <- pmax(.x$father_edu, .x$mother_edu)
-    .x$country <- pisa_countrynames[as.character(.x$CNT)]
-  
-    if (any(unique(.x$wave) %in% c("pisa2012", "pisa2015"))) {
[…]

### 2017-07-05 playing_around/reardon_attempt.R

@@ -1,0 +1,36 @@
+# This is, so far, a failed attempt at calculating the 90/10
+# percentiles for an ordinal variable. This is an implementation taken
+# from Reardon (2011), so please check the paper first.
+
+library(GoFKernel)
+library(tidyverse)
+
+x <- rep(1:7, each = 10)
+y <- seq(250, 500, length.out = 70)
+
+x_density <- density(x, adjust = 2)
+plot(our_density)
+
+lookup <- tapply(X = our_density$y, INDEX = rep(1:7, each = 73), FUN = mean)
+prop <- c(0, unique(cume_dist(sort(m$high_edu_broad))) * 100)
+
+our_dens_fun <- function(x) x * density(x, adjust = 2, n = length(x))$y
+our_dens_fun(x)
+
+x <- 1
+integrate(our_dens_fun, lower = c_k[1], upper = c_k[1 + 1])
+
+inverse_cdf <- inverse(integrate(our_dens_fun, lower = prop[x], upper = prop[x + 1]))
+
+# given a ck and ck_prev
+theta_bottom <- inverse_cdf(prop[x])
+theta_top <- inverse_cdf(prop[x + 1])
+
+# A(1) from paper
+fn <- function (x) {
+  x * our_density(x)
+}
+
+theta_hat <- integrate(fn, theta_bottom, theta_top) / (ck - ck_prev)
+
+lm(y ~ 1 + theta_hat + (theta_hat^2 + ck_bla bla))

### 2017-07-14 paper/paper.Rnw
Redid mistake from stan branch

@@
   library(ggthemes)
+  library(matrixStats)
 
@@
       .x$dv <- apply(.x[test_vars], 1, median, na.rm = T)
+      .x$dv <- rowMedians(as.matrix(.x[test_vars]), na.rm = T)
 
@@
 But fast-forward until today, we've understood the relationship much better.
+Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years.
+The idea of meritocracy has motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century.
+When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family SES and student performance are tightly linked.
+The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements.
+For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.
+But fast-forward until today, we've understood the relationship much better.
 
@@
 Although this sounds logical, it's importance is not understood entirely.
+% Here, instead of James Heckman, you should cite the literature on developmental psychology
+% which had already discovered much of what heckman was saying.
 
@@
 Moreover, many have suggested different explanations and the topic has been very contested \citep{brooks1997}.
+% And schools can do a great deal to diminish it
+The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
+This means that inequality starts strong before the schooling of a child begins and schools can help diminish the influence of the origin of a child.
+The impact of inequality can be eased significantly if children are enrolled in high quality learning environments, specially for children coming from unfavourable conditions.
+But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\).
+Although this sounds logical, it's importance is not understood entirely.
+
+The implications of this cumulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\).
+That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns.
+Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.
 
@@
 In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
+From Denmark, for example, the literature has learned a great deal about how to improve social mobility.
+But that was possible by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility specially for less advantaged children \citep{bjorklund2009, jaeger2007}.
+The important finding came when we discovered the main reason behind their social escalator: it's educational system.
+For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential.
+It does so very strongly for children which otherwise would be doomed to have a bleak future.
+Moreover, it tracks students into different curricula very late compared to other Europen countries (age 16), something which has been linked to less educational inequality.
+These two traits make the Danish system uniquely effective.
+Preschool education attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable.
+In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
 
@@
 The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
+% widening over the years? Gosta, page 3
+% income achievement gap, gosta page 3
 
@@
 Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.
+The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years.
+The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence \citep{anna2016_global} is emerging with a very similar landscape.
+The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940.
+This time-trend has allowed researchers to study achievement gaps in a lengthy period of time.
+Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
+Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}.
+The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
 
@@
 However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008}.
+\citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's.
+In fact, the hard numbers suggest that the gap widened by 40-50\%.
+The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of   75\%.
+Given that the studies before 1970 are less reliable in terms of comparability and sampling design, the author computes all results for both before/after 1970.
+To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2001.
+Using longitudinal data, \citet{bradbury2015} find similar results.
+Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25.
+Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of  1.25-1.50 standard deviations.
+One important drawback of these studies is that they don't present the uncertainty of this estimates.
+Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
+It could be that the gap is at  1.25 standard deviations, varying up to 1.75 and down to 0.80.
+Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.
 
 \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
+Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation.
+However likely, this is a highly contested topic.
+A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality.
+However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008}\footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
+.
+% explain more in detail white income inequality is linked to to the achievement gap
+% gosta page 3
 
@@
 Their conclusions are very speculative and have no empirical support, which is why this is still an open question.
+The reasons the authors find a reversal in the trend could be numerous and should be studied very closely.
+The authors incorporate a number of country-level indicators to try explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment.
+In fact, the point out that in this same period (1998 - 2010) the SES gap in preschool enrollment decreased substantially.
+Their conclusions are very speculative and have no empirical support, which is why this is still an open question.
+
+%
 
@@
 In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
+One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little curricular tracking.
+These four countries have no major jump in school selection, quite the opposite to the average European country.
+A thorough review by \citet{werfhorst_mijs} sheds some light on the subject.
+First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps.
+\citet{hanushek_woesmann_tracking}, use the Progress in International Reading Literacy Study (PIRLS), the Trends in International Mathematics and Science Study (TIMSS) and the Programme for International Student Assessment (PISA) surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
+The results suggest that early tracking increases educational inequality.
+While less clear, there is also a tendency for early tracking to reduce mean performance.
+\citet{micklewright} using PISA but a different empirical strategy, find that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
+In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
 
@@
 That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.
+Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States.
+We will do this by studying the changes in the achievement gap in a time-frame of 15 years.
+This will help asses the magnitude of the inequality as well as the rate of growth/decline.
+This study not only documents the size and changes of the SES gap, but pays particular attention to the source of the changes.
+That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.
 
 More formally, We're interested in studying
+More formally, we're interested in studying
 a) Which countries are experiencing changes in the achievement gap;
@@
 After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
+This shows that it is important to study the changes in the achievement gap over time.
+This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.
+a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
+After recording a SES gap increase of about 40\% in only 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
 This shows that it is important to study the changes in the achievement gap over time.
@@
 If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.
+c) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
+The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind.
+We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.
+This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving.
+For example, if whether bottom performers are stagnated and the top is gaining ground, or is it the other way around.
+Or is it the case that both SES groups are distancing themselves at the same rate.
 
@@
 With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system.
+None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.
+PISA collects data through a two-stage stratified sampling design.
+With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests.
+The sample size for PISA 2000 is \Sexpr{country_rows[1]}, \Sexpr{country_rows[2]} for PISA 2003, \Sexpr{country_rows[3]} for PISA 2006, \Sexpr{country_rows[4]} for PISA 2009, \Sexpr{country_rows[5]} for PISA 2012, and \Sexpr{country_rows[6]} for PISA 2015.
+Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance.
+With the recent inclusion of PISA 2015, we have a new opportunity to study six waves that make up a time-series analysis of 15 years.
 None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.
@@
 PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.
+Another social background variable is the International Socio-Economic Index of Occupational Status (ISEI).
+This variable attempts to capture the social status of the family, without asking for income information.
+This index variable was developed by \citet{ganzeboom1996, ganzeboom2010} and it attempts to measure occupational status using a continuous measure.
+This indicator is a very reliable alternative to the  classical Erikson-Goldthorpe-Portocarero classification \citep{erikson1979}.
+It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}.
+PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.
 
@@
 To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.
+
+% Talk in more detail about the ESCS index - is it a factor analysis? What's the methodology like?
 
@@
 
+% We then use it to calculate the mean achievement and (standard error or standard deviation)?
 For each country in each wave, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps.
@@
 
+% Justify why you use the AGE variable
 \begin{equation}
@@
 \end{equation}
+
+% Change the hat for something else because it makes the notation confusing.
 
@@
 
+% Try plotting this for each country and wave
+% Like, every country as a row and 6 columns for each wave.
 \begin{figure}
@@
 
+% As well, updated y hat inside the equation with the updated y from the
+% equation above. And then change the hat from the left hand side with a new indicator.
+
 \begin{equation}
@@
 
+% Explain more in detail how the reliability indicator is calculated.
+
 After constructing the adjusted test score measurement, we estimate the SES dummy by calculating the weighted 90th and 10th quantiles for the SES composite index, and then generating a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.
@@
 
+% Replace the Y_i with the chosen y from the previos equation.
 \begin{equation}
 Y_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
+Y_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
 \end{equation}
+
+% Try running also a linear model together with the multilevel model and ompcare results
+% Brunno suggested this because countries which might have a very different SES estimate
+% might be moved more towards the average.
 
@@
 For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.
+We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends.
+We fit a multilevel rather than an OLS model because by pooling the information together, we weight countries appropriately to their sample size.
+Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible.
+For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.
+% Explain more in detail that it's a dummy and not three groups.
 
@@
 
+% Add confidence intervals
 \begin{figure}
@@
 We plot this relationship in figure 8.
+Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's a tradeoff between the average country performance and the widening of the achievement gap and if it stands over time.
+One particular topic that has been researched in the past is whether improving the overall performance of a country compromises the equality of the educational system.
+For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance.
+This, however, can come at a price.
+If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance.
+We plot this relationship in figure 8.
 
@@
 
+% Add descriptive table with average test scores, degree of tracking, level of income inequality, among other things.
+
 Looking at figure 9, we see a strong correspondence between the two variables.

### 2017-07-29 paper/paper.Rnw
For some reason, some text was repeated twice - deleted now + added new citation

@@
 But fast-forward until today, we've understood the relationship much better.
-Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years.
-The idea of meritocracy has motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century.
-When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family SES and student performance are tightly linked.
-The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements.
-For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.
-But fast-forward until today, we've understood the relationship much better.
-
-The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
-This means that inequality starts strong before the schooling of a child begins and schools can do so much to close it.
+
+Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive and personality definition \citep{duyme1999, waldfogel1996}.
+Economists and educators have long neglected this branch of literature but the work of James Heckman has brought our attention to it.
+For example, he has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
+This means that inequality starts strong before the schooling of a child begins and schools have the potential to equalize the learning ground.
 These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions.
@@
 Although this sounds logical, it's importance is not understood entirely.
-% Here, instead of James Heckman, you should cite the literature on developmental psychology
-% which had already discovered much of what heckman was saying.
 
@@
 Moreover, many have suggested different explanations and the topic has been very contested \citep{brooks1997}.
-% And schools can do a great deal to diminish it
-The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
-This means that inequality starts strong before the schooling of a child begins and schools can help diminish the influence of the origin of a child.
-The impact of inequality can be eased significantly if children are enrolled in high quality learning environments, specially for children coming from unfavourable conditions.
-But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\).
-Although this sounds logical, it's importance is not understood entirely.
-
-The implications of this cumulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\).
-That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns.
-Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.
 
@@
 The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable.
-In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
-From Denmark, for example, the literature has learned a great deal about how to improve social mobility.
-But that was possible by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility specially for less advantaged children \citep{bjorklund2009, jaeger2007}.
-The important finding came when we discovered the main reason behind their social escalator: it's educational system.
-For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential.
-It does so very strongly for children which otherwise would be doomed to have a bleak future.
-Moreover, it tracks students into different curricula very late compared to other Europen countries (age 16), something which has been linked to less educational inequality.
-These two traits make the Danish system uniquely effective.
-Preschool education attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable.
 In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
@@
 Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.
+
 The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years.
@@
 A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality.
-However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008}.
-\citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's.
-In fact, the hard numbers suggest that the gap widened by 40-50\%.
-The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of   75\%.
-Given that the studies before 1970 are less reliable in terms of comparability and sampling design, the author computes all results for both before/after 1970.
-To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2001.
-Using longitudinal data, \citet{bradbury2015} find similar results.
-Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25.
-Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of  1.25-1.50 standard deviations.
-One important drawback of these studies is that they don't present the uncertainty of this estimates.
-Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
-It could be that the gap is at  1.25 standard deviations, varying up to 1.75 and down to 0.80.
-Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.
-
-\footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
-Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation.
-However likely, this is a highly contested topic.
-A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality.
-However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008}\footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
+However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
 .
@@
 In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
-One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little curricular tracking.
-These four countries have no major jump in school selection, quite the opposite to the average European country.
-A thorough review by \citet{werfhorst_mijs} sheds some light on the subject.
-First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps.
-\citet{hanushek_woesmann_tracking}, use the Progress in International Reading Literacy Study (PIRLS), the Trends in International Mathematics and Science Study (TIMSS) and the Programme for International Student Assessment (PISA) surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
-The results suggest that early tracking increases educational inequality.
-While less clear, there is also a tendency for early tracking to reduce mean performance.
-\citet{micklewright} using PISA but a different empirical strategy, find that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
-In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
 
@@
 That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.
-Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States.
-We will do this by studying the changes in the achievement gap in a time-frame of 15 years.
-This will help asses the magnitude of the inequality as well as the rate of growth/decline.
-This study not only documents the size and changes of the SES gap, but pays particular attention to the source of the changes.
-That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.
-
-More formally, We're interested in studying
+
 More formally, we're interested in studying
@@
 If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.
-c) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
-The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind.
-We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.
-This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving.
-For example, if whether bottom performers are stagnated and the top is gaining ground, or is it the other way around.
-Or is it the case that both SES groups are distancing themselves at the same rate.
 
@@
 With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system.
-None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.
-PISA collects data through a two-stage stratified sampling design.
-With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests.
-The sample size for PISA 2000 is \Sexpr{country_rows[1]}, \Sexpr{country_rows[2]} for PISA 2003, \Sexpr{country_rows[3]} for PISA 2006, \Sexpr{country_rows[4]} for PISA 2009, \Sexpr{country_rows[5]} for PISA 2012, and \Sexpr{country_rows[6]} for PISA 2015.
-Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance.
-With the recent inclusion of PISA 2015, we have a new opportunity to study six waves that make up a time-series analysis of 15 years.
 None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.
@@
 This indicator is a very reliable alternative to the  classical Erikson-Goldthorpe-Portocarero classification \citet{erikson1979}.
-It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}.
-PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.
-Another social background variable is the International Socio-Economic Index of Occupational Status (ISEI).
-This variable attempts to capture the social status of the family, without asking for income information.
-This index variable was developed by \citet{ganzeboom1996, ganzeboom2010} and it attempts to measure occupational status using a continuous measure.
-This indicator is a very reliable alternative to the  classical Erikson-Goldthorpe-Portocarero classification \citep{erikson1979}.
 It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}.
@@
 
-Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's an efficiency/equality tradeoff and if it stands over time.
-One particular topic that has been researched in the past is whether improving the overall cognitive development of a country compromises the distribution of the countries overall performance.
-For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance.
-This, however, can come at a price.
-If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance.
-We plot this relationship in figure 8.
 Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's a tradeoff between the average country performance and the widening of the achievement gap and if it stands over time.

### 2017-07-29 paper/paper.Rnw
Fixed math notation + added corrections from gosta dn bruno

@@
 Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation.
+\citet{reardon2011} offers several possible links, with the most reasonable being that family investment patterns have changed so that high income families now invest more resources on their children.
+This explanation lies in the fact that increasing income became more strongly correlated with other positive family traits related to time allocation and welfare services.
+
 However likely, this is a highly contested topic.
 A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality.
-However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
+However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970 where income inequality was increasing.}.
 .
-% explain more in detail white income inequality is linked to to the achievement gap
-% gosta page 3
 
@@
 Their conclusions are very speculative and have no empirical support, which is why this is still an open question.
-
-%
 
@@
 This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming all mentioned variables into comparable metrics across waves.
+
+The ESCS was derived from a principal component analysis of standardised variables, taking the factor scores for the first principal component as measures of the PISA index of economic, social and cultural status.
+All countries and economies (both OECD and partner countries/economies) contributed equally to the principal component analysis, while in previous cycles, the principal component analysis was based on OECD countries only.
+However, for the purpose of reporting the ESCS scale has been transformed with zero being the score of an average OECD student and one being the standard deviation across equally weighted OECD countries \citep{pisa_2015_results}.
+
 To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.
-
-% Talk in more detail about the ESCS index - is it a factor analysis? What's the methodology like?
 
@@
 
-% We then use it to calculate the mean achievement and (standard error or standard deviation)?
 For each country in each wave, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps.
@@
 
-for each wave,  where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (because all children should be 15 year olds) weighted by the student sample weights.
-
-We then adjust \begin{math} \hat{Y_i} \end{math} by
+for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011}) weighted by the student sample weights.
+
+We then adjust \begin{math} \hat{\gamma_i} \end{math} by
 
@@
 \begin{equation}
-\hat{Y_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
+\hat{\gamma_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
 \end{equation}
-
-% Change the hat for something else because it makes the notation confusing.
 
@@
 
-In order to correct gap estimates for measurement error, I adjust student test scores(\begin{math} \hat{Y_i} \end{math}) by
-
-% As well, updated y hat inside the equation with the updated y from the
-% equation above. And then change the hat from the left hand side with a new indicator.
+In order to correct gap estimates for measurement error, I adjust student test scores(\begin{math} \hat{\gamma_i} \end{math}) by
 
 \begin{equation}
-\hat{Y_i} = \hat{Y_i} * \frac{1}{\sqrt{r}}
+\hat{\gamma_i} = \hat{\gamma_i} * \frac{1}{\sqrt{r}}
 \end{equation}

### 2017-07-29 paper/paper.Rnw
Finished some comments - only text changes

@@
 This is different from previous work because it concentrates solely on 15 year old children, something not done for the achievement gap, and it attempts to capture the evolution of the achievement gap for countries with different educational systems.
-In addition, extend the previous findings by studying the relationship between the gap and the average performance of each country in order to detect if there's a tradeoff between efficiency and equality over time separately for each country.
+In addition, extend the previous findings by studying the relationship between the gap and the average performance of each country in order to detect if there's a tradeoff between good performance and equality over time separately for each country.
 The advantages of this study are twofold.
@@
 This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving.
-If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.
+If the bottom performers are stagnated whereas the top is gaining ground, the performance-equality trade off dilemma of tracking might be much more credible than other explanations.
 
@@
 where \begin{math}r\end{math} is the reliability score of the wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}.
+The realibility estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of Person separation reliability such as internal consistency.
 Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own.
@@
 
-% Explain more in detail how the reliability indicator is calculated.
-
 After constructing the adjusted test score measurement, we estimate the SES dummy by calculating the weighted 90th and 10th quantiles for the SES composite index, and then generating a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.
@@
 \begin{equation}
-Y_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
-Y_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
+\gamma_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
+\gamma_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
 \end{equation}
@@
 We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size.
-Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible.
-For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.
-We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends.
-We fit a multilevel rather than an OLS model because by pooling the information together, we weight countries appropriately to their sample size.
-Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible.
+Given that including the SES \emph{dummy} reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible.
 For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.
@@
 
-\subsection{Efficiency-equality tradeoff}
+\subsection{Performance-equality tradeoff}
 <<graphing_achievement_disparity, cache = T>>=

### 2017-08-01 paper/paper.Rnw
deleted a very slow line for a faster line
Implemented reardon approach - it's not very reliable

@@
 <<loading_data-recoding>>=
-pisa_all <- read_rds("./data/pisa_listcol.Rdata")
+pisa_all <- read_rds("./paper/data/pisa_listcol.Rdata")
 
@@
 
+<<>>=
+
+library(perccalc)
+
+d <- map(1:6, ~ {
+
+df <- pisa_all2$value[[.x]]
+
+df_try <-
+  df %>%
+  filter(country == "France") %>%
+  select(country, high_edu_broad, matches("^PV*.MATH$"), W_FSTUWT) %>%
+  mutate(high_edu_broad = factor(high_edu_broad, ordered = TRUE))
+
+math_score <-
+  df_try %>%
+  select_if(is_double) %>%
+  as.matrix() %>%
+  rowMedians(na.rm = TRUE)
+
+df_try <-
+  df_try %>%
+  select(country, high_edu_broad, W_FSTUWT) %>%
+  add_column(math_score)
+
+perc_diff(df_try, high_edu_broad, math_score, percentiles = c(90, 10))
+}) %>% 
+  setNames(seq(2000, 2015, 3)) %>%
+  enframe() %>%
+  unnest() %>%
+  slice(seq(1, 12, 2))
+
+d %>%
+  ggplot(aes(name, value)) +
+  geom_point() +
+  geom_line(aes(group = 1))
+
+@
+
 <<functions_for_modelling>>=
-
-# Function calculates the bottom 30th quantile for the bottom educated and the 70th quantile
-# for the top educated. If the quantiles can't be estimated, it returns two NA's instead
-quantile_missing <- function(df, weights, probs) {
-
-    quan <- try(Hmisc::wtd.quantile(
-      df$escs_trend,
-      weights = df[[weights]],
-      probs = probs
-      ))
-
-    if (any("try-error" %in% class(quan))) {
-      return(c(NA, NA))
-      } else {
-     return(c(quan[1], quan[2]))
-    }
-}
 
@@
 
-# It returns a dataframe for each survey with all countries and respective coefficients and
-# standard errors.
-test_diff <- function(df, reliability, test, probs) {
-
-    map2(df, reliability, function(.x, .y) {
-
-      conf <- if (unique(.x$wave) == "pisa2015") pisa2015_conf else pisa_conf
-      weights_var <- conf$variables$weightFinal
-
-      country_split <- split(.x, .x$country)
-
-      country_list <- map(country_split, function(country) {
-        print(unique(country$country))
-
-        quan <- quantile_missing(country, weights_var, probs)
-
-        # It's very important to create a variable that returns the number of observations of this dummy
-        # For each country. Possibly to weight by the number of observations.
-        country$escs_dummy <-
-          with(country, case_when(escs_trend >= quan[2] ~ 1,
-                                  escs_trend <= quan[1] ~ 0))
-        country
-      })
-
-      .x <-
-        enframe(country_list) %>%
-        unnest(value)
-
-      .x <-
-        .x %>%
-        dplyr::select(wave,
-                      matches(paste0("^PV.*", test, "$")),
-                      escs_dummy,
-                      country,
-                      one_of(weights_var),
-                      AGE)
-
-      message(paste(unique(.x$wave), "data ready"))
-
-      test_vars <- paste0("PV", seq_len(conf$parameters$PVreps), test)
-      .x[test_vars] <- map(.x[test_vars], ~ ifelse(.x == 9997, NA, .x))
-
-      # Calculate median math score of all PV's
-      .x$dv <- apply(.x[test_vars], 1, median, na.rm = T)
-      .x$dv <- rowMedians(as.matrix(.x[test_vars]), na.rm = T)
-
-      # Should I estimate the model separately by country?
-      mod1 <- lm(dv ~ AGE,
-                 weights = .x[[weights_var]],
-                 data = .x,
-                 na.action = "na.exclude")
-
-      # Take residuals of model and divide by rmse. Multiply that by
-      # 1 / sqrt(reliability of each survey), which is .y in the loop.
-      .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)
-
-      mod2 <-
-        lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
-             data = .x,
-             weights = .x[[weights_var]])
-
-      # Take the country coefficients (absolute coefficients)
-      country_coef <-
-        coef(mod2)$country %>%
-        rownames_to_column() %>%
-        gather(escs_dummy, Mean, -rowname) %>%
-        mutate(escs_dummy = dplyr::recode(escs_dummy,
-                                          `(Intercept)` = "0",
-                                          `escs_dummy` = "1"))
-
-      # Take the absolute country standard errors
-      se <-
-        se.coef(mod2)$country %>%
-        as.data.frame() %>%
-        rownames_to_column() %>%
-        gather(escs_dummy, s.e., -rowname) %>%
-        mutate(escs_dummy = dplyr::recode(escs_dummy,
-                                          `(Intercept)` = "0",
-                                          `escs_dummy` = "1"))
-
-      results <-
-        inner_join(country_coef, se, by = c("rowname", "escs_dummy")) %>%
-        rename(country = rowname) %>%
-        arrange(country, escs_dummy)
-
-      message(paste0(unique(.x$wave), " modeling done"))
-      results
-    })
-}
-
-# Adapted from: https://github.com/jtleek/slipper/blob/master/R/slipper.R
-# Returns a tibble with the actual expr + the bootstrapped expr.
-bootstrapper <- function(df, expr, B = 100, n = nrow(df), replacement = TRUE) {
-  bootstrapper_(df, lazyeval::lazy(expr), B, n, replacement)
-}
-
-bootstrapper_ <- function(df, expr, B = 500, n = nrow(df), replacement = TRUE) {
-  obs_val = lazyeval::lazy_eval(expr, data = df)
-  boot_val = replicate(B, {
-    newdata = sample_n(df, n, replace = replacement)
-    lazyeval::lazy_eval(expr, data = newdata)
-  })
-  out = tibble(type = c("observed", "bootstrap"), 
-               value = c(obs_val, mean(boot_val, na.rm = T)))
-  return(out)
-}
-
-# For example
-# bootstrapper(mtcars, mean(mpg), B = 200)
-@
-
-% This chunk takes about 3-4 hours to run. Once I had them ready I saved them so you can read them
-% as rds. If you run the 'compile pdf' this chunk is already cached (DONT CHANGE ANYTHING FROM
-% THIS CHUNK, CAUSE OTHERWISE IT WILL RERUN FOR 3-4 HOURS). If you want to run the code without
-% compiling sweave, comment the test_diff expressions and read as rds.
-
-% Remember to uncomment when compiling sweave again, so the chunk remains intact.
-<<modeling>>=
 adapted_year_data <-
@@
 })
+
+# It returns a dataframe for each survey with all countries and respective coefficients and
+# standard errors.
+test_diff <- function(df, reliability, test, probs) {
+
+    map2(df, reliability, function(.x, .y) {
+
+      conf <- if (unique(.x$wave) == "pisa2015") pisa2015_conf else pisa_conf
+      weights_var <- conf$variables$weightFinal
+
+      .x <-
+        .x %>%
+        dplyr::select(wave,
+                      matches(paste0("^PV.*", test, "$")),
+                      high_edu_broad,
+                      country,
+                      one_of(weights_var),
+                      AGE) %>%
+        mutate(high_edu_broad = factor(high_edu_broad, ordered = TRUE))
+
+      message(paste(unique(.x$wave), "data ready"))
+
+      test_vars <- paste0("PV", seq_len(conf$parameters$PVreps), test)
+      .x[test_vars] <- map(.x[test_vars], ~ ifelse(.x == 9997, NA, .x))
+
+      # Calculate median math score of all PV's
+      .x$dv <- rowMedians(as.matrix(.x[test_vars]), na.rm = T)
+
+      # Should I estimate the model separately by country?
+      mod1 <- lm(dv ~ AGE,
+                 weights = .x[[weights_var]],
+                 data = .x,
+                 na.action = "na.exclude")
+
+      # Take residuals of model and divide by rmse. Multiply that by
+      # 1 / sqrt(reliability of each survey), which is .y in the loop.
+      .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)
+
+      error_perc_diff <- function(data) {
+        result <- try(perc_diff(data, high_edu_broad, adj_pvnum, W_FSTUWT, percentiles = probs))
+
+        if (class(result) == "try-error") {
+          c(NA, NA)
+        } else {
+          result
+        }
+      }
+
+      .x <- .x %>% filter(country %in% countries)
+
+      mod2 <- 
+        .x %>%
+        split(.x$country) %>%
+        map(~ error_perc_diff(.x)) %>%
+        Filter(function(x) !all(is.na(x)), .)
+
+      mod2 %>%
+        reduce(bind_rows) %>%
+        mutate(country = names(mod2))
+    })
+}
+
+results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
+
+results_math %>%
+  setNames(years) %>%
+  enframe(name = "year") %>%
+  unnest() %>%
+  filter(country != "United States") %>%
+  mutate(low = difference - 1.96 * se,
+         high = difference + 1.96 * se) %>%
+  ggplot(aes(year, difference)) +
+  geom_point() +
+  geom_line(aes(group = country)) +
+  geom_errorbar(aes(ymin = low, ymax = high)) +
+  facet_wrap(~ country)
+
+# .x <- adapted_year_data[[4]]
+# .y <- reliability_pisa[3]
+# test <- "MATH"
+# probs <- c(0.9, 0.1)
+
+# Adapted from: https://github.com/jtleek/slipper/blob/master/R/slipper.R
+# Returns a tibble with the actual expr + the bootstrapped expr.
+bootstrapper <- function(df, expr, B = 100, n = nrow(df), replacement = TRUE) {
+  bootstrapper_(df, lazyeval::lazy(expr), B, n, replacement)
+}
+
+bootstrapper_ <- function(df, expr, B = 500, n = nrow(df), replacement = TRUE) {
+  obs_val = lazyeval::lazy_eval(expr, data = df)
+  boot_val = replicate(B, {
+    newdata = sample_n(df, n, replace = replacement)
+    lazyeval::lazy_eval(expr, data = newdata)
+  })
+  out = tibble(type = c("observed", "bootstrap"), 
+               value = c(obs_val, mean(boot_val, na.rm = T)))
+  return(out)
+}
+
+# For example
+# bootstrapper(mtcars, mean(mpg), B = 200)
+@
+
+% This chunk takes about 3-4 hours to run. Once I had them ready I saved them so you can read them
+% as rds. If you run the 'compile pdf' this chunk is already cached (DONT CHANGE ANYTHING FROM
+% THIS CHUNK, CAUSE OTHERWISE IT WILL RERUN FOR 3-4 HOURS). If you want to run the code without
+% compiling sweave, comment the test_diff expressions and read as rds.
+
+% Remember to uncomment when compiling sweave again, so the chunk remains intact.
+<<modeling>>=
 

### 2017-08-02 paper/paper.Rnw
Updated to run lm separate for each country when adjusting cognitive variable

--- 2017-07-29 paper/paper.Rnw
@@
 # standard errors.
+
 test_diff <- function(df, reliability, test, probs) {
@@
       # Calculate median math score of all PV's
-      .x$dv <- apply(.x[test_vars], 1, median, na.rm = T)
       .x$dv <- rowMedians(as.matrix(.x[test_vars]), na.rm = T)
@@
       # Should I estimate the model separately by country?
+      separate_countries <-
+        .x %>%
+        split(.x$country) %>%
+        map(function(each_country) {
         mod1 <- lm(dv ~ AGE,
-                 weights = .x[[weights_var]],
-                 data = .x,
+                   weights = each_country[[weights_var]],
+                   data = each_country,
                    na.action = "na.exclude")
@@
       # 1 / sqrt(reliability of each survey), which is .y in the loop.
-      .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)
+      each_country$adj_pvnum <- resid(mod1)/rmse(mod1, each_country) * 1 / sqrt(.y)
+      each_country
+      })
+
+      .x <-
+        separate_countries %>%
+        reduce(bind_rows)
 

### 2017-08-16 paper/paper.Rnw
Downloaded all data from pre-primary

--- 2017-07-29 paper/paper.Rnw
@@
 
-results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
-results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
-results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
-results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
-results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
-results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
-
-# results_math <- read_rds("./data/delete.Rdata")
-# results_read <- read_rds("./data/delete_read.Rdata")
-# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
-# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
-# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
-# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
+# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
+# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
+# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
+# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
+# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
+# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
+
+results_math <- read_rds("./paper/data/delete.Rdata")
+results_read <- read_rds("./paper/data/delete_read.Rdata")
+results_math_topmid <- read_rds("./paper/data/delete_math_topmid.Rdata")
+results_read_topmid <- read_rds("./paper/data/delete_read_topmid.Rdata")
+results_math_midbottom <- read_rds("./paper/data/delete_math_midbottom.Rdata")
+results_read_midbottom <- read_rds("./paper/data/delete_read_midbottom.Rdata")
 # US is missing for reading
@@
   theme(text = element_text(size = 14))
+
+rm(var_years)
 @
@@
   enframer()
+
+rm(escs_data)
 @
@@
       caption = "SES sample size and ISCED composition")
+
+rm(sample_tables_topbottom, sample_tables_topmid, sample_tables_midbottom,
+   avg_performance, summary_table, table_coming)
 @
@@
 Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.
+
+<<deleting_used_data>>=
+rm(diff_data, us_math_2000, us_math_2000_upper, us_math_2000_lower)
+@
 
@@
   theme_minimal()
+
+rm(full_data, avg_sd_increase_high, avg_sd_increase_low)
 @
@@
   }
-}) %>% setNames(seq(2000, 2015, 3))
+}) %>%
+  setNames(seq(2000, 2015, 3))
 
@@
   geom_smooth()
+
+rm(country_scores)
 @
@@
 
+<<>>=
+
+# 2000
+# SC03Q01 School public/private
+# 1 Public
+# 2 Private
+# 7 N/A
+# 8 M/R
+# 9 Mis
+
+# 2003
+# SC03Q01 (8) Public or private
+# 1 Public
+# 2 Private
+# 7 N/A
+# 8 Invalid
+# 9 Miss
+
+# 2006
+# SC02Q01 (8) Public or private
+#  1 Public
+#  2 Private
+#  7 N/A
+#  8 Invalid
+#  9 Missing
+
+# 2009
+# SC02Q01
+# 1 public
+# 2 private
+
+# 2012
+# SC01Q01 Public or private Num
+# 1 Public
+# 2 Private
+# 7 N/A
+# 8 Invalid
+# 9 Missing
+
+# 2015
+# SC013Q01TA
+# 1 Public
+# 2 Private
+# 7 N/A
+# 8 Invalid
+# 9 Missing
+
+# 2000: ST25Q01
+# 1 <ISCED 2A>
+# 2 <ISCED 2B>
+# 3 <ISCED 2C>
+# 4 <ISCED 3A>
+# 5 <ISCED 3B>
+# 6 <ISCED 3C>
+# 7 N/A
+# 8 M/R
+# 9 Mis
+
+# 2003: ISCEDO
+# 2006: ISCEDO
+# 2009: ISCEDO
+# 2012: ISCEDO
+# 2015: ISCEDO
+
+# 1	General
+# 2	Pre-Vocational
+# 3	Vocational
+# 4	Modular
+
+options(tibble.max_extra_cols = 10)
+library(WDI)
+
+wd_extract <- function(var, name) {
+  wdi_data <-
+    WDI(indicator = var, start = 2000, end = 2015) %>%
+    select(-1) %>%
+    as_tibble() %>%
+    mutate(year = as.character(year))
+
+  names(wdi_data)[2] <- name
+  wdi_data
+}
+
+pop_preprimary <- wd_extract("SP.PRE.TOTL.IN", "pop_preprimary")
+teacher_pupil_ratio <- wd_extract("SE.PRE.ENRL.TC.ZS", "teacher_pupil_ratio")
+perc_teacher_trained <- wd_extract("UIS.TRTP.0", "perc_teach")
+preprimary_priv_enrol <- wd_extract("SE.PRE.PRIV.ZS", "preprimary_priv_enrol")
+
+difference_data <-
+  complete_data_topbottom %>%
+  filter(type_test == "math") %>%
+  select(wave, country, difference) %>%
+  rename(year = wave) %>%
+  mutate(year = as.character(year))
+
+library(inequalityintsvy)
+
+difference_data %>%
+  left_join(primary_enrolment) %>%
+  left_join(pop_tot) %>%
+  lmer(difference ~ enrolment + (1 | country), data = .) %>%
+  arm::display()
+
+@
+
 <<next_steps>>=

### 2017-08-17 paper/paper.Rnw
Trying out the 75/25 gap

@@
 library(matrixStats)
+library(WDI)
 
@@
     pisa_all2$value[[6]] %>%
-    rename(escs_trend = ESCS)
+    rename(escs_trend = escs_trend)
 @
@@
 
-# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
-# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
-# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
-# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
-# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
-# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
-
-results_math <- read_rds("./paper/data/delete.Rdata")
-results_read <- read_rds("./paper/data/delete_read.Rdata")
-results_math_topmid <- read_rds("./paper/data/delete_math_topmid.Rdata")
-results_read_topmid <- read_rds("./paper/data/delete_read_topmid.Rdata")
-results_math_midbottom <- read_rds("./paper/data/delete_math_midbottom.Rdata")
-results_read_midbottom <- read_rds("./paper/data/delete_read_midbottom.Rdata")
+results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.75))
+results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.75))
+results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.75))
+results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.75))
+results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.5))
+results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.5))
+
+# results_math <- read_rds("./paper/data/delete.Rdata")
+# results_read <- read_rds("./paper/data/delete_read.Rdata")
+# results_math_topmid <- read_rds("./paper/data/delete_math_topmid.Rdata")
+# results_read_topmid <- read_rds("./paper/data/delete_read_topmid.Rdata")
+# results_math_midbottom <- read_rds("./paper/data/delete_math_midbottom.Rdata")
+# results_read_midbottom <- read_rds("./paper/data/delete_read_midbottom.Rdata")
 # US is missing for reading
@@
 
-escs_data <- escs_dummy_creator(adapted_year_data, c(0.1, 0.9))
+escs_data <- escs_dummy_creator(adapted_year_data, c(0.25, 0.75))
 
@@
 
-complete_data_topbottom <- mutate(complete_data_topbottom, type = "90th/10th SES gap")
-complete_data_topmid <- mutate(complete_data_topmid, type = "90th/50th SES gap")
-complete_data_midbottom <- mutate(complete_data_midbottom, type = "50th/10th SES gap")
+complete_data_topbottom <- mutate(complete_data_topbottom, type = "75th/25th SES gap")
+complete_data_topmid <- mutate(complete_data_topmid, type = "75th/50th SES gap")
+complete_data_midbottom <- mutate(complete_data_midbottom, type = "50th/25th SES gap")
 @
@@
 # 90/10 gaps acros countries
+
+df <- complete_data_topbottom
+
 diff_increase_fun <- function(df) {
@@
 us_math_2000 <-
-  filter(complete_data_topbottom, wave == 2000, country == "United States", type_test == "math") %>%
+  filter(complete_data_topbottom,
+         wave == 2000, country == "United States",
+         type_test == "math") %>%
   .[["difference"]]
@@
 us_math_2000_lower <-
-  filter(complete_data_topbottom, wave == 2000, country == "United States", type_test == "math") %>%
+  filter(complete_data_topbottom,
+         wave == 2000, country == "United States",
+         type_test == "math") %>%
   .[["lower"]]
@@
 us_math_2000_upper <-
-  filter(complete_data_topbottom, wave == 2000, country == "United States", type_test == "math") %>%
+  filter(complete_data_topbottom,
+         wave == 2000, country == "United States",
+         type_test == "math") %>%
   .[["upper"]]
@@
 
-<<>>=
+<<eval = F>>=
 
@@
 
-options(tibble.max_extra_cols = 10)
-library(WDI)
+library(readxl)
 
@@
 
+tracking_data <-
+  readxl::read_xlsx("./paper/data/tracking.xlsx", sheet = "all_data") %>%
+  map_if(is_double, round, 2) %>%
+  as_tibble
+
 difference_data <-

### 2017-08-18 paper/paper.Rnw
Trivial commit

@@
 
-results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.75))
-results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.75))
-results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.75))
-results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.75))
-results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.5))
-results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.5))
-
-# results_math <- read_rds("./paper/data/delete.Rdata")
-# results_read <- read_rds("./paper/data/delete_read.Rdata")
-# results_math_topmid <- read_rds("./paper/data/delete_math_topmid.Rdata")
-# results_read_topmid <- read_rds("./paper/data/delete_read_topmid.Rdata")
-# results_math_midbottom <- read_rds("./paper/data/delete_math_midbottom.Rdata")
-# results_read_midbottom <- read_rds("./paper/data/delete_read_midbottom.Rdata")
+# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.75))
+# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.75))
+# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.75))
+# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.75))
+# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.5))
+# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.5))
+
+# results_math_75 <- results_math
+# results_read_75 <- results_read
+# results_math_topmid_75 <- results_math_topmid
+# results_read_topmid_75 <- results_read_topmid
+# results_math_midbottom_75 <- results_math_midbottom
+# results_read_midbottom_75 <- results_read_midbottom
+
+results_math <- read_rds("./paper/data/delete.Rdata")
+results_read <- read_rds("./paper/data/delete_read.Rdata")
+results_math_topmid <- read_rds("./paper/data/delete_math_topmid.Rdata")
+results_read_topmid <- read_rds("./paper/data/delete_read_topmid.Rdata")
+results_math_midbottom <- read_rds("./paper/data/delete_math_midbottom.Rdata")
+results_read_midbottom <- read_rds("./paper/data/delete_read_midbottom.Rdata")
 # US is missing for reading
@@
 <<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
-# 90/10 gaps acros countries
+# 90/10 gaps across countries
 
@@
 
+p <- 
   complete_data_topbottom %>%
   mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
-  filter(type_test == "read", !is.na(country)) %>%
+  filter(type_test == "math", !is.na(country)) %>%
   ggplot(aes(as.character(wave), difference)) +
   geom_point() +
-  geom_linerange(aes(ymin = 0, ymax = difference))+
-  geom_line(stat = "smooth", method = "lm", aes(group = 1),
+  facet_wrap(~ country, ncol = 5) +
+  geom_linerange(aes(ymin = 0, ymax = difference))
+
+p + geom_line(data = complete_data_topbottom %>%
+                mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
+                filter(!is.na(country)),
+              stat = "smooth", method = "lm", aes(group = 1),
               formula = y ~ splines::ns(x, 2), size = 0.7,
               colour = "red") +
-  facet_wrap(~ country, ncol = 5) +
   scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +

### 2017-08-18 paper/paper.Rnw
Fixed pooled trend line - now with math-read pooled

--- 2017-08-16 paper/paper.Rnw
@@
 <<default_conf>>=
+  library(matrixStats)
   library(knitr)
@@
   library(ggthemes)
-  library(matrixStats)
 
@@
 <<loading_data-recoding>>=
-pisa_all <- read_rds("./paper/data/pisa_listcol.Rdata")
+pisa_all <- read_rds("./data/pisa_listcol.Rdata")
 
@@
 
-# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
-# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
-# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
-# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
-# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
-# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
-
-results_math <- read_rds("./paper/data/delete.Rdata")
-results_read <- read_rds("./paper/data/delete_read.Rdata")
-results_math_topmid <- read_rds("./paper/data/delete_math_topmid.Rdata")
-results_read_topmid <- read_rds("./paper/data/delete_read_topmid.Rdata")
-results_math_midbottom <- read_rds("./paper/data/delete_math_midbottom.Rdata")
-results_read_midbottom <- read_rds("./paper/data/delete_read_midbottom.Rdata")
+results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
+results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
+results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
+results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
+results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
+results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
+
+# results_math <- read_rds("./data/delete.Rdata")
+# results_read <- read_rds("./data/delete_read.Rdata")
+# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
+# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
+# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
+# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
 # US is missing for reading
@@
   theme(text = element_text(size = 14))
-
-rm(var_years)
 @
@@
   enframer()
-
-rm(escs_data)
 @
@@
       caption = "SES sample size and ISCED composition")
-
-rm(sample_tables_topbottom, sample_tables_topmid, sample_tables_midbottom,
-   avg_performance, summary_table, table_coming)
 @
@@
 
+pooled_trendline <-
   complete_data_topbottom %>%
   mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
-  filter(type_test == "read", !is.na(country)) %>%
+  filter(!is.na(country))
+
+pooled_trendline %>%
+  filter(type_test == "math") %>%
   ggplot(aes(as.character(wave), difference)) +
@@
   geom_linerange(aes(ymin = 0, ymax = difference))+
-  geom_line(stat = "smooth", method = "lm", aes(group = 1),
+  geom_line(data = pooled_trendline, stat = "smooth", method = "lm", aes(group = 1),
             formula = y ~ splines::ns(x, 2), size = 0.7,
@@
 Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.
-
-<<deleting_used_data>>=
-rm(diff_data, us_math_2000, us_math_2000_upper, us_math_2000_lower)
-@
 
@@
   theme_minimal()
-
-rm(full_data, avg_sd_increase_high, avg_sd_increase_low)
 @
@@
   }
-}) %>%
-  setNames(seq(2000, 2015, 3))
+}) %>% setNames(seq(2000, 2015, 3))
 
@@
   geom_smooth()
-
-rm(country_scores)
 @
@@
 
+\section{Appendix}
+
 <<>>=
-
-# 2000
-# SC03Q01 School public/private
-# 1 Public
-# 2 Private
-# 7 N/A
-# 8 M/R
-# 9 Mis
-
-# 2003
-# SC03Q01 (8) Public or private
-# 1 Public
-# 2 Private
-# 7 N/A
-# 8 Invalid
-# 9 Miss
-
-# 2006
-# SC02Q01 (8) Public or private
-#  1 Public
-#  2 Private
-#  7 N/A
-#  8 Invalid
-#  9 Missing
-
-# 2009
-# SC02Q01
-# 1 public
-# 2 private
-
-# 2012
-# SC01Q01 Public or private Num
-# 1 Public
-# 2 Private
-# 7 N/A
-# 8 Invalid
-# 9 Missing
-
-# 2015
-# SC013Q01TA
-# 1 Public
-# 2 Private
-# 7 N/A
-# 8 Invalid
-# 9 Missing
-
-# 2000: ST25Q01
-# 1 <ISCED 2A>
-# 2 <ISCED 2B>
-# 3 <ISCED 2C>
-# 4 <ISCED 3A>
-# 5 <ISCED 3B>
-# 6 <ISCED 3C>
-# 7 N/A
-# 8 M/R
-# 9 Mis
-
-# 2003: ISCEDO
-# 2006: ISCEDO
-# 2009: ISCEDO
-# 2012: ISCEDO
-# 2015: ISCEDO
-
-# 1	General
-# 2	Pre-Vocational
-# 3	Vocational
-# 4	Modular
-
-options(tibble.max_extra_cols = 10)
-library(WDI)
-
-wd_extract <- function(var, name) {
-  wdi_data <-
-    WDI(indicator = var, start = 2000, end = 2015) %>%
-    select(-1) %>%
-    as_tibble() %>%
-    mutate(year = as.character(year))
-
-  names(wdi_data)[2] <- name
-  wdi_data
-}
-
-pop_preprimary <- wd_extract("SP.PRE.TOTL.IN", "pop_preprimary")
-teacher_pupil_ratio <- wd_extract("SE.PRE.ENRL.TC.ZS", "teacher_pupil_ratio")
-perc_teacher_trained <- wd_extract("UIS.TRTP.0", "perc_teach")
-preprimary_priv_enrol <- wd_extract("SE.PRE.PRIV.ZS", "preprimary_priv_enrol")
-
-difference_data <-
-  complete_data_topbottom %>%
-  filter(type_test == "math") %>%
-  select(wave, country, difference) %>%
-  rename(year = wave) %>%
-  mutate(year = as.character(year))
-
-library(inequalityintsvy)
-
-difference_data %>%
-  left_join(primary_enrolment) %>%
-  left_join(pop_tot) %>%
-  lmer(difference ~ enrolment + (1 | country), data = .) %>%
-  arm::display()
+results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
+results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
+results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
+results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
+results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
+results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
+
+# results_math <- read_rds("./data/delete.Rdata")
+# results_read <- read_rds("./data/delete_read.Rdata")
+# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
+# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
+# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
+# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
 

### 2017-08-18 paper/paper.Rnw
Fixed trendline for optional reading plot

@@
 # See below that reading graph is upon request. This is the one.
-complete_data_topbottom %>%
-  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
-  filter(type_test == "read", !is.na(country)) %>%
+
+pooled_trendline %>%
+  filter(type_test == "read") %>%
   ggplot(aes(as.character(wave), difference)) +
@@
 perc_graph(top_bottom_perc, "read", "90/10 achievement gap", "Percentage change from 2000 to 2015")
-
 @

### 2017-08-18 paper/paper.Rnw
Added 80/20 and 70/30 graphs at the end - still need to check if they work

@@
 @
-\caption{Time trends of achievement gap and average achiement between 2000 and 2015}
+\caption{Time trends of achievement gap and average achievement between 2000 and 2015}
 \end{center}
@@
 
-<<>>=
-results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
-results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
-results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
-results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
-results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
-results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
+<<estimating_othergaps>>=
+results_math_80 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.8, 0.2))
+results_read_80 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.8, 0.2))
+results_math_topmid_80 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.8, 0.5))
+results_read_topmid_80 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.8, 0.5))
+results_math_midbottom_80 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.2))
+results_read_midbottom_80 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.2))
+
+results_math_70 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.70, 0.30))
+results_read_70 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.70, 0.30))
+results_math_topmid_70 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.70, 0.5))
+results_read_topmid_70 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.70, 0.5))
+results_math_midbottom_70 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.30))
+results_read_midbottom_70 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.30))
 
@@
 
-@
+# results_math <- read_rds("./data/delete.Rdata")
+# results_read <- read_rds("./data/delete_read.Rdata")
+# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
+# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
+# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
+# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
+
+@
+
+\begin{figure}
+\begin{center}
+<<graph_all_gaps>>=
+complete_data_topbottom_80 <-
+  pisa_preparer(results_math_80, results_read_80) %>%
+  mutate(type = "80th/20th SES gap")
+
+complete_data_topmid_80 <-
+  pisa_preparer(results_math_topmid_80, results_read_topmid_80) %>%
+  mutate(type = "80th/50th SES gap")
+
+complete_data_midbottom_80 <-
+  pisa_preparer(results_math_midbottom_80, results_read_midbottom_80) %>%
+  mutate(type = "50th/20th SES gap")
+
+######
+
+complete_data_topbottom_70 <-
+  pisa_preparer(results_math_70, results_read_70) %>%
+  mutate(type = "70th/30th SES gap")
+
+complete_data_topmid_70 <-
+  pisa_preparer(results_math_topmid_70, results_read_topmid_70) %>%
+  mutate(type = "70th/50th SES gap")
+
+complete_data_midbottom_70 <-
+  pisa_preparer(results_math_midbottom_70, results_read_midbottom_70) %>%
+  mutate(type = "50th/30th SES gap")
+
+#####
+
+complete_gaps <-
+  complete_data_topbottom %>%
+  bind_rows(complete_data_topbottom_80)
+  bind_rows(complete_data_topbottom_70)
+
+complete_gaps %>%
+  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
+  filter(type_test == "math",
+         country %in% countries) %>%
+  ggplot(aes(as.character(wave), difference, colour = type, linetype = type, group = type)) +
+  geom_line(stat = "smooth", method = "lm", size = 0.7) +
+  facet_wrap(~ country, ncol = 5) +
+  scale_colour_discrete(name = "Type of gaps") +
+  scale_linetype_discrete(name = "Type of gaps") +
+  scale_y_continuous(name = "Gaps in SD", expand = c(0, 0), lim = c(0, 3)) +
+  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
+  ggthemes::theme_few() +
+  theme(panel.spacing = unit(1, "lines"),
+        panel.grid.major.y = element_line(colour = "grey"),
+        legend.title = ) +
+  ggtitle("Evolution of the different achievement gaps")
+@
+\caption{Evolution of the achievement gap for several gaps}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<percentage_change_8020>>=
+perc_increase_fun(complete_data_topbottom_80) %>%
+  perc_graph("math",
+             "80/20 achievement gap",
+             "Percentage change from 2000 to 2015")
+@
+\caption{Percentage change in the 80/20 achievement gap from 2000 to 2015}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<percentage_change_7030>>=
+perc_increase_fun(complete_data_topbottom_70) %>%
+  perc_graph("math",
+             "70/30 achievement gap",
+             "Percentage change from 2000 to 2015")
+@
+\caption{Percentage change in the 70/30 achievement gap from 2000 to 2015}
+\end{center}
+\end{figure}
 

### 2017-08-19 paper/paper.Rnw
Added 80/20-70/30 gaps in appendix - all working when restarted Rstudio

@@
   library(matrixStats)
+  library(tidyverse)
   library(knitr)
@@
   library(modelr)
-  library(tidyverse)
   library(ggrepel)
@@
 \begin{equation}
-\gamma_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
 \gamma_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
@@
 
+<<next_steps>>=
+# Next steps:
+
+# Continue by doing the multilevel models to see what explains what. Include
+# all indicators from the reardon/russian girl paper.
+
+# Graph the increase in each country vs the increase/decrease of the economic inequality indicators
+# Specially the 90/10
+
+# Calculate how big is the gap between reading and math
+
+# Continue with the PIRLS to see if there are specific patterns in 4th and 8th graders gap.
+
+# Get each country trendline adjusted for the inequality indicators and place in the same country graph.
+
+# Should I add the parent's education in the lm model to see how trends change adjusted for that?
+
+# A weak welfare system, together with income inequality, what's their pattern?
+# What if we put the school differentiation/tracking aspect in? Are there country groups based on these
+# patterns.
+
+# In countries where there is high differentiation/tracking, is there a jump in the evolution of the gap between PIRLS/TIMSS and PISA?
+@
+
+\bibliography{mybibliography.bib}
+
+\pagebreak
+
 \section{Appendix}
@@
 
+write_rds(results_math_80, "./data/math_80")
+write_rds(results_read_80, "./data/read_80")
+write_rds(results_math_topmid_80, "./data/math_topmid_80")
+write_rds(results_read_topmid_80, "./data/read_topmid_80")
+write_rds(results_math_midbottom_80, "./data/math_midbottom_80")
+write_rds(results_read_midbottom_80, "./data/read_midbottom_80")
+
+write_rds(results_math_70, "./data/math_70")
+write_rds(results_read_70, "./data/read_70")
+write_rds(results_math_topmid_70, "./data/math_topmid_70")
+write_rds(results_read_topmid_70, "./data/read_topmid_70")
+write_rds(results_math_midbottom_70, "./data/math_midbottom_70")
+write_rds(results_read_midbottom_70, "./data/read_midbottom_70")
+
 # results_math <- read_rds("./data/delete.Rdata")
@@
 
+<<eval = F>>=
+results_math_80 <- read_rds("./paper/data/math_80.Rdata")
+results_read_80 <- read_rds("./paper/data/read_80.Rdata")
+results_math_topmid_80 <- read_rds("./paper/data/math_topmid_80.Rdata")
+results_read_topmid_80 <- read_rds("./paper/data/read_topmid_80.Rdata")
+results_math_midbottom_80 <- read_rds("./paper/data/math_midbottom_80.Rdata")
+results_read_midbottom_80 <- read_rds("./paper/data/read_midbottom_80.Rdata")
+
+results_math_70 <- read_rds("./paper/data/math_70.Rdata")
+results_read_70 <- read_rds("./paper/data/read_70.Rdata")
+results_math_topmid_70 <- read_rds("./paper/data/math_topmid_70.Rdata")
+results_read_topmid_70 <- read_rds("./paper/data/read_topmid_70.Rdata")
+results_math_midbottom_70 <- read_rds("./paper/data/math_midbottom_70.Rdata")
+results_read_midbottom_70 <- read_rds("./paper/data/read_midbottom_70.Rdata")
+@
+
 \begin{figure}
 \begin{center}
-<<graph_all_gaps>>=
+<<graph_all_gaps,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
 complete_data_topbottom_80 <-
@@
   complete_data_topbottom %>%
-  bind_rows(complete_data_topbottom_80)
+  bind_rows(complete_data_topbottom_80) %>%
   bind_rows(complete_data_topbottom_70)
@@
         panel.grid.major.y = element_line(colour = "grey"),
-        legend.title = ) +
+        legend.position = "bottom") +
   ggtitle("Evolution of the different achievement gaps")
@@
 \begin{center}
-<<percentage_change_8020>>=
+<<percentage_change_8020,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
 perc_increase_fun(complete_data_topbottom_80) %>%
@@
 \begin{center}
-<<percentage_change_7030>>=
+<<percentage_change_7030,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
 perc_increase_fun(complete_data_topbottom_70) %>%
@@
 
-<<next_steps>>=
-# Next steps:
-
-# Continue by doing the multilevel models to see what explains what. Include
-# all indicators from the reardon/russian girl paper.
-
-# Graph the increase in each country vs the increase/decrease of the economic inequality indicators
-# Specially the 90/10
-
-# Calculate how big is the gap between reading and math
-
-# Continue with the PIRLS to see if there are specific patterns in 4th and 8th graders gap.
-
-# Get each country trendline adjusted for the inequality indicators and place in the same country graph.
-
-# Should I add the parent's education in the lm model to see how trends change adjusted for that?
-
-# A weak welfare system, together with income inequality, what's their pattern?
-# What if we put the school differentiation/tracking aspect in? Are there country groups based on these
-# patterns.
-
-# In countries where there is high differentiation/tracking, is there a jump in the evolution of the gap between PIRLS/TIMSS and PISA?
-@
-
-\bibliography{mybibliography.bib}
-
 \end{document}

### 2017-08-19 paper/paper.tex
Updated README with step-by-step of repo + deleted all useless scripts from before
Added 80/20-70/30 gaps in appendix - all working when restarted Rstudio

@@
+% Things to remember:
+% The working directory is not where the data is.
+% Change working directory to "/Users/cimentadaj/Downloads/gitrepo/Achievement_Gap_PISA/paper"
+% for everything below to work.
+
+% There's a code chunk named modeling which lasts about 3-4 hours to run. I explain how to get
+% around that in a comment above the code chunk. For down there.
+
 \documentclass[11pt, a4paper]{article}\usepackage[]{graphicx}\usepackage[]{color}
@@
 
-% This chunk needs to have cache = TRUE once you want to run it for like 2 hours until 
-% the results_* models have been created once and the cache can save them
+% pisa_listcol.Rdata is a dataset produced by the source() expression from the above
+% code chunk
+
+% This chunk takes about 3-4 hours to run. Once I had them ready I saved them so you can read them
+% as rds. If you run the 'compile pdf' this chunk is already cached (DONT CHANGE ANYTHING FROM
+% THIS CHUNK, CAUSE OTHERWISE IT WILL RERUN FOR 3-4 HOURS). If you want to run the code without
+% compiling sweave, comment the test_diff expressions and read as rds.
+
+% Remember to uncomment when compiling sweave again, so the chunk remains intact.
 
@@
 
-The work of James Heckman, for example, has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
-This means that inequality starts strong before the schooling of a child begins and schools can do so much to close it.
+Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive and personality definition \citep{duyme1999, waldfogel1996}.
+Economists and educators have long neglected this branch of literature but the work of James Heckman has brought our attention to it.
+For example, he has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
+This means that inequality starts strong before the schooling of a child begins and schools have the potential to equalize the learning ground.
 These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions.
@@
 
+The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years.
+The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence \citep{anna2016_global} is emerging with a very similar landscape.
+The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940.
+This time-trend has allowed researchers to study achievement gaps in a lengthy period of time.
+Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
+Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}.
+The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
+
 Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation.
+\citet{reardon2011} offers several possible links, with the most reasonable being that family investment patterns have changed so that high income families now invest more resources on their children.
+This explanation lies in the fact that increasing income became more strongly correlated with other positive family traits related to time allocation and welfare services.
+
 However likely, this is a highly contested topic.
 A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality.
-However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008}.
-
-\footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970.}.
+However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970 where income inequality was increasing.}.
+.
 
@@
 They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
+Their conclusions are very speculative and have no empirical support, which is why this is still an open question.
+The reasons the authors find a reversal in the trend could be numerous and should be studied very closely.
+The authors incorporate a number of country-level indicators to try explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment.
+In fact, the point out that in this same period (1998 - 2010) the SES gap in preschool enrollment decreased substantially.
 Their conclusions are very speculative and have no empirical support, which is why this is still an open question.
@@
 
-More formally, We're interested in studying
+More formally, we're interested in studying
 a) Which countries are experiencing changes in the achievement gap;
@@
 This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.
+a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
+After recording a SES gap increase of about 40\% in only 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
+This shows that it is important to study the changes in the achievement gap over time.
+This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.
 
@@
 This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming all mentioned variables into comparable metrics across waves.
+
+The ESCS was derived from a principal component analysis of standardised variables, taking the factor scores for the first principal component as measures of the PISA index of economic, social and cultural status.
+All countries and economies (both OECD and partner countries/economies) contributed equally to the principal component analysis, while in previous cycles, the principal component analysis was based on OECD countries only.
+However, for the purpose of reporting the ESCS scale has been transformed with zero being the score of an average OECD student and one being the standard deviation across equally weighted OECD countries \citep{pisa_2015_results}.
+
 To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.
@@
 
-for each wave,  where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (because all children should be 15 year olds) weighted by the student sample weights.
+for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011}) weighted by the student sample weights.
 
-We then adjust \begin{math} \hat{Y_i} \end{math} by
+We then adjust \begin{math} \hat{\gamma_i} \end{math} by
 
@@
 \begin{equation}
-\hat{Y_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
+\hat{\gamma_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
 \end{equation}
@@
 
-% Show graph of variance over time
-% I'm showing standard deviations, is it better to show variance?
-
+% Try plotting this for each country and wave
+% Like, every country as a row and 6 columns for each wave.
 \begin{figure}
@@
 
-In order to correct gap estimates for measurement error, I adjust student test scores(\begin{math} \hat{Y_i} \end{math}) by
+In order to correct gap estimates for measurement error, I adjust student test scores(\begin{math} \hat{\gamma_i} \end{math}) by
 
 \begin{equation}
-\hat{Y_i} = \hat{Y_i} * \frac{1}{\sqrt{r}}
+\hat{\gamma_i} = \hat{\gamma_i} * \frac{1}{\sqrt{r}}
 \end{equation}
@@
 
+% Replace the Y_i with the chosen y from the previos equation.
 \begin{equation}
-Y_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{within country} \ j
+\gamma_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
 \end{equation}
+
+% Try running also a linear model together with the multilevel model and ompcare results
+% Brunno suggested this because countries which might have a very different SES estimate
+% might be moved more towards the average.
 
@@
 We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size.
-Given that including the SES dummy reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible.
+Given that including the SES \emph{dummy} reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible.
 For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.
+% Explain more in detail that it's a dummy and not three groups.
 
@@
 \subsection{Descriptives}
+
+% Previous table that had % of low and high education for low and high SES dummies
 
@@
 
+% New table only has sample size, average score and s.e.
 % latex table generated in R 3.4.0 by xtable 1.8-2 package
@@
 
+% Code chunks for the same plots as above for reading. Available upon request.
+
 Something reassuring is that mathematics and reading (not presented) follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase.
@@
 
-% How to output data frames are latex tables?
-
 \begin{figure}
@@
 
-Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's an efficiency/equality tradeoff and if it stands over time.
-One particular topic that has been researched in the past is whether improving the overall cognitive development of a country compromises the distribution of the countries overall performance.
+Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's a tradeoff between the average country performance and the widening of the achievement gap and if it stands over time.
+One particular topic that has been researched in the past is whether improving the overall performance of a country compromises the equality of the educational system.
 For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance.
@@
 
-\subsection{Efficiency-equality tradeoff}
+\subsection{Performance-equality tradeoff}
 
@@
 
-\caption{Time trends of achievement gap and average achiement between 2000 and 2015}
+\caption{Time trends of achievement gap and average achievement between 2000 and 2015}
 \end{center}
 \end{figure}
+
+% Add descriptive table with average test scores, degree of tracking, level of income inequality, among other things.
 
@@
 
+\pagebreak
+
+\section{Appendix}
+
+\begin{figure}
+\begin{center}
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/graph_all_gaps-1}
+
+}
+
+\caption{Evolution of the achievement gap for several gaps}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/percentage_change_8020-1}
+
+}
+
+\caption{Percentage change in the 80/20 achievement gap from 2000 to 2015}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/percentage_change_7030-1}
+
+}
+
+\caption{Percentage change in the 70/30 achievement gap from 2000 to 2015}
+\end{center}
+\end{figure}
+
 \end{document}

### 2017-08-22 paper/paper.Rnw
Explored tracking and build some nice models in the end

--- 2017-08-18 paper/paper.Rnw
@@
 <<default_conf>>=
-  library(matrixStats)
 library(knitr)
@@
 library(modelr)
-  library(tidyverse)
 library(ggrepel)
@@
 library(ggthemes)
+library(matrixStats)
+library(WDI)
+library(tidyverse)
 
@@
 
-results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
-results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
-results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
-results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
-results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
-results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
-
-# results_math <- read_rds("./data/delete.Rdata")
-# results_read <- read_rds("./data/delete_read.Rdata")
-# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
-# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
-# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
-# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
+# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.75))
+# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.75))
+# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.75))
+# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.75))
+# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.5))
+# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.5))
+
+# results_math_75 <- results_math
+# results_read_75 <- results_read
+# results_math_topmid_75 <- results_math_topmid
+# results_read_topmid_75 <- results_read_topmid
+# results_math_midbottom_75 <- results_math_midbottom
+# results_read_midbottom_75 <- results_read_midbottom
+
+results_math <- read_rds("./paper/data/delete.Rdata")
+results_read <- read_rds("./paper/data/delete_read.Rdata")
+results_math_topmid <- read_rds("./paper/data/delete_math_topmid.Rdata")
+results_read_topmid <- read_rds("./paper/data/delete_read_topmid.Rdata")
+results_math_midbottom <- read_rds("./paper/data/delete_math_midbottom.Rdata")
+results_read_midbottom <- read_rds("./paper/data/delete_read_midbottom.Rdata")
 # US is missing for reading
@@
 <<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
-# 90/10 gaps acros countries
+# 90/10 gaps across countries
+
+df <- complete_data_topbottom
+
 diff_increase_fun <- function(df) {
@@
 
-pooled_trendline <-
+p <- 
   complete_data_topbottom %>%
   mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
-  filter(!is.na(country))
-
-pooled_trendline %>%
-  filter(type_test == "math") %>%
+  filter(type_test == "math", !is.na(country)) %>%
   ggplot(aes(as.character(wave), difference)) +
   geom_point() +
-  geom_linerange(aes(ymin = 0, ymax = difference))+
-  geom_line(data = pooled_trendline, stat = "smooth", method = "lm", aes(group = 1),
+  facet_wrap(~ country, ncol = 5) +
+  geom_linerange(aes(ymin = 0, ymax = difference))
+
+p + geom_line(data = complete_data_topbottom %>%
+                mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
+                filter(!is.na(country)),
+              stat = "smooth", method = "lm", aes(group = 1),
               formula = y ~ splines::ns(x, 2), size = 0.7,
               colour = "red") +
-  facet_wrap(~ country, ncol = 5) +
   scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
@@
   ggtitle("Evolution of the 90/10 achievement gap")
+
 @
@@
 \end{figure}
+
+<<>>=
+m <- complete_data_topbottom %>%
+  filter(type_test == "math") %>%
+  select(wave, country, difference) %>%
+  left_join(tracking_data, by = c("country" = "cntry_name")) %>%
+  left_join(central_examination) %>%
+  mutate(num_tracks = factor(tracks15y))
+
+m %>%
+  filter(country %in% countries) %>%
+  split(.$country) %>%
+  map(~ lm(difference ~ wave + ztrack, data = .x))
+
+p + geom_smooth(data = complete_data_topbottom %>%
+                mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
+                filter(!is.na(country)),
+                method = "lm", se = FALSE,
+                aes(group = 1), size = 0.7,
+                colour = "red") +
+  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
+  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
+  ggthemes::theme_few() +
+  theme(panel.spacing = unit(1, "lines"),
+        panel.grid.major.y = element_line(colour = "grey")) +
+  ggtitle("Evolution of the 90/10 achievement gap")
+@
 
@@
 # See below that reading graph is upon request. This is the one.
-
-pooled_trendline %>%
-  filter(type_test == "read") %>%
+complete_data_topbottom %>%
+  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
+  filter(type_test == "read", !is.na(country)) %>%
   ggplot(aes(as.character(wave), difference)) +
@@
   geom_linerange(aes(ymin = 0, ymax = difference))+
-  geom_line(data = pooled_trendline, stat = "smooth", method = "lm", aes(group = 1),
+  geom_line(stat = "smooth", method = "lm", aes(group = 1),
             formula = y ~ splines::ns(x, 2), size = 0.7,
@@
   perc_graph(top_bottom_perc, "read", "90/10 achievement gap", "Percentage change from 2000 to 2015")
+
 @
@@
 
-\section{Appendix}
+<<eval = F>>=
+
+# 2000
+# SC03Q01 School public/private
+# 1 Public
+# 2 Private
+# 7 N/A
+# 8 M/R
+# 9 Mis
+
+# 2003
+# SC03Q01 (8) Public or private
+# 1 Public
+# 2 Private
+# 7 N/A
+# 8 Invalid
+# 9 Miss
+
+# 2006
+# SC02Q01 (8) Public or private
+#  1 Public
+#  2 Private
+#  7 N/A
+#  8 Invalid
+#  9 Missing
+
+# 2009
+# SC02Q01
+# 1 public
+# 2 private
+
+# 2012
+# SC01Q01 Public or private Num
+# 1 Public
+# 2 Private
+# 7 N/A
+# 8 Invalid
+# 9 Missing
+
+# 2015
+# SC013Q01TA
+# 1 Public
+# 2 Private
+# 7 N/A
+# 8 Invalid
+# 9 Missing
+
+# 2000: ST25Q01
+# 1 <ISCED 2A>
+# 2 <ISCED 2B>
+# 3 <ISCED 2C>
+# 4 <ISCED 3A>
+# 5 <ISCED 3B>
+# 6 <ISCED 3C>
+# 7 N/A
+# 8 M/R
+# 9 Mis
+
+# 2003: ISCEDO
+# 2006: ISCEDO
+# 2009: ISCEDO
+# 2012: ISCEDO
+# 2015: ISCEDO
+
+# 1	General
+# 2	Pre-Vocational
+# 3	Vocational
+# 4	Modular
+
+library(readxl)
+
+wd_extract <- function(var, name) {
+  wdi_data <-
+    WDI(indicator = var, start = 2000, end = 2015) %>%
+    select(-1) %>%
+    as_tibble() %>%
+    mutate(year = as.character(year))
+
+  names(wdi_data)[2] <- name
+  wdi_data
+}
+
+pop_preprimary <- wd_extract("SP.PRE.TOTL.IN", "pop_preprimary")
+teacher_pupil_ratio <- wd_extract("SE.PRE.ENRL.TC.ZS", "teacher_pupil_ratio")
+perc_teacher_trained <- wd_extract("UIS.TRTP.0", "perc_teach")
+preprimary_priv_enrol <- wd_extract("SE.PRE.PRIV.ZS", "preprimary_priv_enrol")
+
+tracking_data <-
+  readxl::read_xlsx("./paper/data/tracking.xlsx", sheet = "all_data") %>%
+  map_if(is_double, round, 2) %>%
+  as_tibble
+
+difference_data <-
+  complete_data_topbottom %>%
+  filter(type_test == "math") %>%
+  select(wave, country, difference) %>%
+  rename(year = wave) %>%
+  mutate(year = as.character(year))
+
+library(inequalityintsvy)
+
+central_examination <-
+  tribble(
+  ~ country, ~central_examination,
+  "Australia", 1, 
+  "Austria", 0, 
+  "Belgium", 0,
+  "Bulgaria", 1,
+  "Canada", rbinom(1, 1, prob = 0.51),
+  "Czech Republic", 1,
+  "Denmark", 1,
+  "Finland", 1,
+  "France", 1,
+  "Germany", 0,
+  "United Kingdom", 1,
+  "Greece", 0,
+  "Hong Kong", 1,
+  "Hungary", 1,
+  "Iceland", 1,
+  "Ireland", 1,
+  "Israel", 1,
+  "Italy", 1,
+  "Japan", 1,
+  "Korea", 1,
+  "Latvia", 1,
+  "Liechtenstein", 1,
+  "Luxembourg", 1,
+  "Netherlands", 1,
+  "New Zealand", 1,
+  "Norway", 1,
+  "Poland", 1,
+  "Portugal", 0,
+  "Russia", 1,
+  "Slovakia", 1,
+  "Slovenia", 1,
+  "Spain", 0,
+  "Sweden", 0,
+  "Switzerland", 0,
+  "Turkey", 1,
+  "United States", 1
+  )
+
+# The tracking variables explain very nicely the difference in test scores
+# for all countries pooled, up to 50% R squared.
+
+ready_data <-
+  difference_data %>%
+  left_join(tracking_data, by = c("country" = "cntry_name")) %>%
+  left_join(central_examination) %>%
+  mutate(num_tracks = factor(tracks15y),
+         age_selection = ifelse(selage >= 15, 1, 0))
+
+mod_tracking <-
+  ready_data %>%
+  lm(difference ~
+       num_tracks +
+       age_selection +
+       length +
+       central_examination, data = .)
+
+new_data <-
+  data_grid(ready_data,
+            selage, num_tracks, central_examination, .model = mod_tracking) %>%
+  filter(complete.cases(.))
+
+new_data$pred <- predict(mod_tracking, newdata = new_data)
+
+new_data %>%
+  ggplot(aes(selage, pred, colour = num_tracks)) +
+  geom_point()
+
+new_data %>%
+  ggplot(aes(num_tracks, pred, fill = as.character(num_tracks))) +
+  geom_col(position = "dodge") +
+  facet_wrap(~ central_examination) +
+  scale_fill_fivethirtyeight()
+
+# What happened if you adjut the trend splines for the tracking data?
+@
 
 <<>>=
-results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
-results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
-results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
-results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
-results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
-results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
-
-# results_math <- read_rds("./data/delete.Rdata")
-# results_read <- read_rds("./data/delete_read.Rdata")
-# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
-# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
-# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
-# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
-
+
+# One idea I had was to make a cumulative sum of the trend,
+# so the cumulative inequality over time for each country
+# and see how each tracking feature explains this cumulative
+# pattern, and it does so very well, with each one
+# separately (because I have only 30 observations) explaining
+# about 22% each. Age selection, up to 30% alone.
+
+ready_data_two <-
+  difference_data %>%
+  left_join(tracking_data, by = c("country" = "cntry_name")) %>%
+  left_join(central_examination) %>%
+  mutate(num_tracks = factor(tracks15y),
+         age_selection = ifelse(selage >= 15, 1, 0))
+
+ready_data_two <-
+  ready_data_two %>%
+  group_by(country) %>%
+  summarize(
+    cum_diff = sum(difference, na.rm = TRUE),
+    length = unique(length),
+    num_tracks = unique(num_tracks),
+    age_selection = unique(age_selection),
+    central_examination = unique(central_examination),
+    avg_track = unique(ztrack)
+         )
+
+mod_tracking <-
+  ready_data_two %>%
+  lm(cum_diff ~ age_selection, data = .)
 @

### 2017-08-23 paper/paper.Rnw
Visualized both counterfactual difference under different tracking and cumulative change

@@
   left_join(central_examination) %>%
-  mutate(num_tracks = factor(tracks15y),
+  mutate(num_tracks = factor(car::recode(tracks15y, "5 = 4")),
          age_selection = ifelse(selage >= 15, 1, 0))
 
-mod_tracking <-
+library(rstanarm)
+
+mod_tracking <- stan_glmer(
+  difference ~
+    num_tracks +
+    age_selection + (1 | country),
+  family = gaussian(),
+  data = ready_data,
+  prior = normal(0, 2),
+  prior_intercept = normal(0, 5),
+  prior_aux = cauchy(0, 1))
+
+# mod_tracking <-
+#   ready_data %>%
+#   lmer(difference ~
+#        num_tracks +
+#        age_selection + (1 | country), data = .)
+
+# new_data <-
+#   data_grid(ready_data,
+#             selage, num_tracks, central_examination, .model = mod_tracking) %>%
+#   filter(complete.cases(.))
+
+new_data <-
   ready_data %>%
-  lm(difference ~
-       num_tracks +
-       age_selection +
-       length +
-       central_examination, data = .)
-
-new_data <-
-  data_grid(ready_data,
-            selage, num_tracks, central_examination, .model = mod_tracking) %>%
+  filter(!is.na(age_selection), !is.na(num_tracks)) %>%
+  expand(country, age_selection, num_tracks) %>%
   filter(complete.cases(.))
 
-new_data$pred <- predict(mod_tracking, newdata = new_data)
-
-new_data %>%
-  ggplot(aes(selage, pred, colour = num_tracks)) +
-  geom_point()
-
-new_data %>%
-  ggplot(aes(num_tracks, pred, fill = as.character(num_tracks))) +
-  geom_col(position = "dodge") +
-  facet_wrap(~ central_examination) +
-  scale_fill_fivethirtyeight()
+predictions <- posterior_predict(mod_tracking, newdata = new_data)
+sd_obs <- colSds(predictions)
+
+new_data$pred <- colMeans(predictions)
+new_data$pred_low <- new_data$pred - (2 * sd_obs)
+new_data$pred_high <- new_data$pred + (2 * sd_obs)
+
+sum_ready_data <-
+  ready_data %>%
+  group_by(country, num_tracks, age_selection) %>%
+  summarize(avg_diff = mean(difference),
+            diff_low = avg_diff - 2 * sd(difference),
+            diff_high = avg_diff + 2 * sd(difference)) %>%
+  ungroup()
+
+new_data %>% left_join(sum_ready_data) %>%
+  filter(country %in% countries) %>%
+  ggplot(aes(num_tracks %>% as.factor, avg_diff, fill = num_tracks)) +
+  geom_errorbar(aes(ymin = diff_low, ymax = diff_high), width = 0.2, colour = "red") +
+  geom_point(colour = "red") +
+  scale_y_continuous(limits = c(0, 4)) +
+  geom_col(aes(y = pred, alpha = 0.4)) +
+  geom_errorbar(aes(ymin = pred_low, ymax = pred_high), width = 0.2) +
+  facet_wrap(country ~ age_selection) +
+  scale_fill_wsj()
 
@@
 ready_data_two <-
-  difference_data %>%
-  left_join(tracking_data, by = c("country" = "cntry_name")) %>%
-  left_join(central_examination) %>%
-  mutate(num_tracks = factor(tracks15y),
-         age_selection = ifelse(selage >= 15, 1, 0))
-
-ready_data_two <-
-  ready_data_two %>%
+  ready_data %>%
   group_by(country) %>%
@@
 
+# mod_tracking <-
+#   ready_data_two %>%
+#   lm(cum_diff ~ age_selection, data = .)
+
 mod_tracking <-
   ready_data_two %>%
-  lm(cum_diff ~ age_selection, data = .)
+  filter(!is.na(age_selection)) %>%
+  stan_glm(cum_diff ~ age_selection, data = .,
+           family = "gaussian",
+           prior = normal(0, 2),
+           prior_intercept = normal(0, 5),
+           prior_aux = cauchy(0, 1),
+           refresh = -1)
+
+ready_data_two %>%
+  filter(!is.na(age_selection)) %>%
+  mutate(pred_mean = posterior_predict(mod_tracking) %>% colMeans(),
+         pred_median = posterior_predict(mod_tracking) %>% colMedians()) %>%
+  select(country, cum_diff, pred_mean, pred_median) %>%
+  gather(cat, value, -country) %>%
+  ggplot(aes(country, value, fill = cat)) +
+  geom_col(position = "dodge") +
+  coord_flip()
+
 @

### 2017-08-24 paper/paper.Rnw
Added a bunch of posterior predictive checks for the models

@@
 library(WDI)
+library(rstanarm)
 library(tidyverse)
@@
   left_join(central_examination) %>%
-  mutate(num_tracks = factor(car::recode(tracks15y, "5 = 4")),
-         age_selection = ifelse(selage >= 15, 1, 0))
-
-library(rstanarm)
-
-mod_tracking <- stan_glmer(
+  mutate(num_tracks = ifelse(tracks15y == 1, 1, 0) %>% as.factor,
+         age_selection = ifelse(selage >= 15, 1, 0)) %>%
+  filter(!is.na(num_tracks), !is.na(age_selection), !is.na(difference))
+
+library(brms)
+
+mod_tracking <-
+  brms::brm(
   difference ~
@@
   data = ready_data,
-  prior = normal(0, 2),
-  prior_intercept = normal(0, 5),
-  prior_aux = cauchy(0, 1))
-
-# mod_tracking <-
-#   ready_data %>%
-#   lmer(difference ~
-#        num_tracks +
-#        age_selection + (1 | country), data = .)
-
-# new_data <-
-#   data_grid(ready_data,
-#             selage, num_tracks, central_examination, .model = mod_tracking) %>%
-#   filter(complete.cases(.))
+  prior = c(set_prior("cauchy(2, 0.5)", coef = "num_tracks1"),
+            set_prior("cauchy(13, 1)", coef = "age_selection")
+            ),
+  warmup = 1000, iter = 2000, chains = 5
+  )
+
+# It under predicts the y value rather than over predict
+# It has problems predicting the central tendency of the model
 
@@
   ready_data %>%
-  filter(!is.na(age_selection), !is.na(num_tracks)) %>%
+  filter(country %in% countries) %>%
   expand(country, age_selection, num_tracks) %>%
@@
 
-new_data$pred <- colMeans(predictions)
-new_data$pred_low <- new_data$pred - (2 * sd_obs)
-new_data$pred_high <- new_data$pred + (2 * sd_obs)
+new_data <-
+  new_data %>%
+  mutate(
+    pred = colMeans(predictions),
+    pred_low = pred - (2 * sd_obs),
+    pred_high = pred + (2 * sd_obs)
+  )
 
@@
 
-new_data %>% left_join(sum_ready_data) %>%
+new_data %>%
+  left_join(sum_ready_data) %>%
   filter(country %in% countries) %>%
@@
   scale_fill_wsj()
+
+y_rep <- posterior_predict(mod_tracking)
+
+replication <-
+  map(1:2000, ~ {
+  ready_data %>%
+  mutate(pred_sample = y_rep[sample(1:4000, 1), , drop = TRUE]) %>%
+  summarize(perc_sample = mean(pred_sample > difference),
+            minus_sample = mean(pred_sample - difference))
+})
+
+replication %>%
+  enframe() %>%
+  unnest() %>%
+  qplot(minus_sample, geom = "histogram", data = .)
+
+replication %>%
+  enframe() %>%
+  unnest() %>%
+  summarize(mean(between(minus_sample, -0.05, 0.05)))
+
+pp_check(mod_tracking, nsamples = 100)
+
+pp_check(mod_tracking, type = "error_scatter", nsamples = 100)
 

### 2017-08-25 paper/paper.Rnw
Added posterior checks for cumulative difference model - model sucks but at least we know it

@@
   unnest() %>%
-  qplot(minus_sample, geom = "histogram", data = .)
+  qplot(perc_sample, geom = "histogram", data = .)
 
@@
   unnest() %>%
-  summarize(mean(between(minus_sample, -0.05, 0.05)))
+  summarize(mean(perc_sample < 0.5))
 
@@
 
-# What happened if you adjut the trend splines for the tracking data?
 @
@@
 
+vars_unique <- c("length", "num_tracks", "age_selection", "central_examination", "ztrack")
+
 ready_data_two <-
@@
   group_by(country) %>%
-  summarize(
-    cum_diff = sum(difference, na.rm = TRUE),
-    length = unique(length),
-    num_tracks = unique(num_tracks),
-    age_selection = unique(age_selection),
-    central_examination = unique(central_examination),
-    avg_track = unique(ztrack)
-         )
-
-# mod_tracking <-
-#   ready_data_two %>%
-#   lm(cum_diff ~ age_selection, data = .)
+  summarize_at(vars(vars_unique), unique) %>%
+  mutate(cum_diff = ready_data %>% group_by(country) %>% summarize(m = sum(difference)) %>% pull(m))
+
+mod_tracking <-
+  ready_data_two %>%
+  lm(cum_diff ~ num_tracks + age_selection, data = .) %>%
+  arm::display()
 
@@
   filter(!is.na(age_selection)) %>%
-  stan_glm(cum_diff ~ age_selection, data = .,
-           family = "gaussian",
-           prior = normal(0, 2),
-           prior_intercept = normal(0, 5),
-           prior_aux = cauchy(0, 1),
-           refresh = -1)
+  brm(
+    cum_diff ~ age_selection,
+    family = gaussian(),
+    data = .,
+    prior = set_prior("cauchy(13, 1)", coef = "age_selection"),
+    warmup = 1000, iter = 2000, chains = 5
+  )
+
+pp_check(mod_tracking, nsamples = 100)
+
+y_rep <- posterior_predict(mod_tracking)
+
+replication <-
+  map(1:2000, ~ {
+  ready_data_two %>%
+  mutate(pred_sample = y_rep[sample(1:4000, 1), , drop = TRUE]) %>%
+  summarize(perc_sample = mean(pred_sample > cum_diff),
+            minus_sample = mean(pred_sample - cum_diff))
+})
+
+y_pred_data <-
+  replication %>%
+  enframe() %>%
+  unnest()
+
+y_pred_data %>% qplot(perc_sample, data = ., geom = "histogram", bins = 50)
+
+y_pred_data %>% {quantile(.$perc_sample)}
+
+y_pred_data %>%
+  summarize(mean_perc = mean(perc_sample < 0.5))
+
+bayes_R2(mod_tracking)
+
+# Too few data points to predict integers, the distribution is full of empty spaces
+# The predictions are too wide! Only 41% of the predictions are between 0.45 and 0.55
+# accuracy. There's more underprediction than over prediction
 
@@
   filter(!is.na(age_selection)) %>%
-  mutate(pred_mean = posterior_predict(mod_tracking) %>% colMeans(),
-         pred_median = posterior_predict(mod_tracking) %>% colMedians()) %>%
-  select(country, cum_diff, pred_mean, pred_median) %>%
+  mutate(pred_median = posterior_predict(mod_tracking) %>% colMedians()) %>%
+  select(country, cum_diff, pred_median) %>%
   gather(cat, value, -country) %>%
@@
   coord_flip()
-
 @

### 2017-08-28 paper/paper.Rnw
Two models are complete - next writeup

@@
 
-# 2000: ST25Q01
-# 1 <ISCED 2A>
-# 2 <ISCED 2B>
-# 3 <ISCED 2C>
-# 4 <ISCED 3A>
-# 5 <ISCED 3B>
-# 6 <ISCED 3C>
-# 7 N/A
-# 8 M/R
-# 9 Mis
-
-# 2003: ISCEDO
-# 2006: ISCEDO
-# 2009: ISCEDO
-# 2012: ISCEDO
-# 2015: ISCEDO
-
-# 1	General
-# 2	Pre-Vocational
-# 3	Vocational
-# 4	Modular
-
-library(readxl)
-
-wd_extract <- function(var, name) {
-  wdi_data <-
-    WDI(indicator = var, start = 2000, end = 2015) %>%
-    select(-1) %>%
-    as_tibble() %>%
-    mutate(year = as.character(year))
-
-  names(wdi_data)[2] <- name
-  wdi_data
-}
-
-pop_preprimary <- wd_extract("SP.PRE.TOTL.IN", "pop_preprimary")
-teacher_pupil_ratio <- wd_extract("SE.PRE.ENRL.TC.ZS", "teacher_pupil_ratio")
-perc_teacher_trained <- wd_extract("UIS.TRTP.0", "perc_teach")
-preprimary_priv_enrol <- wd_extract("SE.PRE.PRIV.ZS", "preprimary_priv_enrol")
-
 tracking_data <-
-  readxl::read_xlsx("./paper/data/tracking.xlsx", sheet = "all_data") %>%
+  read_xlsx("./paper/data/tracking.xlsx", sheet = "all_data") %>%
   map_if(is_double, round, 2) %>%
@@
   left_join(central_examination) %>%
+  left_join(school_data_ready) %>%
   mutate(num_tracks = ifelse(tracks15y == 1, 1, 0) %>% as.factor,
-         age_selection = ifelse(selage >= 15, 1, 0)) %>%
+         age_selection = selage) %>%
   filter(!is.na(num_tracks), !is.na(age_selection), !is.na(difference))
+
+ready_data_age <- mutate(ready_data, age_selection = ifelse(selage >= 15, 1, 0))
 
@@
   family = gaussian(),
-  data = ready_data,
+  data = ready_data_age,
   prior = c(set_prior("cauchy(2, 0.5)", coef = "num_tracks1"),
@@
 new_data <-
-  ready_data %>%
+  ready_data_age %>%
   filter(country %in% countries) %>%
@@
 sum_ready_data <-
-  ready_data %>%
+  ready_data_age %>%
   group_by(country, num_tracks, age_selection) %>%
@@
 
-pp_check(mod_tracking, nsamples = 100)
-
-pp_check(mod_tracking, type = "error_scatter", nsamples = 100)
-
+pp_check(mod_tracking, type = "error_binned", nsamples = 1000)
+# The models tends to overpredict high values and underpredict low values.
 @
@@
   ready_data_two %>%
-  lm(cum_diff ~ num_tracks + age_selection, data = .) %>%
-  arm::display()
+  lm(cum_diff ~ age_selection, data = .) %>%
+  arm::display(detail = T)
 
@@
   )
-
-pp_check(mod_tracking, nsamples = 100)
+pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
+pp_check(mod_tracking, type = "ecdf_overlay", nsamples = 10)
+pp_check(mod_tracking, type = "boxplot", nsamples = 30)
+pp_check(mod_tracking, type = "stat", nsamples = 200) # very nice to compare central tendency
+
+pp_check(mod_tracking, type = "error_scatter_avg", nsamples = 1000)
+
+pp_check(mod_tracking, type = "loo_pit")
+# over prediction
+# central tendency is not captured very well
+# more under-prediction than over prediction
 

### 2017-08-28 paper/paper.Rnw
included bayesian modelling - new specification of models - final model

@@
 <<default_conf>>=
+  library(matrixStats)
+  library(tidyverse)
   library(knitr)
@@
   library(ggthemes)
-library(matrixStats)
-library(WDI)
-library(rstanarm)
-library(tidyverse)
 
@@
     pisa_all2$value[[6]] %>%
-    rename(escs_trend = escs_trend)
+    rename(escs_trend = ESCS)
 @
@@
 
-# results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.75))
-# results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.75))
-# results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.75))
-# results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.75))
-# results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.25, 0.5))
-# results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.25, 0.5))
-
-# results_math_75 <- results_math
-# results_read_75 <- results_read
-# results_math_topmid_75 <- results_math_topmid
-# results_read_topmid_75 <- results_read_topmid
-# results_math_midbottom_75 <- results_math_midbottom
-# results_read_midbottom_75 <- results_read_midbottom
-
-results_math <- read_rds("./paper/data/delete.Rdata")
-results_read <- read_rds("./paper/data/delete_read.Rdata")
-results_math_topmid <- read_rds("./paper/data/delete_math_topmid.Rdata")
-results_read_topmid <- read_rds("./paper/data/delete_read_topmid.Rdata")
-results_math_midbottom <- read_rds("./paper/data/delete_math_midbottom.Rdata")
-results_read_midbottom <- read_rds("./paper/data/delete_read_midbottom.Rdata")
+results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
+results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
+results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
+results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
+results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
+results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
+
+# results_math <- read_rds("./data/delete.Rdata")
+# results_read <- read_rds("./data/delete_read.Rdata")
+# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
+# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
+# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
+# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
 # US is missing for reading
@@
 
-escs_data <- escs_dummy_creator(adapted_year_data, c(0.25, 0.75))
+escs_data <- escs_dummy_creator(adapted_year_data, c(0.1, 0.9))
 
@@
 
-complete_data_topbottom <- mutate(complete_data_topbottom, type = "75th/25th SES gap")
-complete_data_topmid <- mutate(complete_data_topmid, type = "75th/50th SES gap")
-complete_data_midbottom <- mutate(complete_data_midbottom, type = "50th/25th SES gap")
+complete_data_topbottom <- mutate(complete_data_topbottom, type = "90th/10th SES gap")
+complete_data_topmid <- mutate(complete_data_topmid, type = "90th/50th SES gap")
+complete_data_midbottom <- mutate(complete_data_midbottom, type = "50th/10th SES gap")
 @
@@
 <<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
-# 90/10 gaps across countries
-
-df <- complete_data_topbottom
-
+# 90/10 gaps acros countries
 diff_increase_fun <- function(df) {
@@
 
-p <- 
+pooled_trendline <-
   complete_data_topbottom %>%
   mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
-  filter(type_test == "math", !is.na(country)) %>%
+  filter(!is.na(country))
+
+pooled_trendline %>%
+  filter(type_test == "math") %>%
   ggplot(aes(as.character(wave), difference)) +
   geom_point() +
-  facet_wrap(~ country, ncol = 5) +
-  geom_linerange(aes(ymin = 0, ymax = difference))
-
-p + geom_line(data = complete_data_topbottom %>%
-                mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
-                filter(!is.na(country)),
-              stat = "smooth", method = "lm", aes(group = 1),
+  geom_linerange(aes(ymin = 0, ymax = difference))+
+  geom_line(data = pooled_trendline, stat = "smooth", method = "lm", aes(group = 1),
             formula = y ~ splines::ns(x, 2), size = 0.7,
             colour = "red") +
+  facet_wrap(~ country, ncol = 5) +
   scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
@@
   ggtitle("Evolution of the 90/10 achievement gap")
-
 @
@@
 \end{figure}
-
-<<>>=
-m <- complete_data_topbottom %>%
-  filter(type_test == "math") %>%
-  select(wave, country, difference) %>%
-  left_join(tracking_data, by = c("country" = "cntry_name")) %>%
-  left_join(central_examination) %>%
-  mutate(num_tracks = factor(tracks15y))
-
-m %>%
-  filter(country %in% countries) %>%
-  split(.$country) %>%
-  map(~ lm(difference ~ wave + ztrack, data = .x))
-
-p + geom_smooth(data = complete_data_topbottom %>%
-                mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
-                filter(!is.na(country)),
-                method = "lm", se = FALSE,
-                aes(group = 1), size = 0.7,
-                colour = "red") +
-  scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
-  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
-  ggthemes::theme_few() +
-  theme(panel.spacing = unit(1, "lines"),
-        panel.grid.major.y = element_line(colour = "grey")) +
-  ggtitle("Evolution of the 90/10 achievement gap")
-@
 
@@
 us_math_2000 <-
-  filter(complete_data_topbottom,
-         wave == 2000, country == "United States",
-         type_test == "math") %>%
+  filter(complete_data_topbottom, wave == 2000, country == "United States", type_test == "math") %>%
   .[["difference"]]
@@
 us_math_2000_lower <-
-  filter(complete_data_topbottom,
-         wave == 2000, country == "United States",
-         type_test == "math") %>%
+  filter(complete_data_topbottom, wave == 2000, country == "United States", type_test == "math") %>%
   .[["lower"]]
@@
 us_math_2000_upper <-
-  filter(complete_data_topbottom,
-         wave == 2000, country == "United States",
-         type_test == "math") %>%
+  filter(complete_data_topbottom, wave == 2000, country == "United States", type_test == "math") %>%
   .[["upper"]]
@@
 
+<<next_steps>>=
+# Next steps:
+
+# Continue by doing the multilevel models to see what explains what. Include
+# all indicators from the reardon/russian girl paper.
+
+# Graph the increase in each country vs the increase/decrease of the economic inequality indicators
+# Specially the 90/10
+
+# Calculate how big is the gap between reading and math
+
+# Continue with the PIRLS to see if there are specific patterns in 4th and 8th graders gap.
+
+# Get each country trendline adjusted for the inequality indicators and place in the same country graph.
+
+# Should I add the parent's education in the lm model to see how trends change adjusted for that?
+
+# A weak welfare system, together with income inequality, what's their pattern?
+# What if we put the school differentiation/tracking aspect in? Are there country groups based on these
+# patterns.
+
+# In countries where there is high differentiation/tracking, is there a jump in the evolution of the gap between PIRLS/TIMSS and PISA?
+@
+
+\bibliography{mybibliography.bib}
+
+\pagebreak
+
+\section{Appendix}
+
+<<estimating_othergaps>>=
+results_math_80 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.8, 0.2))
+results_read_80 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.8, 0.2))
+results_math_topmid_80 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.8, 0.5))
+results_read_topmid_80 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.8, 0.5))
+results_math_midbottom_80 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.2))
+results_read_midbottom_80 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.2))
+
+results_math_70 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.70, 0.30))
+results_read_70 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.70, 0.30))
+results_math_topmid_70 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.70, 0.5))
+results_read_topmid_70 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.70, 0.5))
+results_math_midbottom_70 <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.30))
+results_read_midbottom_70 <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.30))
+
+write_rds(results_math_80, "./data/math_80")
+write_rds(results_read_80, "./data/read_80")
+write_rds(results_math_topmid_80, "./data/math_topmid_80")
+write_rds(results_read_topmid_80, "./data/read_topmid_80")
+write_rds(results_math_midbottom_80, "./data/math_midbottom_80")
+write_rds(results_read_midbottom_80, "./data/read_midbottom_80")
+
+write_rds(results_math_70, "./data/math_70")
+write_rds(results_read_70, "./data/read_70")
+write_rds(results_math_topmid_70, "./data/math_topmid_70")
+write_rds(results_read_topmid_70, "./data/read_topmid_70")
+write_rds(results_math_midbottom_70, "./data/math_midbottom_70")
+write_rds(results_read_midbottom_70, "./data/read_midbottom_70")
+
+# results_math <- read_rds("./data/delete.Rdata")
+# results_read <- read_rds("./data/delete_read.Rdata")
+# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
+# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
+# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
+# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
+
+# results_math <- read_rds("./data/delete.Rdata")
+# results_read <- read_rds("./data/delete_read.Rdata")
+# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
+# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
+# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
+# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
+
+@
+
+<<eval = F>>=
+results_math_80 <- read_rds("./paper/data/math_80.Rdata")
+results_read_80 <- read_rds("./paper/data/read_80.Rdata")
+results_math_topmid_80 <- read_rds("./paper/data/math_topmid_80.Rdata")
+results_read_topmid_80 <- read_rds("./paper/data/read_topmid_80.Rdata")
+results_math_midbottom_80 <- read_rds("./paper/data/math_midbottom_80.Rdata")
+results_read_midbottom_80 <- read_rds("./paper/data/read_midbottom_80.Rdata")
+
+results_math_70 <- read_rds("./paper/data/math_70.Rdata")
+results_read_70 <- read_rds("./paper/data/read_70.Rdata")
+results_math_topmid_70 <- read_rds("./paper/data/math_topmid_70.Rdata")
+results_read_topmid_70 <- read_rds("./paper/data/read_topmid_70.Rdata")
+results_math_midbottom_70 <- read_rds("./paper/data/math_midbottom_70.Rdata")
+results_read_midbottom_70 <- read_rds("./paper/data/read_midbottom_70.Rdata")
+@
+
+\begin{figure}
+\begin{center}
+<<graph_all_gaps,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+complete_data_topbottom_80 <-
+  pisa_preparer(results_math_80, results_read_80) %>%
+  mutate(type = "80th/20th SES gap")
+
+complete_data_topmid_80 <-
+  pisa_preparer(results_math_topmid_80, results_read_topmid_80) %>%
+  mutate(type = "80th/50th SES gap")
+
+complete_data_midbottom_80 <-
+  pisa_preparer(results_math_midbottom_80, results_read_midbottom_80) %>%
+  mutate(type = "50th/20th SES gap")
+
+######
+
+complete_data_topbottom_70 <-
+  pisa_preparer(results_math_70, results_read_70) %>%
+  mutate(type = "70th/30th SES gap")
+
+complete_data_topmid_70 <-
+  pisa_preparer(results_math_topmid_70, results_read_topmid_70) %>%
+  mutate(type = "70th/50th SES gap")
+
+complete_data_midbottom_70 <-
+  pisa_preparer(results_math_midbottom_70, results_read_midbottom_70) %>%
+  mutate(type = "50th/30th SES gap")
+
+#####
+
+complete_gaps <-
+  complete_data_topbottom %>%
+  bind_rows(complete_data_topbottom_80) %>%
+  bind_rows(complete_data_topbottom_70)
+
+complete_gaps %>%
+  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
+  filter(type_test == "math",
+         country %in% countries) %>%
+  ggplot(aes(as.character(wave), difference, colour = type, linetype = type, group = type)) +
+  geom_line(stat = "smooth", method = "lm", size = 0.7) +
+  facet_wrap(~ country, ncol = 5) +
+  scale_colour_discrete(name = "Type of gaps") +
+  scale_linetype_discrete(name = "Type of gaps") +
+  scale_y_continuous(name = "Gaps in SD", expand = c(0, 0), lim = c(0, 3)) +
+  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
+  ggthemes::theme_few() +
+  theme(panel.spacing = unit(1, "lines"),
+        panel.grid.major.y = element_line(colour = "grey"),
+        legend.position = "bottom") +
+  ggtitle("Evolution of the different achievement gaps")
+@
+\caption{Evolution of the achievement gap for several gaps}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<percentage_change_8020,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+perc_increase_fun(complete_data_topbottom_80) %>%
+  perc_graph("math",
+             "80/20 achievement gap",
+             "Percentage change from 2000 to 2015")
+@
+\caption{Percentage change in the 80/20 achievement gap from 2000 to 2015}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<percentage_change_7030,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+perc_increase_fun(complete_data_topbottom_70) %>%
+  perc_graph("math",
+             "70/30 achievement gap",
+             "Percentage change from 2000 to 2015")
+@
+\caption{Percentage change in the 70/30 achievement gap from 2000 to 2015}
+\end{center}
+\end{figure}
+
 <<eval = F>>=
@@
 # 9 Missing
+library(readxl)
 
@@
 difference_data <-
-  complete_data_topbottom %>%
-  filter(type_test == "math") %>%
+  complete_gaps %>%
+  filter(type_test == "math", type == "90th/10th SES gap") %>%
   select(wave, country, difference) %>%
@@
 
-ready_data_age <- mutate(ready_data, age_selection = ifelse(selage >= 15, 1, 0))
+ready_data_age <- mutate(ready_data, age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)
 
 library(brms)
-
+# This linear model explains the pooled 90/10 differences
 mod_tracking <-
@@
     difference ~
-    num_tracks +
-    age_selection + (1 | country),
+      ztrack +
+      zvoc +
+      ztrack:zvoc, family = gaussian(),
+  data = ready_data_age,
+  warmup = 1000, iter = 2000, chains = 5
+  )
+
+mod_tracking
+bayes_R2(mod_tracking)
+marginal_effects(mod_tracking)
+
+pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
+pp_check(mod_tracking, type = "scatter_avg", nsamples = 100)
+pp_check(mod_tracking, type = "boxplot", nsamples = 20)
+pp_check(mod_tracking, type = "error_hist", nsamples = 6)
+pp_check(mod_tracking, type = "ecdf_overlay")
+
+pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "zvoc")
+pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "ztrack")
+
+# Check whether the previous model can be generalized to a multilevel
+# and see how much it can explain between country var
+mod_tracking <-
+  brms::brm(
+    difference ~
+      ztrack +
+      zvoc +
+      ztrack:zvoc + (1 | country),
   family = gaussian(),
   data = ready_data_age,
-  prior = c(set_prior("cauchy(2, 0.5)", coef = "num_tracks1"),
-            set_prior("cauchy(13, 1)", coef = "age_selection")
-            ),
-  warmup = 1000, iter = 2000, chains = 5
+  prior = c(set_prior("normal(0, 1)", coef = "ztrack"),
+            set_prior("normal(0, 1)", coef = "zvoc")),
+  warmup = 1000, iter = 2000, chains = 4
   )
@@
 # It has problems predicting the central tendency of the model
+
+# The code bloew needs to be adapted to the model from above (
+# this is from previous models)
 
@@
 pp_check(mod_tracking, type = "error_binned", nsamples = 1000)
-# The models tends to overpredict high values and underpredict low values.
 @
@@
 
-vars_unique <- c("length", "num_tracks", "age_selection", "central_examination", "ztrack")
+vars_unique <- c("length", "num_tracks", "age_selection", "central_examination", "ztrack",
+                 "zvoc")
 
@@
   ready_data_two %>%
-  lm(cum_diff ~ age_selection, data = .) %>%
+  lm(cum_diff ~ ztrack + zvoc, data = .) %>%
   arm::display(detail = T)
@@
   brm(
-    cum_diff ~ age_selection,
+    cum_diff ~ ztrack + zvoc,
     family = gaussian(),
     data = .,
-    prior = set_prior("cauchy(13, 1)", coef = "age_selection"),
     warmup = 1000, iter = 2000, chains = 5
   )
+
+bayes_R2(mod_tracking)
+
 pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
@@
 
+# update all of this for the previous model!
 y_rep <- posterior_predict(mod_tracking)
@@
 
-\bibliography{mybibliography.bib}
-
 \end{document}

### 2017-08-28 paper/paper.Rnw
Add comment

@@
   coord_flip()
+
+# See how these models explain for the 90/80/70 gaps.
 @

### 2017-08-29 paper/paper.Rnw
Finshing last mode

@@
   library(matrixStats)
-  library(tidyverse)
   library(knitr)
@@
   library(ggthemes)
+  library(tidyverse)
 
@@
   left_join(central_examination) %>%
-  left_join(school_data_ready) %>%
   mutate(num_tracks = ifelse(tracks15y == 1, 1, 0) %>% as.factor,
@@
 library(brms)
+# It under predicts the y value rather than over predict
+# It has problems predicting the central tendency of the model
 # This linear model explains the pooled 90/10 differences
+
 mod_tracking <-
@@
 mod_tracking
-bayes_R2(mod_tracking)
-marginal_effects(mod_tracking)
 
@@
 
-# Check whether the previous model can be generalized to a multilevel
-# and see how much it can explain between country var
-mod_tracking <-
-  brms::brm(
-    difference ~
-      ztrack +
-      zvoc +
-      ztrack:zvoc + (1 | country),
-  family = gaussian(),
+bayes_R2(mod_tracking)
+marginal_effects(mod_tracking,
+                 effects = "ztrack:zvoc",
+                 int_conditions = list(zvoc = quantile))
+
+# I don't know how to estimate VIF's with bayesian reasoning
+model_formula <- difference ~ num_tracks + age_selection + zvoc + I(length > 0)
+mod_tracking_ext <-
+  brms::brm(model_formula, family = gaussian(),
   data = ready_data_age,
-  prior = c(set_prior("normal(0, 1)", coef = "ztrack"),
-            set_prior("normal(0, 1)", coef = "zvoc")),
-  warmup = 1000, iter = 2000, chains = 4
+  warmup = 1000, iter = 2000, chains = 5
   )
 
-# It under predicts the y value rather than over predict
-# It has problems predicting the central tendency of the model
-
-# The code bloew needs to be adapted to the model from above (
-# this is from previous models)
+bayes_R2(mod_tracking_ext)
+
+# This is the closest I could get to getting vif's:
+lm(model_formula, data = ready_data_age) %>%
+  vif()
 
@@
   ready_data_age %>%
-  filter(country %in% countries) %>%
-  expand(country, age_selection, num_tracks) %>%
-  filter(complete.cases(.))
-
-predictions <- posterior_predict(mod_tracking, newdata = new_data)
+  with(crossing(country = countries,
+                num_tracks = unique(num_tracks),
+                age_selection = unique(age_selection),
+                zvoc = mean(zvoc, na.rm = TRUE),
+                length = typical(length > 0)))
+
+predictions <- posterior_predict(mod_tracking_ext, newdata = new_data)
 sd_obs <- colSds(predictions)
@@
   left_join(sum_ready_data) %>%
-  filter(country %in% countries) %>%
   ggplot(aes(num_tracks %>% as.factor, avg_diff, fill = num_tracks)) +
@@
   scale_fill_wsj()
-
-y_rep <- posterior_predict(mod_tracking)
-
-replication <-
-  map(1:2000, ~ {
-  ready_data %>%
-  mutate(pred_sample = y_rep[sample(1:4000, 1), , drop = TRUE]) %>%
-  summarize(perc_sample = mean(pred_sample > difference),
-            minus_sample = mean(pred_sample - difference))
-})
-
-replication %>%
-  enframe() %>%
-  unnest() %>%
-  qplot(perc_sample, geom = "histogram", data = .)
-
-replication %>%
-  enframe() %>%
-  unnest() %>%
-  summarize(mean(perc_sample < 0.5))
-
-pp_check(mod_tracking, type = "error_binned", nsamples = 1000)
 @
@@
   brm(
-    cum_diff ~ ztrack + zvoc,
+    cum_diff ~ I(age_selection >= 15),
     family = gaussian(),
@@
 
-bayes_R2(mod_tracking)
-
 # Too few data points to predict integers, the distribution is full of empty spaces
@@
 
+m_data <-
   ready_data_two %>%
   filter(!is.na(age_selection)) %>%
-  mutate(pred_median = posterior_predict(mod_tracking) %>% colMedians()) %>%
-  select(country, cum_diff, pred_median) %>%
-  gather(cat, value, -country) %>%
-  ggplot(aes(country, value, fill = cat)) +
-  geom_col(position = "dodge") +
+  mutate(pred_median = posterior_predict(mod_tracking) %>% colMeans(),
+         ratio_pred = pred_median/cum_diff) %>%
+  select(country, ratio_pred, cum_diff, pred_median)
+
+m_data %>%
+  ggplot(aes(ratio_pred)) +
+  geom_histogram(bins = 50)
+
+m_data %>%
+  filter(ratio_pred < 1) %>%
+  ggplot(aes(reorder(country, -ratio_pred), 1 - ratio_pred)) +
+  geom_col() +
+  scale_y_continuous(breaks = seq(0, 0.30, 0.05), limits = c(0, 0.3)) +
   coord_flip()
+
+# Greece, France would have a decrease of about 20-25% of their cumulative
+# inequality when adjusting for the age of selection.
+
+# There is both under and over prediction but the over predictions are very extreme
 

### 2017-08-30 paper/paper.Rnw
Created graph that shows all gaps and the modeling - two models left

@@
 
-difference_data <-
-  complete_gaps %>%
-  filter(type_test == "math", type == "90th/10th SES gap") %>%
-  select(wave, country, difference) %>%
-  rename(year = wave) %>%
-  mutate(year = as.character(year))
-
 library(inequalityintsvy)
@@
 
-ready_data <-
-  difference_data %>%
+dif_data <- function(my_data, tracking, central, which_gap = "90th/10th SES gap") {
+  my_data %>%
+    filter(type_test == "math", type == which_gap) %>%
+    select(wave, country, difference) %>%
+    rename(year = wave) %>%
+    mutate(year = as.character(year)) %>%
     left_join(tracking_data, by = c("country" = "cntry_name")) %>%
@@
 
-ready_data_age <- mutate(ready_data, age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)
-
-library(brms)
-# It under predicts the y value rather than over predict
-# It has problems predicting the central tendency of the model
-# This linear model explains the pooled 90/10 differences
+}
+
+ready_data <- dif_data(complete_gaps, tracking_data, central_examination)
+
+ready_data_age <-
+  dif_data(complete_gaps, tracking_data, central_examination) %>%
+  mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)
+
+gaps <- c("90th/10th SES gap", "80th/20th SES gap", "70th/30th SES gap")
+
+all_gaps_models <-
+  map(gaps, function(gap) {
+    ready_data_age <-
+      dif_data(complete_gaps, tracking_data, central_examination, which_gap = gap) %>%
+      mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)
 
@@
     mod_tracking
+})
+
+model_data_merged <-
+  map(all_gaps_models, ~ fixef(.x)[, c(1, 3, 4)] %>% as_tibble) %>%
+  setNames(gaps) %>%
+  enframe() %>%
+  unnest(value) %>%
+  mutate_if(is_double, round, 2) %>%
+  transmute(model = name,
+           terms = all_gaps_models[[1]] %>% fixef %>% row.names %>% rep(3),
+           estimate = Estimate,
+           lower = `2.5%ile`,
+           upper = `97.5%ile`)
+
+model_data_merged %>%
+  ggplot(aes(terms, estimate, colour = model)) +
+  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
+  geom_point() +
+  geom_hline(yintercept = 0) +
+  scale_colour_discrete(name = NULL) +
+  scale_y_continuous(limits = c(-0.5, 2)) +
+  ggtitle("Coef from linear model of inequality on degree of tracking and vocational tracking") +
+  coord_flip()
+
+library(brms)
+# It under predicts the y value rather than over predict
+# It has problems predicting the central tendency of the model
+# This linear model explains the pooled 90/10 differences
+
+mod_tracking <-
+  brms::brm(
+    difference ~
+      ztrack +
+      zvoc +
+      ztrack:zvoc, family = gaussian(),
+  data = ready_data_age,
+  warmup = 1000, iter = 2000, chains = 5
+  )
+
+bayes_R2(mod_tracking)
+
+plot_interaction <-
+  marginal_effects(mod_tracking,
+                 effects = "ztrack:zvoc",
+                 int_conditions = list(zvoc = quantile))
 
@@
 pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "ztrack")
-
-bayes_R2(mod_tracking)
-marginal_effects(mod_tracking,
-                 effects = "ztrack:zvoc",
-                 int_conditions = list(zvoc = quantile))
 
@@
 pp_check(mod_tracking, type = "stat", nsamples = 200) # very nice to compare central tendency
-
 pp_check(mod_tracking, type = "error_scatter_avg", nsamples = 1000)
@@
 # update all of this for the previous model!
-y_rep <- posterior_predict(mod_tracking)
-
-replication <-
-  map(1:2000, ~ {
-  ready_data_two %>%
-  mutate(pred_sample = y_rep[sample(1:4000, 1), , drop = TRUE]) %>%
-  summarize(perc_sample = mean(pred_sample > cum_diff),
-            minus_sample = mean(pred_sample - cum_diff))
-})
-
-y_pred_data <-
-  replication %>%
-  enframe() %>%
-  unnest()
-
-y_pred_data %>% qplot(perc_sample, data = ., geom = "histogram", bins = 50)
-
-y_pred_data %>% {quantile(.$perc_sample)}
-
-y_pred_data %>%
-  summarize(mean_perc = mean(perc_sample < 0.5))
+# y_rep <- posterior_predict(mod_tracking)
+# 
+# replication <-
+#   map(1:2000, ~ {
+#   ready_data_two %>%
+#   mutate(pred_sample = y_rep[sample(1:4000, 1), , drop = TRUE]) %>%
+#   summarize(perc_sample = mean(pred_sample > cum_diff),
+#             minus_sample = mean(pred_sample - cum_diff))
+# })
+# 
+# y_pred_data <-
+#   replication %>%
+#   enframe() %>%
+#   unnest()
+# 
+# y_pred_data %>% qplot(perc_sample, data = ., geom = "histogram", bins = 50)
+# 
+# y_pred_data %>% {quantile(.$perc_sample)}
+# 
+# y_pred_data %>%
+#   summarize(mean_perc = mean(perc_sample < 0.5))
 
@@
 m_data %>%
-  filter(ratio_pred < 1) %>%
   ggplot(aes(reorder(country, -ratio_pred), 1 - ratio_pred)) +
   geom_col() +
-  scale_y_continuous(breaks = seq(0, 0.30, 0.05), limits = c(0, 0.3)) +
+  scale_y_continuous(breaks = seq(-0.3, 0.3, 0.5), limits = c(-0.3, 0.3)) +
   coord_flip()

### 2017-09-04 paper/paper.Rnw
Produced first two tables from tracking modeling - everything works after restart

@@
 
-<<eval = F>>=
+<<first_model_table>>=
 
@@
 library(readxl)
+library(inequalityintsvy)
+library(brms)
 
@@
 
-library(inequalityintsvy)
-
-central_examination <-
-  tribble(
-  ~ country, ~central_examination,
-  "Australia", 1, 
-  "Austria", 0, 
-  "Belgium", 0,
-  "Bulgaria", 1,
-  "Canada", rbinom(1, 1, prob = 0.51),
-  "Czech Republic", 1,
-  "Denmark", 1,
-  "Finland", 1,
-  "France", 1,
-  "Germany", 0,
-  "United Kingdom", 1,
-  "Greece", 0,
-  "Hong Kong", 1,
-  "Hungary", 1,
-  "Iceland", 1,
-  "Ireland", 1,
-  "Israel", 1,
-  "Italy", 1,
-  "Japan", 1,
-  "Korea", 1,
-  "Latvia", 1,
-  "Liechtenstein", 1,
-  "Luxembourg", 1,
-  "Netherlands", 1,
-  "New Zealand", 1,
-  "Norway", 1,
-  "Poland", 1,
-  "Portugal", 0,
-  "Russia", 1,
-  "Slovakia", 1,
-  "Slovenia", 1,
-  "Spain", 0,
-  "Sweden", 0,
-  "Switzerland", 0,
-  "Turkey", 1,
-  "United States", 1
-  )
-
-# The tracking variables explain very nicely the difference in test scores
-# for all countries pooled, up to 50% R squared.
+# central_examination <-
+#   tribble(
+#   ~ country, ~central_examination,
+#   "Australia", 1, 
+#   "Austria", 0, 
+#   "Belgium", 0,
+#   "Bulgaria", 1,
+#   "Canada", rbinom(1, 1, prob = 0.51),
+#   "Czech Republic", 1,
+#   "Denmark", 1,
+#   "Finland", 1,
+#   "France", 1,
+#   "Germany", 0,
+#   "United Kingdom", 1,
+#   "Greece", 0,
+#   "Hong Kong", 1,
+#   "Hungary", 1,
+#   "Iceland", 1,
+#   "Ireland", 1,
+#   "Israel", 1,
+#   "Italy", 1,
+#   "Japan", 1,
+#   "Korea", 1,
+#   "Latvia", 1,
+#   "Liechtenstein", 1,
+#   "Luxembourg", 1,
+#   "Netherlands", 1,
+#   "New Zealand", 1,
+#   "Norway", 1,
+#   "Poland", 1,
+#   "Portugal", 0,
+#   "Russia", 1,
+#   "Slovakia", 1,
+#   "Slovenia", 1,
+#   "Spain", 0,
+#   "Sweden", 0,
+#   "Switzerland", 0,
+#   "Turkey", 1,
+#   "United States", 1
+#   )
 
@@
 
-ready_data <- dif_data(complete_gaps, tracking_data, central_examination)
-
 ready_data_age <-
@@
 
+ivs <- c("num_tracks", "age_selection", "I(length > 0)", "zvoc")
+
+formula_seq <-
+  lapply(1:length(ivs), function(x) {
+    paste0("difference ~ ", paste0(ivs[1:x], collapse = " + ") %>% paste0(" + (1|year)"))
+  }) %>%
+  lapply(as.formula)
+
+all_track_models <- 
+  map(formula_seq, ~ {
+
+  mod_tracking <-
+  brms::brm(
+    .x, family = gaussian(),
+    data = ready_data_age,
+    warmup = 1000, iter = 2000, chains = 5,
+    control = list(adapt_delta = 0.90)
+  )
+
+  mod_tracking
+
+})
+
+tidy_dfs <-
+  map(all_track_models, ~ {
+
+  broom::tidy(.x) %>%
+    .[grepl("^b_", .$term), ] %>%
+    map_if(is_double, round, 2) %>%
+    as_tibble %>%
+    transmute(term,
+              estimate = paste0(estimate, " (", lower, "/", upper, ")"))
+  })
+
+tidy_dfs_two <-
+  reduce(tidy_dfs, full_join, by = "term") %>%
+  setNames(c(" ", paste("Model", 1:4))) %>%
+  mutate(" " = c(
+    "Intercept",
+    "Num of tracks = 1",
+    "Age at track >= 15",
+    "% curriculum tracked",
+    "Vocational Index"
+  )) %>%
+  .[c(2:5, 1), ]
+
+r_square <-
+  paste0("\\hline ", paste0("R square: & ",
+         paste0(map(all_track_models, ~ bayes_R2(.x)[1] %>% round(2) * 100) %>% paste0("\\%"),
+                       collapse = " & ")), " \\\\")
+n_obs <-
+  paste0(paste0(" Sample size: & ",
+         paste0(map(all_track_models, nobs), collapse = " & ")), " \\\\ \\hline")
+
+command_to_row <- paste0(r_square, n_obs)
+
+addtorow_two <- list(pos = list(pos = 5), command = command_to_row)
+
+xtable(tidy_dfs_two) %>% print.xtable(
+  include.rownames = FALSE,
+  add.to.row = addtorow_two
+  )
+@
+
+<<>>=
+# This is the closest I could get to getting vif's fro the last model:
+
+lm(difference ~ num_tracks + age_selection + I(length > 0) + zvoc,
+   data = ready_data_age) %>%
+  vif()
+
+mod_tracking_ext <-
+  brms::brm(formula_seq[[4]],
+            family = gaussian(),
+            data = ready_data_age,
+            warmup = 1000, iter = 2000, chains = 5,
+            control = list(adapt_delta = 0.90)
+  )
+
+new_data <-
+  ready_data_age %>%
+  with(crossing(country = countries,
+                num_tracks = unique(num_tracks),
+                age_selection = unique(age_selection),
+                zvoc = mean(zvoc, na.rm = TRUE),
+                length = typical(length > 0),
+                year = "2009"))
+
+predictions <- posterior_predict(mod_tracking_ext, newdata = new_data)
+sd_obs <- colSds(predictions)
+
+new_data <-
+  new_data %>%
+  mutate(
+    pred = colMeans(predictions),
+    pred_low = pred - (2 * sd_obs),
+    pred_high = pred + (2 * sd_obs)
+  )
+
+sum_ready_data <-
+  ready_data_age %>%
+  group_by(country, num_tracks, age_selection) %>%
+  summarize(avg_diff = mean(difference),
+            diff_low = avg_diff - 2 * sd(difference),
+            diff_high = avg_diff + 2 * sd(difference)) %>%
+  ungroup()
+
+new_data %>%
+  left_join(sum_ready_data) %>%
+  mutate(age_selection = ifelse(age_selection == 1, ">= 15", "< 15")) %>%
+  ggplot(aes(num_tracks, avg_diff, fill = num_tracks)) +
+  geom_errorbar(aes(ymin = diff_low, ymax = diff_high), width = 0.2, colour = "red") +
+  geom_point(colour = "red") +
+  geom_col(aes(y = pred, alpha = 0.4)) +
+  geom_errorbar(aes(ymin = pred_low, ymax = pred_high), width = 0.2) +
+  facet_wrap(country ~ age_selection) +
+  scale_y_continuous(name = "90/10 SD gap", limits = c(0, 4)) +
+  scale_alpha(guide = F) +
+  scale_x_discrete(name = "Number of tracks", labels = c("> 1 track", "1 track")) +
+  scale_fill_manual(values = c("orangered2", "blue3"), guide = F) +
+  ggtitle("Predicted 90/10 gap for different combinations of number of tracks and age of selection")
+@
+
+<<>>=
+pred_avg_diff<-
+  new_data %>%
+  left_join(sum_ready_data) %>%
+  filter(!is.na(avg_diff) | num_tracks == 1 & age_selection == 1) %>%
+  select(country, num_tracks, age_selection, pred, avg_diff)
+
+track_should_change <- pred_avg_diff$country[pred_avg_diff$country %>% duplicated]
+
+avg_diff_data <-
+  pred_avg_diff %>%
+  filter(country %in% track_should_change) %>%
+  transmute(country,
+            avg_diff = ifelse(avg_diff %>% is.na, pred, avg_diff),
+            combn = paste0(num_tracks, "_", age_selection)) %>%
+  filter(combn != "1_1")
+
+final_avg_diff_table <-
+  avg_diff_data %>%
+  left_join(pred_avg_diff %>%
+              filter(country %in% track_should_change & (num_tracks == 1 & age_selection == 1)) %>%
+              select(country, pred), by = "country") %>%
+  select(country, combn, avg_diff, pred) %>%
+  separate(combn, c("num_tracks", "age_selection"), sep = "_") %>%
+  transmute(
+    Country = country,
+    `Number of tracks` = ifelse(num_tracks == 0, "More than one track", "One track"),
+    `Age of selection` = ifelse(age_selection == 1, "15 or more", "Less than 15"),
+    `Actual inequality` = avg_diff %>% round(3),
+    `Predicted inequality` = pred %>% round(3),
+    `% reduction` = (((`Actual inequality` - `Predicted inequality`)/ `Actual inequality`) * 100) %>%
+      ceiling %>% paste0("%")
+    )
+
+xtable(final_avg_diff_table) %>% print.xtable(
+  include.rownames = FALSE
+  )
+
+@
+
+<<>>=
+
 gaps <- c("90th/10th SES gap", "80th/20th SES gap", "70th/30th SES gap")
-
 all_gaps_models <-
@@
           zvoc +
-          ztrack:zvoc, family = gaussian(),
+          ztrack:zvoc +
+          (1|year), family = gaussian(),
       data = ready_data_age,
-      warmup = 1000, iter = 2000, chains = 5
+      warmup = 1000, iter = 2000, chains = 5,
+      control = list(adapt_delta = 0.95)
       )
@@
   scale_y_continuous(limits = c(-0.5, 2)) +
-  ggtitle("Coef from linear model of inequality on degree of tracking and vocational tracking") +
+  scale_x_discrete(labels = c("Intercept",
+                              "Tracking Index",
+                              "Vocational * Tracking Index",
+                              "Vocational Index")) +
+  labs(y = "Estimate from linear model", x = "Predictors") +
+  ggtitle("Estimation of tracking/vocational models for different gaps") +
   coord_flip()
 
-library(brms)
 # It under predicts the y value rather than over predict
@@
 # This linear model explains the pooled 90/10 differences
+
+ready_data <- dif_data(complete_gaps, tracking_data, central_examination)
 
@@
 
-pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
-pp_check(mod_tracking, type = "scatter_avg", nsamples = 100)
-pp_check(mod_tracking, type = "boxplot", nsamples = 20)
-pp_check(mod_tracking, type = "error_hist", nsamples = 6)
-pp_check(mod_tracking, type = "ecdf_overlay")
-
-pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "zvoc")
-pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "ztrack")
-
-# I don't know how to estimate VIF's with bayesian reasoning
-model_formula <- difference ~ num_tracks + age_selection + zvoc + I(length > 0)
-mod_tracking_ext <-
-  brms::brm(model_formula, family = gaussian(),
-  data = ready_data_age,
-  warmup = 1000, iter = 2000, chains = 5
-  )
-
-bayes_R2(mod_tracking_ext)
-
-# This is the closest I could get to getting vif's:
-lm(model_formula, data = ready_data_age) %>%
-  vif()
-
-new_data <-
-  ready_data_age %>%
-  with(crossing(country = countries,
-                num_tracks = unique(num_tracks),
-                age_selection = unique(age_selection),
-                zvoc = mean(zvoc, na.rm = TRUE),
-                length = typical(length > 0)))
-
-predictions <- posterior_predict(mod_tracking_ext, newdata = new_data)
-sd_obs <- colSds(predictions)
-
-new_data <-
-  new_data %>%
-  mutate(
-    pred = colMeans(predictions),
-    pred_low = pred - (2 * sd_obs),
-    pred_high = pred + (2 * sd_obs)
-  )
-
-sum_ready_data <-
-  ready_data_age %>%
-  group_by(country, num_tracks, age_selection) %>%
-  summarize(avg_diff = mean(difference),
-            diff_low = avg_diff - 2 * sd(difference),
-            diff_high = avg_diff + 2 * sd(difference)) %>%
-  ungroup()
-
-new_data %>%
-  left_join(sum_ready_data) %>%
-  ggplot(aes(num_tracks %>% as.factor, avg_diff, fill = num_tracks)) +
-  geom_errorbar(aes(ymin = diff_low, ymax = diff_high), width = 0.2, colour = "red") +
-  geom_point(colour = "red") +
-  scale_y_continuous(limits = c(0, 4)) +
-  geom_col(aes(y = pred, alpha = 0.4)) +
-  geom_errorbar(aes(ymin = pred_low, ymax = pred_high), width = 0.2) +
-  facet_wrap(country ~ age_selection) +
-  scale_fill_wsj()
+plot(plot_interaction, plot = FALSE)[[1]] +
+  labs(x = "Tracking Index (SD)", y = "Difference in test scores (SD)") +
+  scale_colour_discrete(
+    name = "Vocational index \n quantiles",
+    labels = seq(100, 0, -25) %>% paste0("%")
+    ) +
+  scale_fill_discrete(guide = FALSE) +
+  ggtitle("Interaction between Tracking and Vocational index")
+
+# pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
+# pp_check(mod_tracking, type = "scatter_avg", nsamples = 100)
+# pp_check(mod_tracking, type = "boxplot", nsamples = 20)
+# pp_check(mod_tracking, type = "error_hist", nsamples = 6)
+# pp_check(mod_tracking, type = "ecdf_overlay")
+# 
+# pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "zvoc")
+# pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "ztrack")
+
 @

### 2017-09-05 paper/paper.Rnw
Finished all tracking tables. Reorganizing next

@@
     `Age of selection` = ifelse(age_selection == 1, "15 or more", "Less than 15"),
-    `Actual inequality` = avg_diff %>% round(3),
-    `Predicted inequality` = pred %>% round(3),
-    `% reduction` = (((`Actual inequality` - `Predicted inequality`)/ `Actual inequality`) * 100) %>%
+    `Inequality` = avg_diff %>% round(3),
+    `Predicted` = pred %>% round(3),
+    `% reduction` = (((`Inequality` - `Predicted`)/ `Inequality`) * 100) %>%
       ceiling %>% paste0("%")
@@
 
-<<>>=
-
+<<interaction_model>>=
+ready_data <- dif_data(complete_gaps, tracking_data, central_examination)
+
+mod_tracking <-
+  brms::brm(
+    difference ~
+      ztrack +
+      zvoc +
+      ztrack:zvoc, family = gaussian(),
+  data = ready_data_age,
+  warmup = 1000, iter = 2000, chains = 5
+  )
+
+bayes_R2(mod_tracking)
+
+plot_interaction <-
+  marginal_effects(mod_tracking,
+                 effects = "ztrack:zvoc",
+                 int_conditions = list(zvoc = quantile))
+
+plot(plot_interaction, plot = FALSE)[[1]] +
+  labs(x = "Tracking Index (SD)", y = "Difference in test scores (SD)") +
+  scale_colour_discrete(
+    name = "Vocational index \n quantiles",
+    labels = seq(100, 0, -25) %>% paste0("%")
+    ) +
+  scale_fill_discrete(guide = FALSE) +
+  ggtitle("Interaction between Tracking and Vocational index")
+
+# pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
+# pp_check(mod_tracking, type = "scatter_avg", nsamples = 100)
+# pp_check(mod_tracking, type = "boxplot", nsamples = 20)
+# pp_check(mod_tracking, type = "error_hist", nsamples = 6)
+# pp_check(mod_tracking, type = "ecdf_overlay")
+# 
+# pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "zvoc")
+# pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "ztrack")
+@
+
+<<all_gaps_plot>>=
 gaps <- c("90th/10th SES gap", "80th/20th SES gap", "70th/30th SES gap")
@@
 # This linear model explains the pooled 90/10 differences
-
-ready_data <- dif_data(complete_gaps, tracking_data, central_examination)
-
-mod_tracking <-
-  brms::brm(
-    difference ~
-      ztrack +
-      zvoc +
-      ztrack:zvoc, family = gaussian(),
-  data = ready_data_age,
-  warmup = 1000, iter = 2000, chains = 5
-  )
-
-bayes_R2(mod_tracking)
-
-plot_interaction <-
-  marginal_effects(mod_tracking,
-                 effects = "ztrack:zvoc",
-                 int_conditions = list(zvoc = quantile))
-
-plot(plot_interaction, plot = FALSE)[[1]] +
-  labs(x = "Tracking Index (SD)", y = "Difference in test scores (SD)") +
-  scale_colour_discrete(
-    name = "Vocational index \n quantiles",
-    labels = seq(100, 0, -25) %>% paste0("%")
-    ) +
-  scale_fill_discrete(guide = FALSE) +
-  ggtitle("Interaction between Tracking and Vocational index")
-
-# pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
-# pp_check(mod_tracking, type = "scatter_avg", nsamples = 100)
-# pp_check(mod_tracking, type = "boxplot", nsamples = 20)
-# pp_check(mod_tracking, type = "error_hist", nsamples = 6)
-# pp_check(mod_tracking, type = "ecdf_overlay")
-# 
-# pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "zvoc")
-# pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "ztrack")
-
-@
-
-<<>>=
+@
+
+<<cumulative_modeling_table>>=
 
@@
 
-ready_data_two <-
-  ready_data %>%
+all_gaps_models_cum <-
+  map(gaps, function(gap) {
+
+    gap_data <- dif_data(complete_gaps, tracking_data, central_examination, which_gap = gap)
+
+    ready_data_age <-
+      gap_data %>%
+      mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor) %>%
       group_by(country) %>%
       summarize_at(vars(vars_unique), unique) %>%
-  mutate(cum_diff = ready_data %>% group_by(country) %>% summarize(m = sum(difference)) %>% pull(m))
+      mutate(cum_diff =
+               gap_data %>%
+               group_by(country) %>%
+               summarize(m = sum(difference)) %>%
+               pull(m))
 
     mod_tracking <-
-  ready_data_two %>%
-  lm(cum_diff ~ ztrack + zvoc, data = .) %>%
-  arm::display(detail = T)
-
-mod_tracking <-
-  ready_data_two %>%
+      ready_data_age %>%
       filter(!is.na(age_selection)) %>%
       brm(
-    cum_diff ~ I(age_selection >= 15),
+        cum_diff ~ ztrack + zvoc,
         family = gaussian(),
@@
 
-bayes_R2(mod_tracking)
-
-pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
-pp_check(mod_tracking, type = "ecdf_overlay", nsamples = 10)
-pp_check(mod_tracking, type = "boxplot", nsamples = 30)
-pp_check(mod_tracking, type = "stat", nsamples = 200) # very nice to compare central tendency
-pp_check(mod_tracking, type = "error_scatter_avg", nsamples = 1000)
-
-pp_check(mod_tracking, type = "loo_pit")
+    mod_tracking
+})
+
+tidy_dfs <-
+  map(all_gaps_models_cum, ~ {
+
+  broom::tidy(.x) %>%
+    .[grepl("^b_", .$term), ] %>%
+    mutate_if(is_double, round, 2) %>%
+    transmute(term,
+              estimate = paste0(estimate, " (", lower, "/", upper, ")"))
+  })
+
+tidy_dfs_two <-
+  reduce(tidy_dfs, full_join, by = "term") %>%
+  setNames(c(" ", gsub(" SES gap", "", gaps))) %>%
+  mutate(" " = c(
+    "Intercept",
+    "Track Index",
+    "Vocational Index"
+  )) %>%
+  .[c(2, 3, 1), ]
+
+r_square <-
+  paste0("\\hline ", paste0("R square: & ",
+         paste0(map(all_gaps_models_cum, ~ bayes_R2(.x)[1] %>% round(2) * 100) %>% paste0("\\%"),
+                       collapse = " & ")), " \\\\")
+n_obs <-
+  paste0(paste0(" Sample size: & ",
+         paste0(map(all_gaps_models_cum, nobs), collapse = " & ")), " \\\\ \\hline")
+
+command_to_row <- paste0(r_square, n_obs)
+
+addtorow_two <- list(pos = list(pos = 3), command = command_to_row)
+
+xtable(tidy_dfs_two) %>%
+  print.xtable(
+    include.rownames = FALSE,
+    add.to.row = addtorow_two
+  )
+@
+
+<<>>=
+# pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
+# pp_check(mod_tracking, type = "ecdf_overlay", nsamples = 10)
+# pp_check(mod_tracking, type = "boxplot", nsamples = 30)
+# pp_check(mod_tracking, type = "stat", nsamples = 200) # very nice to compare central tendency
+# pp_check(mod_tracking, type = "error_scatter_avg", nsamples = 1000)
+# 
+# pp_check(mod_tracking, type = "loo_pit")
 # over prediction
@@
   ready_data_two %>%
-  filter(!is.na(age_selection)) %>%
-  mutate(pred_median = posterior_predict(mod_tracking) %>% colMeans(),
+  filter(!is.na(ztrack), !is.na(zvoc)) %>%
+  mutate(pred_median = posterior_predict(all_gaps_models_cum[[1]]) %>% colMeans(),
          ratio_pred = pred_median/cum_diff) %>%

### 2017-09-05 paper/paper.Rnw
Reorganized chunks and tested paper is running fine - still need to prettify output

--- 2017-08-30 paper/paper.Rnw
@@
   library(ggthemes)
+  library(readxl)
+  library(inequalityintsvy)
+  library(brms)
   library(tidyverse)
@@
 Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap.
-Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentially, are countries that have the lower leven of inequalities among the ones presented here.
-
-<<next_steps>>=
-# Next steps:
-
-# Continue by doing the multilevel models to see what explains what. Include
-# all indicators from the reardon/russian girl paper.
-
-# Graph the increase in each country vs the increase/decrease of the economic inequality indicators
-# Specially the 90/10
-
-# Calculate how big is the gap between reading and math
-
-# Continue with the PIRLS to see if there are specific patterns in 4th and 8th graders gap.
-
-# Get each country trendline adjusted for the inequality indicators and place in the same country graph.
-
-# Should I add the parent's education in the lm model to see how trends change adjusted for that?
-
-# A weak welfare system, together with income inequality, what's their pattern?
-# What if we put the school differentiation/tracking aspect in? Are there country groups based on these
-# patterns.
-
-# In countries where there is high differentiation/tracking, is there a jump in the evolution of the gap between PIRLS/TIMSS and PISA?
-@
-
-\bibliography{mybibliography.bib}
-
-\pagebreak
-
-\section{Appendix}
+Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentially, are countries that have the lower level of inequalities among the ones presented here.
+
+% 90/80/70 gaps merged
 
@@
 
-\begin{figure}
-\begin{center}
-<<graph_all_gaps,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+<<merge_all_gaps>>=
 complete_data_topbottom_80 <-
@@
 
-######
+####
 
@@
 
-#####
+####
 
@@
   bind_rows(complete_data_topbottom_70)
-
+@
+
+% end here
+
+% <<>>=
+% # 2000
+% # SC03Q01 School public/private
+% # 1 Public
+% # 2 Private
+% # 7 N/A
+% # 8 M/R
+% # 9 Mis
+% 
+% # 2003
+% # SC03Q01 (8) Public or private
+% # 1 Public
+% # 2 Private
+% # 7 N/A
+% # 8 Invalid
+% # 9 Miss
+% 
+% # 2006
+% # SC02Q01 (8) Public or private
+% #  1 Public
+% #  2 Private
+% #  7 N/A
+% #  8 Invalid
+% #  9 Missing 
+% 
+% # 2009
+% # SC02Q01
+% # 1 public
+% # 2 private
+% 
+% # 2012
+% # SC01Q01 Public or private Num
+% # 1 Public
+% # 2 Private
+% # 7 N/A
+% # 8 Invalid
+% # 9 Missing
+% 
+% # 2015
+% # SC013Q01TA
+% # 1 Public
+% # 2 Private
+% # 7 N/A
+% # 8 Invalid
+% # 9 Missing
+% 
+% # central_examination <-
+% #   tribble(
+% #   ~ country, ~central_examination,
+% #   "Australia", 1, 
+% #   "Austria", 0, 
+% #   "Belgium", 0,
+% #   "Bulgaria", 1,
+% #   "Canada", rbinom(1, 1, prob = 0.51),
+% #   "Czech Republic", 1,
+% #   "Denmark", 1,
+% #   "Finland", 1,
+% #   "France", 1,
+% #   "Germany", 0,
+% #   "United Kingdom", 1,
+% #   "Greece", 0,
+% #   "Hong Kong", 1,
+% #   "Hungary", 1,
+% #   "Iceland", 1,
+% #   "Ireland", 1,
+% #   "Israel", 1,
+% #   "Italy", 1,
+% #   "Japan", 1,
+% #   "Korea", 1,
+% #   "Latvia", 1,
+% #   "Liechtenstein", 1,
+% #   "Luxembourg", 1,
+% #   "Netherlands", 1,
+% #   "New Zealand", 1,
+% #   "Norway", 1,
+% #   "Poland", 1,
+% #   "Portugal", 0,
+% #   "Russia", 1,
+% #   "Slovakia", 1,
+% #   "Slovenia", 1,
+% #   "Spain", 0,
+% #   "Sweden", 0,
+% #   "Switzerland", 0,
+% #   "Turkey", 1,
+% #   "United States", 1
+% #   )
+% 
+% @
+% 
+% <<complete_tracking_model>>=
+% tracking_data <-
+%   read_xlsx("./data/tracking.xlsx", sheet = "all_data") %>%
+%   map_if(is_double, round, 2) %>%
+%   as_tibble
+% 
+% dif_data <- function(my_data, tracking, central, which_gap = "90th/10th SES gap") {
+%   my_data %>%
+%     filter(type_test == "math", type == which_gap) %>%
+%     select(wave, country, difference) %>%
+%     rename(year = wave) %>%
+%     mutate(year = as.character(year)) %>%
+%     left_join(tracking_data, by = c("country" = "cntry_name")) %>%
+%     left_join(central_examination) %>%
+%     mutate(num_tracks = ifelse(tracks15y == 1, 1, 0) %>% as.factor,
+%            age_selection = selage) %>%
+%     filter(!is.na(num_tracks), !is.na(age_selection), !is.na(difference))
+% 
+% }
+% formula_sequence <- function(dv, ivs, random_effect) {
+%   formula_seq <-
+%     lapply(1:length(ivs), function(x) {
+%       paste0(dv ," ~ ", paste0(ivs[1:x], collapse = " + ") %>% paste0(" + ", random_effect))
+%     }) %>%
+%     lapply(as.formula)
+% 
+%   formula_seq  
+% }
+% stan_model_builder <- function(dv, iv, random, data) {
+%   all_formulas <- formula_sequence(dv, iv, random)
+% 
+%   mod_tracking <-
+%     map(all_formulas, ~ {
+%       brms::brm(
+%         .x,
+%         family = gaussian(),
+%         data = data,
+%         warmup = 1000, iter = 2000, chains = 5
+%       )
+%     })
+%   
+% }
+% stan_extractor <- function(models) {
+%   coef_list <-
+%     map(models, ~ {
+%     broom::tidy(.x) %>%
+%       .[grepl("^b_", .$term), ] %>%
+%       map_if(is_double, round, 2) %>%
+%       as_tibble %>%
+%       transmute(term,
+%                 estimate = paste0(estimate, " (", lower, "/", upper, ")"))
+%   })
+%   
+%   coef_list
+% }
+% stan_table_builder <- function(models_extracted) {
+%   intermediate_table <-
+%     reduce(models_extracted, full_join, by = "term") %>%
+%     setNames(c(" ", paste("Model", seq_len(models_extracted %>% length))))
+%   
+%   row_order <-
+%     setdiff(seq_len(nrow(intermediate_table)),
+%             grep("Intercept", intermediate_table$` `)) %>%
+%     `c`(grep("Intercept", intermediate_table$` `))
+% 
+%     intermediate_table[row_order, ]
+% }
+% stan_table <- function(models) {
+% 
+%   table_ready <-
+%     models %>%
+%     stan_extractor() %>%
+%     stan_table_builder()
+%   
+%   table_ready
+% }
+% 
+% table_details <- function(models) {
+%   
+%   random_effect <- nrow(models[[length(models)]]$ranef) > 0
+%   
+%   where_to_place <- 
+%     models[[length(models)]]$formula$formula %>%
+%     as.character %>%
+%     .[-(1:2)] %>% 
+%     stringr::str_split(" \\+ ") %>%
+%     pluck(1) %>%
+%     .[if (random_effect) -length(.) else 1:length(.)] %>%
+%     length
+%   
+%   r_square <-
+%     paste0("\\hline ",
+%            paste0("R square: & ",
+%                   paste0(map(models, ~ bayes_R2(.x)[1] %>% round(2) * 100) %>% paste0("\\%"),
+%                                      collapse = " & ")), " \\\\")
+%   n_obs <-
+%     paste0(paste0(" Sample size: & ",
+%                   paste0(map(models, nobs), collapse = " & ")), " \\\\ \\hline")
+%   
+%   command_to_row <- paste0(r_square, n_obs)
+%   addtorow_two <- list(pos = list(pos = where_to_place + 1), command = command_to_row)
+%   
+%   addtorow_two
+% }
+% 
+% ready_data_age <-
+%   dif_data(complete_gaps, tracking_data, central_examination) %>%
+%   mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)
+% 
+% all_track_models <-
+%   stan_model_builder(
+%   dv = "difference",
+%   iv = c("num_tracks", "age_selection", "I(length > 0)", "zvoc"),
+%   random = "(1 | year)",
+%   data = ready_data_age
+% )
+% 
+% all_track_models %>%
+%   stan_table() %>%
+%   mutate(" " = c(
+%     "Only 1 track",
+%     "Age selection >= 15",
+%     "% of curric tracked",
+%     "Vocational Index",
+%     "Intercept"
+%   )) %>%
+%   xtable() %>%
+%   print.xtable(
+%   include.rownames = FALSE,
+%   add.to.row = table_details(all_track_models)
+%   )
+% @
+% 
+% <<graph_complete_tracking_model>>=
+% # This is the closest I could get to getting vif's fro the last model:
+% 
+% lm(difference ~ num_tracks + age_selection + I(length > 0) + zvoc,
+%    data = ready_data_age) %>%
+%   vif()
+% 
+% mod_tracking_ext <-
+%   brms::brm(formula_seq[[4]],
+%             family = gaussian(),
+%             data = ready_data_age,
+%             warmup = 1000, iter = 2000, chains = 5,
+%             control = list(adapt_delta = 0.90)
+%   )
+% 
+% new_data <-
+%   ready_data_age %>%
+%   with(crossing(country = countries,
+%                 num_tracks = unique(num_tracks),
+%                 age_selection = unique(age_selection),
+%                 zvoc = mean(zvoc, na.rm = TRUE),
+%                 length = typical(length > 0),
+%                 year = "2009"))
+% 
+% predictions <- posterior_predict(mod_tracking_ext, newdata = new_data)
+% sd_obs <- colSds(predictions)
+% 
+% new_data <-
+%   new_data %>%
+%   mutate(
+%     pred = colMeans(predictions),
+%     pred_low = pred - (2 * sd_obs),
+%     pred_high = pred + (2 * sd_obs)
+%   )
+% 
+% sum_ready_data <-
+%   ready_data_age %>%
+%   group_by(country, num_tracks, age_selection) %>%
+%   summarize(avg_diff = mean(difference),
+%             diff_low = avg_diff - 2 * sd(difference),
+%             diff_high = avg_diff + 2 * sd(difference)) %>%
+%   ungroup()
+% 
+% new_data %>%
+%   left_join(sum_ready_data) %>%
+%   mutate(age_selection = ifelse(age_selection == 1, ">= 15", "< 15")) %>%
+%   ggplot(aes(num_tracks, avg_diff, fill = num_tracks)) +
+%   geom_errorbar(aes(ymin = diff_low, ymax = diff_high), width = 0.2, colour = "red") +
+%   geom_point(colour = "red") +
+%   geom_col(aes(y = pred, alpha = 0.4)) +
+%   geom_errorbar(aes(ymin = pred_low, ymax = pred_high), width = 0.2) +
+%   facet_wrap(country ~ age_selection) +
+%   scale_y_continuous(name = "90/10 SD gap", limits = c(0, 4)) +
+%   scale_alpha(guide = F) +
+%   scale_x_discrete(name = "Number of tracks", labels = c("> 1 track", "1 track")) +
+%   scale_fill_manual(values = c("orangered2", "blue3"), guide = F) +
+%   ggtitle("Predicted 90/10 gap for different combinations of number of tracks and age of selection")
+% @
+% 
+% <<counterfactual_tracking>>=
+% pred_avg_diff<-
+%   new_data %>%
+%   left_join(sum_ready_data) %>%
+%   filter(!is.na(avg_diff) | num_tracks == 1 & age_selection == 1) %>%
+%   select(country, num_tracks, age_selection, pred, avg_diff)
+% 
+% track_should_change <- pred_avg_diff$country[pred_avg_diff$country %>% duplicated]
+% 
+% avg_diff_data <-
+%   pred_avg_diff %>%
+%   filter(country %in% track_should_change) %>%
+%   transmute(country,
+%             avg_diff = ifelse(avg_diff %>% is.na, pred, avg_diff),
+%             combn = paste0(num_tracks, "_", age_selection)) %>%
+%   filter(combn != "1_1")
+% 
+% final_avg_diff_table <-
+%   avg_diff_data %>%
+%   left_join(pred_avg_diff %>%
+%               filter(country %in% track_should_change & (num_tracks == 1 & age_selection == 1)) %>%
+%               select(country, pred), by = "country") %>%
+%   select(country, combn, avg_diff, pred) %>%
+%   separate(combn, c("num_tracks", "age_selection"), sep = "_") %>%
+%   transmute(
+%     Country = country,
+%     `Number of tracks` = ifelse(num_tracks == 0, "More than one track", "One track"),
+%     `Age of selection` = ifelse(age_selection == 1, "15 or more", "Less than 15"),
+%     `Inequality` = avg_diff %>% round(3),
+%     `Predicted` = pred %>% round(3),
+%     `% reduction` = (((`Inequality` - `Predicted`)/ `Inequality`) * 100) %>%
+%       ceiling %>% paste0("%")
+%     )
+% 
+% xtable(final_avg_diff_table) %>% print.xtable(
+%   include.rownames = FALSE
+%   )
+% 
+% @
+% 
+% <<interaction_model>>=
+% ready_data <- dif_data(complete_gaps, tracking_data, central_examination)
+% 
+% stan_models <-
+%   stan_model_builder(
+%   dv = "difference",
+%   iv = c("ztrack", "zvoc", "ztrack:zvoc"),
+%   random = "(1 | year)",
+%   data = ready_data
+%   )
+% 
+% table_ready <-
+%   stan_models %>%
+%   stan_table() %>%
+%   mutate(" " = c(
+%     "Tracking Index",
+%     "Vocational Index",
+%     "Trackin * Vocational Index",
+%     "Intercept"
+%   )) %>%
+%   xtable() %>%
+%   print.xtable(
+%     include.rownames = FALSE,
+%     add.to.row = table_details(stan_models)
+%   )
+% 
+% plot_interaction <-
+%   marginal_effects(stan_models[[length(stan_models)]],
+%                  effects = "ztrack:zvoc",
+%                  int_conditions = list(zvoc = quantile))
+% 
+% plot(plot_interaction, plot = FALSE)[[1]] +
+%   labs(x = "Tracking Index (SD)", y = "Difference in test scores (SD)") +
+%   scale_colour_discrete(
+%     name = "Vocational index \n quantiles",
+%     labels = seq(100, 0, -25) %>% paste0("%")
+%     ) +
+%   scale_fill_discrete(guide = FALSE) +
+%   ggtitle("Interaction between Tracking and Vocational index")
+% 
+% # pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
+% # pp_check(mod_tracking, type = "scatter_avg", nsamples = 100)
+% # pp_check(mod_tracking, type = "boxplot", nsamples = 20)
+% # pp_check(mod_tracking, type = "error_hist", nsamples = 6)
+% # pp_check(mod_tracking, type = "ecdf_overlay")
+% # 
+% # pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "zvoc")
+% # pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "ztrack")
+% @
+% 
+% <<all_gaps_plot>>=
+% gaps <- c("90th/10th SES gap", "80th/20th SES gap", "70th/30th SES gap")
+% all_gaps_models <-
+%   map(gaps, function(gap) {
+%     ready_data_age <-
+%       dif_data(complete_gaps, tracking_data, central_examination, which_gap = gap) %>%
+%       mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)
+%     
+%     mod_tracking <-
+%       brms::brm(
+%         difference ~
+%           ztrack +
+%           zvoc +
+%           ztrack:zvoc +
+%           (1|year), family = gaussian(),
+%       data = ready_data_age,
+%       warmup = 1000, iter = 2000, chains = 5,
+%       control = list(adapt_delta = 0.95)
+%       )
+%     
+%     mod_tracking
+% })
+% 
+% model_data_merged <-
+%   map(all_gaps_models, ~ fixef(.x)[, c(1, 3, 4)] %>% as_tibble) %>%
+%   setNames(gaps) %>%
+%   enframe() %>%
+%   unnest(value) %>%
+%   mutate_if(is_double, round, 2) %>%
+%   transmute(model = name,
+%            terms = all_gaps_models[[1]] %>% fixef %>% row.names %>% rep(3),
+%            estimate = Estimate,
+%            lower = `2.5%ile`,
+%            upper = `97.5%ile`)
+% 
+% model_data_merged %>%
+%   ggplot(aes(terms, estimate, colour = model)) +
+%   geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
+%   geom_point() +
+%   geom_hline(yintercept = 0) +
+%   scale_colour_discrete(name = NULL) +
+%   scale_y_continuous(limits = c(-0.5, 2)) +
+%   scale_x_discrete(labels = c("Intercept",
+%                               "Tracking Index",
+%                               "Vocational * Tracking Index",
+%                               "Vocational Index")) +
+%   labs(y = "Estimate from linear model", x = "Predictors") +
+%   ggtitle("Estimation of tracking/vocational models for different gaps") +
+%   coord_flip()
+% 
+% # It under predicts the y value rather than over predict
+% # It has problems predicting the central tendency of the model
+% # This linear model explains the pooled 90/10 differences
+% @
+% 
+% <<cumulative_modeling_table>>=
+% 
+% # One idea I had was to make a cumulative sum of the trend,
+% # so the cumulative inequality over time for each country
+% # and see how each tracking feature explains this cumulative
+% # pattern, and it does so very well, with each one
+% # separately (because I have only 30 observations) explaining
+% # about 22% each. Age selection, up to 30% alone.
+% 
+% vars_unique <- c("length", "num_tracks", "age_selection", "central_examination", "ztrack",
+%                  "zvoc")
+% 
+% all_gaps_models_cum <-
+%   map(gaps, function(gap) {
+%     
+%     gap_data <- dif_data(complete_gaps, tracking_data, central_examination, which_gap = gap)
+%     
+%     ready_data_age <-
+%       gap_data %>%
+%       mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor) %>%
+%       group_by(country) %>%
+%       summarize_at(vars(vars_unique), unique) %>%
+%       mutate(cum_diff =
+%                gap_data %>%
+%                group_by(country) %>%
+%                summarize(m = sum(difference)) %>%
+%                pull(m))
+%     
+%     mod_tracking <-
+%       ready_data_age %>%
+%       filter(!is.na(age_selection)) %>%
+%       brm(
+%         cum_diff ~ ztrack + zvoc,
+%         family = gaussian(),
+%         data = .,
+%         warmup = 1000, iter = 2000, chains = 5
+%       )
+% 
+%     mod_tracking
+% })
+% 
+% all_gaps_models_cum %>%
+%   stan_table() %>%
+%   mutate(" " = c(
+%     "Track Index",
+%     "Vocational Index",
+%     "Intercept"
+%   )) %>%
+%   setNames(c(" ", gsub(" SES gap", "", gaps))) %>%
+%   xtable() %>%
+%   print.xtable(
+%     include.rownames = FALSE,
+%     add.to.row = table_details(all_gaps_models_cum)
+%   )
+% 
+% @
+% 
+% <<next_steps>>=
+% # Next steps:
+% 
+% # Continue by doing the multilevel models to see what explains what. Include
+% # all indicators from the reardon/russian girl paper.
+%   
+% # Graph the increase in each country vs the increase/decrease of the economic inequality indicators
+% # Specially the 90/10
+% 
+% # Calculate how big is the gap between reading and math
+%   
+% # Continue with the PIRLS to see if there are specific patterns in 4th and 8th graders gap.
+%   
+% # Get each country trendline adjusted for the inequality indicators and place in the same country graph.
+% 
+% # Should I add the parent's education in the lm model to see how trends change adjusted for that?
+% 
+% # A weak welfare system, together with income inequality, what's their pattern?
+% # What if we put the school differentiation/tracking aspect in? Are there country groups based on these
+% # patterns.
+% 
+% # In countries where there is high differentiation/tracking, is there a jump in the evolution of the gap between PIRLS/TIMSS and PISA?
+% @
+
+\bibliography{mybibliography.bib}
+
+\pagebreak
+
+\section{Appendix}
+
+\begin{figure}
+\begin{center}
+<<graph_all_gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
 complete_gaps %>%
@@
 <<eval = F>>=
-
-# 2000
-# SC03Q01 School public/private
-# 1 Public
-# 2 Private
-# 7 N/A
-# 8 M/R
-# 9 Mis
-
-# 2003
-# SC03Q01 (8) Public or private
-# 1 Public
-# 2 Private
-# 7 N/A
-# 8 Invalid
-# 9 Miss
-
-# 2006
-# SC02Q01 (8) Public or private
-#  1 Public
-#  2 Private
-#  7 N/A
-#  8 Invalid
-#  9 Missing
-
-# 2009
-# SC02Q01
-# 1 public
-# 2 private
-
-# 2012
-# SC01Q01 Public or private Num
-# 1 Public
-# 2 Private
-# 7 N/A
-# 8 Invalid
-# 9 Missing
-
-# 2015
-# SC013Q01TA
-# 1 Public
-# 2 Private
-# 7 N/A
-# 8 Invalid
-# 9 Missing
-library(readxl)
-
-tracking_data <-
-  read_xlsx("./paper/data/tracking.xlsx", sheet = "all_data") %>%
-  map_if(is_double, round, 2) %>%
-  as_tibble
-
-library(inequalityintsvy)
-
-central_examination <-
-  tribble(
-  ~ country, ~central_examination,
-  "Australia", 1, 
-  "Austria", 0, 
-  "Belgium", 0,
-  "Bulgaria", 1,
-  "Canada", rbinom(1, 1, prob = 0.51),
-  "Czech Republic", 1,
-  "Denmark", 1,
-  "Finland", 1,
-  "France", 1,
-  "Germany", 0,
-  "United Kingdom", 1,
-  "Greece", 0,
-  "Hong Kong", 1,
-  "Hungary", 1,
-  "Iceland", 1,
-  "Ireland", 1,
-  "Israel", 1,
-  "Italy", 1,
-  "Japan", 1,
-  "Korea", 1,
-  "Latvia", 1,
-  "Liechtenstein", 1,
-  "Luxembourg", 1,
-  "Netherlands", 1,
-  "New Zealand", 1,
-  "Norway", 1,
-  "Poland", 1,
-  "Portugal", 0,
-  "Russia", 1,
-  "Slovakia", 1,
-  "Slovenia", 1,
-  "Spain", 0,
-  "Sweden", 0,
-  "Switzerland", 0,
-  "Turkey", 1,
-  "United States", 1
-  )
-
-# The tracking variables explain very nicely the difference in test scores
-# for all countries pooled, up to 50% R squared.
-
-dif_data <- function(my_data, tracking, central, which_gap = "90th/10th SES gap") {
-  my_data %>%
-    filter(type_test == "math", type == which_gap) %>%
-    select(wave, country, difference) %>%
-    rename(year = wave) %>%
-    mutate(year = as.character(year)) %>%
-    left_join(tracking_data, by = c("country" = "cntry_name")) %>%
-    left_join(central_examination) %>%
-    mutate(num_tracks = ifelse(tracks15y == 1, 1, 0) %>% as.factor,
-           age_selection = selage) %>%
-    filter(!is.na(num_tracks), !is.na(age_selection), !is.na(difference))
-
-}
-
-ready_data <- dif_data(complete_gaps, tracking_data, central_examination)
-
-ready_data_age <-
-  dif_data(complete_gaps, tracking_data, central_examination) %>%
-  mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)
-
-gaps <- c("90th/10th SES gap", "80th/20th SES gap", "70th/30th SES gap")
-
-all_gaps_models <-
-  map(gaps, function(gap) {
-    ready_data_age <-
-      dif_data(complete_gaps, tracking_data, central_examination, which_gap = gap) %>%
-      mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)
-
-    mod_tracking <-
-      brms::brm(
-        difference ~
-          ztrack +
-          zvoc +
-          ztrack:zvoc, family = gaussian(),
-      data = ready_data_age,
-      warmup = 1000, iter = 2000, chains = 5
-      )
-
-    mod_tracking
-})
-
-model_data_merged <-
-  map(all_gaps_models, ~ fixef(.x)[, c(1, 3, 4)] %>% as_tibble) %>%
-  setNames(gaps) %>%
-  enframe() %>%
-  unnest(value) %>%
-  mutate_if(is_double, round, 2) %>%
-  transmute(model = name,
-           terms = all_gaps_models[[1]] %>% fixef %>% row.names %>% rep(3),
-           estimate = Estimate,
-           lower = `2.5%ile`,
-           upper = `97.5%ile`)
-
-model_data_merged %>%
-  ggplot(aes(terms, estimate, colour = model)) +
-  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
-  geom_point() +
-  geom_hline(yintercept = 0) +
-  scale_colour_discrete(name = NULL) +
-  scale_y_continuous(limits = c(-0.5, 2)) +
-  ggtitle("Coef from linear model of inequality on degree of tracking and vocational tracking") +
-  coord_flip()
-
-library(brms)
-# It under predicts the y value rather than over predict
-# It has problems predicting the central tendency of the model
-# This linear model explains the pooled 90/10 differences
-
-mod_tracking <-
-  brms::brm(
-    difference ~
-      ztrack +
-      zvoc +
-      ztrack:zvoc, family = gaussian(),
-  data = ready_data_age,
-  warmup = 1000, iter = 2000, chains = 5
-  )
-
-bayes_R2(mod_tracking)
-
-plot_interaction <-
-  marginal_effects(mod_tracking,
-                 effects = "ztrack:zvoc",
-                 int_conditions = list(zvoc = quantile))
-
-pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
-pp_check(mod_tracking, type = "scatter_avg", nsamples = 100)
-pp_check(mod_tracking, type = "boxplot", nsamples = 20)
-pp_check(mod_tracking, type = "error_hist", nsamples = 6)
-pp_check(mod_tracking, type = "ecdf_overlay")
-
-pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "zvoc")
-pp_check(mod_tracking, type = "error_scatter_avg_vs_x", x = "ztrack")
-
-# I don't know how to estimate VIF's with bayesian reasoning
-model_formula <- difference ~ num_tracks + age_selection + zvoc + I(length > 0)
-mod_tracking_ext <-
-  brms::brm(model_formula, family = gaussian(),
-  data = ready_data_age,
-  warmup = 1000, iter = 2000, chains = 5
-  )
-
-bayes_R2(mod_tracking_ext)
-
-# This is the closest I could get to getting vif's:
-lm(model_formula, data = ready_data_age) %>%
-  vif()
-
-new_data <-
-  ready_data_age %>%
-  with(crossing(country = countries,
-                num_tracks = unique(num_tracks),
-                age_selection = unique(age_selection),
-                zvoc = mean(zvoc, na.rm = TRUE),
-                length = typical(length > 0)))
-
-predictions <- posterior_predict(mod_tracking_ext, newdata = new_data)
-sd_obs <- colSds(predictions)
-
-new_data <-
-  new_data %>%
-  mutate(
-    pred = colMeans(predictions),
-    pred_low = pred - (2 * sd_obs),
-    pred_high = pred + (2 * sd_obs)
-  )
-
-sum_ready_data <-
-  ready_data_age %>%
-  group_by(country, num_tracks, age_selection) %>%
-  summarize(avg_diff = mean(difference),
-            diff_low = avg_diff - 2 * sd(difference),
-            diff_high = avg_diff + 2 * sd(difference)) %>%
-  ungroup()
-
-new_data %>%
-  left_join(sum_ready_data) %>%
-  ggplot(aes(num_tracks %>% as.factor, avg_diff, fill = num_tracks)) +
-  geom_errorbar(aes(ymin = diff_low, ymax = diff_high), width = 0.2, colour = "red") +
-  geom_point(colour = "red") +
-  scale_y_continuous(limits = c(0, 4)) +
-  geom_col(aes(y = pred, alpha = 0.4)) +
-  geom_errorbar(aes(ymin = pred_low, ymax = pred_high), width = 0.2) +
-  facet_wrap(country ~ age_selection) +
-  scale_fill_wsj()
-@
-
-<<>>=
-
-# One idea I had was to make a cumulative sum of the trend,
-# so the cumulative inequality over time for each country
-# and see how each tracking feature explains this cumulative
-# pattern, and it does so very well, with each one
-# separately (because I have only 30 observations) explaining
-# about 22% each. Age selection, up to 30% alone.
-
-vars_unique <- c("length", "num_tracks", "age_selection", "central_examination", "ztrack",
-                 "zvoc")
-
-ready_data_two <-
-  ready_data %>%
-  group_by(country) %>%
-  summarize_at(vars(vars_unique), unique) %>%
-  mutate(cum_diff = ready_data %>% group_by(country) %>% summarize(m = sum(difference)) %>% pull(m))
-
-mod_tracking <-
-  ready_data_two %>%
-  lm(cum_diff ~ ztrack + zvoc, data = .) %>%
-  arm::display(detail = T)
-
-mod_tracking <-
-  ready_data_two %>%
-  filter(!is.na(age_selection)) %>%
-  brm(
-    cum_diff ~ I(age_selection >= 15),
-    family = gaussian(),
-    data = .,
-    warmup = 1000, iter = 2000, chains = 5
-  )
-
-bayes_R2(mod_tracking)
-
-pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
-pp_check(mod_tracking, type = "ecdf_overlay", nsamples = 10)
-pp_check(mod_tracking, type = "boxplot", nsamples = 30)
-pp_check(mod_tracking, type = "stat", nsamples = 200) # very nice to compare central tendency
-pp_check(mod_tracking, type = "error_scatter_avg", nsamples = 1000)
-
-pp_check(mod_tracking, type = "loo_pit")
+# pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)
+# pp_check(mod_tracking, type = "ecdf_overlay", nsamples = 10)
+# pp_check(mod_tracking, type = "boxplot", nsamples = 30)
+# pp_check(mod_tracking, type = "stat", nsamples = 200) # very nice to compare central tendency
+# pp_check(mod_tracking, type = "error_scatter_avg", nsamples = 1000)
+# 
+# pp_check(mod_tracking, type = "loo_pit")
 # over prediction
@@
 
-<<next_steps>>=
-# Next steps:
-
-# Continue by doing the multilevel models to see what explains what. Include
-# all indicators from the reardon/russian girl paper.
-
-# Graph the increase in each country vs the increase/decrease of the economic inequality indicators
-# Specially the 90/10
-
-# Calculate how big is the gap between reading and math
-
-# Continue with the PIRLS to see if there are specific patterns in 4th and 8th graders gap.
-
-# Get each country trendline adjusted for the inequality indicators and place in the same country graph.
-
-# Should I add the parent's education in the lm model to see how trends change adjusted for that?
-
-# A weak welfare system, together with income inequality, what's their pattern?
-# What if we put the school differentiation/tracking aspect in? Are there country groups based on these
-# patterns.
-
-# In countries where there is high differentiation/tracking, is there a jump in the evolution of the gap between PIRLS/TIMSS and PISA?
-@
-
 \end{document}

### 2017-09-06 paper/paper.Rnw
Finished commenting the results of the modeling

--- 2017-09-05 paper/paper.Rnw
@@
 Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap.
-Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentially, are countries that have the lower leven of inequalities among the ones presented here.
-
-<<next_steps>>=
-# Next steps:
-
-# Continue by doing the multilevel models to see what explains what. Include
-# all indicators from the reardon/russian girl paper.
-
-# Graph the increase in each country vs the increase/decrease of the economic inequality indicators
-# Specially the 90/10
-
-# Calculate how big is the gap between reading and math
-
-# Continue with the PIRLS to see if there are specific patterns in 4th and 8th graders gap.
-
-# Get each country trendline adjusted for the inequality indicators and place in the same country graph.
-
-# Should I add the parent's education in the lm model to see how trends change adjusted for that?
-
-# A weak welfare system, together with income inequality, what's their pattern?
-# What if we put the school differentiation/tracking aspect in? Are there country groups based on these
-# patterns.
-
-# In countries where there is high differentiation/tracking, is there a jump in the evolution of the gap between PIRLS/TIMSS and PISA?
-@
-
-\bibliography{mybibliography.bib}
-
-\pagebreak
-
-\section{Appendix}
+Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentially, are countries that have the lower level of inequalities among the ones presented here.
+
+\section{The role of curricular tracking in explaining inequalities}
+
+% 90/80/70 gaps merged
 
@@
   bind_rows(complete_data_topbottom_70)
-
-complete_gaps %>%
-  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
-  filter(type_test == "math",
-         country %in% countries) %>%
-  ggplot(aes(as.character(wave), difference, colour = type, linetype = type, group = type)) +
-  geom_line(stat = "smooth", method = "lm", size = 0.7) +
-  facet_wrap(~ country, ncol = 5) +
-  scale_colour_discrete(name = "Type of gaps") +
-  scale_linetype_discrete(name = "Type of gaps") +
-  scale_y_continuous(name = "Gaps in SD", expand = c(0, 0), lim = c(0, 3)) +
-  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
-  ggthemes::theme_few() +
-  theme(panel.spacing = unit(1, "lines"),
-        panel.grid.major.y = element_line(colour = "grey"),
-        legend.position = "bottom") +
-  ggtitle("Evolution of the different achievement gaps")
-@
-\caption{Evolution of the achievement gap for several gaps}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<percentage_change_8020,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
-perc_increase_fun(complete_data_topbottom_80) %>%
-  perc_graph("math",
-             "80/20 achievement gap",
-             "Percentage change from 2000 to 2015")
-@
-\caption{Percentage change in the 80/20 achievement gap from 2000 to 2015}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<percentage_change_7030,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
-perc_increase_fun(complete_data_topbottom_70) %>%
-  perc_graph("math",
-             "70/30 achievement gap",
-             "Percentage change from 2000 to 2015")
-@
-\caption{Percentage change in the 70/30 achievement gap from 2000 to 2015}
-\end{center}
-\end{figure}
-
-<<first_model_table>>=
-
+@
+
+% end here
+
+<<preparing_data>>=
+tracking_data <-
+  read_xlsx("./data/tracking.xlsx", sheet = "all_data") %>%
+  map_if(is_double, round, 2) %>%
+  as_tibble
+
+dif_data <- function(my_data, tracking, central, which_gap = "90th/10th SES gap") {
+  my_data %>%
+    filter(type_test == "math", type == which_gap) %>%
+    select(wave, country, difference) %>%
+    rename(year = wave) %>%
+    mutate(year = as.character(year)) %>%
+    left_join(tracking_data, by = c("country" = "cntry_name")) %>%
+    left_join(central_examination) %>%
+    mutate(num_tracks = ifelse(tracks15y == 1, 1, 0) %>% as.factor,
+           age_selection = selage) %>%
+    filter(!is.na(num_tracks), !is.na(age_selection), !is.na(difference))
+
+}
+formula_sequence <- function(dv, ivs, random_effect) {
+  formula_seq <-
+    lapply(1:length(ivs), function(x) {
+      paste0(dv ," ~ ", paste0(ivs[1:x], collapse = " + ") %>% paste0(" + ", random_effect))
+    }) %>%
+    lapply(as.formula)
+
+  formula_seq
+}
+stan_model_builder <- function(dv, iv, random, data) {
+  all_formulas <- formula_sequence(dv, iv, random)
+
+  mod_tracking <-
+    map(all_formulas, ~ {
+      brms::brm(
+        .x,
+        family = gaussian(),
+        data = data,
+        warmup = 1000, iter = 2000, chains = 5
+      )
+    })
+
+}
+stan_extractor <- function(models) {
+  coef_list <-
+    map(models, ~ {
+    broom::tidy(.x) %>%
+      .[grepl("^b_", .$term), ] %>%
+      map_if(is_double, round, 2) %>%
+      as_tibble %>%
+      transmute(term,
+                estimate = paste0(estimate, " (", lower, "/", upper, ")"))
+  })
+
+  coef_list
+}
+stan_table_builder <- function(models_extracted) {
+  intermediate_table <-
+    reduce(models_extracted, full_join, by = "term") %>%
+    setNames(c(" ", paste("Model", seq_len(models_extracted %>% length))))
+
+  row_order <-
+    setdiff(seq_len(nrow(intermediate_table)),
+            grep("Intercept", intermediate_table$` `)) %>%
+    `c`(grep("Intercept", intermediate_table$` `))
+
+    intermediate_table[row_order, ]
+}
+stan_table <- function(models) {
+
+  table_ready <-
+    models %>%
+    stan_extractor() %>%
+    stan_table_builder()
+
+  table_ready
+}
+
+table_details <- function(models) {
+
+  random_effect <- nrow(models[[length(models)]]$ranef) > 0
+
+  where_to_place <-
+    models[[length(models)]]$formula$formula %>%
+    as.character %>%
+    .[-(1:2)] %>%
+    stringr::str_split(" \\+ ") %>%
+    pluck(1) %>%
+    .[if (random_effect) -length(.) else 1:length(.)] %>%
+    length
+
+  r_square <-
+    paste0("\\hline ",
+           paste0("R square: & ",
+                  paste0(map(models, ~ bayes_R2(.x)[1] %>% round(2) * 100) %>% paste0("\\%"),
+                                     collapse = " & ")), " \\\\")
+  n_obs <-
+    paste0(paste0(" Sample size: & ",
+                  paste0(map(models, nobs), collapse = " & ")), " \\\\ \\hline")
+
+  command_to_row <- paste0(r_square, n_obs)
+  addtorow_two <- list(pos = list(pos = where_to_place + 1), command = command_to_row)
+
+  addtorow_two
+}
+
+ready_data_age <-
+  dif_data(complete_gaps, tracking_data, central_examination) %>%
+  mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)
+@
+
+The reason why countries differ in their evolution of inequality is still unknown.
+There is ample evidence showing the inequality between countries, to a certain extent, can be explained by the degree of tracking (Hanushek and Woessman).
+But we are also interested in what explains the between country evolution of inequality, that is, the explanation as to why in certain countries it is increasing more than others over time.
+
+This is clearly a difficult task given than, at least the country-level, not many features of the educational system change in a 15-year time span, resulting in a practically invariant variable from which to explain the level of inequality (without any variance, we can't explain anything.).
+However, there are other strategies that can at least partially help us to understand the phenomena, and these are precisely the ones we follow.
+
+Following the literature on the topic, we suspect that this evolution is in part very tightly related to the degree of stratification of the educational system.
+In a very simple frame-work, Hanushek and Woessman found that tracking indeed exarcebates inequality in the life course of children.
+Following the same strategy, we believe that tracking is also allowing the same gap in 15 year olds to grow over time, acompannied by other things.
+Yet, instead of concetrating only on the age of selection (just as they do), we explore a more fine grained definition of tracking, which is possible through the work of Thijs and Van der Werfhorst().
+Excluding the age of first selection into tracks, we also use the the number of tracks in the country, the percentage of the entire curriculum that is tracked and a vocational index \footnote{This index is a factor loading from a principal factor analysis of the percentage of students in upper secondary vocational education (taken from two sources, to reduce measurement error)}.
+
+Considering that our summarized data has very few observartions \footnote{We have per year, per country, one single inequality indicator, leaving the total sample size at \sexpr{nrow(ready_data_age)}}, we want to asses the uncertainty of our model by using a fully bayesian hierarchichal linear model.
+The benefits of this approach are twofold: first, we can asses the model fit much more intuitively, something very important given our sample size, and secondly, given that our countries are clustered into years, we can control for country heterogeneity within years, removing any bias in the standard errors.
+
+In the final model, we regress the inequality of a country on a dummy variable where 1 equals only 1 track vs more than 1 track, whether the age of first selection is 15 or more vs below 15, a dummy stating whether the percentage of tracked curriculum is above 1, and the standardized vocational index.
+We are reluctant to include more variable in the model, first to keep it more parsimonious, and secondly to prevent overfitting due to our sample size.
+We present the results in the table below.
+
+<<>>=
 # 2000
@@
 # 9 Missing
-library(readxl)
-library(inequalityintsvy)
-library(brms)
-
-tracking_data <-
-  read_xlsx("./paper/data/tracking.xlsx", sheet = "all_data") %>%
-  map_if(is_double, round, 2) %>%
-  as_tibble
 
@@
 
-dif_data <- function(my_data, tracking, central, which_gap = "90th/10th SES gap") {
-  my_data %>%
-    filter(type_test == "math", type == which_gap) %>%
-    select(wave, country, difference) %>%
-    rename(year = wave) %>%
-    mutate(year = as.character(year)) %>%
-    left_join(tracking_data, by = c("country" = "cntry_name")) %>%
-    left_join(central_examination) %>%
-    mutate(num_tracks = ifelse(tracks15y == 1, 1, 0) %>% as.factor,
-           age_selection = selage) %>%
-    filter(!is.na(num_tracks), !is.na(age_selection), !is.na(difference))
-
-}
-
-ready_data_age <-
-  dif_data(complete_gaps, tracking_data, central_examination) %>%
-  mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)
-
-ivs <- c("num_tracks", "age_selection", "I(length > 0)", "zvoc")
-
-formula_seq <-
-  lapply(1:length(ivs), function(x) {
-    paste0("difference ~ ", paste0(ivs[1:x], collapse = " + ") %>% paste0(" + (1|year)"))
-  }) %>%
-  lapply(as.formula)
-
+@
+
+<<complete_tracking_model>>=
 all_track_models <-
-  map(formula_seq, ~ {
-
-  mod_tracking <-
-  brms::brm(
-    .x, family = gaussian(),
-    data = ready_data_age,
-    warmup = 1000, iter = 2000, chains = 5,
-    control = list(adapt_delta = 0.90)
+  stan_model_builder(
+  dv = "difference",
+  iv = c("num_tracks", "age_selection", "I(length > 0)", "zvoc"),
+  random = "(1 | year)",
+  data = ready_data_age
 )
 
-  mod_tracking
-
-})
-
-tidy_dfs <-
-  map(all_track_models, ~ {
-
-  broom::tidy(.x) %>%
-    .[grepl("^b_", .$term), ] %>%
-    map_if(is_double, round, 2) %>%
-    as_tibble %>%
-    transmute(term,
-              estimate = paste0(estimate, " (", lower, "/", upper, ")"))
-  })
-
-tidy_dfs_two <-
-  reduce(tidy_dfs, full_join, by = "term") %>%
-  setNames(c(" ", paste("Model", 1:4))) %>%
+all_track_models %>%
+  stan_table() %>%
   mutate(" " = c(
-    "Intercept",
-    "Num of tracks = 1",
-    "Age at track >= 15",
-    "% curriculum tracked",
-    "Vocational Index"
+    "Only 1 track",
+    "Age selection >= 15",
+    "% of curric tracked",
+    "Vocational Index",
+    "Intercept"
   )) %>%
-  .[c(2:5, 1), ]
-
-r_square <-
-  paste0("\\hline ", paste0("R square: & ",
-         paste0(map(all_track_models, ~ bayes_R2(.x)[1] %>% round(2) * 100) %>% paste0("\\%"),
-                       collapse = " & ")), " \\\\")
-n_obs <-
-  paste0(paste0(" Sample size: & ",
-         paste0(map(all_track_models, nobs), collapse = " & ")), " \\\\ \\hline")
-
-command_to_row <- paste0(r_square, n_obs)
-
-addtorow_two <- list(pos = list(pos = 5), command = command_to_row)
-
-xtable(tidy_dfs_two) %>% print.xtable(
+  xtable() %>%
+  print(
   include.rownames = FALSE,
-  add.to.row = addtorow_two
+  add.to.row = table_details(all_track_models)
   )
@@
 
-<<>>=
+Given that the classical p values are inexistent in bayesian modelling, I present uncertainty intervals next to each coefficient.
+The first model only includes the 1 track dummy.
+This variable shows that countries that have one track, have on average, half a standard deviation less inequality than other countries.
+This covariate alone explains about 23\% of the variance of the evolution of the 90/10 achievement gap.
+The second model now includes the dummy of whether the country has an age selection of 15 or more and we can see that countries with late selection into tracking have about -.38 less standard deviations in inequality.
+Also note that the number of tracks dummy was also reduced, suggesting these two variables are explaining similar things.
+This is expected given that late tracking is usually associated with fewer tracks, such as in the scandinavian countries.
+The third model now includes whether the country has any degree of tracking, meaning whether the country has above 0\% of the curriculum tracked.
+The coding of the variable comes from visual inspection of the distribution, where some countries, such as the United States and Canada, have untracked curriculums.
+We see now relationship at all, but after including the vocational index, the model is completely reversed with countries with tracked curriculum having about 0.6 more standard deviations in inequality that countries with no tracking.
+
+We also note that vocational enrollment is associated with less inequality.
+Not surprisingly, the age of selection coefficient, also increased due to the correlation between age of selection and vocational enrollment.
+Countries with high vocational enrollment are also those with early age selection, such as Germany and Austria.
+These two variables should be inspected further, perhaps with an interaction, given than both coefficients are sizable and very correlated.
+This final model explains over one third of the variance in the evolution of the 90/10 SES gap, with an R square of 40\%.
+The uncertainty of each of the covariates varies a lot with the percentage of curriculum tracked dummy ranging all the way from 0.9 to 0.3.
+Nonetheless, we see that the other coefficients are very strong and have reasonable intervals.
+It is very important to consider the uncertainty of our estimates, to give confidence in our model fit.
+So far, this model seems to fit the data quite reasonably.
+\footnote{These variables are very correlated, but no collinearity is present with a maximum VIF of 2.2.}
+
+All of these countries have a different tracking structure, with some setups being more egalitarian than others.
+Using the previous model we make a simulation and predict the level of the 90/10 achievement gap for all possible tracking setups for all countries.
+Below we plot all these combinations.
+The x axis has the number of tracks that a country has, the y axis has the predicted 90/10 achievement gap and below each country name we have whether the country has an age selection of 15 or more or below.
+In addition, to confirm how accurate is the model, we also plot the actual level of inequality in the actual tracking setup of the country.
+Let's take Germany for example.
+
+<<graph_complete_tracking_model>>=
 # This is the closest I could get to getting vif's fro the last model:
 
-lm(difference ~ num_tracks + age_selection + I(length > 0) + zvoc,
+our_formula <- difference ~ num_tracks + age_selection + I(length > 0) + zvoc
+lm(our_formula,
    data = ready_data_age) %>%
@@
 mod_tracking_ext <-
-  brms::brm(formula_seq[[4]],
+  brms::brm(our_formula,
             family = gaussian(),
@@
   ggplot(aes(num_tracks, avg_diff, fill = num_tracks)) +
+  geom_point(colour = "red") +
   geom_errorbar(aes(ymin = diff_low, ymax = diff_high), width = 0.2, colour = "red") +
-  geom_point(colour = "red") +
   geom_col(aes(y = pred, alpha = 0.4)) +
@@
 
-<<>>=
+The red point for Germany is in the '> 1 track' and '< 15' setup, meaning that Germany has a tracking system that has more than 1 track and an age of selection of below 15.
+We see that the model predicts the level of inequality for that set up quite accurately, relative to the actual level (red point).
+But if the setup would be in the ideal '1 track' and age of selection above or equal to 15, the inequality would be much lower.
+Generally speaking, this applies to virtually every country in the plot.
+Virtually for all countries which are not in this ideal tracking, a switch to the ideal would reflect a reduction in the achievement gap.
+
+For those countries which are not in the 'ideal tracking' setup, we compute below the percentage change from switching to the ideal tracking. The results are in the table below.
+
+<<counterfactual_tracking>>=
 pred_avg_diff<-
@@
 
-xtable(final_avg_diff_table) %>% print.xtable(
+xtable(final_avg_diff_table) %>%
+  print(
   include.rownames = FALSE
@@
 
+The biggest reductions are for both Hungary and Germany, with 46\% and 36\% respectively.
+France, which had a particularly steep increase in the last 15 years, would see a reduction of the 90/10 SES gap of 31\% and Austria, with a very similar tracking structure as the German one, would see their achievement gap reduced by 20\%.
+The very suprising result is the Dutch case, for which a switch from the 'worst' tracking setup to the 'best' case would result in an increase of the gap of 20\%.
+
+This result could mean two things.
+First and foremost, the Dutch educational system has built other mechanisms that, despite the tracking system, help children from social origin escalate socially.
+This explanation is not very speculative given than in most mobility research, the Netherlands stands out as a very equal country relative to other countries with a similary tracking structure, like Germany or Hungary (cite Gosta study and something from Breen and Goldthorpe).
+Another explanation, and something which needs to assessed, is whether our model is predicting accurately.
+If we look at the plot from above, we see that for some countries, the model is a bit off from the actual level of inequality.
+Moreover, although we see that if all countries switched to the 'ideal' tracking setup we would see a vast reduction in inequality, we also see that each of these predictions is followed by a wide uncertainty interval.
+Considering that the results are very consistent for every country, and in most countries the prediction matches the actual value, this could simply be a consequence of low sample size.
+Whether these results stand the test of a different empirical strategy or a higher sample size, we leave up to future research.
+
+A more summarized empirical strategy is to actually standardize all the tracking indicator into one index, something that Bol and Van der Werfhorst() already do quite nicely.
+Using this standardized tracking variable and the same bayesian multievel model, we now model the tracking and vocational index as before, but add an interaction between two, given that in the previous model we recorded a strong shift in the model when tracking-related variables and the vocational model we both included in the same model.
+Below are the results for this model.
+
 <<interaction_model>>=
+
+sig_vs_sig <- function(coef1, se1, coef2, se2) {
+  diff <- c(coef1, se1) - c(coef2, se2)
+  abs(diff[1]) - 2 * diff[2]
+}
+
 ready_data <- dif_data(complete_gaps, tracking_data, central_examination)
 
-mod_tracking <-
-  brms::brm(
-    difference ~
-      ztrack +
-      zvoc +
-      ztrack:zvoc, family = gaussian(),
-  data = ready_data_age,
-  warmup = 1000, iter = 2000, chains = 5
+stan_models <-
+  stan_model_builder(
+  dv = "difference",
+  iv = c("ztrack", "zvoc", "ztrack:zvoc"),
+  random = "(1 | year)",
+  data = ready_data
   )
 
-bayes_R2(mod_tracking)
-
+stan_models %>%
+  stan_table() %>%
+  mutate(" " = c(
+    "Tracking Index",
+    "Vocational Index",
+    "Trackin * Vocational Index",
+    "Intercept"
+  )) %>%
+  xtable() %>%
+  print.xtable(
+    include.rownames = FALSE,
+    add.to.row = table_details(stan_models)
+  )
+@
+
+We confirm the results from the previous models, where more tracking is associated with an achievement gap of about 0.4 standard deviations wider and more vocational enrollment is associated with a reduction of about 0.2 standard deviations, half the slope of tracking.
+Meaning that the tracking effect might be offsetting the effect of the vocational index given that it's half the size of the slope.
+Both these coefficients are actually significantly different from each other, so it is possible.
+Before we explore this visually, let's note that the interaction between the two is very reliable and significant, considering that small sample size.
+Below we graph the interaction for different quantiles of the vocational index.
+
+<<interaction_plot>>=
 plot_interaction <-
-  marginal_effects(mod_tracking,
+  marginal_effects(stan_models[[length(stan_models)]],
                  effects = "ztrack:zvoc",
@@
 plot(plot_interaction, plot = FALSE)[[1]] +
-  labs(x = "Tracking Index (SD)", y = "Difference in test scores (SD)") +
+  labs(x = "Tracking Index (SD)", y = "90/10 Achievement gap (SD)") +
   scale_colour_discrete(
@@
 
+The results are very interesting.
+We see that for lower levels of tracking (meaning little to no tracking) theres a very low achievement gap, regardless of whether the country has very high or very low vocational enrollment.
+However, once tracking enters the picture, vocational enrollment can be a strong equalizer.
+We see that the bottom quantile of vocational enrollment has almost twice the achievement gap as the top quantile for vocational enrollment.
+Moreover, our earlier interpretation (that tracking is more important than vocational enrollment for reducing the achievement gap) still holds and is very evident here.
+Despite the equalizing power of vocational enrollment, a high level of tracking with the highest level of vocational enrollment still leaves a country with an achievement of over 1.5 standard deviations.
+However, a country with low levels of tracking and \textit{any} level of vocational enrollment is nearly 1.3 standard deviations.
+A key finding from these results is that tracking is reasonably explaining the 90/10 achievement gap, and we should consider it as a very important component in the reduction of the achievement gap.
+However, it should not be studied in isolation from other features such as the vocational enrollment.
+We see a very sizable change in the achievement gap for a given country with the same level of tracking and high and low levels of vocational enrollment.
+
+% We need to define key numbers, to refer in the abstract, like the achievement gap closes % percent in the counterfactual and there's a difference of % percent between same level of tracking but different vocational enrollments.
+
+To test whether these results hold for other samples, we compute the same models as before but also for the 80/20 and 70/30 SES gap. Up next we plot these coefficients for the three SES gaps.
+
 <<all_gaps_plot>>=
 gaps <- c("90th/10th SES gap", "80th/20th SES gap", "70th/30th SES gap")
+
 all_gaps_models <-
@@
 
+With the exception of the Intercept, the coefficients of the three models look pretty much the same.
+The tracking index seems a but stronger for the 90/10 model but the uncertainty intervals overlap with the other two models.
+These results give our claims much more credibility.
+
+Finally, until now, we've studied how we can explain the evolution of the 90/10 achievement gap, by clustering on the year dimension of each country and then performing a linear multilevel model.
+This approach was adopted given that tracking hardly changes over time, making it impossible to model the change over time directly.
+However, there is another way of estimating the evolution of the achievement gap, which we try up next.
+
+Instead of estimating the evolution over time, we calculate the cumulative change over time for each country, summarizing the 6 year-time span of achievement gaps into one cumulative inequality gap.
+This approach dramatically reduces the total sample size to 34, the number of total countries available \footnote{We've been presenting a selected number of countries given the lack of space, but for every estimation we include all countries available.}.
+It is for this reason that I only include at most 2 covariates in the model, the tracking and vocational index, which both summarize the previous models and keep the model parsimonious.
+
 <<cumulative_modeling_table>>=
@@
 
-tidy_dfs <-
-  map(all_gaps_models_cum, ~ {
-
-  broom::tidy(.x) %>%
-    .[grepl("^b_", .$term), ] %>%
-    mutate_if(is_double, round, 2) %>%
-    transmute(term,
-              estimate = paste0(estimate, " (", lower, "/", upper, ")"))
-  })
-
-tidy_dfs_two <-
-  reduce(tidy_dfs, full_join, by = "term") %>%
+all_gaps_models_cum %>%
+  stan_table() %>%
+  mutate(" " = c(
+    "Track Index",
+    "Vocational Index",
+    "Intercept"
+  )) %>%
   setNames(c(" ", gsub(" SES gap", "", gaps))) %>%
-  mutate(" " = c(
-    "Intercept",
-    "Track Index",
-    "Vocational Index"
-  )) %>%
-  .[c(2, 3, 1), ]
-
-r_square <-
-  paste0("\\hline ", paste0("R square: & ",
-         paste0(map(all_gaps_models_cum, ~ bayes_R2(.x)[1] %>% round(2) * 100) %>% paste0("\\%"),
-                       collapse = " & ")), " \\\\")
-n_obs <-
-  paste0(paste0(" Sample size: & ",
-         paste0(map(all_gaps_models_cum, nobs), collapse = " & ")), " \\\\ \\hline")
-
-command_to_row <- paste0(r_square, n_obs)
-
-addtorow_two <- list(pos = list(pos = 3), command = command_to_row)
-
-xtable(tidy_dfs_two) %>%
+  xtable() %>%
   print.xtable(
     include.rownames = FALSE,
-    add.to.row = addtorow_two
+    add.to.row = table_details(all_gaps_models_cum)
   )
-@
-
-<<>>=
+
+@
+
+We see that 1 standard deviation increase in the tracking index widens the 90/10 SES gap by 1.66 standard deviations over time, and the vocational index reduces it by 1 standard deviation and these two covariates explain nearly one third of the cumulative gap.
+As we decrease the achievement gap (80/20, 70/30) these coefficients become smaller, but keep their strength and their predictive power.
+Comparing this model to the previous models, we see a very similar landscape, but an even more reassuring one, given than we're tackling the evolution of the achievement gap directly with this indicator.
+We can safely say that these results, at worse, are supporting the previous findings and making the claim even more credible.
+
+<<next_steps>>=
+# Next steps:
+
+# Continue by doing the multilevel models to see what explains what. Include
+# all indicators from the reardon/russian girl paper.
+
+# Graph the increase in each country vs the increase/decrease of the economic inequality indicators
+# Specially the 90/10
+
+# Calculate how big is the gap between reading and math
+
+# Continue with the PIRLS to see if there are specific patterns in 4th and 8th graders gap.
+
+# Get each country trendline adjusted for the inequality indicators and place in the same country graph.
+
+# Should I add the parent's education in the lm model to see how trends change adjusted for that?
+
+# A weak welfare system, together with income inequality, what's their pattern?
+# What if we put the school differentiation/tracking aspect in? Are there country groups based on these
+# patterns.
+
+# In countries where there is high differentiation/tracking, is there a jump in the evolution of the gap between PIRLS/TIMSS and PISA?
+@
+
+\bibliography{mybibliography.bib}
+
+\pagebreak
+
+\section{Appendix}
+
+\begin{figure}
+\begin{center}
+<<graph_all_gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+complete_gaps %>%
+  mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
+  filter(type_test == "math",
+         country %in% countries) %>%
+  ggplot(aes(as.character(wave), difference, colour = type, linetype = type, group = type)) +
+  geom_line(stat = "smooth", method = "lm", size = 0.7) +
+  facet_wrap(~ country, ncol = 5) +
+  scale_colour_discrete(name = "Type of gaps") +
+  scale_linetype_discrete(name = "Type of gaps") +
+  scale_y_continuous(name = "Gaps in SD", expand = c(0, 0), lim = c(0, 3)) +
+  scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
+  ggthemes::theme_few() +
+  theme(panel.spacing = unit(1, "lines"),
+        panel.grid.major.y = element_line(colour = "grey"),
+        legend.position = "bottom") +
+  ggtitle("Evolution of the different achievement gaps")
+@
+\caption{Evolution of the achievement gap for several gaps}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<percentage_change_8020,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+perc_increase_fun(complete_data_topbottom_80) %>%
+  perc_graph("math",
+             "80/20 achievement gap",
+             "Percentage change from 2000 to 2015")
+@
+\caption{Percentage change in the 80/20 achievement gap from 2000 to 2015}
+\end{center}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<percentage_change_7030,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+perc_increase_fun(complete_data_topbottom_70) %>%
+  perc_graph("math",
+             "70/30 achievement gap",
+             "Percentage change from 2000 to 2015")
+@
+\caption{Percentage change in the 70/30 achievement gap from 2000 to 2015}
+\end{center}
+\end{figure}
+
+<<eval = F>>=
 # pp_check(mod_tracking, type = "dens_overlay", nsamples = 100)

### 2017-09-07 paper/paper.Rnw
Fixed all tables in the output document

@@
 
-Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive and personality definition \citep{duyme1999, waldfogel1996}.
+Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive and personality definition \citep{duyme1999, waldfogel2006}.
 Economists and educators have long neglected this branch of literature but the work of James Heckman has brought our attention to it.
@@
 
-dif_data <- function(my_data, tracking, central, which_gap = "90th/10th SES gap") {
+dif_data <- function(my_data, tracking, which_gap = "90th/10th SES gap") {
   my_data %>%
@@
     left_join(tracking_data, by = c("country" = "cntry_name")) %>%
-    left_join(central_examination) %>%
     mutate(num_tracks = ifelse(tracks15y == 1, 1, 0) %>% as.factor,
@@
 ready_data_age <-
-  dif_data(complete_gaps, tracking_data, central_examination) %>%
+  dif_data(complete_gaps, tracking_data) %>%
   mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)
@@
 
-Considering that our summarized data has very few observartions \footnote{We have per year, per country, one single inequality indicator, leaving the total sample size at \sexpr{nrow(ready_data_age)}}, we want to asses the uncertainty of our model by using a fully bayesian hierarchichal linear model.
+Considering that our summarized data has very few observartions \footnote{We have per year, per country, one single inequality indicator, leaving the total sample size at \Sexpr{nrow(ready_data_age)}}, we want to asses the uncertainty of our model by using a fully bayesian hierarchichal linear model.
 The benefits of this approach are twofold: first, we can asses the model fit much more intuitively, something very important given our sample size, and secondly, given that our countries are clustered into years, we can control for country heterogeneity within years, removing any bias in the standard errors.
@@
 
-<<complete_tracking_model>>=
+<<complete_tracking_model, results = 'hide'>>=
 all_track_models <-
@@
 )
-
+@
+
+<<complete_tracking_results, include = TRUE>>=
 all_track_models %>%
@@
 
-<<graph_complete_tracking_model>>=
+<<graph_complete_tracking_model, results = 'hide'>>=
 # This is the closest I could get to getting vif's fro the last model:
@@
   )
-
+@
+
+<<graph_complete_tracking_graph, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
 new_data <-
@@
 
-The red point for Germany is in the '> 1 track' and '< 15' setup, meaning that Germany has a tracking system that has more than 1 track and an age of selection of below 15.
+The red point for Germany is in the '\textgreater 1 track' and '\textless 15' setup, meaning that Germany has a tracking system that has more than 1 track and an age of selection of below 15.
 We see that the model predicts the level of inequality for that set up quite accurately, relative to the actual level (red point).
@@
 
-<<counterfactual_tracking>>=
+<<counterfactual_tracking, include = TRUE>>=
 pred_avg_diff<-
@@
 
-<<interaction_model>>=
+<<interaction_model, results = 'hide'>>=
 
@@
 
-ready_data <- dif_data(complete_gaps, tracking_data, central_examination)
+ready_data <- dif_data(complete_gaps, tracking_data)
 
@@
   )
-
+@
+
+<<interaction_table,  include = TRUE>>=
 stan_models %>%
@@
 
-<<interaction_plot>>=
+<<interaction_plot, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
 plot_interaction <-
@@
 
-<<all_gaps_plot>>=
+<<all_gaps_model, results = 'hide'>>=
 gaps <- c("90th/10th SES gap", "80th/20th SES gap", "70th/30th SES gap")
@@
     ready_data_age <-
-      dif_data(complete_gaps, tracking_data, central_examination, which_gap = gap) %>%
+      dif_data(complete_gaps, tracking_data, which_gap = gap) %>%
       mutate(age_selection = ifelse(selage >= 15, 1, 0) %>% as.factor)
@@
 })
-
+@
+
+<<all_gaps_plot, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
 model_data_merged <-
@@
 
-<<cumulative_modeling_table>>=
+<<cumulative_modeling, results = 'hide'>>=
 
@@
 
-vars_unique <- c("length", "num_tracks", "age_selection", "central_examination", "ztrack",
+vars_unique <- c("length", "num_tracks", "age_selection", "ztrack",
                  "zvoc")
@@
 
-    gap_data <- dif_data(complete_gaps, tracking_data, central_examination, which_gap = gap)
+    gap_data <- dif_data(complete_gaps, tracking_data, which_gap = gap)
 
@@
 })
-
+@
+
+<<cumulative_modeling_table, include = TRUE>>=
 all_gaps_models_cum %>%

### 2017-09-07 paper/paper.tex
Fixed all tables in the output document

@@
 Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap.
-Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentially, are countries that have the lower leven of inequalities among the ones presented here.
+Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentially, are countries that have the lower level of inequalities among the ones presented here.
+
+\section{The role of curricular tracking in explaining inequalities}
+
+% 90/80/70 gaps merged
+
+% end here
+
+The reason why countries differ in their evolution of inequality is still unknown.
+There is ample evidence showing the inequality between countries, to a certain extent, can be explained by the degree of tracking (Hanushek and Woessman).
+But we are also interested in what explains the between country evolution of inequality, that is, the explanation as to why in certain countries it is increasing more than others over time.
+
+This is clearly a difficult task given than, at least the country-level, not many features of the educational system change in a 15-year time span, resulting in a practically invariant variable from which to explain the level of inequality (without any variance, we can't explain anything.).
+However, there are other strategies that can at least partially help us to understand the phenomena, and these are precisely the ones we follow.
+
+Following the literature on the topic, we suspect that this evolution is in part very tightly related to the degree of stratification of the educational system.
+In a very simple frame-work, Hanushek and Woessman found that tracking indeed exarcebates inequality in the life course of children.
+Following the same strategy, we believe that tracking is also allowing the same gap in 15 year olds to grow over time, acompannied by other things.
+Yet, instead of concetrating only on the age of selection (just as they do), we explore a more fine grained definition of tracking, which is possible through the work of Thijs and Van der Werfhorst().
+Excluding the age of first selection into tracks, we also use the the number of tracks in the country, the percentage of the entire curriculum that is tracked and a vocational index \footnote{This index is a factor loading from a principal factor analysis of the percentage of students in upper secondary vocational education (taken from two sources, to reduce measurement error)}.
+
+Considering that our summarized data has very few observartions \footnote{We have per year, per country, one single inequality indicator, leaving the total sample size at 195}, we want to asses the uncertainty of our model by using a fully bayesian hierarchichal linear model.
+The benefits of this approach are twofold: first, we can asses the model fit much more intuitively, something very important given our sample size, and secondly, given that our countries are clustered into years, we can control for country heterogeneity within years, removing any bias in the standard errors.
+
+In the final model, we regress the inequality of a country on a dummy variable where 1 equals only 1 track vs more than 1 track, whether the age of first selection is 15 or more vs below 15, a dummy stating whether the percentage of tracked curriculum is above 1, and the standardized vocational index.
+We are reluctant to include more variable in the model, first to keep it more parsimonious, and secondly to prevent overfitting due to our sample size.
+We present the results in the table below.
+
+% latex table generated in R 3.4.0 by xtable 1.8-2 package
+% Thu Sep  7 15:04:53 2017
+\begin{table}[ht]
+\centering
+\begin{tabular}{lllll}
+  \hline
+  & Model 1 & Model 2 & Model 3 & Model 4 \\ 
+  \hline
+Only 1 track & -0.53 (-0.64/-0.41) & -0.31 (-0.44/-0.18) & -0.31 (-0.45/-0.17) & -0.21 (-0.34/-0.07) \\ 
+  Age selection $>$= 15 &  & -0.38 (-0.51/-0.25) & -0.38 (-0.51/-0.26) & -0.57 (-0.71/-0.43) \\ 
+  \% of curric tracked &  &  & 0 (-0.23/0.23) & 0.6 (0.3/0.89) \\ 
+  Vocational Index &  &  &  & -0.27 (-0.36/-0.19) \\ 
+  Intercept & 1.77 (1.69/1.84) & 1.94 (1.85/2.03) & 1.94 (1.69/2.19) & 1.54 (1.27/1.81) \\ 
+   \hline R square: & 23\% & 32\% & 32\% & 40\% \\ Sample size: & 195 & 195 & 195 & 183 \\ \hline \hline
+\end{tabular}
+\end{table}
+
+Given that the classical p values are inexistent in bayesian modelling, I present uncertainty intervals next to each coefficient.
+The first model only includes the 1 track dummy.
+This variable shows that countries that have one track, have on average, half a standard deviation less inequality than other countries.
+This covariate alone explains about 23\% of the variance of the evolution of the 90/10 achievement gap.
+The second model now includes the dummy of whether the country has an age selection of 15 or more and we can see that countries with late selection into tracking have about -.38 less standard deviations in inequality.
+Also note that the number of tracks dummy was also reduced, suggesting these two variables are explaining similar things.
+This is expected given that late tracking is usually associated with fewer tracks, such as in the scandinavian countries.
+The third model now includes whether the country has any degree of tracking, meaning whether the country has above 0\% of the curriculum tracked.
+The coding of the variable comes from visual inspection of the distribution, where some countries, such as the United States and Canada, have untracked curriculums.
+We see now relationship at all, but after including the vocational index, the model is completely reversed with countries with tracked curriculum having about 0.6 more standard deviations in inequality that countries with no tracking.
+
+We also note that vocational enrollment is associated with less inequality.
+Not surprisingly, the age of selection coefficient, also increased due to the correlation between age of selection and vocational enrollment.
+Countries with high vocational enrollment are also those with early age selection, such as Germany and Austria.
+These two variables should be inspected further, perhaps with an interaction, given than both coefficients are sizable and very correlated.
+This final model explains over one third of the variance in the evolution of the 90/10 SES gap, with an R square of 40\%.
+The uncertainty of each of the covariates varies a lot with the percentage of curriculum tracked dummy ranging all the way from 0.9 to 0.3.
+Nonetheless, we see that the other coefficients are very strong and have reasonable intervals.
+It is very important to consider the uncertainty of our estimates, to give confidence in our model fit.
+So far, this model seems to fit the data quite reasonably.
+\footnote{These variables are very correlated, but no collinearity is present with a maximum VIF of 2.2.}
+
+All of these countries have a different tracking structure, with some setups being more egalitarian than others.
+Using the previous model we make a simulation and predict the level of the 90/10 achievement gap for all possible tracking setups for all countries.
+Below we plot all these combinations.
+The x axis has the number of tracks that a country has, the y axis has the predicted 90/10 achievement gap and below each country name we have whether the country has an age selection of 15 or more or below.
+In addition, to confirm how accurate is the model, we also plot the actual level of inequality in the actual tracking setup of the country.
+Let's take Germany for example.
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/graph_complete_tracking_graph-1}
+
+}
+
+The red point for Germany is in the '\textgreater 1 track' and '\textless 15' setup, meaning that Germany has a tracking system that has more than 1 track and an age of selection of below 15.
+We see that the model predicts the level of inequality for that set up quite accurately, relative to the actual level (red point).
+But if the setup would be in the ideal '1 track' and age of selection above or equal to 15, the inequality would be much lower.
+Generally speaking, this applies to virtually every country in the plot.
+Virtually for all countries which are not in this ideal tracking, a switch to the ideal would reflect a reduction in the achievement gap.
+
+For those countries which are not in the 'ideal tracking' setup, we compute below the percentage change from switching to the ideal tracking. The results are in the table below.
+
+% latex table generated in R 3.4.0 by xtable 1.8-2 package
+% Thu Sep  7 15:05:38 2017
+\begin{table}[ht]
+\centering
+\begin{tabular}{lllrrl}
+  \hline
+Country & Number of tracks & Age of selection & Inequality & Predicted & \% reduction \\ 
+  \hline
+Austria & More than one track & Less than 15 & 1.58 & 1.27 & 20\% \\ 
+  France & More than one track & 15 or more & 1.84 & 1.27 & 31\% \\ 
+  Germany & More than one track & Less than 15 & 1.95 & 1.27 & 35\% \\ 
+  Hungary & More than one track & Less than 15 & 2.32 & 1.26 & 46\% \\ 
+  Italy & More than one track & Less than 15 & 1.48 & 1.26 & 15\% \\ 
+  Netherlands & More than one track & Less than 15 & 1.06 & 1.27 & -19\% \\ 
+  Poland & More than one track & 15 or more & 1.44 & 1.27 & 12\% \\ 
+   \hline
+\end{tabular}
+\end{table}
+
+The biggest reductions are for both Hungary and Germany, with 46\% and 36\% respectively.
+France, which had a particularly steep increase in the last 15 years, would see a reduction of the 90/10 SES gap of 31\% and Austria, with a very similar tracking structure as the German one, would see their achievement gap reduced by 20\%.
+The very suprising result is the Dutch case, for which a switch from the 'worst' tracking setup to the 'best' case would result in an increase of the gap of 20\%.
+
+This result could mean two things.
+First and foremost, the Dutch educational system has built other mechanisms that, despite the tracking system, help children from social origin escalate socially.
+This explanation is not very speculative given than in most mobility research, the Netherlands stands out as a very equal country relative to other countries with a similary tracking structure, like Germany or Hungary (cite Gosta study and something from Breen and Goldthorpe).
+Another explanation, and something which needs to assessed, is whether our model is predicting accurately.
+If we look at the plot from above, we see that for some countries, the model is a bit off from the actual level of inequality.
+Moreover, although we see that if all countries switched to the 'ideal' tracking setup we would see a vast reduction in inequality, we also see that each of these predictions is followed by a wide uncertainty interval.
+Considering that the results are very consistent for every country, and in most countries the prediction matches the actual value, this could simply be a consequence of low sample size.
+Whether these results stand the test of a different empirical strategy or a higher sample size, we leave up to future research.
+
+A more summarized empirical strategy is to actually standardize all the tracking indicator into one index, something that Bol and Van der Werfhorst() already do quite nicely.
+Using this standardized tracking variable and the same bayesian multievel model, we now model the tracking and vocational index as before, but add an interaction between two, given that in the previous model we recorded a strong shift in the model when tracking-related variables and the vocational model we both included in the same model.
+Below are the results for this model.
+
+% latex table generated in R 3.4.0 by xtable 1.8-2 package
+% Thu Sep  7 15:07:49 2017
+\begin{table}[ht]
+\centering
+\begin{tabular}{llll}
+  \hline
+  & Model 1 & Model 2 & Model 3 \\ 
+  \hline
+Tracking Index & 0.26 (0.21/0.32) & 0.33 (0.27/0.4) & 0.41 (0.34/0.48) \\ 
+  Vocational Index &  & -0.19 (-0.26/-0.11) & -0.22 (-0.29/-0.15) \\ 
+  Trackin * Vocational Index &  &  & -0.12 (-0.17/-0.06) \\ 
+  Intercept & 1.6 (1.53/1.67) & 1.65 (1.58/1.72) & 1.71 (1.63/1.78) \\ 
+   \hline R square: & 24\% & 30\% & 34\% \\ Sample size: & 195 & 183 & 183 \\ \hline \hline
+\end{tabular}
+\end{table}
+
+We confirm the results from the previous models, where more tracking is associated with an achievement gap of about 0.4 standard deviations wider and more vocational enrollment is associated with a reduction of about 0.2 standard deviations, half the slope of tracking.
+Meaning that the tracking effect might be offsetting the effect of the vocational index given that it's half the size of the slope.
+Both these coefficients are actually significantly different from each other, so it is possible.
+Before we explore this visually, let's note that the interaction between the two is very reliable and significant, considering that small sample size.
+Below we graph the interaction for different quantiles of the vocational index.
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/interaction_plot-1}
+
+}
+
+The results are very interesting.
+We see that for lower levels of tracking (meaning little to no tracking) theres a very low achievement gap, regardless of whether the country has very high or very low vocational enrollment.
+However, once tracking enters the picture, vocational enrollment can be a strong equalizer.
+We see that the bottom quantile of vocational enrollment has almost twice the achievement gap as the top quantile for vocational enrollment.
+Moreover, our earlier interpretation (that tracking is more important than vocational enrollment for reducing the achievement gap) still holds and is very evident here.
+Despite the equalizing power of vocational enrollment, a high level of tracking with the highest level of vocational enrollment still leaves a country with an achievement of over 1.5 standard deviations.
+However, a country with low levels of tracking and \textit{any} level of vocational enrollment is nearly 1.3 standard deviations.
+A key finding from these results is that tracking is reasonably explaining the 90/10 achievement gap, and we should consider it as a very important component in the reduction of the achievement gap.
+However, it should not be studied in isolation from other features such as the vocational enrollment.
+We see a very sizable change in the achievement gap for a given country with the same level of tracking and high and low levels of vocational enrollment.
+
+% We need to define key numbers, to refer in the abstract, like the achievement gap closes % percent in the counterfactual and there's a difference of % percent between same level of tracking but different vocational enrollments.
+
+To test whether these results hold for other samples, we compute the same models as before but also for the 80/20 and 70/30 SES gap. Up next we plot these coefficients for the three SES gaps.
+
+{\centering \includegraphics[width=5.5in,height=5in]{figure/all_gaps_plot-1}
+
+}
+
+With the exception of the Intercept, the coefficients of the three models look pretty much the same.
+The tracking index seems a but stronger for the 90/10 model but the uncertainty intervals overlap with the other two models.
+These results give our claims much more credibility.
+
+Finally, until now, we've studied how we can explain the evolution of the 90/10 achievement gap, by clustering on the year dimension of each country and then performing a linear multilevel model.
+This approach was adopted given that tracking hardly changes over time, making it impossible to model the change over time directly.
+However, there is another way of estimating the evolution of the achievement gap, which we try up next.
+
+Instead of estimating the evolution over time, we calculate the cumulative change over time for each country, summarizing the 6 year-time span of achievement gaps into one cumulative inequality gap.
+This approach dramatically reduces the total sample size to 34, the number of total countries available \footnote{We've been presenting a selected number of countries given the lack of space, but for every estimation we include all countries available.}.
+It is for this reason that I only include at most 2 covariates in the model, the tracking and vocational index, which both summarize the previous models and keep the model parsimonious.
+
+% latex table generated in R 3.4.0 by xtable 1.8-2 package
+% Thu Sep  7 15:11:51 2017
+\begin{table}[ht]
+\centering
+\begin{tabular}{llll}
+  \hline
+  & 90th/10th & 80th/20th & 70th/30th \\ 
+  \hline
+Track Index & 1.66 (0.85/2.49) & 1.11 (0.46/1.78) & 1 (0.37/1.67) \\ 
+  Vocational Index & -1.04 (-1.95/-0.12) & -0.9 (-1.64/-0.14) & -0.87 (-1.6/-0.16) \\ 
+  Intercept & 9.37 (8.6/10.12) & 4.6 (3.99/5.19) & 3.89 (3.3/4.47) \\ 
+   \hline R square: & 30\% & 25\% & 23\% \\ Sample size: & 32 & 32 & 32 \\ \hline \hline
+\end{tabular}
+\end{table}
+
+We see that 1 standard deviation increase in the tracking index widens the 90/10 SES gap by 1.66 standard deviations over time, and the vocational index reduces it by 1 standard deviation and these two covariates explain nearly one third of the cumulative gap.
+As we decrease the achievement gap (80/20, 70/30) these coefficients become smaller, but keep their strength and their predictive power.
+Comparing this model to the previous models, we see a very similar landscape, but an even more reassuring one, given than we're tackling the evolution of the achievement gap directly with this indicator.
+We can safely say that these results, at worse, are supporting the previous findings and making the claim even more credible.
 

### 2017-09-08 paper/paper.Rnw
Reread and corrected half the paper

@@
 Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years.
-The idea of meritocracy and intergenerational transfers have motivated, in conservative terms, much of the research on social mobility and social inequalities, for a good part of the 20th century and 21st century.
-When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family SES and student performance are tightly linked.
+The idea of meritocracy and intergenerational transfers has motivated, for a good part of the 20th and 21st century, much of the research on social mobility and social inequalities.
+When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family socio-economic status (SES) and student performance are tightly linked.
 The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements.
@@
 
-Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive and personality definition \citep{duyme1999, waldfogel2006}.
-Economists and educators have long neglected this branch of literature but the work of James Heckman has brought our attention to it.
+Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive development and defining personality traits \citep{duyme1999, waldfogel2006}.
+Economists and educators have long neglected these findings but the work of James Heckman has brought our attention to it.
 For example, he has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
@@
 The implications of this cumulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\).
-That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is the cheaper than investing in later education and it brings about the largest returns.
-Despite all of these findings and investments to reduce the gaps, we still find that the relationship between parental education or socio-economic status is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.
-What we don't know is how and why this relationship is formed from such an early age.
-Moreover, many have suggested different explanations and the topic has been very contested \citep{brooks1997}.
-
-It is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital.
+That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is cheaper than investing at later stages and it brings about the largest returns.
+Despite all of these findings and strategies to reduce the gaps, we still find that the relationship between children's parental education and future destination is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.
+
+For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital.
 Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity.
 The field of economics of education has also helped to understand that education is by far the investment that yields the best returns.
-This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market to improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
+This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
 Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about.
@@
 Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities.
-We still don't know, in comparative terms, which countries have big or small cognitive gaps resulting from SES origins.
+We have very little evidence, in comparative terms, for which countries have big or small cognitive gaps resulting from SES origins.
 And even more importantly, we haven't assesed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.
 
-Virtually in all countries, be it developed or developing, there is inequality of opportunity.
+However, the closest we have is the vast literature on social mobility.
+Thanks to these findings, we know that virtually in all countries, be it developed or developing, there is inequality of opportunity.
 But there is considerable variation in the magnitud of this effect.
@@
 The important finding came when we discovered the main reason behind their social escalator: it's educational system.
-For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impacts in helping children reach their fullest potential.
-It does so very strongly for children which otherwise would be doomed to have a bleak future.
-Moreover, it tracks students into different curricula very late compared to other Europen countries (age 16), something which has been linked to less educational inequality.
+For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impact on improving opportunities.
+The education system is completely subsidized for all children, otherwise giving opportunities to children wouldn't have been able to pay.
+In addition to this, Denmark is recognized as a world-leader in terms of public support for its child-care system as it spends around 2\% of GDP, and has among the highest enrollment rates for children under 6 years old \citep{esping_waldfogel2012}.
+Moreover, it tracks students into different curricula very late compared to other Europen countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}.
 These two traits make the Danish system uniquely effective.
@@
 
-The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years.
-The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence \citep{anna2016_global} is emerging with a very similar landscape.
-The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940.
-This time-trend has allowed researchers to study achievement gaps in a lengthy period of time.
-Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
-Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}.
-The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
-
 Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation.
@@
 However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970 where income inequality was increasing.}.
-.
 
@@
 They also calculated the same changes for a number of personality traits such as self-control and externalizing behaviour and found similar results.
-In contrast to \citet{reardon2011}, he finds that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates.
+In contrast, \citet{reardon2011} finds that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates.
 Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap).
@@
 
-The reasons the authors find a reversal in the trend could be numerous and should be studied very closely.
-The authors incorporate a number of country-level indicators to try and explaind this change and suggest that the reversal is likely due to the high increase of preschool enrollment.
-They suggest similarly that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
-Their conclusions are very speculative and have no empirical support, which is why this is still an open question.
-The reasons the authors find a reversal in the trend could be numerous and should be studied very closely.
-The authors incorporate a number of country-level indicators to try explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment.
-In fact, the point out that in this same period (1998 - 2010) the SES gap in preschool enrollment decreased substantially.
-Their conclusions are very speculative and have no empirical support, which is why this is still an open question.
+The reasons why the authors find a reversal in the trend could be numerous and should be studied very closely.
+The authors incorporate a number of country-level indicators to explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment.
+They build on their previous argument by suggesting that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
+Their conclusions, although suggestive, are very speculative and have no empirical support, which is why this is still an open question.
 
@@
 
-Another limitation of their study is that their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
+Another limitation of the work of \citet{bradbury2015} is that their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
 They do an amazing work at making the four surveys comparable and their findings are indeed very reliable.
@@
 Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
-In comparison to the literature on achievement gaps, they find that the U.S has a gap of   1.20 SD in 2001 which increase to around ~ 1.30 in the year 2006 while Germany has a decreasing gap from around   1.25 to   1 standard deviation in the same year-span.
+In comparison to the literature on achievement gaps, they find that the U.S has a gap of 1.20 SD in 2001 which increase to around 1.30 in the year 2006 while Germany has a decreasing gap from around 1.25 to 1 standard deviation in the same year-span.
 However, these numbers vary a lot and carry a great deal of overlapping uncertainty.
@@
 This is revealing because it doesn't really pay off to look at a general average once each country has their own distinctive gap and evolution.
-Another important finding is that the achievement gap is clearly no universal and should be studied in context.
-
-One clear limitation of their study (as well as \citet{reardon2011}) is that they adjust for the age of each child in all studies.
+This highlights the notion that the increasing achievement gap is clearly not universal and should be studied in context.
+
+One limitation of their study (as well as \citet{reardon2011}) is that they adjust for the age of each child in all studies.
 Although for their modeling purposes this is the correct thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
@@
 The explanation, although very debated, has been gaining much support in recent years.
-If we remember correctly, countries with high levels of curricular differentiation, the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}.
+If we remember correctly, for countries with high levels of curricular differentiation the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}.
 Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}.
@@
 
-\section{Research questions}
-
-This study seeks to evaluate the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries.
+\section{Research framework}
+
+\subsection{Research questions}
+
+This study seeks to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries and search for a likely explanation.
 This is different from previous work because it concentrates solely on 15 year old children, something not done for the achievement gap, and it attempts to capture the evolution of the achievement gap for countries with different educational systems.
-In addition, extend the previous findings by studying the relationship between the gap and the average performance of each country in order to detect if there's a tradeoff between good performance and equality over time separately for each country.
-The advantages of this study are twofold.
+In addition, it extends the previous findings by studying the relationship between the gap and the average performance of each country in order to detect if there's an over-time tradeoff between good performance and equality over time separately for each country.
+The advantages of this study are threefold.
 
@@
 
+Third, we attempt to model why the achievement gap is changing over time. In contrast to other studies, this is a novelty given that most research has been done cross-sectionally between countries.
+
 More formally, we're interested in studying
 a) Which countries are experiencing changes in the achievement gap;
-b) the rate at which each country-gap is widening/narrowing and;
-c) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.
-d) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
+b) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind;
+c) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
+d) What explains this cross-country evolution;
 
@@
 a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
-After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
-This shows that it is important to study the changes in the achievement gap over time.
-This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.
-a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
-After recording a SES gap increase of about 40\% in only 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data.
-This shows that it is important to study the changes in the achievement gap over time.
-This will bring a useful comparison and pinpoint which countries are doing the best to reduce the gap.
-
-b) We want to compare the percentage change at which the gap widened/narrowed from the first to the last year available.
+After recording a SES gap increase of about 40\% in only 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap over time.
+We will compare the percentage change at which the gap widened/narrowed from the first to the last year available.
 This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
@@
 
-c) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
+b) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
 The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind.
 We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.
-This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving.
-If the bottom performers are stagnated whereas the top is gaining ground, the performance-equality trade off dilemma of tracking might be much more credible than other explanations.
-
-d) We want to test the notion that better performing countries have a smaller degree of inequality than other countries.
+
+c) We want to investigate whether better performing countries have lower levels of inequality than other countries.
 \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
@@
 
+d) Tightly linked to the previous point, we want to test whether several dimensions of the tracking setup explain why the gaps are changing.
+This will support the previous points in suggesting that the dynamics of the over-time achievement gap are endogenously intertwined with the effect that tracking has on the specific changes of the high and low SES groups.
+
+\section{Hypothesis}
+
+Following the previous questions and the literature on educational inequality and tracking we pose three hypothesis for answering questions (b), (c) and (d) given that question (a) is purely an exploratory exercise.
+
+\begin{itemize}
+\item Hypothesis 1: In countries where there is a high degree of tracking we should expect the top and bottom to be evolving at the rate given that tracking is thought to maximize the learning experience of both groups.
+For countries with low levels of tracking we should expect the top and bottom groups to be closer to each other and for the gap to be closing since the both groups will be in the same track (this doesn't mean that the top is loweing their score but that the bottom is most likely catching up).
+
+\item Hypothesis 2: Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality.
+Countries with high performance should have less inequality and countries with low performance should have greater inequality
+
+\item Hypothesis 3: if any (or both) of the previous hypothesis are confirmed, then tracking should play an important role on the evolution of the achievement gap.
+We hypothesize that the degree of tracking is tightly related to changes in the gap, and the more tracking, the more inequality.
+Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labour market opportunities.
+\end{itemize}
+
 \section{Methods}
-
 \subsection{Data}
@@
 This variable attempts to capture the social status of the family, without asking for income information.
-This index variable was developed by \citet{ganzeboom1996, ganzeboom2010} and it attempts to measure occupational status using a continuous measure.
-This indicator is a very reliable alternative to the  classical Erikson-Goldthorpe-Portocarero classification \citet{erikson1979}.
+This index variable was developed by \citet{ganzeboom1996} and later refined by \citep{ganzeboom2010} and it attempts to measure occupational status using a continuous measure.
+This indicator is a very reliable alternative to the  classical Erikson-Goldthorpe-Portocarero classification \citep{erikson1979}.
 It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}.
@@
 Yet one of the most relevant variables for our study is a composite SES index created by the PISA team.
-The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measured the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}.
+The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measures the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}.
 This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming all mentioned variables into comparable metrics across waves.

### 2017-09-08 paper/paper.Rnw
Revised all the vocabulary/punctuation - further analysis, graphs and text left

@@
 
-Aside from SES, the other most relevant variables are test scores for Mathematics and Literacy.
+Aside from SES, the other most relevant variables are test scores for mathematics and literacy.
 PISA does not provide a single test result for each respondent.
@@
 
-A more intuitivity is explanation is this: suppose we have \(\mu_i\), the average student test score in Mathematics for student \(i\).
+A more intuitivity is explanation is this: suppose we have \(\mu_i\), the average student test score in mathematics for student \(i\).
 Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\) for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test.
@@
 
-The aim of this paper is to identify and disaggregate country trends in the achievement gap for several countries.
+The aim of this paper is to identify, disaggregate and explain country trends in the achievement gap for several countries.
 To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004, hanushek_woesmann_tracking, saw2016, bradbury2015, byun2010}.
@@
 SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.
-The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, as others have used, for example, the 5th, 50th and 95th \citep{micklewright}.
+The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, as others have used, for example, the 95th, 50th and 5th \citep{micklewright}.
 We use the conventional 90/10 cutoffs in the literature following the standard set by \citet{reardon2011}.
@@
 
-for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011}) weighted by the student sample weights.
+for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011} \footnote{This does not mess up our analysis by masking age-specific gaps because all students in the sample are 15 year olds.
+Control for aging is simply to control for monthly differences in ages.}) weighted by the student sample weights.
 
@@
 
-In order to correct gap estimates for measurement error, I adjust student test scores(\begin{math} \hat{\gamma_i} \end{math}) by
+In order to correct the gap estimates for measurement error, we adjust student test scores(\begin{math} \hat{\gamma_i} \end{math}) by
 
@@
 where \begin{math}r\end{math} is the reliability score of the wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}.
-The realibility estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of Person separation reliability such as internal consistency.
+The realibility estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of person separation reliability such as internal consistency.
 Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own.
@@
 
-We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting their very equal countries.
+We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting they're very equal countries.
 However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average, which is 0.
@@
 For example, France, Austria and surprisingly Sweden have very steep slopes.
-France increasde the gap by roughly \Sexpr{diff_data$math[["France"]]} standard deviations, Austria by \Sexpr{diff_data$math[["Austria"]]} and Sweden by \Sexpr{diff_data$math[["Sweden"]]}.
+France increased the gap by roughly \Sexpr{diff_data$math[["France"]]} standard deviations, Austria by \Sexpr{diff_data$math[["Austria"]]} and Sweden by \Sexpr{diff_data$math[["Sweden"]]}.
 This pattern happens similarly for literacy.
@@
 It's nearly 0.5 standard deviations in 2000 and it increased only by \Sexpr{diff_data$math[["Canada"]]} in 2015.
-We can't really say Italy has had a significant since it has a sort of wiggly pattern that looks like it cancels itself out.
+We can't really say Italy has had a significant increase since it has a sort of wiggly pattern that looks like it cancels itself out.
 
 On the other hand, we do have some countries which show a decrease in the SES achievement gap.
-United Kingdom decreased by about \Sexpr{diff_data$math[["United Kingdom"]]}, Poland by \Sexpr{diff_data$math[["Poland"]]}, and Denmark by \Sexpr{diff_data$math[["Denmark"]]}.
+Poland decreased by about \Sexpr{diff_data$math[["Poland"]]} and Denmark by \Sexpr{diff_data$math[["Denmark"]]}.
 However, the most notable cases are the United States and Germany.
@@
 This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy.
+Do note that the trendline is for both mathematics and literacy.
+If it were only for mathematics, then the trendline would probably be even more extreme.
+Despite this, pooling both subjects gives even more strength to the analysis given that the results stand for both subjects together \footnote{Same graph is availabe for reading upon request, where we find pretty much the same results if not even more steep/downward slopes.}.
 
@@
 The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015.
-The actual difference is of about \Sexpr{diff_data$math[["Sweden"]]}.
 Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.
@@
 @
+
 We find that the initial gap for the U.S in 2000 is \Sexpr{round(us_math_2000, 2)} varying between \Sexpr{round(us_math_2000_lower, 2)} and \Sexpr{round(us_math_2000_upper, 2)}.
-This gap is much higher than found in previous studies, but again, we don't know much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely.
+This gap is much higher than found in previous studies, but again, we don't know how much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely.
 All country-level studies pool many studies together and control for age, removing the age-specific gap effect.
@@
 However, it's important to disentangle where is this gap originating from.
-Is this because the top are improving why the bottom decreases?
+Is this because the top are improving while the bottom decreases?
 Or is it that the bottom is catching up?
@@
 
-Figure 4 shows the evolution for both Low and High SES groups separetly.
+Figure 4 shows the evolution for both low and high SES groups separately.
 The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other.
@@
 We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.
+
+% Make country summaries: Like the US is decreaing but because of bla bla (summing up everything up until now). Look at the graph and build rough summaries.
 
@@
 
-We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany.
-We see a negative relationship where the faster the low SES catch up the slower the High SES decrease.
+We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (top-left panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany.
+Regardless of this, we see an overall negative relationship where countries in which the low SES group is catching up, the high SES group goes down.
 This is very interesting because it suggests that there's a trend for countries to have the bottom SES groups are climbing upwards, a trait of meritocracy.
+However, this means that the top group comes down, rather than stagnate or increase.
 Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel).
-On the other hand, we do see more ocuntries where the both groups are increasing (top right panel).
+On the other hand, we do see more countries where the both groups are increasing (top right panel).
 We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.
+To sum up, we see no definitive evidence of equality in all of these countries.
+Regardless of different tracking setups, we see that there's a tradeoff: either the bottom comes up or the top comes down or the top goes even further up and the bottom further down.
+
+% Make country summary of each country for the tracking hypothesis #3.
+Like for example, countries in the bottom-right quadrant have low catching up and high coming down, do they have a common tracking structure?
+Do the same for the three quadrants.
 
@@
 
-Given that we have the data for the average performance for each country and the achievement gap, we can also investigate whether there's a tradeoff between the average country performance and the widening of the achievement gap and if it stands over time.
-One particular topic that has been researched in the past is whether improving the overall performance of a country compromises the equality of the educational system.
+\subsection{Equality-efficiency tradeoff}
+
+Moving on, we now discuss if there's a tradeoff between an educational system being efficient, i.e raising the overall performance of the country and whether that challenge compromises equality at the same time.
 For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance.
@@
 If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance.
-We plot this relationship in figure 8.
-
-\subsection{Performance-equality tradeoff}
+Given that we have the data for the average performance for each country and the achievement gap, we can investigate this relationship in detail.
+We first being by plotting all countries pooled for each year in figure 8.
+
 <<graphing_achievement_disparity, cache = T>>=
@@
 However, what they didn't really go in depth was whether this corelation is strongest for some countries over time.
+However, we should not get trapped into an ecological fallacy and assume this relationship is identical within each country.
 With the data available we can look at the evolution separately for many countries.
@@
 This pattern is also seen in Austria, Australia and Germany.
-The other obvious pattern is the opposite trend.
+Yet we also see the opposite trend.
 We see countries like Poland, Finland, Canada and less pronounced, Italy, where this pattern is evident.
@@
 
-\section{The role of curricular tracking in explaining inequalities}
+These results actually confirm that there exists a partial ecological fallacy.
+We see countries like Poland, Canada and Germany where the pattern is just as the aggregate trend, namely that as the average performance increases, inequality seems to be reduced in parallel.
+Conversely, France, Austria and Australia, to name a few, portray the opposite: as average performance decreases inequality increases.
+With these results we confirm our second hypothesis, that there is negative relationship.
+But we also unconver important variation within countries in the relationship between average achievement and dispersion of achievement.
+
+\subsection{The role of curricular tracking in explaining inequalities}
 
@@
 
+We now move to the final section of the paper, where we discuss what is a likely explanation to the patterns from before.
 The reason why countries differ in their evolution of inequality is still unknown.
-There is ample evidence showing the inequality between countries, to a certain extent, can be explained by the degree of tracking (Hanushek and Woessman).
+There is ample evidence showing the inequality between countries, to a certain extent, can be explained by the degree of tracking \citep{hanushek_woesmann_tracking}.
 But we are also interested in what explains the between country evolution of inequality, that is, the explanation as to why in certain countries it is increasing more than others over time.
 
-This is clearly a difficult task given than, at least the country-level, not many features of the educational system change in a 15-year time span, resulting in a practically invariant variable from which to explain the level of inequality (without any variance, we can't explain anything.).
+This is clearly a difficult task given that, at least at the country-level, not many features of the educational system change in a 15-year time span, resulting in a practically invariant variable from which to explain the level of inequality (without any variance, we can't explain anything.).
 However, there are other strategies that can at least partially help us to understand the phenomena, and these are precisely the ones we follow.
@@
 Following the literature on the topic, we suspect that this evolution is in part very tightly related to the degree of stratification of the educational system.
-In a very simple frame-work, Hanushek and Woessman found that tracking indeed exarcebates inequality in the life course of children.
+In a very simple frame-work, \citet{hanushek_woesmann_tracking} found that tracking indeed exarcebates inequality in the life course of children.
 Following the same strategy, we believe that tracking is also allowing the same gap in 15 year olds to grow over time, acompannied by other things.
-Yet, instead of concetrating only on the age of selection (just as they do), we explore a more fine grained definition of tracking, which is possible through the work of Thijs and Van der Werfhorst().
-Excluding the age of first selection into tracks, we also use the the number of tracks in the country, the percentage of the entire curriculum that is tracked and a vocational index \footnote{This index is a factor loading from a principal factor analysis of the percentage of students in upper secondary vocational education (taken from two sources, to reduce measurement error)}.
+Yet, instead of concentrating only on the age of selection (just as they do), we explore a more fine grained definition of tracking, which is possible through the work of \citet{bol2013}.
+Aside from the age at first selection into tracks, we also use the the number of tracks in the country, the percentage of the entire curriculum that is tracked and a vocational index \footnote{This index is a factor loading from a principal factor analysis of the percentage of students in upper secondary vocational education (taken from two sources, to reduce measurement error).
+We take all of the data from \citet{bol2013}.}.
 
@@
 
-In the final model, we regress the inequality of a country on a dummy variable where 1 equals only 1 track vs more than 1 track, whether the age of first selection is 15 or more vs below 15, a dummy stating whether the percentage of tracked curriculum is above 1, and the standardized vocational index.
+In the final model, we regress the inequality of a country on a dummy variable where 1 equals only 1 track vs more than 1 track, whether the age of first selection is 15 or more vs below 15, a dummy stating whether the percentage of tracked curriculum is above 0 (that is, any tracking), and the standardized vocational index \footnote{We create dummies to avoid multicollinearity.
+The final model has a maximum VIF of 2.2.}.
 We are reluctant to include more variable in the model, first to keep it more parsimonious, and secondly to prevent overfitting due to our sample size.
@@
 
-Given that the classical p values are inexistent in bayesian modelling, I present uncertainty intervals next to each coefficient.
+Given that the classical p values are inexistent in bayesian modelling, we present uncertainty intervals next to each coefficient.
 The first model only includes the 1 track dummy.
@@
 The second model now includes the dummy of whether the country has an age selection of 15 or more and we can see that countries with late selection into tracking have about -.38 less standard deviations in inequality.
-Also note that the number of tracks dummy was also reduced, suggesting these two variables are explaining similar things.
-This is expected given that late tracking is usually associated with fewer tracks, such as in the scandinavian countries.
+Also note that the number of tracks dummy was now reduced, suggesting these two variables are explaining similar things.
+This is expected given that late tracking is usually associated with fewer tracks, such as in the Scandinavian countries.
 The third model now includes whether the country has any degree of tracking, meaning whether the country has above 0\% of the curriculum tracked.
 The coding of the variable comes from visual inspection of the distribution, where some countries, such as the United States and Canada, have untracked curriculums.
-We see now relationship at all, but after including the vocational index, the model is completely reversed with countries with tracked curriculum having about 0.6 more standard deviations in inequality that countries with no tracking.
+For this last variable we see no relationship at all, but after including the vocational index, the model is completely reversed with countries with tracked curriculum having about 0.6 more standard deviations in inequality that countries with no tracking.
 
@@
 Countries with high vocational enrollment are also those with early age selection, such as Germany and Austria.
-These two variables should be inspected further, perhaps with an interaction, given than both coefficients are sizable and very correlated.
+These two variables should be inspected further, perhaps with an interaction, given that both coefficients are sizable and very correlated.
 This final model explains over one third of the variance in the evolution of the 90/10 SES gap, with an R square of 40\%.
@@
 Nonetheless, we see that the other coefficients are very strong and have reasonable intervals.
-It is very important to consider the uncertainty of our estimates, to give confidence in our model fit.
+It is very important to consider the uncertainty of our estimates, to gain confidence in our model fit.
 So far, this model seems to fit the data quite reasonably.
-\footnote{These variables are very correlated, but no collinearity is present with a maximum VIF of 2.2.}
 
@@
 Below we plot all these combinations.
-The x axis has the number of tracks that a country has, the y axis has the predicted 90/10 achievement gap and below each country name we have whether the country has an age selection of 15 or more or below.
-In addition, to confirm how accurate is the model, we also plot the actual level of inequality in the actual tracking setup of the country.
+The X axis has the number of tracks that a country has, the Y axis has the predicted 90/10 achievement gap and below each country name we have whether the country has an age selection of 15 or more or below 15.
+In addition, to confirm how accurate the model is, we also plot the actual level of inequality in the actual tracking setup of the country.
 Let's take Germany for example.
@@
 
-The red point for Germany is in the '\textgreater 1 track' and '\textless 15' setup, meaning that Germany has a tracking system that has more than 1 track and an age of selection of below 15.
+The red point for Germany is in the '\textgreater 1 track' and '\textless 15 age' setup, meaning that Germany has a tracking system that has more than 1 track and an age of selection of below 15.
 We see that the model predicts the level of inequality for that set up quite accurately, relative to the actual level (red point).
 But if the setup would be in the ideal '1 track' and age of selection above or equal to 15, the inequality would be much lower.
-Generally speaking, this applies to virtually every country in the plot.
-Virtually for all countries which are not in this ideal tracking, a switch to the ideal would reflect a reduction in the achievement gap.
-
-For those countries which are not in the 'ideal tracking' setup, we compute below the percentage change from switching to the ideal tracking. The results are in the table below.
+Generally speaking, this applies to all countries in the plot.
+Virtually for all countries which are not in this 'ideal' tracking, a switch to the 'ideal' would reflect a reduction of the achievement gap.
+
+% Comment on the above graph for particular countries! Make some observations
+
+For those countries which are not in the 'ideal' tracking setup, we compute below the percentage change from switching to the ideal tracking.
 
@@
 The biggest reductions are for both Hungary and Germany, with 46\% and 36\% respectively.
-France, which had a particularly steep increase in the last 15 years, would see a reduction of the 90/10 SES gap of 31\% and Austria, with a very similar tracking structure as the German one, would see their achievement gap reduced by 20\%.
-The very suprising result is the Dutch case, for which a switch from the 'worst' tracking setup to the 'best' case would result in an increase of the gap of 20\%.
-
-This result could mean two things.
-First and foremost, the Dutch educational system has built other mechanisms that, despite the tracking system, help children from social origin escalate socially.
-This explanation is not very speculative given than in most mobility research, the Netherlands stands out as a very equal country relative to other countries with a similary tracking structure, like Germany or Hungary (cite Gosta study and something from Breen and Goldthorpe).
-Another explanation, and something which needs to assessed, is whether our model is predicting accurately.
+France, which had a particularly steep increase in the last 15 years, would see a reduction of the 90/10 SES gap of 31\% and Austria, with a very similar tracking structure as the German case, would see their achievement gap reduced by 20\%.
+The very suprising result is the Dutch case, for which a switch from the 'worst' tracking setup to the 'best' case would result in an increase of the gap of 19\%.
+
+This last result could mean two things.
+First and foremost, the Dutch educational system has built other mechanisms that, despite the tracking system, help children from low social origins escalate upwards.
+This explanation is not very trivial considering that in most mobility research, the Netherlands stands out as a very equal country relative to other countries with a similary tracking structure, like Germany \citep{shavit1993, esping2012}.
+Another possible explanation, and something which needs to assessed, is whether our model is predicting accurately.
 If we look at the plot from above, we see that for some countries, the model is a bit off from the actual level of inequality.
 Moreover, although we see that if all countries switched to the 'ideal' tracking setup we would see a vast reduction in inequality, we also see that each of these predictions is followed by a wide uncertainty interval.
+
 Considering that the results are very consistent for every country, and in most countries the prediction matches the actual value, this could simply be a consequence of low sample size.
@@
 
-A more summarized empirical strategy is to actually standardize all the tracking indicator into one index, something that Bol and Van der Werfhorst() already do quite nicely.
-Using this standardized tracking variable and the same bayesian multievel model, we now model the tracking and vocational index as before, but add an interaction between two, given that in the previous model we recorded a strong shift in the model when tracking-related variables and the vocational model we both included in the same model.
+A more summarized empirical strategy is to actually standardize all the tracking indicator into one index, something that \citet{bol2013} already do quite nicely.
+Using this standardized tracking variable and the same bayesian multilevel model, we now model the tracking and vocational index as before, but add an interaction between two, given that in the previous model we recorded a strong shift in the model when tracking-related variables and the vocational model where both included in the same model.
 Below are the results for this model.
@@
 Both these coefficients are actually significantly different from each other, so it is possible.
-Before we explore this visually, let's note that the interaction between the two is very reliable and significant, considering that small sample size.
+Before we explore it visually, note that the interaction between the two is very reliable and significant, considering that small sample size.
 Below we graph the interaction for different quantiles of the vocational index.
@@
 Moreover, our earlier interpretation (that tracking is more important than vocational enrollment for reducing the achievement gap) still holds and is very evident here.
-Despite the equalizing power of vocational enrollment, a high level of tracking with the highest level of vocational enrollment still leaves a country with an achievement of over 1.5 standard deviations.
-However, a country with low levels of tracking and \textit{any} level of vocational enrollment is nearly 1.3 standard deviations.
-A key finding from these results is that tracking is reasonably explaining the 90/10 achievement gap, and we should consider it as a very important component in the reduction of the achievement gap.
-However, it should not be studied in isolation from other features such as the vocational enrollment.
+Despite the equalizing power of vocational enrollment, a high level of tracking with the highest level of vocational enrollment still leaves a country with an achievement gap of over 1.5 standard deviations.
+However, a country with low levels of tracking and \textit{any} level of vocational enrollment is nearly 1.3 standard deviations at best.
+A key finding from these results is that tracking is a reasonable explanation of the 90/10 achievement gap, and we should consider it as a very important component in the reduction of the achievement gap.
+However, these results highlight that it should not be studied in isolation from other features such as the degree of vocational enrollment.
 We see a very sizable change in the achievement gap for a given country with the same level of tracking and high and low levels of vocational enrollment.
+
+% Talk about the mechanisms why tracking might be explaining this, although tracking doesn't change over time.
+
+% Because it might've caused it at the beginning and now it just exarcebates it.
+
+% Because when combined with vocational enrollment, it might be exarcebarting even more, if enrollment is not high.
 
@@
 With the exception of the Intercept, the coefficients of the three models look pretty much the same.
-The tracking index seems a but stronger for the 90/10 model but the uncertainty intervals overlap with the other two models.
+The tracking index seems slightly stronger for the 90/10 model but the uncertainty intervals overlap with the other two models.
 These results give our claims much more credibility.
@@
 Instead of estimating the evolution over time, we calculate the cumulative change over time for each country, summarizing the 6 year-time span of achievement gaps into one cumulative inequality gap.
-This approach dramatically reduces the total sample size to 34, the number of total countries available \footnote{We've been presenting a selected number of countries given the lack of space, but for every estimation we include all countries available.}.
-It is for this reason that I only include at most 2 covariates in the model, the tracking and vocational index, which both summarize the previous models and keep the model parsimonious.
+This approach dramatically reduces the total sample size to 34, the number of total countries available \footnote{We've been presenting a selected number of countries given the lack of space, but for every model/estimation we include all countries available.}.
+It is for this reason that we only include at most 2 covariates in the model, the tracking and vocational index, which both summarizes the previous models and keeps it parsimonious.
 

### 2017-09-09 paper/paper.Rnw
Added extra commenting of analysis for some sections

@@
 
-% Make country summaries: Like the US is decreaing but because of bla bla (summing up everything up until now). Look at the graph and build rough summaries.
-
-Finally, we pay attention to the flat summary lines.
+We now pay attention to the flat summary lines.
 These lines should be interpreted with caution because it doesn't mean that the gap is not increasing.
@@
 
-Ths source of this gap can be deconstructed further into two broader gaps.
+After analyzing the trend for each country and where this trend is coming from we start to find some interesting results.
+We see that the United States is closing the achievement gap at a very rapid rate, but at the same time we notice that it's because the top SES group is going down faster than the bottom is catching up.
+In contrast, we see countries like Germany and Poland where the gap is closing as well but because bottom group is catching up faster than the high SES group is coming down.
+In fact, the Polish case is the most egalitarian of all the countries presented here; the bottom group is catching up at a rapid pace and the top group is maintaining their high levels of performance, fostering a trully equalizing effect.
+
+Quite interestingly, in countries where the gap is widening, both groups are distancing themselves at a very similar pace.
+Finland, France and Austria, show this exact pattern.
+Give or take, we have a solid pattern that when the gap widens, it's because there's top-to-bottom inequality.
+But when the gap is being reduced, the source can vary between top and bottom changes.
+
+To help assess these results even further, the source of this gap can be deconstructed further into two broader gaps.
 We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group.
@@
 We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.
-To sum up, we see no definitive evidence of equality in all of these countries.
-Regardless of different tracking setups, we see that there's a tradeoff: either the bottom comes up or the top comes down or the top goes even further up and the bottom further down.
-
-% Make country summary of each country for the tracking hypothesis #3.
-Like for example, countries in the bottom-right quadrant have low catching up and high coming down, do they have a common tracking structure?
-Do the same for the three quadrants.
+
+To sum up, we find no evidence in support of our first hypothesis.
+We uncover that the tracking setup of a country doesn't imminently mean a certain degree of equality/inequality.
+Countries where the top SES group is increasing and the bottom is decreasing (top-left panel) have dramatically different tracking setups.
+Austria is an early tracker with several tracks, while Sweden is a late tracker with only 1 track.
+On the opossite panel (bottom-right), where we find countries that the bottom group is catching up while the top is coming down, we find similar results.
+Denmark and Germany are opposite poles in terms of tracking, while the U.S is distinctive in that it has little formal curricular stratification.
+We see no definitive evidence of top-to-bottom equality in all of these countries (except for Poland).
+We find that regardless of the different tracking setups, we see that there's a tradeoff: either the bottom comes up and the top comes down or the top goes even further up and the bottom further down.
 
@@
 
-These results actually confirm that there exists a partial ecological fallacy.
+These results actually confirm our second hypothesis, that there's a negative relationship between the two variables.
 We see countries like Poland, Canada and Germany where the pattern is just as the aggregate trend, namely that as the average performance increases, inequality seems to be reduced in parallel.
 Conversely, France, Austria and Australia, to name a few, portray the opposite: as average performance decreases inequality increases.
-With these results we confirm our second hypothesis, that there is negative relationship.
-But we also unconver important variation within countries in the relationship between average achievement and dispersion of achievement.
+But we also unconver important variation within countries in the relationship between average achievement and dispersion of achievement (have a look at the Netherlands, Italy and the U.S).
 
@@
 But if the setup would be in the ideal '1 track' and age of selection above or equal to 15, the inequality would be much lower.
-Generally speaking, this applies to all countries in the plot.
-Virtually for all countries which are not in this 'ideal' tracking, a switch to the 'ideal' would reflect a reduction of the achievement gap.
-
-% Comment on the above graph for particular countries! Make some observations
-
+
+These graph reveals some interesting patterns.
+Generally speaking, we find that the simulation shows that virtually all countries which are not in this 'ideal' tracking, that would switch to the 'ideal' setup would see a reduction of the achievement gap.
+Yet we also find the opposite result.
+Countries which are in the 'ideal' tracking setup that would change to the 'worst' setup, all countries would see a widening of the achievement gap.
+On average, we find that if all countries switched to the 'ideal' setup, we would see an average reduction of 20\%.
+This is a very important figure considering that \citet{reardon2011} found that the US gap increased by about 40-50\% in 30 years.
+Had all countries in the 'ideal' tracking switched to the 'worst' tracking, we would see an average widening of nearly 72\%.
 For those countries which are not in the 'ideal' tracking setup, we compute below the percentage change from switching to the ideal tracking.
@@
 <<counterfactual_tracking, include = TRUE>>=
-pred_avg_diff<-
+pred_avg_diff <-
   new_data %>%
@@
 track_should_change <- pred_avg_diff$country[pred_avg_diff$country %>% duplicated]
+track_should_not_change <- setdiff(pred_avg_diff$country, track_should_change)
 
@@
 
+@
+
+<<other_counterfactual, eval = F>>=
+
+# This is for when calculating the mean 'widening' of the gap if countries with 'ideal' tracking swtiched to the 'worse' tracking
+
+pred_avg_diff <-
+  new_data %>%
+  left_join(sum_ready_data) %>%
+  filter(!is.na(avg_diff) | as.numeric(num_tracks) > 1 & age_selection == 0) %>%
+  select(country, num_tracks, age_selection, pred, avg_diff)
+
+avg_increase <-
+  pred_avg_diff %>%
+  filter(country %in% track_should_not_change) %>%
+  fill(avg_diff, .direction = "up") %>%
+  group_by(country, num_tracks = ifelse(as.numeric(num_tracks) > 1, "more", "one")) %>%
+  summarize(pred = mean(pred, na.rm = TRUE),
+            avg_diff = mean(avg_diff, na.rm = TRUE)) %>%
+  filter(num_tracks == "more") %>%
+  mutate(reduction = ((pred - avg_diff) / avg_diff) * 100) %>%
+  pull(reduction) %>%
+  mean
+
+avg_increase
 @

### 2017-09-09 paper/paper.Rnw
Fixed all tables/added some analysis - fixing graphs

@@
   mutate(`0_n` = as.character(`0_n`),
-         `1_n` = as.character(`1_n`)) %>%
-  xtable::xtable()
+         `1_n` = as.character(`1_n`))
 
@@
 
-print(table_coming,
-      scalebox = '0.75',
-      floating = FALSE,
+xtable(table_coming,
+       caption = "SES sample size and ISCED composition"
+       ) %>%
+  print(
     add.to.row = addtorow,
@@
     include.rownames = FALSE,
-      hline.after = nrow(table_coming),
-      caption.placement = "top",
-      caption = "SES sample size and ISCED composition")
+    hline.after = nrow(table_coming)
+  )
 @
@@
 Despite this, pooling both subjects gives even more strength to the analysis given that the results stand for both subjects together \footnote{Same graph is availabe for reading upon request, where we find pretty much the same results if not even more steep/downward slopes.}.
+We also plot the 80/20 and 70/30 SES gaps and find that very similar patterns as in the 90/10 SES gap (see figure 10 in the appendix).
 
@@
 Despite this, most countries show a clear sign of either decreasing or increasing.
+We plot the same graph for the 80/20 and 70/30 SES gaps in the appendix and the results hold for these achievement gaps as well.
 
@@
   ggplot(aes(as.factor(wave), gap_size, group = ses, colour = ses, shape = ses)) +
-  geom_point(size = 2, alpha = 0.4) +
-  geom_line(alpha = 0.4) +
+  geom_point(size = 2) +
+  geom_line() +
   geom_line(stat = "smooth", method = "lm", aes(group = 1),
@@
   ggplot(aes(as.factor(wave), mean_math, group = ses, colour = ses, shape = ses)) +
-  geom_line(alpha = 0.4) +
-  geom_point(size = 2, alpha = 0.4) +
+  geom_line() +
+  geom_point(size = 2) +
   coord_cartesian(ylim = c(-1, 2)) +
@@
   ggplot(aes(as.character(wave), value, group = type, colour = type, shape = type)) +
-  geom_point(size = 2, alpha = 0.4) +
-  geom_line(alpha = 0.4) +
+  geom_point(size = 2) +
+  geom_line() +
   scale_y_continuous(name = "Test scores in SD") +
@@
   scale_shape_discrete(name = NULL, labels = c("Achievement gap", "Avg performance")) +
-  ggtitle("Evolution of the top/bottom groups by different gaps") +
+  ggtitle("Evolution of the relationship between achievement and inequality") +
   facet_wrap(~ country, ncol = 5) +
@@
   )) %>%
-  xtable() %>%
+  xtable(
+    caption = "Explaining 90/10 achievement gap - all countries pooled - Multilevel model"
+  ) %>%
   print(
@@
   ggplot(aes(num_tracks, avg_diff, fill = num_tracks)) +
-  geom_point(colour = "red") +
+  geom_col(aes(y = pred, colour = num_tracks)) +
+  geom_errorbar(aes(ymin = pred_low, ymax = pred_high), width = 0.2) +
+  geom_point(colour = "red", shape = 8) +
   geom_errorbar(aes(ymin = diff_low, ymax = diff_high), width = 0.2, colour = "red") +
-  geom_col(aes(y = pred, alpha = 0.4)) +
-  geom_errorbar(aes(ymin = pred_low, ymax = pred_high), width = 0.2) +
   facet_wrap(country ~ age_selection) +
@@
   scale_x_discrete(name = "Number of tracks", labels = c("> 1 track", "1 track")) +
-  scale_fill_manual(values = c("orangered2", "blue3"), guide = F) +
+  scale_fill_manual(values = c("white", "white"), guide = F) +
+  scale_colour_manual(values = c("black", "black"), guide = F) +
   ggtitle("Predicted 90/10 gap for different combinations of number of tracks and age of selection")
@@
 
-The red point for Germany is in the '\textgreater 1 track' and '\textless 15 age' setup, meaning that Germany has a tracking system that has more than 1 track and an age of selection of below 15.
-We see that the model predicts the level of inequality for that set up quite accurately, relative to the actual level (red point).
+The star point for Germany is in the '\textgreater 1 track' and '\textless 15 age' setup, meaning that Germany has a tracking system that has more than 1 track and an age of selection of below 15.
+We see that the model predicts the level of inequality for that set up quite accurately (height of the bar), relative to the actual level (the star point).
 But if the setup would be in the ideal '1 track' and age of selection above or equal to 15, the inequality would be much lower.
@@
 These graph reveals some interesting patterns.
-Generally speaking, we find that the simulation shows that virtually all countries which are not in this 'ideal' tracking, that would switch to the 'ideal' setup would see a reduction of the achievement gap.
+Generally speaking, we find that the simulation predicts that virtually all countries which are not in this 'ideal' tracking, that would switch to the 'ideal' setup, would experience a reduction of the achievement gap.
 Yet we also find the opposite result.
-Countries which are in the 'ideal' tracking setup that would change to the 'worst' setup, all countries would see a widening of the achievement gap.
+Countries which are in the 'ideal' setup that would change to the 'worst' setup would see a widening of the achievement gap.
 On average, we find that if all countries switched to the 'ideal' setup, we would see an average reduction of 20\%.
@@
     )
-
-xtable(final_avg_diff_table) %>%
+final_avg_diff_table %>%
+  xtable(
+    caption = "Simulation: reduction of achievement gap if countries switched to 'ideal' tracking"
+  ) %>%
   print(
@@
 
+% This is for when calculating the mean 'widening' of the gap if countries with 'ideal' tracking swtiched to the 'worse' tracking
+
 <<other_counterfactual, eval = F>>=
-
-# This is for when calculating the mean 'widening' of the gap if countries with 'ideal' tracking swtiched to the 'worse' tracking
 
@@
 <<interaction_model, results = 'hide'>>=
-
-sig_vs_sig <- function(coef1, se1, coef2, se2) {
-  diff <- c(coef1, se1) - c(coef2, se2)
-  abs(diff[1]) - 2 * diff[2]
-}
-
 ready_data <- dif_data(complete_gaps, tracking_data)
@@
   )) %>%
-  xtable() %>%
+  xtable(
+    caption = "What explains 90/10 achievement gap? Tracking and Vocational interaction"
+  ) %>%
   print.xtable(
@@
 
+<<coefficents_sif_diff, eval = F>>=
+
+sig_vs_sig <- function(coef1, se1, coef2, se2) {
+  diff <- c(coef1, se1) - c(coef2, se2)
+  abs(diff[1]) - 2 * diff[2]
+}
+
+sig(0.41, , -0.22, ) # the two arguments missing are the SE's of both coefficients.
+# take from the model from above and finish calculate (already did before, they're sig different from each other but I don't want to rerun again! :P)
+
+@
+
 <<interaction_plot, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
@@
 
-plot(plot_interaction, plot = FALSE)[[1]] +
+data_interaction <-
+  plot(plot_interaction, plot = FALSE)[[1]] %>% ggplot_build() %>% pluck(1) %>% pluck(1)
+
+order_colour <-
+  data_interaction %>%
+  distinct(group, colour) %>%
+  mutate(colour = rev(colour)) %>%
+  pull(colour)
+
+data_interaction %>%
+  mutate(linetype = group %>% as.factor,
+         colour = factor(colour, levels = order_colour)) %>%
+  ggplot(aes(x, y)) +
+  geom_line(aes(colour = colour, linetype = linetype), size = 1) +
+  geom_ribbon(aes(ymin = ymin, ymax = ymax, fill = colour), alpha = 0.4) +
   labs(x = "Tracking Index (SD)", y = "90/10 Achievement gap (SD)") +
   scale_colour_discrete(
+    name = "Vocational index \n quantiles",
+    labels = seq(100, 0, -25) %>% paste0("%")
+  ) +
+  scale_linetype_discrete(
     name = "Vocational index \n quantiles",
@@
 
-% Talk about the mechanisms why tracking might be explaining this, although tracking doesn't change over time.
-
-% Because it might've caused it at the beginning and now it just exarcebates it.
-
-% Because when combined with vocational enrollment, it might be exarcebarting even more, if enrollment is not high.
+The exact mechanisms through whick the tracking setup is explaining these inequalities is still very speculative.
+We could argue that tracking might've introduced the inequality once it was implemented.
+Another possibility is that tracking, althought not changing, is currently exarcebating it every year.
+Another explanation, which we find strong evidence for, is that the tracking setup interacts very closely with the vocational setup.
+These explanations are very speculative and further research should carry the task of looking for a more concrete explanation.
 
@@
   setNames(c(" ", gsub(" SES gap", "", gaps))) %>%
-  xtable() %>%
+  xtable(
+    caption = "What explains the cumulative achievement gaps? Models for three different achievement gaps"
+  ) %>%
   print.xtable(

### 2017-09-11 paper/paper.Rnw
Fixed most graph/table positons - ran everything, so far good

@@
 % Like, every country as a row and 6 columns for each wave.
-\begin{figure}
-\begin{center}
-<<variance_pisa, include = T, out.height = '3in', out.width = '3.5in', fig.align = 'center'>>=
+% \begin{figure}
+% \begin{center}
+<<variance_pisa, include = T, out.height = '3in', out.width = '3.5in', fig.align = 'center', fig.cap = 'Standard deviation of test scores across all waves'>>=
 
@@
 @
-\caption{Standard deviation of test scores across all waves.}
-\end{center}
-\end{figure}
+% \caption{Standard deviation of test scores across all waves}
+% \end{center}
+% \end{figure}
 
@@
 
-\subsection{Descriptives}
+\subsection{Evolution of the achievement gap}
 
@@
 
-\subsection{Evolution of the achievement gap}
 % TODO: Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
 
-\begin{figure}
-\begin{center}
-<<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+% \begin{figure}
+% \begin{center}
+<<graphing_9010gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Evolution of the achievement gap for selected countries'>>=
 # 90/10 gaps acros countries
@@
 @
-\caption{Evolution of the achievement gap for selected countries}
-\end{center}
-\end{figure}
+% \caption{Evolution of the achievement gap for selected countries}
+% \end{center}
+% \end{figure}
 
@@
 
-\begin{figure}
-\begin{center}
-<<graphing_perc_increase, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+% \begin{figure}
+% \begin{center}
+<<graphing_perc_increase, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Evolution of the achievement gap for separate SES groups'>>=
 perc_graph <- function(df, test, title, subtitle = NULL) {
@@
 @
-\caption{Percentage change in the achievement gap from 2000 to 2015}
-\end{center}
-\end{figure}
+% \caption{Percentage change in the achievement gap from 2000 to 2015}
+% \end{center}
+% \end{figure}
 
@@
 % Add confidence intervals
-\begin{figure}
-\begin{center}
-<<rate_increase/decrease, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+% \begin{figure}
+% \begin{center}
+<<rate_increase/decrease, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Evolution of the achievement gap for separate SES groups'>>=
 # You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
@@
 @
-\caption{Evolution of the achievement gap for separate SES groups}
-\end{center}
-\end{figure}
+% \caption{Evolution of the achievement gap for separate SES groups}
+% \end{center}
+% \end{figure}
 
@@
 
-\begin{figure}
-\begin{center}
-<<graphing_allgaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+% \begin{figure}
+% \begin{center}
+<<graphing_allgaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Evolution of 90/50/10 SES gaps for selected countries'>>=
 
@@
 @
-\caption{Evolution of 90/50/10 SES gaps for selected countries}
-\end{center}
-\end{figure}
+% \caption{Evolution of 90/50/10 SES gaps for selected countries}
+% \end{center}
+% \end{figure}
 
@@
 
-\begin{figure}
-\begin{center}
-<<rate_change_graph, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+% \begin{figure}
+% \begin{center}
+<<rate_change_graph, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Rate at which top/bottom SES groups are catching up for the 90/10 SES gap'>>=
 
@@
 @
-\caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
-\end{center}
-\end{figure}
+% \caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
+% \end{center}
+% \end{figure}
 
@@
 
-\begin{figure}
-\begin{center}
-<<graphing_ses_growth, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+% \begin{figure}
+% \begin{center}
+<<graphing_ses_growth, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Evolution of the SES gaps for top/bottom groups'>>=
 # Graphing how the top/bottom are evolving over time instead of absolute difference
@@
 @
-\caption{Evolution of the SES gaps for top/bottom groups}
-\end{center}
-\end{figure}
+% \caption{Evolution of the SES gaps for top/bottom groups}
+% \end{center}
+% \end{figure}
 
@@
 
-\begin{figure}
-\begin{center}
-<<avgperformance_inequality, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+% \begin{figure}
+% \begin{center}
+<<avgperformance_inequality, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Relationship between 90/10 achievement gap and average performance in mathematics'>>=
 data_to_plot %>%
@@
 @
-\caption{Relationship between 90/10 achievement gap and average performance in mathematics}
-\end{center}
-\end{figure}
+% \caption{Relationship between 90/10 achievement gap and average performance in mathematics}
+% \end{center}
+% \end{figure}
 
@@
 % Add confidence intervals
-\begin{figure}
-\begin{center}
-<<country_achievement_and_inequality, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+% \begin{figure}
+% \begin{center}
+<<country_achievement_and_inequality, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Time trends of achievement gap and average achievement between 2000 and 2015'>>=
 data_to_plot %>%
@@
 @
-\caption{Time trends of achievement gap and average achievement between 2000 and 2015}
-\end{center}
-\end{figure}
+% \caption{Time trends of achievement gap and average achievement between 2000 and 2015}
+% \end{center}
+% \end{figure}
 
@@
   xtable(
-    caption = "Explaining 90/10 achievement gap - all countries pooled - Multilevel model"
+    caption = "Explaining 90/10 achievement gap - Multilevel model"
   ) %>%
@@
 
-<<graph_complete_tracking_graph, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+<<graph_complete_tracking_graph, include = T, fig.width = 9.5, out.height = '7in', out.width = '6.5in', fig.align = 'center', fig.cap = 'Simulation: predicted 90/10 achievement gap for different tracking setups'>>=
 new_data <-
@@
 
-<<interaction_plot, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+% \begin{figure}[!htbp]
+% \begin{center}
+<<interaction_plot, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Interaction between tracking and vocational values. Legend ordered relative to the lines'>>=
 plot_interaction <-
@@
   distinct(group, colour) %>%
-  mutate(colour = rev(colour)) %>%
+  mutate(colour = colour) %>%
   pull(colour)
@@
 data_interaction %>%
-  mutate(linetype = group %>% as.factor,
+  mutate(linetype = factor(group, levels = 5:1),
          colour = factor(colour, levels = order_colour)) %>%
@@
     name = "Vocational index \n quantiles",
-    labels = seq(100, 0, -25) %>% paste0("%")
+    labels = seq(0, 100, 25) %>% paste0("%")
   ) +
@@
 @
+% \caption{Interaction between tracking and vocational values. Legend ordered relative to the lines.}
+% \end{figure}
+% \end{center}
 
@@
 
-<<all_gaps_plot, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+% \begin{figure}[!htbp]
+% \begin{center}
+<<all_gaps_plot, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'What explains achievement gaps? Model comparisons'>>=
 model_data_merged <-
@@
 model_data_merged %>%
-  ggplot(aes(terms, estimate, colour = model)) +
+  ggplot(aes(terms, estimate, colour = model, shape = model)) +
   geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1) +
-  geom_point() +
+  geom_point(size = 2.5) +
   geom_hline(yintercept = 0) +
-  scale_colour_discrete(name = NULL) +
   scale_y_continuous(limits = c(-0.5, 2)) +
@@
                               "Vocational Index")) +
+  scale_colour_discrete(name = NULL) +
+  scale_shape_discrete(name = NULL) +
   labs(y = "Estimate from linear model", x = "Predictors") +
@@
 @
+% \caption{What explains achievement gaps? Model comparisons}
+% \end{figure}
+% \end{center}
 
@@
 
-\begin{figure}
-\begin{center}
-<<graph_all_gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+<<graph_all_gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Evolution of the achievement gap for several gaps'>>=
 complete_gaps %>%
@@
 @
-\caption{Evolution of the achievement gap for several gaps}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<percentage_change_8020,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+
+<<percentage_change_8020,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Percentage change in the 80/20 achievement gap from 2000 to 2015'>>=
 perc_increase_fun(complete_data_topbottom_80) %>%
@@
 @
-\caption{Percentage change in the 80/20 achievement gap from 2000 to 2015}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<percentage_change_7030,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center'>>=
+
+<<percentage_change_7030,  include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Percentage change in the 70/30 achievement gap from 2000 to 2015'>>=
 perc_increase_fun(complete_data_topbottom_70) %>%
@@
 @
-\caption{Percentage change in the 70/30 achievement gap from 2000 to 2015}
-\end{center}
-\end{figure}
 

### 2017-09-11 paper/paper.tex
Fixed most graph/table positons - ran everything, so far good

@@
 % Like, every country as a row and 6 columns for each wave.
-\begin{figure}
-\begin{center}
+% \begin{figure}
+% \begin{center}
+\begin{figure}
 
@@
 
-\caption{Standard deviation of test scores across all waves.}
-\end{center}
-\end{figure}
+\caption[Standard deviation of test scores across all waves]{Standard deviation of test scores across all waves}\label{fig:variance_pisa}
+\end{figure}
+
+% \caption{Standard deviation of test scores across all waves}
+% \end{center}
+% \end{figure}
 
@@
 % latex table generated in R 3.4.0 by xtable 1.8-2 package
-% Wed Jul  5 17:51:02 2017
-\scalebox{0.75}{
+% Sat Sep  9 20:13:40 2017
+\begin{table}[ht]
+\centering
 \begin{tabular}{lllrrlrr}
@@
 \end{tabular}
-}
+\caption{SES sample size and ISCED composition} 
+\end{table}
 
@@
 
-\subsection{Evolution of the achievement gap}
 % TODO: Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
 
-\begin{figure}
-\begin{center}
+% \begin{figure}
+% \begin{center}
+\begin{figure}
 
@@
 
-\caption{Evolution of the achievement gap for selected countries}
-\end{center}
-\end{figure}
+\caption[Evolution of the achievement gap for selected countries]{Evolution of the achievement gap for selected countries}\label{fig:graphing_9010gaps}
+\end{figure}
+
+% \caption{Evolution of the achievement gap for selected countries}
+% \end{center}
+% \end{figure}
 
@@
 For example, France, Austria and surprisingly Sweden have very steep slopes.
-France increasde the gap by roughly 0.9 standard deviations, Austria by 0.6 and Sweden by 0.6.
+France increased the gap by roughly 0.9 standard deviations, Austria by 0.6 and Sweden by 0.6.
 This pattern happens similarly for literacy.
@@
 It's nearly 0.5 standard deviations in 2000 and it increased only by 0.2 in 2015.
-We can't really say Italy has had a significant since it has a sort of wiggly pattern that looks like it cancels itself out.
+We can't really say Italy has had a significant increase since it has a sort of wiggly pattern that looks like it cancels itself out.
 
 On the other hand, we do have some countries which show a decrease in the SES achievement gap.
-United Kingdom decreased by about 0, Poland by -0.4, and Denmark by -0.2.
+Poland decreased by about -0.4 and Denmark by -0.2.
 However, the most notable cases are the United States and Germany.
@@
 This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy.
+Do note that the trendline is for both mathematics and literacy.
+If it were only for mathematics, then the trendline would probably be even more extreme.
+Despite this, pooling both subjects gives even more strength to the analysis given that the results stand for both subjects together \footnote{Same graph is availabe for reading upon request, where we find pretty much the same results if not even more steep/downward slopes.}.
+We also plot the 80/20 and 70/30 SES gaps and find that very similar patterns as in the 90/10 SES gap (see figure 10 in the appendix).
 
@@
 The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015.
-The actual difference is of about 0.6.
 Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.
@@
 We find that the initial gap for the U.S in 2000 is 2.17 varying between 2.13 and 2.21.
-This gap is much higher than found in previous studies, but again, we don't know much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely.
+This gap is much higher than found in previous studies, but again, we don't know how much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely.
 All country-level studies pool many studies together and control for age, removing the age-specific gap effect.
@@
 
-\begin{figure}
-\begin{center}
+% \begin{figure}
+% \begin{center}
+\begin{figure}
 
@@
 
-\caption{Percentage change in the achievement gap from 2000 to 2015}
-\end{center}
-\end{figure}
+\caption[Evolution of the achievement gap for separate SES groups]{Evolution of the achievement gap for separate SES groups}\label{fig:graphing_perc_increase}
+\end{figure}
+
+% \caption{Percentage change in the achievement gap from 2000 to 2015}
+% \end{center}
+% \end{figure}
 
@@
 % Add confidence intervals
-\begin{figure}
-\begin{center}
+% \begin{figure}
+% \begin{center}
+\begin{figure}
 
@@
 
-\caption{Evolution of the achievement gap for separate SES groups}
-\end{center}
-\end{figure}
-
-Figure 4 shows the evolution for both Low and High SES groups separetly.
+\caption[Evolution of the achievement gap for separate SES groups]{Evolution of the achievement gap for separate SES groups}\label{fig:rate_increase/decrease}
+\end{figure}
+
+% \caption{Evolution of the achievement gap for separate SES groups}
+% \end{center}
+% \end{figure}
+
+Figure 4 shows the evolution for both low and high SES groups separately.
 The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other.
@@
 
-Finally, we pay attention to the flat summary lines.
+We now pay attention to the flat summary lines.
 These lines should be interpreted with caution because it doesn't mean that the gap is not increasing.
@@
 
-\caption{Evolution of 90/50/10 SES gaps for selected countries}
-\end{center}
-\end{figure}
+\caption[Evolution of 90/50/10 SES gaps for selected countries]{Evolution of 90/50/10 SES gaps for selected countries}\label{fig:graphing_allgaps}
+\end{figure}
+
+% \caption{Evolution of 90/50/10 SES gaps for selected countries}
+% \end{center}
+% \end{figure}
 
@@
 
-\caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
-\end{center}
-\end{figure}
-
-We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (red panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany.
-We see a negative relationship where the faster the low SES catch up the slower the High SES decrease.
+\caption[Rate at which top/bottom SES groups are catching up for the 90/10 SES gap]{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}\label{fig:rate_change_graph}
+\end{figure}
+
+% \caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
+% \end{center}
+% \end{figure}
+
+We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (top-left panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany.
+Regardless of this, we see an overall negative relationship where countries in which the low SES group is catching up, the high SES group goes down.
 This is very interesting because it suggests that there's a trend for countries to have the bottom SES groups are climbing upwards, a trait of meritocracy.
+However, this means that the top group comes down, rather than stagnate or increase.
 Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel).
-On the other hand, we do see more ocuntries where the both groups are increasing (top right panel).
+On the other hand, we do see more countries where the both groups are increasing (top right panel).
 We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.
 
+To sum up, we find no evidence in support of our first hypothesis.
+We uncover that the tracking setup of a country doesn't imminently mean a certain degree of equality/inequality.
+Countries where the top SES group is increasing and the bottom is decreasing (top-left panel) have dramatically different tracking setups.
+Austria is an early tracker with several tracks, while Sweden is a late tracker with only 1 track.
+On the opossite panel (bottom-right), where we find countries that the bottom group is catching up while the top is coming down, we find similar results.
+Denmark and Germany are opposite poles in terms of tracking, while the U.S is distinctive in that it has little formal curricular stratification.
+We see no definitive evidence of top-to-bottom equality in all of these countries (except for Poland).
+We find that regardless of the different tracking setups, we see that there's a tradeoff: either the bottom comes up and the top comes down or the top goes even further up and the bottom further down.
+
 But, similar as before, we don't know which group is causing the decrease/increase of the gap in the 90th/50th and 50th/10th gap. For that we plot the evolution of these gaps in figure 7.
 
-\begin{figure}
-\begin{center}
+% \begin{figure}
+% \begin{center}
+\begin{figure}
 
@@
 
-\caption{Evolution of the SES gaps for top/bottom groups}
-\end{center}
-\end{figure}
+\caption[Evolution of the SES gaps for top/bottom groups]{Evolution of the SES gaps for top/bottom groups}\label{fig:graphing_ses_growth}
+\end{figure}
+
+% \caption{Evolution of the SES gaps for top/bottom groups}
+% \end{center}
+% \end{figure}
 
@@
 If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance.
-We plot this relationship in figure 8.
-
-\subsection{Performance-equality tradeoff}
-
-\begin{figure}
-\begin{center}
+Given that we have the data for the average performance for each country and the achievement gap, we can investigate this relationship in detail.
+We first being by plotting all countries pooled for each year in figure 8.
+
+% \begin{figure}
+% \begin{center}
+\begin{figure}
 
@@
 
-\caption{Relationship between 90/10 achievement gap and average performance in mathematics}
-\end{center}
-\end{figure}
+\caption[Relationship between 90/10 achievement gap and average performance in mathematics]{Relationship between 90/10 achievement gap and average performance in mathematics}\label{fig:avgperformance_inequality}
+\end{figure}
+
+% \caption{Relationship between 90/10 achievement gap and average performance in mathematics}
+% \end{center}
+% \end{figure}
 
@@
 
-\caption{Time trends of achievement gap and average achievement between 2000 and 2015}
-\end{center}
-\end{figure}
+\caption[Time trends of achievement gap and average achievement between 2000 and 2015]{Time trends of achievement gap and average achievement between 2000 and 2015}\label{fig:country_achievement_and_inequality}
+\end{figure}
+
+% \caption{Time trends of achievement gap and average achievement between 2000 and 2015}
+% \end{center}
+% \end{figure}
 
@@
 
-\section{The role of curricular tracking in explaining inequalities}
+These results actually confirm our second hypothesis, that there's a negative relationship between the two variables.
+We see countries like Poland, Canada and Germany where the pattern is just as the aggregate trend, namely that as the average performance increases, inequality seems to be reduced in parallel.
+Conversely, France, Austria and Australia, to name a few, portray the opposite: as average performance decreases inequality increases.
+But we also unconver important variation within countries in the relationship between average achievement and dispersion of achievement (have a look at the Netherlands, Italy and the U.S).
+
+\subsection{The role of curricular tracking in explaining inequalities}
 
@@
 % latex table generated in R 3.4.0 by xtable 1.8-2 package
-% Thu Sep  7 15:04:53 2017
+% Sun Sep 10 14:55:49 2017
 \begin{table}[ht]
@@
 \end{tabular}
+\caption{Explaining 90/10 achievement gap - Multilevel model} 
 \end{table}
 
-Given that the classical p values are inexistent in bayesian modelling, I present uncertainty intervals next to each coefficient.
+Given that the classical p values are inexistent in bayesian modelling, we present uncertainty intervals next to each coefficient.
 The first model only includes the 1 track dummy.
@@
 Below we plot all these combinations.
-The x axis has the number of tracks that a country has, the y axis has the predicted 90/10 achievement gap and below each country name we have whether the country has an age selection of 15 or more or below.
-In addition, to confirm how accurate is the model, we also plot the actual level of inequality in the actual tracking setup of the country.
+The X axis has the number of tracks that a country has, the Y axis has the predicted 90/10 achievement gap and below each country name we have whether the country has an age selection of 15 or more or below 15.
+In addition, to confirm how accurate the model is, we also plot the actual level of inequality in the actual tracking setup of the country.
 Let's take Germany for example.
 
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graph_complete_tracking_graph-1}
-
-}
-
-The red point for Germany is in the '\textgreater 1 track' and '\textless 15' setup, meaning that Germany has a tracking system that has more than 1 track and an age of selection of below 15.
-We see that the model predicts the level of inequality for that set up quite accurately, relative to the actual level (red point).
+\begin{figure}
+
+{\centering \includegraphics[width=6.5in,height=7in]{figure/graph_complete_tracking_graph-1}
+
+}
+
+\caption[Simulation]{Simulation: predicted 90/10 achievement gap for different tracking setups}\label{fig:graph_complete_tracking_graph}
+\end{figure}
+
+The star point for Germany is in the '\textgreater 1 track' and '\textless 15 age' setup, meaning that Germany has a tracking system that has more than 1 track and an age of selection of below 15.
+We see that the model predicts the level of inequality for that set up quite accurately (height of the bar), relative to the actual level (the star point).
 But if the setup would be in the ideal '1 track' and age of selection above or equal to 15, the inequality would be much lower.
-Generally speaking, this applies to virtually every country in the plot.
-Virtually for all countries which are not in this ideal tracking, a switch to the ideal would reflect a reduction in the achievement gap.
-
-For those countries which are not in the 'ideal tracking' setup, we compute below the percentage change from switching to the ideal tracking. The results are in the table below.
+
+These graph reveals some interesting patterns.
+Generally speaking, we find that the simulation predicts that virtually all countries which are not in this 'ideal' tracking, that would switch to the 'ideal' setup, would experience a reduction of the achievement gap.
+Yet we also find the opposite result.
+Countries which are in the 'ideal' setup that would change to the 'worst' setup would see a widening of the achievement gap.
+On average, we find that if all countries switched to the 'ideal' setup, we would see an average reduction of 20\%.
+This is a very important figure considering that \citet{reardon2011} found that the US gap increased by about 40-50\% in 30 years.
+Had all countries in the 'ideal' tracking switched to the 'worst' tracking, we would see an average widening of nearly 72\%.
+For those countries which are not in the 'ideal' tracking setup, we compute below the percentage change from switching to the ideal tracking.
 
 % latex table generated in R 3.4.0 by xtable 1.8-2 package
-% Thu Sep  7 15:05:38 2017
+% Sat Sep  9 20:14:00 2017
 \begin{table}[ht]
@@
 \end{tabular}
+\caption{Simulation: reduction of achievement gap if countries switched to 'ideal' tracking} 
 \end{table}
 
+% This is for when calculating the mean 'widening' of the gap if countries with 'ideal' tracking swtiched to the 'worse' tracking
+
 The biggest reductions are for both Hungary and Germany, with 46\% and 36\% respectively.
-France, which had a particularly steep increase in the last 15 years, would see a reduction of the 90/10 SES gap of 31\% and Austria, with a very similar tracking structure as the German one, would see their achievement gap reduced by 20\%.
-The very suprising result is the Dutch case, for which a switch from the 'worst' tracking setup to the 'best' case would result in an increase of the gap of 20\%.
-
-This result could mean two things.
-First and foremost, the Dutch educational system has built other mechanisms that, despite the tracking system, help children from social origin escalate socially.
-This explanation is not very speculative given than in most mobility research, the Netherlands stands out as a very equal country relative to other countries with a similary tracking structure, like Germany or Hungary (cite Gosta study and something from Breen and Goldthorpe).
-Another explanation, and something which needs to assessed, is whether our model is predicting accurately.
+France, which had a particularly steep increase in the last 15 years, would see a reduction of the 90/10 SES gap of 31\% and Austria, with a very similar tracking structure as the German case, would see their achievement gap reduced by 20\%.
+The very suprising result is the Dutch case, for which a switch from the 'worst' tracking setup to the 'best' case would result in an increase of the gap of 19\%.
+
+This last result could mean two things.
+First and foremost, the Dutch educational system has built other mechanisms that, despite the tracking system, help children from low social origins escalate upwards.
+This explanation is not very trivial considering that in most mobility research, the Netherlands stands out as a very equal country relative to other countries with a similary tracking structure, like Germany \citep{shavit1993, esping2012}.
+Another possible explanation, and something which needs to assessed, is whether our model is predicting accurately.
 If we look at the plot from above, we see that for some countries, the model is a bit off from the actual level of inequality.
 Moreover, although we see that if all countries switched to the 'ideal' tracking setup we would see a vast reduction in inequality, we also see that each of these predictions is followed by a wide uncertainty interval.
+
 Considering that the results are very consistent for every country, and in most countries the prediction matches the actual value, this could simply be a consequence of low sample size.
@@
 % latex table generated in R 3.4.0 by xtable 1.8-2 package
-% Thu Sep  7 15:07:49 2017
+% Sat Sep  9 20:16:18 2017
 \begin{table}[ht]
@@
 \end{tabular}
+\caption{What explains 90/10 achievement gap? Tracking and Vocational interaction} 
 \end{table}
@@
 Both these coefficients are actually significantly different from each other, so it is possible.
-Before we explore this visually, let's note that the interaction between the two is very reliable and significant, considering that small sample size.
+Before we explore it visually, note that the interaction between the two is very reliable and significant, considering that small sample size.
 Below we graph the interaction for different quantiles of the vocational index.
 
+% \begin{figure}[!htbp]
+% \begin{center}
+\begin{figure}
+
 {\centering \includegraphics[width=5.5in,height=5in]{figure/interaction_plot-1}
@@
 }
+
+\caption[Interaction between tracking and vocational values]{Interaction between tracking and vocational values. Legend ordered relative to the lines}\label{fig:interaction_plot}
+\end{figure}
+
+% \caption{Interaction between tracking and vocational values. Legend ordered relative to the lines.}
+% \end{figure}
+% \end{center}
 
@@
 Moreover, our earlier interpretation (that tracking is more important than vocational enrollment for reducing the achievement gap) still holds and is very evident here.
-Despite the equalizing power of vocational enrollment, a high level of tracking with the highest level of vocational enrollment still leaves a country with an achievement of over 1.5 standard deviations.
-However, a country with low levels of tracking and \textit{any} level of vocational enrollment is nearly 1.3 standard deviations.
-A key finding from these results is that tracking is reasonably explaining the 90/10 achievement gap, and we should consider it as a very important component in the reduction of the achievement gap.
-However, it should not be studied in isolation from other features such as the vocational enrollment.
+Despite the equalizing power of vocational enrollment, a high level of tracking with the highest level of vocational enrollment still leaves a country with an achievement gap of over 1.5 standard deviations.
+However, a country with low levels of tracking and \textit{any} level of vocational enrollment is nearly 1.3 standard deviations at best.
+A key finding from these results is that tracking is a reasonable explanation of the 90/10 achievement gap, and we should consider it as a very important component in the reduction of the achievement gap.
+However, these results highlight that it should not be studied in isolation from other features such as the degree of vocational enrollment.
 We see a very sizable change in the achievement gap for a given country with the same level of tracking and high and low levels of vocational enrollment.
 
+The exact mechanisms through whick the tracking setup is explaining these inequalities is still very speculative.
+We could argue that tracking might've introduced the inequality once it was implemented.
+Another possibility is that tracking, althought not changing, is currently exarcebating it every year.
+Another explanation, which we find strong evidence for, is that the tracking setup interacts very closely with the vocational setup.
+These explanations are very speculative and further research should carry the task of looking for a more concrete explanation.
+
 % We need to define key numbers, to refer in the abstract, like the achievement gap closes % percent in the counterfactual and there's a difference of % percent between same level of tracking but different vocational enrollments.
@@
 
+% \begin{figure}[!htbp]
+% \begin{center}
+\begin{figure}
+
 {\centering \includegraphics[width=5.5in,height=5in]{figure/all_gaps_plot-1}
@@
 
+\caption[What explains achievement gaps? Model comparisons]{What explains achievement gaps? Model comparisons}\label{fig:all_gaps_plot}
+\end{figure}
+
+% \caption{What explains achievement gaps? Model comparisons}
+% \end{figure}
+% \end{center}
+
 With the exception of the Intercept, the coefficients of the three models look pretty much the same.
-The tracking index seems a but stronger for the 90/10 model but the uncertainty intervals overlap with the other two models.
+The tracking index seems slightly stronger for the 90/10 model but the uncertainty intervals overlap with the other two models.
 These results give our claims much more credibility.
@@
 Instead of estimating the evolution over time, we calculate the cumulative change over time for each country, summarizing the 6 year-time span of achievement gaps into one cumulative inequality gap.
-This approach dramatically reduces the total sample size to 34, the number of total countries available \footnote{We've been presenting a selected number of countries given the lack of space, but for every estimation we include all countries available.}.
-It is for this reason that I only include at most 2 covariates in the model, the tracking and vocational index, which both summarize the previous models and keep the model parsimonious.
+This approach dramatically reduces the total sample size to 34, the number of total countries available \footnote{We've been presenting a selected number of countries given the lack of space, but for every model/estimation we include all countries available.}.
+It is for this reason that we only include at most 2 covariates in the model, the tracking and vocational index, which both summarizes the previous models and keeps it parsimonious.
 
 % latex table generated in R 3.4.0 by xtable 1.8-2 package
-% Thu Sep  7 15:11:51 2017
+% Sat Sep  9 20:16:22 2017
 \begin{table}[ht]
@@
 \end{tabular}
+\caption{What explains the cumulative achievement gaps? Models for three different achievement gaps} 
 \end{table}
@@
 \begin{figure}
-\begin{center}
 
@@
 
-\caption{Evolution of the achievement gap for several gaps}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
+\caption[Evolution of the achievement gap for several gaps]{Evolution of the achievement gap for several gaps}\label{fig:graph_all_gaps}
+\end{figure}
+
+\begin{figure}
 
@@
 
-\caption{Percentage change in the 80/20 achievement gap from 2000 to 2015}
-\end{center}
-\end{figure}
-
-\begin{figure}
-\begin{center}
+\caption[Percentage change in the 80/20 achievement gap from 2000 to 2015]{Percentage change in the 80/20 achievement gap from 2000 to 2015}\label{fig:percentage_change_8020}
+\end{figure}
+
+\begin{figure}
 
@@
 
-\caption{Percentage change in the 70/30 achievement gap from 2000 to 2015}
-\end{center}
+\caption[Percentage change in the 70/30 achievement gap from 2000 to 2015]{Percentage change in the 70/30 achievement gap from 2000 to 2015}\label{fig:percentage_change_7030}
 \end{figure}

### 2017-09-12 paper/paper.Rnw
Finalized first draft of paper - ran everything: good

@@
 \usepackage{graphicx}
+\usepackage{titling}
 \usepackage[round, colon]{natbib}
@@
 
-\title{The not so widening achievement gap: an interntational comparison of the evolution of the SES achievement gap from 2000-2015}
+\title{Do educational institutions explain SES gaps?: an interntational comparison of the SES achievement gap from 2000-2015}
 \author{Jorge Cimentada}
@@
 
-\tableofcontents
+\begin{abstract}
+The literature on achievement inequality has only recently started studying the evolution of the socio-economic achievement gap in cognitive abilities and it has unconvered a wide variety of results.
+The main finding comes from American research and it shows that the gap has been widening by about 40-50\% in the last 30 year.
+Other international comparisons show that there's important variation among developed countries, but the research on this area is scarce.
+This paper looks to study whether there are specific patterns in the evolution of the gap in a comparative perspective.
+We find that there is considerable variation in the way in which the gap is evolving, with the U.S and Germany closing at a rapid pace and France widening at a similar rate.
+We find that this gap is strongly correlated with a countrys overall performance and simulations show that switching to less tracking can help decrease a countries SES gap by about 20\%.
+We also find that tracking and vocational enrollment explain over one third of the evolution of the achievement gap and show that the relationship is conditioned by a strong interaction: low tracking is associated with a small achievement gap, but high levels of tracking coped with high vocational enrollment can also remedy the adversive effects of tracking.
+\end{abstract}
 
@@
 
-\section{Literature Review}
 % You have a bunch of citations to take from here
@@
 When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family socio-economic status (SES) and student performance are tightly linked.
-The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements.
-For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.
+The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements \footnote{For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.}.
 But fast-forward until today, we've understood the relationship much better.
@@
 Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive development and defining personality traits \citep{duyme1999, waldfogel2006}.
-Economists and educators have long neglected these findings but the work of James Heckman has brought our attention to it.
-For example, he has clearly showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
+Economists and educators have long neglected these findings until the work of James Heckman brought our attention to it.
+For example, he has showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
 This means that inequality starts strong before the schooling of a child begins and schools have the potential to equalize the learning ground.
 These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions.
-But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies from time \(t-1\).
+But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies of time \(t-1\).
 Although this sounds logical, it's importance is not understood entirely.
@@
 Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity.
-The field of economics of education has also helped to understand that education is by far the investment that yields the best returns.
-This is not only for the child and his/her family, but to society as a whole, as it boosts economic activity, it helps the labour market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
+The field of economics of education has also helped to understand that early high-quality education is by far the best investment that yields the highest returns.
+This is not only for the child and his/her family, but to society as a whole as it boosts economic activity, it helps the labour market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
 Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about.
@@
 Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities.
-We have very little evidence, in comparative terms, for which countries have big or small cognitive gaps resulting from SES origins.
+We have very little comparative evidence on which countries have big or small cognitive gaps resulting from SES origins.
 And even more importantly, we haven't assesed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.
 
-However, the closest we have is the vast literature on social mobility.
+Despite these short-comings, we have indirect evidence of this coming from the vast literature on social mobility.
 Thanks to these findings, we know that virtually in all countries, be it developed or developing, there is inequality of opportunity.
@@
 From Denmark, for example, the literature has learned a great deal about how to improve social mobility.
-But that was possible by first learning that Denmark is by far one of the most mobile countries, boosting the chances of upward mobility specially for less advantaged children \citep{bjorklund2009, jaeger2007}.
+But that was possible by first learning that Denmark is by far one of the most mobile countries; this is the case specially for families of low-ses origin \citep{bjorklund2009, jaeger2007}.
 The important finding came when we discovered the main reason behind their social escalator: it's educational system.
 For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impact on improving opportunities.
-The education system is completely subsidized for all children, otherwise giving opportunities to children wouldn't have been able to pay.
+The education system is completely subsidized for all children, otherwise giving opportunities to families who wouldn't have been able to pay.
 In addition to this, Denmark is recognized as a world-leader in terms of public support for its child-care system as it spends around 2\% of GDP, and has among the highest enrollment rates for children under 6 years old \citep{esping_waldfogel2012}.
-Moreover, it tracks students into different curricula very late compared to other Europen countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}.
+Moreover, it separates students into different tracks very late compared to other Europen countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}.
 These two traits make the Danish system uniquely effective.
@@
 Given that the studies before 1970 are less reliable in terms of comparability and sampling design, the author computes all results for both before/after 1970.
-To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the U.S gap between the 90th and 10th income percentile is at about 1.25 standard deviations in the year 2001.
+To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the 90/10 income gap in the U.S is at about 1.25 standard deviations in the year 2001.
 Using longitudinal data, \citet{bradbury2015} find similar results.
@@
 Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of  1.25-1.50 standard deviations.
-One important drawback of these studies is that they don't present the uncertainty of this estimates.
+One important drawback of these studies is that they don't present the uncertainty of these estimates.
 Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
@@
 In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend.
-This follow-up study concentrated solely on the kindergarten children in which they took a snapshot of the achievement gap in for 1998, 2006 and 2010.
+This follow-up study concentrated solely on kindergarten children for the years 1998, 2006 and 2010.
 They found that the 90th/10th income gap in readiness closed modestly.
@@
 Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap).
-It should be noted that perhaps the reversal of trend in \citet{reardon2011} would be obvious if data were available for years after 2000,  which is specifically the time that \citet{reardon_portilla} find the reversal.
+It should be noted that perhaps the reversal of the trend in \citet{reardon2011} would had been evident if data were available for years after 2000, the time-point from which \citet{reardon_portilla} find the reversal.
 
 The reasons why the authors find a reversal in the trend could be numerous and should be studied very closely.
-The authors incorporate a number of country-level indicators to explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment.
+They incorporate a number of country-level indicators to explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment.
 They build on their previous argument by suggesting that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
@@
 
-\citet{anna2016_global}, building on the work of \citet{anna2016} and \citet{reardon_portilla} pooled together all the previously mentioned data, together with over 10 more studies ranging from the year 1964 until 2015 in order to discover differences between and across countries.
+\citet{anna2016_global}, building on the work of \citet{anna2016} and \citet{reardon_portilla} pooled together all the previously mentioned data, together with over 10 more studies ranging from 1964 to 2015 in order to discover differences between and across countries.
 With over 50 years of data, and over 100 countries, she finds that there seems to be a general pattern of increasing achievement gap.
@@
 The explanation, although very debated, has been gaining much support in recent years.
-If we remember correctly, for countries with high levels of curricular differentiation the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}.
+For countries with high levels of curricular differentiation the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}.
 Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}.
-Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts (because there have been tracking reforms in many countries, introducing as well as eliminating tracking structures) nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
+Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
 
@@
 c) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
-d) What explains this cross-country evolution;
+d) What explains this cross-country evolution.
 
@@
 
-\section{Hypothesis}
+\subsection{Hypothesis}
 
@@
 \begin{itemize}
-\item Hypothesis 1: In countries where there is a high degree of tracking we should expect the top and bottom to be evolving at the rate given that tracking is thought to maximize the learning experience of both groups.
-For countries with low levels of tracking we should expect the top and bottom groups to be closer to each other and for the gap to be closing since the both groups will be in the same track (this doesn't mean that the top is loweing their score but that the bottom is most likely catching up).
+\item Hypothesis 1: In countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups.
+For countries with low levels of tracking we should expect the top and bottom groups to be closer to each other and for the gap to be closing since the both groups will be in the same track (this doesn't mean that the top is lowering their score but that the bottom is most likely catching up).
 
@@
 
-To investigate the above mentioned questions I will be using the Programme for International Student Assessment (PISA).
+To investigate the above mentioned questions we will be using the Programme for International Student Assessment (PISA).
 PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students.
@@
 
-Aside from SES, the other most relevant variables are test scores for mathematics and literacy.
+Aside from SES, the other relevant variables are test scores for mathematics and literacy.
 PISA does not provide a single test result for each respondent.
@@
 The actual calculation of the achievement gap varies substantially and different strategies have been implemented.
-For example, \citet{micklewright} calculates the difference in achievement by crudely subtracting the gap between the 95th and 5th percentile of the mathematics distribution.
+For example, \citet{micklewright} calculates the difference in achievement by crudely subtracting the difference between the 95th and 5th percentile of the mathematics distribution.
 Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy.
@@
 for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011} \footnote{This does not mess up our analysis by masking age-specific gaps because all students in the sample are 15 year olds.
-Control for aging is simply to control for monthly differences in ages.}) weighted by the student sample weights.
+Controling for age is simply to adjust for monthly differences in ages.}) weighted by the student sample weights.
 
@@
 
-where \begin{math}r\end{math} is the reliability score of the wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}.
+where \begin{math}r\end{math} is the reliability score of the PISA wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}.
 The realibility estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of person separation reliability such as internal consistency.
@@
 At the moment of writing this paper, they were unavailable.
-For this reason we implement the analysis following the original work of \citet{reardon2011}}.
+For these reasons we implement the analysis following the original work of \citet{reardon2011}}.
 Each PISA survey provides a reliability indicator which we use accordingly.
@@
 
-Finally, we calculate the SES achievement gap for each country by extracting the fitted \(\alpha\) and \(\beta\) for each country \(j\) and calculating the difference of the predicted score of high and low SES.
+Finally, we calculate the SES achievement gap for each country by extracting the fitted \(\alpha\) and \(\beta\) for each country \(j\) and calculating the difference of the predicted score of High and Low SES.
 
@@
 
-The first table shows a description of the sample size and mean score of both top and bottom SES groups for only the first and last time point.
+Table 1 shows a description of the sample size and mean score of both top and bottom SES groups for only the first and last time point.
 One main concern from the planned analysis is that getting the top 90th percentile and bottom 10th percentile would result in a small sample size.
@@
 We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting they're very equal countries.
-However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average, which is 0.
+However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average.
 We also see some countries with major changes from year 2000 to 2015, with, for example, Australia decreasing the low SES gap from 0.22 standard deviations to -0.15, well below the country average.
@@
 Do note that the trendline is for both mathematics and literacy.
-If it were only for mathematics, then the trendline would probably be even more extreme.
+If it were only for mathematics, then the trendline would probably be even more pronounced.
 Despite this, pooling both subjects gives even more strength to the analysis given that the results stand for both subjects together \footnote{Same graph is availabe for reading upon request, where we find pretty much the same results if not even more steep/downward slopes.}.
-We also plot the 80/20 and 70/30 SES gaps and find that very similar patterns as in the 90/10 SES gap (see figure 10 in the appendix).
+We also plot the 80/20 and 70/30 SES gaps and find very similar patterns (see figure 13 in the appendix).
 
@@
 We find that the initial gap for the U.S in 2000 is \Sexpr{round(us_math_2000, 2)} varying between \Sexpr{round(us_math_2000_lower, 2)} and \Sexpr{round(us_math_2000_upper, 2)}.
-This gap is much higher than found in previous studies, but again, we don't know how much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely.
+This gap is much higher than found in previous studies, but again, we don't know how much previous estimates vary and nobody has really studied the 15-year old achievement gap in isolation.
 All country-level studies pool many studies together and control for age, removing the age-specific gap effect.
 
-Figure 3 below takes a more direct approach and looks at the percentage change from the first and last timepoint available.
+Figure 3 takes a more direct approach and looks at the percentage change from the first and last timepoint available.
 Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.
@@
 In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%.
-The benefit of presenting these estimates this way is that we can actually asses the uncertainty of each calculation, and we do find that some of these estimates have wide variability.
+The benefit of presenting these estimates this way is that we can actually asses the uncertainty of each calculation, and we do find that they have wide variability.
 Despite this, most countries show a clear sign of either decreasing or increasing.
-We plot the same graph for the 80/20 and 70/30 SES gaps in the appendix and the results hold for these achievement gaps as well.
+We plot the same graph for the 80/20 (figure 14) and 70/30 (figure 15) SES gaps in the appendix and the results hold for these achievement gaps as well.
 
@@
 The previos two graphs for reading are available upon request}.
-This means that the result is not an artifact of chance.
+This means that the result is not an artifact of chance alone.
 However, it's important to disentangle where is this gap originating from.
@@
 Or is it that the bottom is catching up?
-Up next we plot the same graph but show the divergent patterns between high/low SES origins.
+Up next we plot the same graph but show the divergent patterns between High/Low SES origins.
 
@@
 We find plenty of variation in these patterns.
-For example, in the United States the top seems to be equalizing much stronger than than the bottom is catching up.
+For example, in the United States the top seems to be equalizing much stronger than the bottom is catching up.
 The U.K seems to be following the same path as the United States as well.
 On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster.
-The Netherlands shows a similar pattern as the U.S, U.K patterns but upon closer inspection the explanation is very different.
+The Netherlands shows a similar pattern as the U.S and U.K patterns but upon closer inspection the explanation is very different.
 We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.
@@
 We now pay attention to the flat summary lines.
-These lines should be interpreted with caution because it doesn't mean that the gap is not increasing.
+These lines should be interpreted with caution because it doesn't mean that the gap is not changing.
 Denmark, Belgium, Netherlands, Norway and Spain show a flat line because the gap is growing very little, if at all.
 On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate.
-These results highlight the importance of not only summarizing average achievement gaps.
-The source of these gaps vary greatly between countries and we see how each of these patterns contibutes to the overall inequality of a country.
-
-After analyzing the trend for each country and where this trend is coming from we start to find some interesting results.
+These results highlight the importance of not only summarizing average achievement gaps: the source of the gap varies greatly between countries and we see how each of these patterns contibutes to the overall inequality of a country.
+
+After analyzing the trend of each country and where it's coming from, we start to find some interesting results.
 We see that the United States is closing the achievement gap at a very rapid rate, but at the same time we notice that it's because the top SES group is going down faster than the bottom is catching up.
-In contrast, we see countries like Germany and Poland where the gap is closing as well but because bottom group is catching up faster than the high SES group is coming down.
-In fact, the Polish case is the most egalitarian of all the countries presented here; the bottom group is catching up at a rapid pace and the top group is maintaining their high levels of performance, fostering a trully equalizing effect.
+In contrast, we see countries like Germany and Poland where the gap is closing as well but because the bottom group is catching up faster than the high SES group is coming down.
+In fact, the Polish case is the most egalitarian of all: the bottom group is catching up at a rapid pace and the top group is maintaining their high levels of performance, fostering a trully equalizing effect.
 
@@
 Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group.
-The big takeaway from the U.S is that the decrease of inequality is coming from the shrinking of the 50/10 gap.
+The surprising result is that the reductiong of U.S gap is coming from the shrinking of the 50/10 gap and not the 90/10 gap.
+These results are puzzling.
+Does this mean that the overall distribution of inequality in the U.S is being reduced toward the benefit of the Low SES group?
+The reduction is not uniform across all SES groups.
+We see that the 90/10 gap is closing, yet the the 90/50 gap is very stable over time.
+Despite this, the 50/10 is also shrinking, meaning that the 90 and 50 SES groups are shrinking down relative to the bottom while the bottom is coming up.
+This is something we can only speculate but the evidence points toward something like that.
 
@@
 We see no definitive evidence of top-to-bottom equality in all of these countries (except for Poland).
-We find that regardless of the different tracking setups, we see that there's a tradeoff: either the bottom comes up and the top comes down or the top goes even further up and the bottom further down.
+We find that regardless of the different tracking setups, there's a tradeoff: either the bottom comes up and the top comes down or the top goes even further up and the bottom further down.
 
@@
 
-Moving on, we now discuss if there's a tradeoff between an educational system being efficient, i.e raising the overall performance of the country and whether that challenge compromises equality at the same time.
+Moving on, we now discuss if there's a tradeoff between an educational system being efficient, i.e raising the overall performance of the country, and whether that challenge compromises equality at the same time.
 For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance.
 This, however, can come at a price.
-If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied with an increase in the overall performance.
+If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied by an increase in the overall performance.
 Given that we have the data for the average performance for each country and the achievement gap, we can investigate this relationship in detail.
@@
 This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}.
-However, what they didn't really go in depth was whether this corelation is strongest for some countries over time.
 However, we should not get trapped into an ecological fallacy and assume this relationship is identical within each country.
-With the data available we can look at the evolution separately for many countries.
+With the available data we can look at the evolution separately for countries.
 We plot it up next in figure 9.
@@
 
-\subsection{The role of curricular tracking in explaining inequalities}
+\subsection{The role of curricular tracking in explaining the achievement gap}
 
@@
 
-We now move to the final section of the paper, where we discuss what is a likely explanation to the patterns from before.
+We now move to the final section of the paper, where we discuss what is a likely explanation to these patterns.
 The reason why countries differ in their evolution of inequality is still unknown.
@@
 
-This is clearly a difficult task given that, at least at the country-level, not many features of the educational system change in a 15-year time span, resulting in a practically invariant variable from which to explain the level of inequality (without any variance, we can't explain anything.).
+This is clearly a difficult task given that, at least at the country-level, not many features of the educational system change in a 15-year time span, resulting in a practically invariant variable from which to explain the level of inequality.
 However, there are other strategies that can at least partially help us to understand the phenomena, and these are precisely the ones we follow.
@@
 Following the same strategy, we believe that tracking is also allowing the same gap in 15 year olds to grow over time, acompannied by other things.
-Yet, instead of concentrating only on the age of selection (just as they do), we explore a more fine grained definition of tracking, which is possible through the work of \citet{bol2013}.
+Yet, instead of concentrating only on the age of selection (as they do), we explore a more fine grained definition of tracking, which is possible through the work of \citet{bol2013}.
 Aside from the age at first selection into tracks, we also use the the number of tracks in the country, the percentage of the entire curriculum that is tracked and a vocational index \footnote{This index is a factor loading from a principal factor analysis of the percentage of students in upper secondary vocational education (taken from two sources, to reduce measurement error).
@@
 The final model has a maximum VIF of 2.2.}.
-We are reluctant to include more variable in the model, first to keep it more parsimonious, and secondly to prevent overfitting due to our sample size.
-We present the results in the table below.
+We are reluctant to include more variables in the model, first to keep preserve parsimony, and secondly to prevent overfitting due to our sample size.
+We present the results in table 2.
 
@@
 The first model only includes the 1 track dummy.
-This variable shows that countries that have one track, have on average, half a standard deviation less inequality than other countries.
+This variable shows that countries that have one track, have on average, half a standard deviation of less inequality than other countries.
 This covariate alone explains about 23\% of the variance of the evolution of the 90/10 achievement gap.
@@
 We also note that vocational enrollment is associated with less inequality.
-Not surprisingly, the age of selection coefficient, also increased due to the correlation between age of selection and vocational enrollment.
+Not surprisingly, the age of selection coefficient also increased due to the correlation between age of selection and vocational enrollment.
 Countries with high vocational enrollment are also those with early age selection, such as Germany and Austria.
@@
 Using the previous model we make a simulation and predict the level of the 90/10 achievement gap for all possible tracking setups for all countries.
-Below we plot all these combinations.
+In figure 10 we plot all these combinations.
 The X axis has the number of tracks that a country has, the Y axis has the predicted 90/10 achievement gap and below each country name we have whether the country has an age selection of 15 or more or below 15.
@@
 
-<<graph_complete_tracking_graph, include = T, fig.width = 9.5, out.height = '7in', out.width = '6.5in', fig.align = 'center', fig.cap = 'Simulation: predicted 90/10 achievement gap for different tracking setups'>>=
+<<graph_complete_tracking_graph, include = T, fig.height = 9, fig.width = 9, out.height = '7in', out.width = '6.5in', fig.align = 'center', fig.cap = 'Simulation: predicted 90/10 achievement gap for different tracking setups'>>=
 new_data <-
@@
 Had all countries in the 'ideal' tracking switched to the 'worst' tracking, we would see an average widening of nearly 72\%.
-For those countries which are not in the 'ideal' tracking setup, we compute below the percentage change from switching to the ideal tracking.
+For those countries which are not in the 'ideal' tracking setup, we compute the percentage change from switching to the ideal tracking in table 3.
 
@@
 Using this standardized tracking variable and the same bayesian multilevel model, we now model the tracking and vocational index as before, but add an interaction between two, given that in the previous model we recorded a strong shift in the model when tracking-related variables and the vocational model where both included in the same model.
-Below are the results for this model.
+Table 4 shows the results for this model.
 
@@
 We confirm the results from the previous models, where more tracking is associated with an achievement gap of about 0.4 standard deviations wider and more vocational enrollment is associated with a reduction of about 0.2 standard deviations, half the slope of tracking.
-Meaning that the tracking effect might be offsetting the effect of the vocational index given that it's half the size of the slope.
+Could it be that the tracking effect might be offsetting the effect of the vocational index given that it's half the size of the slope?.
 Both these coefficients are actually significantly different from each other, so it is possible.
-Before we explore it visually, note that the interaction between the two is very reliable and significant, considering that small sample size.
-Below we graph the interaction for different quantiles of the vocational index.
+Before we explore it visually, note that the interaction between the two is very reliable and significant, considering the small sample size.
+Figure 11 plots the interaction for different quantiles of the vocational index.
 
@@
 % \begin{center}
-<<interaction_plot, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Interaction between tracking and vocational values. Legend ordered relative to the lines'>>=
+<<interaction_plot, include = T, fig.width = 8, out.height = '5in', out.width = '6.5in', fig.align = 'center', fig.cap = 'Interaction between tracking and vocational values.
+Legend ordered relative to the lines'>>=
 plot_interaction <-
@@
   pull(colour)
+
+legend_labels <- seq(0, 100, 25) %>% paste0("%")
+legend_labels[1] <- paste0(legend_labels[1], " (no voc enrollment)")
+legend_labels[5] <- paste0(legend_labels[5], " (high voc enrollment)")
 
@@
     name = "Vocational index \n quantiles",
-    labels = seq(0, 100, 25) %>% paste0("%")
+    labels = legend_labels
   ) +
@@
 The results are very interesting.
-We see that for lower levels of tracking (meaning little to no tracking) theres a very low achievement gap, regardless of whether the country has very high or very low vocational enrollment.
+We see that for lower levels of tracking (meaning little to no tracking) there's a very low achievement gap, regardless of whether the country has very high or very low vocational enrollment.
 However, once tracking enters the picture, vocational enrollment can be a strong equalizer.
@@
 
-To test whether these results hold for other samples, we compute the same models as before but also for the 80/20 and 70/30 SES gap. Up next we plot these coefficients for the three SES gaps.
+To test whether these results hold for other samples, we compute the same models as before but also for the 80/20 and 70/30 SES gap. Figure 12 plots these coefficients for the three SES gaps.
 
@@
 Instead of estimating the evolution over time, we calculate the cumulative change over time for each country, summarizing the 6 year-time span of achievement gaps into one cumulative inequality gap.
-This approach dramatically reduces the total sample size to 34, the number of total countries available \footnote{We've been presenting a selected number of countries given the lack of space, but for every model/estimation we include all countries available.}.
+This approach dramatically reduces the total sample size to 32, the number of total countries available \footnote{We've been presenting a selected number of countries given the lack of space, but for every model/estimation we include all countries available.}.
 It is for this reason that we only include at most 2 covariates in the model, the tracking and vocational index, which both summarizes the previous models and keeps it parsimonious.
@@
 
+\section{Conclusion}
+
+The literature on achievement inequality has recently started studying the evolution of the achievement gap and it has unconvered a wide variety of results.
+The main finding is that the U.S gap has been widening over time by about 40-50\% from 1970 until the 2000.
+Others find that this recent trend has reversed, at least for an age-specific group since the 2000's.
+From an international perspective, recent studies have shown that there is wide variability in this trend, with some countries showing decreases while others increases.
+This paper looks to investigate the magnitud of these gaps across selected countries, and attempts to explain why they are happening.
+
+We start off by looking at the evolution of the achievement gap and document the same variation for the 15-year old achievement gap across countries.
+The United States has been closing its 90/10 gap by about 50\% with a wide interval going down to nearly 35\%.
+This reduction is also present in Germany, quite shockingly a country with a very institutionalized tracking system.
+Other countries such as France and Austria experience the contrary with persistent increases in the achievement gap.
+Moreover, once we disaggregate these patterns into a more fine-graned analysis, we discover that the dynamics are very different.
+The U.S gap is shrinking because both the 90th and 50th SES groups are actually going down while the bottom group is coming up.
+These results are evident because the 90/10 and 50/10 gap are shrinking at a very rapid rate, whereas the 90/50 gap is flat in the same 15 year interval.
+Germany, which has a very similar reduction, portrays a different explanation.
+All three achievement gaps are shrinking suggesting a more uniform reduction.
+On the other hand, for countries which are experiencing a widening of the gap, we see that it's happening at all instances of the distribution; all SES groups show a very steep increase.
+
+We then assess whether this gap is correlated with the country's average performance, and find a very strong correlation at the aggregate level (that is, all year-country combination pooled).
+Yet when we look at this relationship within each country, we find exactly the same pattern: countries with high performance show low inequality, whereas countries with low performance show high inequality.
+This gives greater support to the work of \citet{werfhorst_mijs}.
+Further research should attempt to disentangle the causality between these two opposite forces: is efficiency trumping inequality, or is it the other way around?
+
+We complete the analysis by studying what explains the evolution of this inequality and we find that the degree of tracking of a country seems to be a possible explanation, but it shouldn't be studied without factoring in the degree of vocational enrollment.
+We find that if a country has a low level of tracking, vocational enrollment will not help decrease the achievement gap.
+Conversely, once tracking enters the picture, vocational enrollment can help unease the burden of inequality and reduce the achievement gap very strongly.
+In a similar vein, we use this model to peform a simulation for each country: if every country switched to little or no tracking, what would be the resulting reduction in the 90/10 achievement gap?
+We find that if all countries switched, we would find that on average countries would see their gap shrink by about 20\%, with countries like Hungary and Germany experiencing reductions of 46\% and 36\% respectively.
+We contrast the former simulation with another counterfactual scenario: what would be the average gap increase if all countries switched to a system where tracking predominates?
+We record an average widening of 72\%, something that suggests that increasing tracking is more than three times worse than reducing it.
+
+We conclude by showing that tracking and vocational enrollment are playing an important role in explaining the evolution of the achievement gap.
+We find that these results are very suggestive that institutional features, more specifically, stratified curriculums, might be playing a very sizable role in the promotion of inequality.
+After having ran a battery of tests and different empirical strategies, the results are consistent and point out that we should pay attention to the source of the achievement gaps and study whether particular SES groups are contributing more to the gap than others.
+Aside from the results, this paper also revelead some puzzling questions that we leave up to future research to study.
+
 \bibliography{mybibliography.bib}
@@
 
-<<graph_all_gaps, include = T, out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Evolution of the achievement gap for several gaps'>>=
+<<graph_all_gaps, include = T, fig.pos = '!htbp', out.height = '5in', out.width = '5.5in', fig.align = 'center', fig.cap = 'Evolution of the achievement gap for several gaps'>>=
 complete_gaps %>%

### 2017-09-12 paper/paper.tex
Finalized first draft of paper - ran everything: good

@@
 We find that the initial gap for the U.S in 2000 is 2.17 varying between 2.13 and 2.21.
-This gap is much higher than found in previous studies, but again, we don't know how much previous estimates vary and nobody has really studied the 15-year old achievement gap precisely.
+This gap is much higher than found in previous studies, but again, we don't know how much previous estimates vary and nobody has really studied the 15-year old achievement gap in isolation.
 All country-level studies pool many studies together and control for age, removing the age-specific gap effect.
 
-Figure 3 below takes a more direct approach and looks at the percentage change from the first and last timepoint available.
+Figure 3 takes a more direct approach and looks at the percentage change from the first and last timepoint available.
 Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.
@@
 
-{\centering \includegraphics[width=5.5in,height=5in]{figure/interaction_plot-1}
+{\centering \includegraphics[width=6.5in,height=5in]{figure/interaction_plot-1}
 
@@
 
-\begin{figure}
+\begin{figure}[!htbp]
 

### 2017-10-06 paper/paper.Rnw
Remove PISA2012 packae because wasn't used

@@
   library(PISA2009lite)
-  library(PISA2012lite)
   library(intsvy)

### 2017-10-06 transform_data.R

@@ -4,7 +4,6 @@ library(PISA2000lite)
 library(PISA2003lite)
 library(PISA2006lite)
 library(PISA2009lite)
-library(PISA2012lite)
 library(intsvy)
 library(cimentadaj) # For pisa_countrynames
 library(countrycode) # For region variable

### 2017-10-06 playing_around/reardon_attempt.R
Useless file - now use the perccalc package which implements this

@@ -1,36 +1,0 @@
-# This is, so far, a failed attempt at calculating the 90/10
-# percentiles for an ordinal variable. This is an implementation taken
-# from Reardon (2011), so please check the paper first.
-
-library(GoFKernel)
-library(tidyverse)
-
-x <- rep(1:7, each = 10)
-y <- seq(250, 500, length.out = 70)
-
-x_density <- density(x, adjust = 2)
-plot(our_density)
-
-lookup <- tapply(X = our_density$y, INDEX = rep(1:7, each = 73), FUN = mean)
-prop <- c(0, unique(cume_dist(sort(m$high_edu_broad))) * 100)
-
-our_dens_fun <- function(x) x * density(x, adjust = 2, n = length(x))$y
-our_dens_fun(x)
-
-x <- 1
-integrate(our_dens_fun, lower = c_k[1], upper = c_k[1 + 1])
-
-inverse_cdf <- inverse(integrate(our_dens_fun, lower = prop[x], upper = prop[x + 1]))
-
-# given a ck and ck_prev
-theta_bottom <- inverse_cdf(prop[x])
-theta_top <- inverse_cdf(prop[x + 1])
-
-# A(1) from paper
-fn <- function (x) {
-  x * our_density(x)
-}
-
-theta_hat <- integrate(fn, theta_bottom, theta_top) / (ck - ck_prev)
-
-lm(y ~ 1 + theta_hat + (theta_hat^2 + ck_bla bla))

### 2017-10-09 paper/paper.Rnw
Added direct call to script that creates data from .Rnw
Spell-checked the paper

@@
   library(countrycode) # For region variable
+  library(SAScii)
   library(car)
-  library(SAScii)
   library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
@@
   library(tidyverse)
-
-  # source("./transform_data.R")
 
@@
 \begin{abstract}
-The literature on achievement inequality has only recently started studying the evolution of the socio-economic achievement gap in cognitive abilities and it has unconvered a wide variety of results.
+The literature on achievement inequality has only recently started studying the evolution of the socio-economic achievement gap in cognitive abilities and it has uncovered a wide variety of results.
 The main finding comes from American research and it shows that the gap has been widening by about 40-50\% in the last 30 year.
@@
 We find that there is considerable variation in the way in which the gap is evolving, with the U.S and Germany closing at a rapid pace and France widening at a similar rate.
-We find that this gap is strongly correlated with a countrys overall performance and simulations show that switching to less tracking can help decrease a countries SES gap by about 20\%.
-We also find that tracking and vocational enrollment explain over one third of the evolution of the achievement gap and show that the relationship is conditioned by a strong interaction: low tracking is associated with a small achievement gap, but high levels of tracking coped with high vocational enrollment can also remedy the adversive effects of tracking.
+We find that this gap is strongly correlated with a country's overall performance and simulations show that switching to less tracking can help decrease a countries SES gap by about 20\%.
+We also find that tracking and vocational enrollment explain over one third of the evolution of the achievement gap and show that the relationship is conditioned by a strong interaction: low tracking is associated with a small achievement gap, but high levels of tracking coped with high vocational enrollment can also remedy the aversive effects of tracking.
 \end{abstract}
@@
 % code chunk
+
+<<creating_data, eval = F>>=
+# For now I don't run it, but it should be run when you leave the paper as done.
+
+# This file downloads PISA 2012 and PISA 2015 and saves them in ./data/pisa_listcol.Rdata
+source("./transform_data.R")
+@
+
 <<loading_data-recoding>>=
@@
 # When turning 0:6 to numeric, it becomes 1:7 that's why
-# I recode 8:9 to NA. This, however, didn't work for last two surveys
+# I recode 8:9 to NA. This, however, did not work for last two surveys
 
@@
   # Next we'll merge the ESCS data with the PISA data. As explained above, the 6th data (PISA
-  # 2015) doesn't need to be merged so I exclude it with this vector
+  # 2015) does not need to be merged so I exclude it with this vector
   exclude <- -6
@@
 This means that inequality starts strong before the schooling of a child begins and schools have the potential to equalize the learning ground.
-These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavourable conditions.
-But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiencies of time \(t-1\).
+These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavorable conditions.
+But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiences of time \(t-1\).
 Although this sounds logical, it's importance is not understood entirely.
@@
 For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital.
-Social mobility has deepend our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity.
+Social mobility has deepen our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity.
 The field of economics of education has also helped to understand that early high-quality education is by far the best investment that yields the highest returns.
-This is not only for the child and his/her family, but to society as a whole as it boosts economic activity, it helps the labour market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
+This is not only for the child and his/her family, but to society as a whole as it boosts economic activity, it helps the labor market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
 Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about.
 Naturally, we want to do that in order to reverse it and help every child reach its fullest potential.
-Despite our good intentions, we've concentrated little on the magnitud of the gaps, specially in terms of cognitive abilities.
+Despite our good intentions, we've concentrated little on the magnitude of the gaps, specially in terms of cognitive abilities.
 We have very little comparative evidence on which countries have big or small cognitive gaps resulting from SES origins.
-And even more importantly, we haven't assesed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.
+And even more importantly, we haven't assessed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.
 
@@
 Thanks to these findings, we know that virtually in all countries, be it developed or developing, there is inequality of opportunity.
-But there is considerable variation in the magnitud of this effect.
+But there is considerable variation in the magnitude of this effect.
 For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries \citep{esping2012, breen2007, shavit1993}.
@@
 For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impact on improving opportunities.
-The education system is completely subsidized for all children, otherwise giving opportunities to families who wouldn't have been able to pay.
+The education system is completely subsidized for all children, otherwise giving opportunities to families who would not have been able to pay.
 In addition to this, Denmark is recognized as a world-leader in terms of public support for its child-care system as it spends around 2\% of GDP, and has among the highest enrollment rates for children under 6 years old \citep{esping_waldfogel2012}.
-Moreover, it separates students into different tracks very late compared to other Europen countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}.
+Moreover, it separates students into different tracks very late compared to other European countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}.
 These two traits make the Danish system uniquely effective.
@@
 
-\citet{reardon2011} finds that the increase in the gaps has occured predominantely from the 1970's until the 2000's.
+\citet{reardon2011} finds that the increase in the gaps has occurred predominantly from the 1970's until the 2000's.
 In fact, the hard numbers suggest that the gap widened by 40-50\%.
@@
 One important drawback of these studies is that they don't present the uncertainty of these estimates.
-Not necessarily to gauge their statistical significance, but to simply asses how much we can trust their accuracy.
+Not necessarily to gauge their statistical significance, but to simply assess how much we can trust their accuracy.
 It could be that the gap is at  1.25 standard deviations, varying up to 1.75 and down to 0.80.
@@
 A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality.
-However, if the achiement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts betwen 1950 and 1970 where income inequality was increasing.}.
+However, if the achievement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts between 1950 and 1970 where income inequality was increasing.}.
 
@@
 They found that the 90th/10th income gap in readiness closed modestly.
-Furthermore, using data from fall and spring in the same kindergarden year, they calculated that the gap narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010.
-They also calculated the same changes for a number of personality traits such as self-control and externalizing behaviour and found similar results.
-In contrast, \citet{reardon2011} finds that in a 30-year span the gap was systematicaly increasing at a rate of 0.02, something reasonably close to the previous estimates.
+Furthermore, using data from fall and spring in the same kindergarten year, they calculated that the gap narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010.
+They also calculated the same changes for a number of personality traits such as self-control and externalizing behavior and found similar results.
+In contrast, \citet{reardon2011} finds that in a 30-year span the gap was systematically increasing at a rate of 0.02, something reasonably close to the previous estimates.
 Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap).
@@
 The core finding behind their study is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada.
-They find that once the achievement gap is present in early school entry, it doesn't seem to narrow or widen very much over the life course.
+They find that once the achievement gap is present in early school entry, it does not seem to narrow or widen very much over the life course.
 In fact, they calculate that the quality of early education can only explain about  30-40\% of the high school SES gap.
-This suggests that once the achievement gap is present before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous RCT's show that high quality schooling can indeed help ease the gap, in some instances even close it \citep{campbell2002}.}.
+This suggests that once the achievement gap is present before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous Randomized Controlled Trials (RCT) show that high quality schooling can indeed help ease the gap, in some instances even close it \citep{campbell2002}.}.
 One exception is the United Kingdom, which they found to be a country that helps close the gap in early primary years.
@@
 They do an amazing work at making the four surveys comparable and their findings are indeed very reliable.
-But we should be careful at intepreting these findings causally and should be taken as suggestive at best.
+But we should be careful at interpreting these findings causally and should be taken as suggestive at best.
 For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citet{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in the gap using much more comparable surveys.
@@
 However, once she disentangles the relationship by country, she finds a sizable amount of heterogeneity, with some countries seeing the achievement gap narrowing, others no change at all, while others record a steady increase.
-This is revealing because it doesn't really pay off to look at a general average once each country has their own distinctive gap and evolution.
+This is revealing because it does not really pay off to look at a general average once each country has their own distinctive gap and evolution.
 This highlights the notion that the increasing achievement gap is clearly not universal and should be studied in context.
@@
 Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}.
-Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts nor the same between ages, because tracking/no tracking might exarcebate the achievement gap.
+Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts nor the same between ages, because tracking/no tracking might exacerbate the achievement gap.
 
@@
 Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States.
-This will help asses the mangitude of the inequality as well as the rate of growth/decline.
+This will help assess the magnitude of the inequality as well as the rate of growth/decline.
 This study not only documents the size and changes of the SES gap, but pays particular attention to the source of the changes.
@@
 This pattern is not so obvious, however.
-For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score.
+For example, countries with high levels of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score.
 But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with a widening achievement gap.
@@
 \item Hypothesis 1: In countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups.
-For countries with low levels of tracking we should expect the top and bottom groups to be closer to each other and for the gap to be closing since the both groups will be in the same track (this doesn't mean that the top is lowering their score but that the bottom is most likely catching up).
+For countries with low levels of tracking we should expect the top and bottom groups to be closer to each other and for the gap to be closing since the both groups will be in the same track (this does not mean that the top is lowering their score but that the bottom is most likely catching up).
 
@@
 We hypothesize that the degree of tracking is tightly related to changes in the gap, and the more tracking, the more inequality.
-Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labour market opportunities.
+Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labor market opportunities.
 \end{itemize}
@@
 PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students.
-Currrently, PISA has six waves starting in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 developed/developing countries.
+Currently, PISA has six waves starting in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 developed/developing countries.
 
@@
 The sample size for PISA 2000 is \Sexpr{country_rows[1]}, \Sexpr{country_rows[2]} for PISA 2003, \Sexpr{country_rows[3]} for PISA 2006, \Sexpr{country_rows[4]} for PISA 2009, \Sexpr{country_rows[5]} for PISA 2012, and \Sexpr{country_rows[6]} for PISA 2015.
-Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance.
+Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevant background information that can be matched to the students performance.
 With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system.
@@
 
-The ESCS was derived from a principal component analysis of standardised variables, taking the factor scores for the first principal component as measures of the PISA index of economic, social and cultural status.
+The ESCS was derived from a principal component analysis of standardized variables, taking the factor scores for the first principal component as measures of the PISA index of economic, social and cultural status.
 All countries and economies (both OECD and partner countries/economies) contributed equally to the principal component analysis, while in previous cycles, the principal component analysis was based on OECD countries only.
@@
 
-To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.
+To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indexes can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled indexes for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.
 
@@
 
-A more intuitivity is explanation is this: suppose we have \(\mu_i\), the average student test score in mathematics for student \(i\).
+A more intuitive explanation is this: suppose we have \(\mu_i\), the average student test score in mathematics for student \(i\).
 Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\) for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test.
@@
 for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011} \footnote{This does not mess up our analysis by masking age-specific gaps because all students in the sample are 15 year olds.
-Controling for age is simply to adjust for monthly differences in ages.}) weighted by the student sample weights.
+Controlling for age is simply to adjust for monthly differences in ages.}) weighted by the student sample weights.
 
@@
 We plot the standard deviation of the mathematics test score for all waves in figure 1.
-The plot suggests that it's something we shouldn't be deeply concerned with.
+The plot suggests that it's something we should not be deeply concerned with.
 The standard deviation of each wave seems to be following a very similar pattern with a not so drastic exception of the year 2000.
@@
 Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap.
-This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citep{reardon2011}.
+This can be very misleading and suggest erroneous interpretations regarding trends in the of the gaps over time \citep{reardon2011}.
 PISA has tried to make sure the tests are comparable across waves \citep{pisa2012_technical}, but we still have to adjust for this imprecision.
@@
 
-where \begin{math}r\end{math} is the reliability score of the PISA wave \footnote{Other proceedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}.
-The realibility estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of person separation reliability such as internal consistency.
+where \begin{math}r\end{math} is the reliability score of the PISA wave \footnote{Other procedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}.
+The reliability estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of person separation reliability such as internal consistency.
 Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own.
@@
           # In some countries the quan can't be estimated because of very few obs.
-          # The function doesn't stop but returns two NA's.
+          # The function does not stop but returns two NA's.
           quan <- quantile_missing(.x, weights_var, probs)
@@
       unsplit_df %>%
-        count(country, escs_dummy) %>%
+        dplyr::count(country, escs_dummy) %>%
         filter(!is.na(escs_dummy)) %>%
@@
 For example, France has an increase of \Sexpr{diff_data$read[["France"]]}.
-For such a short period of time, the magnitud of these increases are reasonably big.
+For such a short period of time, the magnitude of these increases are reasonably big.
 
@@
 If it were only for mathematics, then the trendline would probably be even more pronounced.
-Despite this, pooling both subjects gives even more strength to the analysis given that the results stand for both subjects together \footnote{Same graph is availabe for reading upon request, where we find pretty much the same results if not even more steep/downward slopes.}.
+Despite this, pooling both subjects gives even more strength to the analysis given that the results stand for both subjects together \footnote{Same graph is available for reading upon request, where we find pretty much the same results if not even more steep/downward slopes.}.
 We also plot the 80/20 and 70/30 SES gaps and find very similar patterns (see figure 13 in the appendix).
@@
 Take the case of Sweden.
-The slope doesn't look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015.
+The slope does not look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015.
 Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.
@@
 France had an average increase of about 80\% since 2000 varying down to 40\%, whereas Germany had a similar figure but decreasing.
-Many of the countries we saw before that didn't have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap.
+Many of the countries we saw before that did not have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap.
 In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%.
-The benefit of presenting these estimates this way is that we can actually asses the uncertainty of each calculation, and we do find that they have wide variability.
+The benefit of presenting these estimates this way is that we can actually assess the uncertainty of each calculation, and we do find that they have wide variability.
 Despite this, most countries show a clear sign of either decreasing or increasing.
@@
 Something reassuring is that mathematics and reading (not presented) follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase.
-The previos two graphs for reading are available upon request}.
+The previous two graphs for reading are available upon request}.
 This means that the result is not an artifact of chance alone.
@@
 We now pay attention to the flat summary lines.
-These lines should be interpreted with caution because it doesn't mean that the gap is not changing.
+These lines should be interpreted with caution because it does not mean that the gap is not changing.
 Denmark, Belgium, Netherlands, Norway and Spain show a flat line because the gap is growing very little, if at all.
 On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate.
-These results highlight the importance of not only summarizing average achievement gaps: the source of the gap varies greatly between countries and we see how each of these patterns contibutes to the overall inequality of a country.
+These results highlight the importance of not only summarizing average achievement gaps: the source of the gap varies greatly between countries and we see how each of these patterns contributes to the overall inequality of a country.
 
@@
 In contrast, we see countries like Germany and Poland where the gap is closing as well but because the bottom group is catching up faster than the high SES group is coming down.
-In fact, the Polish case is the most egalitarian of all: the bottom group is catching up at a rapid pace and the top group is maintaining their high levels of performance, fostering a trully equalizing effect.
+In fact, the Polish case is the most egalitarian of all: the bottom group is catching up at a rapid pace and the top group is maintaining their high levels of performance, fostering a truly equalizing effect.
 
@@
 Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger in mathematics.
-Furthermore, we find, just as in the 90th/10th SES gap, that there isn't a significant decrease \footnote{With the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation.}.
+Furthermore, we find, just as in the 90th/10th SES gap, that there is not a significant decrease \footnote{With the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation.}.
 France shows a similar pattern as Denmark in that the 50th/10th SES gap is the main contributer and that the three gaps have the similar trend.
@@
 Both the 90th/50th/10th gaps seem to contribute rather similarly to the overall 90th/10th gap (although 50th/10th is marginally higher) but more interestingly, the 90th/50th gap has stagnated in the last 15 years.
-The top/middle groups have a relatively big gap and it hasn't changed at all.
+The top/middle groups have a relatively big gap and it has not changed at all.
 On the other hand, the steep decrease we saw for the U.S is coming exclusively from the 50th/10th gap shrinking.
@@
 Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group.
-The surprising result is that the reductiong of U.S gap is coming from the shrinking of the 50/10 gap and not the 90/10 gap.
+The surprising result is that the reduction of U.S gap is coming from the shrinking of the 50/10 gap and not the 90/10 gap.
 These results are puzzling.
@@
 However, this means that the top group comes down, rather than stagnate or increase.
-Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decresing (bottom left panel).
+Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decreasing (bottom left panel).
 On the other hand, we do see more countries where the both groups are increasing (top right panel).
@@
 To sum up, we find no evidence in support of our first hypothesis.
-We uncover that the tracking setup of a country doesn't imminently mean a certain degree of equality/inequality.
+We uncover that the tracking setup of a country does not imminently mean a certain degree of equality/inequality.
 Countries where the top SES group is increasing and the bottom is decreasing (top-left panel) have dramatically different tracking setups.
 Austria is an early tracker with several tracks, while Sweden is a late tracker with only 1 track.
-On the opossite panel (bottom-right), where we find countries that the bottom group is catching up while the top is coming down, we find similar results.
+On the opposite panel (bottom-right), where we find countries that the bottom group is catching up while the top is coming down, we find similar results.
 Denmark and Germany are opposite poles in terms of tracking, while the U.S is distinctive in that it has little formal curricular stratification.
@@
 This, however, can come at a price.
-If tracking actually benefits the top but doesn't increase the performance of the bottom tracks then we see a net increase of inequality accompanied by an increase in the overall performance.
+If tracking actually benefits the top but does not increase the performance of the bottom tracks then we see a net increase of inequality accompanied by an increase in the overall performance.
 Given that we have the data for the average performance for each country and the achievement gap, we can investigate this relationship in detail.
@@
 We see that as the average performance has been decreasing, the average achievement gap has been decreasing as well.
-In fact, among the few countries we should find something that resembles this pattern should be the U.S because it doesn't have a strict institutionalized tracking system.
+In fact, among the few countries we should find something that resembles this pattern should be the U.S because it does not have a strict institutionalized tracking system.
 However, the explanation behind this pattern is very speculative.
@@
 Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap.
-Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentially, are countries that have the lower level of inequalities among the ones presented here.
+Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentally, are countries that have the lower level of inequalities among the ones presented here.
 
@@
 Conversely, France, Austria and Australia, to name a few, portray the opposite: as average performance decreases inequality increases.
-But we also unconver important variation within countries in the relationship between average achievement and dispersion of achievement (have a look at the Netherlands, Italy and the U.S).
+But we also uncovered important variation within countries in the relationship between average achievement and dispersion of achievement (have a look at the Netherlands, Italy and the U.S).
 
@@
 Following the literature on the topic, we suspect that this evolution is in part very tightly related to the degree of stratification of the educational system.
-In a very simple frame-work, \citet{hanushek_woesmann_tracking} found that tracking indeed exarcebates inequality in the life course of children.
-Following the same strategy, we believe that tracking is also allowing the same gap in 15 year olds to grow over time, acompannied by other things.
+In a very simple frame-work, \citet{hanushek_woesmann_tracking} found that tracking indeed exacerbates inequality in the life course of children.
+Following the same strategy, we believe that tracking is also allowing the same gap in 15 year olds to grow over time, accompanied by other things.
 Yet, instead of concentrating only on the age of selection (as they do), we explore a more fine grained definition of tracking, which is possible through the work of \citet{bol2013}.
@@
 
-Considering that our summarized data has very few observartions \footnote{We have per year, per country, one single inequality indicator, leaving the total sample size at \Sexpr{nrow(ready_data_age)}}, we want to asses the uncertainty of our model by using a fully bayesian hierarchichal linear model.
-The benefits of this approach are twofold: first, we can asses the model fit much more intuitively, something very important given our sample size, and secondly, given that our countries are clustered into years, we can control for country heterogeneity within years, removing any bias in the standard errors.
+Considering that our summarized data has very few observations \footnote{We have per year, per country, one single inequality indicator, leaving the total sample size at \Sexpr{nrow(ready_data_age)}}, we want to assess the uncertainty of our model by using a fully bayesian hierarchical linear model.
+The benefits of this approach are twofold: first, we can assess the model fit much more intuitively, something very important given our sample size, and secondly, given that our countries are clustered into years, we can control for country heterogeneity within years, removing any bias in the standard errors.
 
@@
 
-Given that the classical p values are inexistent in bayesian modelling, we present uncertainty intervals next to each coefficient.
+Given that the classical p values are inexistent in bayesian modeling, we present uncertainty intervals next to each coefficient.
 The first model only includes the 1 track dummy.
@@
 France, which had a particularly steep increase in the last 15 years, would see a reduction of the 90/10 SES gap of 31\% and Austria, with a very similar tracking structure as the German case, would see their achievement gap reduced by 20\%.
-The very suprising result is the Dutch case, for which a switch from the 'worst' tracking setup to the 'best' case would result in an increase of the gap of 19\%.
+The very surprising result is the Dutch case, for which a switch from the 'worst' tracking setup to the 'best' case would result in an increase of the gap of 19\%.
 
@@
 First and foremost, the Dutch educational system has built other mechanisms that, despite the tracking system, help children from low social origins escalate upwards.
-This explanation is not very trivial considering that in most mobility research, the Netherlands stands out as a very equal country relative to other countries with a similary tracking structure, like Germany \citep{shavit1993, esping2012}.
+This explanation is not very trivial considering that in most mobility research, the Netherlands stands out as a very equal country relative to other countries with a similar tracking structure, like Germany \citep{shavit1993, esping2012}.
 Another possible explanation, and something which needs to assessed, is whether our model is predicting accurately.
@@
 
-The exact mechanisms through whick the tracking setup is explaining these inequalities is still very speculative.
+The exact mechanisms through which the tracking setup is explaining these inequalities is still very speculative.
 We could argue that tracking might've introduced the inequality once it was implemented.
-Another possibility is that tracking, althought not changing, is currently exarcebating it every year.
+Another possibility is that tracking, although not changing, is currently exacerbating it every year.
 Another explanation, which we find strong evidence for, is that the tracking setup interacts very closely with the vocational setup.
@@
 
-The literature on achievement inequality has recently started studying the evolution of the achievement gap and it has unconvered a wide variety of results.
+The literature on achievement inequality has recently started studying the evolution of the achievement gap and it has uncovered a wide variety of results.
 The main finding is that the U.S gap has been widening over time by about 40-50\% from 1970 until the 2000.
@@
 From an international perspective, recent studies have shown that there is wide variability in this trend, with some countries showing decreases while others increases.
-This paper looks to investigate the magnitud of these gaps across selected countries, and attempts to explain why they are happening.
+This paper looks to investigate the magnitude of these gaps across selected countries, and attempts to explain why they are happening.
 
@@
 Other countries such as France and Austria experience the contrary with persistent increases in the achievement gap.
-Moreover, once we disaggregate these patterns into a more fine-graned analysis, we discover that the dynamics are very different.
+Moreover, once we disaggregate these patterns into a more fine-grained analysis, we discover that the dynamics are very different.
 The U.S gap is shrinking because both the 90th and 50th SES groups are actually going down while the bottom group is coming up.
@@
 
-We complete the analysis by studying what explains the evolution of this inequality and we find that the degree of tracking of a country seems to be a possible explanation, but it shouldn't be studied without factoring in the degree of vocational enrollment.
+We complete the analysis by studying what explains the evolution of this inequality and we find that the degree of tracking of a country seems to be a possible explanation, but it should not be studied without factoring in the degree of vocational enrollment.
 We find that if a country has a low level of tracking, vocational enrollment will not help decrease the achievement gap.
 Conversely, once tracking enters the picture, vocational enrollment can help unease the burden of inequality and reduce the achievement gap very strongly.
-In a similar vein, we use this model to peform a simulation for each country: if every country switched to little or no tracking, what would be the resulting reduction in the 90/10 achievement gap?
+In a similar vein, we use this model to perform a simulation for each country: if every country switched to little or no tracking, what would be the resulting reduction in the 90/10 achievement gap?
 We find that if all countries switched, we would find that on average countries would see their gap shrink by about 20\%, with countries like Hungary and Germany experiencing reductions of 46\% and 36\% respectively.
@@
 After having ran a battery of tests and different empirical strategies, the results are consistent and point out that we should pay attention to the source of the achievement gaps and study whether particular SES groups are contributing more to the gap than others.
-Aside from the results, this paper also revelead some puzzling questions that we leave up to future research to study.
+Aside from the results, this paper also revealed some puzzling questions that we leave up to future research to study.
 

### 2017-10-09 paper/paper.tex
Spell-checked the paper

@@
 The sample size for PISA 2000 is 127,388, 276,165 for PISA 2003, 398,750 for PISA 2006, 515,958 for PISA 2009, 480,810 for PISA 2012, and 519,334 for PISA 2015.
-Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevent background information that can be matched to the students performance.
+Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevant background information that can be matched to the students performance.
 With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system.
@@
 For example, France has an increase of 0.6.
-For such a short period of time, the magnitud of these increases are reasonably big.
+For such a short period of time, the magnitude of these increases are reasonably big.
 
@@
 
-Considering that our summarized data has very few observartions \footnote{We have per year, per country, one single inequality indicator, leaving the total sample size at 195}, we want to asses the uncertainty of our model by using a fully bayesian hierarchichal linear model.
-The benefits of this approach are twofold: first, we can asses the model fit much more intuitively, something very important given our sample size, and secondly, given that our countries are clustered into years, we can control for country heterogeneity within years, removing any bias in the standard errors.
+Considering that our summarized data has very few observations \footnote{We have per year, per country, one single inequality indicator, leaving the total sample size at 195}, we want to assess the uncertainty of our model by using a fully bayesian hierarchical linear model.
+The benefits of this approach are twofold: first, we can assess the model fit much more intuitively, something very important given our sample size, and secondly, given that our countries are clustered into years, we can control for country heterogeneity within years, removing any bias in the standard errors.
 

### 2017-10-09 paper/paper.Rnw
Updated hypothesis section + some text

@@
 For example, he has showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
-This means that inequality starts strong before the schooling of a child begins and schools have the potential to equalize the learning ground.
-These inequalities can be narrowed significantly with high quality learning environments, specially for children coming from unfavorable conditions.
-But one of the most important findings from \citet{cunha2006} is that the cognitive level of a child in time \(t\) is a direct function of the experiences of time \(t-1\).
+To explain these, and other findings, \citet{cunha2006} suggest that the cognitive level of a child in time \(t\) is a direct function of the experiences of time \(t-1\).
 Although this sounds logical, it's importance is not understood entirely.
-
-The implications of this cumulative model suggests that it is much less cost-effective to invest in children in time \(t\) that in \(t-1\).
-That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is cheaper than investing at later stages and it brings about the largest returns.
+The model implies that investments on children are not perfect substitutes at different time points but rather complements.
+There are specific periods of skill formation where it is much more cost-effective to invest in, which are the early stages of the life course.
+In fact, when tested against the data, \citet{cunha2006} show that their explanation fits the results of most of the empirical literature very well.
+
+For this reason, children who receive less (or worse) investments have a cumulative disadvantage against children who do receive quality investments.
+This leads, as early as preschool, to an unequal start in life.
+That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is cheaper than investing at later stages and it brings about the largest returns \citep{elango15}.
 Despite all of these findings and strategies to reduce the gaps, we still find that the relationship between children's parental education and future destination is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.
@@
 Social mobility has deepen our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity.
-The field of economics of education has also helped to understand that early high-quality education is by far the best investment that yields the highest returns.
+The field of economics of education has also helped to understand that early high-quality education is by far the best investment in terms of cognitive development and acquiring personality skills.
+Research has constantly found that early education helps low-income children boost their performance to levels similar to their well-off peers \citep{heckman_oecd}.
+Yet the positive results are not only limited to the cognitive realm: studies have found that that early stimuli is associated with less propensity to criminal activity, more stable relationships, higher levels of persistance, higher income and improved labor market opportunities in late stages in life \citep{heckman_oecd}.
+
 This is not only for the child and his/her family, but to society as a whole as it boosts economic activity, it helps the labor market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
@@
 From Denmark, for example, the literature has learned a great deal about how to improve social mobility.
-But that was possible by first learning that Denmark is by far one of the most mobile countries; this is the case specially for families of low-ses origin \citep{bjorklund2009, jaeger2007}.
-The important finding came when we discovered the main reason behind their social escalator: it's educational system.
+But that was possible by first learning that Denmark is by far one of the most mobile countries compared to other European countries; this is the case specially for families of low-ses origin \citep{bjorklund2009, jaeger2007}.
+The important finding came when research emerged the main reason behind their social escalator: it's educational system.
 For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impact on improving opportunities.
@@
 Moreover, it separates students into different tracks very late compared to other European countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}.
-These two traits make the Danish system uniquely effective.
-The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable.
+These two traits make the Danish system effective in promoting equality relative to the average European country.
+The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable \citep{heckman_oecd}.
 In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
@@
 
-However likely, this is a highly contested topic.
-A recent \href{http://www.epi.org/files/pdf/101972.pdf}{report} by the Economic Policy Institute finds that the black-white income gap has been growing since 1979, very similar to the increase of income inequality.
-However, if the achievement gap would be a function of income inequality, we should expect for the white-black achievement gap to be widening rather than narrowing, and those are not the findings given by \citet{reardon2011} and \citet{magnuson2008} \footnote{The black-white achievement gap narrowed significantly for cohorts between 1950 and 1970 where income inequality was increasing.}.
-
 In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend.
@@
 
-\subsection{Research questions}
-
-This study seeks to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries and search for a likely explanation.
-This is different from previous work because it concentrates solely on 15 year old children, something not done for the achievement gap, and it attempts to capture the evolution of the achievement gap for countries with different educational systems.
-In addition, it extends the previous findings by studying the relationship between the gap and the average performance of each country in order to detect if there's an over-time tradeoff between good performance and equality over time separately for each country.
-The advantages of this study are threefold.
-
-First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the kindergarten gap in \citet{reardon_portilla}.
-As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
-Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.
-
-Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States.
-This will help assess the magnitude of the inequality as well as the rate of growth/decline.
-This study not only documents the size and changes of the SES gap, but pays particular attention to the source of the changes.
-That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.
-
-Third, we attempt to model why the achievement gap is changing over time. In contrast to other studies, this is a novelty given that most research has been done cross-sectionally between countries.
-
-More formally, we're interested in studying
-a) Which countries are experiencing changes in the achievement gap;
-b) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind;
-c) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
-d) What explains this cross-country evolution.
-
-We develop each question separately for more detail.
-
-a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
-After recording a SES gap increase of about 40\% in only 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap over time.
-We will compare the percentage change at which the gap widened/narrowed from the first to the last year available.
-This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
-Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}
-
-b) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
+\subsection{Research questions and hypothesis}
+
+The overaching aim of this paper is to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participating countries and search for a likely explanation for the evolution.
+
+We develop the subquestions and their corresponding hypothesis separately in more detail.
+
+\begin{enumerate}
+
+\item The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker than we though.
+After recording a SES gap increase of about 40\% in 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap.
+First, we will concentrate on the evolution of the gap only for 15 year olds (which as we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap), which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the kindergarten gap in \citet{reardon_portilla}.
+Secondly, we will compare the percentage change at which the gap widened/narrowed from the first to the last year available.
+This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature.
+\footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
+Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}.
+We posit no specific hypothesis for this question given that it's purely explanatory.
+
+\begin{enumerate}
+\item Research question:
+\begin{enumerate}
+\item Which countries are experiencing changes in the achievement gap?
+\end{enumerate}
+\end{enumerate}
+\end{enumerate}
+
+b)
+
+The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
 The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind.
 We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.
-
-c) We want to investigate whether better performing countries have lower levels of inequality than other countries.
+Following that strategy will put us closer to understanding the source of the evolution of the achievement gap.
+
+Research question:
+* Is the gap originating because the top is gaining ground, the top is falling behind or because of a dyanmic interaction between the two?
+
+Hypothesis 1:
+* The theorical arugment in favor of tracking posits that in countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups.
+We expect to find the opposite, i.e for countries with a highly tracked curriculum, we should find more inequality and children from low-ses groups at a greater disadvantage.
+Conversely, in countries with low tracking we should expect for the achievement gap to be narrowing.
+
+c)
+
+Our study itself is composed of several cross-sectional surveys, which is limited to single cross-sectional snapshots, but given we have 6 times points for most countries we can model the evolution of the achievement gap over time, something not done before in an international context.
+We want to test whether several dimensions of the tracking setup and vocational enrollment explain the changes in the gaps, something already discussed in \citet{werfhorst_mijs} and \citet{hanushek_woesmann_tracking}.
+
+Research question:
+* Are tracking and vocational enrollment related to the evolution of the gap?
+
+Hypothesis 2:
+* Building on the previous hypotheiss, tracking should play an important role on the evolution of the achievement gap.
+We hypothesize that the degree of tracking of an educational system is tightly related to changes in the gap, and the more tracking, the more inequality.
+Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labor market opportunities.
+
+d)
+
+We want to investigate whether better performing countries have lower levels of inequality than other countries.
 \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
@@
 
-d) Tightly linked to the previous point, we want to test whether several dimensions of the tracking setup explain why the gaps are changing.
-This will support the previous points in suggesting that the dynamics of the over-time achievement gap are endogenously intertwined with the effect that tracking has on the specific changes of the high and low SES groups.
-
-\subsection{Hypothesis}
-
-Following the previous questions and the literature on educational inequality and tracking we pose three hypothesis for answering questions (b), (c) and (d) given that question (a) is purely an exploratory exercise.
-
-\begin{itemize}
-\item Hypothesis 1: In countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups.
-For countries with low levels of tracking we should expect the top and bottom groups to be closer to each other and for the gap to be closing since the both groups will be in the same track (this does not mean that the top is lowering their score but that the bottom is most likely catching up).
-
-\item Hypothesis 2: Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality.
+Research question:
+* Are achievement and inequality correlated? and if so, has the relationship weakeaned/strengthend over time?
+
+Hypothesis 3:
+* Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality.
 Countries with high performance should have less inequality and countries with low performance should have greater inequality
-
-\item Hypothesis 3: if any (or both) of the previous hypothesis are confirmed, then tracking should play an important role on the evolution of the achievement gap.
-We hypothesize that the degree of tracking is tightly related to changes in the gap, and the more tracking, the more inequality.
-Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labor market opportunities.
-\end{itemize}
 

### 2017-10-10 paper/paper.Rnw
Rewrote hypothesis section + introduction

@@
 
-\title{Do educational institutions explain SES gaps?: an interntational comparison of the SES achievement gap from 2000-2015}
+\title{Do educational institutions explain SES gaps? an international comparison of the SES achievement gap from 2000-2015}
 \author{Jorge Cimentada}
@@
 
-% You have a bunch of citations to take from here
+The impact of inequality, be it economic or of non pecuniary causes, is a crucial determinant for the opportunities of a child.
+Evidence from disciplines as disparate as epidemiology, psychology, sociology and economics keep piling evidence that suggests that inequality in the life course is something present in virtually all societies.
+This is something that politicians and policymakers are aware of and, yet despite their many efforts, inequality has not abated.
+
+The literature has found conclusive evidence showing that, for example, the United States has increased both economic inequality \citep{piketty2003}, inequality of opportunity \citep{chetty2014} and inequality of achievement \citep{reardon2011}.
+All of these findings, however, are from the United States.
+Other countries such as the United Kingdom or other developed countries have their fair share of studies when there is data available.
+However, very rarely can we compare levels of inequality (of any kind), across several countries with a great degree of confidence.
+
+This paper looks to study the cognitive achievement gap between the 90th and 10th percentile of the SES distribution.
+This gap can be termed as the SES achievement gap between the top and bottom SES groups.
+Using the PISA database, we calculate the achievement gap for every country and for all six available PISA waves, building a country panel that shows a time trend of 15 years for each country.
+
+More concretely, we aim to study the evolution of the achievement gap.
+We aim to answer two broad question: Which countries are changing over time while others are stagnated?
+Equally important, what explains these changes?
+Building on the literature on educational institutions we suggest the degree of tracking of a country and it's level of vocational enrollments as possible explanations for the changes of a country over time.
+
+Findings suggest that there's reasonable variation in the evolution of the gap.
+For example, the United States has been closing its 90/10 gap by about 50\% with a wide interval going down to nearly 35\%.
+This reduction is also present in other countries such as Germany and Poland.
+Other countries such as France and Austria experience the contrary with persistent increases in the achievement gap.
+Robustness checks shows that also the 80/20 and 70/30 replicate the results for nearly all countries.
+
+We then assess whether this gap is correlated with the country's average performance, and find a very strong negative correlation for all country pooled.
+This suggests that countries with high levels of achievement have low levels of inequality.
+Yet these results are aggregated for every year.
+A more disaggregate analysis shows that this pattern is also present for every country over time, although some exceptions are encountered, like the United States where achievement and inequality as positively correlated.
+
+To finish, we find that the degree of tracking of a country seems to be a likely explanation to the changes in the achievement gap, but it should not be studied without factoring in the degree of vocational enrollment.
+We find that if a country has a low level of tracking, vocational enrollment will not help decrease the achievement gap.
+Conversely, once tracking enters the picture, vocational enrollment can help unease the burden of inequality and reduce the achievement gap very strongly.
+Simulations show that if every country switched to little or no tracking their gaps would shrink by about 20\%.
+
+These findings have two specific implications.
+First, inequality should not be 'generalized' or 'summarized' with one single indicator for many countries.
+There are widespread differences between countries, where each one should be studied in context.
+Secondly, now that we know that, further research should attempt to explain why some countries follow such different evolutions.
+We propose a possible explanation, the degree of tracking and vocational enrollment of the country.
+Tracking seems to exacerbate inequality, but if coped with a reasonable level of vocational enrollment it can alleviate the negative effects of tracking.
+Further research should confirm these results hold for other empirical tests and if so, policymakers should incorporate vocational enrollment not only as a 'labor market' solution to youth unemployment but as a compensatory mechanism for highly tracked curriculums.
+
+\section{Literature Review}
 Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years.
@@
 Research has constantly found that early education helps low-income children boost their performance to levels similar to their well-off peers \citep{heckman_oecd}.
-Yet the positive results are not only limited to the cognitive realm: studies have found that that early stimuli is associated with less propensity to criminal activity, more stable relationships, higher levels of persistance, higher income and improved labor market opportunities in late stages in life \citep{heckman_oecd}.
+Yet the positive results are not only limited to the cognitive realm: studies have found that that early stimuli is associated with less propensity to criminal activity, more stable relationships, higher levels of persistence, higher income and improved labor market opportunities in late stages in life \citep{heckman_oecd}.
 
@@
 
-\section{Research framework}
-
-\subsection{Research questions and hypothesis}
-
-The overaching aim of this paper is to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participating countries and search for a likely explanation for the evolution.
-
+\section{Research questions and hypothesis}
+
+The overarching aim of this paper is to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participating countries and search for a likely explanation for the evolution.
+
+\noindent
 We develop the subquestions and their corresponding hypothesis separately in more detail.
@@
 
-\item The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker than we though.
-After recording a SES gap increase of about 40\% in 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap.
+\item The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker than we though after recording a SES gap increase of about 40\% in 30 years.
+\citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap.
 First, we will concentrate on the evolution of the gap only for 15 year olds (which as we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap), which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the kindergarten gap in \citet{reardon_portilla}.
@@
 
-\begin{enumerate}
+\begin{itemize}
 \item Research question:
 \begin{enumerate}
+
 \item Which countries are experiencing changes in the achievement gap?
+
 \end{enumerate}
-\end{enumerate}
-\end{enumerate}
-
-b)
-
-The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
-The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind.
+\end{itemize}
+
+\item The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
+The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because the bottom performers are falling behind or because both are changing at the same time.
 We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.
@@
 
-Research question:
-* Is the gap originating because the top is gaining ground, the top is falling behind or because of a dyanmic interaction between the two?
-
-Hypothesis 1:
-* The theorical arugment in favor of tracking posits that in countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups.
-We expect to find the opposite, i.e for countries with a highly tracked curriculum, we should find more inequality and children from low-ses groups at a greater disadvantage.
+\begin{itemize}
+\item Research question:
+\begin{enumerate}
+
+\item Is the gap originating because the top is gaining ground, the top is falling behind or because of a dynamic interaction between the two?
+
+\item Hypothesis 1: The theoretical argument in favor of tracking posits that in countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups.
+We expect to find the opposite, i.e for countries with a highly tracked curriculum, we should find more inequality and children from low-SES groups at a greater disadvantage.
 Conversely, in countries with low tracking we should expect for the achievement gap to be narrowing.
 
-c)
-
-Our study itself is composed of several cross-sectional surveys, which is limited to single cross-sectional snapshots, but given we have 6 times points for most countries we can model the evolution of the achievement gap over time, something not done before in an international context.
-We want to test whether several dimensions of the tracking setup and vocational enrollment explain the changes in the gaps, something already discussed in \citet{werfhorst_mijs} and \citet{hanushek_woesmann_tracking}.
-
-Research question:
-* Are tracking and vocational enrollment related to the evolution of the gap?
-
-Hypothesis 2:
-* Building on the previous hypotheiss, tracking should play an important role on the evolution of the achievement gap.
+\end{enumerate}
+\end{itemize}
+
+\item Our study itself is composed of six cross-sectional surveys, which limits the design to single country snapshots, but given that most countries participated in all six surveys we can model the evolution of the achievement gap over time.
+We want to test whether several dimensions of the tracking setup and vocational enrollment explain the changes in the gaps, something already discussed in \citet{werfhorst_mijs} and \citet{hanushek_woesmann_tracking} but not empirically tested before in an international context.
+
+\begin{itemize}
+\item Research question:
+\begin{enumerate}
+
+\item Are tracking and vocational enrollment related to the evolution of the gap?
+
+\item Hypothesis 2: Building on the previous hypothesis, tracking should play an important role on the evolution of the achievement gap.
 We hypothesize that the degree of tracking of an educational system is tightly related to changes in the gap, and the more tracking, the more inequality.
@@
 
-d)
-
-We want to investigate whether better performing countries have lower levels of inequality than other countries.
+\end{enumerate}
+\end{itemize}
+
+\item Lastly, we want to investigate whether better performing countries have lower levels of inequality than other countries.
 \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
@@
 
-Research question:
-* Are achievement and inequality correlated? and if so, has the relationship weakeaned/strengthend over time?
-
-Hypothesis 3:
-* Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality.
+\begin{itemize}
+\item Research question:
+\begin{enumerate}
+
+\item Are achievement and inequality correlated? and if so, has the relationship weakened/strengthen over time?
+
+\item Hypothesis 3: Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality.
 Countries with high performance should have less inequality and countries with low performance should have greater inequality
+
+\end{enumerate}
+\end{itemize}
+
+\end{enumerate}
 
@@
 None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.
+This paper takes advantage of these six waves to build a country pseudo-panel where each country-wave will have its corresponding achievement gap, making it possible to study changes in nearly 15 years for each country.
 

### 2017-10-10 paper/paper.tex
Rewrote hypothesis section + introduction

@@
 Social mobility has deepen our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity.
-The field of economics of education has also helped to understand that early high-quality education is by far the best investment that yields the highest returns.
+The field of economics of education has also helped to understand that early high-quality education is by far the best investment in terms of cognitive development and acquiring personality skills.
+Research has constantly found that early education helps low-income children boost their performance to levels similar to their well-off peers \citep{heckman_oecd}.
+Yet the positive results are not only limited to the cognitive realm: studies have found that that early stimuli is associated with less propensity to criminal activity, more stable relationships, higher levels of persistence, higher income and improved labor market opportunities in late stages in life \citep{heckman_oecd}.
+
 This is not only for the child and his/her family, but to society as a whole as it boosts economic activity, it helps the labor market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
@@
 
-\section{Research framework}
-
-\subsection{Research questions}
-
-This study seeks to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participant countries and search for a likely explanation.
-This is different from previous work because it concentrates solely on 15 year old children, something not done for the achievement gap, and it attempts to capture the evolution of the achievement gap for countries with different educational systems.
-In addition, it extends the previous findings by studying the relationship between the gap and the average performance of each country in order to detect if there's an over-time tradeoff between good performance and equality over time separately for each country.
-The advantages of this study are threefold.
-
-First, we concentrate on the evolution of the gap for only 15 year olds, which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the kindergarten gap in \citet{reardon_portilla}.
-As we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap.
-Moreover, in almost all countries with a tracked curriculum children are either at or in the process of tracking by the age 15, meaning that we will be able to link whether tracked countries are the most variable in their evolution of achievement gaps.
-
-Second, we will be able, for the first time, to study the evolution of the achievement gap for several countries other than the United States.
-This will help assess the magnitude of the inequality as well as the rate of growth/decline.
-This study not only documents the size and changes of the SES gap, but pays particular attention to the source of the changes.
-That is, we study whether bottom performers are falling behind, if top performers are gaining advantages, or if both phenomenas are simultaneously at play.
-
-Third, we attempt to model why the achievement gap is changing over time. In contrast to other studies, this is a novelty given that most research has been done cross-sectionally between countries.
-
-More formally, we're interested in studying
-a) Which countries are experiencing changes in the achievement gap;
-b) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind;
-c) The size of the achievement in a comparative perspective and its relationship to overall achievement levels;
-d) What explains this cross-country evolution.
-
-We develop each question separately for more detail.
-
-a) The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though.
-After recording a SES gap increase of about 40\% in only 30 years, \citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap over time.
-We will compare the percentage change at which the gap widened/narrowed from the first to the last year available.
-This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
-Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}
-
-b) The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
-The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind.
+\section{Research questions and hypothesis}
+
+The overarching aim of this paper is to study the evolution of the high/low SES achievement gap in the past 15 years for all PISA participating countries and search for a likely explanation for the evolution.
+
+\noindent
+We develop the subquestions and their corresponding hypothesis separately in more detail.
+
+\begin{enumerate}
+
+\item The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker than we though after recording a SES gap increase of about 40\% in 30 years.
+\citet{reardon_portilla} stress that they also found a significant decrease in only 15 years of data, showing how important it is to study the changes in the achievement gap.
+First, we will concentrate on the evolution of the gap only for 15 year olds (which as we've seen before, there are reasons to think that specific age-groups have seen changes in the achievement gap), which will serve as a comparison to the single year-country snapshot of \citet{anna2016} and the evolution of the kindergarten gap in \citet{reardon_portilla}.
+Secondly, we will compare the percentage change at which the gap widened/narrowed from the first to the last year available.
+This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature.
+\footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time.
+Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders using PIRLS.}.
+We posit no specific hypothesis for this question given that it's purely explanatory.
+
+\begin{itemize}
+\item Research question:
+\begin{enumerate}
+
+\item Which countries are experiencing changes in the achievement gap?
+
+\end{enumerate}
+\end{itemize}
+
+\item The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment.
+The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because the bottom performers are falling behind or because both are changing at the same time.
 We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time.
-
-c) We want to investigate whether better performing countries have lower levels of inequality than other countries.
+Following that strategy will put us closer to understanding the source of the evolution of the achievement gap.
+
+\begin{itemize}
+\item Research question:
+\begin{enumerate}
+
+\item Is the gap originating because the top is gaining ground, the top is falling behind or because of a dynamic interaction between the two?
+
+\item Hypothesis 1: The theoretical argument in favor of tracking posits that in countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups.
+We expect to find the opposite, i.e for countries with a highly tracked curriculum, we should find more inequality and children from low-SES groups at a greater disadvantage.
+Conversely, in countries with low tracking we should expect for the achievement gap to be narrowing.
+
+\end{enumerate}
+\end{itemize}
+
+\item Our study itself is composed of six cross-sectional surveys, which limits the design to single country snapshots, but given that most countries participated in all six surveys we can model the evolution of the achievement gap over time.
+We want to test whether several dimensions of the tracking setup and vocational enrollment explain the changes in the gaps, something already discussed in \citet{werfhorst_mijs} and \citet{hanushek_woesmann_tracking} but not empirically tested before in an international context.
+
+\begin{itemize}
+\item Research question:
+\begin{enumerate}
+
+\item Are tracking and vocational enrollment related to the evolution of the gap?
+
+\item Hypothesis 2: Building on the previous hypothesis, tracking should play an important role on the evolution of the achievement gap.
+We hypothesize that the degree of tracking of an educational system is tightly related to changes in the gap, and the more tracking, the more inequality.
+Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labor market opportunities.
+
+\end{enumerate}
+\end{itemize}
+
+\item Lastly, we want to investigate whether better performing countries have lower levels of inequality than other countries.
 \citet{werfhorst_mijs} emphasize that there is empirical evidence that suggests this.
@@
 
-d) Tightly linked to the previous point, we want to test whether several dimensions of the tracking setup explain why the gaps are changing.
-This will support the previous points in suggesting that the dynamics of the over-time achievement gap are endogenously intertwined with the effect that tracking has on the specific changes of the high and low SES groups.
-
-\subsection{Hypothesis}
-
-Following the previous questions and the literature on educational inequality and tracking we pose three hypothesis for answering questions (b), (c) and (d) given that question (a) is purely an exploratory exercise.
-
 \begin{itemize}
-\item Hypothesis 1: In countries where there is a high degree of tracking we should expect the top and bottom to be evolving at a similar rate given that tracking is thought to maximize the learning experience of both groups.
-For countries with low levels of tracking we should expect the top and bottom groups to be closer to each other and for the gap to be closing since the both groups will be in the same track (this does not mean that the top is lowering their score but that the bottom is most likely catching up).
-
-\item Hypothesis 2: Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality.
+\item Research question:
+\begin{enumerate}
+
+\item Are achievement and inequality correlated? and if so, has the relationship weakened/strengthen over time?
+
+\item Hypothesis 3: Building on \citet{werfhorst_mijs}, then there should be a negative relationship between the average country performance and their level of inequality.
 Countries with high performance should have less inequality and countries with low performance should have greater inequality
 
-\item Hypothesis 3: if any (or both) of the previous hypothesis are confirmed, then tracking should play an important role on the evolution of the achievement gap.
-We hypothesize that the degree of tracking is tightly related to changes in the gap, and the more tracking, the more inequality.
-Moreover, the more vocational tracking, the less inequality considering that it gives short term returns in terms of labor market opportunities.
+\end{enumerate}
 \end{itemize}
+
+\end{enumerate}
 

### 2018-01-23 paper/paper.Rnw
Last commit before delete local folder - everything moved to thesis/ repo

@@
-% Things to remember:
-% The working directory is not where the data is.
-% Change working directory to "/Users/cimentadaj/Downloads/gitrepo/Achievement_Gap_PISA/paper"
-% for everything below to work.
-
-% There's a code chunk named modeling which lasts about 3-4 hours to run. I explain how to get
-% around that in a comment above the code chunk. For down there.
-
-\documentclass[11pt, a4paper]{article}
-\bibliographystyle{apalike}
-\pagestyle{headings}
-
-\usepackage{geometry}
- \geometry{
- a4paper,
- total={170mm,257mm},
- left=20mm,
- top=20mm,
- }
-\usepackage{pdflscape}
-\usepackage{amsmath}
-\usepackage{subcaption}
-\usepackage{graphics}
-\usepackage{graphicx}
-\usepackage{titling}
-\usepackage[round, colon]{natbib}
-\usepackage[colorlinks]{hyperref}
-\usepackage{booktabs}
-\AtBeginDocument{%
-  \hypersetup{
-    citecolor=blue,
-    linkcolor=blue,   
-    urlcolor=blue}}
-
-\title{Do educational institutions explain SES gaps? an international comparison of the SES achievement gap from 2000-2015}
-\author{Jorge Cimentada}
-
-\begin{document}
-\setlength{\parindent}{2em}
-\setlength{\parskip}{1em}
-\showboxdepth=5
-\showboxbreadth=5
-
-\maketitle
-
-<<working directory, echo = F>>=
-options(scipen = 213141)
-
-opts_chunk$set(echo = FALSE,
-               message = FALSE,
-               cache = TRUE,
-               warning = FALSE,
-               include = FALSE,
-               cache.lazy = FALSE,
-               results = 'asis')
-@
-
-<<default_conf>>=
-  library(matrixStats)
-  library(knitr)
-  library(arm)
-  library(saves)
-  library(haven)
-  library(PISA2000lite)
-  library(PISA2003lite)
-  library(PISA2006lite)
-  library(PISA2009lite)
-  library(intsvy)
-  library(cimentadaj) # devtools::install_github("cimentadaj/cimentadaj")
-  library(countrycode) # For region variable
-  library(SAScii)
-  library(car)
-  library(inequalityintsvy) # devtools::install_github("cimentadaj/inequalityintsvy")
-  library(lme4)
-  library(modelr)
-  library(ggrepel)
-  library(xtable)
-  library(ggthemes)
-  library(readxl)
-  library(inequalityintsvy)
-  library(brms)
-  library(tidyverse)
-
-  # Conf for PISA_2015
-  pisa2015_conf <- list(variables = list(pvlabelpref = "PV",
-                                         pvlabelsuff = "READ",
-                                         weightFinal = "W_FSTUWT",
-                                         weightBRR = "W_FSTURWT"),
-          parameters = list(cutoffs = c(357.77, 420.07, 482.38, 544.68, 606.99, 669.30),
-                                          percentiles = c(5, 10, 25, 75, 90, 95),
-                                          PVreps = 10,
-                                          BRRreps = 80,
-                                          weights = "BRR",
-                                          replication_scheme = 'pisa')
-  )
-
-  countries <- c("Finland",
-                 "France",
-                 # "New Zealand",
-                 "Austria",
-                 "Australia",
-                 "Sweden",
-                 # "Czech Republic",
-                 "Canada",
-                 "Hungary",
-                 # "Iceland",
-                 "Netherlands",
-                 "Spain",
-                 # "Belgium",
-                 "Italy",
-                 # "Norway",
-                 "United Kingdom",
-                 # "Greece",
-                 "Denmark",
-                 # "Israel",
-                 "Poland",
-                 "United States",
-                 "Germany"
-                 # "Turkey",
-                 # "Russia"
-                 )
-@
-
-\begin{abstract}
-The literature on achievement inequality has only recently started studying the evolution of the socio-economic achievement gap in cognitive abilities and it has uncovered a wide variety of results.
-The main finding comes from American research and it shows that the gap has been widening by about 40-50\% in the last 30 year.
-Other international comparisons show that there's important variation among developed countries, but the research on this area is scarce.
-This paper looks to study whether there are specific patterns in the evolution of the gap in a comparative perspective.
-We find that there is considerable variation in the way in which the gap is evolving, with the U.S and Germany closing at a rapid pace and France widening at a similar rate.
-We find that this gap is strongly correlated with a country's overall performance and simulations show that switching to less tracking can help decrease a countries SES gap by about 20\%.
-We also find that tracking and vocational enrollment explain over one third of the evolution of the achievement gap and show that the relationship is conditioned by a strong interaction: low tracking is associated with a small achievement gap, but high levels of tracking coped with high vocational enrollment can also remedy the aversive effects of tracking.
-\end{abstract}
-
-% pisa_listcol.Rdata is a dataset produced by the source() expression from the above
-% code chunk
-
-<<creating_data, eval = F>>=
-# For now I don't run it, but it should be run when you leave the paper as done.
-
-# This file downloads PISA 2012 and PISA 2015 and saves them in ./data/pisa_listcol.Rdata
-source("./transform_data.R")
-@
-
-<<loading_data-recoding>>=
-pisa_all <- read_rds("./data/pisa_listcol.Rdata")
-
-pisa_all2 <- pisa_all
-
-years <- seq(2000, 2015, 3)
-
-db <- paste0("pisa", years)
-pisa_all2$value <- map2(pisa_all2$value, db, ~ { .x$wave <- .y; .x})
-pisa_all2$value[[1]]$CNT <- pisa_all2$value[[1]]$COUNTRY
-
-pisa_all2$value <- map(pisa_all2$value, ~ {
-
-# 2000 to 2015
-# The coding is from 0 to 6, where 0 is no schooling and 6 is
-# BA or above.
-
-# When turning 0:6 to numeric, it becomes 1:7 that's why
-# I recode 8:9 to NA. This, however, did not work for last two surveys
-
-  .x$father_edu <- car::recode(as.numeric(.x$FISCED), "8:9 = NA")
-  .x$mother_edu <- car::recode(as.numeric(.x$MISCED), "8:9 = NA")
-  .x$high_edu_broad <- pmax(.x$father_edu, .x$mother_edu)
-  .x$country <- pisa_countrynames[as.character(.x$CNT)]
-
-  if (any(unique(.x$wave) %in% c("pisa2012", "pisa2015"))) {
-    # These two surveys were from 0:6 so I had to add + 1
-    # so that it equals 1:7 as all other surveys.
-    .x$father_edu <- .x$father_edu + 1
-    .x$mother_edu <- .x$mother_edu + 1
-    .x$high_edu_broad <- .x$high_edu_broad + 1
-  }
-  .x
-})
-
-reliability_pisa <-
-  c("2000" = 0.81,
-    "2003" = 0.85,
-    "2006" = 0.78,
-    "2009" = 0.74,
-    "2012" = 0.82,
-    "2015" = 0.74) # 2015 imputed
-
-@
-
-<<escs_trend>>=
-  # Rescaled trend ESCS data to merge.
-  # This only has data for seq(2000, 2012, 3) because
-  # PISA 2015 has the ESCS trend variable.
-  dir <- tempdir()
-  file_name <- "escs_trend.zip"
-  download.file("http://vs-web-fs-1.oecd.org/pisa/trend_escs_SPSS.zip",
-                destfile = file.path(dir, file_name))
-  unzip(file.path(dir, file_name), exdir = dir)
-  escs_trend <- map(file.path(dir, list.files(dir, pattern = ".sav")), haven::read_spss)
-  file.remove(file.path(dir, list.files(dir)))
-
-  escs_trend <-
-    map(escs_trend, ~ {
-    mutate(.x, cnt = pisa_countrynames[cnt]) %>%
-    rename(country = cnt)
-  })
-@
-
-<<merge_escs_pisa>>=
-  # Next we'll merge the ESCS data with the PISA data. As explained above, the 6th data (PISA
-  # 2015) does not need to be merged so I exclude it with this vector
-  exclude <- -6
-
-  # Loop in parallel to the PISA data, the ESCS data and the year vector (which is seq(2012, 2015, 3))
-  pisa_all2$value[exclude] <-
-    pmap(list(pisa_all2$value[exclude], escs_trend, years[exclude]), function(.x, .y, .z) {
-
-      # The escs data needs to have the key variables the same class as the
-      # same data.
-      escs <-
-        .y %>% mutate(schoolid = as.numeric(schoolid),
-                      stidstd = as.numeric(stidstd))
-
-      # .z is the corresponding year that will be created as a column
-      # And perform the same transformation of the key variables as in the ESCS data
-      data_trend <-
-        .x %>%
-        mutate(
-          year = .z,
-          schoolid = as.numeric(as.character(SCHOOLID)),
-          stidstd = as.numeric(as.character(STIDSTD))
-        ) %>%
-        left_join(escs,
-                  by = c("country", "schoolid", "stidstd"))
-
-      message(paste(unique(.x$wave), "done"))
-
-      data_trend
-    })
-
-  pisa_all2$value[[6]] <-
-    pisa_all2$value[[6]] %>%
-    rename(escs_trend = ESCS)
-@
-
-<<functions_for_modelling>>=
-
-# Function calculates the bottom 30th quantile for the bottom educated and the 70th quantile
-# for the top educated. If the quantiles can't be estimated, it returns two NA's instead
-quantile_missing <- function(df, weights, probs) {
-
-    quan <- try(Hmisc::wtd.quantile(
-      df$escs_trend,
-      weights = df[[weights]],
-      probs = probs
-      ))
-
-    if (any("try-error" %in% class(quan))) {
-      return(c(NA, NA))
-      } else {
-     return(c(quan[1], quan[2]))
-    }
-}
-
-# Producing the plot to get the difference between the top 30% of the high educated
-# vs the bottom 30% of the low educated. This function loops through each dataset/country
-# and survey reliability and estimates the difference while also extracting the s.e. of each
-# difference.
-
-# It returns a dataframe for each survey with all countries and respective coefficients and
-# standard errors.
-test_diff <- function(df, reliability, test, probs) {
-
-    map2(df, reliability, function(.x, .y) {
-
-      conf <- if (unique(.x$wave) == "pisa2015") pisa2015_conf else pisa_conf
-      weights_var <- conf$variables$weightFinal
-
-      country_split <- split(.x, .x$country)
-
-      country_list <- map(country_split, function(country) {
-        print(unique(country$country))
-
-        quan <- quantile_missing(country, weights_var, probs)
-
-        # It's very important to create a variable that returns the number of observations of this dummy
-        # For each country. Possibly to weight by the number of observations.
-        country$escs_dummy <-
-          with(country, case_when(escs_trend >= quan[2] ~ 1,
-                                  escs_trend <= quan[1] ~ 0))
-        country
-      })
-
-      .x <-
-        enframe(country_list) %>%
-        unnest(value)
-
-      .x <-
-        .x %>%
-        dplyr::select(wave,
-                      matches(paste0("^PV.*", test, "$")),
-                      escs_dummy,
-                      country,
-                      one_of(weights_var),
-                      AGE)
-
-      message(paste(unique(.x$wave), "data ready"))
-
-      test_vars <- paste0("PV", seq_len(conf$parameters$PVreps), test)
-      .x[test_vars] <- map(.x[test_vars], ~ ifelse(.x == 9997, NA, .x))
-
-      # Calculate median math score of all PV's
-      .x$dv <- rowMedians(as.matrix(.x[test_vars]), na.rm = T)
-
-      # Should I estimate the model separately by country?
-      mod1 <- lm(dv ~ AGE,
-                 weights = .x[[weights_var]],
-                 data = .x,
-                 na.action = "na.exclude")
-
-      # Take residuals of model and divide by rmse. Multiply that by
-      # 1 / sqrt(reliability of each survey), which is .y in the loop.
-      .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)
-
-      mod2 <-
-        lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
-             data = .x,
-             weights = .x[[weights_var]])
-
-      # Take the country coefficients (absolute coefficients)
-      country_coef <-
-        coef(mod2)$country %>%
-        rownames_to_column() %>%
-        gather(escs_dummy, Mean, -rowname) %>%
-        mutate(escs_dummy = dplyr::recode(escs_dummy,
-                                          `(Intercept)` = "0",
-                                          `escs_dummy` = "1"))
-
-      # Take the absolute country standard errors
-      se <-
-        se.coef(mod2)$country %>%
-        as.data.frame() %>%
-        rownames_to_column() %>%
-        gather(escs_dummy, s.e., -rowname) %>%
-        mutate(escs_dummy = dplyr::recode(escs_dummy,
-                                          `(Intercept)` = "0",
-                                          `escs_dummy` = "1"))
-
-      results <-
-        inner_join(country_coef, se, by = c("rowname", "escs_dummy")) %>%
-        rename(country = rowname) %>%
-        arrange(country, escs_dummy)
-
-      message(paste0(unique(.x$wave), " modeling done"))
-      results
-    })
-}
-
-# Adapted from: https://github.com/jtleek/slipper/blob/master/R/slipper.R
-# Returns a tibble with the actual expr + the bootstrapped expr.
-bootstrapper <- function(df, expr, B = 100, n = nrow(df), replacement = TRUE) {
-  bootstrapper_(df, lazyeval::lazy(expr), B, n, replacement)
-}
-
-bootstrapper_ <- function(df, expr, B = 500, n = nrow(df), replacement = TRUE) {
-  obs_val = lazyeval::lazy_eval(expr, data = df)
-  boot_val = replicate(B, {
-    newdata = sample_n(df, n, replace = replacement)
-    lazyeval::lazy_eval(expr, data = newdata)
-  })
-  out = tibble(type = c("observed", "bootstrap"), 
-               value = c(obs_val, mean(boot_val, na.rm = T)))
-  return(out)
-}
-
-# For example
-# bootstrapper(mtcars, mean(mpg), B = 200)
-@
-
-% This chunk takes about 3-4 hours to run. Once I had them ready I saved them so you can read them
-% as rds. If you run the 'compile pdf' this chunk is already cached (DONT CHANGE ANYTHING FROM
-% THIS CHUNK, CAUSE OTHERWISE IT WILL RERUN FOR 3-4 HOURS). If you want to run the code without
-% compiling sweave, comment the test_diff expressions and read as rds.
-
-% Remember to uncomment when compiling sweave again, so the chunk remains intact.
-<<modeling>>=
-adapted_year_data <-
-    map(pisa_all2$value, ~ {
-      if (unique(.x$wave) == "pisa2000") {
-        # pisa2000 has a different coding so here I recode 6 to 7 so that in all waves the top edu
-        # is 7 and the bottom is 1
-        .x <-
-          mutate(.x, new_hisced = as.character(dplyr::recode(as.numeric(high_edu_broad), `6` = 7)))
-      } else {
-        .x <-
-          mutate(.x, new_hisced = as.character(high_edu_broad))
-      }
-      .x
-})
-
-results_math <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.9))
-results_read <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.9))
-results_math_topmid <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.5, 0.9))
-results_read_topmid <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.5, 0.9))
-results_math_midbottom <- test_diff(adapted_year_data, reliability_pisa, "MATH", c(0.1, 0.5))
-results_read_midbottom <- test_diff(adapted_year_data, reliability_pisa, "READ", c(0.1, 0.5))
-
-# results_math <- read_rds("./data/delete.Rdata")
-# results_read <- read_rds("./data/delete_read.Rdata")
-# results_math_topmid <- read_rds("./data/delete_math_topmid.Rdata")
-# results_read_topmid <- read_rds("./data/delete_read_topmid.Rdata")
-# results_math_midbottom <- read_rds("./data/delete_math_midbottom.Rdata")
-# results_read_midbottom <- read_rds("./data/delete_read_midbottom.Rdata")
-# US is missing for reading
-
-# Cache is not working properly for the code above, so I just load the saved cached file
-# load("./paper/cache/modeling_9a0b38d1d53fa243b0242580f0672fa5.RData")
-@
-
-\section{Introduction}
-
-The impact of inequality, be it economic or of non pecuniary causes, is a crucial determinant for the opportunities of a child.
-Evidence from disciplines as disparate as epidemiology, psychology, sociology and economics keep piling evidence that suggests that inequality in the life course is something present in virtually all societies.
-This is something that politicians and policymakers are aware of and, yet despite their many efforts, inequality has not abated.
-
-The literature has found conclusive evidence showing that, for example, the United States has increased both economic inequality \citep{piketty2003}, inequality of opportunity \citep{chetty2014} and inequality of achievement \citep{reardon2011}.
-All of these findings, however, are from the United States.
-Other countries such as the United Kingdom or other developed countries have their fair share of studies when there is data available.
-However, very rarely can we compare levels of inequality (of any kind), across several countries with a great degree of confidence.
-
-This paper looks to study the cognitive achievement gap between the 90th and 10th percentile of the SES distribution.
-This gap can be termed as the SES achievement gap between the top and bottom SES groups.
-Using the PISA database, we calculate the achievement gap for every country and for all six available PISA waves, building a country panel that shows a time trend of 15 years for each country.
-
-More concretely, we aim to study the evolution of the achievement gap.
-We aim to answer two broad question: Which countries are changing over time while others are stagnated?
-Equally important, what explains these changes?
-Building on the literature on educational institutions we suggest the degree of tracking of a country and it's level of vocational enrollments as possible explanations for the changes of a country over time.
-
-Findings suggest that there's reasonable variation in the evolution of the gap.
-For example, the United States has been closing its 90/10 gap by about 50\% with a wide interval going down to nearly 35\%.
-This reduction is also present in other countries such as Germany and Poland.
-Other countries such as France and Austria experience the contrary with persistent increases in the achievement gap.
-Robustness checks shows that also the 80/20 and 70/30 replicate the results for nearly all countries.
-
-We then assess whether this gap is correlated with the country's average performance, and find a very strong negative correlation for all country pooled.
-This suggests that countries with high levels of achievement have low levels of inequality.
-Yet these results are aggregated for every year.
-A more disaggregate analysis shows that this pattern is also present for every country over time, although some exceptions are encountered, like the United States where achievement and inequality as positively correlated.
-
-To finish, we find that the degree of tracking of a country seems to be a likely explanation to the changes in the achievement gap, but it should not be studied without factoring in the degree of vocational enrollment.
-We find that if a country has a low level of tracking, vocational enrollment will not help decrease the achievement gap.
-Conversely, once tracking enters the picture, vocational enrollment can help unease the burden of inequality and reduce the achievement gap very strongly.
-Simulations show that if every country switched to little or no tracking their gaps would shrink by about 20\%.
-
-These findings have two specific implications.
-First, inequality should not be 'generalized' or 'summarized' with one single indicator for many countries.
-There are widespread differences between countries, where each one should be studied in context.
-Secondly, now that we know that, further research should attempt to explain why some countries follow such different evolutions.
-We propose a possible explanation, the degree of tracking and vocational enrollment of the country.
-Tracking seems to exacerbate inequality, but if coped with a reasonable level of vocational enrollment it can alleviate the negative effects of tracking.
-Further research should confirm these results hold for other empirical tests and if so, policymakers should incorporate vocational enrollment not only as a 'labor market' solution to youth unemployment but as a compensatory mechanism for highly tracked curriculums.
-
-\section{Literature Review}
-Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years.
-The idea of meritocracy and intergenerational transfers has motivated, for a good part of the 20th and 21st century, much of the research on social mobility and social inequalities.
-When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family socio-economic status (SES) and student performance are tightly linked.
-The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements \footnote{For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.}.
-But fast-forward until today, we've understood the relationship much better.
-
-Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive development and defining personality traits \citep{duyme1999, waldfogel2006}.
-Economists and educators have long neglected these findings until the work of James Heckman brought our attention to it.
-For example, he has showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
-To explain these, and other findings, \citet{cunha2006} suggest that the cognitive level of a child in time \(t\) is a direct function of the experiences of time \(t-1\).
-Although this sounds logical, it's importance is not understood entirely.
-The model implies that investments on children are not perfect substitutes at different time points but rather complements.
-There are specific periods of skill formation where it is much more cost-effective to invest in, which are the early stages of the life course.
-In fact, when tested against the data, \citet{cunha2006} show that their explanation fits the results of most of the empirical literature very well.
-
-For this reason, children who receive less (or worse) investments have a cumulative disadvantage against children who do receive quality investments.
-This leads, as early as preschool, to an unequal start in life.
-That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is cheaper than investing at later stages and it brings about the largest returns \citep{elango15}.
-Despite all of these findings and strategies to reduce the gaps, we still find that the relationship between children's parental education and future destination is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.
-
-For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital.
-Social mobility has deepen our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity.
-The field of economics of education has also helped to understand that early high-quality education is by far the best investment in terms of cognitive development and acquiring personality skills.
-Research has constantly found that early education helps low-income children boost their performance to levels similar to their well-off peers \citep{heckman_oecd}.
-Yet the positive results are not only limited to the cognitive realm: studies have found that that early stimuli is associated with less propensity to criminal activity, more stable relationships, higher levels of persistence, higher income and improved labor market opportunities in late stages in life \citep{heckman_oecd}.
-
-This is not only for the child and his/her family, but to society as a whole as it boosts economic activity, it helps the labor market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
-Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about.
-Naturally, we want to do that in order to reverse it and help every child reach its fullest potential.
-Despite our good intentions, we've concentrated little on the magnitude of the gaps, specially in terms of cognitive abilities.
-We have very little comparative evidence on which countries have big or small cognitive gaps resulting from SES origins.
-And even more importantly, we haven't assessed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.
-
-Despite these short-comings, we have indirect evidence of this coming from the vast literature on social mobility.
-Thanks to these findings, we know that virtually in all countries, be it developed or developing, there is inequality of opportunity.
-But there is considerable variation in the magnitude of this effect.
-For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries \citep{esping2012, breen2007, shavit1993}.
-
-From Denmark, for example, the literature has learned a great deal about how to improve social mobility.
-But that was possible by first learning that Denmark is by far one of the most mobile countries compared to other European countries; this is the case specially for families of low-ses origin \citep{bjorklund2009, jaeger2007}.
-The important finding came when research emerged the main reason behind their social escalator: it's educational system.
-For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impact on improving opportunities.
-The education system is completely subsidized for all children, otherwise giving opportunities to families who would not have been able to pay.
-In addition to this, Denmark is recognized as a world-leader in terms of public support for its child-care system as it spends around 2\% of GDP, and has among the highest enrollment rates for children under 6 years old \citep{esping_waldfogel2012}.
-Moreover, it separates students into different tracks very late compared to other European countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}.
-These two traits make the Danish system effective in promoting equality relative to the average European country.
-The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable \citep{heckman_oecd}.
-In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
-
-The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years.
-The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape.
-The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940.
-This time-trend has allowed researchers to study achievement gaps in a lengthy period of time.
-Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
-Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}.
-The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
-% widening over the years? Gosta, page 3
-% income achievement gap, gosta page 3
-
-\citet{reardon2011} finds that the increase in the gaps has occurred predominantly from the 1970's until the 2000's.
-In fact, the hard numbers suggest that the gap widened by 40-50\%.
-The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of   75\%.
-Given that the studies before 1970 are less reliable in terms of comparability and sampling design, the author computes all results for both before/after 1970.
-To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the 90/10 income gap in the U.S is at about 1.25 standard deviations in the year 2001.
-Using longitudinal data, \citet{bradbury2015} find similar results.
-Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25.
-Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of  1.25-1.50 standard deviations.
-One important drawback of these studies is that they don't present the uncertainty of these estimates.
-Not necessarily to gauge their statistical significance, but to simply assess how much we can trust their accuracy.
-It could be that the gap is at 1.25 standard deviations, varying up to 1.75 and down to 0.80.
-Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.
-
-Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation.
-\citet{reardon2011} offers several possible links, with the most reasonable being that family investment patterns have changed so that high income families now invest more resources on their children.
-This explanation lies in the fact that increasing income became more strongly correlated with other positive family traits related to time allocation and welfare services.
-
-In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend.
-This follow-up study concentrated solely on kindergarten children for the years 1998, 2006 and 2010.
-They found that the 90th/10th income gap in readiness closed modestly.
-Furthermore, using data from fall and spring in the same kindergarten year, they calculated that the gap narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010.
-They also calculated the same changes for a number of personality traits such as self-control and externalizing behavior and found similar results.
-In contrast, \citet{reardon2011} finds that in a 30-year span the gap was systematically increasing at a rate of 0.02, something reasonably close to the previous estimates.
-Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap).
-It should be noted that perhaps the reversal of the trend in \citet{reardon2011} would had been evident if data were available for years after 2000, the time-point from which \citet{reardon_portilla} find the reversal.
-
-The reasons why the authors find a reversal in the trend could be numerous and should be studied very closely.
-They incorporate a number of country-level indicators to explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment.
-They build on their previous argument by suggesting that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
-Their conclusions, although suggestive, are very speculative and have no empirical support, which is why this is still an open question.
-
-Motivated by these recent results, other authors have taken this analysis to an international context in order to discover between-country trends.
-The work of \citet{bradbury2015} employs a unique comparative analysis of the achievement gap between Australia, United Kingdom, United States and Canada.
-Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is not only the first study that uses panel data to study achievement gaps, but to also do it between countries}.
-The core finding behind their study is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada.
-They find that once the achievement gap is present in early school entry, it does not seem to narrow or widen very much over the life course.
-In fact, they calculate that the quality of early education can only explain about  30-40\% of the high school SES gap.
-This suggests that once the achievement gap is present before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous Randomized Controlled Trials (RCT) show that high quality schooling can indeed help ease the gap, in some instances even close it \citep{campbell2002}.}.
-One exception is the United Kingdom, which they found to be a country that helps close the gap in early primary years.
-This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health and income support.
-
-One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula.
-These four countries have no major jump in school selection, quite the opposite to the average European country.
-A thorough review by \citet{werfhorst_mijs} sheds some light on the subject.
-First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps.
-\citet{hanushek_woesmann_tracking}, use the Progress in International Reading Literacy Study (PIRLS), the Trends in International Mathematics and Science Study (TIMSS) and the Programme for International Student Assessment (PISA) surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
-The results suggest that early tracking increases educational inequality.
-While less clear, there is also a tendency for early tracking to reduce mean performance.
-\citet{micklewright} using PISA but a different empirical strategy, find that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
-In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
-
-Another limitation of the work of \citet{bradbury2015} is that their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
-They do an amazing work at making the four surveys comparable and their findings are indeed very reliable.
-But we should be careful at interpreting these findings causally and should be taken as suggestive at best.
-For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citet{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in the gap using much more comparable surveys.
-However it should be noted that these studies tackle a completely different question from the above, namely to study cross-sectional differences between many countries, instead of over-time analysis of student gaps.
-Nonetheless, they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
-
-\citet{anna2016}, again using PIRLS, TIMSS and PISA, assess whether there are patterns of cross-national variation in the achievement gap.
-In other words, does the achievement gap differ between countries?
-Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
-In comparison to the literature on achievement gaps, they find that the U.S has a gap of 1.20 SD in 2001 which increase to around 1.30 in the year 2006 while Germany has a decreasing gap from around 1.25 to 1 standard deviation in the same year-span.
-However, these numbers vary a lot and carry a great deal of overlapping uncertainty.
-
-They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation and central exams, among other things.
-The correlations are very suggestive  as explanatory mechanisms but they are very cautious in drawing causality, specially because they are all reasonably correlated among each other.
-
-One interesting question that is still missing from the literature is whether these country gaps have evolved over time.
-With their data, they only have 3 countries which are present in all waves but they also have very few waves because their question of interest (income categories) was only asked in three time points.
-That is why their study is more about between country gaps rather than the magnitude and evolution of the gaps.
-
-\citet{anna2016_global}, building on the work of \citet{anna2016} and \citet{reardon_portilla} pooled together all the previously mentioned data, together with over 10 more studies ranging from 1964 to 2015 in order to discover differences between and across countries.
-With over 50 years of data, and over 100 countries, she finds that there seems to be a general pattern of increasing achievement gap.
-However, once she disentangles the relationship by country, she finds a sizable amount of heterogeneity, with some countries seeing the achievement gap narrowing, others no change at all, while others record a steady increase.
-This is revealing because it does not really pay off to look at a general average once each country has their own distinctive gap and evolution.
-This highlights the notion that the increasing achievement gap is clearly not universal and should be studied in context.
-
-One limitation of their study (as well as \citet{reardon2011}) is that they adjust for the age of each child in all studies.
-Although for their modeling purposes this is the correct thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
-However, as we've noted before, \citet{bradbury2015} find that the achievement gap is very stable across the life time}, these modeling strategies are masking age-specific achievement gaps by controlling for age.
-We clearly see in \citet{reardon_portilla} that there are age-specific gaps, and they do change at a fast pace in very little time.
-
-In fact, the evolution of high/low SES gaps for preschool children might be much less marked than the same gap for high school children.
-The explanation, although very debated, has been gaining much support in recent years.
-For countries with high levels of curricular differentiation the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}.
-Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}.
-Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts nor the same between ages, because tracking/no tracking might exacerbate the achievement gap.
-
 % TODO: Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)

### 2018-01-23 paper/paper.tex
Last commit before delete local folder - everything moved to thesis/ repo

@@
-% Things to remember:
-% The working directory is not where the data is.
-% Change working directory to "/Users/cimentadaj/Downloads/gitrepo/Achievement_Gap_PISA/paper"
-% for everything below to work.
-
-% There's a code chunk named modeling which lasts about 3-4 hours to run. I explain how to get
-% around that in a comment above the code chunk. For down there.
-
-\documentclass[11pt, a4paper]{article}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-\let\hlipl\hlkwb
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-\bibliographystyle{apalike}
-\pagestyle{headings}
-
-\usepackage{geometry}
- \geometry{
- a4paper,
- total={170mm,257mm},
- left=20mm,
- top=20mm,
- }
-\usepackage{pdflscape}
-\usepackage{amsmath}
-\usepackage{subcaption}
-\usepackage{graphics}
-\usepackage{graphicx}
-\usepackage{titling}
-\usepackage[round, colon]{natbib}
-\usepackage[colorlinks]{hyperref}
-\usepackage{booktabs}
-\AtBeginDocument{%
-  \hypersetup{
-    citecolor=blue,
-    linkcolor=blue,   
-    urlcolor=blue}}
-
-\title{Do educational institutions explain SES gaps? an international comparison of the SES achievement gap from 2000-2015}
-\author{Jorge Cimentada}
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\begin{document}
-\setlength{\parindent}{2em}
-\setlength{\parskip}{1em}
-\showboxdepth=5
-\showboxbreadth=5
-
-\maketitle
-
-\begin{abstract}
-The literature on achievement inequality has only recently started studying the evolution of the socio-economic achievement gap in cognitive abilities and it has uncovered a wide variety of results.
-The main finding comes from American research and it shows that the gap has been widening by about 40-50\% in the last 30 year.
-Other international comparisons show that there's important variation among developed countries, but the research on this area is scarce.
-This paper looks to study whether there are specific patterns in the evolution of the gap in a comparative perspective.
-We find that there is considerable variation in the way in which the gap is evolving, with the U.S and Germany closing at a rapid pace and France widening at a similar rate.
-We find that this gap is strongly correlated with a country's overall performance and simulations show that switching to less tracking can help decrease a countries SES gap by about 20\%.
-We also find that tracking and vocational enrollment explain over one third of the evolution of the achievement gap and show that the relationship is conditioned by a strong interaction: low tracking is associated with a small achievement gap, but high levels of tracking coped with high vocational enrollment can also remedy the aversive effects of tracking.
-\end{abstract}
-
-% pisa_listcol.Rdata is a dataset produced by the source() expression from the above
-% code chunk
-
-% This chunk takes about 3-4 hours to run. Once I had them ready I saved them so you can read them
-% as rds. If you run the 'compile pdf' this chunk is already cached (DONT CHANGE ANYTHING FROM
-% THIS CHUNK, CAUSE OTHERWISE IT WILL RERUN FOR 3-4 HOURS). If you want to run the code without
-% compiling sweave, comment the test_diff expressions and read as rds.
-
-% Remember to uncomment when compiling sweave again, so the chunk remains intact.
-
-\section{Introduction}
-
-The impact of inequality, be it economic or of non pecuniary causes, is a crucial determinant for the opportunities of a child.
-Evidence from disciplines as disparate as epidemiology, psychology, sociology and economics keep piling evidence that suggests that inequality in the life course is something present in virtually all societies.
-This is something that politicians and policymakers are aware of and, yet despite their many efforts, inequality has not abated.
-
-The literature has found conclusive evidence showing that, for example, the United States has increased both economic inequality \citep{piketty2003}, inequality of opportunity \citep{chetty2014} and inequality of achievement \citep{reardon2011}.
-All of these findings, however, are from the United States.
-Other countries such as the United Kingdom or other developed countries have their fair share of studies when there is data available.
-However, very rarely can we compare levels of inequality (of any kind), across several countries with a great degree of confidence.
-
-This paper looks to study the cognitive achievement gap between the 90th and 10th percentile of the SES distribution.
-This gap can be termed as the SES achievement gap between the top and bottom SES groups.
-Using the PISA database, we calculate the achievement gap for every country and for all six available PISA waves, building a country panel that shows a time trend of 15 years for each country.
-
-More concretely, we aim to study the evolution of the achievement gap.
-We aim to answer two broad question: Which countries are changing over time while others are stagnated?
-Equally important, what explains these changes?
-Building on the literature on educational institutions we suggest the degree of tracking of a country and it's level of vocational enrollments as possible explanations for the changes of a country over time.
-
-Findings suggest that there's reasonable variation in the evolution of the gap.
-For example, the United States has been closing its 90/10 gap by about 50\% with a wide interval going down to nearly 35\%.
-This reduction is also present in other countries such as Germany and Poland.
-Other countries such as France and Austria experience the contrary with persistent increases in the achievement gap.
-Robustness checks shows that also the 80/20 and 70/30 replicate the results for nearly all countries.
-
-We then assess whether this gap is correlated with the country's average performance, and find a very strong negative correlation for all country pooled.
-This suggests that countries with high levels of achievement have low levels of inequality.
-Yet these results are aggregated for every year.
-A more disaggregate analysis shows that this pattern is also present for every country over time, although some exceptions are encountered, like the United States where achievement and inequality as positively correlated.
-
-To finish, we find that the degree of tracking of a country seems to be a likely explanation to the changes in the achievement gap, but it should not be studied without factoring in the degree of vocational enrollment.
-We find that if a country has a low level of tracking, vocational enrollment will not help decrease the achievement gap.
-Conversely, once tracking enters the picture, vocational enrollment can help unease the burden of inequality and reduce the achievement gap very strongly.
-Simulations show that if every country switched to little or no tracking their gaps would shrink by about 20\%.
-
-These findings have two specific implications.
-First, inequality should not be 'generalized' or 'summarized' with one single indicator for many countries.
-There are widespread differences between countries, where each one should be studied in context.
-Secondly, now that we know that, further research should attempt to explain why some countries follow such different evolutions.
-We propose a possible explanation, the degree of tracking and vocational enrollment of the country.
-Tracking seems to exacerbate inequality, but if coped with a reasonable level of vocational enrollment it can alleviate the negative effects of tracking.
-Further research should confirm these results hold for other empirical tests and if so, policymakers should incorporate vocational enrollment not only as a 'labor market' solution to youth unemployment but as a compensatory mechanism for highly tracked curriculums.
-
-\section{Literature Review}
-Educational inequality and its long term impacts is a topic that has been prominent in the social science literature for the past 30 years.
-The idea of meritocracy and intergenerational transfers has motivated, for a good part of the 20th and 21st century, much of the research on social mobility and social inequalities.
-When James Coleman released his famous Coleman report \citep{coleman1966}, it raised the topic to starlight by suggesting that family socio-economic status (SES) and student performance are tightly linked.
-The topic has been studied extensively since the report was released and several authors have contested whether the relationship is an invariable social law or a product of institutional arrangements \footnote{For a detailed review of the long literature on educational inequality, please see \citet{gamoran2001}.}.
-But fast-forward until today, we've understood the relationship much better.
-
-Developmental psychology has been studying the subject for a long time and they find that the early stages in a child's life course are extremely important, if not the most, for cognitive development and defining personality traits \citep{duyme1999, waldfogel2006}.
-Economists and educators have long neglected these findings until the work of James Heckman brought our attention to it.
-For example, he has showed that cognitive and non-cognitive inequalities are present before entering school \citep{heckman2006}.
-To explain these, and other findings, \citet{cunha2006} suggest that the cognitive level of a child in time \(t\) is a direct function of the experiences of time \(t-1\).
-Although this sounds logical, it's importance is not understood entirely.
-The model implies that investments on children are not perfect substitutes at different time points but rather complements.
-There are specific periods of skill formation where it is much more cost-effective to invest in, which are the early stages of the life course.
-In fact, when tested against the data, \citet{cunha2006} show that their explanation fits the results of most of the empirical literature very well.
-
-For this reason, children who receive less (or worse) investments have a cumulative disadvantage against children who do receive quality investments.
-This leads, as early as preschool, to an unequal start in life.
-That is why the whole debate about narrowing inequality between high and low SES families should start in early education -- it is cheaper than investing at later stages and it brings about the largest returns \citep{elango15}.
-Despite all of these findings and strategies to reduce the gaps, we still find that the relationship between children's parental education and future destination is present in virtually all empirical studies of social mobility and inequalities \citep{breen1997, breen2007, waldfogel2006, bradbury2015, chetty2016}.
-
-For this reason, it is incomplete to study educational inequalities without considering the prevalence of intergenerational transmission of human and social capital.
-Social mobility has deepen our knowledge on the relationship between intergenerational transmission of SES and inequality of opportunity.
-The field of economics of education has also helped to understand that early high-quality education is by far the best investment in terms of cognitive development and acquiring personality skills.
-Research has constantly found that early education helps low-income children boost their performance to levels similar to their well-off peers \citep{heckman_oecd}.
-Yet the positive results are not only limited to the cognitive realm: studies have found that that early stimuli is associated with less propensity to criminal activity, more stable relationships, higher levels of persistence, higher income and improved labor market opportunities in late stages in life \citep{heckman_oecd}.
-
-This is not only for the child and his/her family, but to society as a whole as it boosts economic activity, it helps the labor market improve job conditions and maintain a rapid economic growth \citep{hanushek2007}.
-Since \citet{coleman1966} we've spent most of our time studying the mechanisms through which this inequality comes about.
-Naturally, we want to do that in order to reverse it and help every child reach its fullest potential.
-Despite our good intentions, we've concentrated little on the magnitude of the gaps, specially in terms of cognitive abilities.
-We have very little comparative evidence on which countries have big or small cognitive gaps resulting from SES origins.
-And even more importantly, we haven't assessed whether policy efforts to reduce inequality have actually had an impact in reducing the achievement gap over time.
-
-Despite these short-comings, we have indirect evidence of this coming from the vast literature on social mobility.
-Thanks to these findings, we know that virtually in all countries, be it developed or developing, there is inequality of opportunity.
-But there is considerable variation in the magnitude of this effect.
-For example, the Scandinavian countries, particularly Denmark and Sweden, prove to be very mobile countries \citep{esping2012, breen2007, shavit1993}.
-
-From Denmark, for example, the literature has learned a great deal about how to improve social mobility.
-But that was possible by first learning that Denmark is by far one of the most mobile countries compared to other European countries; this is the case specially for families of low-ses origin \citep{bjorklund2009, jaeger2007}.
-The important finding came when research emerged the main reason behind their social escalator: it's educational system.
-For example, research by \citet{esping_waldfogel2012} and \citet{bauchmuller2014}, show that the Danish early education system has important and longstanding impact on improving opportunities.
-The education system is completely subsidized for all children, otherwise giving opportunities to families who would not have been able to pay.
-In addition to this, Denmark is recognized as a world-leader in terms of public support for its child-care system as it spends around 2\% of GDP, and has among the highest enrollment rates for children under 6 years old \citep{esping_waldfogel2012}.
-Moreover, it separates students into different tracks very late compared to other European countries (age 16), something which has been linked to less educational inequality \citep{hanushek_woesmann_tracking}.
-These two traits make the Danish system effective in promoting equality relative to the average European country.
-The early schooling experience attempts to level children to the same degree and this process is not stopped by different curricular tracks because it starts at around age 16 when cognitive abilities are less malleable \citep{heckman_oecd}.
-In short, the importance of this finding, is that we should step back and first study whether the effect is there, how big is it, and then proceed to find the causes behind it.
-
-The first attempt to study the evolution of the achievement gap has found that the gap in cognitive abilities between High-SES and Low-SES kids has been widening over the years.
-The literature on the topic has mainly concentrated on studying the American case \citep{reardon2011} but other international evidence is emerging with a very similar landscape.
-The United States is usually the case of study because it is the only country where cognitive testing is present as early as 1940.
-This time-trend has allowed researchers to study achievement gaps in a lengthy period of time.
-Using this information, \citet{reardon2011} is the first to investigate the evolution of the cognitive gap and the results are very surprising.
-Not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black gap \citep{magnuson2008}.
-The gaps have actually reversed and now a days we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
-% widening over the years? Gosta, page 3
-% income achievement gap, gosta page 3
-
-\citet{reardon2011} finds that the increase in the gaps has occurred predominantly from the 1970's until the 2000's.
-In fact, the hard numbers suggest that the gap widened by 40-50\%.
-The author also estimates the rate of change using data as early as 1940 and finds an even higher increase of   75\%.
-Given that the studies before 1970 are less reliable in terms of comparability and sampling design, the author computes all results for both before/after 1970.
-To provide a definitive answer to the size of the gap, \citet{reardon2011} concludes that the 90/10 income gap in the U.S is at about 1.25 standard deviations in the year 2001.
-Using longitudinal data, \citet{bradbury2015} find similar results.
-Their empirical analysis suggests that for American 14 year olds, the gap is above 1 standard deviation but lower than 1.25.
-Surprisingly, \citet{duncan2011} find similarly to the previous studies and confirm a gap of  1.25-1.50 standard deviations.
-One important drawback of these studies is that they don't present the uncertainty of these estimates.
-Not necessarily to gauge their statistical significance, but to simply assess how much we can trust their accuracy.
-It could be that the gap is at 1.25 standard deviations, varying up to 1.75 and down to 0.80.
-Right now we don't know the upper/lower bound of this estimate, making it difficult to compare when new evidence arises.
-
-Interestingly, this widening of the achievement gap has been paralleled by a growth of income inequality, a very suggestive explanation.
-\citet{reardon2011} offers several possible links, with the most reasonable being that family investment patterns have changed so that high income families now invest more resources on their children.
-This explanation lies in the fact that increasing income became more strongly correlated with other positive family traits related to time allocation and welfare services.
-
-In a follow-up study, \citet{reardon_portilla} unexpectedly uncovered a new finding: the reversal of the trend.
-This follow-up study concentrated solely on kindergarten children for the years 1998, 2006 and 2010.
-They found that the 90th/10th income gap in readiness closed modestly.
-Furthermore, using data from fall and spring in the same kindergarten year, they calculated that the gap narrowed at a rate of 0.01 and 0.008 standard deviations per year for mathematics and literacy between 1998 and 2010.
-They also calculated the same changes for a number of personality traits such as self-control and externalizing behavior and found similar results.
-In contrast, \citet{reardon2011} finds that in a 30-year span the gap was systematically increasing at a rate of 0.02, something reasonably close to the previous estimates.
-Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap).
-It should be noted that perhaps the reversal of the trend in \citet{reardon2011} would had been evident if data were available for years after 2000, the time-point from which \citet{reardon_portilla} find the reversal.
-
-The reasons why the authors find a reversal in the trend could be numerous and should be studied very closely.
-They incorporate a number of country-level indicators to explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment.
-They build on their previous argument by suggesting that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
-Their conclusions, although suggestive, are very speculative and have no empirical support, which is why this is still an open question.
-
-Motivated by these recent results, other authors have taken this analysis to an international context in order to discover between-country trends.
-The work of \citet{bradbury2015} employs a unique comparative analysis of the achievement gap between Australia, United Kingdom, United States and Canada.
-Their research design is very distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is not only the first study that uses panel data to study achievement gaps, but to also do it between countries}.
-The core finding behind their study is that the American achievement gap is much wider than in any other comparison country, specially Australia and Canada.
-They find that once the achievement gap is present in early school entry, it does not seem to narrow or widen very much over the life course.
-In fact, they calculate that the quality of early education can only explain about  30-40\% of the high school SES gap.
-This suggests that once the achievement gap is present before entering school, it carries a social-scar effect \footnote{However, schooling could be preventing the gap to widen even more, and very rigorous Randomized Controlled Trials (RCT) show that high quality schooling can indeed help ease the gap, in some instances even close it \citep{campbell2002}.}.
-One exception is the United Kingdom, which they found to be a country that helps close the gap in early primary years.
-This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health and income support.
-
-One limitation of their study is that they concentrate on countries which have a very particular educational structure, namely the fact that there is little formal stratification in terms of curricula.
-These four countries have no major jump in school selection, quite the opposite to the average European country.
-A thorough review by \citet{werfhorst_mijs} sheds some light on the subject.
-First and foremost, they gather substantive evidence showing that countries which have a highly tracked curriculum tend to have high levels of inequality, measured in terms of achievement gaps.
-\citet{hanushek_woesmann_tracking}, use the Progress in International Reading Literacy Study (PIRLS), the Trends in International Mathematics and Science Study (TIMSS) and the Programme for International Student Assessment (PISA) surveys to gauge whether highly tracked countries do indeed increase inequality after students pass the age at first selection of tracking.
-The results suggest that early tracking increases educational inequality.
-While less clear, there is also a tendency for early tracking to reduce mean performance.
-\citet{micklewright} using PISA but a different empirical strategy, find that countries which have a high level of tracking, are distinctively unequal in the difference between the top 95th and bottom 5th performers.
-In fact, the difference in test scores between these two groups is about 10 times higher then the average annual gain of a year of schooling.
-
-Another limitation of the work of \citet{bradbury2015} is that their analysis is based on four surveys that have significant differences in terms of questions, sampling and populations and cannot be easily compared.
-They do an amazing work at making the four surveys comparable and their findings are indeed very reliable.
-But we should be careful at interpreting these findings causally and should be taken as suggestive at best.
-For this reason we should also pay particular attention to studies such as \citet{anna2016} and \citet{anna2016_global} which have attempted not only to compare gaps between countries, but to evaluate whether there is a general increase in the gap using much more comparable surveys.
-However it should be noted that these studies tackle a completely different question from the above, namely to study cross-sectional differences between many countries, instead of over-time analysis of student gaps.
-Nonetheless, they do provide support for the overall finding that the achievement gap is certainly not narrowing over time.
-
-\citet{anna2016}, again using PIRLS, TIMSS and PISA, assess whether there are patterns of cross-national variation in the achievement gap.
-In other words, does the achievement gap differ between countries?
-Their work suggests that there is considerable variation in the achievement gap between top and bottom earning families across many developed countries.
-In comparison to the literature on achievement gaps, they find that the U.S has a gap of 1.20 SD in 2001 which increase to around 1.30 in the year 2006 while Germany has a decreasing gap from around 1.25 to 1 standard deviation in the same year-span.
-However, these numbers vary a lot and carry a great deal of overlapping uncertainty.
-
-They go even further and link this achievement gap to several country-level indicators related to income inequality, school differentiation and central exams, among other things.
-The correlations are very suggestive  as explanatory mechanisms but they are very cautious in drawing causality, specially because they are all reasonably correlated among each other.
-
-One interesting question that is still missing from the literature is whether these country gaps have evolved over time.
-With their data, they only have 3 countries which are present in all waves but they also have very few waves because their question of interest (income categories) was only asked in three time points.
-That is why their study is more about between country gaps rather than the magnitude and evolution of the gaps.
-
-\citet{anna2016_global}, building on the work of \citet{anna2016} and \citet{reardon_portilla} pooled together all the previously mentioned data, together with over 10 more studies ranging from 1964 to 2015 in order to discover differences between and across countries.
-With over 50 years of data, and over 100 countries, she finds that there seems to be a general pattern of increasing achievement gap.
-However, once she disentangles the relationship by country, she finds a sizable amount of heterogeneity, with some countries seeing the achievement gap narrowing, others no change at all, while others record a steady increase.
-This is revealing because it does not really pay off to look at a general average once each country has their own distinctive gap and evolution.
-This highlights the notion that the increasing achievement gap is clearly not universal and should be studied in context.
-
-One limitation of their study (as well as \citet{reardon2011}) is that they adjust for the age of each child in all studies.
-Although for their modeling purposes this is the correct thing to do \footnote{The differences in achievement could simply be due to changes in cognitive abilities across the lifetime.
-However, as we've noted before, \citet{bradbury2015} find that the achievement gap is very stable across the life time}, these modeling strategies are masking age-specific achievement gaps by controlling for age.
-We clearly see in \citet{reardon_portilla} that there are age-specific gaps, and they do change at a fast pace in very little time.
-
-In fact, the evolution of high/low SES gaps for preschool children might be much less marked than the same gap for high school children.
-The explanation, although very debated, has been gaining much support in recent years.
-For countries with high levels of curricular differentiation the transition from early schooling into the tracking system has been found to increase inequality of learning \citep{hanushek_woesmann_tracking}.
-Moreover, the vast sociological literature on educational transitions systematically finds that tracking tends to foster between-track inequality rather than erode their differences by tackling their specific needs \citep{werfhorst_mijs}.
-Based on this, we cannot simply assume that the achievement gap has been neither constant across cohorts nor the same between ages, because tracking/no tracking might exacerbate the achievement gap.
-
 % TODO: Talk about how the achievement gap has been widening in Malaysia and South Korea (or Japan?)
@@
 
-To investigate the above mentioned questions we will be using the Programme for International Student Assessment (PISA).
-PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students.
-Currently, PISA has six waves starting in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 developed/developing countries.
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
+\begin{alltt}
+\hlstd{country_rows} \hlkwb{<-}
+  \hlkwd{map_dbl}\hlstd{(adapted_year_data, nrow)} \hlopt{%>%}
+  \hlkwd{format}\hlstd{(}\hlkwc{big.mark} \hlstd{=} \hlstr{","}\hlstd{)}
+\end{alltt}
 
-PISA collects data through a two-stage stratified sampling design.
-With the help of official governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests.
-The sample size for PISA 2000 is 127,388, 276,165 for PISA 2003, 398,750 for PISA 2006, 515,958 for PISA 2009, 480,810 for PISA 2012, and 519,334 for PISA 2015.
-Together with the subject tests, PISA collects personal information from students, their families and their school environment (including teacher surveys), that serves as relevant background information that can be matched to the students performance.
-With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system.
-None of the literature cited so far has used the last PISA wave, which was recently released in December 2016.
-This paper takes advantage of these six waves to build a country pseudo-panel where each country-wave will have its corresponding achievement gap, making it possible to study changes in nearly 15 years for each country.
-
-To identify a family's socio-economic status PISA collects several variables that measure different dimensions.
-Classically, they ask student's their parent's educational level.
-Scholars have considered this to be a reliable recall given that we expect fifteen year olds to know their parent's level of education \citep{reardon2011}.
-This question has been asked in every wave and holds a somewhat similar coding across time, although the first two waves have small differences.
-In spite of this, another limitation is the fact that parent's education is measured using the ISCED classification, something that has changed over time.
-For example, until PISA 2009, the preferred framework was ISCED 1997, whereas the next wave switched to the newly developed ISCED 2011 classification.
-Both these classification schemes have equivalent look-up tables, but this requires a detailed inspection of the codings.
-
-Another social background variable is the International Socio-Economic Index of Occupational Status (ISEI).
-This variable attempts to capture the social status of the family, without asking for income information.
-This index variable was developed by \citet{ganzeboom1996} and later refined by \citep{ganzeboom2010} and it attempts to measure occupational status using a continuous measure.
-This indicator is a very reliable alternative to the  classical Erikson-Goldthorpe-Portocarero classification \citep{erikson1979}.
-It has been scaled for comparability between waves and some authors have used it for inequality studies, finding expected results to be consistent with the social mobility literature \citep{anna2016_global}.
-PISA also includes a plethora of indicators on family wealth, home educational resources, the number of books in the home, among many other material resources in the household.
-
-Yet one of the most relevant variables for our study is a composite SES index created by the PISA team.
-The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measures the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) \citep{oecd_glance_2002}.
-This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming all mentioned variables into comparable metrics across waves.
-
-The ESCS was derived from a principal component analysis of standardized variables, taking the factor scores for the first principal component as measures of the PISA index of economic, social and cultural status.
-All countries and economies (both OECD and partner countries/economies) contributed equally to the principal component analysis, while in previous cycles, the principal component analysis was based on OECD countries only.
-However, for the purpose of reporting the ESCS scale has been transformed with zero being the score of an average OECD student and one being the standard deviation across equally weighted OECD countries \citep{pisa_2015_results}.
-
-To the best of our knowledge, this is the first paper that uses the newly-released ESCS index \citep{pisa_2015_results} which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indexes can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled indexes for Trend Analyses}.} In other words, the ESCS index does not need any transformation or coding updates because it is ready for comparison over time.
-
-Aside from SES, the other relevant variables are test scores for mathematics and literacy.
-PISA does not provide a single test result for each respondent.
-Instead, it provides a \emph{series} of 'plausible values' that the child could actually score.
-As explained in the PISA manual \citep{pisa2012_technical}, these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured (the true distribution of the possible scores a student can achieve).
-
-A more intuitive explanation is this: suppose we have \(\mu_i\), the average student test score in mathematics for student \(i\).
-Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\) for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test.
-This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\).
-The reason why we use this procedure is because estimating a single estimate \(\mu_i\) is plagued with measurement error, among other types of bias \citep[see][]{wu2005}.
-The number of plausible values for PISA waves are usually five (although ten for PISA 2015) random draws from this distribution.
-In practice each student has 5 scores for each test, that resemble their distribution.
-Those values are continuous, ranging from 0 to 500, with a mean of 250.
-
-\subsection{Data analysis}
-
-The aim of this paper is to identify, disaggregate and explain country trends in the achievement gap for several countries.
-To represent the SES gap, most of the literature on achievement gaps has concentrated on indicators such as parental education, parental occupational status, income achievement gaps and actual SES achievement gaps \citep{fryer2004, hanushek_woesmann_tracking, saw2016, bradbury2015, byun2010}.
-The actual calculation of the achievement gap varies substantially and different strategies have been implemented.
-For example, \citet{micklewright} calculates the difference in achievement by crudely subtracting the difference between the 95th and 5th percentile of the mathematics distribution.
-Although in principle you should be able to capture some type of SES effect like this, theoretically, it should be much more accurate to difference out the mean score of, for example, parental education or some other SES proxy.
-\citet{saw2016}, for instance, used parental education as a proxy of SES, whereas \citet{byun2010} use a similar SES index as ours, but created by them.
-
-\citet{reardon_portilla}, \citet{anna2016} and \citet{anna2016_global} used a different method developed by \citet{reardon2011} which we partially adopt in this paper.
-SES achievement gaps are measured as the difference in standardized achievement between the 90th and 10th percentiles of the chosen SES variable.
-The rule of thumb to choose the 90th, 50th and 10th percentile is arbitrary, as others have used, for example, the 95th, 50th and 5th \citep{micklewright}.
-We use the conventional 90/10 cutoffs in the literature following the standard set by \citet{reardon2011}.
-
-For each country in each wave, SES disparities in achievement are measured as the gap in standardized achievement between the 90th and 10th percentiles of each country’s distribution of each SES variable, following Reardon’s (2011) method for income achievement gaps.
-The original strategy of \citet{reardon2011} is as follows: first, achievement is standardized (see below for a mathematical explanation of the standardization).
-We then use it to calculate the mean achievement (and standard error) for each category of the SES variable of interest (parent's education, income categories, etc..).
-"Category means are plotted at their percentile ranks and cubic models are fit through the points using weighted least squares.
-Finally, achievement at each country’s 90th and 10th SES percentiles is interpolated from the model" \citep{anna2016_global}.
-The result is a SES gap from an ordinal variable of interest.
-
-As mentioned before, PISA does not provide a single achievement indicator.
-Instead, we take the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.
-
-To standardize the test score we fit a linear model
-
-% Justify why you use the AGE variable
-\begin{equation}
-Y_i = \alpha + \beta_1 * AGE_i + \epsilon_i, \quad \epsilon_i ~ N(0, \sigma^2)
-\end{equation}
-
-for each wave, where \begin{math}Y_i\end{math} is the median student test score and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as \citet{reardon2011} \footnote{This does not mess up our analysis by masking age-specific gaps because all students in the sample are 15 year olds.
-Controlling for age is simply to adjust for monthly differences in ages.}) weighted by the student sample weights.
-
-We then adjust \begin{math} \hat{\gamma_i} \end{math} by
-
-% I think the denominator could be wrong! Because I include y_hat and that's what the equation is estimating
-\begin{equation}
-\hat{\gamma_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
-\end{equation}
-
-% # Take residuals of model and divide by rmse. Multiply that by
-% # 1 / sqrt(reliability of each survey), which is .y in the loop.
-% .x$adj_pvnum <- resid(mod1)/rmse(mod1, .x) * 1 / sqrt(.y)
-
-Where \begin{math}\hat{\epsilon_i}\end{math} is the residual for student \begin{math}i\end{math} and the denominator is the root mean square error of the model.
-
-This new standardized variable has a mean of zero.
-Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves because the test scores have now the same metric across time.
-However, if the variance of academic achievement changes over time, then standardizing the overall score at each country-wave pair actually makes the transformation biased.
-That is, by standardizing we're forcing the standard deviation of test scores to be zero across all waves.
-But if the true deviations of the median academic achievement changes over time, then the estimated trend in the SES gaps will be underestimated, or vice versa.
-
-% Try plotting this for each country and wave
-% Like, every country as a row and 6 columns for each wave.
-% \begin{figure}
-% \begin{center}
-\begin{figure}
-
-{\centering \includegraphics[width=3.5in,height=3in]{figure/variance_pisa-1}
-
-}
-
-\caption[Standard deviation of test scores across all waves]{Standard deviation of test scores across all waves}\label{fig:variance_pisa}
-\end{figure}
-
-% \caption{Standard deviation of test scores across all waves}
-% \end{center}
-% \end{figure}
-
-We plot the standard deviation of the mathematics test score for all waves in figure 1.
-The plot suggests that it's something we should not be deeply concerned with.
-The standard deviation of each wave seems to be following a very similar pattern with a not so drastic exception of the year 2000.
-
-Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap.
-This can be very misleading and suggest erroneous interpretations regarding trends in the of the gaps over time \citep{reardon2011}.
-PISA has tried to make sure the tests are comparable across waves \citep{pisa2012_technical}, but we still have to adjust for this imprecision.
-
-In order to correct the gap estimates for measurement error, we adjust student test scores(\begin{math} \hat{\gamma_i} \end{math}) by
-
-\begin{equation}
-\hat{\gamma_i} = \hat{\gamma_i} * \frac{1}{\sqrt{r}}
-\end{equation}
-
-where \begin{math}r\end{math} is the reliability score of the PISA wave \footnote{Other procedures multiply each country by their own reliability measure for each year-subject pair \citep{anna2016_global}.
-The reliability estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of person separation reliability such as internal consistency.
-Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and PISA 2015 has to yet release their own.
-At the moment of writing this paper, they were unavailable.
-For these reasons we implement the analysis following the original work of \citet{reardon2011}}.
-Each PISA survey provides a reliability indicator which we use accordingly.
-This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.
-
-After constructing the adjusted test score measurement, we estimate the SES dummy by calculating the weighted 90th and 10th quantiles for the SES composite index, and then generating a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile.
-
-We then fit
-
-% Replace the Y_i with the chosen y from the previos equation.
-\begin{equation}
-\gamma_i = \alpha_{j[i]} + \beta_{j[i]} * SES_i + \epsilon_i,\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
-\end{equation}
-
-% Try running also a linear model together with the multilevel model and ompcare results
-% Brunno suggested this because countries which might have a very different SES estimate
-% might be moved more towards the average.
-
-% mod2 <-
-%   lmer(adj_pvnum ~ escs_dummy + (1 + escs_dummy | country),
-%        data = .x,
-%        weights = .x[[weights_var]])
-
-where \(SES_i\) is whether the student is in the top/bottom SES dummy and we allow both \(\alpha_i\) and \(\beta_i\) to vary by country \(j\).
-We implement this separately for each wave and weight by the wave-specific student sample weights.
-
-Finally, we calculate the SES achievement gap for each country by extracting the fitted \(\alpha\) and \(\beta\) for each country \(j\) and calculating the difference of the predicted score of High and Low SES.
-
-\begin{equation}
-\begin{split}
-\quad \text{High SES}_j = \alpha_j + \beta_j \\
-\quad \text{Low SES}_j = \alpha_j \\
-\quad \text{SES gap}_j = \text{High SES}_j - \text{Low SES}_j
-\end{split}
-\end{equation}
-
-We also calculate the standard error of this difference and use it to create uncertainty intervals in the estimation of time trends.
-We fit a multilevel rather than a linear model because by pooling the information together, we weight countries appropriately to their sample size.
-Given that including the SES \emph{dummy} reduces the sample size considerably, we want to be able to estimate each country-difference as accurately as possible.
-For further exploration, we also compute the 90th/50th and 50th/10th SES gaps following the same method outlined above.
-% Explain more in detail that it's a dummy and not three groups.
-
-\section{Analysis and results}
-
-\subsection{Evolution of the achievement gap}
-
-% Previous table that had % of low and high education for low and high SES dummies
-
-% <<table_escs, include = TRUE>>=
-% summary_table <-
-%   left_join(summary_data, sample_tables_topbottom) %>%
-%   filter(name %in% c("2000", "2015")) %>%
-%   arrange(country, name)
-% 
-% table_coming <-
-%   summary_table %>%
-%   select(-perc) %>%
-%   gather(cats, vals, -(name:escs_dummy)) %>%
-%   unite(all_vals, escs_dummy, cats, sep = "_") %>%
-%   spread(all_vals, vals) %>%
-%   map_at(c(3, 4, 6, 7), ~ {
-%     .x[is.na(.x)] <- 0
-%     paste0(round(.x, 0), "%")
-%   }) %>%
-%   map_if(is_double, as.character) %>%
-%   data.frame() %>%
-%   xtable::xtable()
-% 
-% addtorow <- list()
-% addtorow$pos <- list(0)
-% addtorow$command <- paste("\\hline \\
-%                           & & & Low SES & & & High SES \\\\
-%                           \\cmidrule(l){3-5}
-%                           \\cmidrule(l){6-8}",
-%                           paste0(
-%                             c("Year",
-%                               "Countries",
-%                               "\\% Low Edu",
-%                               "\\% High Edu",
-%                               "Sample size",
-%                               "\\% Low Edu",
-%                               "\\% High Edu",
-%                               "Sample size"),
-%                             collapse = " & "),
-%                           "\\\\ \\hline")
-% 
-% print(table_coming,
-%       scalebox = '0.75',
-%       floating = FALSE,
-%       add.to.row = addtorow,
-%       include.colnames = FALSE,
-%       include.rownames = FALSE,
-%       hline.after = nrow(table_coming),
-%       caption.placement = "top",
-%       caption = "SES sample size and ISCED composition")
-% @
-
-% New table only has sample size, average score and s.e.
-% latex table generated in R 3.4.0 by xtable 1.8-2 package
-% Sat Sep  9 20:13:40 2017
-\begin{table}[ht]
-\centering
-\begin{tabular}{lllrrlrr}
-   \hline \
-                          & & & Low SES & & & High SES \\
-                          \cmidrule(l){3-5}
-                          \cmidrule(l){6-8} Year & Countries & N & Avg score & S.E & N & Avg score & S.E \\ \hline2000 & Australia & 294 & 0.22 & 0.07 & 270 & 1.22 & 0.09 \\ 
-  2000 & Austria & 294 & -0.04 & 0.11 & 235 & 1.15 & 0.15 \\ 
-  2000 & Canada & 1970 & 0.30 & 0.05 & 1357 & 0.90 & 0.07 \\ 
-  2000 & Denmark & 230 & -0.01 & 0.13 & 228 & 1.17 & 0.17 \\ 
-  2000 & Finland & 261 & 0.42 & 0.12 & 267 & 0.89 & 0.15 \\ 
-  2000 & France & 253 & 0.04 & 0.04 & 258 & 1.21 & 0.05 \\ 
-  2000 & Germany & 250 & -0.64 & 0.04 & 303 & 1.81 & 0.05 \\ 
-  2000 & Hungary & 291 & -0.46 & 0.09 & 254 & 1.66 & 0.12 \\ 
-  2000 & Italy & 278 & -0.31 & 0.05 & 278 & 0.81 & 0.06 \\ 
-  2000 & Netherlands & 129 & 0.46 & 0.08 & 158 & 1.25 & 0.11 \\ 
-  2000 & Poland & 209 & -0.31 & 0.04 & 166 & 1.25 & 0.06 \\ 
-  2000 & Spain & 318 & -0.27 & 0.05 & 340 & 1.09 & 0.07 \\ 
-  2000 & Sweden & 241 & 0.12 & 0.10 & 246 & 1.06 & 0.13 \\ 
-  2000 & United Kingdom & 531 & 0.16 & 0.04 & 415 & 1.33 & 0.06 \\ 
-  2000 & United States & 222 & -0.43 & 0.02 & 164 & 1.74 & 0.03 \\ 
-  2015 & Australia & 1694 & -0.15 & 0.04 & 1235 & 1.28 & 0.05 \\ 
-  2015 & Austria & 676 & -0.30 & 0.07 & 705 & 1.53 & 0.10 \\ 
-  2015 & Canada & 2215 & 0.20 & 0.03 & 1863 & 0.99 & 0.05 \\ 
-  2015 & Denmark & 1013 & 0.09 & 0.08 & 595 & 1.08 & 0.10 \\ 
-  2015 & Finland & 575 & 0.07 & 0.08 & 584 & 1.16 & 0.11 \\ 
-  2015 & France & 570 & -0.37 & 0.02 & 615 & 1.66 & 0.03 \\ 
-  2015 & Germany & 545 & -0.09 & 0.02 & 582 & 1.45 & 0.03 \\ 
-  2015 & Hungary & 466 & -0.72 & 0.07 & 589 & 1.79 & 0.09 \\ 
-  2015 & Italy & 947 & -0.20 & 0.03 & 1043 & 1.23 & 0.04 \\ 
-  2015 & Netherlands & 521 & 0.13 & 0.04 & 525 & 1.24 & 0.06 \\ 
-  2015 & Poland & 446 & 0.10 & 0.03 & 448 & 1.24 & 0.05 \\ 
-  2015 & Spain & 608 & -0.25 & 0.03 & 703 & 1.24 & 0.04 \\ 
-  2015 & Sweden & 527 & -0.20 & 0.06 & 542 & 1.32 & 0.09 \\ 
-  2015 & United Kingdom & 1387 & -0.03 & 0.03 & 1388 & 1.15 & 0.04 \\ 
-  2015 & United States & 585 & -0.39 & 0.01 & 544 & 1.22 & 0.01 \\ 
-   \hline
-\end{tabular}
-\caption{SES sample size and ISCED composition} 
-\end{table}
-
-Table 1 shows a description of the sample size and mean score of both top and bottom SES groups for only the first and last time point.
-One main concern from the planned analysis is that getting the top 90th percentile and bottom 10th percentile would result in a small sample size.
-The table suggests that we have a reasonable number of respondents to actually estimate gaps accurately.
-Moreover, we can see that in all instances the bottom SES group has a lower score than all top SES groups.
-
-We see that in some countries like Finland and Sweden the average low SES scores is actually above the average score of 0, suggesting they're very equal countries.
-However, most countries, like Spain, Poland, Hungary and Germany, their average low SES scores are much lower than the country average.
-We also see some countries with major changes from year 2000 to 2015, with, for example, Australia decreasing the low SES gap from 0.22 standard deviations to -0.15, well below the country average.
-In the next section we take a look at this in a more detailed fashion, considering all years in between.
-
-% TODO: Calculate the difference between 90/10 and then the ratio of this with the annual gain of one year of schooling (\citet{micklewright} did it).
-
-% \begin{figure}
-% \begin{center}
-\begin{figure}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_9010gaps-1}
-
-}
-
-\caption[Evolution of the achievement gap for selected countries]{Evolution of the achievement gap for selected countries}\label{fig:graphing_9010gaps}
-\end{figure}
-
-% \caption{Evolution of the achievement gap for selected countries}
-% \end{center}
-% \end{figure}
-
-We start by look at the achievement for some countries in figure 2.
-We plot only mathematics for each country and also a quadratic trend spline for \emph{both mathematics and literacy} pooled.
-Some countries have increased their achievement very strongly.
-For example, France, Austria and surprisingly Sweden have very steep slopes.
-France increased the gap by roughly 0.9 standard deviations, Austria by 0.6 and Sweden by 0.6.
-This pattern happens similarly for literacy.
-For example, France has an increase of 0.6.
-For such a short period of time, the magnitude of these increases are reasonably big.
-
-Given that no one has estimated the evolution of the gap we can't cross-check how other empirical estimations put France at.
-However, we can take \citet{micklewright} as the closest reference which also finds that France was a low dispersion country in 2000.
-However, there's no evidence on what happened over time.
-Luckily, the work of \citet{bernardi2016} did study the relationship of social origin effects (broadly speaking, not in terms of achievement gaps) in France and found that they increased in the last decades.
-
-Other countries have reasonable increases such as Finland and Hungary, with increases of nearly 0.6 and 0.4, respectively.
-Aside from these countries, there are other cases that have no change at all, specifically, Canada, Netherlands and Spain.
-Canada excels here not only because the gap has been stable over time, but because it has the smallest gap of all countries presented here.
-It's nearly 0.5 standard deviations in 2000 and it increased only by 0.2 in 2015.
-We can't really say Italy has had a significant increase since it has a sort of wiggly pattern that looks like it cancels itself out.
-
-On the other hand, we do have some countries which show a decrease in the SES achievement gap.
-Poland decreased by about -0.4 and Denmark by -0.2.
-However, the most notable cases are the United States and Germany.
-These two countries show a very high level of dispersion in the year 2000 with SES gaps of over 2 standard deviations.
-But in the 15-year time trend both countries reduced the gaps by -0.6 and -0.9 respectively.
-Their distinctively big gaps in 2000 also show up in the work of \citet{micklewright}.
-This corroborates the findings of \citet{reardon_portilla}, which found a decreasing gap for kindergarteners.
-This decline is evident also for 15 year olds, suggesting it might be more of an institutional change rather than a specific grade-level policy.
-Do note that the trendline is for both mathematics and literacy.
-If it were only for mathematics, then the trendline would probably be even more pronounced.
-Despite this, pooling both subjects gives even more strength to the analysis given that the results stand for both subjects together \footnote{Same graph is available for reading upon request, where we find pretty much the same results if not even more steep/downward slopes.}.
-We also plot the 80/20 and 70/30 SES gaps and find very similar patterns (see figure 13 in the appendix).
-
-Analyzing this graph we might naively conclude that these trends are not very steep and they should not be very important to consider.
-However, remember that the Y axis is measured in standard deviations.
-Small changes are actually very huge in practical terms.
-Take the case of Sweden.
-The slope does not look that steep but in reality it increased the gap from 1 standard deviation in 2000 to around 1.5 standard deviation in 2015.
-Now that we know that, the trends of Poland, United States, France and Germany are particularly shocking.
-
-We find that the initial gap for the U.S in 2000 is 2.17 varying between 2.13 and 2.21.
-This gap is much higher than found in previous studies, but again, we don't know how much previous estimates vary and nobody has really studied the 15-year old achievement gap in isolation.
-All country-level studies pool many studies together and control for age, removing the age-specific gap effect.
-
-Figure 3 takes a more direct approach and looks at the percentage change from the first and last timepoint available.
-Each data point has been computed together with it's 50\% and 95\% uncertainty interval \footnote{Each of these uncertainty intervals were computed using a 500-replicate bootstrap}.
-
-% How to output data frames are latex tables?
-
-% \begin{figure}
-% \begin{center}
-\begin{figure}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_perc_increase-1}
-
-}
-
-\caption[Evolution of the achievement gap for separate SES groups]{Evolution of the achievement gap for separate SES groups}\label{fig:graphing_perc_increase}
-\end{figure}
-
-% \caption{Percentage change in the achievement gap from 2000 to 2015}
-% \end{center}
-% \end{figure}
-
-Generally speaking, we see that most countries increased their achievement gap over time.
-France had an average increase of about 80\% since 2000 varying down to 40\%, whereas Germany had a similar figure but decreasing.
-Many of the countries we saw before that did not have a very steep slope, such as Hungary or Australia, in fact had increases of about 40\% of their gap.
-In contrast, we see that the U.S and Poland had also quite significant decreases of about 40\%.
-The benefit of presenting these estimates this way is that we can actually assess the uncertainty of each calculation, and we do find that they have wide variability.
-Despite this, most countries show a clear sign of either decreasing or increasing.
-We plot the same graph for the 80/20 (figure 14) and 70/30 (figure 15) SES gaps in the appendix and the results hold for these achievement gaps as well.
-
-% Code chunks for the same plots as above for reading. Available upon request.
-
-Something reassuring is that mathematics and reading (not presented) follow basically the same trend across all countries \footnote{Reading had an even bigger decrease for similar countries, and a much smaller increase.
-The previous two graphs for reading are available upon request}.
-This means that the result is not an artifact of chance alone.
-However, it's important to disentangle where is this gap originating from.
-Is this because the top are improving while the bottom decreases?
-Or is it that the bottom is catching up?
-Up next we plot the same graph but show the divergent patterns between High/Low SES origins.
-
-\subsection{Source of the achievement gap}
-
-% Add confidence intervals
-% \begin{figure}
-% \begin{center}
-\begin{figure}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/rate_increase/decrease-1}
-
-}
-
-\caption[Evolution of the achievement gap for separate SES groups]{Evolution of the achievement gap for separate SES groups}\label{fig:rate_increase/decrease}
-\end{figure}
-
-% \caption{Evolution of the achievement gap for separate SES groups}
-% \end{center}
-% \end{figure}
-
-Figure 4 shows the evolution for both low and high SES groups separately.
-The middle line is a summary line showing whether one of the groups is growing/shrinking faster than the other.
-We find plenty of variation in these patterns.
-For example, in the United States the top seems to be equalizing much stronger than the bottom is catching up.
-The U.K seems to be following the same path as the United States as well.
-On the other hand, in Poland the gap seems to be closing because the bottom SES group is catching up much faster.
-The Netherlands shows a similar pattern as the U.S and U.K patterns but upon closer inspection the explanation is very different.
-We see that the slope is negative (like in the previous two countries) but that's because the low SES group is decreasing at a greater rate than the high SES is decreasing.
-
-We now pay attention to the flat summary lines.
-These lines should be interpreted with caution because it does not mean that the gap is not changing.
-Denmark, Belgium, Netherlands, Norway and Spain show a flat line because the gap is growing very little, if at all.
-On the other hand, Sweden, France, Finland, New Zealand, Austria, among other countries, show a flat line because both groups are distancing themselves at a very similar rate.
-These results highlight the importance of not only summarizing average achievement gaps: the source of the gap varies greatly between countries and we see how each of these patterns contributes to the overall inequality of a country.
-
-After analyzing the trend of each country and where it's coming from, we start to find some interesting results.
-We see that the United States is closing the achievement gap at a very rapid rate, but at the same time we notice that it's because the top SES group is going down faster than the bottom is catching up.
-In contrast, we see countries like Germany and Poland where the gap is closing as well but because the bottom group is catching up faster than the high SES group is coming down.
-In fact, the Polish case is the most egalitarian of all: the bottom group is catching up at a rapid pace and the top group is maintaining their high levels of performance, fostering a truly equalizing effect.
-
-Quite interestingly, in countries where the gap is widening, both groups are distancing themselves at a very similar pace.
-Finland, France and Austria, show this exact pattern.
-Give or take, we have a solid pattern that when the gap widens, it's because there's top-to-bottom inequality.
-But when the gap is being reduced, the source can vary between top and bottom changes.
-
-To help assess these results even further, the source of this gap can be deconstructed further into two broader gaps.
-We want to know if these gaps have increased because the 90th SES group has distanced itself from the 50th SES group or because the middle has been also distancing itself from 10th SES group.
-For example, we're interested in searching which gap is bigger between countries, the 90th/50th or the 50th/10th?
-Moreover, are some of these gaps contributing more to the 90th/10th gap we saw earlier?
-Up next we plot all the combinations of the 90th/50th/10th gaps and pay attention to the magnitude of each gap in figure 5.
-Because we visualize three plots for every country, we only plot a selected number of countries that represent the overall patterns from the previous graphs \footnote{United States and Germany show a decrease, France shows a steep increase and Denmark shows no significant change.}.
-
-% \begin{figure}
-% \begin{center}
-\begin{figure}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_allgaps-1}
-
-}
-
-\caption[Evolution of 90/50/10 SES gaps for selected countries]{Evolution of 90/50/10 SES gaps for selected countries}\label{fig:graphing_allgaps}
-\end{figure}
-
-% \caption{Evolution of 90/50/10 SES gaps for selected countries}
-% \end{center}
-% \end{figure}
-
-We start off with Denmark.
-Generally speaking, we find that the 50th/10th gap is contributing more than the 90th/50th gap as the gap is bigger in mathematics.
-Furthermore, we find, just as in the 90th/10th SES gap, that there is not a significant decrease \footnote{With the exception of reading in the 50th/10th SES gap which seemed to decrease from nearly 1.5 to 1 standard deviation.}.
-France shows a similar pattern as Denmark in that the 50th/10th SES gap is the main contributer and that the three gaps have the similar trend.
-
-Germany and U.S, on the other hand, show a very different landscape.
-For example, Germany, shows that the 90th/10th gap is primarily driven by the 50th/10th SES gap and that the slopes of the 90th/50th gap is much weaker than the other two gaps.
-This means that top/middle SES groups have shrunk much less than the other groups.
-The big catching up is actually coming from the middle/bottom SES groups.
-However, we don't know if this is because the bottom is catching up or that the middle is coming down.
-Based on the gap plots from above, we can infer that it is because the bottom is catching up rapidly because against the 90th SES group, it is increasing steeply every year.
-
-The U.S shows a different pattern from all other countries.
-Both the 90th/50th/10th gaps seem to contribute rather similarly to the overall 90th/10th gap (although 50th/10th is marginally higher) but more interestingly, the 90th/50th gap has stagnated in the last 15 years.
-The top/middle groups have a relatively big gap and it has not changed at all.
-On the other hand, the steep decrease we saw for the U.S is coming exclusively from the 50th/10th gap shrinking.
-The reason, nonetheless, might be different from the German case, where it's probable that the bottom is catching up.
-Here, as in the 90th/10th gap, the middle might be 'dumbing' down closer to the low SES group.
-The surprising result is that the reduction of U.S gap is coming from the shrinking of the 50/10 gap and not the 90/10 gap.
-These results are puzzling.
-Does this mean that the overall distribution of inequality in the U.S is being reduced toward the benefit of the Low SES group?
-The reduction is not uniform across all SES groups.
-We see that the 90/10 gap is closing, yet the the 90/50 gap is very stable over time.
-Despite this, the 50/10 is also shrinking, meaning that the 90 and 50 SES groups are shrinking down relative to the bottom while the bottom is coming up.
-This is something we can only speculate but the evidence points toward something like that.
-
-In a more convenient way, we can inspect whether there is a lot of variance in the rate at which top/bottom groups are changing in figure 6.
-
-% \begin{figure}
-% \begin{center}
-\begin{figure}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/rate_change_graph-1}
-
-}
-
-\caption[Rate at which top/bottom SES groups are catching up for the 90/10 SES gap]{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}\label{fig:rate_change_graph}
-\end{figure}
-
-% \caption{Rate at which top/bottom SES groups are catching up for the 90/10 SES gap}
-% \end{center}
-% \end{figure}
-
-We see that there are more countries where High SES is increasing inequality faster than low SES is catching up (top-left panel) but there's also a fair share where low SES is catching up faster, such as the United States and Germany.
-Regardless of this, we see an overall negative relationship where countries in which the low SES group is catching up, the high SES group goes down.
-This is very interesting because it suggests that there's a trend for countries to have the bottom SES groups are climbing upwards, a trait of meritocracy.
-However, this means that the top group comes down, rather than stagnate or increase.
-Equally important, we see very few countries where low SES are decreasing (so going even further down) and High SES is also decreasing (bottom left panel).
-On the other hand, we do see more countries where the both groups are increasing (top right panel).
-We plot all remaining countries in the back, together with a trendline and we see that the relation holds for all PISA countries.
-
-To sum up, we find no evidence in support of our first hypothesis.
-We uncover that the tracking setup of a country does not imminently mean a certain degree of equality/inequality.
-Countries where the top SES group is increasing and the bottom is decreasing (top-left panel) have dramatically different tracking setups.
-Austria is an early tracker with several tracks, while Sweden is a late tracker with only 1 track.
-On the opposite panel (bottom-right), where we find countries that the bottom group is catching up while the top is coming down, we find similar results.
-Denmark and Germany are opposite poles in terms of tracking, while the U.S is distinctive in that it has little formal curricular stratification.
-We see no definitive evidence of top-to-bottom equality in all of these countries (except for Poland).
-We find that regardless of the different tracking setups, there's a tradeoff: either the bottom comes up and the top comes down or the top goes even further up and the bottom further down.
-
-But, similar as before, we don't know which group is causing the decrease/increase of the gap in the 90th/50th and 50th/10th gap. For that we plot the evolution of these gaps in figure 7.
-
-% \begin{figure}
-% \begin{center}
-\begin{figure}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graphing_ses_growth-1}
-
-}
-
-\caption[Evolution of the SES gaps for top/bottom groups]{Evolution of the SES gaps for top/bottom groups}\label{fig:graphing_ses_growth}
-\end{figure}
-
-% \caption{Evolution of the SES gaps for top/bottom groups}
-% \end{center}
-% \end{figure}
-
-With Denmark there's no average change across all years.
-However, for France, we see that the gap is growing much faster for the 50th/10th SES gap.
-For Germany we see that the gap is closing in all three groups although much more strongly for the 50th/10th SES gap.
-It looks like the 50th/10th gap is mainly driving the changes in the gaps for most of the countries that actually see a change over time.
-For the U.S, there's very little change in the 90th/50th SES gap, with most of the change coming from the 50th/10th gap.
-
-\subsection{Equality-efficiency tradeoff}
-
-Moving on, we now discuss if there's a tradeoff between an educational system being efficient, i.e raising the overall performance of the country, and whether that challenge compromises equality at the same time.
-For example, in order to improve the average score in, let's say, mathematics, tracking can improve the average score by making children in the higher trackers perform better, raising the overall performance.
-This, however, can come at a price.
-If tracking actually benefits the top but does not increase the performance of the bottom tracks then we see a net increase of inequality accompanied by an increase in the overall performance.
-Given that we have the data for the average performance for each country and the achievement gap, we can investigate this relationship in detail.
-We first being by plotting all countries pooled for each year in figure 8.
-
-% \begin{figure}
-% \begin{center}
-\begin{figure}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/avgperformance_inequality-1}
-
-}
-
-\caption[Relationship between 90/10 achievement gap and average performance in mathematics]{Relationship between 90/10 achievement gap and average performance in mathematics}\label{fig:avgperformance_inequality}
-\end{figure}
-
-% \caption{Relationship between 90/10 achievement gap and average performance in mathematics}
-% \end{center}
-% \end{figure}
-
-For the past 15 years we see a very strong correlation suggesting that countries that perform very well in PISA have actually very low achievement gaps (or the other way around) \footnote{We graph all countries-years pooled as well and the relationship is identical}.
-The correlations are between -0.73 and -0.88.
-This confirms the theoretical and empirical frameworks suggested by \citet{werfhorst_mijs}.
-However, we should not get trapped into an ecological fallacy and assume this relationship is identical within each country.
-With the available data we can look at the evolution separately for countries.
-We plot it up next in figure 9.
-
-% Add confidence intervals
-% \begin{figure}
-% \begin{center}
-\begin{figure}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/country_achievement_and_inequality-1}
-
-}
-
-\caption[Time trends of achievement gap and average achievement between 2000 and 2015]{Time trends of achievement gap and average achievement between 2000 and 2015}\label{fig:country_achievement_and_inequality}
-\end{figure}
-
-% \caption{Time trends of achievement gap and average achievement between 2000 and 2015}
-% \end{center}
-% \end{figure}
-
-% Add descriptive table with average test scores, degree of tracking, level of income inequality, among other things.
-
-Looking at figure 9, we see a strong correspondence between the two variables.
-In countries like France we see quite a functional form between both inequality and performance.
-As average performance plummets, the achievement gap is increasing.
-This pattern is also seen in Austria, Australia and Germany.
-Yet we also see the opposite trend.
-We see countries like Poland, Finland, Canada and less pronounced, Italy, where this pattern is evident.
-The surprising case here is the United States.
-We see that as the average performance has been decreasing, the average achievement gap has been decreasing as well.
-In fact, among the few countries we should find something that resembles this pattern should be the U.S because it does not have a strict institutionalized tracking system.
-However, the explanation behind this pattern is very speculative.
-
-Finally, the last notable pattern is that in some countries the average performance is so high that it actually overperforms the achievement gap.
-Only Finland, Canada and Netherlands (three very equal countries) have their average performance in standard deviations above their level of inequality, which coincidentally, are countries that have the lower level of inequalities among the ones presented here.
-
-These results actually confirm our second hypothesis, that there's a negative relationship between the two variables.
-We see countries like Poland, Canada and Germany where the pattern is just as the aggregate trend, namely that as the average performance increases, inequality seems to be reduced in parallel.
-Conversely, France, Austria and Australia, to name a few, portray the opposite: as average performance decreases inequality increases.
-But we also uncovered important variation within countries in the relationship between average achievement and dispersion of achievement (have a look at the Netherlands, Italy and the U.S).
-
-\subsection{The role of curricular tracking in explaining the achievement gap}
-
-% 90/80/70 gaps merged
-
-% end here
-
-We now move to the final section of the paper, where we discuss what is a likely explanation to these patterns.
-The reason why countries differ in their evolution of inequality is still unknown.
-There is ample evidence showing the inequality between countries, to a certain extent, can be explained by the degree of tracking \citep{hanushek_woesmann_tracking}.
-But we are also interested in what explains the between country evolution of inequality, that is, the explanation as to why in certain countries it is increasing more than others over time.
-
-This is clearly a difficult task given that, at least at the country-level, not many features of the educational system change in a 15-year time span, resulting in a practically invariant variable from which to explain the level of inequality.
-However, there are other strategies that can at least partially help us to understand the phenomena, and these are precisely the ones we follow.
-
-Following the literature on the topic, we suspect that this evolution is in part very tightly related to the degree of stratification of the educational system.
-In a very simple frame-work, \citet{hanushek_woesmann_tracking} found that tracking indeed exacerbates inequality in the life course of children.
-Following the same strategy, we believe that tracking is also allowing the same gap in 15 year olds to grow over time, accompanied by other things.
-Yet, instead of concentrating only on the age of selection (as they do), we explore a more fine grained definition of tracking, which is possible through the work of \citet{bol2013}.
-Aside from the age at first selection into tracks, we also use the the number of tracks in the country, the percentage of the entire curriculum that is tracked and a vocational index \footnote{This index is a factor loading from a principal factor analysis of the percentage of students in upper secondary vocational education (taken from two sources, to reduce measurement error).
-We take all of the data from \citet{bol2013}.}.
-
-Considering that our summarized data has very few observations \footnote{We have per year, per country, one single inequality indicator, leaving the total sample size at 195}, we want to assess the uncertainty of our model by using a fully bayesian hierarchical linear model.
-The benefits of this approach are twofold: first, we can assess the model fit much more intuitively, something very important given our sample size, and secondly, given that our countries are clustered into years, we can control for country heterogeneity within years, removing any bias in the standard errors.
-
-In the final model, we regress the inequality of a country on a dummy variable where 1 equals only 1 track vs more than 1 track, whether the age of first selection is 15 or more vs below 15, a dummy stating whether the percentage of tracked curriculum is above 0 (that is, any tracking), and the standardized vocational index \footnote{We create dummies to avoid multicollinearity.
-The final model has a maximum VIF of 2.2.}.
-We are reluctant to include more variables in the model, first to keep preserve parsimony, and secondly to prevent overfitting due to our sample size.
-We present the results in table 2.
-
-% latex table generated in R 3.4.0 by xtable 1.8-2 package
-% Sun Sep 10 14:55:49 2017
-\begin{table}[ht]
-\centering
-\begin{tabular}{lllll}
-  \hline
-  & Model 1 & Model 2 & Model 3 & Model 4 \\ 
-  \hline
-Only 1 track & -0.53 (-0.64/-0.41) & -0.31 (-0.44/-0.18) & -0.31 (-0.45/-0.17) & -0.21 (-0.34/-0.07) \\ 
-  Age selection $>$= 15 &  & -0.38 (-0.51/-0.25) & -0.38 (-0.51/-0.26) & -0.57 (-0.71/-0.43) \\ 
-  \% of curric tracked &  &  & 0 (-0.23/0.23) & 0.6 (0.3/0.89) \\ 
-  Vocational Index &  &  &  & -0.27 (-0.36/-0.19) \\ 
-  Intercept & 1.77 (1.69/1.84) & 1.94 (1.85/2.03) & 1.94 (1.69/2.19) & 1.54 (1.27/1.81) \\ 
-   \hline R square: & 23\% & 32\% & 32\% & 40\% \\ Sample size: & 195 & 195 & 195 & 183 \\ \hline \hline
-\end{tabular}
-\caption{Explaining 90/10 achievement gap - Multilevel model} 
-\end{table}
-
-Given that the classical p values are inexistent in bayesian modeling, we present uncertainty intervals next to each coefficient.
-The first model only includes the 1 track dummy.
-This variable shows that countries that have one track, have on average, half a standard deviation of less inequality than other countries.
-This covariate alone explains about 23\% of the variance of the evolution of the 90/10 achievement gap.
-The second model now includes the dummy of whether the country has an age selection of 15 or more and we can see that countries with late selection into tracking have about -.38 less standard deviations in inequality.
-Also note that the number of tracks dummy was now reduced, suggesting these two variables are explaining similar things.
-This is expected given that late tracking is usually associated with fewer tracks, such as in the Scandinavian countries.
-The third model now includes whether the country has any degree of tracking, meaning whether the country has above 0\% of the curriculum tracked.
-The coding of the variable comes from visual inspection of the distribution, where some countries, such as the United States and Canada, have untracked curriculums.
-For this last variable we see no relationship at all, but after including the vocational index, the model is completely reversed with countries with tracked curriculum having about 0.6 more standard deviations in inequality that countries with no tracking.
-
-We also note that vocational enrollment is associated with less inequality.
-Not surprisingly, the age of selection coefficient also increased due to the correlation between age of selection and vocational enrollment.
-Countries with high vocational enrollment are also those with early age selection, such as Germany and Austria.
-These two variables should be inspected further, perhaps with an interaction, given that both coefficients are sizable and very correlated.
-This final model explains over one third of the variance in the evolution of the 90/10 SES gap, with an R square of 40\%.
-The uncertainty of each of the covariates varies a lot with the percentage of curriculum tracked dummy ranging all the way from 0.9 to 0.3.
-Nonetheless, we see that the other coefficients are very strong and have reasonable intervals.
-It is very important to consider the uncertainty of our estimates, to gain confidence in our model fit.
-So far, this model seems to fit the data quite reasonably.
-
-All of these countries have a different tracking structure, with some setups being more egalitarian than others.
-Using the previous model we make a simulation and predict the level of the 90/10 achievement gap for all possible tracking setups for all countries.
-In figure 10 we plot all these combinations.
-The X axis has the number of tracks that a country has, the Y axis has the predicted 90/10 achievement gap and below each country name we have whether the country has an age selection of 15 or more or below 15.
-In addition, to confirm how accurate the model is, we also plot the actual level of inequality in the actual tracking setup of the country.
-Let's take Germany for example.
-
-\begin{figure}
-
-{\centering \includegraphics[width=6.5in,height=7in]{figure/graph_complete_tracking_graph-1}
-
-}
-
-\caption[Simulation]{Simulation: predicted 90/10 achievement gap for different tracking setups}\label{fig:graph_complete_tracking_graph}
-\end{figure}
-
-The star point for Germany is in the '\textgreater 1 track' and '\textless 15 age' setup, meaning that Germany has a tracking system that has more than 1 track and an age of selection of below 15.
-We see that the model predicts the level of inequality for that set up quite accurately (height of the bar), relative to the actual level (the star point).
-But if the setup would be in the ideal '1 track' and age of selection above or equal to 15, the inequality would be much lower.
-
-These graph reveals some interesting patterns.
-Generally speaking, we find that the simulation predicts that virtually all countries which are not in this 'ideal' tracking, that would switch to the 'ideal' setup, would experience a reduction of the achievement gap.
-Yet we also find the opposite result.
-Countries which are in the 'ideal' setup that would change to the 'worst' setup would see a widening of the achievement gap.
-On average, we find that if all countries switched to the 'ideal' setup, we would see an average reduction of 20\%.
-This is a very important figure considering that \citet{reardon2011} found that the US gap increased by about 40-50\% in 30 years.
-Had all countries in the 'ideal' tracking switched to the 'worst' tracking, we would see an average widening of nearly 72\%.
-For those countries which are not in the 'ideal' tracking setup, we compute the percentage change from switching to the ideal tracking in table 3.
-
-% latex table generated in R 3.4.0 by xtable 1.8-2 package
-% Sat Sep  9 20:14:00 2017
-\begin{table}[ht]
-\centering
-\begin{tabular}{lllrrl}
-  \hline
-Country & Number of tracks & Age of selection & Inequality & Predicted & \% reduction \\ 
-  \hline
-Austria & More than one track & Less than 15 & 1.58 & 1.27 & 20\% \\ 
-  France & More than one track & 15 or more & 1.84 & 1.27 & 31\% \\ 
-  Germany & More than one track & Less than 15 & 1.95 & 1.27 & 35\% \\ 
-  Hungary & More than one track & Less than 15 & 2.32 & 1.26 & 46\% \\ 
-  Italy & More than one track & Less than 15 & 1.48 & 1.26 & 15\% \\ 
-  Netherlands & More than one track & Less than 15 & 1.06 & 1.27 & -19\% \\ 
-  Poland & More than one track & 15 or more & 1.44 & 1.27 & 12\% \\ 
-   \hline
-\end{tabular}
-\caption{Simulation: reduction of achievement gap if countries switched to 'ideal' tracking} 
-\end{table}
-
-% This is for when calculating the mean 'widening' of the gap if countries with 'ideal' tracking swtiched to the 'worse' tracking
-
-The biggest reductions are for both Hungary and Germany, with 46\% and 36\% respectively.
-France, which had a particularly steep increase in the last 15 years, would see a reduction of the 90/10 SES gap of 31\% and Austria, with a very similar tracking structure as the German case, would see their achievement gap reduced by 20\%.
-The very surprising result is the Dutch case, for which a switch from the 'worst' tracking setup to the 'best' case would result in an increase of the gap of 19\%.
-
-This last result could mean two things.
-First and foremost, the Dutch educational system has built other mechanisms that, despite the tracking system, help children from low social origins escalate upwards.
-This explanation is not very trivial considering that in most mobility research, the Netherlands stands out as a very equal country relative to other countries with a similar tracking structure, like Germany \citep{shavit1993, esping2012}.
-Another possible explanation, and something which needs to assessed, is whether our model is predicting accurately.
-If we look at the plot from above, we see that for some countries, the model is a bit off from the actual level of inequality.
-Moreover, although we see that if all countries switched to the 'ideal' tracking setup we would see a vast reduction in inequality, we also see that each of these predictions is followed by a wide uncertainty interval.
-
-Considering that the results are very consistent for every country, and in most countries the prediction matches the actual value, this could simply be a consequence of low sample size.
-Whether these results stand the test of a different empirical strategy or a higher sample size, we leave up to future research.
-
-A more summarized empirical strategy is to actually standardize all the tracking indicator into one index, something that \citet{bol2013} already do quite nicely.
-Using this standardized tracking variable and the same bayesian multilevel model, we now model the tracking and vocational index as before, but add an interaction between two, given that in the previous model we recorded a strong shift in the model when tracking-related variables and the vocational model where both included in the same model.
-Table 4 shows the results for this model.
-
-% latex table generated in R 3.4.0 by xtable 1.8-2 package
-% Sat Sep  9 20:16:18 2017
-\begin{table}[ht]
-\centering
-\begin{tabular}{llll}
-  \hline
-  & Model 1 & Model 2 & Model 3 \\ 
-  \hline
-Tracking Index & 0.26 (0.21/0.32) & 0.33 (0.27/0.4) & 0.41 (0.34/0.48) \\ 
-  Vocational Index &  & -0.19 (-0.26/-0.11) & -0.22 (-0.29/-0.15) \\ 
-  Trackin * Vocational Index &  &  & -0.12 (-0.17/-0.06) \\ 
-  Intercept & 1.6 (1.53/1.67) & 1.65 (1.58/1.72) & 1.71 (1.63/1.78) \\ 
-   \hline R square: & 24\% & 30\% & 34\% \\ Sample size: & 195 & 183 & 183 \\ \hline \hline
-\end{tabular}
-\caption{What explains 90/10 achievement gap? Tracking and Vocational interaction} 
-\end{table}
-
-We confirm the results from the previous models, where more tracking is associated with an achievement gap of about 0.4 standard deviations wider and more vocational enrollment is associated with a reduction of about 0.2 standard deviations, half the slope of tracking.
-Could it be that the tracking effect might be offsetting the effect of the vocational index given that it's half the size of the slope?.
-Both these coefficients are actually significantly different from each other, so it is possible.
-Before we explore it visually, note that the interaction between the two is very reliable and significant, considering the small sample size.
-Figure 11 plots the interaction for different quantiles of the vocational index.
-
-% \begin{figure}[!htbp]
-% \begin{center}
-\begin{figure}
-
-{\centering \includegraphics[width=6.5in,height=5in]{figure/interaction_plot-1}
-
-}
-
-\caption[Interaction between tracking and vocational values]{Interaction between tracking and vocational values. Legend ordered relative to the lines}\label{fig:interaction_plot}
-\end{figure}
-
-% \caption{Interaction between tracking and vocational values. Legend ordered relative to the lines.}
-% \end{figure}
-% \end{center}
-
-The results are very interesting.
-We see that for lower levels of tracking (meaning little to no tracking) there's a very low achievement gap, regardless of whether the country has very high or very low vocational enrollment.
-However, once tracking enters the picture, vocational enrollment can be a strong equalizer.
-We see that the bottom quantile of vocational enrollment has almost twice the achievement gap as the top quantile for vocational enrollment.
-Moreover, our earlier interpretation (that tracking is more important than vocational enrollment for reducing the achievement gap) still holds and is very evident here.
-Despite the equalizing power of vocational enrollment, a high level of tracking with the highest level of vocational enrollment still leaves a country with an achievement gap of over 1.5 standard deviations.
-However, a country with low levels of tracking and \textit{any} level of vocational enrollment is nearly 1.3 standard deviations at best.
-A key finding from these results is that tracking is a reasonable explanation of the 90/10 achievement gap, and we should consider it as a very important component in the reduction of the achievement gap.
-However, these results highlight that it should not be studied in isolation from other features such as the degree of vocational enrollment.
-We see a very sizable change in the achievement gap for a given country with the same level of tracking and high and low levels of vocational enrollment.
-
-The exact mechanisms through which the tracking setup is explaining these inequalities is still very speculative.
-We could argue that tracking might've introduced the inequality once it was implemented.
-Another possibility is that tracking, although not changing, is currently exacerbating it every year.
-Another explanation, which we find strong evidence for, is that the tracking setup interacts very closely with the vocational setup.
-These explanations are very speculative and further research should carry the task of looking for a more concrete explanation.
-
-% We need to define key numbers, to refer in the abstract, like the achievement gap closes % percent in the counterfactual and there's a difference of % percent between same level of tracking but different vocational enrollments.
-
-To test whether these results hold for other samples, we compute the same models as before but also for the 80/20 and 70/30 SES gap. Figure 12 plots these coefficients for the three SES gaps.
-
-% \begin{figure}[!htbp]
-% \begin{center}
-\begin{figure}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/all_gaps_plot-1}
-
-}
-
-\caption[What explains achievement gaps? Model comparisons]{What explains achievement gaps? Model comparisons}\label{fig:all_gaps_plot}
-\end{figure}
-
-% \caption{What explains achievement gaps? Model comparisons}
-% \end{figure}
-% \end{center}
-
-With the exception of the Intercept, the coefficients of the three models look pretty much the same.
-The tracking index seems slightly stronger for the 90/10 model but the uncertainty intervals overlap with the other two models.
-These results give our claims much more credibility.
-
-Finally, until now, we've studied how we can explain the evolution of the 90/10 achievement gap, by clustering on the year dimension of each country and then performing a linear multilevel model.
-This approach was adopted given that tracking hardly changes over time, making it impossible to model the change over time directly.
-However, there is another way of estimating the evolution of the achievement gap, which we try up next.
-
-Instead of estimating the evolution over time, we calculate the cumulative change over time for each country, summarizing the 6 year-time span of achievement gaps into one cumulative inequality gap.
-This approach dramatically reduces the total sample size to 32, the number of total countries available \footnote{We've been presenting a selected number of countries given the lack of space, but for every model/estimation we include all countries available.}.
-It is for this reason that we only include at most 2 covariates in the model, the tracking and vocational index, which both summarizes the previous models and keeps it parsimonious.
-
-% latex table generated in R 3.4.0 by xtable 1.8-2 package
-% Sat Sep  9 20:16:22 2017
-\begin{table}[ht]
-\centering
-\begin{tabular}{llll}
-  \hline
-  & 90th/10th & 80th/20th & 70th/30th \\ 
-  \hline
-Track Index & 1.66 (0.85/2.49) & 1.11 (0.46/1.78) & 1 (0.37/1.67) \\ 
-  Vocational Index & -1.04 (-1.95/-0.12) & -0.9 (-1.64/-0.14) & -0.87 (-1.6/-0.16) \\ 
-  Intercept & 9.37 (8.6/10.12) & 4.6 (3.99/5.19) & 3.89 (3.3/4.47) \\ 
-   \hline R square: & 30\% & 25\% & 23\% \\ Sample size: & 32 & 32 & 32 \\ \hline \hline
-\end{tabular}
-\caption{What explains the cumulative achievement gaps? Models for three different achievement gaps} 
-\end{table}
-
-We see that 1 standard deviation increase in the tracking index widens the 90/10 SES gap by 1.66 standard deviations over time, and the vocational index reduces it by 1 standard deviation and these two covariates explain nearly one third of the cumulative gap.
-As we decrease the achievement gap (80/20, 70/30) these coefficients become smaller, but keep their strength and their predictive power.
-Comparing this model to the previous models, we see a very similar landscape, but an even more reassuring one, given than we're tackling the evolution of the achievement gap directly with this indicator.
-We can safely say that these results, at worse, are supporting the previous findings and making the claim even more credible.
-
-\section{Conclusion}
-
-The literature on achievement inequality has recently started studying the evolution of the achievement gap and it has uncovered a wide variety of results.
-The main finding is that the U.S gap has been widening over time by about 40-50\% from 1970 until the 2000.
-Others find that this recent trend has reversed, at least for an age-specific group since the 2000's.
-From an international perspective, recent studies have shown that there is wide variability in this trend, with some countries showing decreases while others increases.
-This paper looks to investigate the magnitude of these gaps across selected countries, and attempts to explain why they are happening.
-
-We start off by looking at the evolution of the achievement gap and document the same variation for the 15-year old achievement gap across countries.
-The United States has been closing its 90/10 gap by about 50\% with a wide interval going down to nearly 35\%.
-This reduction is also present in Germany, quite shockingly a country with a very institutionalized tracking system.
-Other countries such as France and Austria experience the contrary with persistent increases in the achievement gap.
-Moreover, once we disaggregate these patterns into a more fine-grained analysis, we discover that the dynamics are very different.
-The U.S gap is shrinking because both the 90th and 50th SES groups are actually going down while the bottom group is coming up.
-These results are evident because the 90/10 and 50/10 gap are shrinking at a very rapid rate, whereas the 90/50 gap is flat in the same 15 year interval.
-Germany, which has a very similar reduction, portrays a different explanation.
-All three achievement gaps are shrinking suggesting a more uniform reduction.
-On the other hand, for countries which are experiencing a widening of the gap, we see that it's happening at all instances of the distribution; all SES groups show a very steep increase.
-
-We then assess whether this gap is correlated with the country's average performance, and find a very strong correlation at the aggregate level (that is, all year-country combination pooled).
-Yet when we look at this relationship within each country, we find exactly the same pattern: countries with high performance show low inequality, whereas countries with low performance show high inequality.
-This gives greater support to the work of \citet{werfhorst_mijs}.
-Further research should attempt to disentangle the causality between these two opposite forces: is efficiency trumping inequality, or is it the other way around?
-
-We complete the analysis by studying what explains the evolution of this inequality and we find that the degree of tracking of a country seems to be a possible explanation, but it should not be studied without factoring in the degree of vocational enrollment.
-We find that if a country has a low level of tracking, vocational enrollment will not help decrease the achievement gap.
-Conversely, once tracking enters the picture, vocational enrollment can help unease the burden of inequality and reduce the achievement gap very strongly.
-In a similar vein, we use this model to perform a simulation for each country: if every country switched to little or no tracking, what would be the resulting reduction in the 90/10 achievement gap?
-We find that if all countries switched, we would find that on average countries would see their gap shrink by about 20\%, with countries like Hungary and Germany experiencing reductions of 46\% and 36\% respectively.
-We contrast the former simulation with another counterfactual scenario: what would be the average gap increase if all countries switched to a system where tracking predominates?
-We record an average widening of 72\%, something that suggests that increasing tracking is more than three times worse than reducing it.
-
-We conclude by showing that tracking and vocational enrollment are playing an important role in explaining the evolution of the achievement gap.
-We find that these results are very suggestive that institutional features, more specifically, stratified curriculums, might be playing a very sizable role in the promotion of inequality.
-After having ran a battery of tests and different empirical strategies, the results are consistent and point out that we should pay attention to the source of the achievement gaps and study whether particular SES groups are contributing more to the gap than others.
-Aside from the results, this paper also revealed some puzzling questions that we leave up to future research to study.
-
-\bibliography{mybibliography.bib}
-
-\pagebreak
-
-\section{Appendix}
-
-\begin{figure}[!htbp]
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/graph_all_gaps-1}
-
-}
-
-\caption[Evolution of the achievement gap for several gaps]{Evolution of the achievement gap for several gaps}\label{fig:graph_all_gaps}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/percentage_change_8020-1}
-
-}
-
-\caption[Percentage change in the 80/20 achievement gap from 2000 to 2015]{Percentage change in the 80/20 achievement gap from 2000 to 2015}\label{fig:percentage_change_8020}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=5.5in,height=5in]{figure/percentage_change_7030-1}
-
-}
-
-\caption[Percentage change in the 70/30 achievement gap from 2000 to 2015]{Percentage change in the 70/30 achievement gap from 2000 to 2015}\label{fig:percentage_change_7030}
-\end{figure}
-
-\end{document}
+{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in map\_dbl(adapted\_year\_data, nrow) \%>\% format(big.mark = "{},"{}): could not find function "{}\%>\%"{}}}\end{kframe}
+\end{knitrout}

### 2019-10-21 code/02-analysis.R
Adds school level dummy for top/worst

@@ -59,7 +59,7 @@ load_escs <- function(raw_data_dir, recode_cntrys) {
     mutate(name = seq(2000, 2012, 3))
 }
 
-harmonize_student <- function(raw_student, recode_cntrys, final_countries) {
+harmonize_student <- function(raw_student, recode_cntrys) {
 
   # PISA 2000
   raw_student$value[[1]] <-
@@ -288,10 +288,9 @@ harmonize_student <- function(raw_student, recode_cntrys, final_countries) {
 
       .x <-
         .x %>%
-        mutate(SCHOOLID = as.character(SCHOOLID)) %>% 
-        filter(country %in% final_countries)
+        mutate(SCHOOLID = as.character(SCHOOLID))
         
-        .x
+      .x
     })
 
   raw_student
@@ -711,8 +710,19 @@ create_escs_dummy <- function(merged_student_school, probs) {
           with(country, case_when(escs_trend >= quan[2] ~ 1,
                                   escs_trend <= quan[1] ~ 0))
 
+
+        quan_all <- quantile_missing(country, weights_var, seq(0, 1, .1))
+
+        if (all(is.na(quan_all))) {
+          country$quantiles_escs <- NA_real_
+        } else {
+          country$quantiles_escs <-
+            as.numeric(cut(country$escs_trend, quan_all, labels = 1:10))
+        }
+
         country
       })
+
       rm(country_split)
 
       .x <-
@@ -769,7 +779,7 @@ create_scores <- function(created_escs_dummy, reliability) {
 
         # Take residuals of model and divide by rmse. Multiply that by
         # 1 / sqrt(reliability of each survey), which is .y in the loop.
-        test_name <- paste0("adj_pvnum_", test_score)
+        test_name <- tolower(test_score)
         .x[[test_name]] <- resid(mod1) / modelr::rmse(mod1, .x) * 1 / sqrt(.y)
 
         .x[, test_name]
@@ -1602,35 +1612,6 @@ select_cols_student <- function(harmonized_student) {
 
 }
 
-select_cols_student <- function(harmonized_student) {
-
-  harmonized_student$value <-
-    map(harmonized_student$value, ~ {
-
-      escs_col <- if (unique(.x$wave) == "pisa2015") "ESCS" else " "
-
-      .x %>%
-        select(country,
-               wave,
-               SCHOOLID,
-               stu_id,
-               ## escs_dummy,
-               starts_with("PV"),
-               contains(escs_col),
-               contains("W_FSTUWT"),
-               AGE,
[…]

### 2019-10-21 code/02-analysis.R
Weights analysis added

@@ -279,12 +279,18 @@ harmonize_student <- function(raw_student, recode_cntrys) {
                matches("ISCEDO"),
                # The ESCS index from pisa 2015
                matches("ESCS"),
-               matches("ST16Q03|ST37Q01|ST019CQ01T|ST013Q01TA"))
+               matches("ST16Q03|ST37Q01|ST019CQ01T|ST013Q01TA"),
+               # Weight column
+               matches("W_FSTUWT"),
+               )
 
       # Harmonize SCHOOLID variable for all waves
       if (unique(.x$wave) == "pisa2015") {
-        .x <- rename(.x, SCHOOLID = CNTSCHID)
+        .x <- rename(.x,
+                     SCHOOLID = CNTSCHID)
       }
+
+      .x <- rename(.x, stu_weight = W_FSTUWT)
 
       .x <-
         .x %>%
@@ -691,13 +697,7 @@ create_escs_dummy <- function(merged_student_school, probs) {
   merged_student_school <-
     map(separate_waves, function(.x) {
 
-      conf <- if (unique(.x$wave) == "pisa2015") pisa2015_conf else pisa_conf
-      weights_var <- conf$variables$weightFinal
-
-      ## # Harmonize education variable
-      ## if (unique(.x$wave) == "pisa2000") {
-      ##   .x$high_edu_broad <- dplyr::recode(.x$high_edu_broad, `6` = 7)
-      ## }
+      weights_var <- "stu_weight"
 
       country_split <- split(.x, .x$country)
 
@@ -749,7 +749,7 @@ create_scores <- function(created_escs_dummy, reliability) {
   all_waves <- map2(separate_waves, reliability, function(.x, .y) {
 
     conf <- if (unique(.x$wave) == "pisa2015") pisa2015_conf else pisa_conf
-    weights_var <- conf$variables$weightFinal
+    weights_var <- "stu_weight"
 
     # Look over each test and create an adjusted version of it and
     # then cbind them together to the original dataset
@@ -805,8 +805,7 @@ sample_size_calc <- function(df, probs, selected = FALSE, cnts = NULL) {
   cnt_to_bind <-
     map(df, function(df) {
 
-      conf <- if (unique(df$wave) == "pisa2015") pisa2015_conf else pisa_conf
-      weights_var <- conf$variables$weightFinal
+      weights_var <- "stu_weight"
 
       print(unique(df$wave))
 
@@ -1604,7 +1603,8 @@ select_cols_student <- function(harmonized_student) {
[…]

### 2019-10-21 code/plan.R

@@ -113,18 +113,18 @@ recode_cntrys <-
 
 # Function calculates the bottom 30th quantile for the bottom educated and the 70th quantile
 # for the top educated. If the quantiles cannot be estimated, it returns two NA's instead
-quantile_missing <- function(df, weights, probs) {
+quantile_missing <- function(df, weights, probs, ch_var = "escs_trend") {
 
   quan <- try(Hmisc::wtd.quantile(
-    df$escs_trend,
+    df[[ch_var]],
     weights = df[[weights]],
     probs = probs
   ))
 
   if (any("try-error" %in% class(quan))) {
-    return(c(NA, NA))
+    return(rep(NA, times = length(probs)))
   } else {
-    return(c(quan[1], quan[2]))
+    return(quan)
   }
 }
 
@@ -633,8 +633,7 @@ plan <-
 
     harmonized_student = harmonize_student(
       raw_student,
-      recode_cntrys,
-      final_countries
+      recode_cntrys
     ),
     rm_raw = delete_raw_data(), # Delete the student raw data after harmonized
     # Check how we're doing with memory
@@ -652,7 +651,7 @@ plan <-
     merged_student_escs = merge_student_escs(selected_cols_student, loaded_escs),
 
     # This is the step when the tibble with a list column becomes an entire
-    # data frame with over 1M students for all waves with the stuent and
+    # data frame with over 1M students for all waves with the student and
     # school data.
     merged_student_school = merge_student_school(merged_student_escs,
                                                  harmonized_school),
@@ -670,20 +669,24 @@ plan <-
     ############################################################################
 
     ## escs_dummy_data now contains the dummy column for the 90th/10th
+    ## and the quantile_cuts with 10% quantile groups for the escs variable
     created_escs_dummy = create_escs_dummy(merged_student_school, c(0.1, 0.9)),
 
     # This function adjusts and standardizes the test scores
     created_scores = create_scores(created_escs_dummy, reliability_pisa),
 
+    # Only keep countries with more than 50% of data for all years
+    filtered_data = filter(created_scores, country %in% final_countries) %>% sample_frac(10),
+
     # These two below are the final datasets for the modelling.
 
     ## data_modelling now contains the adjusted math/read column for all students
     ## tests scores
-    data_modelling = select(created_scores,
+    data_modelling = select(filtered_data,
                             wave,
                             country,
-                            adj_pvnum_MATH,
-                            adj_pvnum_READ,
+                            math,
+                            read,
[…]

### 2019-10-21 code/02-analysis.R
Fixes weight problems

@@ -1649,7 +1649,7 @@ impute_missing <- function(all_data) {
   all_data_imput$imputations[[1]]
 }
 
-generate_models <- function(all_data, dv, group, aut_var) {
+generate_models <- function(data_modelling, dv, group, aut_var) {
 
   model_formula <-
     as.formula(
@@ -1674,18 +1674,15 @@ generate_models <- function(all_data, dv, group, aut_var) {
                        "hisei",
                        "native")
 
-  mod_df <-
-    all_data %>%
-    filter(escs_dummy == group)
-
-  mod_df <-
-    mod_df %>% mutate_at(vars(ends_with("aut")), center) %>%
+  data_modelling <-
+    data_modelling %>%
+    filter(escs_dummy == group) %>% 
+    mutate_at(vars(ends_with("aut")), center) %>%
     rename(prop_cert = PROPCERT,
            private = SCHLTYPE) %>%
-    select(all.vars(model_formula), fixed_variables) %>%
-    filter(complete.cases(.))
-  ## mutate(high_edu_broad = recode(high_edu_broad, `2` = 3))
-  ## high_edu_broad = as.character(high_edu_broad))
+    select(all.vars(model_formula), fixed_variables, stu_weight) %>%
+    filter(complete.cases(.)) %>%
+    mutate(num_stu = num_stu / 100)
 
   all_formulas <- formula_gen(model_formula)
 
@@ -1698,10 +1695,13 @@ generate_models <- function(all_data, dv, group, aut_var) {
   len <- length(all_formulas)
   all_formulas[[len + 1]] <- update(all_formulas[[len]], for_fixed)
 
-  all_mods <- map(all_formulas,
-                  ~ lmer(.x, data = mod_df,
-                         weights = mod_df[["stu_weight"]],
-                         control = lmerControl(optimizer = "Nelder_Mead")))
+  all_mods <- map(all_formulas, ~ {
+    lmer(.x,
+         data = data_modelling,
+         weights = stu_weight,
+         control = lmerControl(optimizer = "Nelder_Mead"))
+  })
+  
   all_mods
 }
 
@@ -1761,7 +1761,7 @@ create_school_dummy <- function(filtered_data, probs) {
   filtered_data
 }
 
-generate_models_schools <- function(all_data, dv, group, aut_var) {
+generate_models_schools <- function(data_modelling, dv, group, aut_var) {
 
   model_formula <-
     as.formula(
[…]

### 2019-10-21 code/plan.R

@@ -676,7 +676,7 @@ plan <-
     created_scores = create_scores(created_escs_dummy, reliability_pisa),
 
     # Only keep countries with more than 50% of data for all years
-    filtered_data = filter(created_scores, country %in% final_countries) %>% sample_frac(10),
+    filtered_data = filter(created_scores, country %in% final_countries),
 
     # These two below are the final datasets for the modelling.
 
@@ -774,7 +774,8 @@ plan <-
                                    num_stu,
                                    government_fund,
                                    hisei,
-                                   SCHLTYPE),
+                                   SCHLTYPE,
+                                   stu_weight),
     aut_allcnt = target(
       generate_models(data_modelling_allcnt,
                       dv = math_read,
@@ -807,12 +808,10 @@ plan <-
     aut_interact = target(
       generate_models_interaction(filtered_data,
                                   dv = math_read,
-                                  group = group_vals,
                                   aut_var = aut_val,
                                   interact = type_interact
                                   ),
       transform = cross(math_read = c("math", "read"),
-                        group_vals = c(0, 1),
                         aut_val = !!autonomy_measures,
                         type_interact = c("quantiles_escs", "quantiles_escs_chr"))
     ),

### 2019-10-24 _drake.R
Adds gap plots

@@ -16,7 +16,8 @@ dconf <- drake_config(
   ## See https://github.com/ropensci/drake/issues/960
   env = new.env(parent = globalenv()),
   ## lock_envir = TRUE,
-  memory_strategy = "memory"
+  memory_strategy = "memory",
+  garbage_collection = TRUE
 )
 
 source(here("code/read_raw_data.R"))

### 2019-10-24 code/02-analysis.R

@@ -299,6 +299,30 @@ harmonize_student <- function(raw_student, recode_cntrys) {
       .x
     })
 
+  raw_student$value <-
+    map(raw_student$value, ~ {
+
+      escs_col <- if (unique(.x$wave) == "pisa2015") "ESCS" else " "
+
+      .x %>%
+        select(country,
+               wave,
+               SCHOOLID,
+               stu_id,
+               ## escs_dummy,
+               starts_with("PV"),
+               contains(escs_col),
+               contains("W_FSTUWT"),
+               AGE,
+               gender,
+               high_edu_broad,
+               books_hh,
+               hisei,
+               native,
+               stu_weight)
+
+    })
+
   raw_student
 }
 
@@ -379,7 +403,7 @@ harmonize_school <- function(loaded_school) {
       mutate(SCHOOLID = as.character(SCHOOLID))
   }
 
-  # PISA 20000
+  # PISA 2000
   loaded_school$value[[1]] <-
     loaded_school$value[[1]] %>%
     as_tibble() %>%
@@ -540,7 +564,12 @@ harmonize_school <- function(loaded_school) {
 
   loaded_school$value <- map(loaded_school$value, ~ {
     .x %>%
-      mutate(SCHLTYPE = as.character(SCHLTYPE),
+      mutate(PROPCERT = ifelse(PROPCERT > 9000, NA_real_, PROPCERT),
+             SCHLTYPE = case_when(SCHLTYPE %in% c("1", "2") ~ "Private",
+                                  SCHLTYPE %in% "3" ~ "Public",
+                                  SCHLTYPE %in% c("Government", "Public") ~ "Public",
+                                  str_detect(SCHLTYPE, "^Private") ~ "Private",
+                                  TRUE ~ NA_character_),
              academic_content_aut = rowMeans(select(., course_aut, content_aut, textbook_aut)),
              personnel_aut = rowMeans(select(., hiring_aut, salary_aut))) %>% 
       mutate_all(unclass) %>%
@@ -568,16 +597,16 @@ harmonize_school <- function(loaded_school) {
   loaded_school
 }
 
-plot_autonomy <- function(harmonized_school, cntrys) {
+plot_autonomy_public <- function(harmonized_school, cntrys, autonomy_var) {
+  autonomy_var <- enquo(autonomy_var)
 
   summarize_aut <- function(df_pisa, wave) {
     df_pisa %>%
[…]

### 2019-10-24 code/plan.R

@@ -25,7 +25,6 @@ pisa2015_conf <- list(
 # Final countries used in the analysis (30)
 # which have at least 50% of all years available (6 years)
 final_countries <- c("Finland",
-                     "France",
                      "Austria",
                      "Australia",
                      "Sweden",
@@ -61,7 +60,7 @@ final_countries <- c("Finland",
 
 # Selected countries for plotting
 countries <- c("Finland",
-               "France",
+               # "France",
                # "New Zealand",
                "Austria",
                "Australia",
@@ -140,23 +139,24 @@ quantile_missing <- function(df, weights, probs, ch_var = "escs_trend") {
 
 # It returns a dataframe for each survey with all countries and respective coefficients and
 # standard errors.
-test_diff <- function(df, test) {
+calculate_gap <- function(data_modelling, test) {
+  separate_waves <- split(data_modelling, data_modelling$wave)
 
-  map(df, function(.x) {
-    conf <- if (unique(.x$wave) == "pisa2015") pisa2015_conf else pisa_conf
-    weights_var <- conf$variables$weightFinal
+  separate_gaps <- map(separate_waves, function(.x) {
+    weights_var <- "stu_weight"
 
     mod_formula <-
       as.formula(
         paste0(
-          paste0("adj_pvnum_", test), " ~ escs_dummy + (1 + escs_dummy | country)"
+          test, " ~ escs_dummy + (1 + escs_dummy | country)"
         )
       )
 
     mod2 <-
       lmer(mod_formula,
            data = .x,
-           weights = .x[[weights_var]])
+           weights = .x[[weights_var]],
+           control = lmerControl(optimizer = "Nelder_Mead"))
 
     # Take the country coefficients (absolute coefficients)
     country_coef <-
@@ -185,6 +185,8 @@ test_diff <- function(df, test) {
     message(paste0(unique(.x$wave), " modeling done"))
     list(results, mod2)
   })
+
+  separate_gaps
 }
 
 
@@ -624,94 +626,124 @@ plan <-
     # See the notes in /code/read_raw_data.R. It's called raw_data.
 
     # All read_* functions return a list with the data for each wave
-    loaded_school = load_school(raw_data_dir),
+    loaded_school = target(
+      load_school(raw_data_dir),
+      format = "fst"
+     ),
     # This is the socio-economic index harmonized to be used across all waves
-    loaded_escs = load_escs(raw_data_dir, recode_cntrys),
[…]

### 2019-10-24 paper/achievement_gap.Rmd
Adds mockup of article

---
title: Fake title
runninghead: Cimentada
author:
- name: Jorge Cimentada
  num: 1
address:
- num: 1
  org: Laboratory of Digital and Computational Demography, Max Planck Institute of Demographic Research
corrauth: "Konrad-Zuse-Str. 1. Rostock, Germany"
email: cimentada@demogr.mpg.de
abstract: "The literature on achievement inequality has recently started to focus on the dynamics of the socio-economic achievement gap in cognitive abilities. The main findings come from research in the U.S. revealing that the 90th/10th income achievement gap has widened 50% in the last 30 years. This chapter aims to investigate whether there are patterns in the evolution of the achievement gap from a comparative perspective. Using 15 years of data in 32 countries from the Program for International Student Assessment (PISA), I find that there is considerable variation in the way in which the gap between the average score of students above (and at) the 90th percentile and below (and) the 10th percentile is evolving. The prime examples come from the U.S. and Germany closing at about 50% and 30% in the last 15 years while France is widening at a similar rate. I find that curricular tracking and vocational enrollment explain 40% of the variance in the achievement gap between countries and show that the relationship is conditioned by a strong interaction. Low curricular tracking is associated with a small achievement gap, whereas high levels of curricular tracking is associated with wide achievement gaps. However, once tracking is coupled with high vocational enrollment this can remedy the potential adverse effects and reduce the gap by .6 standard deviation. I use simulations to show that switching to less curricular tracking can help decrease a country’s SES gap by about 10% while switching to more tracking would increase the achievement gap by about 51% percent."
keywords: academic achievement, education inequality, school autonomy, international comparison
classoption:
  - Royal
  - times
bibliography: bibfile
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
---

# The Article Header Information

YAML header:

```
output:
  rticles::sim_article:
    keep_tex: TRUE
```

Configure the YAML header including the following elements:

* `title`: Title

* `runninghead`: Author last names, use _et al._ if there are three or more authors.

* `author`: List of author(s) containing `name` and `num`

* `corrauth`: Corresponding author's name and address.

* `email`: Correspondence email

* `abstract`: Limited to 200 words

* `keywords`: Keywords for the article

* `bibliography`: BibTeX `.bib` file

* `bibliographystyle`: sageh or sagev

* `classoption`: options of the `sagej` class

## Remarks

2. `bibliographystyle`

3. `classoption`

4. Keywords are separated by commas.

# The Body of the Article

## Mathematics

Use mathematics in Rmarkdown as usual.

## Figures and Tables

Figures are supported from R code:

```{r plot-ref, fig.cap = "Fancy Caption\\label{fig:plot}", out.width='100%'}
x = rnorm(10)
y = rnorm(10)
plot(x, y)
```

...and can be referenced (Figure \ref{fig:plot}) by including the `\\label{}` tag in the `fig.cap` attribute of the R chunk: `fig.cap = "Fancy Caption\\label{fig:plot}"`. It is a quirky hack at the moment, see [here](https://github.com/yihui/knitr/issues/323).

Analogously, use Rmarkdown to produce tables as usual:

```{r, results = "asis"}
if (!require("xtable")) install.packages("xtable")
xt <- xtable(head(cars), caption = "A table", label = "tab:table")
print(xt, comment = FALSE)
```

Referenced via \ref{tab:table}. You can also use the YAML option `header-includes` to includes custom \LaTeX packages for tables (keep in mind that `pandoc` uses `longtables` by default, and it is hardcoded; some things may require including the package `longtable`). E.g., using `ctable`:
```
header-includes:
- \usepackage{ctable}
```
Then, just write straight-up \LaTeX code and reference is as usual (`\ref{tab:ctable}`):
```
\ctable[cap = {Short caption},
        caption = {A caption for this table.},
        label={tab:ctable},]
        {cc}
        {
        \tnote[$\ast$]{Footnote 1}
        \tnote[$\dagger$]{Other footnote}
        \tnote[b]{Mistakes are possible.}
        }{
        \FL
        COL 1\tmark[a] & COL 2\tmark[$\ast$]
        \ML
        6.92\tmark[$\dagger$] & 0.09781 \\
        6.93\tmark[$\dagger$] & 0.09901 \\
        97 & 2000
        \LL
}
```

It is also possible to set the `YAML` option `longtable: true` and use markdown tables (or the `knitr::kable` function): `knitr::kable(head(cars))` produces the same table as the `xtable` example presented before.

## Cross-referencing

The use of the Rmarkdown equivalent of the \LaTeX cross-reference system
for figures, tables, equations, etc., is encouraged (using `[@<name>]`, equivalent of `\ref{<name>}` and `\label{<name>}`). That works well for citations in Rmarkdown, not so well for figures and tables. In that case, it is possible to revert to standard \LaTeX syntax.

## Double Spacing

If you need to double space your document for submission please use the `doublespace` option in the header.

# Bibliography

Link a `.bib` document via the YAML header, and bibliography will be printed at the very end (as usual). The default bibliography style is provided by Wiley as in `WileyNJD-AMA.bst`, do not delete that file.

Use the Rmarkdown equivalent of the \LaTeX citation system using `[@<name>]`. Example: [@Taylor1937], [@Knupp1999; @Kamm2000].

To include all citation from the `.bib` file, add `\nocite{*}` before the end of the document.

# Further information

All \LaTeX enviroments supported by the main template are supported here as well; see the `.tex` sample file [here](http://onlinelibrary.wiley.com/journal/10.1002/(ISSN)1097-0258/homepage/la_tex_class_file.htm) for more details and example.

### 2019-10-24 paper/achievement_gap.tex
Adds mockup of article

\documentclass[Royal,times,sageh]{sagej}

\usepackage{moreverb,url,natbib, multirow, tabularx}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}

\begin{document}

\title{Fake title}

\runninghead{Cimentada}

\author{Jorge Cimentada\affilnum{1}}

\affiliation{\affilnum{1}{Laboratory of Digital and Computational Demography, Max Planck Institute
of Demographic Research}}

\corrauth{Konrad-Zuse-Str. 1. Rostock, Germany}

\email{\href{mailto:cimentada@demogr.mpg.de}{\nolinkurl{cimentada@demogr.mpg.de}}}

\begin{abstract}
The literature on achievement inequality has recently started to focus
on the dynamics of the socio-economic achievement gap in cognitive
abilities. The main findings come from research in the U.S. revealing
that the 90th/10th income achievement gap has widened 50\% in the last
30 years. This chapter aims to investigate whether there are patterns in
the evolution of the achievement gap from a comparative perspective.
Using 15 years of data in 32 countries from the Program for
International Student Assessment (PISA), I find that there is
considerable variation in the way in which the gap between the average
score of students above (and at) the 90th percentile and below (and) the
10th percentile is evolving. The prime examples come from the U.S. and
Germany closing at about 50\% and 30\% in the last 15 years while France
is widening at a similar rate. I find that curricular tracking and
vocational enrollment explain 40\% of the variance in the achievement
gap between countries and show that the relationship is conditioned by a
strong interaction. Low curricular tracking is associated with a small
achievement gap, whereas high levels of curricular tracking is
associated with wide achievement gaps. However, once tracking is coupled
with high vocational enrollment this can remedy the potential adverse
effects and reduce the gap by .6 standard deviation. I use simulations
to show that switching to less curricular tracking can help decrease a
country's SES gap by about 10\% while switching to more tracking would
increase the achievement gap by about 51\% percent.
\end{abstract}

\keywords{academic achievement, education inequality, school autonomy,
international comparison}

\maketitle

\hypertarget{the-article-header-information}{%
\section{The Article Header
Information}\label{the-article-header-information}}

YAML header:

\begin{verbatim}
output:
  rticles::sim_article:
    keep_tex: TRUE
\end{verbatim}

Configure the YAML header including the following elements:

\begin{itemize}
\item
  \texttt{title}: Title
\item
  \texttt{runninghead}: Author last names, use \emph{et al.} if there
  are three or more authors.
\item
  \texttt{author}: List of author(s) containing \texttt{name} and
  \texttt{num}
\item
  \texttt{corrauth}: Corresponding author's name and address.
\item
  \texttt{email}: Correspondence email
\item
  \texttt{abstract}: Limited to 200 words
\item
  \texttt{keywords}: Keywords for the article
\item
  \texttt{bibliography}: BibTeX \texttt{.bib} file
\item
  \texttt{bibliographystyle}: sageh or sagev
\item
  \texttt{classoption}: options of the \texttt{sagej} class
\end{itemize}

\hypertarget{remarks}{%
\subsection{Remarks}\label{remarks}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\item
  \texttt{bibliographystyle}
\item
  \texttt{classoption}
\item
  Keywords are separated by commas.
\end{enumerate}

\hypertarget{the-body-of-the-article}{%
\section{The Body of the Article}\label{the-body-of-the-article}}

\hypertarget{mathematics}{%
\subsection{Mathematics}\label{mathematics}}

Use mathematics in Rmarkdown as usual.

\hypertarget{figures-and-tables}{%
\subsection{Figures and Tables}\label{figures-and-tables}}

Figures are supported from R code:

\begin{verbatim}
x = rnorm(10)
y = rnorm(10)
plot(x, y)
\end{verbatim}

\begin{figure}
\includegraphics[width=1\linewidth]{achievement_gap_files/figure-latex/plot-ref-1} \caption{Fancy Caption\label{fig:plot}}\label{fig:plot-ref}
\end{figure}

\ldots and can be referenced (Figure \ref{fig:plot}) by including the
\texttt{\textbackslash{}\textbackslash{}label\{\}} tag in the
\texttt{fig.cap} attribute of the R chunk:
\texttt{fig.cap\ =\ "Fancy\ Caption\textbackslash{}\textbackslash{}label\{fig:plot\}"}.
It is a quirky hack at the moment, see
\href{https://github.com/yihui/knitr/issues/323}{here}.

Analogously, use Rmarkdown to produce tables as usual:

\begin{verbatim}
if (!require("xtable")) install.packages("xtable")
\end{verbatim}

\begin{verbatim}
## Loading required package: xtable
\end{verbatim}

\begin{verbatim}
xt <- xtable(head(cars), caption = "A table", label = "tab:table")
print(xt, comment = FALSE)
\end{verbatim}

\begin{table}[ht]
\centering
\begin{tabular}{rrr}
  \hline
 & speed & dist \\ 
  \hline
1 & 4.00 & 2.00 \\ 
  2 & 4.00 & 10.00 \\ 
  3 & 7.00 & 4.00 \\ 
  4 & 7.00 & 22.00 \\ 
  5 & 8.00 & 16.00 \\ 
  6 & 9.00 & 10.00 \\ 
   \hline
\end{tabular}
\caption{A table} 
\label{tab:table}
\end{table}

Referenced via \ref{tab:table}. You can also use the YAML option
\texttt{header-includes} to includes custom \LaTeX packages for tables
(keep in mind that \texttt{pandoc} uses \texttt{longtables} by default,
and it is hardcoded; some things may require including the package
\texttt{longtable}). E.g., using \texttt{ctable}:

\begin{verbatim}
header-includes:
- \usepackage{ctable}
\end{verbatim}

Then, just write straight-up \LaTeX code and reference is as usual
(\texttt{\textbackslash{}ref\{tab:ctable\}}):

\begin{verbatim}
\ctable[cap = {Short caption},
        caption = {A caption for this table.},
        label={tab:ctable},]
        {cc}
        {
        \tnote[$\ast$]{Footnote 1}
        \tnote[$\dagger$]{Other footnote}
        \tnote[b]{Mistakes are possible.}
        }{
        \FL
        COL 1\tmark[a] & COL 2\tmark[$\ast$]
        \ML
        6.92\tmark[$\dagger$] & 0.09781 \\
        6.93\tmark[$\dagger$] & 0.09901 \\
        97 & 2000
        \LL
}
\end{verbatim}

It is also possible to set the \texttt{YAML} option
\texttt{longtable:\ true} and use markdown tables (or the
\texttt{knitr::kable} function): \texttt{knitr::kable(head(cars))}
produces the same table as the \texttt{xtable} example presented before.

\hypertarget{cross-referencing}{%
\subsection{Cross-referencing}\label{cross-referencing}}

The use of the Rmarkdown equivalent of the \LaTeX cross-reference system
for figures, tables, equations, etc., is encouraged (using
\texttt{{[}@\textless{}name\textgreater{}{]}}, equivalent of
\texttt{\textbackslash{}ref\{\textless{}name\textgreater{}\}} and
\texttt{\textbackslash{}label\{\textless{}name\textgreater{}\}}). That
works well for citations in Rmarkdown, not so well for figures and
tables. In that case, it is possible to revert to standard
\LaTeX syntax.

\hypertarget{double-spacing}{%
\subsection{Double Spacing}\label{double-spacing}}

If you need to double space your document for submission please use the
\texttt{doublespace} option in the header.

\hypertarget{bibliography}{%
\section{Bibliography}\label{bibliography}}

Link a \texttt{.bib} document via the YAML header, and bibliography will
be printed at the very end (as usual). The default bibliography style is
provided by Wiley as in \texttt{WileyNJD-AMA.bst}, do not delete that
file.

Use the Rmarkdown equivalent of the \LaTeX citation system using
\texttt{{[}@\textless{}name\textgreater{}{]}}. Example:
\citep{Taylor1937}, \citep{Knupp1999, Kamm2000}.

To include all citation from the \texttt{.bib} file, add
\texttt{\textbackslash{}nocite\{*\}} before the end of the document.

\hypertarget{further-information}{%
\section{Further information}\label{further-information}}

All \LaTeX enviroments supported by the main template are supported here
as well; see the \texttt{.tex} sample file
\href{http://onlinelibrary.wiley.com/journal/10.1002/(ISSN)1097-0258/homepage/la_tex_class_file.htm}{here}
for more details and example.

\bibliographystyle{sageh}
\bibliography{bibfile}

\end{document}

### 2019-10-24 code/02-analysis.R

@@ -1195,7 +1195,6 @@ plot_evolution_gaps <- function(complete_data_topbottom) {
     complete_data_topbottom %>%
     filter(type_test == "math") %>%
     dplyr::select(wave, country, difference) %>%
-    filter(country %in% countries) %>%
     split(.$country) %>%
     map(lm_data) %>%
     enframer("country") %>%
@@ -1285,7 +1284,7 @@ perc_increase_fun <- function(df) {
   data_ready
 }
 
-perc_graph <- function(df, test, title, subtitle = NULL, ordered_cnt) {
+plot_perc_change <- function(df, test, title, subtitle = NULL, ordered_cnt) {
   df %>%
     enframe(name = "country") %>%
     mutate(value = map(value, ~ dplyr::select(.x, -country))) %>% 
@@ -1316,10 +1315,28 @@ perc_graph <- function(df, test, title, subtitle = NULL, ordered_cnt) {
     theme(axis.text.x = element_text(angle = 45, hjust = 1))
 }
 
-evolution_ses_groups <- function(complete_data_topbottom, ordered_cnt) {
-  # You left off here. Plot the gap for 1 and 0 for all countries in the graph before.
+plot_evolution_ses <- function(complete_data_topbottom) {
+
+  lm_data <- function(df) {
+    lm(log(difference) ~ wave, data = df) %>%
+      broom::tidy() %>%
+      mutate(estimate = exp(estimate))
+  }
+
+
+  ordered_cnt <-
+    complete_data_topbottom %>%
+    filter(type_test == "math") %>%
+    dplyr::select(wave, country, difference) %>%
+    split(.$country) %>%
+    map(lm_data) %>%
+    enframer("country") %>%
+    filter(term == "wave") %>%
+    arrange(-estimate) %>%
+    pull(country)
+
+
   complete_data_topbottom %>%
-    filter(country %in% ordered_cnt) %>%
     mutate(country = factor(country, levels = ordered_cnt, ordered = TRUE)) %>%
     dplyr::select(wave, country, matches("*._math$")) %>%
     gather(ses, gap_size, matches("^\\d_mean_math$")) %>%
@@ -1332,17 +1349,20 @@ evolution_ses_groups <- function(complete_data_topbottom, ordered_cnt) {
     geom_line(stat = "smooth", method = "lm", aes(group = 1),
               formula = y ~ splines::ns(x, 1), linetype = "longdash",
[…]

### 2019-10-24 code/plan.R

@@ -28,7 +28,6 @@ final_countries <- c("Finland",
                      "Austria",
                      "Australia",
                      "Sweden",
-                     "Slovakia",
                      "Czech Republic",
                      "Canada",
                      "Hungary",
@@ -215,7 +214,7 @@ f_ind <- function(.x) .x[[1]]
 enframer <- function(df, col_name = "name") {
   df %>%
     enframe(name = col_name) %>%
-    unnest()
+    unnest(cols = value)
 }
 
 print_table <- function(x, ...) {
@@ -721,13 +720,13 @@ plan <-
     ############################################################################
 
     plotted_academic_aut = plot_autonomy_public(harmonized_school,
-                                                countries,
+                                                final_countries,
                                                 academic_content_aut),
     plotted_personnel_aut = plot_autonomy_public(harmonized_school,
-                                                 countries,
+                                                 final_countries,
                                                  personnel_aut),
     plotted_budget_aut = plot_autonomy_public(harmonized_school,
-                                              countries,
+                                              final_countries,
                                               budget_aut),
 
     calculated_corr_aut = calculate_corr_aut(harmonized_school),
@@ -737,13 +736,14 @@ plan <-
 
     p1_evolution_gaps = plot_evolution_gaps(complete_data_topbottom),
     top_bottom_perc = perc_increase_fun(complete_data_topbottom),
-    p2_perc_change = perc_graph(
+    p2_perc_change = plot_perc_change(
       top_bottom_perc,
       "math",
       "90/10 achievement gap",
       "Percentage change from 2000 to 2015",
       countries
     ),
+    p3_evolution_ses = plot_evolution_ses(complete_data_topbottom),
 
     ############################# Modelling ####################################
     ############################################################################

### 2019-10-24 paper/achievement_gap.Rmd
Starts adding plots to paper

@@
 
-# The Article Header Information
-
-YAML header:
-
-```
-output:
-  rticles::sim_article:
-    keep_tex: TRUE
+```{r }
+knitr::opts_chunk$set(
+  warning = FALSE,
+  message = FALSE,
+  echo = FALSE,
+  ## fig.width = 6,
+  ## fig.asp = 0.618,
+  out.width = "70%",
+  fig.align = "center",
+  progress = FALSE,
+  verbose = FALSE,
+  results = "asis"
+)
 ```
 
-Configure the YAML header including the following elements:
-
-* `title`: Title
-
-* `runninghead`: Author last names, use _et al._ if there are three or more authors.
-
-* `author`: List of author(s) containing `name` and `num`
-
-* `corrauth`: Corresponding author's name and address.
-
-* `email`: Correspondence email
-
-* `abstract`: Limited to 200 words
-
-* `keywords`: Keywords for the article
-
-* `bibliography`: BibTeX `.bib` file
-
-* `bibliographystyle`: sageh or sagev
-
-* `classoption`: options of the `sagej` class
-
-## Remarks
-
-2. `bibliographystyle`
-
-3. `classoption`
-
-4. Keywords are separated by commas.
-
 # The Body of the Article
-
-## Mathematics
-
-Use mathematics in Rmarkdown as usual.
-
-## Figures and Tables
-
-Figures are supported from R code:
-
-```{r plot-ref, fig.cap = "Fancy Caption\\label{fig:plot}", out.width='100%'}
-x = rnorm(10)
-y = rnorm(10)
-plot(x, y)
-```
-
-...and can be referenced (Figure \ref{fig:plot}) by including the `\\label{}` tag in the `fig.cap` attribute of the R chunk: `fig.cap = "Fancy Caption\\label{fig:plot}"`.
-It is a quirky hack at the moment, see [here](https://github.com/yihui/knitr/issues/323).
-
-Analogously, use Rmarkdown to produce tables as usual:
-
-```{r, results = "asis"}
-if (!require("xtable")) install.packages("xtable")
-xt <- xtable(head(cars), caption = "A table", label = "tab:table")
-print(xt, comment = FALSE)
-```
-
-Referenced via \ref{tab:table}.
-You can also use the YAML option `header-includes` to includes custom \LaTeX packages for tables (keep in mind that `pandoc` uses `longtables` by default, and it is hardcoded; some things may require including the package `longtable`).
-E.g., using `ctable`:
-```
-header-includes:
-- \usepackage{ctable}
-```
-Then, just write straight-up \LaTeX code and reference is as usual (`\ref{tab:ctable}`):
-```
-\ctable[cap = {Short caption},
-        caption = {A caption for this table.},
-        label={tab:ctable},]
-        {cc}
-        {
-        \tnote[$\ast$]{Footnote 1}
-        \tnote[$\dagger$]{Other footnote}
-        \tnote[b]{Mistakes are possible.}
-        }{
-        \FL
-        COL 1\tmark[a] & COL 2\tmark[$\ast$]
-        \ML
-        6.92\tmark[$\dagger$] & 0.09781 \\
-        6.93\tmark[$\dagger$] & 0.09901 \\
-        97 & 2000
-        \LL
-}
-```
-
-It is also possible to set the `YAML` option `longtable: true` and use markdown tables (or the `knitr::kable` function): `knitr::kable(head(cars))` produces the same table as the `xtable` example presented before.
-
-## Cross-referencing
-
-The use of the Rmarkdown equivalent of the \LaTeX cross-reference system
-for figures, tables, equations, etc., is encouraged (using `[@<name>]`, equivalent of `\ref{<name>}` and `\label{<name>}`).
-That works well for citations in Rmarkdown, not so well for figures and tables.
-In that case, it is possible to revert to standard \LaTeX syntax.
-
-## Double Spacing
-
-If you need to double space your document for submission please use the `doublespace` option in the header.
 
@@
 
+```{r, fig.height = 90, fig.width = 60, out.width = "95%"}
+p1_plot <- readd(p1_evolution_gaps)
+p1_plot
+```
+
 # Further information

### 2019-10-24 paper/achievement_gap.tex
Starts adding plots to paper

@@
 
-\hypertarget{the-article-header-information}{%
-\section{The Article Header
-Information}\label{the-article-header-information}}
-
-YAML header:
-
 \begin{verbatim}
-output:
-  rticles::sim_article:
-    keep_tex: TRUE
+knitr::opts_chunk$set(
+  warning = FALSE,
+  message = FALSE,
+  echo = FALSE,
+  ## fig.width = 6,
+  ## fig.asp = 0.618,
+  out.width = "70%",
+  fig.align = "center",
+  progress = FALSE,
+  verbose = FALSE,
+  results = "asis"
+)
 \end{verbatim}
-
-Configure the YAML header including the following elements:
-
-\begin{itemize}
-\item
-  \texttt{title}: Title
-\item
-  \texttt{runninghead}: Author last names, use \emph{et al.} if there
-  are three or more authors.
-\item
-  \texttt{author}: List of author(s) containing \texttt{name} and
-  \texttt{num}
-\item
-  \texttt{corrauth}: Corresponding author's name and address.
-\item
-  \texttt{email}: Correspondence email
-\item
-  \texttt{abstract}: Limited to 200 words
-\item
-  \texttt{keywords}: Keywords for the article
-\item
-  \texttt{bibliography}: BibTeX \texttt{.bib} file
-\item
-  \texttt{bibliographystyle}: sageh or sagev
-\item
-  \texttt{classoption}: options of the \texttt{sagej} class
-\end{itemize}
-
-\hypertarget{remarks}{%
-\subsection{Remarks}\label{remarks}}
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{1}
-\item
-  \texttt{bibliographystyle}
-\item
-  \texttt{classoption}
-\item
-  Keywords are separated by commas.
-\end{enumerate}
 
@@
 \section{The Body of the Article}\label{the-body-of-the-article}}
-
-\hypertarget{mathematics}{%
-\subsection{Mathematics}\label{mathematics}}
-
-Use mathematics in Rmarkdown as usual.
-
-\hypertarget{figures-and-tables}{%
-\subsection{Figures and Tables}\label{figures-and-tables}}
-
-Figures are supported from R code:
-
-\begin{verbatim}
-x = rnorm(10)
-y = rnorm(10)
-plot(x, y)
-\end{verbatim}
-
-\begin{figure}
-\includegraphics[width=1\linewidth]{achievement_gap_files/figure-latex/plot-ref-1} \caption{Fancy Caption\label{fig:plot}}\label{fig:plot-ref}
-\end{figure}
-
-\ldots and can be referenced (Figure \ref{fig:plot}) by including the
-\texttt{\textbackslash{}\textbackslash{}label\{\}} tag in the
-\texttt{fig.cap} attribute of the R chunk:
-\texttt{fig.cap\ =\ "Fancy\ Caption\textbackslash{}\textbackslash{}label\{fig:plot\}"}.
-It is a quirky hack at the moment, see
-\href{https://github.com/yihui/knitr/issues/323}{here}.
-
-Analogously, use Rmarkdown to produce tables as usual:
-
-\begin{verbatim}
-if (!require("xtable")) install.packages("xtable")
-\end{verbatim}
-
-\begin{verbatim}
-## Loading required package: xtable
-\end{verbatim}
-
-\begin{verbatim}
-xt <- xtable(head(cars), caption = "A table", label = "tab:table")
-print(xt, comment = FALSE)
-\end{verbatim}
-
-\begin{table}[ht]
-\centering
-\begin{tabular}{rrr}
-  \hline
- & speed & dist \\ 
-  \hline
-1 & 4.00 & 2.00 \\ 
-  2 & 4.00 & 10.00 \\ 
-  3 & 7.00 & 4.00 \\ 
-  4 & 7.00 & 22.00 \\ 
-  5 & 8.00 & 16.00 \\ 
-  6 & 9.00 & 10.00 \\ 
-   \hline
-\end{tabular}
-\caption{A table} 
-\label{tab:table}
-\end{table}
-
-Referenced via \ref{tab:table}. You can also use the YAML option
-\texttt{header-includes} to includes custom \LaTeX packages for tables
-(keep in mind that \texttt{pandoc} uses \texttt{longtables} by default,
-and it is hardcoded; some things may require including the package
-\texttt{longtable}). E.g., using \texttt{ctable}:
-
-\begin{verbatim}
-header-includes:
-- \usepackage{ctable}
-\end{verbatim}
-
-Then, just write straight-up \LaTeX code and reference is as usual
-(\texttt{\textbackslash{}ref\{tab:ctable\}}):
-
-\begin{verbatim}
-\ctable[cap = {Short caption},
-        caption = {A caption for this table.},
-        label={tab:ctable},]
-        {cc}
-        {
-        \tnote[$\ast$]{Footnote 1}
-        \tnote[$\dagger$]{Other footnote}
-        \tnote[b]{Mistakes are possible.}
-        }{
-        \FL
-        COL 1\tmark[a] & COL 2\tmark[$\ast$]
-        \ML
-        6.92\tmark[$\dagger$] & 0.09781 \\
-        6.93\tmark[$\dagger$] & 0.09901 \\
-        97 & 2000
-        \LL
-}
-\end{verbatim}
-
-It is also possible to set the \texttt{YAML} option
-\texttt{longtable:\ true} and use markdown tables (or the
-\texttt{knitr::kable} function): \texttt{knitr::kable(head(cars))}
-produces the same table as the \texttt{xtable} example presented before.
-
-\hypertarget{cross-referencing}{%
-\subsection{Cross-referencing}\label{cross-referencing}}
-
-The use of the Rmarkdown equivalent of the \LaTeX cross-reference system
-for figures, tables, equations, etc., is encouraged (using
-\texttt{{[}@\textless{}name\textgreater{}{]}}, equivalent of
-\texttt{\textbackslash{}ref\{\textless{}name\textgreater{}\}} and
-\texttt{\textbackslash{}label\{\textless{}name\textgreater{}\}}). That
-works well for citations in Rmarkdown, not so well for figures and
-tables. In that case, it is possible to revert to standard
-\LaTeX syntax.
-
-\hypertarget{double-spacing}{%
-\subsection{Double Spacing}\label{double-spacing}}
-
-If you need to double space your document for submission please use the
-\texttt{doublespace} option in the header.
 
@@
 
+\begin{center}\includegraphics[width=0.95\linewidth]{achievement_gap_files/figure-latex/unnamed-chunk-2-1} \end{center}
+
 \hypertarget{further-information}{%

### 2019-10-24 code/01-packages.R

@@ -1,5 +1,6 @@
 library(car)
 library(drake)
+library(rmarkdown)
 library(here)
 library(xtable)
 library(patchwork) # devtools::install_github("thomasp85/patchwork")

### 2019-10-24 code/02-analysis.R

@@ -1190,7 +1190,6 @@ plot_evolution_gaps <- function(complete_data_topbottom) {
       mutate(estimate = exp(estimate))
   }
 
-
   ordered_cnt <-
     complete_data_topbottom %>%
     filter(type_test == "math") %>%
@@ -1204,6 +1203,7 @@ plot_evolution_gaps <- function(complete_data_topbottom) {
 
   diff_data <-
     diff_increase_fun(complete_data_topbottom) %>%
+    map(select, -country) %>% 
     enframer("country") %>%
     filter(country %in% ordered_cnt) %>%
     dplyr::select(country, type_test, diff, contains("math")) %>%
@@ -1218,20 +1218,21 @@ plot_evolution_gaps <- function(complete_data_topbottom) {
   pooled_trendline %>%
     filter(type_test == "math") %>%
     ggplot(aes(as.character(wave), difference)) +
-    geom_point() +
-    geom_linerange(aes(ymin = 0, ymax = difference)) +
-    geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.3) +
+    ## geom_point(size = 20) +
+    ## geom_linerange(aes(ymin = 0, ymax = difference)) +
+    ## geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.3) +
     geom_line(data = pooled_trendline, stat = "smooth", method = "lm", aes(group = 1),
               formula = y ~ splines::ns(x, 2), size = 0.7,
-              colour = "red") +
+              colour = "red",
+              size = 8) +
     facet_wrap(~ country, ncol = 5) +
-    scale_y_continuous(name = "90/10 gap in SD", expand = c(0, 0), lim = c(0, 3)) +
-    scale_x_discrete(name = NULL, breaks = c(2000, 2009, 2015)) +
+    scale_y_continuous(name = "Average Difference in Standardized Test Scores \n Between the 90th/10th Percentiles in Mathematics", expand = c(0, 0), lim = c(0, 3)) +
+    scale_x_discrete(name = NULL, breaks = seq(2000, 2015, 3)) +
     theme_few() +
-    theme(panel.spacing = unit(1, "lines"),
-          panel.grid.major.y = element_line(colour = "grey")) +
-    ggtitle("Evolution of the 90/10 achievement gap")
-
+    theme(panel.grid.major.y = element_line(size = 0.1, color = "grey"),
+          axis.text.x = element_text(size = rel(9), angle = 45, hjust = 1),
[…]

### 2019-10-24 code/plan.R

@@ -827,6 +827,11 @@ plan <-
                              ),
       transform = cross(math_read = c("math", "read"),
                         group_vals = c(0, 1))
+    ),
+    paper = render(
+      knitr_in(here("paper/achievement_gap.Rmd")),
+      output_file = file_out(here("paper/achievement_gap.pdf")),
+      quiet = TRUE
     )
 
     ##   sample_tables_topbottom = sample_size_calc(

### 2020-02-04 paper/achievement_gap.Rmd
Updates paper for EPC

@@
 ---
-title: Fake title
-runninghead: Cimentada
-author:
-- name: Jorge Cimentada
-  num: 1
-address:
-- num: 1
-  org: Laboratory of Digital and Computational Demography, Max Planck Institute of Demographic Research
-corrauth: "Konrad-Zuse-Str. 1. Rostock, Germany"
-email: cimentada@demogr.mpg.de
-abstract: "The literature on achievement inequality has recently started to focus on the dynamics of the socio-economic achievement gap in cognitive abilities.
+title: 'A New Perspective On The International Achievement Gap: Is Academic Autonomy Good For Everyone?'
+author: Jorge Cimentada
+date: "`r format(Sys.time(), '%d %B, %Y')`"
+header-includes:
+  - \usepackage{pdflscape}
+  - \newcommand{\blandscape}{\begin{landscape}}
+  - \newcommand{\elandscape}{\end{landscape}}
+bibliography: bibfile.bib
+output:
+  pdf_document:
+    number_sections: true
+abstract: There is a growing literature and interest on the study of the cognitive achievement gap between the top and bottom SES groups.
+Amidst public concern for this distancing between social classes, researchers have been unable to find an adequate explanation for the increasing cross-country inequality.
+In this paper, I argue that we need to refocus our efforts towards understanding better what correlates with the academic performance of both SES groups separately.
+By shifting attention to the amount of school autonomy that different schools have, I show that school autonomy over academic content, courses and text books is associated with a decrease of test scores of nearly .4 standard deviations for the bottom 10% performers in mathematics and literacy -- a whole grade's worth of knowledge.
+I show that this relationship holds under several specifications, including fixed effect models.
+In contrast, the same relationship turns positive when relating to the top 10% of students but it's much weaker than for the bottom performers.
+These results point out that perhaps an explanation to the changing gaps is not symmetrical between groups but rather group specific.
+The importance of understanding what affects separate SES groups is paramount to understanding the achievement gap and these preliminary results can have important implications in policy making as they speak directly to education policy makers trying to fine tune the autonomy measures of their country.
+---
+
+<!-- --- -->
+<!-- title: Fake title -->
+<!-- runninghead: Cimentada -->
+<!-- author: -->
+<!-- - name: Jorge Cimentada -->
+<!--   num: 1 -->
+<!-- address: -->
+<!-- - num: 1 -->
+<!--   org: Laboratory of Digital and Computational Demography, Max Planck Institute of Demographic Research -->
+<!-- corrauth: "Konrad-Zuse-Str. 1. Rostock, Germany" -->
+<!-- email: cimentada@demogr.mpg.de -->
+<!-- abstract: "The literature on achievement inequality has recently started to focus on the dynamics of the socio-economic achievement gap in cognitive abilities.
 The main findings come from research in the U.S.
@@
 However, once tracking is coupled with high vocational enrollment this can remedy the potential adverse effects and reduce the gap by .6 standard deviation.
-I use simulations to show that switching to less curricular tracking can help decrease a country’s SES gap by about 10% while switching to more tracking would increase the achievement gap by about 51% percent."
-keywords: academic achievement, education inequality, school autonomy, international comparison
-classoption:
-  - Royal
-  - times
-bibliography: bibfile
-bibliographystyle: sageh
-output:
-  rticles::sage_article:
-    keep_tex: yes
----
+I use simulations to show that switching to less curricular tracking can help decrease a country’s SES gap by about 10% while switching to more tracking would increase the achievement gap by about 51% percent." -->
+<!-- keywords: academic achievement, education inequality, school autonomy, international comparison -->
+<!-- classoption: -->
+<!--   - Royal -->
+<!--   - times -->
+<!-- bibliography: bibfile -->
+<!-- bibliographystyle: sageh -->
+<!-- output: -->
+<!--   rticles::sage_article: -->
+<!--     keep_tex: yes -->
+<!-- --- -->
 
-```{r }
+```{r, echo = FALSE}
 knitr::opts_chunk$set(
@@
   echo = FALSE,
-  ## fig.width = 6,
+  fig.width = 6,
   ## fig.asp = 0.618,
@@
 )
+
 ```
 
-# The Body of the Article
+# Background
+## The National And International SES Achievement Gap
 
-# Bibliography
+The cognitive achievement gap between the most and least advantaged children is growing steeply over time [@chmielewskiglobal2019].
+This pattern of increasing inequality between SES groups has already been investigated in countries such as the United States [@reardon2011] and other developed countries [@bradbury2015] but there is still much that is not known on the causes and correlates of the evolution of the SES achievement gap.
+This relationship is of particular importance for policy making as it can help understand the mechanisms under which different SES groups are faring better or worse under increasing economic and social inequality.
 
-Link a `.bib` document via the YAML header, and bibliography will be printed at the very end (as usual).
-The default bibliography style is provided by Wiley as in `WileyNJD-AMA.bst`, do not delete that file.
+Most of the literature on achievement gaps has concentrated on comparing the magnitude of the differences between and within countries on the cognitive achievement gap [@micklewright2006; @reardon2011; @vandenberge2006].
+This literature has shown that, similarly to the income distribution gap [@alvaredo2017; @milanovic2016], the SES cognitive achievement gap is also drastically different between countries.
+For example, evidence points out that the United States has the biggest achievement gap of all countries with a total of 1.25 standard deviations and Iceland has the narrowest gap with a gap centered at around 0.75 standard deviations [@reardon2016].
+Amidst this growing concern, researchers have tried to focus on trying to explain why this gap is so different between countries.
+In particular, many have payed attention to inequality indicators as there seems to be some relationship with income inequality [@chmielewski2016] as well as with social inequality [@duru2005].
+Others have discarded explanations such as the curricular setup of a country and the level of segregation given the weak correlations to the cross-country differences [@duru2005].
 
-Use the Rmarkdown equivalent of the \LaTeX citation system using `[@<name>]`. Example: [@Taylor1937], [@Knupp1999; @Kamm2000].
+However, nearly all of the previous studies focus on the cross-national comparison of the achievement gap.
+In recent years attention has shifted to studying the trends in achievement gap rather than solely focusing on the static SES gaps.
+The first attempts to study the *evolution* of the achievement gap was done in the United States and it documented that the gap in cognitive abilities between high-SES and low-SES children has been widening over the years [@reardon2011].
+Using over 60 years of data on educational testing surveys, @reardon2011 found that not only has the cognitive gap between the 90th income percentile and the 10th income percentile grown over time, but it has grown faster and to be wider than the highly contested white-black achievement gap [@magnuson2008].
+According to his findings, these gaps have actually reversed and we find that the income achievement gap is nearly twice as large as the black-white achievement gap (quite the opposite to 20 years back).
+Interestingly, the widening of the achievement gap has been paralleled by a growth of income inequality, which may be telling.
+@reardon2011 offers several possible links, with the most reasonable being that family investment patterns have changed so that high income families now invest more resources on their children.
+The explanation lies in the fact that increasing income became more strongly correlated with other positive family traits related to time allocation and welfare services.
 
-To include all citation from the `.bib` file, add `\nocite{*}` before the end of the document.
+In a follow-up study, @reardon2016 uncovered a reversal of the trend.
+The follow-up study concentrated solely on kindergarten children in the U.S.
+for the years 1998, 2006 and 2010.
+They found that the 90th/10th income gap in readiness closed modestly.
+Furthermore, using data from fall and spring in the same kindergarten year, they calculated that the gap narrowed at a rate of 0.01 and 0.008 SD per year for mathematics and literacy between 1998 and 2010.
+They also calculated the same changes for a number of personality traits such as self-control and externalizing behavior and found similar results.
+In contrast, @reardon2011 finds that in a 30-year span the gap was systematically increasing at a rate of 0.02, something reasonably close to the previous estimates.
+Their results not only hold for the income achievement gap, but they also found a decline in the white-hispanic gap (although not for the white-black gap).
+The reasons why the authors find a reversal in the trend could be numerous and should be studied closely.
+They discuss a number of country-level indicators to explain this change and suggest that the reversal is likely due to the high increase of preschool enrollment from the low SES group.
+They build on their previous argument by suggesting that in this same period (1998 - 2010) the income achievement gap in early schooling enrollment decreased substantially.
+Their conclusions, although suggestive, are speculative and have no \emph{empirical support} which is why this is still an open question.
 
-```{r, fig.height = 90, fig.width = 60, out.width = "95%"}
-p1_plot <- readd(p1_evolution_gaps)
+There have been other attempts to explain the achievement gaps with indicators such as economic inequality [@dupriez2006], the difference in schooling hours and the tracking system [@duru2005; @dupriez2006], home and family factors [@marks2006] and expanding school access [@chmielewskiglobal2019].
+The work of @dupriez2006 explored the relationship between achievement gaps and economic inequality but without factoring in the multilevel structure of the students nested into schools.
+Moreover, it merely correlated achievement gaps with economic inequality.
+The work of @duru2005]is more comprehensive as it explores several indicators of the school system, among which is the differentiation structure of the secondary school system (tracking).
+However, as noted by @reardon2008, \emph{'our understanding of the causes and patterns of these achievement gaps is far from complete'}.
+For this reason, the review by @werfhorst_mijs gains particular relevance because it documents many instances in which tracking explains inequality between schools (one notable example is the work of @dupriez2006 which finds a strong correlation between tracking and achievement gaps).
+
+Motivated by these recent results, other authors have taken this analysis to an international context in order to discover between-country trends.
+The work of @bradbury2015 employs a unique comparative analysis of the achievement gap between Australia, United Kingdom, United States and Canada.
+Their research design is distinctive in that they use longitudinal data from children as early as age 2 and study the evolution of the achievement gap up until age 14 \footnote{To the best of my knowledge this is not only the first study that uses panel data to study achievement gaps, but to also do it between countries}.
+The core finding of their study is that the American achievement gap is much wider than the gaps in Australia and Canada.
+They find that once the achievement gap is present in early school entry, it does not seem to narrow or widen much over the life course.
+In fact, they estimate that the quality of early childhood education can only explain about 30-40\% of the high school SES gap.
+This suggests that once the achievement gap is present before entering school, it carries a social-scar effect ^[However, schooling could be preventing the gap from widening even more, and rigorous Randomized Controlled Trials (RCT) show that high quality schooling can indeed help ease the gap, in some instances even close it [@campbell2002]].
+One exception is the UK, which they found to be a country that helps close the gap in early primary years.
+This can likely be due to the comprehensive schooling and also the public support by the welfare state in dimensions like health and income support.
+
+In a similar line but using data on more than 80 countries, @chmielewskiglobal2019 compares data from 30 international large-scale assessments over 50 years of data to study the global evolution of the achievement gap.
+Despite big cross-national variation in the evolution of the achievement gap, there is a widespread trend of increasing achievement gaps for the three SES proxies used in the study: parent's education, parent's occupation and books in the household.
+In fact, the hard numbers point out that the achievement gap between the three indicators increased at a rate of 0.007-0.008 standard deviations per year.
+These increases add up to a total of .4 standard deviations in the time span of the study, a sizable increase in the gap.
+It's also reassuring as the average yearly increase of @chmielewskiglobal2019 matches the magnitude of the point estimates from @reardon2011 and @reardon2016 which are around 0.01 and 0.02.
+
+However, one drawback of these over time cross-country analysis is that they adjust for age as most of the studies they use come from children at different stages in their school trajectory.
+However, there is evidence which suggests that achievement gaps are indeed exarcebated differently at different time points in the school trajectory [@hanushek_woesmann_tracking; @werfhorst_mijs; @bradbury2015].
+This means that most of these studies mask age-specific gaps in exchange for overall yearly gaps for each country separately.
+Among the few studies which concentrate on age-specific gaps, there are already quite different results from the main ones discussed in @chmielewskiglobal2019 and @broer2019.
+For example, @reardon2016 and @hanushek2019 find a decrease in the gap and a static gap when focusing in a narrow set of ages^[@hanushek2019 use the Long-Term Trend National Assessment of Educational Progress, the Main National Assessment of Educational Progress and the Programme for International Student Assessment but focus almost exclusively on children between ages 13 and 15].
+
+Age-specific gaps are indeed quite different as they reflect the overall distance between SES groups at the same point in time in different years (achievement gap for 15 year olds in many different years).
+These age-specific gaps has at least one benefit relative to age pooled trend analysis.
+It allows to compare students at the same time point, holding constant any differences which come about with age (in pooled analysis, age is controlled for but not all of the associated differences to inequality across the lifespan, such as for example, the stronger effect of inequality at earlier ages [@kulic2019]).
+By focusing on one age gap, students are at a point where the intensity of inequality is absored similarly by everyone due to the same age in the life-cycle skill formation [@cunha2006].
+
+As can be seen from the evidence here, there is no clear consensus on what might be driving these within country changes.
+But even more, we're far from establishing some general explanations for the between country differences in the gap.
+Instead of focusing on a single or joint explanation for the evolution of the achievement gap, this paper is interested in exploring the relationship between school level indicators and their relationship to the achievement gap separately by SES achievement groups.
+More concretely, what is the role of school autonomy in influencing the achievement gap?
+
+## School Autonomy And The Widening Of The Achievement Gap
+
+The concept of school autonomy and it's relationship to increasing equality has been a predominant topic in policy research.
+Most rigorous studies performed on autonomy interventions document a positive increase in test scores, a decrease in school dropouts and a decrease in grade repetitions among other things [@bruns2011].
+However, most of this evidence is concentrated on small-scale pilots and interventions in particular areas of interest (low SES areas in developing countries, for example) [@gropello2006].
+
+The work of @hanushek2013 is among the first to document internationally that autonomy seems to be negative for low income countries and positive for high income countries.
+Focusing on three different types of school autonomy (curricular, personnel, budget), they find consistent evidence that curricular and personnel autonomy seems to be associated with increasing test scores in high income countries and decreasing test scores for low income countries.
+They argue that the mechanism under which autonomy can have negative effects is when decision makers are opportunistic in their behaviors but also when decision makers are not well prepared to make these decisions.
+@ammermuller2005 actually found that the more autonomy the school has, the more relevance the parent's education and cultural resources gain importance.
+This means that autonomy is associated with a somewhat negative influence on student's performance as they have to rely more on the input from resources outside of school than on their teacher's and school's resources.
+
+However, the mechanism under which @hanushek2013 suggests that autonomy can have negative impact can also be present in developed countries.
+For example, if the worst performing schools have on average lower quality of teaching, then autonomy can theoretically have negative affects as lower performing teachers could deviate from the validated national curriculum and affect the learning experience.
+Similarly, teachers could be lowering the academic standards of lower performing students reinforcing their already poor performance through lower goals.
+In particular, this last mechanism has been discussed in detail in @gamoran1987 and @hattie2002.
+Conversely, having greater academic autonomy for the good performing students can increase their performance by altering the teaching methods to increase the learning rate of these students.
+
+Most school autonomy related research focuses on between country relationships and the average country performance [@hanushek2013; @ammermuller2005; @letendre2003; @stevenson1991] without paying attention whether autonomy can have varying effects within a country at the extremes of the achievement gap.
+School autonomy as a means of school differentiation can have either positive or negative results based on the capacity of the school/teachers to make these decision and on the composition of the students at each school.
+That is why @werfhorst_mijs discuss evidence that standardization in autonomy is often associated with increasing equality in Europe.
+However, there is no clear evidence on whether autonomy is good for all \emph{within} countries.
+Moreover, the argument from @hanushek2013 is that autonomy is beneficial for high income countries without disaggregating whether it's good for some and bad for others.
+
+This study will focus on studying the relationship between academic autonomy, personnel autonomy and budget autonomy and it's relationship to the highly contested SES achievement gap.
+In particular, this paper will investigate whether different types of autonomy can have negative associations with the bottom 10% of students (the bottom group of the SES gap) and whether this same relationship is reversed for the top 10% of students (the top group of the SES gap).
+However, I focus almost exclusively on high income countries to test whether autonomy can also have negative effects within developed economies.
+
+# Empirical Strategy
+
+## Data
+
+To investigate the above mentioned questions, I will use the Programme for International Student Assessment (PISA).
+PISA is a survey carried out every three years that aims to evaluate education systems by testing the skills and knowledge of 15-year-old students.
+Currently, PISA has six waves starting in 2000 up until 2015, where recently, over half a million students were tested in mathematics, literacy and science in over 70 developed/developing countries.
+
+PISA collects data through a two-stage stratified sampling design.
+With the help of governments, PISA randomly chooses 150 schools in each country, where they then randomly pick thirty 15 year olds to undertake the two hour tests.
+Together with the subject tests, PISA collects personal information from students, their families and their school environment, that serves as relevant background information that can be matched to the students performance.
+With the recent inclusion of PISA 2015, these six waves make up a time-series analysis of 15 years, enough to visualize changes in the structure of an educational system.
+None of the studies cited so far has used the last PISA wave, which was released in December 2016.
+This chapter takes advantage of these six waves to build a country pseudo-panel, making it possible to study changes in nearly 15 years for 29 countries.
+In order to maximize country variation, I have included countries which have at least participated in 50\% of all waves.
+
+To identify a student's socio economic status I use the composite SES index created by the PISA team.
+The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI), the highest level of education of the student’s parents, the PISA index of family wealth (which measures the material wealth of the family), the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (mainly about books in the household) [@oecd_glance_2002].
+The variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming all mentioned variables into comparable metrics across waves.
+The ESCS index was derived from a principal component analysis of standardized variables, taking the factor scores for the first principal component as measures of the PISA index of economic, social and cultural status.
+All countries and economies (both OECD and partner countries/economies) were assigned the same weight in the principal component analysis, while in previous cycles, the principal component analysis was based only on OECD countries.
+However, for the purpose of reporting, the ESCS scale has been transformed with zero being the score of an average OECD student and one being the standard deviation across equally weighted OECD countries [@pisa_2015_results].
+To the best of my knowledge this is the first piece of research that uses the newly-released ESCS index [@pisa_2015_results],  which was rescaled so that all ESCS indexes are suitable for over-time analysis \footnote{These rescaled indexes can be found at \href{http://www.oecd.org/pisa/data/2015database/}{http://www.oecd.org/pisa/data/2015database/} under \emph{Rescaled indices for Trend Analyses}.}.
+In other words, the ESCS index does not need any transformation or coding updates as it is ready for comparison over time.
+
+Aside from SES, the other relevant variables are test scores for mathematics and literacy \footnote{The analysis in this paper is mainly concentrated on Mathematics to be able to compare some of the findings with the existent literature which has predominantly focused on this subject.
+Literacy is used as a second test to check if the results hold.
+PISA also tests students in Science but since very little research has been done on this subject related to achievement gaps, it was not included in the analysis}.
+PISA does not provide a single test result for each respondent.
+Instead, it provides a \emph{series} of 'plausible values' that the child could actually score.
+As explained in the PISA manual [@pisa2012_technical], these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured (the true distribution of the possible scores a student can achieve) \footnote{It should be noted that PISA has rotating modules for the main subject of that year.
+This means that the quality of data might be different for the same subject over time}.
+
+A more intuitive explanation is this: suppose we have \(\mu_i\), the average student test score in mathematics for student \(i\).
+Instead of estimating \(\mu_i\) alone, plausible values estimate a distribution of possible \(\mu\text{'s}\) for student \(i\), together with the likelihood of each \(\mu_i\) based on the respondents answers on the test.
+This is defined as the posterior distributions of \(\mu\text{'s}\) for student \(i\).
+The reason why PISA uses this procedure is because estimating a single number \(\mu_i\) is plagued with measurement error, among other types of bias [see @wu2005].
+The number of plausible values for PISA waves are usually five (although ten for PISA 2015) random draws from this distribution.
+In practice, each student has 5 scores for each test, which resembles their distribution.
+Those values are continuous, ranging from 0 to 500, with a mean of 250.
+However, PISA test scores were scaled to have a mean of 500 and a standard deviation of 100 over students in all OECD countries in the first year of focal testing (e.g.
+2000 for mathematics and reading).
+
+As per the independent variables, I will include both student level variable and school level variables.
+For the student level variables, I will include the gender of each student, their parent's level of education (to control for residual variation within each group of the SES index), the occupation index of their parents (centered at 0), whether they're native students or immigrants, the number of books in their household and the specific school programme they belong to (general, pre-vocational or vocational tracks).
+At the school level, I include the location of the school (small town, town, large town, city or large city), whether it's a public or private school, the size of the school in number of students (center with a mean of 0) and the percentage of funding that comes from government funding (centered with a mean of 0).
+In the next section I describe the definition of autonomy measures.
+
+## Methodology And Variables
+
+### Definition Of Autonomy
+
+Each school principal in PISA was asked the question of who has the main responsibility for the certain areas in the school.
+These areas are (1) autonomy over which courses are offered, (2) over the content of the courses, (3) over choosing textbooks, (4) over hiring teachers, (5) over setting their salaries and (6) over budget allocationfor the school.
+For each of these questions, the principal can answer whether this is the responsibility of the teacher's, the principal's, the school's governing board, a regional or local education authority or a national education authority.
+Similarly to @hanushek2013, I define that a school has autonomy over an area if neither a regional or local authority nor the national education authority has any responsibility in setting these areas.
+That is, the decision is taken either by a teacher, a principal, or the school's governing board.
+This definition mimics the already established definition by @hanushek2013 and tackles specifically whether the decision is solely responsibility of the school.
+Table \ref{tab:corr_aut} shows the correlation between these autonomy meaures.
+
+Given that course autonomy, content autonomy and text book autonomy are highly correlated, I also include an 'academic content autonomy' index which is just the average between these three.
+Moreover, I do the same for autonomy over hiring teacher's and setting teacher's salary into the index 'personnel autonomy'.
+Since school budget autonomy does not correlate with any other variables, I leave it as is.
+
+```{r corr_aut}
+corr_aut <- readd(calculated_corr_aut)
+
+aut_names <- c(" ",
+               "Course",
+               "Content",
+               "Textbook",
+               "Hiring",
+               "Salary",
+               "School Budget",
+               "Academic content",
+               "Personnel")
+
+names(corr_aut) <- aut_names
+corr_aut$` ` <- aut_names[-1]
+stargazer(corr_aut,
+          type = "latex",
+          summary = FALSE,
+          header = FALSE,
+          font.size = "small",
+          title = "Correlation coefficients between school autonomy measures",
+          column.sep.width = "1pt",
+          rownames = FALSE,
+          label = paste0("tab:", knitr::opts_current$get("label")),
+          no.space = TRUE)
+```
+
+Table \ref{tab:corr_aut} shows that the index of academic content is highly correlated to autonomy over which courses are offered, which content is offered and autonomy over choosing textbooks (.90, .92 and .88, respectively), which suggests that it represents the concept of academic content autonomy well.
+Similarly, personnel autonomy has correlations of .93 and .80 with autonomy over teacher hiring and teacher salary, which also validates the index of personnel autonomy.
+From now on, the main unit of analysis for the autonomy variables will be the index on academic content autonomy, personnel autonomy and budget autonomy.
+Figure \ref{fig:autonomy_evolution} plots the evolution of academic content autonomy and personnel autonomy for all countries over the 6 pisa waves.
+
+```{r autonomy_evolution, fig.asp = 1.7, fig.cap = "Evolution of school autonomy by country"}
+p1_plot <- readd(plotted_autonomies)
 p1_plot
@@
 
-# Further information
+There are indeed starking differences between countries.
+For example, the United States is experiencing a decrease in autonomy over the past 15 years whereas Denmark is witnessing an increase in autonomy in recent years.
+@hanushek2013 discusses in detail some decentralization reforms happening in some of these countries and how they match the currents trends we see in this plot.
+For example, the increase in autonomy for Germany and the decline in autonomy from the United States reflects underlying reforms to decentralize and centralize decision making respectively.
+For Germany, the increasing autonomy matches the reforms implemented in North Rhine-Westphalia which give more autonomy to schools over hiring teachers [@hanushek2013].
+Whereas in the United the decreasing trend also reflects the expansion of national level standards from the 'No Child Left Behind' reform.
+All in all, the evidence points out to  some important differences in autonomy between countries accompanied with over time dynamics in decision making.
 
-All \LaTeX enviroments supported by the main template are supported here as well; see the `.tex` sample file [here](http://onlinelibrary.wiley.com/journal/10.1002/(ISSN)1097-0258/homepage/la_tex_class_file.htm) for more details and example.
+```{r, fig.asp = 2, fig.width = 6.3, eval = FALSE}
+p2_plot <- readd(plotted_academic_aut)
+p2_plot
+```
+
+### Model
+
+The model that I will use in the analysis is a cross-classified multilevel model where I allow the country-wave intercepts to vary randomly to account for the clustering of students into each country-year combination.
+Note that the multilevel approach is just a strategy to account for for the clustering as it adjusts the standard errors appropriately.
+Yet the specification of a multilevel model is not of particular interest for it's random component as the objective is not to explain cross-country or cross-wave differences in autonomy but rather test whether the relationship between autonomy hold across a sample of countries which are repeated over time, accounting for their nested structure.
+
+The main model can be formally defined as:
+
+\begin{equation}
+y_{icy} = \alpha_{cy} + b_1 * autonomy_{cj} + b_n * x_{icj} + \epsilon_{icy}
+\end{equation}
+
+Where $y_{icy}$ is the mathematics test score for student $i$ in country $c$ and year $y$.
+$\alpha_{cy}$ is the random intercept for each country/year, $b_1$ is the beta coefficient for each autonomy measure (models are run separately) and $x_{icj}$ is the vector of covariates used as controls in all models.
+All models presented are run using the student sample weights to have representative estimates of the populations.
+
+The dependent variable of the analysis will be the standardized mathematics test scores for each student^[I also report results for literacy in the appendix].
+As mentioned before, PISA does not provide a single achievement indicator.
+Instead, I calculate the median of all plausible values for each student \footnote{Since each plausible value is a random draw from a theoretical latent normal distribution of possible student achievement scores, the median should be precise in getting a central measure of the latent distribution.}, resulting in one single score.
+
+To standardize the test scores I fit a linear model for each wave, where test scores is regressed on age measured in months (following the same strategy as @reardon2011 \footnote{This does not mess up the analysis by masking age-specific gaps as all students in the sample are 15 year olds.
+controlling for age is simply to adjust for monthly differences in ages.}) weighted by the student sample weights from PISA.
+With the residuals of this adjusted test score, I standardized the metric to solve the problem of comparability over time ^[PISA 2000 has a slightly different metric over time] for all PISA waves.
+However, another concern is whether test scores measured at different waves have different amounts of measurement error.
+If that is the case, then the amount of bias will not be the same in each measure of the gap.
+This can be misleading and suggest erroneous interpretations regarding trends of the gaps over time [@reardon2011].
+PISA has tried to make sure the tests are comparable across waves but it is still necessary to adjust for this imprecision [@pisa2012_technical].
+Accordingly, each PISA survey provides a reliability indicator for each of the tests which can be used to adjust for the reliability of the scores.
+
+In order to correct for this I calculate $\lambda_i$ which is just the standardized test score \begin{math}  \hat{\gamma_i} \end{math} adjusted by the reliability indicator of each wave.
+More formally, I calculate it through
+
+\begin{equation}
+\hat{\lambda_i} = \hat{\gamma_i} * \frac{1}{\sqrt{r}}
+\end{equation}
+
+Where \begin{math}r\end{math} is the reliability score of the test score in that PISA wave \footnote{Other procedures multiply each country by their own reliability measure for each year-subject pair [@chmielewskiglobal2019].
+With this standardized and adjusted test score I define a dummy for those in the top 10% of the SES distribution and those at the bottom 10% of the SES distribution.
+I develop this standardization and estimation proceedure more in detail in the appendix.
+All models below are run separately for the bottom and top 10% of the SES distribution to test whether the autonomy measures have different dynamics for the two groups.
+
+The model defined above is the standard model presented in the paper.
+However, I run a battery of different models to show the robustness of the results.
+Aside from the standard model, I run the same specification only on public schools, as the results of autonomy and test scores are certainly endogenous to the type of school.
+I also run a model for all available PISA countries (not only the developed subset of countries), a variant of the standard model that uses a linear model with country-year fixed effects (similarly to @hanushek2013), a model which exchanges the top and bottom 10% of students for the top and bottom 10% of schools, a model which runs all autonomy measures together (not separate models by autonomy measure, as was defined above) and a model with all students pooled and a formal interaction term between the SES groups and each autonomy measure.
+These models are presented in detail in the appendix and can be identified by their title.
+However, I also summarize their results in the results section.
+
+## Results
+
+To begin, we visualize the highly contested SES achievement gap to understand the different dynamics between and within countries.
+Figure \ref{fig:evolution_gaps} shows the evolution of the achievement gap for developed countries.
+
+```{r evolution_gaps, fig.asp = 1.7, fig.cap = "evolution of achievement gaps by countries"}
+p3_plot <- readd(p1_evolution_gaps)
+p3_plot
+```
+
+As we can see from the results, some countries have increased their achievement strongly.
+For example, France, Austria and surprisingly Sweden have very steep slopes.
+France experienced an increase in inequality by roughly 0.9 SD, Austria by 0.6 and Sweden by 0.6.
+For such a short period of time, the magnitude of these increases are reasonably big.
+
+Given that no one has estimated the evolution of the gap I cannot cross-check how other empirical estimations put France at.
+However, the work of @micklewright2006 is the closest reference available which also finds that France was a low dispersion country in 2000; there is no evidence on what happened over time.
+Fortunately, the work of @bernardi2016 did study social origin inequalities (broadly speaking, not in terms of achievement gaps) in France and found that they increased since the 2000's.
+
+Other countries have reasonable increases such as Finland and Hungary, with increases of nearly 0.6 and 0.4 standard deviations respectively.
+Aside from these countries, there are other countries which experience no changes at all, specifically, Canada, Netherlands and Spain.
+Canada excels here not only because the gap has been stable over time, but because it has the smallest gap of all countries presented here.
+It is nearly 0.5 SD in 2000 and it increased only by 0.2 in 2015.
+
+On the other hand, there are other countries which experience a decrease in the SES achievement gap.
+Poland decreased by about -0.4 and Denmark by -0.2.
+However, the most notable cases are the United States and Germany.
+These two countries show high levels of dispersion in the year 2000 with SES gaps of over 2 SD.
+But in the 15-year time trend both countries reduced the gaps by -0.6 and -0.9 respectively.
+Their distinctively large gaps in 2000 also show up in the work of @micklewright2006.
+This finding is similar to the one in @reardon2016, in which they found a decreasing gap for kindergartners.
+It is important to highlight that the cohorts in their analysis are different from the ones in this study but also reassures that evidence close to the cohorts in this study also found a decline.
+
+Analyzing figure \ref{fig:evolution_gaps} the reader may get the impression that these trends are not very steep and they should not be relevant in practical terms.
+However, note that the Y axis is measured in standard deviations.
+Small changes are actually large in practical terms.
+For example, evidence from PIRLS shows that the predicted growth of a student for a year of school is of around 0.30 standard deviations [@beaton1996].
+PISA has also documented this type of metric in their annual reports [@oecd2009_sd].
+Take the case of Sweden.
+The slope does not look that steep but in reality it increased the gap from 1 SD in 2000 to around 1.5 SD in 2015.
+With that information in mind, the trends of Poland, United States, France and Germany gain particular relevance.
+
+Table \ref{standard_bottom_math} contains the main models for the bottom 10% of all students and table \ref{standard_top_math} contains the main models for the top 10% of all students.
+Both tables have 2 models per autonomy measure corresponding to models for the index of academic autonomy, personnel autonomy and budget autonomy.
+
+\blandscape
+
+```{r standard_bottom_math, results = "asis"}
+
+# name_models <- "^aut_allcnt_math_1"
+# test <- "Mathematics"
+# student_school <- "students"
+# tit = "Multilevel model with varying intercepts for top 10\\% of students in Mathematics for all countries (not only developed)"
+
+print_models <-
+  function(name_models,
+           test,
+           student_school = "students",
+           tit = "") {
+
+    labs_covart <- c(
+      "Academic autonomy",
+      "Personnel autonomy",
+      "Budget autonomy",
+      "- Gender: Male",
+      "- Edu: Primary",
+      "- Edu: Lower sec",
+      "- Edu: Upper sec I",
+      "- Edu: Upper sec II",
+      "- Edu: University",
+      "- Books in HH: 11-100",
+      "- Books in HH: 101-500",
+      "- Books in HH: >500",
+      "- Occupation index",
+      "- Native student",
+      "- Voc track: Vocational",
+      "- Voc track: General",
+      "- Location: Town",
+      "- Location: Large town",
+      "- Location: City",
+      "- Location: Large city",
+      "- % certified teachers",
+      "- Public (ref: private)",
+      "- Size of school",
+      "- % government funding"
+    )
+
+    labs_covart <-
+      if (grepl("1", name_models)) setdiff(labs_covart, c("Edu: Primary", "Edu: Lower sec")) else labs_covart
+
+    model_names <- cached()[grepl(name_models, cached())]
+    model_names <- determine_aut_order(model_names)
+    all_mods <-
+      map(model_names, ~ readd(.x, character_only = TRUE)[2:3])
+
+    type_mod <- if (grepl("0", name_models)) "bottom" else "top"
+
+    stargazer(
+        all_mods,
+        type = "latex",
+        title = tit,
+        header = FALSE,
+        order = c(
+          "academic_content_aut",
+          "personnel_aut",
+          "budget_aut",
+          "genderMale",
+          "high_edu_broadPrimary",
+          "high_edu_broadLower secondary",
+          "high_edu_broadUpper secondary I",
+          "high_edu_broadUpper secondary II",
+          "high_edu_broadUniversity",
+          "books_hh11-100",
+          "books_hh101-500",
+          "books_hh> 500",
+          "hisei",
+          "nativeNative",
+          "ISCEDOVocational",
+          "ISCEDOGeneral"
+        ),
+        covariate.labels = labs_covart,
+        add.lines = create_lmer_stats(all_mods),
+        dep.var.caption = paste0(test, " test score"),
+        dep.var.labels = paste0("Models restricted to ", type_mod, " 10\\% of ", student_school),
+        font.size = "small",
+        column.sep.width = "1pt",
+        label = knitr::opts_current$get("label"),
+        single.row = TRUE,
+        no.space = TRUE
+      )
+
+    # student_level_pos <- which(grepl("Budget autonomy", mod_table))
+    # school_level_pos <- which(grepl("Voc track: General", mod_table))
+    # 
+    # cat(
+    #   starpolishr::star_insert_row(
+    #   mod_table,
+    #   c("Student-level variables", "School-level variables"),
+    #   insert.after = c(student_level_pos, school_level_pos)
+    # ))
+
+}
+
+print_models("^aut_math_0", "Mathematics", tit = "Multilevel model with varying intercepts for bottom 10\\% of students in Mathematics")
+
+```
+
+\elandscape
+
+Results for the first two models in table \ref{standard_bottom_math} shows that academic autonomy is associated with a decrease in test scores in Mathematics of about -0.06 standard deviations in test scores for students in the bottom 10%.
+These results hold even after controlling for the type of tracking programme that the child is enrolled in, the location of the school, the size of the school, the percentage of funding by the government, the percentage of certified teachers and whether the school is public or private.
+These results speak directly to the work of @stevenson1991 which found that autonomy was detrimental in many cases but was criticized saying that these differences in autonomy was due to within country differences in school types.
+These results show that even after adjusting for different types of schools (private/public) and type of programme (different curricular tracks), the magnitude of the relationship is still big.
+These results seem to be as strong for personnel autonomy (model 4) with a decrease of nearly -0.07 points in the standardized mathematics test score on average for each country in every year.
+
+Finally, the last two models for budget autonomy do not seem to be related to decreasing test scores, as the effect size is only -0.01 with a great deal of uncertainty.
+It is important to highlight that these effect size (-0.06 and -0.07) represent the average increase for the average year.
+If we multiply them to represent the actual time span of PISA (6 waves), it sums up to nearly 0.4 standard deviations.
+These results match very closely the results of @chmielewskiglobal2019 in the evolution of the achievement gap and are quite big considering that magnitudes of over 0.3 standard deviations reflect a gap of about one grade's worth of knowledge.
+To put it simply, greater autonomy towards low performing students seems to be associated with a decrease of .4 standard deviations in test scores over the 15 years of data available.
+
+Having said that, it can be the case that certain students self-select into schools with greater autonomy and that in itself is correlated with poorer performance.
+That is certainly playing a role in this estimation.
+However, it's not totally clear that it is the case given that within countries there is no particular evidence suggesting that public schools within a country have different levels of autonomy depending on performance.
+
+Moving on to table \ref{standard_top_math}, we can explore the results for the top 10% of students.
+The first two models show that academic autonomy seems to be positively associated with increasing student test scores by about .05 standard deviations for the top 10% of students.
+This amounts to a total of .3 standard deviations over the whole PISA waves.
+However, for the personnel and budget autonomy indexes both associations seem to be much weaker (effect sizes of about .01 for both) suggesting that academic autonomy seems to be an important autonomy component over both the good and bad performers yet autonomy over hiring teachers and setting their salaries is only detrimental in the context of bad performers.
+
+\blandscape
+
+```{r standard_top_math, results = "asis"}
+print_models("^aut_math_1", "Mathematics",
+             tit = "Multilevel model with varying intercepts for top 10\\% of students in Mathematics")
+```
+
+\elandscape
+
+Table \ref{standard_bottom_reading} and \ref{standard_top_reading} replicate the results for literacy and show similar magnitudes and associations.
+In table 3 I find that academic autonomy is associated with a decrease in test scores of about -.09 standard deviations and personnel autonomy is associated with a decrease of -.04 standard deviations for the bottom 10% of students.
+Similarly, budget autonomy is unrelated to decreasing of increasing performance.
+However, in table 4 I find exactly the same results as for reading: for the top 10% of students, academic autonomy seems to be related to increasing test scores by about 0.05 standard deviations but not for personnel autonomy.
+However, budget autonomy seems to be associated with an increase in performance of about 0.02 standard deviations.
+The magnitude of this effect size is much lower than in the other models, so it is still relatively weak.
+
+\blandscape
+
+```{r standard_bottom_reading}
+print_models("^aut_read_0",
+             "Reading",
+             tit = "Multilevel model with varying intercepts for bottom 10\\% of students in Literacy")
+```
+
+\elandscape
+
+\blandscape
+
+```{r standard_top_reading}
+print_models("^aut_read_1",
+             "Reading",
+             tit = "Multilevel model with varying intercepts for top 10\\% of students in Literacy")
+```
+
+\elandscape
+
+As robustness, I've ran the previous model under several different specifications which I've compiled in figure \ref{fig:plot_combined_models}.
+
+\newpage
+
+```{r plot_combined_models, fig.asp = 1, fig.cap = "Comparison of autonomy measures under different model variations"}
+p4_plot <- readd(all_mods_plot)
+p4_plot
+```
+
+On the right panel (red) we have all estimates related to the bottom 10% of students whereas on the left panel (blue) we have all estimates related to the top 10% of students.
+The x axis shows all different model specifications: "Public schools" refers to the same model but only restricted to public schools, "Standard model" refers to the previously shown models, "All countries" refers to models with all countries (including low income countries) and "Fixed effects" is an additional specification which uses a linear model with country-year fixed effects (to match @hanushek2013).
+The top left panel shows that the size of the effect sizes is very similar in all model specifications for academic autonomy.
+In contrast, the positive associations for the top 10% of students seem to dissapear once we add country-year fixed effects.
+In the middle panel, we see that for the personnel autonomy index, all models have negative relationships with the bottom 10% of students carrying sizable and robust associations.
+In contrast, for the top 10% of performers the relationship is quite sizable when restricted only to public schools and to all countries (increases of about 0.06-0.10 in standardized test scores) but dissapear when adding country-year fixed effects.
+Moving on to the bottom panel, results show that the effect sizes for the budget autonomy seem to be very small for both the top and bottom 10% of students.
+In summary, the evidence suggests that academic autonomy and personnel autonomy seem to have a somewhat sizable negative relationship with test scores yet this evidence is not symmetrical for the top 10% of students: the coefficients are somewhat sizable under certain specifications but the evidence is mixed.
+The models that estimated all of these changes are present in the appendix section.
+In addition, there are 4 tables concentrated not on the top/bottom 10% of students but on the top/bottom 10% of schools, to tackle directly whethere these results are similar when focusing only on the worst/best schools.
+Results are very similar.
+
+```{r, eval = FALSE}
+p5_plot <- readd(interact_plot)
+p5_plot
+```
+
+# Appendix
+
+\blandscape
+```{r, results = "asis"}
+print_models("^aut_schools_math_0", "Mathematics", "schools",
+             tit = "Multilevel model with varying intercepts for bottom 10\\% of schools in Mathematics")
+```
+\elandscape
+
+\blandscape
+```{r, results = "asis"}
+print_models("^aut_schools_read_0", "Reading", "schools",
+             tit = "Multilevel model with varying intercepts for bottom 10\\% of schools in Literacy")
+```
+\elandscape
+
+\blandscape
+```{r, results = "asis"}
+print_models("^aut_schools_math_1", "Mathematics", "schools",
+             tit = "Multilevel model with varying intercepts for top 10\\% of schools in Mathematics")
+```
+\elandscape
+
+\blandscape
+```{r, results = "asis"}
+print_models("^aut_schools_read_1", "Reading", "schools",
+             tit = "Multilevel model with varying intercepts for top 10\\% of schools in Literacy")
+```
+\elandscape
+
+\blandscape
+```{r standard_allcnt_bottom_math, results = "asis"}
+print_models("^aut_allcnt_math_0",
+             "Mathematics",
+             tit = "Multilevel model with varying intercepts for bottom 10\\% of students in Mathematics for all countries (not only developed)")
+```
+\elandscape
+
+<!-- \blandscape -->
+<!-- ```{r standard_allcnt_top_math, results = "asis"} -->
+<!-- print_models("^aut_allcnt_math_1", -->
+<!--              "Mathematics", -->
+<!--              tit = "Multilevel model with varying intercepts for top 10\\% of students in Mathematics for all countries (not only developed)" -->
+<!--              ) -->
+<!-- ``` -->
+<!-- \elandscape -->
+
+\blandscape
+```{r fixedeff_bottom_math, results = "asis"}
+
+print_models_fixedeff <- function(name_models, test, student_school = "students", tit) {
+
+    labs_covart <- c("Academic autonomy",
+                   "Personnel autonomy",
+                   "Budget autonomy",
+                   "- Gender: Male",
+                   "- Edu: Primary",
+                   "- Edu: Lower sec",
+                   "- Edu: Upper sec I",
+                   "- Edu: Upper sec II",
+                   "- Edu: University",
+                   "- Books in HH: 11-100",
+                   "- Books in HH: 101-500",
+                   "- Books in HH: >500",
+                   "- Occupation index",
+                   "- Native student",
+                   "- Voc track: Vocational",
+                   "- Voc track: General",
+                   "- Location: Town",
+                   "- Location: Large town",
+                   "- Location: City",
+                   "- Location: Large city",
+                   "- % certified teachers",
+                   "- Public (ref: private)",
+                   "- Size of school",
+                   "- % government funding")
+
+  model_names <- cached()[grepl(name_models, cached())]
+  model_names <- determine_aut_order(model_names)
+  all_mods <- map(model_names, ~ readd(.x, character_only = TRUE))
+
+  type_mod <- if (grepl("0", name_models)) "bottom" else "top"
+
+  stargazer(all_mods,
+            type = "latex",
+            title = tit,
+            header = FALSE,
+            order = c("academic_content_aut",
+                      "personnel_aut",
+                      "budget_aut",
+                      "genderMale",
+                      "high_edu_broadPrimary",
+                      "high_edu_broadLower secondary",
+                      "high_edu_broadUpper secondary I",
+                      "high_edu_broadUpper secondary II",
+                      "high_edu_broadUniversity",
+                      "books_hh11-100",
+                      "books_hh101-500",
+                      "books_hh> 500",
+                      "hisei",
+                      "nativeNative",
+                      "ISCEDOVocational",
+                      "ISCEDOGeneral"),
+            covariate.labels = labs_covart,
+            omit = "^as.factor",
+            omit.stat = "f",
+            add.lines = list(c("Country-Year fixed effects", rep("yes", length(flatten(all_mods))))),
+            dep.var.caption = paste0(test, " test score"),
+            dep.var.labels = paste0("Models restricted to ", type_mod, " 10\\% of ", student_school),
+            font.size = "small",
+            column.sep.width = "1pt",
+            label = knitr::opts_current$get("label"),
+            single.row = TRUE,
+            no.space = TRUE)
+
+  # student_level_pos <- which(grepl("Budget autonomy", mod_table))
+  #
+  # mod_table <- starpolishr::star_insert_row(mod_table, "Student-level variables", insert.after = student_level_pos)
+  #
+  # school_level_pos <- which(grepl("Voc track: General", mod_table))
+  #
+  #
+  # mod_table <- starpolishr::star_insert_row(mod_table, "School-level variables", insert.after = school_level_pos)
+  #
+  # mod_table
+
+}
+
+print_models_fixedeff("^aut_fixedeff_math_0",
+             "Mathematics",
+             tit = "Country-wave fixed effect models for bottom 10\\% of students in Mathematics")
+
+```
+\elandscape
+
+\blandscape
+```{r fixedeff_top_math, results = "asis"}
+print_models_fixedeff(
+  "^aut_fixedeff_math_1",
+  "Mathematics",
+  tit = "Country-wave fixed effect models for top 10\\% of students in Mathematics"
+  )
+```
+\elandscape
+
+\blandscape
+```{r standard_autpooled_bottom, results = "asis"}
+
+print_models_pooled <- function(name_models, test, student_school = "students", tit) {
+
+    labs_covart <- c("Academic autonomy",
+                   "Personnel autonomy",
+                   "Budget autonomy",
+                   "- Gender: Male",
+                   "- Edu: Primary",
+                   "- Edu: Lower sec",
+                   "- Edu: Upper sec I",
+                   "- Edu: Upper sec II",
+                   "- Edu: University",
+                   "- Books in HH: 11-100",
+                   "- Books in HH: 101-500",
+                   "- Books in HH: >500",
+                   "- Occupation index",
+                   "- Native student",
+                   "- Voc track: Vocational",
+                   "- Voc track: General",
+                   "- Location: Town",
+                   "- Location: Large town",
+                   "- Location: City",
+                   "- Location: Large city",
+                   "- % certified teachers",
+                   "- Public (ref: private)",
+                   "- Size of school",
+                   "- % government funding")
+
+  model_names <- cached()[grepl(name_models, cached())]
+  model_names <- determine_aut_order(model_names)
+  all_mods <- map(model_names, ~ readd(.x, character_only = TRUE)[4:5])
+
+  type_mod <- if (grepl("0", name_models)) "bottom" else "top"
+
+  labs_covart <- if (grepl("1", name_models)) setdiff(labs_covart, c("Edu: Primary", "Edu: Lower sec")) else labs_covart
+
+  stargazer(all_mods,
+            type = "latex",
+            title = tit,
+            header = FALSE,
+            order = c("academic_content_aut",
+                      "personnel_aut",
+                      "budget_aut",
+                      "genderMale",
+                      "high_edu_broadPrimary",
+                      "high_edu_broadLower secondary",
+                      "high_edu_broadUpper secondary I",
+                      "high_edu_broadUpper secondary II",
+                      "high_edu_broadUniversity",
+                      "books_hh11-100",
+                      "books_hh101-500",
+                      "books_hh> 500",
+                      "hisei",
+                      "nativeNative",
+                      "ISCEDOVocational",
+                      "ISCEDOGeneral"),
+            covariate.labels = labs_covart,
+            add.lines = create_lmer_stats(all_mods),
+            dep.var.caption = paste0(test, " test score"),
+            dep.var.labels = paste0("Models restricted to ", type_mod, " 10\\% of ", student_school),
+            font.size = "small",
+            column.sep.width = "1pt",
+            label = knitr::opts_current$get("label"),
+            single.row = TRUE,
+            no.space = TRUE)
+
+  #   student_level_pos <- which(grepl("Budget autonomy", mod_table))
+  #
+  # mod_table <- starpolishr::star_insert_row(mod_table, "Student-level variables", insert.after = student_level_pos)
+  #
+  # school_level_pos <- which(grepl("Voc track: General", mod_table))
+  #
+  #
+  # mod_table <- starpolishr::star_insert_row(mod_table, "School-level variables", insert.after = school_level_pos)
+  #
+  # mod_table
+
+}
+
+print_models_pooled("aut_allpooled_math_0",
+                    "Mathematics",
+                    tit = "Multilevel model with varying intercepts for bottom 10\\% of students in Mathematics with all autonomy measures pooled")
+
+```
+\elandscape
+
+\blandscape
+```{r standard_autpooled_top, results = "asis"}
+print_models_pooled("aut_allpooled_math_1",
+                    "Mathematics",
+                    tit = "Multilevel model with varying intercepts for top 10\\% of students in Mathematics with all autonomy measures pooled")
+```
+\elandscape
+
+\blandscape
+```{r, results = "asis", eval = FALSE}
+
+print_models_interaction <- function(name_models, test, student_school = "students", tit) {
+
+  labs_covart <- c("Academic autonomy",
+                   "Academic aut * Mid students",
+                   "Academic aut * High students",
+                   "Personnel autonomy",
+                   "Personnel aut * Mid students",
+                   "Personnel aut * High students",
+                   "Budget autonomy",
+                   "Budget aut * Mid students",
+                   "Budget aut * High students",
+                   "- ESCS group: Mid students",
+                   "- ESCS group: High students",
+                   "- Gender: Male",
+                   "- Edu: Primary",
+                   "- Edu: Lower sec",
+                   "- Edu: Upper sec I",
+                   "- Edu: Upper sec II",
+                   "- Edu: University",
+                   "- Books in HH: 11-100",
+                   "- Books in HH: 101-500",
+                   "- Books in HH: >500",
+                   "- Occupation index",
+                   "- Native student",
+                   "- Voc track: Vocational",
+                   "- Voc track: General",
+                   "- Location: Town",
+                   "- Location: Large town",
+                   "- Location: City",
+                   "- Location: Large city",
+                   "- % certified teachers",
+                   "- Public (ref: private)",
+                   "- Size of school",
+                   "- % government funding")
+
+  model_names <- cached()[grepl(name_models, cached())]
+  model_names <- determine_aut_order(model_names)
+  all_mods <- map(model_names, ~ readd(.x, character_only = TRUE))
+
+  type_mod <- if (grepl("0", name_models)) "bottom" else "top"
+
+  stargazer(all_mods,
+            type = "latex",
+            title = tit,
+            header = FALSE,
+            order = c("academic_content_aut",
+                      "academic_content_aut:quantiles_escs_chrMid",
+                      "academic_content_aut:quantiles_escs_chrHigh",
+                      "personnel_aut",
+                      "personnel_content_aut:quantiles_escs_chrMid",
+                      "personnel_content_aut:quantiles_escs_chrHigh",
+                      "budget_aut",
+                      "budget_content_aut:quantiles_escs_chrMid",
+                      "budget_content_aut:quantiles_escs_chrHigh",
+                      "quantiles_escs_chrMid",
+                      "quantiles_escs_chrHigh",
+                      "genderMale",
+                      "high_edu_broadPrimary",
+                      "high_edu_broadLower secondary",
+                      "high_edu_broadUpper secondary I",
+                      "high_edu_broadUpper secondary II",
+                      "high_edu_broadUniversity",
+                      "books_hh11-100",
+                      "books_hh101-500",
+                      "books_hh> 500",
+                      "hisei",
+                      "nativeNative",
+                      "ISCEDOVocational",
+                      "ISCEDOGeneral"),
+            covariate.labels = labs_covart,
+            add.lines = create_lmer_stats(all_mods),
+            dep.var.caption = paste0(test, " test score"),
+            dep.var.labels = "Models applied to all students",
+            font.size = "small",
+            column.sep.width = "1pt",
+            label = knitr::opts_current$get("label"),
+            single.row = TRUE,
+            no.space = TRUE)
+
+  # student_level_pos <- which(grepl("Budget aut \\* High students", mod_table))
+  #
+  # mod_table <- starpolishr::star_insert_row(mod_table, "Student-level variables", insert.after = student_level_pos)
+  #
+  # school_level_pos <- which(grepl("Voc track: General", mod_table))
+  #
+  #
+  # mod_table <- starpolishr::star_insert_row(mod_table, "School-level variables", insert.after = school_level_pos)
+  #
+  # mod_table
+
+}
+
+print_models_interaction("aut_interact_math_.+",
+                         "Mathematics",
+                         tit = "Multilevel model with varying intercepts for all students of students in Mathematics with interaction between student quantiles and autonomy measures")
+
+```
+\elandscape
+
+## Calculating Achievement Gaps {#achievement_appendix}
+
+To \emph{standardize} the test score I fit a linear model
+
+\begin{equation}
+y_i = \alpha + \beta_1 * AGE_i + \epsilon_i, \quad \epsilon_i ~ N(0, \sigma^2)
+\end{equation}
+
+for each wave, where \begin{math}y_i\end{math} is the median student test score for student \begin{math}i\end{math} and \begin{math}AGE_i\end{math} is their age measured in months (following the same strategy as @reardon2011 \footnote{This does not mess up the analysis by masking age-specific gaps as all students in the sample are 15 year olds.
+Controlling for age is simply to adjust for monthly differences in ages.}) weighted by the student sample weights from PISA \footnote{I also tried to run the model for each country-wave separately but the results were very similar and it was more computationally expensive}.
+
+I then calculate \begin{math} \hat{\gamma_i} \end{math} by
+
+\begin{equation}
+\hat{\gamma_i} = \frac{\hat{\epsilon_i}}{\sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y_i})^2}}
+\end{equation}
+
+where \begin{math}\hat{\epsilon_i}\end{math} is the residual for student \begin{math}i\end{math}, $\hat{y_i}$ is the predicted test score for student $i$ and the denominator is the root mean square error of the model.
+
+This new standardized variable has a mean of zero.
+Standardizing the median test score solves the problem of comparability between different tests and across waves as the test scores have now the same metric across time.
+Another concern is whether test scores measured at different waves have different amounts of measurement error.
+If that is the case, then the amount of bias will not be the same in each measure of the gap.
+This can be misleading and suggest erroneous interpretations regarding trends of the gaps over time [@reardon2011].
+PISA has tried to make sure the tests are comparable across waves but it is still necessary to adjust for this imprecision [@pisa2012_technical].
+Accordingly, each PISA survey provides a reliability indicator for each of the tests which can be used to adjust for the reliability of the scores.
+
+In order to correct for this I calculate $\lambda_i$ which is just \begin{math} \hat{\gamma_i} \end{math} adjusted by the reliability indicator of each wave. More formally, I calculate it through
+
+\begin{equation} \label{eq:adjust_eq}
+\hat{\lambda_i} = \hat{\gamma_i} * \frac{1}{\sqrt{r}}
+\end{equation}
+
+where \begin{math}r\end{math} is the reliability score of the test score in that PISA wave ^[Other procedures multiply each country by their own reliability measure for each year-subject pair [@chmielewskiglobal2019].
+The reliability estimates are calculated using Item Response Theory (IRT) analogues of traditional estimates of person separation reliability such as internal consistency.
+Unfortunately, PISA 2000 did not provide any reliability measure separately for each country and at the moment of the writing of this paper, PISA 2015 has yet to release their own.
+For these reasons, I implement the analysis following the original work of [@reardon2011]].
+Note that I implement equation $\eqref{eq:adjust_eq}$ separately by test scores and waves because there is a separate reliability indicator for each one.
+Once that is adjusted, the test scores should be roughly free of any bias in the trend that may arise from differential reliability of the tests.
+
+In order to calculate the SES gaps it is necessary to estimate the thresholds for the 90th and 10th percentile.
+I calculate the thresholds using the SES index separately for each country-wave combination using the specific student sample weights of each one.
+I then generate a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile for each country-wave pair.
+
+I then fit a multilevel model:
+
+\begin{equation}
+\lambda_{ij} = \alpha_{j} + \beta_{j} * SES_i + \epsilon_{ij},\ \text{for} \ i \ \text{= 1, 2, ...,} \ n \ \text{for each country} \ j
+\end{equation}
+
+where \(SES_i\) is whether the student is at or above the 90th percentile (coded as 1) or whether it is at or below the 10th percentile (coded as 0).
+I allow \(\alpha\) and \(\beta\) to vary by country \(j\) in order to obtain gaps for each country.
+I implement this model separately for each wave and weight by the wave-specific student sample weights.
+The previous model allows to calculate the achievement gap for each country by extracting the \(\beta\)'s and \(\alpha\)'s for each country.
+I also calculate the standard error of this difference and generate uncertainty intervals.
+
+# Bibliography

### 2020-02-04 _drake.R

@@ -2,9 +2,9 @@
 # If you use supporting scripts like the ones below,
 # you will need to supply them yourself. Examples:
 # https://github.com/wlandau/drake-examples/tree/master/main/R
-source("code/01-packages.R")  # Load your packages, e.g. library(drake).
-source("code/02-analysis.R") # Define your custom code as a bunch of functions.
-source("code/plan.R")      # Create your drake plan.
+source(here::here("code/01-packages.R"))  # Load your packages, e.g. library(drake).
+source(here::here("code/02-analysis.R")) # Define your custom code as a bunch of functions.
+source(here::here("code/plan.R"))      # Create your drake plan.
 
 # options(clustermq.scheduler = "multicore")
 # _drake.R must end with a call to drake_config().

### 2020-02-04 code/01-packages.R

@@ -1,6 +1,9 @@
 library(car)
 library(drake)
+library(plm)
 library(rmarkdown)
+library(ggpubr)
+library(ggsci)
 library(here)
 library(xtable)
 library(patchwork) # devtools::install_github("thomasp85/patchwork")
@@ -25,3 +28,4 @@ library(fst)
 library(insight)
 library(Amelia)
 library(stargazer)
+library(emmeans)

### 2020-02-04 code/02-analysis.R

@@ -19,7 +19,6 @@ load_school <- function(raw_data_dir) {
 
   colnames(school2012) <- dic_school$varname
 
-
   # PISA 2015
   pisa2015_path <- file.path(raw_data_dir, "pisa2015", "CY6_MS_CMB_SCH_QQQ.sav")
   pisa_countrynames <- c(cimentadaj::pisa_countrynames, "United States" = "USA")
@@ -76,8 +75,17 @@ harmonize_student <- function(raw_student, recode_cntrys) {
            native = as.numeric(ST16Q01),
            native = ifelse(native == 1, "Native", ifelse(native == 2, "Non-native", NA_character_)),
            native = factor(native, levels = c("Non-native", "Native")),
-           stu_id = as.character(STIDSTD)
+           stu_id = as.character(STIDSTD),
+           ISCEDO = as.character(ST25Q01),
+           ISCEDO = case_when(ISCEDO == "<ISCED 2A>" ~ "General",
+                              ISCEDO == "<ISCED 2B>" ~ "Pre-Vocational",
+                              ISCEDO == "<ISCED 2C>" ~ "Pre-Vocational",
+                              ISCEDO == "<ISCED 3A>" ~ "General",
+                              ISCEDO == "<ISCED 3B>" ~ "Vocational",
+                              ISCEDO == "<ISCED 3C>" ~ "Vocational",
+                              TRUE ~ ISCEDO)
            )
+
 
   # PISA 2003
   raw_student$value[[2]] <-
@@ -94,7 +102,8 @@ harmonize_student <- function(raw_student, recode_cntrys) {
            native = as.numeric(ST15Q01),
            native = ifelse(native == 1, "Native", ifelse(native == 2, "Non-native", NA_character_)),
            native = factor(native, levels = c("Non-native", "Native")),
-           stu_id = as.character(STIDSTD)
+           stu_id = as.character(STIDSTD),
+           ISCEDO = as.character(ISCEDO)
            )
 
   # PISA 2006
@@ -112,7 +121,8 @@ harmonize_student <- function(raw_student, recode_cntrys) {
            native = as.numeric(ST11Q01),
            native = ifelse(native == 1, "Native", ifelse(native == 2, "Non-native", NA_character_)),
            native = factor(native, levels = c("Non-native", "Native")),
-           stu_id = as.character(STIDSTD)
+           stu_id = as.character(STIDSTD),
[…]

### 2020-02-04 code/plan.R

@@ -228,6 +228,46 @@ print_table <- function(x, ...) {
 }
 
 
+determine_aut_order <- function(x) {
+  academic_pos <- which(grepl("academic_content", x))
+  personnel_pos <- which(grepl("personnel_", x))
+  budget_pos <- setdiff(seq_along(x), c(academic_pos, personnel_pos))
+  correct_order <- c(academic_pos, personnel_pos, budget_pos)
+
+  x[correct_order]
+}
+
+
+create_lmer_stats <- function(tst_mods) {
+  tst_mods <- flatten(tst_mods)
+
+  all_groups <- lapply(tst_mods, function(.x) {
+    res <- ngrps(.x)
+    res
+  })
+
+  ngrps_mixed <- lapply(transpose(all_groups), as.numeric)
+
+  ngrps_final <- lapply(seq_along(ngrps_mixed), function(.x) {
+    name <- paste0("N. ", tools::toTitleCase(names(ngrps_mixed)[.x]))
+    ngrps_mixed[[.x]] <- c(name, ngrps_mixed[[.x]])
+    ngrps_mixed[[.x]]
+  })
+
+
+  all_iccs <- vapply(tst_mods,
+                     function(.x) round(performance::icc(.x)$ICC_adjusted, 4),
+                     FUN.VALUE = numeric(1)
+                     )
+
+  all_iccs <- list(c("ICC", all_iccs))
+
+  all_stats <- c(ngrps_final, all_iccs)
+
+  all_stats
+}
+
+
 formula_sequence <- function(dv, ivs, random_effect) {
   formula_seq <-
     lapply(1:length(ivs), function(x) {
@@ -608,6 +648,152 @@ sig_vs_sig <- function(coef1, se1, coef2, se2) {
   round(c(coef_diff, se_diff), 2)
 }
 
+model_converter <- function(name_models, model_type) {
+  model_names <- cached()[grepl(name_models, cached())]
+
+  # Because for the fixed effect models we only run 2 models
+  # but for all multilevel models we run 3 (1st is empty model
+  # to record variance). -->
+  if (grepl("fixed", name_models)) index <- 2 else index <- 3
+  raw_math <-
+    set_names(
+      map(model_names, ~ readd(.x, character_only = TRUE)[[index]]),
+      model_names
+    )
+
+  lower <- raw_math[grepl("0", names(raw_math))]
+  upper <- raw_math[grepl("1", names(raw_math))]
+  lower <-
[…]
